### 2018-09-20 archive/accessPEP.Rmd
Initial commit

---
title: "PEP Access"
author: "Malavika Rajeev"
date: "6/25/2018"
output: html_document
---
```{r directory, include=FALSE}
rm(list = ls())
require("knitr")
opts_knit$set(root.dir = "/Users/mrajeev/Dropbox/MadaPEP")
```
## Data

```{r setup, include=FALSE}
## libraries
# for gis 
library(knitr)
library(raster)
library(maptools)
library(maps)
library(GISTools)
library(rgdal)
library(sp)
library(rgdal)
library(gdistance)
# everything else
library(lubridate)
library(reshape2)
library(tidyverse)
library(rjags)
library(coda)
library(lattice)
# pkgs <- c("lattice", "coda")
# require(pkgs, character.only = TRUE)

## gis data {NEED TO FIGURE OUT PROJECTIONS!}
p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
mada.district <- readOGR("data/MadaGIS/MadaPops.shp", p4s = p4s)
ctar.gps <- read.csv("data/ctar_gps.csv")
ctar.gps <- SpatialPoints(cbind(ctar.gps$LONGITUDE, ctar.gps$LATITUDE),
                          proj4string = crs(mada.district))
# ## getting travel time layer
mada.access <- raster("output/study.area.accessibility.tif")
mada.access <- crop(mada.access, mada.district)
plot(mada.access/60, breaks = c(0, 1, 2, 3, 4, 5, 6, 8, 16, max(values(mada.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
     #col = colorRampPalette(c("navyblue", "mediumturquoise", "orange", "red"))(11), axes = FALSE, box = FALSE)
points(ctar.gps, pch = 18, col = "grey", cex = 1)

## getting which districts have a ctar
mada.district$ctar.in <- ifelse(is.na(match(mada.district$district, ctar.gps$District)), 0, 1)
mada.district$ctar.in[mada.district$district %in% c('Manakara Atsimo', 'Toliary-I', 'Taolagnaro',
                                                    'Nosy-Be', "Ambovombe-Androy", "Arivonimamo")] <- 1
mada.district$ctar.in[mada.district$district %in% c('Miarinarivo')] <- 0
## should equal 31
sum(mada.district$ctar.in)
## check to see if it worked
plot(mada.district, col = ifelse(mada.district$ctar.in == 1, "grey", "white"))
points(ctar.gps, col = "blue")

## getting districts which are "covered" by a ctar
mada.district$ctar.close <- poly.counts(ctar.gps, mada.district)
mada.district$ctar.close[mada.district$district %in% c("Antsiranana II", "Antsiranana I", "Mahajanga II", 
                                                    "Mahajanga I", "Maintirano", "Toliary-II", "Toliary-I",
                                                    "Taolagnaro", "Isandra", "Ambohimahasoa", 
                                                    "Lalangina", "Vohibato", "Mankara", "Antsirabe II",
                                                    "Antsirabe I", "Fianarantsoa I", "Fianarantsoa II",
                                                    "Toamasina I", "Toamasina II", "Manakara Atsimo")] <- 1
mada.district$ctar.close[mada.district$reg_code == 11] <- 1
# ## check to see if it worked
# plot(mada.district, col = ifelse(mada.district$ctar.close == 1, "grey", "white"))
# points(ctar.gps, col = "blue")

## mid.pops
mada.district$midpops <- (mada.district$pop2015adj + mada.district$pop2020adj)/2

## ctar data: getting exposure matrix (district by year)
ctar.data <- read.csv("data/SaisieRage_DATA_2018-06-28_0956.csv")
ctar.data %>% 
  filter(type != "contact") %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(mdy(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures
## heatmap by year
ggplot(exposures, aes(x = year, y = district, fill = n)) +
  geom_tile()

## merge district shapefile with the summarized matrix!
exposure.mat <- dcast(exposures, district ~ year)
colnames(exposure.mat) <- c("district", "year2013", "year2014", "year2015", "year2016", "year2017", "year2018")
mada.district@data <- merge(mada.district@data, exposure.mat, by.x = "mdg_dis_co", by.y = "district", all.x = TRUE)
vars <- c("mdg_dis_co", "district", "midpops", "study_area", "ctar.in",
         "ctar.close", "year2014", "year2015", "year2016", "year2017")

mada.district@data %>%
  select(!!vars) %>%
  mutate_at(vars(starts_with("year")), funs(replace(., is.na(.), 0))) %>%
  mutate(avg.bites100k = (select(., starts_with("year")) %>% rowSums())/4/(midpops/1e5)) -> dist.data
dist.data %>% 
  filter(district)
## subset out Analamanga (IPM) + Haute Matsiatra (Fianarantsoa)

ctar.data %>%
  group_by(id_ctar) %>%
  summarize(n = n()) -> ctar.counts

## plotting average bite incidence
pal <- colorRampPalette(c('white','blue'))
mada.district$col <- pal(10)[as.numeric(cut(mada.district$avg.bites100k, 
                                            c(0, 1, 10, 50, 100, 150, 200, 250, 300, 350)))]

plot(mada.district, col = mada.district$col)
points(ctar.gps, col = "grey 50", pch = 20)

## ggplot
gg.mada <- fortify(mada.district, region = "mdg_dis_co")
madamerge <- merge (gg.mada, mada.district@data, by.x = "id", by.y = "mdg_dis_co")
madamerge <- madamerge[order(madamerge$order),]
theme_map <- function(...) {
  theme_minimal() +
  theme(
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    # panel.grid.minor = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.major = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f5f5f2", color = NA), 
    panel.background = element_rect(fill = "#f5f5f2", color = NA), 
    legend.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.border = element_blank(),
    ...
  )
}
ctar.coords <- as.data.frame(coordinates(ctar.gps))
madamerge$avg.bites100k[madamerge$avg.bites100k == 0] <- NA
ggplot () + geom_polygon (data = madamerge, aes(x=long, y=lat, group=group, fill = avg.bites100k), color="grey50", size=0.25) + 
  coord_equal() +
  geom_point(data = ctar.coords, aes(x = coords.x1, y = coords.x2), col = "darkred") +
  scale_fill_gradient(low ="white", high="red", space = "Lab", na.value = "grey") + 
  ggtitle ("Map of total visits by district") +
  theme_map()
```

## Rationale
Access to PEP is geographically limited in Madagascar, with PEP available in only 31 clinics across the country. Therefore the burden of human rabies deaths is likely heterogeneous spatially. We are  limited by the fact that the only available data is on reported animal bites (i.e. there are no data available on confirmed/probable rabies exposures or on unreported bites of any kind other than from one district over a period of 16 months, Rajeev et al. 2018).

Standard decision tree models would estimate rabies exposure incidence (*R*) as:
*R = (B x p_rabid)/p_report*
and the burden of human rabies deaths *(D)* as:
*D = R x (1 - p_report) x p_death*
where *B* is the total reported bites, *p_rabid* is the probability that a reported animal bite is a genuine rabies exposure, *p_report* is the probability that a rabies exposure is reported, and *p_death* is the probability of death given a rabies exposure.

Applying this framework to our data on reported bites at the district level, we see that estimated rabies incidence far exceeds expected incidence in areas with high incidence of reported bites and is lower than expected for areas with little data/low incidence of reported bites (Figure 1).

Aside for Jess: this is the same issue if we fit bites as a function of travel times + pop (lower reported bites in )
```{r standard decision tree }
## adding in expected exposures annually
# rabies incidence per 100kdogs/# of human exps per rabid dog/human:dog ratio = inc per 100k persons
exps.per100k.high <- 1000*0.39/5 
exps.per100k.low <- 600*0.39/25

p_rabid <- c(0.05, 0.1, 0.2, 0.4, 0.8)
p_report<- c(0.95, 0.8, 0.4, 0.2, 0.1)
cols <- colorRampPalette(c("pink", "darkred"))(5)

ordered.bites <- dist.data$avg.bites100k[order(dist.data$study_area, decreasing = TRUE)]

plot(ordered.bites*0.6/0.85, 1:nrow(dist.data), bty = "l", pch = 20, col = "darkred")
points(ordered.bites*0.2/0.85, 1:nrow(dist.data),
       pch = 20, col = "darkred")
segments(y0 = 1:nrow(dist.data), x0 = ordered.bites*0.2/0.85,
         y1 = 1:nrow(dist.data),
         x1 = ordered.bites*0.6/0.85, col = "darkred")
abline(v = c(exps.per100k.high, exps.per100k.low), col = c("red", "red"), lty = 2)
```

## Empirical estimates of p_report for each location across a range of scenarios
We can calculate p_report empirically assuming incidence and p_rabid is uniform:
*p_report = (B x p_rabid)/R*

We constrain p_rabid so that it p_report cannot exceed 1 (i.e. *p_rabid_max = R/B* which would mean that all rabies exposures are reported). In general, as p_rabid gets higher within it's constrained range, we get higher estimates of reporting (Figure 2).

```{r empirical p_reports}
p_rabid_high <- exps.per100k.high/dist.data$avg.bites100k
p_rabid_high[p_rabid_high >= 1] <- 1
p_rabid_low <- exps.per100k.low/dist.data$avg.bites100k
p_rabid_low[p_rabid_low >= 1] <- 1

empirical.reporting.fixed <- function (bites, p_rabid = 0.2, expected) {
    p_rabid_max = expected/bites
    p_rabid_max[p_rabid_max >= p_rabid] <- p_rabid
    p_report <- bites*p_rabid_max/expected
    return(p_report)
}

plot( 1:length(p_rabid_high), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.05, 
                          expected = exps.per100k.low), bty = "l", col = "red", pch = 20, ylim = c(0, 1))
points(1:length(p_rabid_high), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.low), col = "red", pch = 20)
segments(x0 = 1:length(p_rabid_high), y0 = empirical.reporting.fixed(bites =ordered.bites, 
               p_rabid = 0.05, expected = exps.per100k.low), x1 = 1:length(p_rabid_high),
         y1 = empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.low), col = "darkred")

plot( 1:length(p_rabid_high), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.05, 
                          expected = exps.per100k.high), bty = "l", col = "red", pch = 20, ylim = c(0, 1))
points(1:length(p_rabid_high), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.high), col = "red", pch = 20)
segments(x0 = 1:length(p_rabid_high), y0 = empirical.reporting.fixed(bites =ordered.bites, 
               p_rabid = 0.05, expected = exps.per100k.high), x1 = 1:length(p_rabid_high),
         y1 = empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.high), col = "darkred")

plot(dist.data$study_area, empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.05, 
                          expected = exps.per100k.low), bty = "l", col = "red", pch = 20, ylim = c(0, 1))
```

We generate high and low estimates of expected rabies incidence (*R*) using estimates from a recent study that estimated incidence of human exposures per 100,000 dogs in a no-vaccination scenario and given high and low estimates of human:dog ratios.

## Models of reporting by travel time to clinic
To generate more probable estimates of rabies burden given what we know about access to PEP in Maadagascar, we use assumptions of rabies exposure incidence and our data on reported bites to estimate *p_report*:
*B = (R x p_report) + A*

We then estimate a district/commune specific reporting rate (*p_report*) as a function of travel times to the nearest ARMC and the background incidence of non-rabid reported bites (*A*) as a function of population size.

```{r Annual jags model with p_rabid + p_report estimated}
## model
tt.model <- "model {
  # reported bites
    for (j in 1:nlocs){
        # Number of bites
        rabid[j] ~ dpois(expected.mean[j])
        reported[j] ~ dbinom(rho[j], rabid[j])
        bites[j] ~ dpois(reported[j] + A[j])
    }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)
    pop_eff[g] ~ dnorm(0, tau_pop)

    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]

    # Background avg in place g
    log(A[g]) <- B_pop*pop[g] + pop_eff[g]
  }

  tau_loc <- 1/sd
  tau_pop <- 1/sd.pop

  # priors
  B_tt ~ dnorm(0, 0.1)
  sd ~ dunif(0, 100)
  sd.pop ~ dunif(0, 100)
  B_pop ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
}"
```

```{r Data and estimation}
pop <- dist.data$midpops
bites <- round(dist.data$avg.bites100k*dist.data$midpops/1e5)
ttimes <- dist.data$study_area
nlocs = length(bites)

data <- list (bites = bites, pop = pop, 
              expected.mean = exps.per100k.high*dist.data$midpops/1e5,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 0.1, sd.pop = 0.1,
              B_pop = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_pop", "sd.pop")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
par(mar = c(2, 2, 2, 2))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 2), mar = c(5, 5, 2,  2))
plot(pop, rpois(length(bites), exp(check$statistics[,"Mean"]["B_pop"]*pop))
      + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd.pop"]), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])),
     bty = "l")
```

```{r Moramanga commune data}
exposure.mat <- read.csv("data/exposures_by_commune_Moramanga.csv", header = TRUE, row.names = 1)
moramanga <- readOGR("data/MoramangaGIS/MoramangaPops.shp")
ctar.gps <- read.csv("data/ctar_gps.csv")
monthly.bites <-rowSums(exposure.mat)/ncol(exposure.mat)
moramanga$commune.midpops <- (moramanga$pop2015adj + moramanga$pop2020adj)/2
moramanga$monthly.bites <- monthly.bites[match(moramanga$commune, names(monthly.bites))]
moramanga$monthly.incidence <- moramanga$monthly.bites/moramanga$commune.midpops*1e5
# distance!

## travel time layer
travel.times <- raster("output/study.area.accessibility.tif")
mora.access <- mask(travel.times, moramanga)
mora.access <- crop(mora.access, moramanga)
moramanga <- raster::extract(mora.access, moramanga, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
moramanga$comm.long <- coordinates(moramanga)[,1]
moramanga$comm.lat <- coordinates(moramanga)[,2]

# plot it
plot(mora.access/60,
     breaks = c(0, 1, 2, 3, 4, 5, 6, 7,8, 16, max(values(mora.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
             "#ae017e", "#7a0177", "#49006a"), axes = FALSE, box = FALSE)
points(moramanga$comm.long[moramanga$commune == "Moramanga"], moramanga$comm.lat[moramanga$commune == "Moramanga"], pch = 18, col = "grey", cex = 2)

## correlation checks (need to add in distance and proportion of pop within x km/hrs)
cor(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")])
splom(log(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")]), type=c('p','r'))

## data
mora.bites <- merge(moramanga@data, exposure.mat, by.x = "commune", by.y = "row.names")
bites <- round(mora.bites$monthly.bites*12)
ttimes <- mora.bites$study.area.accessibility/60
pop <- mora.bites$commune.midpops
nlocs <- length(bites)

## empirical @ commune level
ordered.bites <- bites[order(ttimes)]
plot(1:length(ordered.bites), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.05, 
                          expected = exps.per100k.low), bty = "l", col = "red", pch = 20, ylim = c(0, 1))
points(1:length(ordered.bites), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.low), col = "red", pch = 20)
segments(x0 = 1:length(ordered.bites), y0 = empirical.reporting.fixed(bites =ordered.bites, 
               p_rabid = 0.05, expected = exps.per100k.low), x1 = 1:length(ordered.bites),
         y1 = empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.low), col = "darkred")

plot( 1:length(ordered.bites), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.05, 
                          expected = exps.per100k.high), bty = "l", col = "red", pch = 20, ylim = c(0, 1))
points(1:length(ordered.bites), empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.high), col = "red", pch = 20)
segments(x0 = 1:length(ordered.bites), y0 = empirical.reporting.fixed(bites =ordered.bites, 
               p_rabid = 0.05, expected = exps.per100k.high), x1 = 1:length(ordered.bites),
         y1 = empirical.reporting.fixed(bites =ordered.bites, p_rabid = 0.95, 
                          expected = exps.per100k.high), col = "darkred")

data <- list (bites = bites, pop = pop, 
              expected.mean = exps.per100k.low*pop/1e5,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1, sd.pop = 1,
              B_pop = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_pop", "sd.pop")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
par(mar = c(2, 2, 2, 2))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 2), mar = c(5, 5, 2,  2))
plot(pop, rpois(length(bites), exp(check$statistics[,"Mean"]["B_pop"]*pop))
      + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd.pop"]), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])),
     bty = "l")
```

More  remote areas are also the areas with the lowest population size, which makes it difficult to distinguish decreases in the incidence of reported bites due to population size (i.e. if incidence increases with population size or alternatively if the dog population increases with population size) vs. access--**need some figure to show this!**. To generate the most conservative estimates of reporting and burden (i.e. highest reporting overall), we estimate a scaling factor for incidence with population size so that incidence in any given place cannot exceed or be less than our maximum and minimum estimates of rabies exposure incidence.

If this doesn't work then: max and min estimates of A for each R (low, high, low scaling, high scaling) and estimate p_report as function of travel time...& estimate p_report empirically for each place this way!

### 1. p_report only
#### Convergence and model diagnostics
#### Estimates
#### Predictive power

### 2. p_report + A
#### Convergence and model diagnostics
#### Estimates
#### Predictive power

### 3. p_report + A + scaling of R
#### Convergence and model diagnostics
#### Estimates
#### Predictive power

### Other factors + next steps
#### Setting stronger priors (scaling factors for reporting and background bites)
#### Clinic effect
#### Accounting for underreporters @ clinic level
#### To commune level
#### A better metric of travel time (i.e. proportion of population living w/in x hours)

## Estimates of burden from model outputs
### 1. Status quo
**Need to account for missing data and best case scenario of dose usage**
### 2. Access expanded to all districts
### 3. Access expanded to all CSB IIs

### 2018-09-20 R/extractTravelTimes.R

@@ -1,0 +1,15 @@
+## Getting district mean travel time to 
+# libraries for gis
+library(raster)
+library(maptools)
+library(maps)
+library(GISTools)
+library(rgdal)
+library(sp)
+library(rgdal)
+library(gdistance)
+ttime.layer <- raster("output/study.area.accessibility.tif")
+mada.district <- readOGR("data/MadaGIS/MadaPops.shp")
+
+mada.out <- raster::extract(ttime.layer, mada.district, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
+writeSpatialShape(mada.out, "data/MadaGIS/MadaPops")

### 2018-09-20 mapData.R

@@ -1,0 +1,105 @@
+## Clean-up
+setwd("~/Dropbox/MadaRabies/MadaPEP")
+rm(list = ls())
+
+## Packages
+# for gis 
+library(knitr)
+library(raster)
+library(maptools)
+library(maps)
+library(GISTools)
+library(rgdal)
+library(sp)
+library(rgdal)
+library(gdistance)
+# everything else
+library(lubridate)
+library(reshape2)
+library(tidyverse)
+library(rjags)
+library(coda)
+library(lattice)
+# pkgs <- c("lattice", "coda")
+# require(pkgs, character.only = TRUE)
+
+## gis data {NEED TO FIGURE OUT PROJECTIONS!}
+p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
+mada.district <- readOGR("data/MadaGIS/MadaPops.shp", p4s = p4s)
+ctar.gps <- read.csv("data/ctar_gps.csv")
+ctar.gps <- SpatialPoints(cbind(ctar.gps$LONGITUDE, ctar.gps$LATITUDE),
+                          proj4string = crs(mada.district))
+ctar.atts <- readOGR("data/MadaGIS/31CTAR.shp")
+# ## getting travel time layer
+mada.access <- raster("output/study.area.accessibility.tif")
+mada.access <- crop(mada.access, mada.district)
+plot(mada.access/60, breaks = c(0, 1, 2, 3, 4, 5, 6, 8, 16, max(values(mada.access)/60, na.rm = TRUE)),
+     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
+#col = colorRampPalette(c("navyblue", "mediumturquoise", "orange", "red"))(11), axes = FALSE, box = FALSE)
+points(ctar.gps, pch = 18, col = "grey", cex = 1)
+
+## getting which districts have a ctar
+mada.district$ctar.in <- ifelse(is.na(match(mada.district$district, ctar.atts$District)), 0, 1)
+mada.district$ctar.in[mada.district$district %in% c('Manakara Atsimo', 'Toliary-I', 'Taolagnaro',
+                                                    'Nosy-Be', "Ambovombe-Androy")] <- 1
+
+## should equal 31
+sum(mada.district$ctar.in)
+
+## mid.pops
[…]

### 2018-09-20 accessPEP4.Rmd
Initial commit

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: "Malavika Rajeev"
date: "July 18th, 2018"
output:
  word_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
css: style.css
---

```{r directory, echo=FALSE, message = FALSE, results = 'hide'}
rm(list = ls())
require("knitr")
opts_knit$set(root.dir = "/Users/mrajeev/Dropbox/MadaPEP")
knitr::opts_chunk$set(fig.pos = 'h')
```
#### Canine rabies
* Annually, an estimated 60,000 people die from canine rabies.
* These deaths are preventable if people receive prompt post-exposure prophylaxis
* In places where rabies is endemic, there are significant barriers to accessing PEP (i.e. socioeconomics, availability)
* In Madagascar, rabies is endemic, however surveillance is poor and little data is available. There are 31 clinics, or CTAR (centre de traitment antirabique), that provide post-exposure prophylaxis in the country (Figure 1).

### Preliminary data from Madagascar Clinics
* Commune level average travel time to the closest CTAR (generated from global travel time layer from Malaria Atlas Project).
* Human population per commune extracted from World Pop
* 16 months of data on reported animal bites from the Moramanga District in Madagascar 
* In the next month, we will have 4 years of data on reported animal bites from 20 of the 31 CTAR.

```{r setup, echo = FALSE, message = FALSE, results = 'hide'}
## Clean-up
## Packages
# for gis 
library(knitr)
library(raster)
library(maptools)
library(maps)
library(GISTools)
library(rgdal)
library(sp)
library(rgdal)
library(gdistance)
# everything else
library(lubridate)
library(reshape2)
library(tidyverse)
library(rjags)
library(coda)
library(lattice)
# pkgs <- c("lattice", "coda")
# require(pkgs, character.only = TRUE)

## gis data {NEED TO FIGURE OUT PROJECTIONS!}
p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
mada.district <- readOGR("data/MadaGIS/MadaPops.shp", p4s = p4s)
ctar.gps <- read.csv("data/ctar_gps.csv")
ctar.gps <- SpatialPoints(cbind(ctar.gps$LONGITUDE, ctar.gps$LATITUDE),
                          proj4string = crs(mada.district))
ctar.atts <- readOGR("data/MadaGIS/31CTAR.shp")
# ## getting travel time layer
mada.access <- raster("output/study.area.accessibility.tif")
mada.access <- mask(mada.access, mada.district)
plot(mada.access/60, breaks = c(0, 1, 2, 3, 4, 5, 6, 8, 16, max(values(mada.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
#col = colorRampPalette(c("navyblue", "mediumturquoise", "orange", "red"))(11), axes = FALSE, box = FALSE)
points(ctar.gps, pch = 18, col = "grey", cex = 1)

## getting which districts have a ctar
mada.district$ctar.in <- ifelse(is.na(match(mada.district$district, ctar.atts$District)), 0, 1)
mada.district$ctar.in[mada.district$district %in% c('Manakara Atsimo', 'Toliary-I', 'Taolagnaro',
                                                    'Nosy-Be', "Ambovombe-Androy")] <- 1

## should equal 31
sum(mada.district$ctar.in)

## mid.pops
mada.district$midpops <- (mada.district$pop2015adj + mada.district$pop2020adj)/2

## ctar data: getting exposure matrix (district by year)
ctar.data <- read.csv("data/SaisieRage_DATA_2018-06-28_0956.csv")
ctar.data %>% 
  filter(type != "contact") %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(mdy(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

## merge district shapefile with the summarized matrix!
exposure.mat <- dcast(exposures, district ~ year)
colnames(exposure.mat) <- c("district", "year2013", "year2014", "year2015", "year2016", "year2017", "year2018")
mada.district@data <- merge(mada.district@data, exposure.mat, by.x = "mdg_dis_co", by.y = "district", all.x = TRUE)
vars <- c("mdg_dis_co", "district", "midpops", "study_area", "ctar.in",
          "year2014", "year2015", "year2016", "year2017")

## Getting all dist data
mada.district@data %>%
  select(!!vars) %>%
  mutate_at(vars(starts_with("year")), funs(replace(., is.na(.), 0))) %>%
  mutate(avg.bites100k = (select(., starts_with("year")) %>% rowSums())/4/(midpops/1e5),
         dogs_min = round(midpops/25), dogs_max = round(midpops/5)) -> dist.data

## districts that are non-reporter and data not available yet CTAR
no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
             "Marolambo", "Sainte-Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano", 
             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
dist.data %>%
  filter(!(district %in% no_data)) -> dist.data
nrow(dist.data)

```

#### Rationale
Standard decision tree models using overall reported bites would estimate rabies exposures ($R$) as:
$$ R = \frac{B\times p_{rabid}}{\rho } $$

where $B$ is the total reported bites, $p_{rabid}$ is the probability that a reported animal bite is a genuine rabies exposure, $\rho$ is the probability that a rabies exposure is reported.

We generate high and low estimates of expected rabies incidence ($R$) using outputs from a recent modeling study that estimated incidence of human exposures per 100,000 dogs in a no-vaccination scenario and given high and low estimates of human:dog ratios. We can do this in Madagascar because there's no vacc going on!

Applying this framework to our data on reported bites in the Madagascar, we see that estimated rabies incidence exceeds expected incidence in areas with high incidence of reported bites and is lower than expected for areas with little data/low incidence of reported bites (Figure 1).

```{r standard decision tree, echo=FALSE, message = FALSE, results = 'hide'}
## adding in expected exposures annually
# rabies incidence per 100kdogs/# of human exps per rabid dog/human:dog ratio = inc per 100k persons
exps.per100k.high <- 1000*0.39/5 
exps.per100k.low <- 600*0.39/25

ordered.bites <- dist.data$avg.bites100k[order(dist.data$study_area)]
par(mar = c(4, 10, 4, 4))
plot(ordered.bites*0.6/0.85, 1:length(ordered.bites), bty = "l", pch = 20, col = "darkred", 
     xlab = "Rabies exposures per 100k persons", ylab = "", axes = FALSE, 
     xlim = c(0, max(ordered.bites*0.6/0.85)+100))
axis(1)
text(dist.data$district[order(dist.data$study_area)], y = 1:length(ordered.bites), x = -10, adj = 1, xpd = TRUE)
points(ordered.bites*0.2/0.85, 1:length(ordered.bites),
       pch = 20, col = "darkred")
segments(y0 = 1:length(ordered.bites), x0 = ordered.bites*0.2/0.85,
         y1 = 1:length(ordered.bites),
         x1 = ordered.bites*0.6/0.85, col = "darkred")
abline(v = c(exps.per100k.high, exps.per100k.low), col = c("red", "red"), lty = 2)
legend("topleft", "Range of expected \n rabies incidence", col = "red", lty = 2, bty = "n",
       inset = c(0.4, 0), xpd = TRUE, x.intersp = 0.2, seg.len = 2)

## Empirical p_report
p_rabid_high <- exps.per100k.high/(dist.data$avg.bites100k)
p_rabid_high[p_rabid_high >= 0.95] <- 0.95
p_rabid_low <- exps.per100k.low/(dist.data$avg.bites100k)
p_rabid_low[p_rabid_low >= 0.95] <- 0.95

plot(dist.data$study_area/60, 
     dist.data$avg.bites100k*p_rabid_high/exps.per100k.high, ylim = c(0, 1), 
     col = "red", pch = 20, bty = "l", xlab = "Travel times (hrs)", ylab = "Reporting")
points(dist.data$study_area/60, 
     dist.data$avg.bites100k*p_rabid_low/exps.per100k.low,  
     col = "darkred")
segments(x0 = dist.data$study_area/60, y0 = dist.data$avg.bites100k*p_rabid_low/exps.per100k.low,
         x1 = dist.data$study_area/60,
         y1 = dist.data$avg.bites100k*p_rabid_high/exps.per100k.high, col = "darkred")
legend("topright", c("High incidence", "Low incidence"), col = c("Red", "darkred"), pch = c(20, 1), 
       bty = "n", inset = c(-0.2, 0), xpd = TRUE)
```

```{r district data}
bites <- dist.data[,6:9]
ttimes <- dist.data$study_area
pop <- dist.data$midpops
times <- 1:4
nlocs <- nrow(dist.data)
exps.high <- exps.per100k.high*dist.data$midpops/1e5
exps.low <- exps.per100k.low*dist.data$midpops/1e5
burn.in = 5000
```

#### Empirical estimates of p_report for each location across a range of scenarios
We can calculate p_report empirically assuming incidence and p_rabid is uniform:
$$ \rho  = \frac{B\times p_{rabid}}{R}$$

We constrain $p_{rabid}$ so that $\rho$ cannot exceed 1:
    $$
           p_{rabid}=
            \begin{cases}
            1, & \text{if}\ \frac{R}{B} > 1 \\
            \frac{R}{B}, & \text{otherwise}
            \end{cases}
    $$
which would give us the highest estimates of reporting for a given incidence level (assumes that the maximum # of bites are reported the $\frac{R}{B} > 1$ and that reporting = 1 when $\frac{R}{B} < 1$). We see that generally reporting decreases with travel time, although this relationship will depend on our assumptions of incidence and $p_{rabid}$ (Figure 2).

#### Models of reporting by travel time to clinic
To generate more probable estimates of rabies burden given what we know about access to PEP in Maadagascar, we estimate a district specific reporting rate as a function of travel time ($T$) to the nearest ARMC:

$$ \rho_j = logistic(B_{tt}T_j+ B_{0})$$
We can specify our expected bites in a few ways:

##### Given an expected rabies exposures incidence and corresponding $p_{rabid}$:

$$ B_j \sim Poisson(\frac{\rho_j R}{p_{rabid_j}})$$
$$
           p_{rabid_j}=
            \begin{cases}
            x, & \text{if}\ \frac{R_j}{B_j} > x \\
            \frac{R_j}{B_j}, & \text{otherwise}
            \end{cases}
$$

We estimate the potential relationship between travel time and reporting ($B_{tt}$ and $B_{0}$) given high and low estimates of incidence and high, mid, and low proportion rabid ($x$ = 0.25, 0.5, or 0.75).

```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
## model
tt.model.1 <- "model {
for (j in 1:nlocs){
  for (t in 1:times){
    # Number of bites
    bites[j, t] ~ dpois(rho[j]*expected.mean[j]/p_rabid[j, t])
  }
}

# spatial effects
  for (g in 1:nlocs){
  # Reporting in place g
  logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
}

# priors
  B_tt ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
}"

est.p.1 <- function (bites, ttimes, times, nlocs, expected.mean, p_rabid.est){
  p_rabid <- expected.mean/bites
  p_rabid[p_rabid >= p_rabid.est] <- p_rabid.est

  data <- list (bites = bites, times = times,
              expected.mean = expected.mean, p_rabid = p_rabid,
              ttimes = ttimes, nlocs = nlocs)

  inits <- list(B_tt = rnorm(1, 0, 1e-6), 
              B_0 = rnorm(1, 0, 1e-6))

  jags_mod <- jags.model(textConnection(tt.model.1), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

  params <- c("B_tt", "B_0")
  samps <- coda.samples(jags_mod, params, n.iter = 10000)
  return(samps)
}
check <- est.p.1(bites = bites, times = times, ttimes = ttimes, nlocs = nlocs,
                        expected.mean = exps.low, p_rabid.est = 0.5)
```

##### Given an expected rabies exposures incidence and non-rabid bites:

$$ B_j \sim Poisson(\rho_j (R_j + N_j))$$
where $N$ is a function of human population size (more people = more people getting bitten by dogs?):
$$ N_j \sim f(pop_j)$$
This assumes that reporting is the same for non-rabid and rabid bites, but that the incidence of non-rabid bites is not uniform across space (as with rabies incidence) and increases with increasing population size.

```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
## model
tt.model.2 <- "model {
# reported bites
for (j in 1:nlocs){
  for (t in 1:times){
    # Number of bites
    bites[j, t] ~ dpois(rho[j]*(expected.mean[j] + N[j]))
  }
}

# spatial effects
  for (g in 1:nlocs){
    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
    # Non-rabid incidence in place g
    log(N[g]) <- B_pop*pop[g] + B_1
  }

# priors
  B_tt ~ dnorm(0, 0.1)
  B_pop ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
  B_1 ~ dnorm(0, 0.1)
}"

est.p.2 <- function (bites, ttimes, times, pop, nlocs, expected.mean){
  data <- list (bites = bites, times = times, pop = pop,
              expected.mean = expected.mean, p_rabid = p_rabid,
              ttimes = ttimes, nlocs = nlocs)

  inits <- list(B_tt = rnorm(1, 0, 1e-6), 
              B_0 = rnorm(1, 0, 1e-6), B_pop = rnorm(1, 0, 1e-6), 
              B_1 = rnorm(1, 0, 1e-6))

  jags_mod <- jags.model(textConnection(tt.model.2), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

  params <- c("B_tt", "B_0", "B_1", "B_pop")
  samps <- coda.samples(jags_mod, params, n.iter = 10000)
  return(samps)
}
```

##### Given an expected rabies exposures incidence and reported non-rabid bites:

$$ B_j \sim Poisson(\rho_jR_j + N_j)$$
where $N_j$ is a function of travel time (closer you are to a clinic, higher incidence of reported non-rabid bites):
$$ N_j \sim f(T_j)$$
This assumes that reporting may not be the same for non-rabid and rabid bites, but that the incidence of non-rabid bites increases with travel time, just as reporting increases with travel time. This may be true, but we will likely run into identifiability issues (i.e. estimating two params which increase with travel time, reporting and reported non-rabid incidence).

```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
## model
tt.model.3 <- "model {
# reported bites
for (j in 1:nlocs){
  for (t in 1:times){
    # Number of bites
    bites[j, t] ~ dpois(rho[j]*expected.mean[j] + N[j])
  }
}

# spatial effects
  for (g in 1:nlocs){
    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
    # Non-rabid incidence in place g
    log(N[g]) <- B_nr*(1/ttimes[g]) + B_1
  }

# priors
  B_tt ~ dnorm(0, 0.1)
  B_nr ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
  B_1 ~ dnorm(0, 0.1)
}"

est.p.3 <- function (bites, ttimes, times, nlocs, expected.mean){
  data <- list (bites = bites, times = times,
              expected.mean = expected.mean,
              ttimes = ttimes, nlocs = nlocs)

  inits <- list(B_tt = rnorm(1, 0, 1e-6), 
              B_0 = rnorm(1, 0, 1e-6), B_nr = rnorm(1, 0, 1e-6), 
              B_1 = rnorm(1, 0, 1e-6))

  jags_mod <- jags.model(textConnection(tt.model.3), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

  params <- c("B_tt", "B_0", "B_1", "B_nr")
  samps <- coda.samples(jags_mod, params, n.iter = 10000)
  return(samps)
}
```

```{r Annual jags model with p_rabid + p_report estimated, high incidence, echo = FALSE, message = FALSE, results = 'hide'}
## models assuming high incidence

plot(0, 0, xlim = c(1, 10), ylim = c(0, 1), bty = "l", xlab = "Travel times (hrs)", ylab = "Reporting")
ttimes.plot <- seq(0.01, 12, by = 0.2)
lowinc.prab25 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs,
                        expected.mean = exp.low, p_rabid.est = 0.25)
check <- summary(window(lowinc.prab25, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "slateblue1")
lowinc.prab5 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs, 
                       expected.mean = exp.low, p_rabid.est = 0.5)
check <- summary(window(lowinc.prab5, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "slateblue2")
lowinc.prab75 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs, 
                        expected.mean = exp.low, p_rabid.est = 0.75)
check <- summary(window(lowinc.prab75, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "slateblue3")
highinc.prab25 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs, 
                         expected.mean = exp.high, p_rabid.est = 0.25)
check <- summary(window(highinc.prab25, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "firebrick1")
highinc.prab5 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs, 
                        expected.mean = exp.high, p_rabid.est = 0.5)
check <- summary(window(highinc.prab5, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "firebrick2")
highinc.prab75 <- est.p(bites = bites, ttimes = ttimes, nlocs = nlocs, 
                         expected.mean = exp.high, p_rabid.est = 0.75)
check <- summary(window(highinc.prab75, start = burn.in))
points(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "firebrick2")
legend("topright", c("Low incidence", "High incidence"), col = c("Blue", "Red"), lty = 1, 
       bty = "n", inset = c(0.25, 0.2), xpd = TRUE)

## parameter estimates
p_rabid <- exp.low/bites
p_rabid[p_rabid >= 0.25] <- 0.25
par(mfrow = c(1, 1), mar = c(5, 5, 2,  2))
## mean and confidence interval
sims = 1000
rho <- plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)
                    + check$statistics[,"Mean"]["B_0"])
bites.sim.high <- matrix(0, nrow = sims, ncol = length(bites))
for (i in 1:sims){
  bites.sim.high[i, ] <- rpois(length(p_rabid), exp.low*rho/p_rabid)
}
rankCI <- function(x){
  upper <- x[order(x)][0.95*length(x)]
  lower <- x[order(x)][0.05*length(x)]
  return(c(lower, upper))
}

upper.CI.high <- apply(bites.sim.high, 2, rankCI)[2,]
lower.CI.high <- apply(bites.sim.high, 2, rankCI)[1,]

plot(bites, pch = 1, col = "red", bty = "l", xlab = "w/ Low estimated incidence",
     ylab = "Total bites", axes = FALSE)
axis(2)
points(colMeans(bites.sim.high), col = "darkred", pch = 20)
segments(x0 = 1:length(bites), y0 = lower.CI.high, x1 = 1:length(bites), 
         y1 = upper.CI.high, col = "darkred")

```

#### Estimating 
```{r Annual jags model, echo = FALSE, message = FALSE, results = 'hide'}
bites = round(moramanga$monthly.bites*12)
ttimes = moramanga$study.area.accessibility/60
pop = moramanga$commune.midpops
nlocs = length(moramanga$commune)
exp.low = exps.per100k.low*moramanga$commune.midpops/1e5
exp.high = exps.per100k.high*moramanga$commune.midpops/1e5
burn.in = 5000

## model
tt.model <- "model {
# reported bites
for (j in 1:nlocs){
# Number of bites
  all[j] ~ dpois(expected.mean[j] + A[j])
  bites[j] ~ dbinom(rho[j], all[j])
}

# spatial effects
  for (g in 1:nlocs){
  # Reporting in place g
  logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
  A[g] <- non_rabid*pop[g]/1e5
}

# priors
  B_tt ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
  non_rabid ~ dunif(10, 1000)
}"

data <- list (bites = bites, 
              expected.mean = exp.high, pop = pop,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 1, 1), B_0 = rnorm(1, 1, 1), 
              non_rabid = rnorm(1, 700, 1))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "B_0", "non_rabid")
samps <- coda.samples(jags_mod, params, n.iter = 10000)
plot(samps)
check <- summary(window(samps, start = burn.in))
check
plot(ttimes.plot, plogis(check$statistics[,"Mean"]["B_tt"]*(1/(ttimes.plot))
                  + check$statistics[,"Mean"]["B_0"]), type = "l", col = "slateblue1")

## mean and confidence interval
sims = 1000
rho <- plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)
                    + check$statistics[,"Mean"]["B_0"])
bites.sim.high <- matrix(0, nrow = sims, ncol = length(bites))
for (i in 1:sims){
  all <- rpois(length(bites), check$statistics[,"Mean"]["non_rabid"]*pop/1e5 + exp.low)
  bites.sim.high[i, ] <- rbinom(length(bites), all, rho)
}
rankCI <- function(x){
  upper <- x[order(x)][0.95*length(x)]
  lower <- x[order(x)][0.05*length(x)]
  return(c(lower, upper))
}

upper.CI.high <- apply(bites.sim.high, 2, rankCI)[2,]
lower.CI.high <- apply(bites.sim.high, 2, rankCI)[1,]

plot(bites, pch = 1, col = "red", bty = "l", xlab = "w/ Low estimated incidence",
     ylab = "Total bites", axes = FALSE)
axis(2)
points(colMeans(bites.sim.high), col = "darkred", pch = 20)
segments(x0 = 1:length(bites), y0 = lower.CI.high, x1 = 1:length(bites), 
         y1 = upper.CI.high, col = "darkred")

```

### 2018-09-20 archive/accessPEP2.Rmd
Initial commit

---
title: "PEP Access"
author: "Malavika Rajeev"
date: "6/25/2018"
output: html_document
---
```{r directory, include=FALSE}
rm(list = ls())
require("knitr")
opts_knit$set(root.dir = "/Users/mrajeev/Dropbox/MadaPEP")
```

## Data [Show figure from paper]
* Commune level average travel time to the closest CTAR (where PEP is available).
* Human population per commune extracted from World Pop
* 16 months of data on reported animal bites from the Moramanga District in Madagascar 
  - Use to calculate the annual # of reported bites per 100,000 persons.

```{r setup, include=FALSE}
## libraries
# for gis 
library(knitr)
library(raster)
library(maptools)
library(maps)
library(GISTools)
library(rgdal)
library(sp)
library(rgdal)
library(gdistance)
# everything else
library(lubridate)
library(reshape2)
library(tidyverse)
library(rjags)
library(coda)
library(lattice)
# pkgs <- c("lattice", "coda")
# require(pkgs, character.only = TRUE)

## gis data {NEED TO FIGURE OUT PROJECTIONS!}
p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

## data
exposure.mat <- read.csv("data/exposures_by_commune_Moramanga.csv", header = TRUE, row.names = 1)
moramanga <- readOGR("data/MoramangaGIS/MoramangaPops.shp")
ctar.gps <- read.csv("data/ctar_gps.csv")

## data
monthly.bites <-rowSums(exposure.mat)/ncol(exposure.mat)
moramanga$commune.midpops <- (moramanga$pop2015adj + moramanga$pop2020adj)/2
moramanga$monthly.bites <- monthly.bites[match(moramanga$commune, names(monthly.bites))]
moramanga$monthly.incidence <- moramanga$monthly.bites/moramanga$commune.midpops*1e5

## travel time layer
travel.times <- raster("output/study.area.accessibility.tif")
mora.access <- mask(travel.times, moramanga)
mora.access <- crop(mora.access, moramanga)
moramanga <- raster::extract(mora.access, moramanga, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
moramanga$comm.long <- coordinates(moramanga)[,1]
moramanga$comm.lat <- coordinates(moramanga)[,2]

# plot it
plot(mora.access/60,
     breaks = c(0, 1, 2, 3, 4, 5, 6, 7,8, 16, max(values(mora.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
             "#ae017e", "#7a0177", "#49006a"), axes = FALSE, box = FALSE)
points(moramanga$comm.long[moramanga$commune == "Moramanga"], moramanga$comm.lat[moramanga$commune == "Moramanga"], pch = 18, col = "grey", cex = 2)

```

## Rationale
Access to PEP is geographically limited in Madagascar, with PEP available in only 31 clinics across the country. Therefore the burden of human rabies deaths is likely heterogeneous spatially.

Standard decision tree models using overall reported bites would estimate rabies exposure incidence ($R$) as:
$$ R = \frac{B\times p_{rabid}}{p_{report}} $$

where *B* is the total reported bites, *p_rabid* is the probability that a reported animal bite is a genuine rabies exposure, *p_report* is the probability that a rabies exposure is reported.

We generate high and low estimates of expected rabies incidence ($R$) using estimates from a recent modeling study that generated incidence of human exposures per 100,000 dogs in a no-vaccination scenario and given high and low estimates of human:dog ratios. We can do this in Madagascar because there's no vacc going on!

Applying this framework to our data on reported bites in the Moramanga District, we see that estimated rabies incidence exceeds expected incidence in areas with high incidence of reported bites and is lower than expected for areas with little data/low incidence of reported bites (Figure 1).

Aside for Jess: this is the same issue if we fit bites as a function of travel times + pop (lower reported bites in )
```{r standard decision tree }
## adding in expected exposures annually
# rabies incidence per 100kdogs/# of human exps per rabid dog/human:dog ratio = inc per 100k persons
exps.per100k.high <- 1000*0.39/5 
exps.per100k.low <- 600*0.39/25

ordered.bites <- moramanga$monthly.incidence[order(moramanga$commune.midpops)]*12
par(mar = c(4, 10, 2, 2))
plot(ordered.bites*0.6/0.85, 1:length(monthly.bites), bty = "l", pch = 20, col = "darkred", 
     xlab = "Rabies exposures per 100k persons", ylab = "", axes = FALSE, 
     xlim = c(0, max(ordered.bites*0.6/0.85)+100))
axis(1)
text(moramanga$commune[order(moramanga$commune.midpops)], y = 1:length(monthly.bites), x = -10, adj = 1, xpd = TRUE)
points(ordered.bites*0.2/0.85, 1:length(monthly.bites),
       pch = 20, col = "darkred")
segments(y0 = 1:length(monthly.bites), x0 = ordered.bites*0.2/0.85,
         y1 = 1:length(monthly.bites),
         x1 = ordered.bites*0.6/0.85, col = "darkred")
abline(v = c(exps.per100k.high, exps.per100k.low), col = c("red", "red"), lty = 2)
legend("topleft", "Range of expected \n rabies incidence", col = "red", lty = 2, bty = "n",
       inset = c(0.4, 0), xpd = TRUE, x.intersp = 0.2, seg.len = 1)
```

## Empirical estimates of p_report for each location across a range of scenarios
We can calculate p_report empirically assuming incidence and p_rabid is uniform:
$$ p_{report} = \frac{B\times p_{rabid}}{R}$$

We constrain $p_{rabid}$ so that p_report cannot exceed 1:
    $$
           p_{rabid}=
            \begin{cases}
            1, & \text{if}\ \frac{R}{B} > 1 \\
            \frac{R}{B}, & \text{otherwise}
            \end{cases}
    $$
which would give us the highest estimates of reporting for a given incidence level (assumes that the maximum # of bites are reported the $\frac{R}{B} > 1$ and that reporting = 1 when $\frac{R}{B} < 1$).

```{r empirical p_reports}
p_rabid_high <- exps.per100k.high/(moramanga$monthly.incidence*12)
p_rabid_high[p_rabid_high >= 0.95] <- 0.95
p_rabid_low <- exps.per100k.low/(moramanga$monthly.incidence*12)
p_rabid_low[p_rabid_low >= 0.95] <- 0.95

## p_report
plot(moramanga$study.area.accessibility/60, 
     moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high, ylim = c(0, 1), 
     col = "red", pch = 20, bty = "l", xlab = "Travel times (hrs)", ylab = "Reporting")
points(moramanga$study.area.accessibility/60, 
     moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low,  
     col = "darkred")
segments(x0 = moramanga$study.area.accessibility/60, y0 = moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low,
         x1 = moramanga$study.area.accessibility/60,
         y1 = moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high, col = "darkred")
legend("topright", c("High incidence", "Low incidence"), col = c("Red", "darkred"), pch = c(20, 1), 
       bty = "n", inset = c(-0.2, 0), xpd = TRUE)
```

## Models of reporting by travel time to clinic
To generate more probable estimates of rabies burden given what we know about access to PEP in Maadagascar, we use assumptions of rabies exposure incidence and our data on reported bites to estimate $p_{report}$ based on our original decision tree:
$B = \frac{R\times p_{report}}{p_{rabid}}$

We  estimate a commune specific reporting rate as a function of travel times to the nearest ARMC and we say that $p_{rabid} scales inversely with the number of reported rabid bites (so that as more rabid bites are reported, more normal bites per rabid bites are reported as well). We say tha

```{r Annual jags model with p_rabid + p_report estimated, high incidence}
### data
bites <- round(moramanga$monthly.bites*12)
ttimes <- moramanga$study.area.accessibility/60
pop <- moramanga$commune.midpops
nlocs <- length(bites)
expected.mean <- exps.per100k.high/2*moramanga$commune.midpops/1e5

## model
tt.model <- "model {
# reported bites
for (j in 1:nlocs){
# Number of bites
  bites[j] ~ dpois(expected.mean[j]*rho[j]/p_rabid[j])
}

# spatial effects
  for (g in 1:nlocs){
  # Spatial random effect
  loc_eff[g] ~ dnorm(0, tau_loc)

  # Reporting in place g
  logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]

  # Background avg in place g
  p_rabid[g] <- 1/(1 + sqrt(expected.mean[g]*rho[g])*B_norm)
}

tau_loc <- 1/sd

# priors
B_tt ~ dnorm(0, 0.1)
sd ~ dnorm(0, 0.1)
B_norm ~ dnorm(0.1, 0.1)
}"

data <- list (bites = bites, 
              expected.mean = expected.mean,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1,
              B_norm = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_norm")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
par(mar = c(2, 2, 2, 2))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 1), mar = c(5, 5, 2,  2))
plot(expected.mean, 1/(1 + sqrt(expected.mean)*check$statistics[,"Mean"]["B_norm"]), bty = "l",
     pch = 20, xlab = "Expected incidence", ylab = "P_rabid")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])),
     bty = "l", pch = 20, xlab = "Travel times (hrs)", ylab = "P_report")

## mean and confidence interval
sims = 1000
bites.sim.high <- matrix(0, nrow = sims, ncol = length(bites))
for (i in 1:sims){
  rho <- plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"]))
  p_rabid <- 1/(1 + sqrt(expected.mean*rho)*check$statistics[,"Mean"]["B_norm"])
  bites.sim.high[i, ] <- rpois(length(p_rabid), expected.mean*rho/p_rabid)
}
rankCI <- function(x){
  upper <- x[order(x)][0.95*length(x)]
  lower <- x[order(x)][0.05*length(x)]
  return(c(lower, upper))
}

upper.CI.high <- apply(bites.sim.high, 2, rankCI)[2,]
lower.CI.high <- apply(bites.sim.high, 2, rankCI)[1,]
```

```{r Annual jags model with p_rabid + p_report estimated, low incidence}
### data
expected.mean <- exps.per100k.low/2*moramanga$commune.midpops/1e5

data <- list (bites = bites, 
              expected.mean = expected.mean,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1,
              B_norm = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_norm")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
par(mar = c(2, 2, 2, 2))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 1), mar = c(5, 5, 2,  2))
plot(expected.mean, 1/(1 + sqrt(expected.mean)*check$statistics[,"Mean"]["B_norm"]), bty = "l",
     pch = 20, xlab = "Expected incidence", ylab = "P_rabid")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])),
     bty = "l", pch = 20, xlab = "Travel times (hrs)", ylab = "P_report")

## mean and confidence interval
sims = 1000
bites.sim.low <- matrix(0, nrow = sims, ncol = length(bites))
for (i in 1:sims){
  rho <- plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"]))
  p_rabid <- 1/(1 + sqrt(expected.mean*rho)*check$statistics[,"Mean"]["B_norm"])
  bites.sim.low[i, ] <- rpois(length(p_rabid), expected.mean*rho/p_rabid)
}
rankCI <- function(x){
  upper <- x[order(x)][0.95*length(x)]
  lower <- x[order(x)][0.05*length(x)]
  return(c(lower, upper))
}

upper.CI.low <- apply(bites.sim.low, 2, rankCI)[2,]
lower.CI.low <- apply(bites.sim.low, 2, rankCI)[1,]

par(mfrow = c(1, 2))
plot(bites, pch = 1, col = "red", bty = "l", xlab = "w/ Low estimated incidence",
     ylab = "Total bites", axes = FALSE)
axis(2)
points(colMeans(bites.sim.low), col = "darkred", pch = 20)
segments(x0 = 1:length(bites), y0 = lower.CI.low, x1 = 1:length(bites), 
         y1 = upper.CI.low, col = "darkred")

plot(bites, pch = 1, col = "red", bty = "l", xlab = "w/ High estimated incidence",
     ylab = "Total bites",axes = FALSE)
axis(2)
points(colMeans(bites.sim.high), col = "red", pch = 20)
segments(x0 = 1:length(bites), y0 = lower.CI.high, x1 = 1:length(bites), 
         y1 = upper.CI.high, col = "red")
```

## Right now: generating the right functional forms of p_report and p_rabid

```{r getting the right functional forms}

## functional form of p_rabid
par(mfrow = c(1,1))
plot(moramanga$monthly.incidence*12, p_rabid_low, bty = "l", pch = 20, ylim = c(0,1), col = "darkred")
points(moramanga$monthly.incidence*12, p_rabid_high, pch = 20, col = "red")

points(moramanga$monthly.incidence*12, 
       1/(1 + sqrt(exps.per100k.low*moramanga$commune.midpops/1e5)*
                     (moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low)*10), col = "darkred")

points(moramanga$monthly.incidence*12, 
       1/(1 + sqrt(exps.per100k.high*moramanga$commune.midpops/1e5)*
                     (moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high)*10), col = "red")

par(mfrow = c(1,1))
plot(plogis(seq(-10, 10, by = 1)), p_rabid_low, bty = "l", pch = 20, ylim = c(0,1), col = "darkred")
points(moramanga$monthly.incidence*12, p_rabid_high, pch = 20, col = "red")

points(moramanga$monthly.incidence*12, 
       1/(1 + sqrt(exps.per100k.low*moramanga$commune.midpops/1e5)*
                     (moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low)*10), col = "darkred")

points(moramanga$monthly.incidence*12, 
       1/(1 + sqrt(exps.per100k.high*moramanga$commune.midpops/1e5)*
                     (moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high)*10), col = "red")
```

*Next step would be to clean the data from Madagascar and see if this works with the bigger data set*

*Question to be able to answer*:
What would happen if we thought incidence scaled with population size in this range? Then generally
we would see more conservative estimates of p_report overall.

## GAM

### 2018-09-20 archive/accessPEP_archive.Rmd
Initial commit

---
title: "PEP Access"
author: "Malavika Rajeev"
date: "6/25/2018"
output: html_document
---
## Background
Access to PEP is geographically limited in Madagascar, with PEP available in only 31 clinics across the country. Therefore the burden of human rabies deaths is likely concentrated in the areas with the least access. However, these areas are also the areas with the lowest population size, which makes it difficult to distinguish decreases in the incidence of reported bites due to population size (i.e. if incidence increases with population size or alternatively if the dog population increases with population size) vs. access. We are further limited by the fact that the only available data is on reported animal bites (i.e. there are no data available on confirmed/probable rabies exposures or on unreported bites of any kind other than from one district over a period of 16 months, Rajeev et al. 2018)

Here we attempt to disentagle these factors by estimating the relationship between travel time to a clinic and reporting rates based on assumptions of dog rabies incidence and background levels of non-rabid bites.

We constrain incidence of human exposures based on estimates of human exposures per 100,000 dogs in a no-vaccination scenario generated by Hampsone et al. 2018. We then set upper and lower bounds on the human:dog ratio (between 5 people per dog to 25 people per dog) to get the maximum and minimum expected annual bites per 100,000 people for each commune.

We then fit a model to our reported bite data (annual reported bites by commune) where bite incidence is predicted by population size. Then we estimate the reporting rate as a function of travel time.

Then the bites we osberve
bites[j,t] = expected.incidence*reporting + background.bites
expected.incidence = f(pop)
reporting = f(ttime)

So that we get the most conservative estimates of the relationship between reporting and travel times (i.e. taking out the maximum observed impact of pop on incidence)

```{r setup, include=FALSE}
## libraries 
# for gis 
library(raster)
library(maptools)
library(maps)
library(GISTools)
library(rgdal)
library(sp)
library(rgdal)
library(gdistance)
# everything else
library(tidyverse)
library(rjags)
library(coda)
library(lattice)
# pkgs <- c("lattice", "coda")
# require(pkgs, character.only = TRUE)

## data
exposure.mat <- read.csv("data/exposures_by_commune_Moramanga.csv", header = TRUE, row.names = 1)
moramanga <- readOGR("data/MoramangaGIS/MoramangaPops.shp")
ctar.gps <- read.csv("data/ctar_gps.csv")
```

```{r matching to gis file}
## Summary stats
monthly.bites <-rowSums(exposure.mat)/ncol(exposure.mat)
moramanga$commune.midpops <- moramanga$pop2015adj + (moramanga$pop2020adj - moramanga$pop2015adj)/2
moramanga$monthly.bites <- monthly.bites[match(moramanga$commune, names(monthly.bites))]
moramanga$monthly.incidence <- moramanga$monthly.bites/moramanga$commune.midpops*1e5
# distance!

## travel time layer
travel.times <- raster("output/study.area.accessibility.tif")
mora.access <- mask(travel.times, moramanga)
mora.access <- crop(mora.access, moramanga)
moramanga <- raster::extract(mora.access, moramanga, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
moramanga$comm.long <- coordinates(moramanga)[,1]
moramanga$comm.lat <- coordinates(moramanga)[,2]

# plot it
plot(mora.access/60,
     breaks = c(0, 1, 2, 3, 4, 5, 6, 7,8, 16, max(values(mora.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
             "#ae017e", "#7a0177", "#49006a"), axes = FALSE, box = FALSE)
points(moramanga$comm.long[moramanga$commune == "Moramanga"], moramanga$comm.lat[moramanga$commune == "Moramanga"], pch = 18, col = "grey", cex = 2)

## correlation checks (need to add in distance and proportion of pop within x km/hrs)
cor(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")])
splom(log(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")]), type=c('p','r'))
```

## Empirical estimates of p_report
```{r empirical}
exposures.per100k.dogs.high <- 1000*0.39
exposures.per100k.dogs.low <- 600*0.39

hdr.high <- 25
hdr.low <- 5

expected.exposures.high <- moramanga$commune.midpops/hdr.low/1e5*exposures.per100k.dogs.high # annual bites assuming low HDR and high exposures
expected.exposures.low <- moramanga$commune.midpops/hdr.high/1e5*exposures.per100k.dogs.low # annual bites assuming high HDR and low exposures
expected.exposures.mid <- (expected.exposures.high + expected.exposures.low)/2
# middle of the road estimate of incidence

reported <- moramanga$monthly.bites*12

p_rabid_max_low <- expected.exposures.low/reported*0.9
p_rabid_max_high <- expected.exposures.high/reported*0.9
p_rabid_max_low[p_rabid_max_low >= 1] <- 1
p_rabid_max_high[p_rabid_max_high > 1] <- 1

plot(moramanga$study.area.accessibility, p_rabid_max_low, ylim = c(0, 1), bty = "l", col = "darkblue") 
plot(moramanga$study.area.accessibility, p_rabid_max_high, ylim = c(0, 1), bty = "l", col = "darkblue")

empirical.reporting <- function (bites, p_rabid = 0.2, expected, samples = 1000) {
    p_rabid_max = expected/bites*0.9
    p_rabid_max[p_rabid_max >= p_rabid] <- p_rabid
    p_report <- rep(NA, length(bites))

    for (i in 1:length(bites)){
      reported_rabies <- rpois(samples, bites[i]*p_rabid_max[i])
      expected_rabies <- rpois(samples, expected[i])
      p_reported <- reported_rabies/expected_rabies
      p_report[i] <- mean(p_reported[p_reported != Inf], na.rm = TRUE)
    }
  return(p_report)
}

empirical.reporting.fixed <- function (bites, p_rabid = 0.2, expected) {
    p_rabid_max = expected/bites*0.9
    p_rabid_max[p_rabid_max >= p_rabid] <- p_rabid
    p_report <- bites*p_rabid_max/expected
    return(p_report)
}

plot(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.05, expected.exposures.low),
     ylim = c(0, 1), bty = "l", col = "darkblue") # best case
points(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.9, expected.exposures.low), 
       col = "blue") # worst case
points(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.9, expected.exposures.high), 
       col = "red", pch = 20) # middle of the ground scenario
points(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.05, expected.exposures.high), 
       col = "darkred", pch = 20) # worst case

## expected to observed
plot(expected.exposures.high, moramanga$monthly.bites*12, 
     xlim = c(0, max(expected.exposures.high)), 
     ylim = c(0, max(moramanga$monthly.bites*12)), bty = "l", pch = 20)
abline(a = 0, b = 1)

plot(moramanga$commune.midpops, p_rabid_max_high)
plot(moramanga$study.area.accessibility, p_rabid_max_high)
plot(moramanga$commune.midpops, p_rabid_max_low)
plot(moramanga$study.area.accessibility, p_rabid_max_low)

```

## JAGs model of p_report

```{r jags model}
p_rabid_max_low[p_rabid_max_low > 0.2] <- 0.2
p_rabid_max_high[p_rabid_max_high > 0.2] <- 0.2

mora.bites <- merge(moramanga@data, exposure.mat, by.x = "commune", by.y = "row.names")
bites <- mora.bites[ , 21:36]
ttimes <- mora.bites$study.area.accessibility/60

## model
tt.model <- "model {
  # reported bites
  for (t in 1:ntimes){
    for (j in 1:nlocs){
        # Number of bites
        expected[j, t] ~ dpois(expected.mean[j])
        reported[j, t] ~ dbinom(rho[j], expected[j, t])
        bites[j, t] ~ dpois(reported[j, t]/p_rabid[j] + B_0)
    }
  }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)

    # Reporting in place g
    logit(rho[g]) <- B_tt*ttimes[g] + loc_eff[g]
  }

  tau_loc <- 1/sd

  # priors
  B_tt ~ dnorm(0, 0.001)
  sd ~ dunif(0, 100)
  B_0 ~ dnorm(0, 0.001)

}"

## data
ntimes = ncol(bites)
nlocs = nrow(bites)

data <- list (bites = bites, expected.mean = expected.exposures.high/12,
              ttimes = ttimes, ntimes = ntimes , nlocs = nlocs, p_rabid = p_rabid_max_low)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1, B_0 = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, n.adapt = 1000)

params <- c("B_tt", "sd", "B_0")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
summary(window(samps, start = burn.in))
plot(samps)
gelman.plot(samps)
gelman.diag(samps)
```

```{r jags model annual}
bites <- rowSums(mora.bites[ , 21:36])
ttimes <- mora.bites$study.area.accessibility/60

## model
tt.model <- "model {
  # reported bites
    for (j in 1:nlocs){
        # Number of bites
        expected[j] ~ dpois(expected.mean[j])
        reported[j] ~ dbinom(rho[j], expected[j])
        bites[j] ~ dpois(reported[j]/p_rabid[j] + 0.01)
    }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)

    # Reporting in place g
    logit(rho[g]) <- B_tt*ttimes[g] + loc_eff[g]
  }

  tau_loc <- 1/sd

  # priors
  B_tt ~ dnorm(5, 0.001)
  sd ~ dunif(0, 100)
}"

## data
nlocs = length(bites)

data <- list (bites = bites, expected.mean = expected.exposures.high,
              ttimes = ttimes, nlocs = nlocs, p_rabid = p_rabid_max_high)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1)

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, n.adapt = 1000)

params <- c("B_tt", "sd")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
summary(window(samps, start = burn.in))
plot(samps)
gelman.plot(samps)
gelman.diag(samps)
```

```{r jags model with incidence scaling with pop}
p_rabid_max_low[p_rabid_max_low > 0.2] <- 0.2
p_rabid_max_high[p_rabid_max_high > 0.2] <- 0.2

mora.bites <- merge(moramanga@data, exposure.mat, by.x = "commune", by.y = "row.names")
bites <- mora.bites[ , 21:36]
ttimes <- mora.bites$study.area.accessibility/60

## model
tt.model <- "model {
  # reported bites
  for (t in 1:ntimes){
    for (j in 1:nlocs){
        # Number of bites
        bites[j, t] ~ dpois(expected.mean[j] + background[j] + 0.01)
    }
  }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)

    # Reporting in place g
    logit(rho[g]) <- B_tt*ttimes[g] + loc_eff[g]

    # Incidence avg in place g
    expected.inc[g] <- B_pop*pop[g]

    # Background avg in place g
    # Proportion rabid
    log(background[g]) <- B_prop*pop[g]
  }

  tau_loc <- 1/sd

  # priors
  B_tt ~ dnorm(0, 0.001)
  sd ~ dunif(0, 100)
  B_pop ~ dnorm(0, 0.001)
  B_0 ~ dnorm (1, 0.001)
  B_prop ~ dnorm(0, 0.001)
}"

## data
ntimes = ncol(bites)
nlocs = nrow(bites)

data <- list (bites = bites, pop = moramanga$commune.midpops, 
              expected.max = round(expected.exposures.high/12 + 3), 
              ttimes = ttimes, ntimes = ntimes , nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1, 
              B_pop = rnorm(1, 1, 1e-6), B_0 = rnorm(1, 1, 1e-6), B_prop = rnorm(1, 1, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, n.adapt = 1000)

params <- c("B_tt", "sd", "B_pop", "B_0", "B_prop")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
summary(window(samps, start = burn.in))
plot(samps)
gelman.plot(samps)
gelman.diag(samps)
```

### Data overall
```{r matching to gis file}
## Summary stats
monthly.bites <-rowSums(exposure.mat)/ncol(exposure.mat)
moramanga$commune.midpops <- moramanga$pop2015adj + (moramanga$pop2020adj - moramanga$pop2015adj)/2
moramanga$monthly.bites <- monthly.bites[match(moramanga$commune, names(monthly.bites))]
moramanga$monthly.incidence <- moramanga$monthly.bites/moramanga$commune.midpops*1e5
# distance!

## travel time layer
travel.times <- raster("output/study.area.accessibility.tif")
mora.access <- mask(travel.times, moramanga)
mora.access <- crop(mora.access, moramanga)
moramanga <- raster::extract(mora.access, moramanga, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
moramanga$comm.long <- coordinates(moramanga)[,1]
moramanga$comm.lat <- coordinates(moramanga)[,2]

# plot it
plot(mora.access/60,
     breaks = c(0, 1, 2, 3, 4, 5, 6, 7,8, 16, max(values(mora.access)/60, na.rm = TRUE)),
     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
             "#ae017e", "#7a0177", "#49006a"), axes = FALSE, box = FALSE)
points(moramanga$comm.long[moramanga$commune == "Moramanga"], moramanga$comm.lat[moramanga$commune == "Moramanga"], pch = 18, col = "grey", cex = 2)
legend("topright", legend = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 16, max(values(mora.access)/60, na.rm = TRUE)), fill = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
             "#ae017e", "#7a0177", "#49006a"))

# correlation checks (need to add in distance and proportion of pop within x km/hrs)
library(GGally)
#kable(cor(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")]))
ggpairs(log(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")]), columnLabels = c("log(Pop)", "log(Travel time)", "log(Monthly Incidence)"))
#splom(log(moramanga@data[,colnames(moramanga@data) %in% c("commune.midpops", "study.area.accessibility", "monthly.incidence")]), type=c('p','r'))
```

## Empirical estimates of p_report
We can empirically calculate estimates of p_report by solving for it at our maximum and minimum estimates of incidence and maximum and minimum estimates of p_rabid to get a range of reporting probabilities for each location (here we set the maximum possible reporting to be 0.95).

```{r empirical}
exposures.per100k.dogs.high <- 1000*0.39
exposures.per100k.dogs.low <- 600*0.39

hdr.high <- 25
hdr.low <- 5

expected.exposures.high <- moramanga$commune.midpops/hdr.low/1e5*exposures.per100k.dogs.high # annual bites assuming low HDR and high exposures
expected.exposures.low <- moramanga$commune.midpops/hdr.high/1e5*exposures.per100k.dogs.low # annual bites assuming high HDR and low exposures
expected.exposures.mid <- (expected.exposures.high + expected.exposures.low)/2
# middle of the road estimate of incidence

reported <- moramanga$monthly.bites*12

p_rabid_max_low <- expected.exposures.low/reported*0.95
p_rabid_max_high <- expected.exposures.high/reported*0.95
p_rabid_max_low[p_rabid_max_low >= 1] <- 1
p_rabid_max_high[p_rabid_max_high > 1] <- 1

par(mfrow = c(1, 1), mar = c(5, 5, 2,  2))
plot(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.05, expected.exposures.low),
     ylim = c(0, 1), bty = "l", col = "darkblue", xlab = "travel time (mi") # best case
plot(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.9, expected.exposures.low), 
       col = "blue") # worst case
plot(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.9, expected.exposures.high), 
       col = "red", pch = 20) # middle of the ground scenario
plot(moramanga$study.area.accessibility, empirical.reporting.fixed(reported, p_rabid = 0.05, expected.exposures.high), 
       col = "darkred", pch = 20) # worst case
# legend("topright", bty = "n", legend = c("Low exposures + low p_rabid", "Low exposures + high p_rabid",
                    # "High exposures + high p_rabid", "High exposures + low p_rabid"),
       # col = c("darkblue", "blue", "red", "darkred"), pch = 20)
```

Here we attempt to disentagle these factors by estimating the relationship between travel time to a clinic and reporting rates based on assumptions of dog rabies incidence and background levels of non-rabid bites.

We constrain incidence of human exposures based on estimates of human exposures per 100,000 dogs in a no-vaccination scenario generated by Hampson et al. 2018. We then set upper and lower bounds on the human:dog ratio (between 5 people per dog to 25 people per dog) to get the maximum and minimum expected annual bites per 100,000 people for each commune.

We then fit a model to our reported bite data (annual reported bites by commune) where bite incidence is predicted by population size, but only within a plausible range of incidences by estimating a scaling factor of incidence with population size. We also assume that background incidence of reported non-rabid dog bites increase with population size (Hampson et al. 2018, Figure 2). Then we estimate the incidence scaling factor by population size, the relationship between travel time and the reporting rate, and the relationship between population size and the proportion rabid, such that:

```{r Annual jags model with incidence scaling with pop and estimating p_rabid}
## model
tt.model <- "model {
  # reported bites
    for (j in 1:nlocs){
        # Number of bites
        rabid[j] ~ dpois(expected.mean[j])
        reported[j] ~ dbinom(rho[j], rabid[j])
        bites[j] ~ dpois(reported[j] + A[j])
    }
  `
  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)
    pop_eff[g] ~ dnorm(0, tau_pop)

    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]

    # Background avg in place g
    log(A[g]) <- B_pop*pop[g] + pop_eff[g]
  }

  tau_loc <- 1/sd
  tau_pop <- 1/sd.pop

  # priors
  B_tt ~ dnorm(0, 0.1)
  sd ~ dunif(0, 100)
  sd.pop ~ dunif(0, 100)
  B_pop ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
}"
```

And we set it so that we get the most conservative estimates of the relationship between reporting and travel times (i.e. taking out the maximum observed impact of pop on incidence)

```{r Data and estimation}
mora.bites <- merge(moramanga@data, exposure.mat, by.x = "commune", by.y = "row.names")
pop <- moramanga$commune.midpops
bites <- round(moramanga$monthly.bites*12)
ttimes <- mora.bites$study.area.accessibility/60
max.sf <- (expected.exposures.high/expected.exposures.mid)[1] # upper constraint of scaling factor
min.sf <- (expected.exposures.low/expected.exposures.mid)[1] # lower constraint of scaling factor
B_pop.max <- min(max.sf/pop)
B_pop.max*pop > max.sf # check to see if max good
B_pop.min <- max(min.sf/pop)
B_pop.min*pop < min.sf # check to see if min is good

## expected to observed
par(mfrow = c(1, 2))
plot(expected.exposures.high, moramanga$monthly.bites*12, col = "red", bty = "l", pch = 20)
lines(expected.exposures.high, expected.exposures.high, col = "red")
plot(expected.exposures.low, moramanga$monthly.bites*12, col = "blue", bty = "l", pch = 20)
lines(expected.exposures.low, expected.exposures.low, col = "blue")

plot(moramanga$commune.midpops, 
     (moramanga$monthly.bites*12)/(expected.exposures.mid*(B_pop.min*moramanga$commune.midpops)), 
      col = "red", bty = "l", pch = 20)
abline(h = 1)
points(moramanga$commune.midpops, 
       moramanga$monthly.bites*12/(expected.exposures.mid*(B_pop.max*moramanga$commune.midpops)), 
       col = "blue", pch = 20)

plot(moramanga$study.area.accessibility/60, 
     (moramanga$monthly.bites*12)/(expected.exposures.mid*(B_pop.min*moramanga$commune.midpops)), 
      col = "red", bty = "l", pch = 20)
abline(h = 1)
points(moramanga$study.area.accessibility/60, 
       moramanga$monthly.bites*12/(expected.exposures.mid*(B_pop.max*moramanga$commune.midpops)), 
       col = "blue", pch = 20)

## maximums 
plot(moramanga$commune.midpops, p_rabid_max_high, bty = "l", pch = 20, 
     ylim = c(0,1), col = "blue") # low incidence
points(moramanga$commune.midpops, p_rabid_max_low, col = "red", pch = 20) # high incidence
```

```{r Estimation}
## potential relationship between reporting and travelt
## data
nlocs = length(bites)

data <- list (bites = bites, pop = moramanga$commune.midpops, 
              expected.mean = expected.exposures.mid,
              ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 5, sd.pop = 5,
              B_pop = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_pop", "sd.pop")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 2), mar = c(5, 5, 2,  2))
plot(pop, rpois(length(bites), exp(check$statistics[,"Mean"]["B_pop"]*pop))
      + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd.pop"]), bty = "l")
plot(pop, plogis(check$statistics[,"Mean"]["B_prop"]*pop), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) 
                    + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])),
     bty = "l")

```

The scaling factor is not identifiable.

### Fixing at minimum scaling factor

```{r Fixing at minimum scaling of inc w/ pop and estimating other params}
## model
tt.model <- "model {
  # reported bites
    for (j in 1:nlocs){
        # Number of bites
        bites[j] ~ dpois(rabid[j]*rho[j] + background[j] + B_0)
    }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)

    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]

    # Incidence avg in place g
    scaling.factor[g] <- B_pop*pop[g]
    rabid[g] <- scaling.factor[g]*expected.mean[g]

    # Background avg in place g
    logit(p_rabid[g]) <- B_prop*pop[g]
    background[g] <- (rabid[g]*rho[g])/p_rabid[g] - rabid[g]
  }

  tau_loc <- 1/sd

  # priors
  B_tt ~ dnorm(0, 0.1)
  sd ~ dunif(0, 100)
  B_prop ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
}"

## data
nlocs = length(bites)

data <- list (bites = bites, pop = moramanga$commune.midpops, 
              expected.mean = expected.exposures.mid,
              B_pop = B_pop.min, ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 5, 
              B_prop = rnorm(1, 0, 1e-6), 
              B_0 = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_prop", "B_0")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
summary(window(samps, start = burn.in))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(1, 1), mar = c(5, 5, 2,  2))
plot(pop, (B_pop.min*pop*expected.exposures.mid)/(pop/1e5), bty = "l")
plot(pop, plogis(check$statistics[,"Mean"]["B_prop"]*pop), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) + rnorm(length(ttimes), mean = 0, sd = check$statistics[,"Mean"]["sd"])), bty = "l")
```

### Fixing at maximum scaling factor

```{r Fixing at maximum scaling of inc w/ pop and estimating other params}
## model
tt.model <- "model {
  # reported bites
    for (j in 1:nlocs){
        # Number of bites
        bites[j] ~ dpois(rabid[j]*rho[j] + background[j] + B_0)
    }

  # spatial effects
  for (g in 1:nlocs){
    # Spatial random effect
    loc_eff[g] ~ dnorm(0, tau_loc)

    # Reporting in place g
    logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]

    # Incidence avg in place g
    scaling.factor[g] <- B_pop*pop[g]
    rabid[g] <- scaling.factor[g]*expected.mean[g]

    # Background avg in place g
    logit(p_rabid[g]) <- B_prop*pop[g]
    background[g] <- (rabid[g]*rho[g])/p_rabid[g] - rabid[g]
  }

  tau_loc <- 1/sd

  # priors
  B_tt ~ dnorm(0, 0.1)
  sd ~ dunif(0, 100)
  B_prop ~ dnorm(0, 0.1)
  B_0 ~ dnorm(0, 0.1)
}"

## data
nlocs = length(bites)

data <- list (bites = bites, pop = moramanga$commune.midpops, 
              expected.mean = expected.exposures.mid,
              B_pop = B_pop.max, ttimes = ttimes, nlocs = nlocs)

inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 5, 
              B_prop = rnorm(1, 0, 1e-6), 
              B_0 = rnorm(1, 0, 1e-6))

jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                       n.adapt = 1000)

params <- c("B_tt", "sd", "B_prop", "B_0")

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
summary(window(samps, start = burn.in))
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

samps <- coda.samples(jags_mod, params, n.iter = 10000)
burn.in <- 5000
check <- summary(window(samps, start = burn.in))
check
plot(samps)
# gelman.plot(samps)
gelman.diag(samps)

## parameter estimates
par(mfrow = c(2, 2), mar = c(4, 6, 2, 6))
plot(pop, (B_pop.max*pop*expected.exposures.mid)/(pop/1e5), bty = "l")
plot(pop, plogis(check$statistics[,"Mean"]["B_prop"]*pop), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes)), bty = "l")
plot(ttimes, plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes) + rnorm(length(ttimes), mean = 0, 
                              sd = check$statistics[,"Mean"]["sd"])), bty = "l")
```

```{r jags model with incidence scaling with pop and fixed p_rabid}
# ## model
# tt.model <- "model {
#   # reported bites
#   for (t in 1:ntimes){
#     for (j in 1:nlocs){
#         # Number of bites
#         bites[j, t] ~ dpois(rabid[j]*rho[j] + background[j] + B_0)
#     }
#   }
# 
#   # spatial effects
#   for (g in 1:nlocs){
#     # Spatial random effect
#     loc_eff[g] ~ dnorm(0, tau_loc)
# 
#     # Reporting in place g
#     logit(rho[g]) <- B_tt*(1/ttimes[g]) + loc_eff[g]
# 
#     # Incidence avg in place g
#     scaling.factor[g] <- B_pop*pop[g]
#     rabid[g] <- scaling.factor[g]*expected.mean[g]
#     
#     # Background avg in place g
#     p_rabid[g] <- ifelse(rabid[g]/avg.reported[g]*0.9 >= p_rabid_max,
#                               p_rabid_max, rabid[g]/avg.reported[g]*0.9)
#     background[g] <- rabid[g]/p_rabid[g] - rabid[g]
#   }
# 
#   tau_loc <- 1/sd
# 
#   # priors
#   B_tt ~ dnorm(0, 0.01)
#   sd ~ dunif(0, 100)
#   B_pop ~ dunif(B_pop.min, B_pop.max)
#   B_0 ~ dnorm(0, 0.01)
# }"

# ## data
# ntimes = ncol(bites)
# nlocs = nrow(bites)
# 
# data <- list (bites = bites, pop = moramanga$commune.midpops, 
#               expected.mean = expected.exposures.mid, avg.reported = moramanga$monthly.bites,
#               p_rabid_max = 0.9, B_pop.min = B_pop.min, B_pop.max = B_pop.max,
#               ttimes = ttimes, ntimes = ntimes , nlocs = nlocs)
# 
# inits <- list(B_tt = rnorm(1, 0, 1e-6), sd = 1, 
#               B_pop = rnorm(1, (B_pop.max + B_pop.min)/2, 1e-6))
# 
# jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 10, n.adapt = 1000)
# 
# params <- c("B_tt", "sd", "B_pop")
# 
# samps <- coda.samples(jags_mod, params, n.iter = 10000)
# burn.in <- 5000
# summary(window(samps, start = burn.in))
# plot(samps)
# gelman.plot(samps)
# gelman.diag(samps)
```

### 2018-09-20 interactiveMap.Rmd
Initial commit

--- 2018-09-20 archive/accessPEP.Rmd
@@
   mutate_at(vars(starts_with("year")), funs(replace(., is.na(.), 0))) %>%
-  mutate(avg.bites100k = (select(., starts_with("year")) %>% rowSums())/4/(midpops/1e5)) -> dist.data
-dist.data %>% 
-  filter(district)
-## subset out Analamanga (IPM) + Haute Matsiatra (Fianarantsoa)
+  mutate(avg.bites100k = (select(., starts_with("year")) %>% rowSums())/4/(midpops/1e5),
+         dogs_min = round(midpops/25), dogs_max = round(dist.data$midpops/5)) -> dist.data
+
+## set to NA all non-reporters!
 

### 2018-09-20 R/R_generic_accessibilty_mapping_script.r

@@ -1,0 +1,79 @@
+# Accessibility Mapping in Google Earth Engine (GEE)
+# 
+# Dan Weiss, Malaria Atlas Project, University of Oxford
+# 2017-11-06
+#
+# This script requires the gdistance package (van Etten, J. R Package gdistance: Distances and Routes on Geographical Grids. Journal of Statistical Software 76, 1-21)
+#
+# This script requires the two user supplied datasets:
+# (a) The friction surface, which is available here:  http://www.map.ox.ac.uk/accessibility_to_cities/
+# (b) A user-supplied .csv of points (i.e., known geographic coordinates) 
+#
+# Notes:
+# (a) All file paths and names should be changed as needed.
+# (b) Important runtime details can be found in the comments.
+# (c) This script is suitable only for analyses of moderately sized areas for most (e.g., up to 10 million km^2 in lower latitude settings - GLOBAL RUNS WILL NOT WORK).
+#     We recommend using Google Earth Engine for larger areas, with the exception of high-latitude areas where custom approaches are typically required.
+#
+# Citation: D.J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M. Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier, 
+# T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E. Battle, S. Bhatt, and P.W. Gething. A global map of travel time to cities to assess 
+# inequalities in accessibility in 2015. (2018). Nature. doi:10.1038/nature25181.
+# 
+
+## Required Packages
+require(gdistance)
+
+# User Defined Variables - used if clipping from the global layer, if no clipping is needed, see lines 54-55 (currently commented out).
+# This could also be accomplished by importing a shapefile (for example) 
+# Geographic Coordinates (WGS84)
+left   <- -2.0
+right  <- 0.0
+bottom <- 50.0
+top    <- 52.0
+transition.matrix.exists.flag <- 0 # if the geo-corrected graph has already been made, this can save time.  Uses the same T.GC.filename as specified using the T.GC.filename variable.
+
+# Input Files
+friction.surface.filename <- 'E:\\accessibility\\friction_surface_2015_v1.0.tif'
[…]

### 2018-09-20 interactiveMap.Rmd
adding push_up file for syncing to shared dropbox with IPM (goes to Mal_analyses/folder)

@@
 ---
-title: "PEP Access"
+title: "CHanging to checkPEP Access"
 author: "Malavika Rajeev"

### 2018-09-20 push_up.sh
adding push_up file for syncing to shared dropbox with IPM (goes to Mal_analyses/folder)

@@ -1,0 +1,7 @@
+#!/bin/bash
+## -r recurse through sub-directories
+## -L transform symlink into reference file/dir
+## -v be verbose about printing messages
+## -z compresses data before transfer and decompresses after transfer
+
+rsync -rLvz --update --exclude '*.git' --exclude '.Rproj*' --exclude archive --exclude '.gitignore' --exclude 'README.md' --exclude 'push_up.sh' --exclude lit ~/Documents/MadaAccess/ ~/Dropbox/MadaPEP/Mal_analyses/

### 2018-10-16 archive/accessPEP3.Rmd
Reorganizing files and creating draft rmd to push to googledrive

--- 2018-09-20 accessPEP4.Rmd
@@
 
-### Preliminary data from Madagascar Clinics
+``` {r label, out.width = "75%", fig.cap = "Figure 1", echo = FALSE, message = FALSE}
+include_graphics("figs/figure2.pdf")
+
+```
+
+### Preliminary data from the Moramanga District 
 * Commune level average travel time to the closest CTAR (generated from global travel time layer from Malaria Atlas Project).
@@
 ```{r setup, echo = FALSE, message = FALSE, results = 'hide'}
-## Clean-up
-## Packages
+## libraries
 # for gis 
 library(knitr)
-library(raster)
-library(maptools)
-library(maps)
-library(GISTools)
-library(rgdal)
-library(sp)
-library(rgdal)
-library(gdistance)
 # everything else
@@
 p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
-mada.district <- readOGR("data/MadaGIS/MadaPops.shp", p4s = p4s)
+
+## data
+exposure.mat <- read.csv("data/exposures_by_commune_Moramanga.csv", header = TRUE, row.names = 1)
+moramanga <- readOGR("data/MoramangaGIS/MoramangaPops.shp")
 ctar.gps <- read.csv("data/ctar_gps.csv")
-ctar.gps <- SpatialPoints(cbind(ctar.gps$LONGITUDE, ctar.gps$LATITUDE),
-                          proj4string = crs(mada.district))
-ctar.atts <- readOGR("data/MadaGIS/31CTAR.shp")
-# ## getting travel time layer
-mada.access <- raster("output/study.area.accessibility.tif")
-mada.access <- mask(mada.access, mada.district)
-plot(mada.access/60, breaks = c(0, 1, 2, 3, 4, 5, 6, 8, 16, max(values(mada.access)/60, na.rm = TRUE)),
-     col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
-#col = colorRampPalette(c("navyblue", "mediumturquoise", "orange", "red"))(11), axes = FALSE, box = FALSE)
-points(ctar.gps, pch = 18, col = "grey", cex = 1)
 
-## getting which districts have a ctar
-mada.district$ctar.in <- ifelse(is.na(match(mada.district$district, ctar.atts$District)), 0, 1)
-mada.district$ctar.in[mada.district$district %in% c('Manakara Atsimo', 'Toliary-I', 'Taolagnaro',
-                                                    'Nosy-Be', "Ambovombe-Androy")] <- 1
+## data
+monthly.bites <-rowSums(exposure.mat)/ncol(exposure.mat)
+moramanga$commune.midpops <- (moramanga$pop2015adj + moramanga$pop2020adj)/2
+moramanga$monthly.bites <- monthly.bites[match(moramanga$commune, names(monthly.bites))]
+moramanga$monthly.incidence <- moramanga$monthly.bites/moramanga$commune.midpops*1e5
 
-## should equal 31
-sum(mada.district$ctar.in)
-
-## mid.pops
-mada.district$midpops <- (mada.district$pop2015adj + mada.district$pop2020adj)/2
-
-## ctar data: getting exposure matrix (district by year)
-ctar.data <- read.csv("data/SaisieRage_DATA_2018-06-28_0956.csv")
-ctar.data %>% 
-  filter(type != "contact") %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(mdy(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), district) %>% # group by year and district
-  summarize(n = n()) -> exposures # get count of exposures
-
-## merge district shapefile with the summarized matrix!
-exposure.mat <- dcast(exposures, district ~ year)
-colnames(exposure.mat) <- c("district", "year2013", "year2014", "year2015", "year2016", "year2017", "year2018")
-mada.district@data <- merge(mada.district@data, exposure.mat, by.x = "mdg_dis_co", by.y = "district", all.x = TRUE)
-vars <- c("mdg_dis_co", "district", "midpops", "study_area", "ctar.in",
-          "year2014", "year2015", "year2016", "year2017")
-
-## Getting all dist data
-mada.district@data %>%
-  select(!!vars) %>%
-  mutate_at(vars(starts_with("year")), funs(replace(., is.na(.), 0))) %>%
-  mutate(avg.bites100k = (select(., starts_with("year")) %>% rowSums())/4/(midpops/1e5),
-         dogs_min = round(midpops/25), dogs_max = round(midpops/5)) -> dist.data
-
-## districts that are non-reporter and data not available yet CTAR
-no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
-             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
-             "Marolambo", "Sainte-Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano", 
-             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
-             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
-             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
-dist.data %>%
-  filter(!(district %in% no_data)) -> dist.data
-nrow(dist.data)
+## travel time layer
+travel.times <- raster("output/study.area.accessibility.tif")
+mora.access <- mask(travel.times, moramanga)
+mora.access <- crop(mora.access, moramanga)
+moramanga <- raster::extract(mora.access, moramanga, fun = mean, na.rm=TRUE, df = TRUE, sp = TRUE)
+moramanga$comm.long <- coordinates(moramanga)[,1]
+moramanga$comm.lat <- coordinates(moramanga)[,2]
+# # plot it
+# plot(mora.access/60,
+#      breaks = c(0, 1, 2, 3, 4, 5, 6, 7,8, 16, max(values(mora.access)/60, na.rm = TRUE)),
+#      col = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", 
+#              "#ae017e", "#7a0177", "#49006a"), axes = FALSE, box = FALSE)
+# points(moramanga$comm.long[moramanga$commune == "Moramanga"], moramanga$comm.lat[moramanga$commune == "Moramanga"], pch = 18, col = "grey", cex = 2)
 
@@
 
-Applying this framework to our data on reported bites in the Madagascar, we see that estimated rabies incidence exceeds expected incidence in areas with high incidence of reported bites and is lower than expected for areas with little data/low incidence of reported bites (Figure 1).
+Applying this framework to our data on reported bites in the Moramanga District, we see that estimated rabies incidence exceeds expected incidence in areas with high incidence of reported bites and is lower than expected for areas with little data/low incidence of reported bites (Figure 1).
 
@@
 
-ordered.bites <- dist.data$avg.bites100k[order(dist.data$study_area)]
+ordered.bites <- moramanga$monthly.incidence[order(moramanga$study.area.accessibility)]*12
 par(mar = c(4, 10, 4, 4))
-plot(ordered.bites*0.6/0.85, 1:length(ordered.bites), bty = "l", pch = 20, col = "darkred", 
+plot(ordered.bites*0.6/0.85, 1:length(monthly.bites), bty = "l", pch = 20, col = "darkred", 
      xlab = "Rabies exposures per 100k persons", ylab = "", axes = FALSE, 
@@
 axis(1)
-text(dist.data$district[order(dist.data$study_area)], y = 1:length(ordered.bites), x = -10, adj = 1, xpd = TRUE)
-points(ordered.bites*0.2/0.85, 1:length(ordered.bites),
+text(moramanga$commune[order(moramanga$study.area.accessibility)], y = 1:length(monthly.bites), x = -10, adj = 1, xpd = TRUE)
+points(ordered.bites*0.2/0.85, 1:length(monthly.bites),
        pch = 20, col = "darkred")
-segments(y0 = 1:length(ordered.bites), x0 = ordered.bites*0.2/0.85,
-         y1 = 1:length(ordered.bites),
+segments(y0 = 1:length(monthly.bites), x0 = ordered.bites*0.2/0.85,
+         y1 = 1:length(monthly.bites),
          x1 = ordered.bites*0.6/0.85, col = "darkred")
@@
 ## Empirical p_report
-p_rabid_high <- exps.per100k.high/(dist.data$avg.bites100k)
+p_rabid_high <- exps.per100k.high/(moramanga$monthly.incidence*12)
 p_rabid_high[p_rabid_high >= 0.95] <- 0.95
-p_rabid_low <- exps.per100k.low/(dist.data$avg.bites100k)
+p_rabid_low <- exps.per100k.low/(moramanga$monthly.incidence*12)
 p_rabid_low[p_rabid_low >= 0.95] <- 0.95
 
-plot(dist.data$study_area/60, 
-     dist.data$avg.bites100k*p_rabid_high/exps.per100k.high, ylim = c(0, 1), 
+plot(moramanga$study.area.accessibility/60, 
+     moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high, ylim = c(0, 1), 
      col = "red", pch = 20, bty = "l", xlab = "Travel times (hrs)", ylab = "Reporting")
-points(dist.data$study_area/60, 
-     dist.data$avg.bites100k*p_rabid_low/exps.per100k.low,  
+points(moramanga$study.area.accessibility/60, 
+     moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low,  
      col = "darkred")
-segments(x0 = dist.data$study_area/60, y0 = dist.data$avg.bites100k*p_rabid_low/exps.per100k.low,
-         x1 = dist.data$study_area/60,
-         y1 = dist.data$avg.bites100k*p_rabid_high/exps.per100k.high, col = "darkred")
+segments(x0 = moramanga$study.area.accessibility/60, y0 = moramanga$monthly.incidence*12*p_rabid_low/exps.per100k.low,
+         x1 = moramanga$study.area.accessibility/60,
+         y1 = moramanga$monthly.incidence*12*p_rabid_high/exps.per100k.high, col = "darkred")
 legend("topright", c("High incidence", "Low incidence"), col = c("Red", "darkred"), pch = c(20, 1), 
        bty = "n", inset = c(-0.2, 0), xpd = TRUE)
-```
 
-```{r district data}
-bites <- dist.data[,6:9]
-ttimes <- dist.data$study_area
-pop <- dist.data$midpops
-times <- 1:4
-nlocs <- nrow(dist.data)
-exps.high <- exps.per100k.high*dist.data$midpops/1e5
-exps.low <- exps.per100k.low*dist.data$midpops/1e5
-burn.in = 5000
 ```
@@
 #### Models of reporting by travel time to clinic
-To generate more probable estimates of rabies burden given what we know about access to PEP in Maadagascar, we estimate a district specific reporting rate as a function of travel time ($T$) to the nearest ARMC:
+To generate more probable estimates of rabies burden given what we know about access to PEP in Maadagascar, we estimate a commune specific reporting rate as a function of travel time ($T$) to the nearest ARMC given an expected rabies exposures incidence and corresponding $p_{rabid}$:
 
-$$ \rho_j = logistic(B_{tt}T_j+ B_{0})$$
-We can specify our expected bites in a few ways:
-
-##### Given an expected rabies exposures incidence and corresponding $p_{rabid}$:
-
-$$ B_j \sim Poisson(\frac{\rho_j R}{p_{rabid_j}})$$
+$$ B \sim Poisson(\frac{\rho R}{p_{rabid}})$$
+$$ \rho = logistic(B_{tt}T+ B_{0})$$
     $$
-           p_{rabid_j}=
+           p_{rabid}=
             \begin{cases}
-            x, & \text{if}\ \frac{R_j}{B_j} > x \\
-            \frac{R_j}{B_j}, & \text{otherwise}
+            x, & \text{if}\ \frac{R}{B} > x \\
+            \frac{R}{B}, & \text{otherwise}
             \end{cases}
@@
 
-```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
+```{r Annual jags model with p_rabid + p_report estimated, high incidence, echo = FALSE, message = FALSE, results = 'hide'}
 ## model
-tt.model.1 <- "model {
+tt.model <- "model {
+# reported bites
 for (j in 1:nlocs){
-  for (t in 1:times){
 # Number of bites
-    bites[j, t] ~ dpois(rho[j]*expected.mean[j]/p_rabid[j, t])
-  }
+  bites[j] ~ dpois(expected.mean[j]*rho[j]/p_rabid[j])
 }
@@
 
-est.p.1 <- function (bites, ttimes, times, nlocs, expected.mean, p_rabid.est){
+est.p <- function (bites, ttimes, nlocs, expected.mean, p_rabid.est){
   p_rabid <- expected.mean/bites
   p_rabid[p_rabid >= p_rabid.est] <- p_rabid.est
-
-  data <- list (bites = bites, times = times,
+  data <- list (bites = bites, 
               expected.mean = expected.mean, p_rabid = p_rabid,
@@
 
-  jags_mod <- jags.model(textConnection(tt.model.1), data = data, inits = inits, n.chains = 5, 
+  jags_mod <- jags.model(textConnection(tt.model), data = data, inits = inits, n.chains = 5, 
                        n.adapt = 1000)
@@
 }
-check <- est.p.1(bites = bites, times = times, ttimes = ttimes, nlocs = nlocs,
-                        expected.mean = exps.low, p_rabid.est = 0.5)
-```
 
-##### Given an expected rabies exposures incidence and non-rabid bites:
-
-$$ B_j \sim Poisson(\rho_j (R_j + N_j))$$
-where $N$ is a function of human population size (more people = more people getting bitten by dogs?):
-$$ N_j \sim f(pop_j)$$
-This assumes that reporting is the same for non-rabid and rabid bites, but that the incidence of non-rabid bites is not uniform across space (as with rabies incidence) and increases with increasing population size.
-
-```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
-## model
-tt.model.2 <- "model {
-# reported bites
-for (j in 1:nlocs){
-  for (t in 1:times){
-    # Number of bites
-    bites[j, t] ~ dpois(rho[j]*(expected.mean[j] + N[j]))
-  }
-}
-
-# spatial effects
-  for (g in 1:nlocs){
-    # Reporting in place g
-    logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
-    # Non-rabid incidence in place g
-    log(N[g]) <- B_pop*pop[g] + B_1
-  }
-
-# priors
-  B_tt ~ dnorm(0, 0.1)
-  B_pop ~ dnorm(0, 0.1)
-  B_0 ~ dnorm(0, 0.1)
-  B_1 ~ dnorm(0, 0.1)
-}"
-
-est.p.2 <- function (bites, ttimes, times, pop, nlocs, expected.mean){
-  data <- list (bites = bites, times = times, pop = pop,
-              expected.mean = expected.mean, p_rabid = p_rabid,
-              ttimes = ttimes, nlocs = nlocs)
-
-  inits <- list(B_tt = rnorm(1, 0, 1e-6), 
-              B_0 = rnorm(1, 0, 1e-6), B_pop = rnorm(1, 0, 1e-6), 
-              B_1 = rnorm(1, 0, 1e-6))
-
-  jags_mod <- jags.model(textConnection(tt.model.2), data = data, inits = inits, n.chains = 5, 
-                       n.adapt = 1000)
-
-  params <- c("B_tt", "B_0", "B_1", "B_pop")
-  samps <- coda.samples(jags_mod, params, n.iter = 10000)
-  return(samps)
-}
-```
-
-##### Given an expected rabies exposures incidence and reported non-rabid bites:
-
-$$ B_j \sim Poisson(\rho_jR_j + N_j)$$
-where $N_j$ is a function of travel time (closer you are to a clinic, higher incidence of reported non-rabid bites):
-$$ N_j \sim f(T_j)$$
-This assumes that reporting may not be the same for non-rabid and rabid bites, but that the incidence of non-rabid bites increases with travel time, just as reporting increases with travel time.
-This may be true, but we will likely run into identifiability issues (i.e.
-estimating two params which increase with travel time, reporting and reported non-rabid incidence).
-
-```{r Annual jags model with p_report estimated, echo = FALSE, message = FALSE, results = 'hide'}
-## model
-tt.model.3 <- "model {
-# reported bites
-for (j in 1:nlocs){
-  for (t in 1:times){
-    # Number of bites
-    bites[j, t] ~ dpois(rho[j]*expected.mean[j] + N[j])
-  }
-}
-
-# spatial effects
-  for (g in 1:nlocs){
-    # Reporting in place g
-    logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
-    # Non-rabid incidence in place g
-    log(N[g]) <- B_nr*(1/ttimes[g]) + B_1
-  }
-
-# priors
-  B_tt ~ dnorm(0, 0.1)
-  B_nr ~ dnorm(0, 0.1)
-  B_0 ~ dnorm(0, 0.1)
-  B_1 ~ dnorm(0, 0.1)
-}"
-
-est.p.3 <- function (bites, ttimes, times, nlocs, expected.mean){
-  data <- list (bites = bites, times = times,
-              expected.mean = expected.mean,
-              ttimes = ttimes, nlocs = nlocs)
-
-  inits <- list(B_tt = rnorm(1, 0, 1e-6), 
-              B_0 = rnorm(1, 0, 1e-6), B_nr = rnorm(1, 0, 1e-6), 
-              B_1 = rnorm(1, 0, 1e-6))
-
-  jags_mod <- jags.model(textConnection(tt.model.3), data = data, inits = inits, n.chains = 5, 
-                       n.adapt = 1000)
-
-  params <- c("B_tt", "B_0", "B_1", "B_nr")
-  samps <- coda.samples(jags_mod, params, n.iter = 10000)
-  return(samps)
-}
-```
-
-```{r Annual jags model with p_rabid + p_report estimated, high incidence, echo = FALSE, message = FALSE, results = 'hide'}
-## models assuming high incidence
+bites = round(moramanga$monthly.bites*12)
+ttimes = moramanga$study.area.accessibility/60
+nlocs = length(moramanga$commune)
+exp.low = exps.per100k.low*moramanga$commune.midpops/1e5
+exp.high = exps.per100k.high*moramanga$commune.midpops/1e5
+burn.in = 5000
 

### 2018-10-16 archive/accessPEP4.Rmd
Reorganizing files and creating draft rmd to push to googledrive

--- 2018-09-20 accessPEP4.Rmd
@@
 require("knitr")
-opts_knit$set(root.dir = "/Users/mrajeev/Dropbox/MadaPEP")
 knitr::opts_chunk$set(fig.pos = 'h')
@@
 pop <- dist.data$midpops
-times <- 1:4
-nlocs <- nrow(dist.data)
+times <- ncol(bites)
+nlocs <- nrow(bites)
 exps.high <- exps.per100k.high*dist.data$midpops/1e5
@@
     logit(rho[g]) <- B_tt*(1/ttimes[g]) + B_0
+
     # Non-rabid incidence in place g
@@
   data <- list (bites = bites, times = times, pop = pop,
-              expected.mean = expected.mean, p_rabid = p_rabid,
+              expected.mean = expected.mean,
               ttimes = ttimes, nlocs = nlocs)
@@
 }
+
+check <- est.p.2(bites = bites, times = times, ttimes = ttimes, pop = pop, nlocs = nlocs,
+                        expected.mean = exps.high)
+plot(check)
+gelman.diag(check)
+check <- summary(window(check, start = burn.in))
+ttimes.plot <- seq(0.01, 12, by = 0.2)*60
+par(mar = c(2,2,2,2))
+plot(ttimes.plot,
+     plogis(check$statistics[,"Mean"]["B_tt"]*(1/ttimes.plot)
+                  + check$statistics[,"Mean"]["B_0"]),
+     type = "l", col = "slateblue1")
+
 ```

### 2018-10-16 archive/interactiveMap.Rmd
Reorganizing files and creating draft rmd to push to googledrive

---
title: "Locating neighborhood diversity in the American metropolis"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    social: menu
    source_code: https://github.com/walkerke/neighborhood_diversity
    theme: simplex

---

<!-- Learn more about flexdashboard at https://rstudio.github.io/flexdashboard -->

```{r setup, include=FALSE}

library(shiny)
library(leaflet) # devtools::install_github('rstudio/leaflet')
# library(highcharter) # devtools::install_github('jbkunst/highcharter')
library(plotly) # devtools::install_github('ropensci/plotly')
library(ggplot2) # devtools::install_github('hadley/ggplot2')
library(sp)
# library(dplyr)
library(flexdashboard) # devtools::install_github('rstudio/flexdashboard')
library(rgeos)
library(mapproj)
library(maptools)
library(readr)
library(ggthemes)
library(viridis)

# Define the list of available metros
lookup <- structure(c(12060L, 12420L, 12580L, 13820L, 14460L, 15380L, 16740L, 
16980L, 17140L, 17460L, 18140L, 19100L, 19740L, 19820L, 25540L, 
26420L, 26900L, 27260L, 28140L, 29820L, 31100L, 31140L, 32820L, 
33100L, 33340L, 33460L, 34980L, 35620L, 36420L, 36740L, 37980L, 
38060L, 38300L, 38900L, 39300L, 40060L, 40380L, 40900L, 
41180L, 41620L, 41700L, 41740L, 41860L, 42660L, 45300L,  
47900L), .Names = c("Atlanta", "Austin", "Baltimore", "Birmingham", 
"Boston", "Buffalo", "Charlotte", "Chicago", "Cincinnati", "Cleveland", 
"Columbus", "Dallas-Fort Worth", "Denver", "Detroit", "Hartford", "Houston", 
"Indianapolis", "Jacksonville", "Kansas City", "Las Vegas", "Los Angeles", 
"Louisville", "Memphis", "Miami", "Milwaukee", "Minneapolis-St. Paul", 
"Nashville", "New York", "Oklahoma City", "Orlando", "Philadelphia", 
"Phoenix", "Pittsburgh", "Portland", "Providence", 
"Richmond", "Rochester", "Sacramento", "St. Louis", "Salt Lake City", 
"San Antonio", "San Diego", "San Francisco-Oakland", "Seattle", "Tampa-St. Petersburg", 
"Washington"))

# Read in data, and subset for the selected metro
full_tracts <- readRDS('full_simp2.rds')
map_data <- read.csv("map_data.csv")
p4s <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
mada.district <- readOGR("data/MadaGIS/MadaPops.shp", p4s = p4s)
ctar.gps <- read.csv("data/ctar_gps.csv")
ctar.gps <- SpatialPoints(cbind(ctar.gps$LONGITUDE, ctar.gps$LATITUDE),
                          proj4string = crs(mada.district))
metro <- reactive({

  m <- dplyr::select(map_data, district, input$var)

  return(m)

})

```

Sidebar {.sidebar}
======================================================================

```{r}

# Define inputs
selectInput('var', label = 'Select data', choices = colnames(map_data), selected = "midpops")

# Set up data download
data_for_dl <- reactive({

  dat <- map_data

})

output$downloadCSV <- downloadHandler(
  filename = 'data.csv', 
  content = function(file) {
    write_csv(data_for_dl(), file)
  }
)

downloadLink('downloadCSV', label = 'Download CSV of district data')

```

Use the __Explore metros__ tab to explore neighborhood diversity for your chosen metropolitan area in 2010.  The red line on the scatterplot represents a locally-weighted estimate of how diversity varies in the metropolitan area by distance from its urban core or cores.  Click and drag on the scatterplot to highlight the corresponding Census tracts on the map below, and click on a Census tract on the map to generate a chart of race and ethnicity counts.

Click the __Compare over time__ tab to examine how locally-weighted estimates of neighborhood diversity by distance from the urban core has varied between the 1990, 2000, and 2010 Censuses, and view maps of these shifts over time. To learn more about the project, click the __About__ tab.

Application author: [Kyle Walker](http://personal.tcu.edu/kylewalker), [Texas Christian University](http://www.tcu.edu)

Explore metros
======================================================================

Row 
-----------------------------------------------------------------------

### Map of Mada

```{r}

# Draw the map without selected tracts
output$map <- renderLeaflet({

    pal <- colorNumeric(viridis(999), NULL)

    map <- leaflet(metro()) %>%
      addProviderTiles('CartoDB.Positron') %>%
      clearShapes() %>%
      addPolygons(stroke = FALSE, smoothFactor = 0, 
                  fillColor = ~pal(input$var), fillOpacity = 0.7, 
                  layerId = ~GEOID) %>%
      addLegend(position = 'bottomright', pal = pal, 
                values = metro()$entropy, title = 'Score')

    map

  })

# Click event for the map (will use to generate chart)
click_tract <- eventReactive(input$map_shape_click, {

    x <- input$map_shape_click

    y <- x$id

    return(y)

  })

# Drag event for the scatterplot; will grab tractids of selected points
sub <- reactive({

    eventdata <- event_data('plotly_selected', source = 'source')

    if (is.null(eventdata)) {

      return(NULL) # do nothing

    } else {

      tracts <- eventdata[['key']]

      if (length(tracts) == 0) {

        tracts <- 'abcdefg' # a hack but it's working - set to something that can't be selected

      }

      if (!(tracts %in% metro()$tractid)) {

        return(NULL) # if there is not a match, do nothing as well

      } else {

        # Give back a sp data frame of the selected tracts
        sub <- metro()[metro()$tractid %in% tracts, ]

        return(sub)

      }

    }

  })

observe({

  req(sub()) # Do this if sub() is not null

  proxy <- leafletProxy('map')

  # Clear old selection on map, and add new selection
  proxy %>%
    clearGroup(group = 'sub') %>%
    addPolygons(data = sub(), fill = FALSE, color = '#00FFFF',
                opacity = 1, group = 'sub') %>%
    fitBounds(lng1 = bbox(sub())[1],
              lat1 = bbox(sub())[2],
              lng2 = bbox(sub())[3],
              lat2 = bbox(sub())[4])

})

observe({

  req(click_tract()) # do this if click_tract() is not null

  # Add the clicked tract to the map in aqua, and remove when a new one is clicked
  map <- leafletProxy('map') %>%
      removeShape('htract') %>%
      addPolygons(data = full_tracts[full_tracts$GEOID == click_tract(), ], fill = FALSE,
                  color = '#FF00FF', opacity = 1, layerId = 'htract')

  })

tract_data <- reactive({

    # Fetch data for the clicked tract
    return(metro()@data[metro()@data$GEOID == click_tract(), ])

  })

leafletOutput('map')

```

Row
-----------------------------------------------------------------------

### Table of data

```{r}

# Here, we draw the diversity gradient with ggplotly
output$scatter <- renderPlotly({

  key <- metro()$tractid # This will uniquely identify tracts for Plotly

  p1a <- ggplot(metro()@data) + 
    geom_point(alpha = 0.4, aes(Distance, Score, key = key)) + 
    theme_minimal(base_size = 14) + 
    stat_smooth(aes(Distance, Score), 
                color = '#FFD700', method = 'loess', span = input$span, se = FALSE, alpha = 1) + 
    xlab('Distance from city hall (miles)') + ylab('')

  g <- ggplotly(p1a, source = 'source') %>% 
    layout(dragmode = 'lasso', 
           yaxis = list(title = 'Diversity score'), 
           margin = list(l = 100), 
           font = list(family = 'Open Sans', size = 16))

  # Need to manually set the hoverinfo to avoid the key appearing in it
  build <- plotly_build(g)

  build$data[[1]]$text <- paste0('Distance: ', as.character(round(metro()$Distance, 2)), '<br>', 
                               'Score: ', as.character(round(metro()$Score, 2)))

  build

})

plotlyOutput('scatter', width = "80%")
```

About
============================================================================

This application is in support of the article in _Urban Studies_, ["Locating neighborhood diversity in the American Metropolis."](http://usj.sagepub.com/content/early/2016/04/29/0042098016643481.abstract)  The article analyzes geographic variations in neighborhood racial and ethnic diversity over time in large metropolitan areas in the United States.

The key metric in this article is the neighborhood-level _entropy index_ (called "diversity score" in the application), which measures the degree of neighborhood diversity for four general racial/ethnic groups: non-Hispanic white, non-Hispanic black, Hispanic, and Asian/Pacific Islander.  The entropy index $E$ is calculated as follows (Farrell and Lee 2011):

$$E = {\sum\limits_{r=1}^{n}Q_r}ln{\dfrac{1}{Q_r}}$$

where $Q_r$ is group $r$'s proportion of the neighborhood population.  The maximum value of $E$, then, is the natural log of the number of groups - which would occur when all groups in a neighborhood are of equal size. Following [Hall and Lee (2010)](http://usj.sagepub.com/content/47/1/3.abstract), [Farrell and Lee (2011)](http://www.sciencedirect.com/science/article/pii/S0049089X11000706), and [Wright et al. (2014)](http://www.tandfonline.com/doi/abs/10.1080/00330124.2012.735924#.Vwxi7fkrLRY), $E$ is scaled by its maximum by dividing by $ln(4)$, setting the range of values from 0 to 1.

To study how neighborhood diversity varies with distance from urban cores in the largest metropolitan areas in the United States, entropy indices are plotted against the distance from the Census tract centroids to their corresponding nearest major city hall.  Locally-weighted regression (LOESS) is then used to produce a "diversity gradient" of estimates of neighborhood diversity by distance from the city center.

This application allows visitors to explore this part of the paper interactively.  The article follows by using local exploratory spatial data analysis techniques to identify how spatial clusters of diversity have shifted over time; this will be the focus of a future application that corresponds to an extension of the study published in _Urban Studies._

Demographic data for the paper come from [Brown University's Longitudinal Tract Database](http://www.s4.brown.edu/us2010/Researcher/LTDB.htm), and geographic data in the paper come from the [National Historical Geographic Information System](https://www.nhgis.org/).  Geographic data in the application are from the [US Census Bureau's Cartographic Boundary Files](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html) and subsequently simplified with the [rmapshaper](https://github.com/ateucher/rmapshaper) R package to improve performance.

The application is built with the [Shiny](http://shiny.rstudio.com) framework for the [R programming language](https://www.r-project.org/). The application layout is produced with the [flexdashboard](http://rstudio.github.io/flexdashboard/index.html) package, and the charts and maps use [Plotly](http://plot.ly), [Leaflet.js](http://leafletjs.com/), [Highcharts](http://www.highcharts.com/), and [ggplot2](http://ggplot2.org/), all accessed through their corresponding R packages.

Research code for the article is available upon request, and will be formally released this summer as a GitHub repository.  Code for the application is available at <https://github.com/walkerke/neighborhood_diversity>.

I welcome feedback and suggestions!  [Please visit my personal website](http://personal.tcu.edu/kylewalker/) for contact information or [connect with me on Twitter](https://twitter.com/kyle_e_walker).

<style>

.section.sidebar {

  background-color: white; 
  font-family: "Open-Sans", "Helvetica Neue", Helvetica, Arial, sans-serif;

}

.js-irs-0 .irs-bar {
border-top-color: #443A83;
border-bottom-color: #443A83;
}

.js-irs-0 .irs-bar-edge {
border-color: #443A83;
}

.js-irs-0 .irs-single, .js-irs-0 .irs-bar-edge, .js-irs-0 .irs-bar {
background: #443A83;
}

.navbar-inverse {
background-color: #443A83;
border-color: #440154;
}

.navbar-inverse .navbar-brand {
color: #a3a9ac;
}

a:hover, a:focus {
color: #440154;
text-decoration: underline;
}

a {
color: #443A83;
text-decoration: none;
}

.navbar-inverse .navbar-nav>li>a {
color: #a3a9ac;
}

</style>

### 2018-10-16 format_docs/draft_styles.Rmd
Reorganizing files and creating draft rmd to push to googledrive

---
title: "Word Reference Style"
author: "Malavika Rajeev"
date: "10/9/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2018-10-16 accessPEP.Rmd
Reorganizing files and creating draft rmd to push to googledrive

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
css: style.css
---

```{r directory, echo=FALSE, message = FALSE, results = 'hide'}
rm(list = ls())
require("knitr")
knitr::opts_chunk$set(fig.pos = 'h')
```

## INTRODUCTION

[Start broadly with inequity?]
To cite = Jess’s, Alegana, Emily from JHU.
Inequities in access to healthcare and interventions are a major driver of the burden of disease in human populations. These inequities can be driven by socioeconomics, lack of awareness and education, and by geographic accessibility. Despite the impact of access in shaping burden, incorporating these factors into estimations of burden is challenging due to limited data and lack of representation in data of the most marginalized groups (i.e. remote or low income populations).

<br>

Human rabies post-exposure prophylaxis prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic. Canine rabies causes an estimated 59,000 human deaths annually that could be prevented by PEP.

<br>

The majority of studies which estimate burden of rabies use probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space. However, these estimations… Given little data on true exposures of rabies how do we do this? Newly developed frameworks which rely on in-country data on key parameters likely better capture burden by disentangling rabid from non-rabid bites and generating incidence from estimates of the dog population and control in the dog population.

<br>

Here we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country with a largely absent control program in the canine population. Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to the patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use predictions of incidence from dynamic models and 
We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence.

## METHODS

### Bite patient data
We used the database of bite patient forms submitted from 27 ARMC between 2011-2014. These were paper forms that were sent to IPM as frequently as monthly to annually by individual clinics. Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database. These data include details on the location of the bite patient and biting animal (*cite baseline data paper*). We also had the number of doses delivered on an annual basis to each clinic and used this information to estimate submission rates of paper forms to IPM for each clinic as described in (*Moramanga paper + baseline data paper*, using minimum # of doses expected based on clinic throughput). As there were many discrepancies between commune names in the GIS administrative layer and the names written on the forms, we used fuzzy matching to find the closest matching commune names and manually checked these.

RIG data?

<br>

### GIS Data
We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then averaged these to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC. Population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level.

<br>

### Model of reporting as a function of travel time
We model the bites reported annually to the clinic ($B_{i,t}$) for each commune using a Bayesian state-space framework as follows:

$$ B_{i,t} =  k_{c,t} \times \rho_i \times(R_{i,t} + N_{i,t}) $$

$k_{c,t}$ is the annual reporting at the catchment level (the product of the estimated proportion of forms that were submitted and the proportion of patients for which we were able to match commune names for that year).

$p_i$ is the reporting of all bites which follows the logistic function:

$$ \rho_{i} = \frac{C}{1 + e^{\frac{A - x}{d}}}$$

where $C$ is a limiting function on the y-axis that measures probability of attendance when distance is zero, $X$ is the log transformed travel times in minutes, $d$ is the distance decay parameter, and A is an asymptote factor at an inflection point of the model. We fix C to 0.90 (estimated from the Moramanga data), as even at minimal travel times, there are people that do not report bites for other reasons than accessibility.

<br>

$N_{i,t}$ is the non-rabid bite incidence as a function of human population size **(Need to think about how to justify this...)**.

<br>

$$ N_{i,t} = Poisson(\beta_pP_i)$$

<br>

$R_{i,t}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

<br>

$$r \times D \times p_{exp}$$

<br>

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range of HDRs observed across Africa(cite?).

<br>

We use RJAGS (...) to fit this model to our reported bites at the commune level over the 4-year period using non-informative priors for $A$ $d$ and $\beta_p$

<br>

### Model evaluation and prediction
We fitted the model to data from n catchments which had above x% clinic level reporting {$k_c$} for all four years.

We used

For three other catchments with reporting above x% we used to out-of-fit predict...
Ability to predict average annual/monthly bite incidence by travel time
Ability to predict out-of-fit to other catchments 
(observed to expected mean incidence over 4 year period (monthly or annual)

<br>

### Sensitivity analyses
Sensitivity to R (@ what level of human exposure incidence!)
Simulate and recover params? @ what *true* incidence does p_i become non-identifiable...

<br>

### Exploring alternative scenarios and estimation of burden
We estimate burden of human rabies deaths as follows:

$$ D_i = p_{death}\times (1-\rho_i) \times R_i $$
where $p_{death}$ is the probability of death given an exposure in the absence of PEP, $\rho_i$ is the estimated commune specific reporting probability derived from the model, and $R_i$ is the expected rabies incidence. **We use Monte Carlo? simulation drawing from a dist of incidence to get an average number of expected deaths.** We do not include potential deaths due to imperfect PEP (i.e. not completing at least 3 doses), delays in adminsitration, or lack of RIG.

We consider three scenarios:
1. The baseline with the current clinic locations (n = 31)
2. Expansion of ARMC to one clinic per district (n = 114)
3. Expansion of ARMC to all CSB IIs (n = ...?)

We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3. We then predict the expected reporting rate from the model given these new travel times and compare the relative decreases in burden for the three scenarios.

<br>

### 2018-10-22 accessPEP.Rmd
preliminary data pipeline
summary for KH & JM Oct 22nd

@@
 
-```{r directory, echo=FALSE, message = FALSE, results = 'hide'}
+```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
 rm(list = ls())
@@
 knitr::opts_chunk$set(fig.pos = 'h')
+rm(list = ls())
+library(RImagePalette)
+library(jpeg)
+library(tidyverse)
+library(lubridate)
+library(magrittr)
+library(rgdal)
 ```
@@
 [Start broadly with inequity?]
-To cite = Jess’s, Alegana, Emily from JHU.
 Inequities in access to healthcare and interventions are a major driver of the burden of disease in human populations.
-These inequities can be driven by socioeconomics, lack of awareness and education, and by geographic accessibility.
-Despite the impact of access in shaping burden, incorporating these factors into estimations of burden is challenging due to limited data and lack of representation in data of the most marginalized groups (i.e.
-remote or low income populations).
+Those most impacted by infectious disease are often those from lower socioeconomic strata and remote populations.
+Despite the impact of these inequities in shaping disease burden, incorporating these factors into estimations of burden is challenging due to the lack of data resolved to these levels...>
 
@@
 
-Human rabies post-exposure prophylaxis prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
-Canine rabies causes an estimated 59,000 human deaths annually that could be prevented by PEP.
+Canine rabies causes an estimated 59,000 human deaths annually.
+Post-exposure prophylaxis prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
+Data on true rabies exposures in humans and incidence in animals is also lacking in most countries.
+Often, the only data available are numbers of bite victims reporting to health facilities.
 
@@
 
-The majority of studies which estimate burden of rabies use probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e.
+The majority of rabies burden studies use probability decision tree frameworks to estimate burden of rabies from these data, often with the key assumptions that overall reported bite incidence (i.e.
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space.
-However, these estimations… Given little data on true exposures of rabies how do we do this?
-Newly developed frameworks which rely on in-country data on key parameters likely better capture burden by disentangling rabid from non-rabid bites and generating incidence from estimates of the dog population and control in the dog population.
+
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence improves upon previous studies which may over estimate rabies burden in areas of high reporting and underestimating in areas with low reporting due to the assumption that overall reported bite incidence is proportional to rabies incidence.
 
@@
 
-Here we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country with a largely absent control program in the canine population.
+Here, we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country with a largely absent control program in the canine population.
 Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to the patients at 31 clinics across the country.
 PEP is not available at any other public clinics or through the private sector.
-Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use predictions of incidence from dynamic models and
-We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence.
+Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use predictions of incidence from the dynamic model developed by Hampson et al.
+2018 in a no control scenario.
+We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reporting in a bayesian state space framework...
 
@@
 where $C$ is a limiting function on the y-axis that measures probability of attendance when distance is zero, $X$ is the log transformed travel times in minutes, $d$ is the distance decay parameter, and A is an asymptote factor at an inflection point of the model.
-We fix C to 0.90 (estimated from the Moramanga data), as even at minimal travel times, there are people that do not report bites for other reasons than accessibility.
+We fix C to 0.90 (estimated from the Moramanga data), as even at minimal travel times, there are people that do not report bites for other reasons than accessibility (Alegana et al.
+2012).
 
@@
 
-We use RJAGS (...) to fit this model to our reported bites at the commune level over the 4-year period using non-informative priors for $A$ $d$ and $\beta_p$
+We use RJAGS (...) to fit this model to our reported bites at the commune level over the 4-year period using non-informative priors (...) for $A$ $d$ and $\beta_p$
 
@@
 <br>
+
+## Preliminary data processing
+
+### Getting catchments
+- Assigning by minimum average commune travel time to clinic (color corresponds to catchment)
+```{r catchments, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+ctar_metadata <- read.csv("data/ctar_metadata.csv")
+img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
+pal <- image_palette(img, 31)
+ctar_metadata$color <- pal
+
+mada_communes <- readOGR("data/MadaGIS/commune_mada.shp")
+catchments <- read.csv("output/catchments2.csv")
+
+left_join(catchments, dplyr::select(ctar_metadata, CTAR, color, id_ctar),
+            by = c("ctar" = "CTAR")) -> catchments
+mada_communes@data <- merge(mada_communes@data, catchments, by.x = "mdg_com_co",
+                            by.y = "comm_code")
+plot(mada_communes, col = as.character(mada_communes$color), border = NA)
+```
+**There are some issues with this at the moment, maybe could fix by weighted average of raster value travel times by population. This is sort of expensive computationally, so running on cluster now!
+
+## Submission of forms
+- Lack of consistent submission of forms, but we think that this for the most part has to do with gaps in sending of forms to IPM
+
+```{r doses, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
+
+ctar_metadata %>% 
+  dplyr::select(id_ctar, starts_with("doses")) %>%
+  gather(year, doses, -id_ctar) %>%
+  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> doses
+
+## getting dose time series by clinic
+ctar_data %>%
+  select(id_ctar, datej0, datej3, datej7, datej28) %>%
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
+  gather(dose, date, -id_ctar) %>%
+  group_by(id_ctar, date) %>%
+  summarise(n = n()) -> dose_ts
+dose_ts <- dose_ts[!is.na(dose_ts$date), ]
+dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
+dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]
+
+## getting theoretical # vials by year
+dose_ts %>%
+  mutate(year = year(date), vials = ceiling(n/2)) %>%
+  group_by(id_ctar, year) %>%
+  summarize(vials = sum(vials)) %>%
+  left_join(doses) -> dose_comp
+
+ggplot(data = dose_ts, aes(date, ctar)) + 
+  geom_tile(aes(fill = n)) +
+  scale_fill_continuous(type = "viridis", name = "Patients")
+
+```
+
+Potential solution: 
+- if monthly: exclude any months with greater than 14 days consecutively with no reporting if 
+- if annually: # days included from above/365 = approximate reporting on an annual basis?
+
+### Commune matching
+- For each district, ranked commune name matches by fuzzy matching.
+For those with partial match distances > 1 and < 10, manually chose matching commune names where possible.
+All others were set as NA.
+- Put this into reporting @ the clinic level (either monthly or annual prop matched)
+
+```{r commune matching, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+commune_matching <- read.csv("data/commune_check_CH.csv")
+mada_communes$match <- tolower(str_replace_all(mada_communes$commune,"[^[:graph:]]", "/"))
+mada_communes$district <- substr(as.character(mada_communes$mdg_com_co), 1, 8)
+
+mada_communes@data %>%
+  select(match, district, mdg_com_co) %>%
+  left_join(commune_matching, .) %>%
+  select(name, match, district, mdg_com_co) -> comm_matches
+
+ctar_data$name <- tolower(str_replace_all(ctar_data$commune,"[^[:graph:]]", "/"))
+left_join(ctar_data, comm_matches) %>%
+  distinct(record_id, .keep_all = TRUE) -> ctar_data
+
+ctar_data %>%
+  group_by(id_ctar) %>%
+  summarise(n = n(), unmatched = sum(is.na(mdg_com_co)),
+            prop_matched = unmatched/n) -> ctar_data_matched
+
+ctar_metadata <- merge(ctar_metadata, ctar_data_matched)
+ctar_metadata <- ctar_metadata[!is.na(ctar_metadata$id_ctar), ]
+ggplot(ctar_metadata, aes(CTAR, prop_matched)) +
+  geom_point() +
+  coord_flip()  +
+  ylab(label = "Proportion of patients matched to commune")
+
+```
+
+- Not too great of a job for most of them! Ask Chantal to go back through and match more?
+
+## Overall plan
+- Pick best catchments (Toamasina, Ambatodrajaka, Toliara, Manakara, Sambava, Soanierana Ivongo, Moramanga, Anstirabe, Ambositra)
+- See if anything comes out even with limited success of commune matching
+
+- Alternatively = by district? Less likely to really capture spatial het in access
+- Alternatively = w/ Moramanga data from 2016-2018 only; only one catchment, and where we've been actively working/contact tracing

### 2018-12-03 accessPEP.Rmd
Changing to simpler model and cleaning data using dplyr

--- 2018-10-16 accessPEP.Rmd
@@
 output:
-  word_document:
-    reference_docx: format_docs/word_styles_reference_01.docx
+  html_document: default
   pdf_document:
@@
       in_header: mystyle.sty
-  html_document: default
+  word_document:
+    reference_docx: format_docs/word_styles_reference_01.docx
 css: style.css
@@
 
-```{r directory, echo=FALSE, message = FALSE, results = 'hide'}
+```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
 rm(list = ls())
-require("knitr")
-knitr::opts_chunk$set(fig.pos = 'h')
+library(knitr)
+library(RImagePalette)
+library(jpeg)
+library(tidyverse)
+library(lubridate)
+library(magrittr)
+library(rgdal)
+library(broom)
+knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
+
 ```
@@
 
-Human rabies post-exposure prophylaxis prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
-Canine rabies causes an estimated 59,000 human deaths annually that could be prevented by PEP.
+Canine rabies causes an estimated 59,000 human deaths annually.
+Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
+Data on true rabies exposures in humans and incidence in animals is also lacking in most countries.
+Often, the only data available are numbers of bite victims reporting to health facilities.
+The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space.
+
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may overestimate rabies burden in areas of high reporting and underestimate in areas with low reporting due to the assumption that overall reported bite incidence is proportional to rabies incidence.
 
@@
 
-The majority of studies which estimate burden of rabies use probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space.
-However, these estimations… Given little data on true exposures of rabies how do we do this?
-Newly developed frameworks which rely on in-country data on key parameters likely better capture burden by disentangling rabid from non-rabid bites and generating incidence from estimates of the dog population and control in the dog population.
-
-<br>
-
-Here we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country with a largely absent control program in the canine population.
-Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to the patients at 31 clinics across the country.
+Here, we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country that lacks a national strategy for controlling rabies in domestic dogs.
+Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country.
 PEP is not available at any other public clinics or through the private sector.
-Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use predictions of incidence from dynamic models and
-We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence.
+Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al.
+2018.
+We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.
 
@@
 ### Bite patient data
-We used the database of bite patient forms submitted from 27 ARMC between 2011-2014.
-These were paper forms that were sent to IPM as frequently as monthly to annually by individual clinics.
+We used the database of bite patient forms submitted from 24 ARMC between 2014 - 2017.
+These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics.
 Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
-These data include details on the location of the bite patient and biting animal (*cite baseline data paper*).
-We also had the number of doses delivered on an annual basis to each clinic and used this information to estimate submission rates of paper forms to IPM for each clinic as described in (*Moramanga paper + baseline data paper*, using minimum # of doses expected based on clinic throughput).
-As there were many discrepancies between commune names in the GIS administrative layer and the names written on the forms, we used fuzzy matching to find the closest matching commune names and manually checked these.
+These data include the administrative district of the bite patient's address.
+As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted.
+We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.
+Figure 1a + 1b?
 
@@
 2012) and also aggreagated to the commune level.
+Figure 2 = networky fig + catchments!
 
@@
 
-### Model of reporting as a function of travel time
-We model the bites reported annually to the clinic ($B_{i,t}$) for each commune using a Bayesian state-space framework as follows:
+### Model of reported bites as a function of travel time
+Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.
 
-$$ B_{i,t} =  k_{c,t} \times \rho_i \times(R_{i,t} + N_{i,t}) $$
+We fit a linear mixed effects model where the annual average bite incidence at the district level was predicted by the average travel time to the nearest CTAR for the district, the population within the district, and a random catchment level effect.
 
-$k_{c,t}$ is the annual reporting at the catchment level (the product of the estimated proportion of forms that were submitted and the proportion of patients for which we were able to match commune names for that year).
-
-$p_i$ is the reporting of all bites which follows the logistic function:
-
-$$ \rho_{i} = \frac{C}{1 + e^{\frac{A - x}{d}}}$$
-
-where $C$ is a limiting function on the y-axis that measures probability of attendance when distance is zero, $X$ is the log transformed travel times in minutes, $d$ is the distance decay parameter, and A is an asymptote factor at an inflection point of the model.
-We fix C to 0.90 (estimated from the Moramanga data), as even at minimal travel times, there are people that do not report bites for other reasons than accessibility.
+For three other catchments with reporting above x% we used to out-of-fit predict...
+Ability to predict average annual/monthly bite incidence by travel time
+Ability to predict out-of-fit to other catchments
 
@@
 
-$N_{i,t}$ is the non-rabid bite incidence as a function of human population size **(Need to think about how to justify this...)**.
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
 
+We calculated deaths as:
+$$deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}$$
+
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
+            \frac{R_i}{B_i}, & \text{otherwise}
+            \end{cases}
+$$
 <br>
 
-$$ N_{i,t} = Poisson(\beta_pP_i)$$
+such that rabid reported bites (i.e.
+$p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_{i} \times \rho_{max}$.
+$p_{rabid}$ is taken from Rajeev et al.
+2018 on contact tracing data from the Moramanga District.
 
-<br>
-
-$R_{i,t}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
+$R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
 
@@
 We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range of HDRs observed across Africa(cite?).
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa(cite?).
 
@@
 
-We use RJAGS (...) to fit this model to our reported bites at the commune level over the 4-year period using non-informative priors for $A$ $d$ and $\beta_p$
-
-<br>
-
-### Model evaluation and prediction
-We fitted the model to data from n catchments which had above x% clinic level reporting {$k_c$} for all four years.
-
-We used
-
-For three other catchments with reporting above x% we used to out-of-fit predict...
-Ability to predict average annual/monthly bite incidence by travel time
-Ability to predict out-of-fit to other catchments 
-(observed to expected mean incidence over 4 year period (monthly or annual)
-
-<br>
-
-### Sensitivity analyses
-Sensitivity to R (@ what level of human exposure incidence!)
-Simulate and recover params? @ what *true* incidence does p_i become non-identifiable...
-
-<br>
-
-### Exploring alternative scenarios and estimation of burden
-We estimate burden of human rabies deaths as follows:
-
-$$ D_i = p_{death}\times (1-\rho_i) \times R_i $$
-where $p_{death}$ is the probability of death given an exposure in the absence of PEP, $\rho_i$ is the estimated commune specific reporting probability derived from the model, and $R_i$ is the expected rabies incidence.
-**We use Monte Carlo?
-simulation drawing from a dist of incidence to get an average number of expected deaths.** We do not include potential deaths due to imperfect PEP (i.e.
-not completing at least 3 doses), delays in adminsitration, or lack of RIG.
-
-We consider three scenarios:
+## Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
 1. The baseline with the current clinic locations (n = 31)
@@
 We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected reporting rate from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
 
 <br>
+
+### Sensitivity Analyses
+- Sensitivity to assumptions of R
+- Sensitivity to assumptions of p_rabid
+
+## Preliminary data processing
+
+### Cleaning data
+```{r first cleaning}
+## read in data
+ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
+
+## percent with wrong dates by ctar
+
+## percent with missing district by ctar
+
+## exclude contacts
+
+```
+
+## Submission of forms
+- Missingness seems to have been due to periods of time without form submission
+- Calculating periods of time
+
+```{r doses}
+ctar_metadata <- read.csv("data/ctar_metadata.csv")
+mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
+
+img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
+pal <- image_palette(img, 31)
+ctar_metadata$color <- pal
+
+ctar_metadata %>% 
+  dplyr::select(id_ctar, starts_with("doses")) %>%
+  gather(year, doses, -id_ctar) %>%
+  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> doses
+
+## getting dose time series by clinic
+ctar_data %>%
+  select(id_ctar, datej0, datej3, datej7, datej28) %>%
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
+  gather(dose, date, -id_ctar) %>%
+  group_by(id_ctar, date) %>%
+  summarise(n = n()) -> dose_ts
+dose_ts <- dose_ts[!is.na(dose_ts$date), ]
+dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
+dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]
+
+## getting theoretical # vials by year
+dose_ts %>%
+  mutate(year = year(date), vials = ceiling(n/2)) %>%
+  group_by(id_ctar, year) %>%
+  summarize(vials = sum(vials)) %>%
+  left_join(doses) -> dose_comp
+
+ggplot(data = dose_ts, aes(date, ctar)) + 
+  geom_tile(aes(fill = n)) +
+  scale_fill_continuous(type = "viridis", name = "Patients")
+
+## getting reporting
+start_date <- ymd("2014-01-01")
+end_date <- ymd("2017-12-31")
+ts <- as_tibble(seq(start_date, end_date, by = "day"))
+
+dose_ts %>%
+  right_join(ts, by = c("date" = "value")) %>%
+  select(id_ctar, date, n) %>%
+  spread(id_ctar, n) %>%
+  replace(., is.na(.), 0) -> doses_wide
+dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])
+
+date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
+for (j in 1:ncol(date_mat)){
+  rle(dose_mat[ , j]) %>%
+    unclass() %>%
+    as.data.frame() %>%
+    mutate(end = cumsum(lengths),
+         start = c(1, lag(end)[-1] + 1)) %>%
+    filter(values == 0, lengths >= 10) -> rles
+  for (i in 1:nrow(rles)){
+    date_mat[rles$start[i]:rles$end[i], j] <- 0
+  }
+}
+
+date_mat <- replace(date_mat, is.na(date_mat), 1)
+date_mat %>% 
+  as_tibble() %>%
+  group_by(date = year(doses_wide$date)) %>%
+  summarise_all(funs(sum(.)/365)) -> clinic_reporting
+colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
+clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date)
+clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
+                                                  ctar_metadata$id_ctar)]
+
+ggplot(clinic_reporting, aes(x = date, y = ctar, fill = prop)) + 
+  geom_tile() +
+  scale_fill_gradient(low = "white", high = "blue")
+
+to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
+to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
+ctar_data$excl <- 0
+ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1
+
+ctar_data %>% 
+  filter(excl != 1) %>% # exclude contacts
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
+  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
+  summarize(n = n()) -> exposures # get count of exposures
+
+## districts that are non-reporter and data not available yet CTAR
+no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
+             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
+             "Marolambo", "Sainte Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano",
+             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
+             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
+             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
+no_data <- mada_district$mdg_dis_co[match(no_data, mada_district$district)]
+mada_district@data %>%
+  select(district = mdg_dis_co, pop = pop2015adj, ttimes = study_area) -> covars
+
+exposures %>%
+  rename(date = year) %>%
+  mutate(id_ctar = as.character(id_ctar)) %>%
+  left_join(clinic_reporting) %>%   # add reporting by clinic + year
+  filter(!(district %in% no_data), prop >= 0.25) %>% # exclude no data & < 25% annual reporting
+  group_by(district) %>%
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites)) %>% # average by ctar + district 
+  left_join(covars) %>% # add pop + ttimes + reporting
+  mutate(incidence = bites/(pop/1e5)) -> exps_cleaned 
+exps_cleaned %$% plot(ttimes, bites)
+exps_cleaned %$% plot(ttimes, incidence, bty = 
+                        "n", pch = 20)
+
+mada_dist_gg <- tidy(mada_district, region = "mdg_dis_co")
+mada_dist_gg <- left_join(mada_dist_gg, exps_cleaned, by = c("id" = "district"))
+
+ggplot() +
+  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = incidence)) +
+  scale_fill_continuous(type = "viridis") +
+  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
+ggplot() +
+  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes)) +
+  scale_fill_continuous(type = "viridis") +
+  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
+
+library(lme4)
+summary(lm(log(incidence) ~ ttimes, data = exps_cleaned))
+```

### 2018-12-04 accessPEP.Rmd
dist models!

@@
 ---
-
-```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
-rm(list = ls())
-library(knitr)
-library(RImagePalette)
-library(jpeg)
-library(tidyverse)
-library(lubridate)
-library(magrittr)
-library(rgdal)
-library(broom)
-knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
-
-```
 
@@
 ## Preliminary data processing
-
-### Cleaning data
-```{r first cleaning}
-## read in data
-ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
-
-## percent with wrong dates by ctar
-
-## percent with missing district by ctar
-
-## exclude contacts
+```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+rm(list = ls())
+library(knitr)
+library(RImagePalette)
+library(jpeg)
+library(tidyverse)
+library(lubridate)
+library(magrittr)
+library(rgdal)
+library(broom)
+knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
 
@@
 
-## Submission of forms
-- Missingness seems to have been due to periods of time without form submission
-- Calculating periods of time
-
-```{r doses}
+### Data
+```{r get data}
+## read in data
+ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
 ctar_metadata <- read.csv("data/ctar_metadata.csv")
 mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
+dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
+names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
 
+## Centers with no data
+no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
+             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
+             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
+ctar_metadata$exclude <- 0
+ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1
+
+## Get color palette from other network fig!
 img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
 pal <- image_palette(img, 31)
-ctar_metadata$color <- pal
+add.alpha <- function(col, alpha=1){
+  if(missing(col))
+    stop("Please provide a vector of colours.")
+  apply(sapply(col, col2rgb)/255, 2, 
+                     function(x) 
+                       rgb(x[1], x[2], x[3], alpha=alpha))  
+}
+ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
+                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
+                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
+                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
+ctar_metadata$fill <- ctar_metadata$color
+ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
+ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
+ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
+                                                           mada_district$district)]
+display_palette(ctar_metadata$color)
+dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
+                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
+mada_district@data <- left_join(mada_district@data, dist_catchments)
+```
 
+```{r catchment figure}
+## get exposure matrix
+ctar_data %>% 
+  group_by(district, id_ctar) %>% # group by district and ctar
+  summarize(n = n()) -> exp_mat # get count of exposures
+dist_points <- cbind(select(mada_district@data,
+                            mdg_dis_co), coordinates(mada_district))
+names(dist_points)[2:3] <- c("dist_long", "dist_lat")
+dist_points %>%
+ left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
+           by = c("mdg_dis_co" = "ctar_dist")) %>%
+ mutate(dist_long = coalesce(dist_long, LONGITUDE), 
+        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points
+
+exp_mat %>% 
+  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
+            by = c("district" = "mdg_dis_co")) %>%
+  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
+                   color, fill, ctar_dist)) -> dist_lines
+dist_lines$size <- log(dist_lines$n + 0.1)*0.2
+
+ctar_data %>%
+  group_by(district) %>%
+  summarize(n = n()) %>%
+  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
+             by = c("district" = "mdg_dis_co")) %>%
+  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
+dist_points$size <- log(dist_points$n + 0.1)*0.25
+
+pdf("figs/network.pdf")
+plot(mada_district, col = mada_district$fill, border = "grey") 
+points(dist_points$dist_long, dist_points$dist_lat, pch = 20, cex = dist_points$size, 
+       col = dist_points$color)
+segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat, 
+         lwd = dist_lines$size*2, col = dist_lines$color)
+dev.off()
+
+legend("topleft", c(">100", ">200", ">400", ">800"), title = "Number of total reported bites", 
+       text.width = 3, pt.cex = circle_size, col = "grey",  pch = 16, inset = c(0.1, 0.1), lwd = 1.5, bty = "n", xpd = TRUE, y.intersp=2)
+
+exp_mat %>% 
+  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
+  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
+  group_by(district) %>%
+  mutate(prop = n/sum(n),
+         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
+ggplot(data = catch_check, aes(CTAR.x, district)) + 
+  geom_tile(aes(fill = prop)) +
+  scale_fill_continuous(type = "viridis", name = "proportion") +
+  geom_text(label = catch_check$catch_match)
+```
+
+```{r}
+## get vials delivered
 ctar_metadata %>% 
   dplyr::select(id_ctar, starts_with("doses")) %>%
-  gather(year, doses, -id_ctar) %>%
-  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> doses
+  gather(year, vials_delivered, -id_ctar) %>%
+  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered
 
 ## getting dose time series by clinic
+########### Need to fix dates first! ###############
 ctar_data %>%
@@
   group_by(id_ctar, year) %>%
-  summarize(vials = sum(vials)) %>%
-  left_join(doses) -> dose_comp
+  summarize(vials_estimated = sum(vials)) %>%
+  left_join(vials_delivered) -> dose_comp
 
-ggplot(data = dose_ts, aes(date, ctar)) + 
+######### Patient time series
+ctar_data %>%
+  select(id_ctar, date_de_consultation) %>%
+  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
+  gather(dose, date_de_consultation, -id_ctar) %>%
+  group_by(id_ctar, date_de_consultation) %>%
+  summarise(n = n()) -> patient_ts
+patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
+patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
+patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]
+
+ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
   geom_tile(aes(fill = n)) +
@@
 
-dose_ts %>%
-  right_join(ts, by = c("date" = "value")) %>%
-  select(id_ctar, date, n) %>%
+patient_ts %>%
+  right_join(ts, by = c("date_de_consultation" = "value")) %>%
+  select(id_ctar, date_de_consultation, n) %>%
   spread(id_ctar, n) %>%
@@
   as_tibble() %>%
-  group_by(date = year(doses_wide$date)) %>%
-  summarise_all(funs(sum(.)/365)) -> clinic_reporting
+  group_by(date = year(doses_wide$date_de_consultation)) %>%
+  summarise_all(funs(sum(.)/n())) -> clinic_reporting
 colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
-clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date)
+clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
 clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                   ctar_metadata$id_ctar)]
+clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
 
-ggplot(clinic_reporting, aes(x = date, y = ctar, fill = prop)) + 
+ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
   geom_tile() +
   scale_fill_gradient(low = "white", high = "blue")
+```
+
+## Submission of forms
+- Missingness seems to have been due to periods of time without form submission
+- Calculating periods of time
+
+```{r clean data}
+## districts that are non-reporter and data not available yet CTAR
+mada_district@data %>%
+select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
 
@@
 
-## districts that are non-reporter and data not available yet CTAR
-no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
-             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
-             "Marolambo", "Sainte Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano",
-             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
-             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
-             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
-no_data <- mada_district$mdg_dis_co[match(no_data, mada_district$district)]
-mada_district@data %>%
-  select(district = mdg_dis_co, pop = pop2015adj, ttimes = study_area) -> covars
-
 exposures %>%
   rename(date = year) %>%
-  mutate(id_ctar = as.character(id_ctar)) %>%
+  left_join(covars) %>% # add pop + ttimes + reporting
   left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(!(district %in% no_data), prop >= 0.25) %>% # exclude no data & < 25% annual reporting
+  filter(exclude == 0, prop > 0.25) %>%
+  group_by(district, date) %>%
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
+exps_cleaned %>%
   group_by(district) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites)) %>% # average by ctar + district 
-  left_join(covars) %>% # add pop + ttimes + reporting
-  mutate(incidence = bites/(pop/1e5)) -> exps_cleaned 
-exps_cleaned %$% plot(ttimes, bites)
-exps_cleaned %$% plot(ttimes, incidence, bty = 
+  summarise(bites = mean(bites)) %>%
+  left_join(covars) -> exps_cleaned
+
+library(lme4)
+
+## poisson with total pop as offset
+mod <- glm(round(bites) ~ ttimes_weighted + pop + offset(log(round(pop))), 
+                 data = exps_cleaned, family = poisson)
+plot(exps_cleaned$bites, predict(mod, type = "response"))
+plot(exps_cleaned$ttimes_weighted, predict(mod, type = "response"))
+
+```
+
+```{r}
+p_rabid = 0.50
+mean_bites <- predict(mod, type = "response", newdata = covars)
+bites <- rpois(length(mean_bites), mean_bites)
+mean_exps <- runif(length(covars$pop), min = 0.39*0.01*exps_cleaned$pop/25, 
+                    max = 0.39*0.01*exps_cleaned$pop/5)
+rabid_exps <- rpois(length(mean_bites), mean_exps)
+p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > 0.95, (0.95*rabid_exps)/bites, p_rabid)
+deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), 0.16)
+rho <- bites*p_rabid_t/rabid_exps
+plot(covars$ttimes_weighted, rho)
+
+```
+
+```{r}
+exps_cleaned %$% plot(ttimes_weighted, bites)
+
+exps_cleaned %>%
+  filter(exclude == 0) %$% 
+  plot(ttimes_weighted, incidence, bty = 
                         "n", pch = 20)
@@
 
+quartz()
 ggplot() +
-  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = incidence)) +
-  scale_fill_continuous(type = "viridis") +
+  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = CTAR)) +
+  scale_colour_viridis_d()
+
 geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
 ggplot() +
-  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes)) +
+  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes_weighted)) +
   scale_fill_continuous(type = "viridis") +
@@
 
-library(lme4)
-summary(lm(log(incidence) ~ ttimes, data = exps_cleaned))
 ```

### 2018-12-04 accessPEP.Rmd
district models

@@
 ## poisson with total pop as offset
-mod <- glm(round(bites) ~ ttimes_weighted + pop + offset(log(round(pop))), 
+mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
                  data = exps_cleaned, family = poisson)
@@
 ```{r}
-p_rabid = 0.50
-mean_bites <- predict(mod, type = "response", newdata = covars)
+get.burden <- function(covs = covars, pop = covars$pop, model = mod,
+                       p_rabid = 0.5, rho_max = 0.9,
+                       max_HDR = 25, min_HDR = 5, 
+                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                       prob_death = 0.16, nsims = 1000) {
+  store_deaths <- store_p_rabid <- matrix(NA, nrow(covs), nsims)
+
+  for (i in 1:nsims) {
+    mean_bites <- predict(model, type = "response", newdata = covs)
     bites <- rpois(length(mean_bites), mean_bites)
-mean_exps <- runif(length(covars$pop), min = 0.39*0.01*exps_cleaned$pop/25, 
-                    max = 0.39*0.01*exps_cleaned$pop/5)
+    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
+                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
     rabid_exps <- rpois(length(mean_bites), mean_exps)
-p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > 0.95, (0.95*rabid_exps)/bites, p_rabid)
-deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), 0.16)
-rho <- bites*p_rabid_t/rabid_exps
-plot(covars$ttimes_weighted, rho)
+    p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid)
+    deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), prob_death)
+    store_deaths[, i] <- deaths
+    store_p_rabid[, i] <- p_rabid_t
+  }
+  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
+  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
+                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
+  results <- cbind(covars, results)
+  return(results)
+}
 
+get.burden.flat <- function(covs = covars, pop = covars$pop, model = mod,
+                       p_rabid = 0.5, rho = 0.85, 
+                       prob_death = 0.16, nsims = 1000) {
+  store_deaths <- matrix(NA, nrow(covs), nsims)
+
+  for (i in 1:nsims) {
+    mean_bites <- predict(model, type = "response", newdata = covs)
+    bites <- rpois(length(mean_bites), mean_bites)
+    unreported <- bites*p_rabid/rho - bites*p_rabid
+    deaths <- rbinom(length(mean_bites), round(unreported), prob_death)
+    store_deaths[, i] <- deaths
+  }
+  results <- get.meanCI(store_deaths)
+  colnames(results) <- c("deaths_mean", "death_lowerCI", "deaths_upperCI")
+  results <- cbind(covars, results)
+  return(results)
+}
+
+get.meanCI <- function(matrix_by_row, nboots = 1000) {
+    mean_rows <- apply(matrix_by_row, 1, mean)
+    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
+    lower <- unlist(CIs)[1, ] 
+    upper <- unlist(CIs)[2, ]
+    summ <- cbind(mean_rows, lower, upper)
+    colnames(summ) <- c("mean", "lower", "upper")
+    return(summ)
+}
+
+boots <- function(x, nboot = 1000){
+  n = length(x)
+  xbar = mean(x)
+  tmpdata = sample(x,n*nboot, replace=TRUE)
+  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
+  bsmeans = colMeans(bootstrapsample)
+  deltastar = bsmeans - xbar
+  d = quantile(deltastar, c(0.05, 0.95))
+  ## Get from quantiles
+  ci = xbar - c(d[2], d[1])
+
+  ## Sorted CIs
+  sorteddeltastar = sort(deltastar)
+  # # Look at the sorted results
+  # Find the .05 and .95 critical values of deltastar
+  d9alt = sorteddeltastar[5]
+  d1alt = sorteddeltastar[950]
+  # Find and print the 80% confidence interval for the mean
+  ciAlt = xbar - c(d1alt,d9alt)
+  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
+  return(ci)
+}
+
+```
+
+```{r}
+results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
+results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
+
+plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
+points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
+segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
+         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
+
+results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.90)
+results_pr60 <- get.burden(p_rabid = 0.20, rho_max = 0.99)
+
+plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
+points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
+segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
+         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
+
+results_pr20 <- get.burden(p_rabid = 0.60, rho_max = 0.90)
+results_pr60 <- get.burden(p_rabid = 0.60, rho_max = 0.99)
+
+plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
+points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
+segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
+         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
 ```
@@
 
-quartz()
 ggplot() +

### 2018-12-18 accessPEP.Rmd
adding in commune level catchments etc

@@
 library(broom)
+select <- dplyr::select
 knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
@@
 ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
+## Getting ctar district and commune
 ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                            mada_district$district)]
+pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
+                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
+
 display_palette(ctar_metadata$color)
@@
 mada_district@data <- left_join(mada_district@data, dist_catchments)
+dist_catchments$ctar_in <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
 ```
@@
 
-pdf("figs/network.pdf")
+pdf("figs/network_1.pdf")
 plot(mada_district, col = mada_district$fill, border = "grey") 
-points(dist_points$dist_long, dist_points$dist_lat, pch = 20, cex = dist_points$size, 
-       col = dist_points$color)
+points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
+       , pch = 1, cex = dist_points$size, 
+       col = dist_lines$color)
 segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
@@
          catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
+hist(catch_check$prop)
 ggplot(data = catch_check, aes(CTAR.x, district)) + 
@@
   geom_text(label = catch_check$catch_match)
+
 ```
@@
   scale_fill_continuous(type = "viridis", name = "Patients")
+
+patient_ts %>%
+  filter(id_ctar == 1) %$%
+  hist(n, breaks = 100, col = "blue", 
+       xlab = "Patients reporting", main = "Soanierana Ivongo")
+patient_ts %>%
+  filter(ctar == "Toamasina") %$%
+  hist(n, breaks = 100, 
+       xlab = "Patients reporting", main = "Toamasina", col = "red")
 
@@
                  data = exps_cleaned, family = poisson)
-plot(exps_cleaned$bites, predict(mod, type = "response"))
-plot(exps_cleaned$ttimes_weighted, predict(mod, type = "response"))
-
+par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
+plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
+     xlab = "Observed bites", ylab = "Predicted bites")
+abline(a = 0, b = 1, lty = 2, col = "blue")
+legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
+       inset = c(0.0, -0.1))
+plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
+     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
+     ylim = c(0, max(exps_cleaned$bites)))
+points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
+     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
+legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
+       pch = 20, bty = "n")
 ```
 
-```{r}
+```{r functions for getting burden}
 get.burden <- function(covs = covars, pop = covars$pop, model = mod,
@@
   }
+
   results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
@@
 plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
-points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
+points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 20)
 segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
@@
 
-results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.90)
-results_pr60 <- get.burden(p_rabid = 0.20, rho_max = 0.99)
+results_base <- get.burden(p_rabid = 0.40, rho_max = 0.98)
 
-plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
-points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
-segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
-         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
+p_rabid_check <- seq(0.10, 0.80, by = 0.1)
+check <- data.frame(p_rabid = 0, district = NA, ttimes = 0, deaths_mean = 0)
+for (i in 1:length(p_rabid_check)) {
+  print(paste(i, "/", length(p_rabid_check)))
+  results <- get.burden(p_rabid = p_rabid_check[i], rho_max = 0.98)
+  p_rabid <- rep(p_rabid_check[i], length(results$deaths_mean))
+  check_one <- cbind(p_rabid, results$district, results$ttimes, results$deaths_mean)
+  colnames(check_one) <- c("p_rabid", "district", "ttimes", "deaths_mean")
+  check <- rbind(check, check_one)
+}
 
-results_pr20 <- get.burden(p_rabid = 0.60, rho_max = 0.90)
-results_pr60 <- get.burden(p_rabid = 0.60, rho_max = 0.99)
+check <- check[-1, ]
+check$base <- rep(results_base$deaths_mean, length(p_rabid_check))
+ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
+  geom_tile(aes(fill = deaths_mean/base)) +
+  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 1, name = "Mean deaths")
 
-plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
-points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 1)
-segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
-         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
-```
+rho_max_check <- seq(0.90, 0.99, by = 0.01)
+check <- data.frame(rho_max = 0, district = NA, ttimes = 0, deaths_mean = 0)
+for (i in 1:length(rho_max_check)) {
+  print(paste(i, "/", length(rho_max_check)))
+  results <- get.burden(p_rabid = 0.4, rho_max = rho_max_check[i])
+  rho_max <- rep(rho_max_check[i], length(results$deaths_mean))
+  check_one <- cbind(rho_max, results$district, results$ttimes, results$deaths_mean)
+  colnames(check_one) <- c("rho_max", "district", "ttimes", "deaths_mean")
+  check <- rbind(check, check_one)
+}
 
-```{r}
-exps_cleaned %$% plot(ttimes_weighted, bites)
+check <- check[-1, ]
+check$base <- rep(results_base$deaths_mean, length(rho_max_check))
+check %>%
+  group_by(district) %>%
+  mutate(min = min(deaths_mean), max = max(deaths_mean), 
+         standardized = (deaths_mean - base)/(max - base)) -> check
 
-exps_cleaned %>%
-  filter(exclude == 0) %$% 
-  plot(ttimes_weighted, incidence, bty = 
-                        "n", pch = 20)
+ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
+  geom_tile(aes(fill = standardized)) +
+  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths") 
+ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
+  geom_tile(aes(fill = deaths_mean)) +
+  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths")
 
@@
 
-geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
-ggplot() +
-  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes_weighted)) +
-  scale_fill_continuous(type = "viridis") +
-  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
+```
 
+## Commune level mods
+```{r commune matching, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+commune_matching <- read.csv("data/commune_check_CH.csv")
+mada_communes <- readOGR("output/communes/communes_extracted.shp")
+mada_communes$match <- tolower(str_replace_all(mada_communes$commune,"[^[:graph:]]", "/"))
+
+mada_communes@data %>%
+  select(match, mtch_ds, mdg_cm_) %>%
+  left_join(commune_matching, .) %>%
+  select(name, match, district, mdg_cm_) -> comm_matches
+
+ctar_data$name <- tolower(str_replace_all(ctar_data$commune,"[^[:graph:]]", "/"))
+left_join(ctar_data, comm_matches) %>%
+  distinct(record_id, .keep_all = TRUE) -> ctar_data
+
+## Looking at proportion matched
+ctar_data %>%
+  group_by(id_ctar) %>%
+  summarise(n = n(), unmatched = sum(is.na(mdg_cm_)),
+            prop_matched = unmatched/n) -> ctar_data_matched
+
+ctar_metadata <- merge(ctar_metadata, ctar_data_matched)
+ctar_metadata <- ctar_metadata[!is.na(ctar_metadata$id_ctar), ]
+ggplot(ctar_metadata, aes(CTAR, prop_matched)) +
+  geom_point() +
+  coord_flip()  +
+  ylab(label = "Proportion of patients matched to commune")
 ```
+
+```{r }
+## Ctar data at commune level
+ctar_data %>% 
+  filter(excl != 1) %>% # exclude contacts
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
+  group_by(year = year(date_de_consultation), commune = mdg_cm_, id_ctar) %>% # group by year and district
+  summarize(n = n()) -> exposures # get count of exposures
+
+comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
+names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")
+comm_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
+                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> comm_catchments
+mada_communes@data <- left_join(mada_communes@data, comm_catchments)
+
+mada_communes@data %>%
+select(commune = mdg_cm_, pop = MDG__201, ttimes_weighted, CTAR, exclude) -> covars
+mada_communes$ctar_in <- ifelse(mada_communes$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
+
+## Catchments with no data
+no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
+             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
+             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
+covars$exclude <- 0
+covars$exclude[covars$CTAR %in% no_data] <- 1
+
+exposures %>%
+  rename(date = year) %>%
+  left_join(covars) %>% # add pop + ttimes + reporting
+  left_join(clinic_reporting) %>%   # add reporting by clinic + year
+  filter(exclude == 0, prop > 0.25) %>%
+  group_by(commune, date) %>%
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
+exps_cleaned %>%
+  group_by(commune) %>%
+  summarise(bites = mean(bites)) %>%
+  left_join(covars) -> exps_cleaned
+
+library(lme4)
+
+## poisson with total pop as offset
+exps_cleaned <- filter(exps_cleaned, ttimes_weighted != "Inf")
+mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
+                 data = exps_cleaned, family = poisson)
+
+par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
+plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
+     xlab = "Observed bites", ylab = "Predicted bites")
+abline(a = 0, b = 1, lty = 2, col = "blue")
+legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
+       inset = c(0.0, -0.1))
+plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
+     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
+     ylim = c(0, max(exps_cleaned$bites)))
+points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
+     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
+legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
+       pch = 20, bty = "n")
+
+ctar_data %>% 
+  filter(excl != 1) %>% # exclude contacts
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
+  group_by(commune = mdg_cm_, id_ctar) %>% # group by year and district
+  summarize(n = n()) ->  exp_mat # get count of exposures
+
+comm_catchments$exclude <- 0
+comm_catchments$exclude[comm_catchments$CTAR %in% no_data] <- 1
+
+exp_mat %>% 
+  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
+  left_join(select(comm_catchments, mdg_cm_, CTAR, exclude), by = c("commune" = "mdg_cm_")) %>%
+  group_by(commune) %>%
+  mutate(prop = n/sum(n),
+         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
+hist(catch_check$prop[catch_check$exclude == 0])
+hist(catch_check$prop[catch_check$exclude == 1])
+
+## getting var for ctar in
+pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
+                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+over(pts, mada_communes)$mtch_ds
+```

### 2018-12-18 accessPEP.Rmd
with old code + likelihood code

@@
 ---
+## Goals
+- To estimate burden of rabies in Madagascar using nationally available data from clinics that provide PEP.
+- To explore the potential impacts of expanded access to PEP
 
-## INTRODUCTION
+## Data
+- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
+  - includes dates and home locations (to the district level) of patient reporting for PEP
+- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
+  - includes dates and home locations (to the commune level) of patients reporting for PEP
+  - also includes rabies exposure status based on a combination of clinic-based triage and contact
+  tracing (Rajeev et al. 2018)
+- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
+  - We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
+2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
+  - We then calculated a weighted average by human population to the commune or district level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
+  - For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
+- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the designated admin level.
 
-[Start broadly with inequity?]
-Inequities in access to healthcare and interventions are a major driver of the burden of disease in human populations.
-Those most impacted by infectious disease are often those from lower socioeconomic strata and remote populations.
-Despite the impact of these inequities in shaping disease burden, incorporating these factors into estimations of burden is challenging due to the lack of data resolved to these levels...>
+```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+rm(list = ls())
+library(knitr)
+library(RImagePalette)
+library(jpeg)
+library(tidyverse)
+library(lubridate)
+library(magrittr)
+library(rgdal)
+library(broom)
+select <- dplyr::select
+knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
+```
 
-<br>
+```{r get data}
+## read in data
+ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
+ctar_metadata <- read.csv("data/ctar_metadata.csv")
+mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
+mada_communes <- readOGR("output/communes/communes_extracted.shp")
 
-Canine rabies causes an estimated 59,000 human deaths annually.
-Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
-Data on true rabies exposures in humans and incidence in animals is also lacking in most countries.
-Often, the only data available are numbers of bite victims reporting to health facilities.
-The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space.
+## catchment data
+dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
+names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
+comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
+names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")
 
-The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
-Using transmission dynamic models to estimate incidence could improve upon previous studies which may overestimate rabies burden in areas of high reporting and underestimate in areas with low reporting due to the assumption that overall reported bite incidence is proportional to rabies incidence.
+## Centers with no data
+no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
+             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
+             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
+ctar_metadata$exclude <- 0
+ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1
 
-<br>
+## Getting ctar district and commune
+ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
+                                                           mada_district$district)]
+pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
+                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
+dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
+comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
+```
 
-Here, we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country that lacks a national strategy for controlling rabies in domestic dogs.
-Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country.
-PEP is not available at any other public clinics or through the private sector.
-Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al.
-2018.
-We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.
+```{r network fig}
+## Network fig
+## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments?
 
-## METHODS
+## Get color palette from other network fig!
+img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
+pal <- image_palette(img, 31)
+add.alpha <- function(col, alpha=1){
+  if(missing(col))
+    stop("Please provide a vector of colours.")
+  apply(sapply(col, col2rgb)/255, 2, 
+                     function(x) 
+                       rgb(x[1], x[2], x[3], alpha=alpha))  
+}
+ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
+                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
+                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
+                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
+ctar_metadata$fill <- ctar_metadata$color
+ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
+ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
 
-### Bite patient data
-We used the database of bite patient forms submitted from 24 ARMC between 2014 - 2017.
-These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics.
-Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
-These data include the administrative district of the bite patient's address.
-As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted.
-We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.
-Figure 1a + 1b?
+display_palette(ctar_metadata$color)
+dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
+                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
+mada_district@data <- left_join(mada_district@data, dist_catchments)
 
-RIG data?
+## get exposure matrix
+ctar_data %>% 
+  group_by(district, id_ctar) %>% # group by district and ctar
+  summarize(n = n()) -> exp_mat # get count of exposures
+dist_points <- cbind(select(mada_district@data,
+                            mdg_dis_co), coordinates(mada_district))
+names(dist_points)[2:3] <- c("dist_long", "dist_lat")
+dist_points %>%
+ left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
+           by = c("mdg_dis_co" = "ctar_dist")) %>%
+ mutate(dist_long = coalesce(dist_long, LONGITUDE), 
+        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points
 
-<br>
+exp_mat %>% 
+  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
+            by = c("district" = "mdg_dis_co")) %>%
+  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
+                   color, fill, ctar_dist)) -> dist_lines
+dist_lines$size <- log(dist_lines$n + 0.1)*0.2
 
-### GIS Data
-We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
-2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
-We then averaged these to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
-For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
-Population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
-2012) and also aggreagated to the commune level.
-Figure 2 = networky fig + catchments!
+ctar_data %>%
+  group_by(district) %>%
+  summarize(n = n()) %>%
+  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
+             by = c("district" = "mdg_dis_co")) %>%
+  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
+dist_points$size <- log(dist_points$n + 0.1)*0.25
 
-<br>
+pdf("figs/network_1.pdf")
+plot(mada_district, col = mada_district$fill, border = "grey") 
+points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
+       , pch = 1, cex = dist_points$size, 
+       col = dist_lines$color)
+segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
+         lwd = dist_lines$size*2, col = dist_lines$color)
+dev.off()
 
-### Model of reported bites as a function of travel time
+exp_mat %>% 
+  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
+  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
+  group_by(district) %>%
+  mutate(prop = n/sum(n),
+         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
+hist(catch_check$prop)
+ggplot(data = catch_check, aes(CTAR.x, district)) + 
+  geom_tile(aes(fill = prop)) +
+  scale_fill_continuous(type = "viridis", name = "proportion") +
+  geom_text(label = catch_check$catch_match)
+```
+
+```{r clinic level reporting}
+######### Patient time series
+ctar_data %>%
+  select(id_ctar, date_de_consultation) %>%
+  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
+  gather(dose, date_de_consultation, -id_ctar) %>%
+  group_by(id_ctar, date_de_consultation) %>%
+  summarise(n = n()) -> patient_ts
+patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
+patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
+patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]
+
+ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
+  geom_tile(aes(fill = n)) +
+  scale_fill_continuous(type = "viridis", name = "Patients")
+
+## getting reporting
+start_date <- ymd("2014-01-01")
+end_date <- ymd("2017-12-31")
+ts <- as_tibble(seq(start_date, end_date, by = "day"))
+
+patient_ts %>%
+  right_join(ts, by = c("date_de_consultation" = "value")) %>%
+  select(id_ctar, date_de_consultation, n) %>%
+  spread(id_ctar, n) %>%
+  replace(., is.na(.), 0) -> doses_wide
+dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])
+
+date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
+for (j in 1:ncol(date_mat)){
+  rle(dose_mat[ , j]) %>%
+    unclass() %>%
+    as.data.frame() %>%
+    mutate(end = cumsum(lengths),
+         start = c(1, lag(end)[-1] + 1)) %>%
+    filter(values == 0, lengths >= 10) -> rles
+  for (i in 1:nrow(rles)){
+    date_mat[rles$start[i]:rles$end[i], j] <- 0
+  }
+}
+
+date_mat <- replace(date_mat, is.na(date_mat), 1)
+date_mat %>% 
+  as_tibble() %>%
+  group_by(date = year(doses_wide$date_de_consultation)) %>%
+  summarise_all(funs(sum(.)/n())) -> clinic_reporting
+colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
+clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
+clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
+                                                  ctar_metadata$id_ctar)]
+clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
+
+ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
+  geom_tile() +
+  scale_fill_gradient(low = "white", high = "blue")
+```
+
+```{r district level exposures}
+## districts that are non-reporter and data not available yet CTAR
+mada_district@data %>%
+select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
+
+to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
+to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
+ctar_data$excl <- 0
+ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1
+
+ctar_data %>% 
+  filter(excl != 1) %>% # exclude contacts
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
+  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
+  summarize(n = n()) -> exposures # get count of exposures
+
+exposures %>%
+  rename(date = year) %>%
+  left_join(covars) %>% # add pop + ttimes + reporting
+  left_join(clinic_reporting) %>%   # add reporting by clinic + year
+  filter(exclude == 0, prop > 0.25) -> exps_cleaned
+
+exps_cleaned %>%
+  group_by(district, date) %>%
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites, na.rm = TRUE)) %>%
+  group_by(district) %>%
+  summarise(bites = mean(bites)) %>%
+  left_join(covars) -> exps_dist
+
+exps_cleaned %>%
+  group_by(ctar, date) %>% ## Need to rename CTAR so clear if it's CTAR in catchment vs. ctar of bites...
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites, na.rm = TRUE)) %>%
+  group_by(ctar) %>%
+  summarise(bites = mean(bites)) -> exps_ctar
+
+```
+
+```{r ttimes + pop at commune level}
+## Need: weighted ttimes, pop, ctar_in, exclude
+comm_catchments %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) -> comm_covars
+comm_covars <- filter(comm_covars, ttimes_weighted != "Inf")
+comm_covars %>% 
+  left_join(select(mada_district@data, mdg_dis_co, district)) -> comm_data
+comm_data$exclude <- dist_catchments$exclude[match(comm_data$mdg_dis_co, dist_catchments$mdg_dis_co)]
+
+## Data to include
+nrow(comm_data) - sum(comm_data$exclude)
+```
+
+```{r likelihood of observing district data given sum of communes in district}
+library()
+
+get.likelihood <- function(commune,district, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
+
+tmp <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_dist,
+             commune = comm_data, hessian = TRUE)
+params <- tmp$par
+B_ttimes <- params[1]
+B_0 <- params[2]
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
+plot(check$ttimes_weighted, check$bites)
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites)
+abline(a = 0, b = 1, col = "blue", lty = 2)
+
+```
+
+## Model of reported bites as a function of travel time
+As bite patient data were only available...
+
 Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.
@@
 - Sensitivity to assumptions of p_rabid
-
-## Preliminary data processing
-```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
-rm(list = ls())
-library(knitr)
-library(RImagePalette)
-library(jpeg)
-library(tidyverse)
-library(lubridate)
-library(magrittr)
-library(rgdal)
-library(broom)
-select <- dplyr::select
-knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
-
-```
-
-### Data
-```{r get data}
-## read in data
-ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
-ctar_metadata <- read.csv("data/ctar_metadata.csv")
-mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
-dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
-names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
-
-## Centers with no data
-no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
-             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
-             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
-ctar_metadata$exclude <- 0
-ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1
-
-## Get color palette from other network fig!
-img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
-pal <- image_palette(img, 31)
-add.alpha <- function(col, alpha=1){
-  if(missing(col))
-    stop("Please provide a vector of colours.")
-  apply(sapply(col, col2rgb)/255, 2, 
-                     function(x) 
-                       rgb(x[1], x[2], x[3], alpha=alpha))  
-}
-ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
-                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
-                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
-                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
-                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
-ctar_metadata$fill <- ctar_metadata$color
-ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
-ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
-## Getting ctar district and commune
-ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
-                                                           mada_district$district)]
-pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
-                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
-ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
-
-display_palette(ctar_metadata$color)
-dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
-                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
-mada_district@data <- left_join(mada_district@data, dist_catchments)
-dist_catchments$ctar_in <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
-```
-
-```{r catchment figure}
-## get exposure matrix
-ctar_data %>% 
-  group_by(district, id_ctar) %>% # group by district and ctar
-  summarize(n = n()) -> exp_mat # get count of exposures
-dist_points <- cbind(select(mada_district@data,
-                            mdg_dis_co), coordinates(mada_district))
-names(dist_points)[2:3] <- c("dist_long", "dist_lat")
-dist_points %>%
- left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
-           by = c("mdg_dis_co" = "ctar_dist")) %>%
- mutate(dist_long = coalesce(dist_long, LONGITUDE), 
-        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points
-
-exp_mat %>% 
-  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
-            by = c("district" = "mdg_dis_co")) %>%
-  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
-                   color, fill, ctar_dist)) -> dist_lines
-dist_lines$size <- log(dist_lines$n + 0.1)*0.2
-
-ctar_data %>%
-  group_by(district) %>%
-  summarize(n = n()) %>%
-  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
-             by = c("district" = "mdg_dis_co")) %>%
-  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
-dist_points$size <- log(dist_points$n + 0.1)*0.25
-
-pdf("figs/network_1.pdf")
-plot(mada_district, col = mada_district$fill, border = "grey") 
-points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
-       , pch = 1, cex = dist_points$size, 
-       col = dist_lines$color)
-segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
-         lwd = dist_lines$size*2, col = dist_lines$color)
-dev.off()
-
-legend("topleft", c(">100", ">200", ">400", ">800"), title = "Number of total reported bites", 
-       text.width = 3, pt.cex = circle_size, col = "grey",  pch = 16, inset = c(0.1, 0.1), lwd = 1.5, bty = "n", xpd = TRUE, y.intersp=2)
-
-exp_mat %>% 
-  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
-  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
-  group_by(district) %>%
-  mutate(prop = n/sum(n),
-         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
-hist(catch_check$prop)
-ggplot(data = catch_check, aes(CTAR.x, district)) + 
-  geom_tile(aes(fill = prop)) +
-  scale_fill_continuous(type = "viridis", name = "proportion") +
-  geom_text(label = catch_check$catch_match)
-
-```
 
@@
 ```
-
-## Commune level mods
-```{r commune matching, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
-commune_matching <- read.csv("data/commune_check_CH.csv")
-mada_communes <- readOGR("output/communes/communes_extracted.shp")
-mada_communes$match <- tolower(str_replace_all(mada_communes$commune,"[^[:graph:]]", "/"))
-
-mada_communes@data %>%
-  select(match, mtch_ds, mdg_cm_) %>%
-  left_join(commune_matching, .) %>%
-  select(name, match, district, mdg_cm_) -> comm_matches
-
-ctar_data$name <- tolower(str_replace_all(ctar_data$commune,"[^[:graph:]]", "/"))
-left_join(ctar_data, comm_matches) %>%
-  distinct(record_id, .keep_all = TRUE) -> ctar_data
-
-## Looking at proportion matched
-ctar_data %>%
-  group_by(id_ctar) %>%
-  summarise(n = n(), unmatched = sum(is.na(mdg_cm_)),
-            prop_matched = unmatched/n) -> ctar_data_matched
-
-ctar_metadata <- merge(ctar_metadata, ctar_data_matched)
-ctar_metadata <- ctar_metadata[!is.na(ctar_metadata$id_ctar), ]
-ggplot(ctar_metadata, aes(CTAR, prop_matched)) +
-  geom_point() +
-  coord_flip()  +
-  ylab(label = "Proportion of patients matched to commune")
-```
-
-```{r }
-## Ctar data at commune level
-ctar_data %>% 
-  filter(excl != 1) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), commune = mdg_cm_, id_ctar) %>% # group by year and district
-  summarize(n = n()) -> exposures # get count of exposures
-
-comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
-names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")
-comm_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
-                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> comm_catchments
-mada_communes@data <- left_join(mada_communes@data, comm_catchments)
-
-mada_communes@data %>%
-select(commune = mdg_cm_, pop = MDG__201, ttimes_weighted, CTAR, exclude) -> covars
-mada_communes$ctar_in <- ifelse(mada_communes$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
-
-## Catchments with no data
-no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
-             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
-             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
-covars$exclude <- 0
-covars$exclude[covars$CTAR %in% no_data] <- 1
-
-exposures %>%
-  rename(date = year) %>%
-  left_join(covars) %>% # add pop + ttimes + reporting
-  left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) %>%
-  group_by(commune, date) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
-exps_cleaned %>%
-  group_by(commune) %>%
-  summarise(bites = mean(bites)) %>%
-  left_join(covars) -> exps_cleaned
-
-library(lme4)
-
-## poisson with total pop as offset
-exps_cleaned <- filter(exps_cleaned, ttimes_weighted != "Inf")
-mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
-                 data = exps_cleaned, family = poisson)
-
-par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
-plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Observed bites", ylab = "Predicted bites")
-abline(a = 0, b = 1, lty = 2, col = "blue")
-legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
-       inset = c(0.0, -0.1))
-plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
-     ylim = c(0, max(exps_cleaned$bites)))
-points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
-legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
-       pch = 20, bty = "n")
-
-ctar_data %>% 
-  filter(excl != 1) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(commune = mdg_cm_, id_ctar) %>% # group by year and district
-  summarize(n = n()) ->  exp_mat # get count of exposures
-
-comm_catchments$exclude <- 0
-comm_catchments$exclude[comm_catchments$CTAR %in% no_data] <- 1
-
-exp_mat %>% 
-  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
-  left_join(select(comm_catchments, mdg_cm_, CTAR, exclude), by = c("commune" = "mdg_cm_")) %>%
-  group_by(commune) %>%
-  mutate(prop = n/sum(n),
-         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
-hist(catch_check$prop[catch_check$exclude == 0])
-hist(catch_check$prop[catch_check$exclude == 1])
-
-## getting var for ctar in
-pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
-                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
-over(pts, mada_communes)$mtch_ds
-```

### 2019-01-07 accessPEP.Rmd
initial code with new model (aggregated at district)

@@
 ---
-## Goals
-- To estimate burden of rabies in Madagascar using nationally available data from clinics that provide PEP.
-- To explore the potential impacts of expanded access to PEP
 
-## Data
+## INTRODUCTION
+
+[Start broadly with inequity?]
+
+<br>
+
+Canine rabies causes an estimated 59,000 human deaths annually.
+Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
+Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries.
+Often, the only data available are numbers of bite victims reporting to health facilities.
+The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
+the more overall bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may overestimate rabies burden in areas of high reporting and underestimate in areas with low reporting due to the assumption that overall reported bite incidence is proportional to rabies incidence.
+
+<br>
+
+Here, we extend this framework to include spatial access to PEP, using bite patient data from 19 clinics in Madagascar, a rabies endemic country that lacks a national strategy for controlling rabies in domestic dogs.
+Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country.
+PEP is not available at any other public clinics or through the private sector.
+Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al.
+2018.
+We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.
+Assuming a range of exposure incidence given endemic rabies, we generate sub-national estimates of rabies burden and explore the impacts of expanding access to PEP in Madagascar.
+
+Other analyses = vials vs. additional deaths averted (given non-rabid exposures) (# vials of PEP per death averted for baseline + expanded1 + expanded2)
+
+## METHODS
+
 - Patient data from 19 ARMC accross the country over four years (2013 - 2017)
@@
 - Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the designated admin level.
+
+### Bite patient data
+We used the database of bite patient forms submitted from 24 ARMC between 2014 - 2017.
+These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics.
+Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
+These data include the administrative district of the bite patient's address.
+As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted.
+We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.
+
+RIG data?
+
+<br>
+
+### GIS Data
+We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
+2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
+We then averaged these to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
+For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
+Population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
+2012) and also aggreagated to the commune level.
+
+<br>
+
+### Model of reported bites as a function of travel time
+While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX).
+In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence as a function of travel time at the commune level (sub-district administrative level).
+Specifically, we modeled bites as follows:
+$$ 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
+$$
+where $\mu_{d}$ is the mean number of bites in district which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$.
+
+As we had data available on reported bites at the commune level from the Moramanga ARMC, we modeled observed commune bites in the same framework, but un-aggregated where:
+$$ 
+\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+$$
+where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
+We only included communes which were designated to be within the catchment for the clinic (so primarily the Moramanga and Anosibe A'nala Districts)??
+We compared our estimates of $\beta_t$ (i.e.
+the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
+
+P_rabid model...how to explain this one...
+
+To better understand if ttimes might capture access, we also looked at whether pop alone can predict bites and distance could predict bites...
+
+Ability to predict average annual/monthly bite incidence by travel time
+Ability to predict out-of-fit to other catchments? Tana + Fianarantsoa?
+
+<br>
+
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
+
+We calculated deaths as:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
+            \frac{R_i \times \rho_{max}}{B_i}, & \text{otherwise}
+            \end{cases}
+$$
+<br>
+such that rabid reported bites (i.e.
+$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_i \times \rho_{max}$).
+$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
+the area with the minimum travel time).
+
+We define $x$ in two different ways:
+1. At the min and max of the range reported in Rajeev et al. 2018 for data from the Moramanga District
+2. As predicted by the model of proportion rabid as a function of travel time estimated from the Moramanga ARMC data.
+
+$R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
+
+<br>
+
+$$r \times D \times p_{exp}$$
+
+<br>
+
+where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018).
+We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
+
+<br>
+
+### Sensitivity Analyses
+- Sensitivity to assumptions of R
+- Sensitivity to assumptions of p_rabid
+
+<br>
+
+## Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
+1. The baseline with the current clinic locations (n = 31)
+2. Expansion of ARMC to one clinic per district (n = 114)
+3. Expansion of ARMC to all CSB IIs (n = ...?)
+
+We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+## Results
 
@@
 
+### Clinic catchments and reporting
 ```{r get data}
@@
 
+### Models of bites as a function of travel times
+
 ```{r district level exposures}
@@
 ```{r likelihood of observing district data given sum of communes in district}
-library()
 
@@
       summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
+      left_join(select(exps_dist, district, dist_bites = bites), 
                 by = c("mdg_dis_co" = "district")) %>%
@@
 
-## Model of reported bites as a function of travel time
-As bite patient data were only available...
+```{r checking impact of ctar in}
+exps_dist$ctar_in <- dist_catchments$ctar_in_dist[match(exps_dist$district,
+                                                        dist_catchments$mdg_dis_co)]
+ggplot() + 
+  geom_boxplot(data = exps_dist, aes(x = as.factor(ctar_in), y = bites/pop))
 
-Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.
+comm_data$ctar_in_comm <- comm_catchments$ctar_in_comm[match(comm_data$mdg_cm_,
+                                                        comm_catchments$mdg_cm_)]
+comm_data$ctar_in_dist <- dist_catchments$ctar_in_dist[match(comm_data$mdg_dis_co,
+                                                        dist_catchments$mdg_dis_co)]
 
-We fit a linear mixed effects model where the annual average bite incidence at the district level was predicted by the average travel time to the nearest CTAR for the district, the population within the district, and a random catchment level effect.
+get.likelihood <- function(commune,district, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_in <- params[2]
+  B_0 <- params[3]
 
-For three other catchments with reporting above x% we used to out-of-fit predict...
-Ability to predict average annual/monthly bite incidence by travel time
-Ability to predict out-of-fit to other catchments
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_dist + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
 
-<br>
+tmp <- optim(par = c(1e-6, 1e-6, 0.1), get.likelihood, district = exps_dist,
+             commune = comm_data, hessian = TRUE)
+params <- tmp$par
+B_ttimes <- params[1]
+B_in <- params[2]
+B_0 <- params[3]
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_comm + B_0)*pop) -> check
+plot(check$ttimes_weighted, check$bites)
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + + B_in*ctar_in_comm + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites)
+abline(a = 0, b = 1, col = "blue", lty = 2)
+
+```
+
+### Model of proportion rabid
+
+```{r Moramanga data}
+## Libraries
+library(ggThemeAssist)
+library(colourpicker)
+library(dplyr)
+library(stringr)
+library(lubridate)
+library(ISOweek)
+library(ggplot2)
+library(knitr)
+library(prettymapr)
+# libraries for gis
+library(raster)
+library(maptools)
+library(maps)
+library(GISTools)
+library(rgdal)
+library(sp)
+library(rgdal)
+library(gdistance)
+
+source("R/utils.R")
+
+## data from WM
+triage <- read.csv("data/moramanga/triage_up_todate.csv", header = TRUE)
+ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
+pep <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "PEP")), header = TRUE)
+hh <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Household")), header = TRUE)
+human <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Human")), header = TRUE)
+animal <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Animal")), header = TRUE)
+
+## Read in new column names
+col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
+match.colnames <- function (dataframe, lookup){
+  colnames(dataframe) <- lookup$new.colnames[match(colnames(dataframe), 
+                                                   lookup$og.colnames)]
+  dataframe <- dataframe[ , !names(dataframe) %in% c("NULL")]
+  return(dataframe)
+}
+
+triage <- match.colnames(triage, subset(col_lookup, form == "triage"))
+ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
+pep <- match.colnames(pep, subset(col_lookup, form == "pep"))
+hh <- match.colnames(hh, subset(col_lookup, form == "hh"))
+animal <- match.colnames(animal, subset(col_lookup, form == "animal"))
+human <- match.colnames(human, subset(col_lookup, form == "human"))
+
+# fix factor names
+match.within <- function (vec_return, mismatch, match){
+  for (i in 1:length(match)){
+    if (is.na(mismatch[i])) next
+    mismatched <- paste0 ("\\<", mismatch[i], "\\>")
+    vec_return <- gsub(mismatched, match[i], vec_return, ignore.case = TRUE)
+  }
+  return (vec_return)
+}
+
+## Changing suspected to probable
+animal$suspect.animal <- match.within(animal$suspect.animal, "Suspected", "Probable")
+human$suspect.exposure <- match.within(human$suspect.exposure, "Suspected", "Probable")
+triage$suspect.animal <- match.within(triage$suspect.animal, "Suspected", "Probable")
+
+## get number of animals + humans bitten by biter.IDs here!
+animals.bitten.byID <- table(animal$biter.ID)
+humans.bitten.byID <- table(human$biter.ID)
+
+## Set-up time series plotting (axes)
+tfunc <- function (date, format_date="%Y-%m-%d", start="01-09-2016", format_start="%d-%m-%Y",
+                   year1=2016, tstep="month", period=FALSE) {
+  get.consec (date, format_date, start, format_start,
+              year1, tstep, period) - 8
+}
+
+## all dates
+triage %>% 
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> triage
+pep$pep.date <- dmy(pep$pep.date)
+ctar %>%
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
+animal %>%
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> animal
+human %>%
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> human
+
+# triage
+triage$commune <- sapply(strsplit(as.character(triage$exposure.location), "\\("), "[", 1)
+triage$commune <- trimws(triage$commune, which = "right")
+triage$district <- sapply(strsplit(as.character(triage$exposure.location), "\\, "), "[", 2)
+triage$district <- gsub(" \\(District\\)", "", triage$district)
+
+# ctar
+ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
+ctar$commune <- trimws(ctar$commune, which = "right")
+ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
+ctar$district <- gsub(" \\(District\\)", "", ctar$district)
+
+## animal and human locations
+animal <- merge(animal, hh, by = "reference.hh")
+human <- merge(human, hh, by = "reference.hh")
+
+## get apprx case dates
+ctar$ctar.ID <- paste0(ctar$ctar.ID, "_", year(ctar$date.reported))
+human$ctar.ID <- paste0(human$ctar.ID, "_", year(human$date.bitten))
+human$reference.ctar <- ctar$reference.ctar[match(human$ctar.ID, ctar$ctar.ID)]
+#human$date.bitten[!is.na(human$reference.ctar)] <- ctar$date.bitten[match(human$reference.ctar[!is.na(human$reference.ctar)], ctar$reference.ctar)]
+
+human %>%
+  group_by(biter.ID) %>% 
+  summarise(sort(date.bitten)[1]) -> date.first.bite
+animal$date.case <- date.first.bite$`sort(date.bitten)[1]`[match(animal$ID, date.first.bite$biter.ID)]
+animal$date.case[is.na(animal$date.case)] <- animal$date.first.bite[is.na(animal$date.case)]
+animal$date.case[is.na(animal$date.case)] <- animal$date.symptoms.started[is.na(animal$date.case)] # fixing animals which don't have a case date (i.e. non-biters, but ones that are still suspect!)
+## what about ones without a date?? recheck!
+
+ctar %>%
+  group_by(reference.triage) %>% 
+  summarise(sort(date.bitten)[1]) -> date.first.bite
+triage$date.case <- date.first.bite$`sort(date.bitten)[1]`[match(triage$reference.triage, date.first.bite$reference.triage)]
+triage$date.case[is.na(triage$date.case)] <-triage$date.reported[is.na(triage$date.case)]
+
+## get all months 
+triage$month <- tfunc(triage$date.reported)
+ctar$month <- tfunc(ctar$date.reported)
+human$month <- tfunc(human$date.bitten)
+animal$month <- tfunc(animal$date.case)
+pep$month <- tfunc(pep$pep.date)
+
+## change all true/false to 0/1
+triage[triage == FALSE] <- 0
+ctar[ctar == FALSE] <- 0
+animal[animal == FALSE] <- 0
+human[human == FALSE] <- 0
+pep[pep == FALSE] <- 0
+pep[pep == TRUE] <- 1
+
+## subset to study period (Sep 2016 - to date)
+## checks for cases where month is NA!
+triage <- subset(triage, month > 0 & month < 25)
+ctar <- subset(ctar, month > 0 & month < 25)
+human <- subset(human, month > 0 & month < 25)
+animal <- subset(animal, month > 0 & month < 25)
+pep <- subset(pep, month > 0 & month < 25)
+
+## linking up
+ctar$traced <- NA
+ctar$traced <- match(ctar$ctar.ID, human$ctar.ID)
+ctar$traced[ctar$traced > 0] <- 1
+ctar$traced[is.na(ctar$traced)] <- 0
+ctar$reference.human <- human$reference.human[match(ctar$ctar.ID, human$ctar.ID)]
+triage$reference.human <- ctar$reference.human[match(triage$reference.triage, ctar$reference.triage)] ## here!
+triage$suspect.animal[!is.na(triage$reference.human)] <- human$suspect.exposure[match(triage$reference.human[!is.na(triage$reference.human)], human$reference.human)]
+
+ctar$suspect.exposure <- triage$suspect.animal[match(ctar$reference.triage, triage$reference.triage)] ## matched to triage!
+
+nonbites <- c("CTAR_201708_000016", "CTAR_201708_000017", "CTAR_201708_000001", "CTAR_201708_000002",
+              "CTAR_201708_000003", "CTAR_201709_000037")
+ctar$traced[ctar$reference.ctar %in% nonbites] <- 1
+triage <- triage[which(triage$reference.triage %in% ctar$reference.triage), ] ## only triage with corresponding exposures
+ctar <- ctar[which(ctar$reference.triage %in% triage$reference.triage), ] ## only ctar with corresponding exposures
+
+## triage data
+triage$triaged <- NA
+triage$triaged[triage$triage.status == "Successful"] <- 1
+triage$triaged[is.na(triage$triaged)] <- 0
+triage$traced <- ctar$traced[match(triage$reference.triage, ctar$reference.triage)]
+triage$traced[which(triage$reference.triage %in% ctar$reference.triage[ctar$traced == 1])] <- 1
+triage$traced[is.na(triage$traced)] <- 0
+nonlinked <- c("TR_201709_000014", "TR_201802_001040", "TR_201802_001045", "TR_201802_001063",  "TR_201802_001107", "TR_201802_001219", "TR_201802_001253", "TR_201802_001332", "TR_201802_001361", "TR_201802_001372", "TR_201709_000031", "TR_201708_000042", "TR_201708_000010", "TR_201708_000001", "TR_201709_000020", "TR_201709_000019", "TR_201709_000035", "TR_201709_000023", "TR_201709_000027", "TR_201710_000025", "TR_201711_000008", "TR_201712_000020", "TR_201712_000027", "TR_201708_000002")
+triage$traced[triage$reference.triage %in% nonlinked] <- 1
+triage$followedup <- triage$traced + triage$triaged
+triage$followedup[triage$followedup > 1] <- 1
+
+## cleaning up statuses
+ctar$suspect.exposure <- as.factor(ctar$suspect.exposure)
+triage$suspect.animal <- as.factor(triage$suspect.animal)
+levels(ctar$suspect.exposure) <- c(levels(ctar$suspect.exposure), "Contact with suspect case")
+ctar$suspect.exposure[ctar$consult.type == "Contact with suspect case"] <- "Contact with suspect case"
+levels(ctar$suspect.exposure)[levels(ctar$suspect.exposure) == ""] <- "Unknown"
+levels(triage$suspect.animal)[levels(triage$suspect.animal) == ""] <- "Unknown"
+ctar$followedup <- triage$followedup[match(ctar$reference.triage, triage$reference.triage)]
+
+## folks that reported after tracing
+ctar$date.traced <- human$date.traced[match(ctar$ctar.ID, human$ctar.ID)]
+date.diff <- ctar$date.traced - ctar$date.reported
+date.diff[is.na(date.diff)] <-0
+after.tracing <- ctar[date.diff < 0,]
+ctar <- ctar[date.diff >= 0, ]
+nrow(after.tracing)
+
+ctar$triaged <- triage$triaged[match(ctar$reference.triage, triage$reference.triage)]
+ctar %>%
+  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
+                          by = c("commune" = "commune", "district" = "mtch_ds")) %>%
+  filter(triaged == 1, consult.type != "Contact with suspect case") %>%
+  dplyr::select(mdg_cm_, suspect.exposure) %>%
+  gather(suspect.exposure, value, -mdg_cm_) %>%
+  group_by(mdg_cm_, value) %>%
+  tally %>%
+  spread(value, n, fill = 0) %>%
+  mutate(total = Confirmed + No + Probable + Unknown, 
+         rabid_upper = Confirmed + Probable + Unknown,
+         rabid_lower = Confirmed + Probable) %>% 
+  left_join(comm_covars) %>%
+  filter(!is.na(ttimes_weighted), district != "Antananarivo Renivohitra") -> prop_rabid
+
+## Model of proportion rabid
+prop_rabid %$% plot(ttimes_weighted, rabid_lower/total)
+mod <- glm(cbind(rabid_lower, total - rabid_lower) ~ ttimes_weighted, data = prop_rabid, 
+           family = binomial)
+prop_rabid %$% plot(rabid_lower/total, predict(mod, type = "response"))
+prop_rabid %$% plot(ttimes_weighted, predict(mod, type = "response"), ylim = c(0, 1))
+prop_rabid %$% points(ttimes_weighted, rabid_lower/total, col = "blue")
+
+mod <- glm(cbind(rabid_upper, total - rabid_upper) ~ ttimes_weighted, data = prop_rabid, 
+           family = binomial)
+prop_rabid %$% plot(rabid_upper/total, predict(mod, type = "response"))
+prop_rabid %$% plot(ttimes_weighted, predict(mod, type = "response"), ylim = c(0, 1))
+prop_rabid %$% points(ttimes_weighted, rabid_upper/total, col = "blue")
+
+```
+
+```{r likelihood of observing moramanga data @ commune level}
+## Bite data in Moramanga
+ctar %>%
+  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
+                          by = c("commune" = "commune", "district" = "mtch_ds")) %>%
+  filter(consult.type == "Animal bite/exposure") %>%
+  group_by(mdg_cm_) %>%
+  summarize(bites = n()) %>%
+  left_join(comm_covars) %>%
+  filter(!is.na(ttimes_weighted), district %in% c("Moramanga", "Anosibe-An'ala")) -> bites
+
+get.likelihood <- function(commune, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  commune %>%
+      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
+  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
+}
+
+tmp <- optim(par = c(1e-6, 0.1), get.likelihood, commune = bites, hessian = TRUE)
+params <- tmp$par
+B_ttimes <- params[1]
+B_0 <- params[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(tmp$hessian)))); se
+
+ex#simulate "standard errors"
+varcov <- solve(tmp$hessian)
+library(MASS)
+
+ntest = 1000; testpar = tmp$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, tmp$par, varcov)
+}
+
+mean(tmp$par[1])
+mean(tmp$par[2])
+#exp(mean(tmp$par[4]))
+
+quantile(sim.par[,1],c(0.025,0.975))
+quantile(sim.par[,2],c(0.025,0.975))
+
+bites %>%
+      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
+plot(check$ttimes_weighted, check$bites)
+plot(check$bites, check$bites_mu)
+abline(a = 0, b = 1, col = "blue", lty = 2)
+ttimes_run <- seq(0, 12*60, by = 30)
+test_pop = 1e5
+plot(ttimes_run, exp(B_ttimes*ttimes_run + B_0)*test_pop, type = "l")
+
+```
 
 ### Estimation of burden and reporting
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
-
-We calculated deaths as:
-$$deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}$$
-
-where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
-$$
-           p_{rabid}=
-            \begin{cases}
-            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
-            \frac{R_i}{B_i}, & \text{otherwise}
-            \end{cases}
-$$
-<br>
-
-such that rabid reported bites (i.e.
-$p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
-$R_{i} \times \rho_{max}$.
-$p_{rabid}$ is taken from Rajeev et al.
-2018 on contact tracing data from the Moramanga District.
-
-$R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
-
-<br>
-
-$$r \times D \times p_{exp}$$
-
-<br>
-
-where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
-2018).
-We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa(cite?).
-
-<br>
-
-## Estimating the impact of expanding PEP Access
-We use this framework to compare three scenarios of PEP provisioning in Madagascar:
-1. The baseline with the current clinic locations (n = 31)
-2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to all CSB IIs (n = ...?)
-
-We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
-
-<br>
-
-### Sensitivity Analyses
-- Sensitivity to assumptions of R
-- Sensitivity to assumptions of p_rabid
-
-```{r}
-## get vials delivered
-ctar_metadata %>% 
-  dplyr::select(id_ctar, starts_with("doses")) %>%
-  gather(year, vials_delivered, -id_ctar) %>%
-  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered
-
-## getting dose time series by clinic
-########### Need to fix dates first! ###############
-ctar_data %>%
-  select(id_ctar, datej0, datej3, datej7, datej28) %>%
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
-  gather(dose, date, -id_ctar) %>%
-  group_by(id_ctar, date) %>%
-  summarise(n = n()) -> dose_ts
-dose_ts <- dose_ts[!is.na(dose_ts$date), ]
-dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
-dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]
-
-## getting theoretical # vials by year
-dose_ts %>%
-  mutate(year = year(date), vials = ceiling(n/2)) %>%
-  group_by(id_ctar, year) %>%
-  summarize(vials_estimated = sum(vials)) %>%
-  left_join(vials_delivered) -> dose_comp
-
-######### Patient time series
-ctar_data %>%
-  select(id_ctar, date_de_consultation) %>%
-  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
-  gather(dose, date_de_consultation, -id_ctar) %>%
-  group_by(id_ctar, date_de_consultation) %>%
-  summarise(n = n()) -> patient_ts
-patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
-patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
-patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]
-
-ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
-  geom_tile(aes(fill = n)) +
-  scale_fill_continuous(type = "viridis", name = "Patients")
-
-patient_ts %>%
-  filter(id_ctar == 1) %$%
-  hist(n, breaks = 100, col = "blue", 
-       xlab = "Patients reporting", main = "Soanierana Ivongo")
-patient_ts %>%
-  filter(ctar == "Toamasina") %$%
-  hist(n, breaks = 100, 
-       xlab = "Patients reporting", main = "Toamasina", col = "red")
-
-## getting reporting
-start_date <- ymd("2014-01-01")
-end_date <- ymd("2017-12-31")
-ts <- as_tibble(seq(start_date, end_date, by = "day"))
-
-patient_ts %>%
-  right_join(ts, by = c("date_de_consultation" = "value")) %>%
-  select(id_ctar, date_de_consultation, n) %>%
-  spread(id_ctar, n) %>%
-  replace(., is.na(.), 0) -> doses_wide
-dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])
-
-date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
-for (j in 1:ncol(date_mat)){
-  rle(dose_mat[ , j]) %>%
-    unclass() %>%
-    as.data.frame() %>%
-    mutate(end = cumsum(lengths),
-         start = c(1, lag(end)[-1] + 1)) %>%
-    filter(values == 0, lengths >= 10) -> rles
-  for (i in 1:nrow(rles)){
-    date_mat[rles$start[i]:rles$end[i], j] <- 0
-  }
-}
-
-date_mat <- replace(date_mat, is.na(date_mat), 1)
-date_mat %>% 
-  as_tibble() %>%
-  group_by(date = year(doses_wide$date_de_consultation)) %>%
-  summarise_all(funs(sum(.)/n())) -> clinic_reporting
-colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
-clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
-clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
-                                                  ctar_metadata$id_ctar)]
-clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
-
-ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
-  geom_tile() +
-  scale_fill_gradient(low = "white", high = "blue")
-```
-
-## Submission of forms
-- Missingness seems to have been due to periods of time without form submission
-- Calculating periods of time
-
-```{r clean data}
-## districts that are non-reporter and data not available yet CTAR
-mada_district@data %>%
-select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
-
-to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
-to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$excl <- 0
-ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1
-
-ctar_data %>% 
-  filter(excl != 1) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
-  summarize(n = n()) -> exposures # get count of exposures
-
-exposures %>%
-  rename(date = year) %>%
-  left_join(covars) %>% # add pop + ttimes + reporting
-  left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) %>%
-  group_by(district, date) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
-exps_cleaned %>%
-  group_by(district) %>%
-  summarise(bites = mean(bites)) %>%
-  left_join(covars) -> exps_cleaned
-
-library(lme4)
-
-## poisson with total pop as offset
-mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
-                 data = exps_cleaned, family = poisson)
-par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
-plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Observed bites", ylab = "Predicted bites")
-abline(a = 0, b = 1, lty = 2, col = "blue")
-legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
-       inset = c(0.0, -0.1))
-plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
-     ylim = c(0, max(exps_cleaned$bites)))
-points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
-legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
-       pch = 20, bty = "n")
-```
 
@@
 ```
+
+### Sensitivity Analyses
+
+### Estimating the impact of expanding PEP Access
+
+## Discussion
+
+### Key findings
+
+### Strengths and Limitations
+- Not accounting for clinic functioning
+- Not explicit data on reporting or rabies incidence!
+- Other factors that drive reporting
+
+### Broader context
+
+### Conclusions

### 2019-01-08 accessPEP.Rmd
adding in some code to show patient daily ts
## don't know if it's worth the pain!

--- 2018-12-18 accessPEP.Rmd
@@
 ---
-## Goals
-- To estimate burden of rabies in Madagascar using nationally available data from clinics that provide PEP.
-- To explore the potential impacts of expanded access to PEP
 
-## Data
+## INTRODUCTION
+
+Access to healthcare is a significant driver of disease burden globally.
+Spatial access to care (cite Jess paper here).
+Rarely incorporated into national estimates of burden due to lack of data and tools for understanding...
+
+<br>
+
+Canine rabies causes an estimated 59,000 human deaths annually.
+Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
+Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries.
+Often, the only data available are numbers of bite victims reporting to health facilities.
+The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
+the more overall bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+
+In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country.
+PEP is not available at any other public clinics or through the private sector.
+Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al.
+2018.
+We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.
+Assuming a range of exposure incidence given endemic rabies, we generate sub-national estimates of rabies burden and explore the impacts of expanding access to PEP in Madagascar.
+
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
+Here, we adapt this framework and use bite patient data from Madagascar to predict rabies burden.
+
+## METHODS
+
+### Bite patient data
+We used the database of bite patient forms submitted from 24 ARMC across ARMC between 2014 - 2017.
+These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics.
+Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
+These data include the administrative district of the bite patient's address.
+As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted.
+We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.
+
 - Patient data from 19 ARMC accross the country over four years (2013 - 2017)
+
   - includes dates and home locations (to the district level) of patient reporting for PEP
@@
   - includes dates and home locations (to the commune level) of patients reporting for PEP
-  - also includes rabies exposure status based on a combination of clinic-based triage and contact
-  tracing (Rajeev et al. 2018)
-- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
-  - We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
+
+RIG data?
+
+<br>
+
+### GIS Data
+We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
-  - We then calculated a weighted average by human population to the commune or district level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
-  - For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
-- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the designated admin level.
+We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
+For each clinic, we defined the catchment area as all **districts?** for which the clinic was the closest ARMC.
+Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
+2012) and also aggreagated to the commune level
+<br>
+
+### Model of reported bites as a function of travel time
+While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX).
+In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
+Incidence at the commune level is then a function of travel times to the closest ARMC.
+Specifically, we modeled bites as follows:
+$$ 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
+$$
+where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+
+As we had data available on reported bites at the commune level from the Moramanga ARMC, we modeled observed commune bites in the same framework, but un-aggregated where:
+$$ 
+\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+$$
+where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
+We only included communes which were designated to be within the catchment for the clinic (so primarily the Moramanga and Anosibe A'nala Districts)??
+We compared our estimates of $\beta_t$ (i.e.
+the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
+
+To better understand if ttimes might capture access, we also looked at whether pop alone can predict bites and distance could predict bites...
+Ability to predict out-of-fit to other catchments? Tana + Fianarantsoa?
+
+<br>
+
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
+
+We calculated deaths as:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
+            \frac{R_i \times \rho_{max}}{B_i}, & \text{otherwise}
+            \end{cases}
+$$
+<br>
+
+such that rabid reported bites (i.e.
+$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_i \times \rho_{max}$).
+$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
+the area with the minimum travel time).
+We look at the range of $x$ reported in Rajeev et al.
+2018 for data from the Moramanga District (0.2 - 0.6).
+
+$R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
+
+<br>
+
+$$r \times D \times p_{exp}$$
+
+<br>
+
+where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018).
+We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
+
+<br>
+
+### Sensitivity Analyses
+- R (increasing or decreasing with pop) (min + max)
+- P-rabid min and max
+- Rho-max min and max
+- Probability of death min and max
+
+<br>
+
+### Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
+
+1. The baseline with the current clinic locations (n = 31)
+2. Expansion of ARMC to one clinic per district (n = 114)
+3. Expansion of ARMC to all CSB IIs, something to describe CSB IIs here, that is primary hospitals (n = X)
+
+We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+## RESULTS
 
@@
 rm(list = ls())
+
+## Libraries
 library(knitr)
@@
 library(broom)
+library(stringr)
+library(ISOweek)
+library(raster)
+library(sp)
+
+source("R/utils.R")
+
 select <- dplyr::select
@@
 
+### Clinic catchments and reporting
+For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).
+In addition, at the clinic level, the majority of bites reported to a given ARMC came from within the catchment (Figure 1C).
+
 ```{r network fig}
@@
 ## Get color palette from other network fig!
-img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
-pal <- image_palette(img, 31)
-add.alpha <- function(col, alpha=1){
-  if(missing(col))
-    stop("Please provide a vector of colours.")
-  apply(sapply(col, col2rgb)/255, 2, 
-                     function(x) 
-                       rgb(x[1], x[2], x[3], alpha=alpha))  
-}
 ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
-                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                 "#99d8c9", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                  "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
@@
                  "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
+ctar_metadata$color[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
 ctar_metadata$fill <- ctar_metadata$color
-ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
-ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
+ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.4)
+display_palette(ctar_metadata$color)
 
-display_palette(ctar_metadata$color)
 dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                      color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
-mada_district@data <- left_join(mada_district@data, dist_catchments)
+mada_district@data <- left_join(mada_district@data, dist_catchments) ## need this?
 
-## get exposure matrix
+## get exposure matrix = first set of points
+## district exposures
 ctar_data %>% 
+  drop_na(id_ctar, district) %>%
+  group_by(district) %>% # group by district
+  summarize(n = n()) -> exps_dist # get count of exposures
+## ctar exposures
+ctar_data %>% 
+  drop_na(id_ctar) %>%
+  group_by(id_ctar) %>% # group by ctar
+  summarize(n = n()) -> exps_ctar
+## ctar by district exposure matrix
+ctar_data %>% 
+  drop_na(id_ctar, district) %>%
   group_by(district, id_ctar) %>% # group by district and ctar
-  summarize(n = n()) -> exp_mat # get count of exposures
+  summarize(n = n()) -> exp_mat# get count of exposures
+
+## To plot district bites
 dist_points <- cbind(select(mada_district@data,
@@
            by = c("mdg_dis_co" = "ctar_dist")) %>%
- mutate(dist_long = coalesce(dist_long, LONGITUDE), 
-        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points
+ mutate(dist_long = coalesce(LONGITUDE, dist_long), 
+        dist_lat = coalesce(LATITUDE, dist_lat)) -> dist_points
 
-exp_mat %>% 
-  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
-            by = c("district" = "mdg_dis_co")) %>%
-  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
-                   color, fill, ctar_dist)) -> dist_lines
-dist_lines$size <- log(dist_lines$n + 0.1)*0.2
-
-ctar_data %>%
-  group_by(district) %>%
-  summarize(n = n()) %>%
+exps_dist %>%
   right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
@@
 
-pdf("figs/network_1.pdf")
+## To plot ctar points (only those not excluded)
+exps_ctar %>%
+ left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE, 
+                  exclude, color, fill)) %>%
+ filter(exclude == 0) -> ctar_points
+
+## To plot lines between ctar + districts
+exp_mat %>% 
+  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
+  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
+                   color, fill, ctar_dist)) -> dist_lines
+dist_lines$size <- log(dist_lines$n + 0.1)*0.2
+
+pdf("figs/network.pdf")
 plot(mada_district, col = mada_district$fill, border = "grey") 
-points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
-       , pch = 1, cex = dist_points$size, 
-       col = dist_lines$color)
-segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
+segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
+         y1 = dist_lines$LATITUDE,
          lwd = dist_lines$size*2, col = dist_lines$color)
+points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
+       cex = log(ctar_points$n + 0.1)*0.25)
+points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
+       col = dist_points$color)
 dev.off()
@@
 exp_mat %>% 
-  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
-  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
-  group_by(district) %>%
+  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
+  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
+            by = c("district" = "mdg_dis_co")) %>%
+  group_by(district) %>% # group by ctar
   mutate(prop = n/sum(n),
-         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
-hist(catch_check$prop)
-ggplot(data = catch_check, aes(CTAR.x, district)) + 
-  geom_tile(aes(fill = prop)) +
-  scale_fill_continuous(type = "viridis", name = "proportion") +
-  geom_text(label = catch_check$catch_match)
+         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
+  filter(catch_match == 1) -> catch_dist
+hist(catch_dist$prop, main = "", 
+     xlab = "Proportion of patients \n reporting to closest ARMC",
+     ylab = "Number of districts", col = "grey50", border = "white", xlim = c(0, 1))
+
+exp_mat %>% 
+  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
+  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
+            by = c("district" = "mdg_dis_co")) %>%
+  group_by(ctar_obs) %>% # group by ctar
+  mutate(prop = n/sum(n),
+         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
+  filter(catch_match == 1) %>%
+  summarize(prop = sum(prop)) -> catch_ctar
+hist(catch_ctar$prop, main = "", 
+     xlab = "Proportion of patients reporting from within catchment",
+     ylab = "Number of ARMC", col = "grey50", border = "white", xlim = c(0, 1))
 ```
+
+We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 10 consecutive days (Figure 2).
+We further excluded any years for which there was less than 25% reporting.
 
@@
   select(id_ctar, date_de_consultation) %>%
+  drop_na(id_ctar) %>%
   mutate(date_de_consultation = ymd(date_de_consultation)) %>%
@@
   summarise(n = n()) -> patient_ts
+
 patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
@@
 
-ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
-  geom_tile(aes(fill = n)) +
-  scale_fill_continuous(type = "viridis", name = "Patients")
+start_date <- ymd("2014-01-01")
+end_date <- ymd("2017-12-31")
 
 ## getting reporting
-start_date <- ymd("2014-01-01")
-end_date <- ymd("2017-12-31")
 ts <- as_tibble(seq(start_date, end_date, by = "day"))
@@
 
-date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
+get.days <- function(dmat = dose_mat, threshold = 10) {
+  date_mat <- matrix(NA, nrow(dmat), ncol(dmat))
   for (j in 1:ncol(date_mat)){
-  rle(dose_mat[ , j]) %>%
+    rle(dmat[ , j]) %>%
       unclass() %>%
@@
              start = c(1, lag(end)[-1] + 1)) %>%
-    filter(values == 0, lengths >= 10) -> rles
+      filter(values == 0, lengths >= threshold) -> rles
+  if(nrow(rles) > 0){
     for (i in 1:nrow(rles)){
         date_mat[rles$start[i]:rles$end[i], j] <- 0
+      }
     }
@@
   colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
+
   clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
@@
   clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
+  clinic_reporting$threshold <- threshold
+  return(list(clinic_reporting, date_mat))
+}
+
+## Color or size dots by their # from that year
+# ctar_data %>%
+#   group_by(date_de_consultation = year(date_de_consultation), id_ctar) %>%
+#   summarize(tot = n()) -> year_by_clinic
+
+clinic_reporting <- get.days(threshold = 10)[[1]]
+# clinic_reporting %>%
+#   mutate(prop10 = prop, prop5 = get.days(threshold = 5)[[1]]$prop, 
+#          prop30 = get.days(threshold = 30)[[1]]$prop) %>%
+#   left_join(year_by_clinic) %>%
+#   ggplot(., aes(x = reorder(ctar, prop10), y = prop10)) + 
+#   geom_point() +
+#   geom_linerange(aes(ymin = prop30, ymax = prop5)) +
+#   facet_wrap(~ date_de_consultation, nrow = 1) +
+#   coord_flip() +
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+#   xlab("") + 
+#   ylab("Proportion of days included") +
+#   scale_color_gradient(low = "white", high = "blue")
+# 
+# date_mat <- as.data.frame(get.days(threshold = 10)[[2]])
+# date_mat$date_de_consultation <- doses_wide$date_de_consultation
+# names(date_mat[1:(ncol(date_mat)-1)]) <- names(doses_wide[2:(ncol(date_mat))])
+# library(reshape2)
+# head(melt(date_mat, id = "date_de_consultation"))
+
+ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
+   geom_tile(aes(fill = n)) +
+   scale_fill_continuous(type = "viridis", name = "Patients") +
+   xlim(start_date, end_date)
 
@@
   scale_fill_gradient(low = "white", high = "blue")
+
 ```
 
-```{r district level exposures}
-## districts that are non-reporter and data not available yet CTAR
-mada_district@data %>%
-select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
+```{r getting Mora data}
+patient_ts %>%
+  group_by(ctar) %>%
+  mutate(exclude = ifelse(n > mean(n) + 6*sd(n), 1, 0)) -> patient_ts
+
+ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
+  geom_jitter(alpha = 0.5, width = 0.25) +
+  coord_flip()
 
@@
 to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$excl <- 0
-ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1
+ctar_data$exclude_bycomment <- 0
+ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
+ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
+patient_ts %>%
+  ungroup() %>%
+  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
+  right_join(ctar_data) -> ctar_check
 
-ctar_data %>% 
-  filter(excl != 1) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
-  summarize(n = n()) -> exposures # get count of exposures
+(ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/
+  length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])
 
-exposures %>%
-  rename(date = year) %>%
-  left_join(covars) %>% # add pop + ttimes + reporting
-  left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) -> exps_cleaned
-
-exps_cleaned %>%
-  group_by(district, date) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) %>%
-  group_by(district) %>%
-  summarise(bites = mean(bites)) %>%
-  left_join(covars) -> exps_dist
-
-exps_cleaned %>%
-  group_by(ctar, date) %>% ## Need to rename CTAR so clear if it's CTAR in catchment vs. ctar of bites...
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) %>%
-  group_by(ctar) %>%
-  summarise(bites = mean(bites)) -> exps_ctar
+ctar_check %>%
+  group_by(id_ctar) %>%
+  summarize(prop_excluded = length(exclude_bydate[exclude_bydate == 1])/n()) %>%
+  ggplot(data = ., aes(y = id_ctar, prop_excluded)) +
+  geom_point()
 
@@
 
-```{r likelihood of observing district data given sum of communes in district}
-library()
+``` {r check with moramanga}
+## ctar for # bites and also checking days to rule out contacts
+col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
+ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
+ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
 
-get.likelihood <- function(commune,district, params) {
+## merge with newest data!
+
+## Set-up time series
+tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
+                   year1 = 2016, tstep = "month", period = FALSE) {
+  get.consec (date, format_date, start, format_start,
+              year1, tstep, period) - 8 ## starts in Sep
+}
+
+## all dates
+ctar %>%
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
+ctar$month <- tfunc(ctar$date.reported)
+## subset to study period (Sep 2016 - to date)
+## checks for cases where month is NA!
+ctar <- subset(ctar, month > 0 & month < 28)
+
+# ctar district and commune
+ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
+ctar$commune <- trimws(ctar$commune, which = "right")
+ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
+ctar$district <- gsub(" \\(District\\)", "", ctar$district)
+
+## change all true/false to 0/1
+ctar[ctar == FALSE] <- 0
+
+## Bite data in Moramanga
+n_months <- max(ctar$month, na.rm = TRUE)
+
+ctar %>%
+  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
+            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
+  filter(consult.type == "Animal bite/exposure") %>%
+  group_by(mdg_cm_) %>%
+  summarize(bites = round(n()/n_months*12)) %>%
+  left_join(comm_covars) %>%
+  filter(!is.na(ttimes_weighted), 
+         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
+
+ctar %>%
+  group_by(date.reported) %>%
+  summarize(n = n()) %>%
+  mutate(exclude = ifelse(n > mean(n) + 2*sd(n), 1, 0)) -> mora_ts
+
+ggplot(data = mora_ts, aes(x = "Moramanga", y = n, color = as.factor(exclude))) +
+  geom_jitter(alpha = 0.5, width = 0.1) +
+  coord_flip()
+
+ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+table(ctar$consult.type, ctar$exclude)
+
+sdev <- seq(2, 10, by = 1)
+check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
+for (i in 1:length(sdev)){
+  ctar %>%
+    group_by(date.reported) %>%
+    summarize(n = n()) %>%
+    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
+  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
+  check <- rbind(check, c(sdev[i], checksd[1:2]))
+}
+
+plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
+points(check$sd, check$prop_bites, col = "red", pch = 20)
+
+```
+
+### Models of bites as a function of travel times
+
+```{r district level exposures}
+## districts that are non-reporter and data not available yet CTAR
+mada_district@data %>%
+select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
+
+ctar_check %>% 
+  filter(exclude_bydate == 0, exclud_bycomment == 0) %>% # exclude contacts
+  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
+  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
+  summarize(n = n()) %>% # get count of exposures
+  left_join(covars) %>% # add pop + ttimes + reporting
+  left_join(clinic_reporting) %>%   # add reporting by clinic + year
+  filter(exclude == 0, prop > 0.25) %>%
+  group_by(district, year) %>%
+  mutate(bites = n/prop) %>%
+  summarise(bites = sum(bites, na.rm = TRUE)) %>%
+  group_by(district) %>%
+  summarise(bites = mean(bites)) %>%
+  left_join(covars) -> exps_check
+
+```
+
+### Model results Moramanga and District
+```{r likelihood of observing moramanga data @ commune level}
+
+get.likelihood.mora <- function(commune, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  commune %>%
+      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
+  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
+}
+
+get.likelihood <- function(commune, district, params) {
   # To test likelihood function
@@
 }
+```
 
-tmp <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_dist,
+```{r likelihood of observing district data given sum of communes in district}
+mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
+B_ttimes_mora <- mora_pars$par[1]
+B_0_mora <- mora_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(mora_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(mora_pars$hessian)
+library(MASS)
+
+ntest = 1000; testpar = mora_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, mora_pars$par, varcov)
+}
+
+B_ttimes_mora_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_mora_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_check,
+             commune = comm_data, hessian = TRUE)
+B_ttimes_mada <- mada_pars$par[1]
+B_0_mada <- mada_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(mada_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(mada_pars$hessian)
+library(MASS)
+
+ntest = 1000; testpar = mada_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, mada_pars$par, varcov)
+}
+
+B_ttimes_mada_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
+par_df <- data.frame(values = c(B_ttimes_mada, B_ttimes_mora, B_0_mada, B_0_mora),  
+                     par = c("B_ttimes", "B_ttimes", "intercept", "intercept"),
+                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
+                                  B_0_mada_CI[2], B_0_mora_CI[2]),
+                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
+                                  CI = B_0_mada_CI[1], B_0_mora_CI[1]),
+                     model = c("Mada", "Mora", "Mada", "Mora"))
+
+ggplot(data = filter(par_df, par == "B_ttimes"), aes(x = model, y = values)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI))
+ggplot(data = filter(par_df, par == "intercept"), aes(x = model, y = values)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI))
+
+ttimes_plot <- seq(0, 15*60, by = 30)
+pop_plot <- 1e5
+
+## Mada
+plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
+     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+
+points(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, pch = 20, col = "darkblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+
+points(ttimes_plot/60, exp(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "purple")
+points(ttimes_plot/60, exp(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "magenta")
+points(ttimes_plot/60, exp(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
+       lty = 2, col = "magenta")
+legend("topright", c("Mada", "Mora", "CIs"), col = c("darkblue", "purple", "grey"), lty = c(1, 1, 2), 
+       bty = "n")
+
+## Data vs. predicted
+comm_data %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
+                   xlab = "Observed mean bites at district (Mada)", ylab = "Model predicted mean bites")
+abline(a = 0, b = 1, col = "blue", lty = 2)
+
+bites %>%
+      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
+plot(check$bites, check$bites_mu, bty = "n", pch = 20, 
+      xlab = "Observed mean bites at commune (Moramanga)", ylab = "Model predicted mean bites")
+abline(a = 0, b = 1, col = "blue", lty = 2)
+
+```
+
+### Adding dist with ctar vs. without
+
+```{r checking impact of ctar in}
+exps_dist$ctar_in <- dist_catchments$ctar_in_dist[match(exps_dist$district,
+                                                        dist_catchments$mdg_dis_co)]
+ggplot() + 
+  geom_boxplot(data = exps_dist, aes(x = as.factor(ctar_in), y = bites/pop))
+
+comm_data$ctar_in_comm <- comm_catchments$ctar_in_comm[match(comm_data$mdg_cm_,
+                                                        comm_catchments$mdg_cm_)]
+comm_data$ctar_in_dist <- dist_catchments$ctar_in_dist[match(comm_data$mdg_dis_co,
+                                                        dist_catchments$mdg_dis_co)]
+
+get.likelihood <- function(commune,district, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_in <- params[2]
+  B_0 <- params[3]
+
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_dist + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
+
+tmp <- optim(par = c(1e-6, 1e-6, 0.1), get.likelihood, district = exps_dist,
              commune = comm_data, hessian = TRUE)
@@
 B_ttimes <- params[1]
-B_0 <- params[2]
+B_in <- params[2]
+B_0 <- params[3]
 
@@
       group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_comm + B_0)*pop) -> check
 plot(check$ttimes_weighted, check$bites)
@@
       group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + + B_in*ctar_in_comm + B_0)*pop) %>%
       summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
+      left_join(select(exps_dist, district, dist_bites = bites), 
                 by = c("mdg_dis_co" = "district")) %>%
@@
 
-## Model of reported bites as a function of travel time
-As bite patient data were only available...
+```{burden by incidence}
 
-Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.
+## Mada
+plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
+     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
+       lty = 2, col = "lightblue")
+abline(h = pop_plot/25*0.01*0.39, col = "darkred") # low rabies incidence
+abline(h = pop_plot/5*0.01*0.39, col = "red") # high rabies incidence
 
-We fit a linear mixed effects model where the annual average bite incidence at the district level was predicted by the average travel time to the nearest CTAR for the district, the population within the district, and a random catchment level effect.
+bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
+R_100k_min <- pop_plot/25*0.01*0.39
+R_100k_max <- pop_plot/5*0.01*0.39
 
-For three other catchments with reporting above x% we used to out-of-fit predict...
-Ability to predict average annual/monthly bite incidence by travel time
-Ability to predict out-of-fit to other catchments
+plot(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, R_100k_min/bites_100k, 0.6))/R_100k_min, 
+     type = "l", bty = "n",
+     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, R_100k_max/bites_100k, 0.2))/R_100k_max,
+     col = "red")
 
-<br>
+ests_lower <- predict(mod_lower, newdata = as.data.frame(ttimes_plot), 
+                                type = "response")
+ests_upper <- predict(mod_upper, newdata = as.data.frame(ttimes_plot), 
+                                type = "response")
+
+plot(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*ests_upper > R_100k_min, R_100k_min/bites_100k,
+                        ests_upper))/R_100k_min, 
+     type = "l", bty = "n",
+     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*ests_lower > R_100k_max, R_100k_max/bites_100k,
+                        ests_lower))/R_100k_max,
+     col = "red")
+
+```
 
 ### Estimation of burden and reporting
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
-
-We calculated deaths as:
-$$deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}$$
-
-where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
-$$
-           p_{rabid}=
-            \begin{cases}
-            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
-            \frac{R_i}{B_i}, & \text{otherwise}
-            \end{cases}
-$$
-<br>
-
-such that rabid reported bites (i.e.
-$p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
-$R_{i} \times \rho_{max}$.
-$p_{rabid}$ is taken from Rajeev et al.
-2018 on contact tracing data from the Moramanga District.
-
-$R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
-
-<br>
-
-$$r \times D \times p_{exp}$$
-
-<br>
-
-where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
-2018).
-We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa(cite?).
-
-<br>
-
-## Estimating the impact of expanding PEP Access
-We use this framework to compare three scenarios of PEP provisioning in Madagascar:
-1. The baseline with the current clinic locations (n = 31)
-2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to all CSB IIs (n = ...?)
-
-We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
-
-<br>
-
-### Sensitivity Analyses
-- Sensitivity to assumptions of R
-- Sensitivity to assumptions of p_rabid
-
-```{r}
-## get vials delivered
-ctar_metadata %>% 
-  dplyr::select(id_ctar, starts_with("doses")) %>%
-  gather(year, vials_delivered, -id_ctar) %>%
-  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered
-
-## getting dose time series by clinic
-########### Need to fix dates first! ###############
-ctar_data %>%
-  select(id_ctar, datej0, datej3, datej7, datej28) %>%
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
-  gather(dose, date, -id_ctar) %>%
-  group_by(id_ctar, date) %>%
-  summarise(n = n()) -> dose_ts
-dose_ts <- dose_ts[!is.na(dose_ts$date), ]
-dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
-dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]
-
-## getting theoretical # vials by year
-dose_ts %>%
-  mutate(year = year(date), vials = ceiling(n/2)) %>%
-  group_by(id_ctar, year) %>%
-  summarize(vials_estimated = sum(vials)) %>%
-  left_join(vials_delivered) -> dose_comp
-
-######### Patient time series
-ctar_data %>%
-  select(id_ctar, date_de_consultation) %>%
-  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
-  gather(dose, date_de_consultation, -id_ctar) %>%
-  group_by(id_ctar, date_de_consultation) %>%
-  summarise(n = n()) -> patient_ts
-patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
-patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
-patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]
-
-ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
-  geom_tile(aes(fill = n)) +
-  scale_fill_continuous(type = "viridis", name = "Patients")
-
-patient_ts %>%
-  filter(id_ctar == 1) %$%
-  hist(n, breaks = 100, col = "blue", 
-       xlab = "Patients reporting", main = "Soanierana Ivongo")
-patient_ts %>%
-  filter(ctar == "Toamasina") %$%
-  hist(n, breaks = 100, 
-       xlab = "Patients reporting", main = "Toamasina", col = "red")
-
-## getting reporting
-start_date <- ymd("2014-01-01")
-end_date <- ymd("2017-12-31")
-ts <- as_tibble(seq(start_date, end_date, by = "day"))
-
-patient_ts %>%
-  right_join(ts, by = c("date_de_consultation" = "value")) %>%
-  select(id_ctar, date_de_consultation, n) %>%
-  spread(id_ctar, n) %>%
-  replace(., is.na(.), 0) -> doses_wide
-dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])
-
-date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
-for (j in 1:ncol(date_mat)){
-  rle(dose_mat[ , j]) %>%
-    unclass() %>%
-    as.data.frame() %>%
-    mutate(end = cumsum(lengths),
-         start = c(1, lag(end)[-1] + 1)) %>%
-    filter(values == 0, lengths >= 10) -> rles
-  for (i in 1:nrow(rles)){
-    date_mat[rles$start[i]:rles$end[i], j] <- 0
-  }
-}
-
-date_mat <- replace(date_mat, is.na(date_mat), 1)
-date_mat %>% 
-  as_tibble() %>%
-  group_by(date = year(doses_wide$date_de_consultation)) %>%
-  summarise_all(funs(sum(.)/n())) -> clinic_reporting
-colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
-clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
-clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
-                                                  ctar_metadata$id_ctar)]
-clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
-
-ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
-  geom_tile() +
-  scale_fill_gradient(low = "white", high = "blue")
-```
-
-## Submission of forms
-- Missingness seems to have been due to periods of time without form submission
-- Calculating periods of time
-
-```{r clean data}
-## districts that are non-reporter and data not available yet CTAR
-mada_district@data %>%
-select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
-
-to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
-to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$excl <- 0
-ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1
-
-ctar_data %>% 
-  filter(excl != 1) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
-  summarize(n = n()) -> exposures # get count of exposures
-
-exposures %>%
-  rename(date = year) %>%
-  left_join(covars) %>% # add pop + ttimes + reporting
-  left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) %>%
-  group_by(district, date) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
-exps_cleaned %>%
-  group_by(district) %>%
-  summarise(bites = mean(bites)) %>%
-  left_join(covars) -> exps_cleaned
-
-library(lme4)
-
-## poisson with total pop as offset
-mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
-                 data = exps_cleaned, family = poisson)
-par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
-plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Observed bites", ylab = "Predicted bites")
-abline(a = 0, b = 1, lty = 2, col = "blue")
-legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
-       inset = c(0.0, -0.1))
-plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
-     ylim = c(0, max(exps_cleaned$bites)))
-points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
-     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
-legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
-       pch = 20, bty = "n")
-```
 
@@
 ```
+
+### Sensitivity Analyses
+
+### Estimating the impact of expanding PEP Access
+
+<br>
+
+## Discussion
+
+### Key findings
+
+### Strengths and Limitations
+- Not accounting for clinic functioning
+- Not explicit data on reporting or rabies incidence!
+- Other factors that drive reporting
+- While we use a signficiant amount of assumptions, these are the same set of assumptions that underlie current rabies estimates, we're just improving them a bit...
+and ours are a likely a little bit more in line with reality @ a sub-national level
+
+### Broader context
+
+### Conclusions

### 2019-01-08 accessPEP.Rmd
added in taking out contacts by distribution of patients per day
Going to probably leave this out though...

@@
   group_by(id_ctar) %>%
-  summarize(prop_excluded = length(exclude_bydate[exclude_bydate == 1])/n()) %>%
-  ggplot(data = ., aes(y = id_ctar, prop_excluded)) +
-  geom_point()
+  summarize(prop_excluded = length(exclude_bydate[exclude_bydate == 1])/n()) -> summ_check
+
+#   ggplot(data = ., aes(y = id_ctar, prop_excluded)) +
+#   geom_point()
 

### 2019-01-09 accessPEP.Rmd
preliminary results on bites + burden

--- 2019-01-08 accessPEP.Rmd
@@
     reference_docx: format_docs/word_styles_reference_01.docx
-css: style.css
 ---
@@
 
-Access to healthcare is a significant driver of disease burden globally.
-Spatial access to care (cite Jess paper here).
-Rarely incorporated into national estimates of burden due to lack of data and tools for understanding...
+Access to healthcare is a significant driver of disease burden globally, and heterogeneities in access to care due to socioeconomic and geographic factors likely shape where and who are the most impacted by a disease, particularly at sub-national levels.
+However, these factors are rarely incorporated into estimates of burden due to limited data.
 
-<br>
-
-Canine rabies causes an estimated 59,000 human deaths annually.
-Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic.
-Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries.
-Often, the only data available are numbers of bite victims reporting to health facilities.
+Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
+However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
+Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
 The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
-the more overall bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+the morebites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+While at a national level, these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
 
-In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country.
+In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
 PEP is not available at any other public clinics or through the private sector.
-Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al.
-2018.
-We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.
-Assuming a range of exposure incidence given endemic rabies, we generate sub-national estimates of rabies burden and explore the impacts of expanding access to PEP in Madagascar.
-
-The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level.
-Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
-Here, we adapt this framework and use bite patient data from Madagascar to predict rabies burden.
+Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
+To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
+Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
+Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
 
@@
 ### Bite patient data
-We used the database of bite patient forms submitted from 24 ARMC across ARMC between 2014 - 2017.
-These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics.
+We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
+These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
 Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
-These data include the administrative district of the bite patient's address.
-As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted.
-We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.
-
-- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
-
-  - includes dates and home locations (to the district level) of patient reporting for PEP
-- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
-  - includes dates and home locations (to the commune level) of patients reporting for PEP
-
-RIG data?
+These data include the administrative district of the bite patient's address and the date of reporting.
+As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted, excluding any periods of time for which there were no forms submitted for 15 consecutive days.
+Clinic level reporting was calculated the proportion of days for which were included based on the 15 day threshold per year.
+As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
+2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
+Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We validated this method using 27 months of bite patient data from the Moramanga ARMC for which we had details on the type of exposure.
 
@@
 We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
-2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
+2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
 We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
-For each clinic, we defined the catchment area as all **districts?** for which the clinic was the closest ARMC.
+For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
 Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
@@
 ### Model of reported bites as a function of travel time
-While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX).
+While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX, *need to think about how to show this*).
 In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
@@
 
-As we had data available on reported bites at the commune level from the Moramanga ARMC, we modeled observed commune bites in the same framework, but un-aggregated where:
+As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
 $$ 
@@
 where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
-We only included communes which were designated to be within the catchment for the clinic (so primarily the Moramanga and Anosibe A'nala Districts)??
+We only included communes which were designated to be within the catchment for the clinic.
 We compared our estimates of $\beta_t$ (i.e.
@@
 
-To better understand if ttimes might capture access, we also looked at whether pop alone can predict bites and distance could predict bites...
-Ability to predict out-of-fit to other catchments? Tana + Fianarantsoa?
+### Model sensitivity/validation
+- Out-of-fit check for Tana + Fianarantsoa?
+- Compare to null models 
+- Look at sensitivity to inclusion criteria of days and exclusion of contacts
 
@@
 ### Estimation of burden and reporting 
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
 
-We calculated deaths as:
+We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) $:
 $$
-deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+  \overline\rho = \frac{\mu \times p_{rabid}}{R}
 $$
-
+or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
+We look at the range of $p_{rabid}$ reported in Rajeev et al.
+2018 for data from the Moramanga District (0.2 - 0.6).
 where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
@@
             \begin{cases}
-            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
-            \frac{R_i \times \rho_{max}}{B_i}, & \text{otherwise}
+            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
+            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
             \end{cases}
@@
 the area with the minimum travel time).
-We look at the range of $x$ reported in Rajeev et al.
-2018 for data from the Moramanga District (0.2 - 0.6).
-
-$R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:
+.
 
@@
 
-$$r \times D \times p_{exp}$$
-
-<br>
-
-where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
-2018).
+The rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018) to generate $R$.
 We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
@@
 
-<br>
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
+We assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
 
-### Sensitivity Analyses
-- R (increasing or decreasing with pop) (min + max)
-- P-rabid min and max
-- Rho-max min and max
-- Probability of death min and max
+*Need to include bits about RIG in here somewhere*
 
@@
 2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to all CSB IIs, something to describe CSB IIs here, that is primary hospitals (n = X)
+3. Expansion of ARMC to all CSB IIs (*something to describe CSB IIs here*), that is primary hospitals (n = X)
 
@@
 We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+### Sensitivity analyses for burden estimates 
+- R (increasing or decreasing with pop) (min + max)
+- P-rabid min and max
+- Probability of death min and max
+- Look at change in raw # of deaths and also change in burden by scenarios
+
+<br>
 
@@
 library(sp)
-
+library(MASS)
+library(reshape2)
 source("R/utils.R")
@@
 select <- dplyr::select
-knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
+knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
+                      results = "hide",
+                      include = TRUE)
 ```
@@
 
-## Centers with no data
+## Centers with no data ## Change this to be so that only ones with zero forms!
 no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
-             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
-             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
+             "Antsiranana", "Marolambo", "Nosy be", "Sainte Marie", "Vangaindrano")
 ctar_metadata$exclude <- 0
@@
 comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
-```
-
-### Clinic catchments and reporting
-For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).
-In addition, at the clinic level, the majority of bites reported to a given ARMC came from within the catchment (Figure 1C).
-
-```{r network fig}
-## Network fig
-## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments?
 
@@
 ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.4)
-display_palette(ctar_metadata$color)
+# display_palette(ctar_metadata$color)
 
@@
 mada_district@data <- left_join(mada_district@data, dist_catchments) ## need this?
+
+## Getting commune covars
+## Need: weighted ttimes, pop, ctar_in, exclude
+comm_catchments %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) %>%
+  left_join(select(mada_district@data, mdg_dis_co, district, ctar_in_dist, dist_catch = CTAR, exclude,
+                   color, fill, ctar_dist)) -> comm_data
+comm_data <- filter(comm_data, ttimes_weighted != "Inf")
+
+# ## Data to include
+# nrow(dist_catchments) - sum(dist_catchments$exclude)
+# nrow(comm_data) - sum(comm_data$exclude)
+
+#### Moramanga data! ###
+## ctar for # bites and also checking days to rule out contacts
+col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
+ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
+ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
+
+## Set-up time series
+tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
+                   year1 = 2016, tstep = "month", period = FALSE) {
+  get.consec (date, format_date, start, format_start,
+              year1, tstep, period) - 8 ## starts in Sep
+}
+
+## all dates
+ctar %>%
+  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
+ctar$month <- tfunc(ctar$date.reported)
+## subset to study period (Sep 2016 - to date)
+## checks for cases where month is NA!
+ctar <- subset(ctar, month > 0 & month < 28)
+
+# ctar district and commune
+ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
+ctar$commune <- trimws(ctar$commune, which = "right")
+ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
+ctar$district <- gsub(" \\(District\\)", "", ctar$district)
+
+## change all true/false to 0/1
+ctar[ctar == FALSE] <- 0
+
+## Bite data in Moramanga
+n_months <- max(ctar$month, na.rm = TRUE)
+
+ctar %>%
+  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
+            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
+  filter(consult.type == "Animal bite/exposure") %>%
+  group_by(mdg_cm_) %>%
+  summarize(bites = round(n()/n_months*12)) %>%
+  left_join(comm_data) %>%
+  filter(!is.na(ttimes_weighted), 
+         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
+```
+
+### Estimating average bite incidence at the district level
+```{r network fig}
+## Network fig
+## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments
 
@@
 dist_lines$size <- log(dist_lines$n + 0.1)*0.2
+```
 
+```{r figure 1A}
 pdf("figs/network.pdf")
@@
 dev.off()
+```
 
+```{r network output, echo=FALSE, fig.cap="Figure 1C", results = "show", out.width = '100%'}
+knitr::include_graphics("figs/network.pdf")
+```
+
+```{r figure 1b, fig.cap = "Figure 1B"}
 exp_mat %>% 
@@
   summarize(prop = sum(prop)) -> catch_ctar
+```
+
+```{r figure 1C, fig.cap = "Figure 1C"}
 hist(catch_ctar$prop, main = "", 
@@
 
-We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 10 consecutive days (Figure 2).
-We further excluded any years for which there was less than 25% reporting.
+For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).
+In addition, at the clinic level, the majority of bites reported to a given ARMC came from within the catchment (Figure 1C).
 
-```{r clinic level reporting}
+```{r clinic level reporting, fig.cap = "Figure 2A"}
 ######### Patient time series
@@
 
-## Color or size dots by their # from that year
-# ctar_data %>%
-#   group_by(date_de_consultation = year(date_de_consultation), id_ctar) %>%
-#   summarize(tot = n()) -> year_by_clinic
+date_mat <- as.data.frame(get.days(threshold = 15)[[2]])
+date_mat$date_de_consultation <- doses_wide$date_de_consultation
+date_mat <-melt(date_mat, id = "date_de_consultation")
+levels(date_mat$variable) <- names(doses_wide)[2:23]
+names(date_mat) <- c("date_de_consultation", "id_ctar", "exclude")
 
-clinic_reporting <- get.days(threshold = 10)[[1]]
-# clinic_reporting %>%
-#   mutate(prop10 = prop, prop5 = get.days(threshold = 5)[[1]]$prop, 
-#          prop30 = get.days(threshold = 30)[[1]]$prop) %>%
-#   left_join(year_by_clinic) %>%
-#   ggplot(., aes(x = reorder(ctar, prop10), y = prop10)) + 
-#   geom_point() +
-#   geom_linerange(aes(ymin = prop30, ymax = prop5)) +
-#   facet_wrap(~ date_de_consultation, nrow = 1) +
-#   coord_flip() +
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-#   xlab("") + 
-#   ylab("Proportion of days included") +
-#   scale_color_gradient(low = "white", high = "blue")
-# 
-# date_mat <- as.data.frame(get.days(threshold = 10)[[2]])
-# date_mat$date_de_consultation <- doses_wide$date_de_consultation
-# names(date_mat[1:(ncol(date_mat)-1)]) <- names(doses_wide[2:(ncol(date_mat))])
-# library(reshape2)
-# head(melt(date_mat, id = "date_de_consultation"))
+patient_ts %>%
+  ungroup() %>%
+  mutate(id_ctar = as.factor(id_ctar)) %>%
+  right_join(date_mat) %>%
+  mutate(n = replace_na(n, 0), 
+         n = ifelse(exclude == 1, n, NA)) -> doses_long
+doses_long$ctar <- ctar_metadata$CTAR[match(doses_long$id_ctar, ctar_metadata$id_ctar)]
 
-ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
+ggplot(data = doses_long, aes(date_de_consultation, ctar)) + 
    geom_tile(aes(fill = n)) +
-   scale_fill_continuous(type = "viridis", name = "Patients") +
+   scale_fill_gradientn(colours = c("white", "purple", "black"), values = c(0, 0.1, 1),
+                        na.value = alpha("lightgrey", 1)) +
    xlim(start_date, end_date)
 
-ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
-  geom_tile() +
-  scale_fill_gradient(low = "white", high = "blue")
+clinic_reporting <- get.days(threshold = 15)[[1]]
 
@@
 
-```{r getting Mora data}
+```{r figure 2B, fig.cap = "Figure 2B"}
+
+clinic_reporting %>%
+  mutate(prop15 = prop, prop5 = get.days(threshold = 5)[[1]]$prop,
+         prop30 = get.days(threshold = 30)[[1]]$prop) %>%
+  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
+  geom_hline(yintercept = 0.25, color = "darkgrey") +
+  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
+               stat = "identity") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
+  xlab("") +
+  ylab("Proportion of days included") +
+  coord_flip()
+
+```
+
+We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure 2A).
+This estimate did vary based on our assumption of the threshold number of consecutive days (Figure 2B), so we looked at sensitivity of parameter estimates to changing this threshold (Supplementary Materials?).
+In all cases, we further excluded any years for which there was less than 25% reporting.
+
+``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
+## Testing with Moramanga data 
+sdev <- seq(2, 10, by = 1)
+check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
+for (i in 1:length(sdev)){
+  ctar %>%
+    group_by(date.reported) %>%
+    summarize(n = n()) %>%
+    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
+  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
+  check <- rbind(check, c(sdev[i], checksd[1:2]))
+}
+
+plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
+points(check$sd, check$prop_bites, col = "red", pch = 20)
+```
+
+```{r figure 3B, fig.cap= "Figure 3B"}
 patient_ts %>%
   group_by(ctar) %>%
-  mutate(exclude = ifelse(n > mean(n) + 6*sd(n), 1, 0)) -> patient_ts
+  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
 
@@
 
-(ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/
-  length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])
+```
+
+We attempted to exclude these by looking at the distribution of patients reporting per day to a given clini, excluding patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We validated this using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
+For the national data for which a subset of patient forms were noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
+We further excluded known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
+
+```{r district level exposures, fig.cap = "Figure 4"}
+## districts that are non-reporter and data not available yet CTAR
+mada_district@data %>%
+select(district = mdg_dis_co, dist_name = district, pop = pop2015adj, ttimes_weighted_dist = ttimes_weighted, 
+       dist_catch = CTAR, exclude, color) -> covars
 
 ctar_check %>% 
-  group_by(id_ctar) %>%
-  summarize(prop_excluded = length(exclude_bydate[exclude_bydate == 1])/n()) %>%
-  ggplot(data = ., aes(y = id_ctar, prop_excluded)) +
-  geom_point()
-
-```
-
-```{r ttimes + pop at commune level}
-## Need: weighted ttimes, pop, ctar_in, exclude
-comm_catchments %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) -> comm_covars
-comm_covars <- filter(comm_covars, ttimes_weighted != "Inf")
-comm_covars %>% 
-  left_join(select(mada_district@data, mdg_dis_co, district)) -> comm_data
-comm_data$exclude <- dist_catchments$exclude[match(comm_data$mdg_dis_co, dist_catchments$mdg_dis_co)]
-
-## Data to include
-nrow(comm_data) - sum(comm_data$exclude)
-```
-
-``` {r check with moramanga}
-## ctar for # bites and also checking days to rule out contacts
-col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
-ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
-ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
-
-## merge with newest data!
-
-## Set-up time series
-tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
-                   year1 = 2016, tstep = "month", period = FALSE) {
-  get.consec (date, format_date, start, format_start,
-              year1, tstep, period) - 8 ## starts in Sep
-}
-
-## all dates
-ctar %>%
-  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
-ctar$month <- tfunc(ctar$date.reported)
-## subset to study period (Sep 2016 - to date)
-## checks for cases where month is NA!
-ctar <- subset(ctar, month > 0 & month < 28)
-
-# ctar district and commune
-ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
-ctar$commune <- trimws(ctar$commune, which = "right")
-ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
-ctar$district <- gsub(" \\(District\\)", "", ctar$district)
-
-## change all true/false to 0/1
-ctar[ctar == FALSE] <- 0
-
-## Bite data in Moramanga
-n_months <- max(ctar$month, na.rm = TRUE)
-
-ctar %>%
-  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
-            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
-  filter(consult.type == "Animal bite/exposure") %>%
-  group_by(mdg_cm_) %>%
-  summarize(bites = round(n()/n_months*12)) %>%
-  left_join(comm_covars) %>%
-  filter(!is.na(ttimes_weighted), 
-         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
-
-ctar %>%
-  group_by(date.reported) %>%
-  summarize(n = n()) %>%
-  mutate(exclude = ifelse(n > mean(n) + 2*sd(n), 1, 0)) -> mora_ts
-
-ggplot(data = mora_ts, aes(x = "Moramanga", y = n, color = as.factor(exclude))) +
-  geom_jitter(alpha = 0.5, width = 0.1) +
-  coord_flip()
-
-ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
-table(ctar$consult.type, ctar$exclude)
-
-sdev <- seq(2, 10, by = 1)
-check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
-for (i in 1:length(sdev)){
-  ctar %>%
-    group_by(date.reported) %>%
-    summarize(n = n()) %>%
-    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
-  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
-  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
-  check <- rbind(check, c(sdev[i], checksd[1:2]))
-}
-
-plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
-points(check$sd, check$prop_bites, col = "red", pch = 20)
-
-```
-
-### Models of bites as a function of travel times
-
-```{r district level exposures}
-## districts that are non-reporter and data not available yet CTAR
-mada_district@data %>%
-select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars
-
-ctar_check %>% 
-  filter(exclude_bydate == 0, exclud_bycomment == 0) %>% # exclude contacts
+  filter(exclude_bydate == 0, exclude_bycomment == 0) %>% # exclude contacts
   mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
@@
   left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) %>%
+  filter(exclude == 0, prop > 0.25) %>% ## filter out excluded catchments and ones with < 25% reporting
   group_by(district, year) %>%
@@
   summarise(bites = sum(bites, na.rm = TRUE)) %>%
+  left_join(covars) -> exps
+
+col_vals <- dist_catchments$color
+names(col_vals) <- dist_catchments$mdg_dis_co
+
+ggplot(exps, aes(x = reorder(dist_name, ttimes_weighted_dist), y = bites/pop*1e5, color = district)) +
+  geom_boxplot() +
+  theme(legend.position = "none") +
+  scale_color_manual(values = col_vals) + 
+  ylab("Annual bites per 100k") +
+  xlab("") +
+  coord_flip()
+
+exps %>%
   group_by(district) %>%
-  summarise(bites = mean(bites)) %>%
-  left_join(covars) -> exps_check
-
+  summarise(bites = mean(bites), n = n()) %>%
+  left_join(covars) -> exps_dist
 ```
 
-### Model results Moramanga and District
-```{r likelihood of observing moramanga data @ commune level}
+<br>
+
+Our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
+
+### Models of bites as a function of travel times
+```{r likelihood of observing moramanga data @ commune level, fig.cap = "Figure 5A"}
 
@@
 
-get.likelihood <- function(commune, district, params) {
+get.likelihood.mada <- function(commune, district, params) {
   # To test likelihood function
@@
 }
-```
 
-```{r likelihood of observing district data given sum of communes in district}
 mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
@@
 varcov <- solve(mora_pars$hessian)
-library(MASS)
 
@@
 
-mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_check,
+mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mada, district = exps_dist,
              commune = comm_data, hessian = TRUE)
@@
 varcov <- solve(mada_pars$hessian)
-library(MASS)
 
@@
 B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
-par_df <- data.frame(values = c(B_ttimes_mada, B_ttimes_mora, B_0_mada, B_0_mora),  
+par_df <- data.frame(model = c("Mada", "Mora", "Mada", "Mora"),
                      par = c("B_ttimes", "B_ttimes", "intercept", "intercept"),
+                     values = c(B_ttimes_mada, B_ttimes_mora, B_0_mada, B_0_mora),
                      upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
@@
                      lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
-                                  CI = B_0_mada_CI[1], B_0_mora_CI[1]),
-                     model = c("Mada", "Mora", "Mada", "Mora"))
-
-ggplot(data = filter(par_df, par == "B_ttimes"), aes(x = model, y = values)) +
-  geom_point() +
-  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI))
-ggplot(data = filter(par_df, par == "intercept"), aes(x = model, y = values)) +
-  geom_point() +
-  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI))
-
+                                  CI = B_0_mada_CI[1], B_0_mora_CI[1]))
+kable(par_df, caption = "Table 1: summary of parameter estimates")
 ttimes_plot <- seq(0, 15*60, by = 30)
@@
 ## Mada
-plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
+plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "l",
      ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
@@
 
-points(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, pch = 20, col = "darkblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
+points(ttimes_plot/60, exp(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "red")
+points(ttimes_plot/60, exp(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "pink")
+points(ttimes_plot/60, exp(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
+       lty = 2, col = "pink")
+legend("topright", c("Mada", "Mora", "CIs"), col = c("darkblue", "red", "grey"), lty = c(1, 1, 2), 
+       bty = "n")
+```
 
-points(ttimes_plot/60, exp(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "purple")
-points(ttimes_plot/60, exp(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "magenta")
-points(ttimes_plot/60, exp(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
-       lty = 2, col = "magenta")
-legend("topright", c("Mada", "Mora", "CIs"), col = c("darkblue", "purple", "grey"), lty = c(1, 1, 2), 
-       bty = "n")
+```{r model table, results = "show"}
+kable(par_df, caption = "Table 1: summary of parameter estimates")
+```
 
+```{r figure 5B, fig.cap="Figure 5B"}
 ## Data vs. predicted
@@
       filter(exclude == 0) %>%
+      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
       group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
       summarize(sum_bites = sum(bites)) %>%
@@
       filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
-                   xlab = "Observed mean bites at district (Mada)", ylab = "Model predicted mean bites")
-abline(a = 0, b = 1, col = "blue", lty = 2)
+sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
+                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
+abline(a = 0, b = 1, col = "grey", lty = 2)
 
-bites %>%
-      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
-plot(check$bites, check$bites_mu, bty = "n", pch = 20, 
-      xlab = "Observed mean bites at commune (Moramanga)", ylab = "Model predicted mean bites")
-abline(a = 0, b = 1, col = "blue", lty = 2)
+mora_bites %>%
+      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
+points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))
 
@@
 
-### Adding dist with ctar vs. without
+<br>
 
-```{r checking impact of ctar in}
-exps_dist$ctar_in <- dist_catchments$ctar_in_dist[match(exps_dist$district,
-                                                        dist_catchments$mdg_dis_co)]
-ggplot() + 
-  geom_boxplot(data = exps_dist, aes(x = as.factor(ctar_in), y = bites/pop))
+We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC.
+Both models produced reasonable fits to the data (Figure 5B,...), however, there was some variation in bite incidence which both models were not able to capture.
 
-comm_data$ctar_in_comm <- comm_catchments$ctar_in_comm[match(comm_data$mdg_cm_,
-                                                        comm_catchments$mdg_cm_)]
-comm_data$ctar_in_dist <- dist_catchments$ctar_in_dist[match(comm_data$mdg_dis_co,
-                                                        dist_catchments$mdg_dis_co)]
+### Model sensitivity analyses
+- compare to null models: flat incidence, incidence solely predicted by pop;
+- sensitivity of param estimates to reporting cut-offs (i.e. days excluded)
+- sensitivity of param estimates to exclusion of contacts
 
-get.likelihood <- function(commune,district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_in <- params[2]
-  B_0 <- params[3]
+### Estimation of burden and reporting
 
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_dist + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-
-tmp <- optim(par = c(1e-6, 1e-6, 0.1), get.likelihood, district = exps_dist,
-             commune = comm_data, hessian = TRUE)
-params <- tmp$par
-B_ttimes <- params[1]
-B_in <- params[2]
-B_0 <- params[3]
-
-comm_data %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_in*ctar_in_comm + B_0)*pop) -> check
-plot(check$ttimes_weighted, check$bites)
-
-comm_data %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + + B_in*ctar_in_comm + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites)
-abline(a = 0, b = 1, col = "blue", lty = 2)
-
-```
-
-```{burden by incidence}
-
+```{r reporting by incidence, fig.cap="Figure 6"}
 ## Mada
-plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
-     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-abline(h = pop_plot/25*0.01*0.39, col = "darkred") # low rabies incidence
-abline(h = pop_plot/5*0.01*0.39, col = "red") # high rabies incidence
-
 bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
@@
 plot(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, R_100k_min/bites_100k, 0.6))/R_100k_min, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
      type = "l", bty = "n",
-     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
+     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
 lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, R_100k_max/bites_100k, 0.2))/R_100k_max,
-     col = "red")
-
-ests_lower <- predict(mod_lower, newdata = as.data.frame(ttimes_plot), 
-                                type = "response")
-ests_upper <- predict(mod_upper, newdata = as.data.frame(ttimes_plot), 
-                                type = "response")
-
-plot(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*ests_upper > R_100k_min, R_100k_min/bites_100k,
-                        ests_upper))/R_100k_min, 
-     type = "l", bty = "n",
-     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+     col = "darkred")
 lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*ests_lower > R_100k_max, R_100k_max/bites_100k,
-                        ests_lower))/R_100k_max,
-     col = "red")
-
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
+     col = "red", lty = 2)
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
+     col = "red", lty = 2)
+polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
+        border = NA, col = alpha("red", 0.2))
+polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
+        border = NA, col = alpha("red", 0.6))
+legend("topright", inset = c(-0.20, 0), c("Max rabies exposures", "Min rabies exposures"), lty = c(2, 1), 
+       col = c("red", "darkred"),
+       bty = "n", xpd = TRUE)
+legend("topright", inset = c(0.01, 0.2), c("p_rabid = 0.2", "p_rabid = 0.6"), border = NA, 
+       fill = c(alpha("red", 0.2), alpha("red", 0.6)), bty = "n", xpd = TRUE)
+abline(v = max(ttimes_plot[bites_100k*0.6 > R_100k_min])/60, col = "grey", lty = 3)
+abline(v = max(ttimes_plot[bites_100k*0.2 > R_100k_max])/60, col = "grey", lty = 3)
 ```
 
-### Estimation of burden and reporting
+Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 6).
+Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5.5 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).
 
 ```{r functions for getting burden}
-get.burden <- function(covs = covars, pop = covars$pop, model = mod,
-                       p_rabid = 0.5, rho_max = 0.9,
-                       max_HDR = 25, min_HDR = 5, 
-                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                       prob_death = 0.16, nsims = 1000) {
-  store_deaths <- store_p_rabid <- matrix(NA, nrow(covs), nsims)
-
-  for (i in 1:nsims) {
-    mean_bites <- predict(model, type = "response", newdata = covs)
-    bites <- rpois(length(mean_bites), mean_bites)
-    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
-                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
-    rabid_exps <- rpois(length(mean_bites), mean_exps)
-    p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid)
-    deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), prob_death)
-    store_deaths[, i] <- deaths
-    store_p_rabid[, i] <- p_rabid_t
-  }
-
-  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
-  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
-                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
-  results <- cbind(covars, results)
-  return(results)
-}
-
-get.burden.flat <- function(covs = covars, pop = covars$pop, model = mod,
-                       p_rabid = 0.5, rho = 0.85, 
-                       prob_death = 0.16, nsims = 1000) {
-  store_deaths <- matrix(NA, nrow(covs), nsims)
-
-  for (i in 1:nsims) {
-    mean_bites <- predict(model, type = "response", newdata = covs)
-    bites <- rpois(length(mean_bites), mean_bites)
-    unreported <- bites*p_rabid/rho - bites*p_rabid
-    deaths <- rbinom(length(mean_bites), round(unreported), prob_death)
-    store_deaths[, i] <- deaths
-  }
-  results <- get.meanCI(store_deaths)
-  colnames(results) <- c("deaths_mean", "death_lowerCI", "deaths_upperCI")
-  results <- cbind(covars, results)
-  return(results)
-}
-
-get.meanCI <- function(matrix_by_row, nboots = 1000) {
-    mean_rows <- apply(matrix_by_row, 1, mean)
-    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
-    lower <- unlist(CIs)[1, ] 
-    upper <- unlist(CIs)[2, ]
-    summ <- cbind(mean_rows, lower, upper)
-    colnames(summ) <- c("mean", "lower", "upper")
-    return(summ)
-}
 
@@
   deltastar = bsmeans - xbar
-  d = quantile(deltastar, c(0.05, 0.95))
+  d = quantile(deltastar, c(0.05, 0.95), na.rm = TRUE)
   ## Get from quantiles
@@
 
+get.meanCI <- function(matrix_by_row, nboots = 1000) {
+    mean_rows <- apply(matrix_by_row, 1, mean)
+    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
+    lower <- unlist(CIs)[1, ] 
+    upper <- unlist(CIs)[2, ]
+    summ <- cbind(mean_rows, lower, upper)
+    colnames(summ) <- c("mean", "lower", "upper")
+    return(summ)
+}
+
+get.burden <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
+                       param_ttimes = B_ttimes_mada,
+                       param_intercept = B_0_mada,
+                       p_rabid = 0.5, rho_max = 0.9,
+                       max_HDR = 25, min_HDR = 5, 
+                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                       prob_death = 0.16, nsims = 1000) {
+
+  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)
+
+  for (i in 1:nsims) {
+    ## for testing
+    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
+    #                    param_ttimes = B_ttimes_mada;
+    #                    param_intercept = B_0_mada;
+    #                    p_rabid = 0.5; rho_max = 0.9;
+    #                    max_HDR = 25; min_HDR = 5;
+    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
+    #                    prob_death = 0.16;
+
+    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
+    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
+                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
+    rabid_exps <- rpois(length(mean_exps), mean_exps)
+    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
+                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
+    reported_rabid <- round(bites*p_rabid_t)
+    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
+    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
+    store_deaths[, i] <- deaths
+    store_p_rabid[, i] <- p_rabid_t
+  }
+
+  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
+  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
+                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
+  results <- cbind(covars, results)
+  return(results)
+}
+
+get.burden.flat <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
+                       param_ttimes = B_ttimes_mada,
+                       param_intercept = B_0_mada,
+                       p_rabid = 0.5, rho = 0.9,
+                       max_HDR = 25, min_HDR = 5, 
+                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                       prob_death = 0.16, nsims = 1000) {
+
+  store_deaths <- matrix(NA, length(ttimes), nsims)
+
+  for (i in 1:nsims) {
+    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
+    reported_rabid <- rbinom(length(ttimes), size = bites, prob = 0.5)
+    total_rabid <- round(reported_rabid/rho)
+    deaths <- rbinom(length(mean_bites), total_rabid - reported_rabid, prob_death)
+    store_deaths[, i] <- deaths
+  }
+
+  results <- get.meanCI(store_deaths)
+  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI")
+  results <- cbind(covars, results)
+  return(results)
+}
 ```
 
-```{r}
+```{r burden results, fig.cap = "Figure 7", cache = TRUE}
 results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
 results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
+results_pr20$pr <- "0.20"
+results_pr60$pr <- "0.60"
+results <- rbind(results_pr20, results_pr60)
 
-plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
-points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 20)
-segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
-         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
-
-results_base <- get.burden(p_rabid = 0.40, rho_max = 0.98)
-
-p_rabid_check <- seq(0.10, 0.80, by = 0.1)
-check <- data.frame(p_rabid = 0, district = NA, ttimes = 0, deaths_mean = 0)
-for (i in 1:length(p_rabid_check)) {
-  print(paste(i, "/", length(p_rabid_check)))
-  results <- get.burden(p_rabid = p_rabid_check[i], rho_max = 0.98)
-  p_rabid <- rep(p_rabid_check[i], length(results$deaths_mean))
-  check_one <- cbind(p_rabid, results$district, results$ttimes, results$deaths_mean)
-  colnames(check_one) <- c("p_rabid", "district", "ttimes", "deaths_mean")
-  check <- rbind(check, check_one)
-}
-
-check <- check[-1, ]
-check$base <- rep(results_base$deaths_mean, length(p_rabid_check))
-ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
-  geom_tile(aes(fill = deaths_mean/base)) +
-  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 1, name = "Mean deaths")
-
-rho_max_check <- seq(0.90, 0.99, by = 0.01)
-check <- data.frame(rho_max = 0, district = NA, ttimes = 0, deaths_mean = 0)
-for (i in 1:length(rho_max_check)) {
-  print(paste(i, "/", length(rho_max_check)))
-  results <- get.burden(p_rabid = 0.4, rho_max = rho_max_check[i])
-  rho_max <- rep(rho_max_check[i], length(results$deaths_mean))
-  check_one <- cbind(rho_max, results$district, results$ttimes, results$deaths_mean)
-  colnames(check_one) <- c("rho_max", "district", "ttimes", "deaths_mean")
-  check <- rbind(check, check_one)
-}
-
-check <- check[-1, ]
-check$base <- rep(results_base$deaths_mean, length(rho_max_check))
-check %>%
-  group_by(district) %>%
-  mutate(min = min(deaths_mean), max = max(deaths_mean), 
-         standardized = (deaths_mean - base)/(max - base)) -> check
-
-ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
-  geom_tile(aes(fill = standardized)) +
-  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths") 
-ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
-  geom_tile(aes(fill = deaths_mean)) +
-  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths")
-
-mada_dist_gg <- tidy(mada_district, region = "mdg_dis_co")
-mada_dist_gg <- left_join(mada_dist_gg, exps_cleaned, by = c("id" = "district"))
-
-ggplot() +
-  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = CTAR)) +
-  scale_colour_viridis_d()
+results %>% 
+  group_by(mdg_dis_co, pr) %>% 
+  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI), lowerCI = sum(deaths_lowerCI)) %>%
+  left_join(dist_catchments) %>% 
+  left_join(select(mada_district@data, mdg_dis_co, district)) %>%
+  ggplot(., aes(y = reorder(district, ttimes_weighted), x = tot, color = pr)) +
+  ggstance::geom_pointrangeh(aes(xmin = lowerCI, xmax = upperCI), fatten = 2) +
+  ylab("") +
+  xlab("Average annual deaths") +
+  scale_color_manual(values = c("0.20" = alpha("red", 0.2), "0.60" = alpha("red", 0.6)), 
+                     name = "Proportion rabid") -> p
+p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
 
@@
 
-### Sensitivity Analyses
+When we estimate burden of deaths stochastically within this range of incidence and given our high and low estimates of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure 7, results presented aggregated at the district level).
+Overall, we estimate average annual deaths between `r round(sum(results_pr60$deaths_mean))` - `r round(sum(results_pr20$deaths_mean))`.
+**Also estimate deaths averted here!**.
 
-### Estimating the impact of expanding PEP Access
+### Burden sensitivity analyses
+- Compare to assuming flat reporting + bite model?  Burden @ district level with bites @ district level?
+- $\rho_{max}$, change by urban vs. rural commune?
+- If rabies incidence, $R$, scales with population size within our estimated range
+- Impacts on # of deaths vs. change in scenarios (i.e. focusing on relative decreases in incidence of deaths based on expanded access rather than raw #s)
 
@@
 - While we use a signficiant amount of assumptions, these are the same set of assumptions that underlie current rabies estimates, we're just improving them a bit...
-and ours are a likely a little bit more in line with reality @ a sub-national level
+and ours are a likely a little bit more in line with reality @ a sub-national level?
 

### 2019-01-09 archive/accessPEP_summary.Rmd
archiving older files

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
css: style.css
---

## Goals
- To estimate burden of rabies in Madagascar using nationally available data from clinics that provide PEP.
- To explore the potential impacts of expanded access to PEP

## Data
- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
  - includes dates and home locations (to the district level) of patient reporting for PEP
- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
  - includes dates and home locations (to the commune level) of patients reporting for PEP
  - also includes rabies exposure status based on a combination of clinic-based triage and contact
  tracing (Rajeev et al. 2018)
- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
  - We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. 
  - We then calculated a weighted average by human population to the commune or district level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. 
  - For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the designated admin level.

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
```

```{r CTAR data}

```

## Cleaning the data
```{r get data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Get color palette from other network fig!
img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
pal <- image_palette(img, 31)
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
## Getting ctar district and commune
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_

display_palette(ctar_metadata$color)
dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments)
dist_catchments$ctar_in <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
```
1. Dealing with underreporting at the clinic level
2. Matching names to the commune level
3. Estimating catchments and weighted travel times by admin level
```{r catchment figure}
## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network_1.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
       , pch = 1, cex = dist_points$size, 
       col = dist_lines$color)
segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
         lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

legend("topleft", c(">100", ">200", ">400", ">800"), title = "Number of total reported bites", 
       text.width = 3, pt.cex = circle_size, col = "grey",  pch = 16, inset = c(0.1, 0.1), lwd = 1.5, bty = "n", xpd = TRUE, y.intersp=2)

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop)
ggplot(data = catch_check, aes(CTAR.x, district)) + 
  geom_tile(aes(fill = prop)) +
  scale_fill_continuous(type = "viridis", name = "proportion") +
  geom_text(label = catch_check$catch_match)

```

```{r}
## get vials delivered
ctar_metadata %>% 
  dplyr::select(id_ctar, starts_with("doses")) %>%
  gather(year, vials_delivered, -id_ctar) %>%
  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered

## getting dose time series by clinic
########### Need to fix dates first! ###############
ctar_data %>%
  select(id_ctar, datej0, datej3, datej7, datej28) %>%
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
  gather(dose, date, -id_ctar) %>%
  group_by(id_ctar, date) %>%
  summarise(n = n()) -> dose_ts
dose_ts <- dose_ts[!is.na(dose_ts$date), ]
dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]

## getting theoretical # vials by year
dose_ts %>%
  mutate(year = year(date), vials = ceiling(n/2)) %>%
  group_by(id_ctar, year) %>%
  summarize(vials_estimated = sum(vials)) %>%
  left_join(vials_delivered) -> dose_comp

######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

patient_ts %>%
  filter(id_ctar == 1) %$%
  hist(n, breaks = 100, col = "blue", 
       xlab = "Patients reporting", main = "Soanierana Ivongo")
patient_ts %>%
  filter(ctar == "Toamasina") %$%
  hist(n, breaks = 100, 
       xlab = "Patients reporting", main = "Toamasina", col = "red")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```
4. Contacts + other issues (i.e. date errors)
5. Corrected incidence

## Models of bites
We assume that the incidence of bites at the district/commune level, reflects differences in access to PEP. 
We fit a generalized linear model with the annual average bites modeled as a rate per population (with a log offset oft the population) for three datasets:
1. National data corrected and averaged to the district level
2. National data corrected and averaged to the commune level
3. Moramanga data

### Results of models
#### Fit
#### Compare relationships

## Estimating burden
We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies.

Using the standard decision tree framework, we assumed that reported bites were proportional to rabies incidence, and calculated deaths, assuming uniform reporting across Madagascar (taken from Moramanga paper = 0.85):
$$
  deaths_{i} = (frac{{p_{rabid} \times B_{i}}{\rho} - p_{rabid} \times B_{i}) \times p_{death}
$$

Alternaetively, we calculated deaths based on assumptions about endemic incidence of rabies exposures:
$$
  deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}
$$

where $R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

<br>

$$r \times D \times p_{exp}$$

<br>

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We estimate the dog population by using a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa(cite?).

The proportion of reported bites that are rabies exposures ($p_{rabid}$) are also constrained by the number of rabies exposures:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
            \frac{R_i}{B_i}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_{i} \times \rho_{max}$. The range of $p_{rabid}$ and $\rho_max$ are taken from Rajeev et al. 2018 on contact tracing data from the Moramanga District.

Finally, we also accounted for differences in reporting probability of rabid vs. non-rabid bites as a function of travel times. Using the Moramanga data, we looked at how travel times predict the $p_{rabid}$ so that:
model here

In our estimation of burden, we predict what p_rabid is but also similarly constrain it so that rabid reported bites cannot exceed total reported bites.

## Estimating the impact of expanding PEP Access
We use this framework to compare three scenarios of PEP provisioning in Madagascar:
1. The baseline with the current clinic locations (n = 31)
2. Expansion of ARMC to one clinic per district (n = 114)
3. Expansion of ARMC to all CSB IIs (n = ...?)

We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3. We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.

<br>

### Sensitivity Analyses
- Sensitivity to assumptions of R
  - pop?
- Sensitivity to assumptions of p_rabid and p_max

```{r catchment figure}
## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network_1.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
       , pch = 1, cex = dist_points$size, 
       col = dist_lines$color)
segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
         lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

legend("topleft", c(">100", ">200", ">400", ">800"), title = "Number of total reported bites", 
       text.width = 3, pt.cex = circle_size, col = "grey",  pch = 16, inset = c(0.1, 0.1), lwd = 1.5, bty = "n", xpd = TRUE, y.intersp=2)

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop)
ggplot(data = catch_check, aes(CTAR.x, district)) + 
  geom_tile(aes(fill = prop)) +
  scale_fill_continuous(type = "viridis", name = "proportion") +
  geom_text(label = catch_check$catch_match)

```

```{r}
## get vials delivered
ctar_metadata %>% 
  dplyr::select(id_ctar, starts_with("doses")) %>%
  gather(year, vials_delivered, -id_ctar) %>%
  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered

## getting dose time series by clinic
########### Need to fix dates first! ###############
ctar_data %>%
  select(id_ctar, datej0, datej3, datej7, datej28) %>%
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
  gather(dose, date, -id_ctar) %>%
  group_by(id_ctar, date) %>%
  summarise(n = n()) -> dose_ts
dose_ts <- dose_ts[!is.na(dose_ts$date), ]
dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]

## getting theoretical # vials by year
dose_ts %>%
  mutate(year = year(date), vials = ceiling(n/2)) %>%
  group_by(id_ctar, year) %>%
  summarize(vials_estimated = sum(vials)) %>%
  left_join(vials_delivered) -> dose_comp

######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

patient_ts %>%
  filter(id_ctar == 1) %$%
  hist(n, breaks = 100, col = "blue", 
       xlab = "Patients reporting", main = "Soanierana Ivongo")
patient_ts %>%
  filter(ctar == "Toamasina") %$%
  hist(n, breaks = 100, 
       xlab = "Patients reporting", main = "Toamasina", col = "red")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```

## Submission of forms
- Missingness seems to have been due to periods of time without form submission
- Calculating periods of time

```{r clean data}
## districts that are non-reporter and data not available yet CTAR
mada_district@data %>%
select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$excl <- 0
ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) %>%
  group_by(district, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
exps_cleaned %>%
  group_by(district) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_cleaned

library(lme4)

## poisson with total pop as offset
mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
                 data = exps_cleaned, family = poisson)
par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
     xlab = "Observed bites", ylab = "Predicted bites")
abline(a = 0, b = 1, lty = 2, col = "blue")
legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
       inset = c(0.0, -0.1))
plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
     ylim = c(0, max(exps_cleaned$bites)))
points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
       pch = 20, bty = "n")
```

```{r functions for getting burden}
get.burden <- function(covs = covars, pop = covars$pop, model = mod,
                       p_rabid = 0.5, rho_max = 0.9,
                       max_HDR = 25, min_HDR = 5, 
                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                       prob_death = 0.16, nsims = 1000) {
  store_deaths <- store_p_rabid <- matrix(NA, nrow(covs), nsims)

  for (i in 1:nsims) {
    mean_bites <- predict(model, type = "response", newdata = covs)
    bites <- rpois(length(mean_bites), mean_bites)
    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
    rabid_exps <- rpois(length(mean_bites), mean_exps)
    p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid)
    deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), prob_death)
    store_deaths[, i] <- deaths
    store_p_rabid[, i] <- p_rabid_t
  }

  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.burden.flat <- function(covs = covars, pop = covars$pop, model = mod,
                       p_rabid = 0.5, rho = 0.85, 
                       prob_death = 0.16, nsims = 1000) {
  store_deaths <- matrix(NA, nrow(covs), nsims)

  for (i in 1:nsims) {
    mean_bites <- predict(model, type = "response", newdata = covs)
    bites <- rpois(length(mean_bites), mean_bites)
    unreported <- bites*p_rabid/rho - bites*p_rabid
    deaths <- rbinom(length(mean_bites), round(unreported), prob_death)
    store_deaths[, i] <- deaths
  }
  results <- get.meanCI(store_deaths)
  colnames(results) <- c("deaths_mean", "death_lowerCI", "deaths_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.meanCI <- function(matrix_by_row, nboots = 1000) {
    mean_rows <- apply(matrix_by_row, 1, mean)
    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
    lower <- unlist(CIs)[1, ] 
    upper <- unlist(CIs)[2, ]
    summ <- cbind(mean_rows, lower, upper)
    colnames(summ) <- c("mean", "lower", "upper")
    return(summ)
}

boots <- function(x, nboot = 1000){
  n = length(x)
  xbar = mean(x)
  tmpdata = sample(x,n*nboot, replace=TRUE)
  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
  bsmeans = colMeans(bootstrapsample)
  deltastar = bsmeans - xbar
  d = quantile(deltastar, c(0.05, 0.95))
  ## Get from quantiles
  ci = xbar - c(d[2], d[1])

  ## Sorted CIs
  sorteddeltastar = sort(deltastar)
  # # Look at the sorted results
  # Find the .05 and .95 critical values of deltastar
  d9alt = sorteddeltastar[5]
  d1alt = sorteddeltastar[950]
  # Find and print the 80% confidence interval for the mean
  ciAlt = xbar - c(d1alt,d9alt)
  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
  return(ci)
}

```

```{r}
results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)

plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 20)
segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")

results_base <- get.burden(p_rabid = 0.40, rho_max = 0.98)

p_rabid_check <- seq(0.10, 0.80, by = 0.1)
check <- data.frame(p_rabid = 0, district = NA, ttimes = 0, deaths_mean = 0)
for (i in 1:length(p_rabid_check)) {
  print(paste(i, "/", length(p_rabid_check)))
  results <- get.burden(p_rabid = p_rabid_check[i], rho_max = 0.98)
  p_rabid <- rep(p_rabid_check[i], length(results$deaths_mean))
  check_one <- cbind(p_rabid, results$district, results$ttimes, results$deaths_mean)
  colnames(check_one) <- c("p_rabid", "district", "ttimes", "deaths_mean")
  check <- rbind(check, check_one)
}

check <- check[-1, ]
check$base <- rep(results_base$deaths_mean, length(p_rabid_check))
ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
  geom_tile(aes(fill = deaths_mean/base)) +
  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 1, name = "Mean deaths")

rho_max_check <- seq(0.90, 0.99, by = 0.01)
check <- data.frame(rho_max = 0, district = NA, ttimes = 0, deaths_mean = 0)
for (i in 1:length(rho_max_check)) {
  print(paste(i, "/", length(rho_max_check)))
  results <- get.burden(p_rabid = 0.4, rho_max = rho_max_check[i])
  rho_max <- rep(rho_max_check[i], length(results$deaths_mean))
  check_one <- cbind(rho_max, results$district, results$ttimes, results$deaths_mean)
  colnames(check_one) <- c("rho_max", "district", "ttimes", "deaths_mean")
  check <- rbind(check, check_one)
}

check <- check[-1, ]
check$base <- rep(results_base$deaths_mean, length(rho_max_check))
check %>%
  group_by(district) %>%
  mutate(min = min(deaths_mean), max = max(deaths_mean), 
         standardized = (deaths_mean - base)/(max - base)) -> check

ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
  geom_tile(aes(fill = standardized)) +
  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths") 
ggplot(data = check, aes(x = as.factor(p_rabid), y = as.factor(reorder(district, ttimes)))) + 
  geom_tile(aes(fill = deaths_mean)) +
  scale_fill_gradient2(high = "red", mid = "white", low = "blue", midpoint = 0, name = "Mean deaths")

mada_dist_gg <- tidy(mada_district, region = "mdg_dis_co")
mada_dist_gg <- left_join(mada_dist_gg, exps_cleaned, by = c("id" = "district"))

ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = CTAR)) +
  scale_colour_viridis_d()

```

## Commune level mods
```{r commune matching, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
commune_matching <- read.csv("data/commune_check_CH.csv")
mada_communes <- readOGR("output/communes/communes_extracted.shp")
mada_communes$match <- tolower(str_replace_all(mada_communes$commune,"[^[:graph:]]", "/"))

mada_communes@data %>%
  select(match, mtch_ds, mdg_cm_) %>%
  left_join(commune_matching, .) %>%
  select(name, match, district, mdg_cm_) -> comm_matches

ctar_data$name <- tolower(str_replace_all(ctar_data$commune,"[^[:graph:]]", "/"))
left_join(ctar_data, comm_matches) %>%
  distinct(record_id, .keep_all = TRUE) -> ctar_data

## Looking at proportion matched
ctar_data %>%
  group_by(id_ctar) %>%
  summarise(n = n(), unmatched = sum(is.na(mdg_cm_)),
            prop_matched = unmatched/n) -> ctar_data_matched

ctar_metadata <- merge(ctar_metadata, ctar_data_matched)
ctar_metadata <- ctar_metadata[!is.na(ctar_metadata$id_ctar), ]
ggplot(ctar_metadata, aes(CTAR, prop_matched)) +
  geom_point() +
  coord_flip()  +
  ylab(label = "Proportion of patients matched to commune")
```

```{r }
## Ctar data at commune level
ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), commune = mdg_cm_, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")
comm_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> comm_catchments
mada_communes@data <- left_join(mada_communes@data, comm_catchments)

mada_communes@data %>%
select(commune = mdg_cm_, pop = MDG__201, ttimes_weighted, CTAR, exclude) -> covars
mada_communes$ctar_in <- ifelse(mada_communes$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)

## Catchments with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
covars$exclude <- 0
covars$exclude[covars$CTAR %in% no_data] <- 1

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) %>%
  group_by(commune, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
exps_cleaned %>%
  group_by(commune) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_cleaned

library(lme4)

## poisson with total pop as offset
exps_cleaned <- filter(exps_cleaned, ttimes_weighted != "Inf")
mod <- glm(round(bites) ~ ttimes_weighted + offset(log(round(pop))), 
                 data = exps_cleaned, family = poisson)

par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
plot(exps_cleaned$bites, predict(mod, type = "response"), bty = "l", pch = 20,
     xlab = "Observed bites", ylab = "Predicted bites")
abline(a = 0, b = 1, lty = 2, col = "blue")
legend("topright", col = "blue", legend = "1:1 line", lty = 2, bty = "n", seg.len = 1, 
       inset = c(0.0, -0.1))
plot(exps_cleaned$ttimes_weighted/60, predict(mod, type = "response"), bty = "l", pch = 20,
     xlab = "Travel times (hrs)", ylab = "Bites", col = alpha("blue", 0.7), 
     ylim = c(0, max(exps_cleaned$bites)))
points(exps_cleaned$ttimes_weighted/60, exps_cleaned$bites, bty = "l", pch = 20,
     xlab = "Travel times (hrs)", col = alpha("darkred", 0.7))
legend("topright", col = c(alpha("blue", 0.7), alpha("darkred", 0.7)), legend = c("Predicted", "Observed"), 
       pch = 20, bty = "n")

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(commune = mdg_cm_, id_ctar) %>% # group by year and district
  summarize(n = n()) ->  exp_mat # get count of exposures

comm_catchments$exclude <- 0
comm_catchments$exclude[comm_catchments$CTAR %in% no_data] <- 1

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(comm_catchments, mdg_cm_, CTAR, exclude), by = c("commune" = "mdg_cm_")) %>%
  group_by(commune) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop[catch_check$exclude == 0])
hist(catch_check$prop[catch_check$exclude == 1])

## getting var for ctar in
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
over(pts, mada_communes)$mtch_ds
```

### 2019-01-09 archive/check.Rmd
archiving older files

---
title: "Draft of Figures for Access Paper"
author: "Malavika Rajeev"
date: "12/19/2018"
output: 
  html_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
```

## Goals of this analysis
- To estimate burden of rabies in Madagascar using nationally available bite patient data 
- To explore the potential impacts of expanded access to PEP in Madagascar

## Slide with R Output
## Data
- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
  - includes dates and home locations (to the district level) of patient reporting for PEP
- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
  - includes dates and home locations (to the commune level) of patients reporting for PEP
  - also includes rabies exposure status based on a combination of clinic-based triage and contact
  tracing (Rajeev et al. 2018)
- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop

```{r data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
mada_communes <- readOGR("output/communes/communes_extracted.shp")

## catchment data
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Getting ctar district and commune
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
``````{r get data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
mada_communes <- readOGR("output/communes/communes_extracted.shp")

## catchment data
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Getting ctar district and commune
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
```

## Summary of analyses: Data
- Cleaning data
  - excluded contacts with suspect cases
  - estimated clinic level underreporting (i.e. incomplete submission of forms)
  - average number of annual bites corrected by clinic level underreporting at the district level
- Travel times
  - calculated weighted mean of travel times by human population summed to the commune level 
  - For each clinic, we defined the catchment area as all communes for which the clinic 
  was the closest ARMC.

## Figure 1: Catchment network and distribution of patient travel

```{r network fig}
## Network fig
## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments?

## Get color palette from other network fig!
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)

display_palette(ctar_metadata$color)
dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments)

## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines %>% 
  group_by(id_ctar) %>%
  summarise(n = sum(n)) %>%
  left_join(select(ctar_metadata, id_ctar, ctar_dist, LATITUDE, LONGITUDE, color)) -> ctar_points
ctar_points$size <- log(ctar_points$n + 0.1)*0.4
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co")) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, pch = 20, cex = dist_points$size, 
       col = ctar_points$color)
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, y1 = dist_lines$LATITUDE, lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  filter(exclude == 0) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop[catch_check$catch_match == 1], main = "", xlab = "Proportion patients reporting \n within catchment", col = "grey50", border = "white")

```

## Figure 2: Adjustment for reporting
```{r clinic level reporting}
######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

# ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
#   geom_tile(aes(fill = n)) +
#   scale_fill_continuous(type = "viridis", name = "Patients")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```

```{r district level exposures}
## districts that are non-reporter and data not available yet CTAR
mada_district@data %>%
select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$excl <- 0
ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) -> exps_cleaned

exps_cleaned %>%
  group_by(district, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(district) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_dist

exps_cleaned %>%
  group_by(ctar, date) %>% ## Need to rename CTAR so clear if it's CTAR in catchment vs. ctar of bites...
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(ctar) %>%
  summarise(bites = mean(bites)) -> exps_ctar

```

```{r ttimes + pop at commune level}
## Need: weighted ttimes, pop, ctar_in, exclude
comm_catchments %>%
  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) -> comm_covars
comm_covars <- filter(comm_covars, ttimes_weighted != "Inf")
comm_covars %>% 
  left_join(select(mada_district@data, mdg_dis_co, district)) -> comm_data
comm_data$exclude <- dist_catchments$exclude[match(comm_data$mdg_dis_co, dist_catchments$mdg_dis_co)]

## Data to include
nrow(comm_data) - sum(comm_data$exclude)
```

## Summary of analyses: Models of bites as a function of travel time, Moramanga District
- Moramanga models to the commune level:
  $$ \mu_{j} = exp(\beta_{t}T_{j} + \beta_{0})\times pop_{j} $$
  $\mu_{j}$ = the mean number of bites in commune $j$ 
- We then estimate the likelihood of observing the bites at the commune level at the Moramanga clinic where bites are a poisson distribution around the mean $\mu_{j}$

## Summary of analyses: Extending this to the district level
- Moramanga models to the commune level:
  $$ \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_{j} + \beta_{0})\times pop_{j} $$
  $\mu_{d}$ is the mean number of bites in district which is the sum of bites at the commune level given **commune level travel times** $d$
- We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$

### 2019-01-09 archive/summary_figs.Rmd
archiving older files

---
title: "Draft of Figures for Access Paper"
author: "Malavika Rajeev"
date: "12/19/2018"
output: 
  html_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
```

## Goals of this analysis
- To estimate burden of rabies in Madagascar using nationally available bite patient data 
- To explore the potential impacts of expanded access to PEP in Madagascar

## Slide with R Output
## Data
- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
  - includes dates and home locations (to the district level) of patient reporting for PEP
- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
  - includes dates and home locations (to the commune level) of patients reporting for PEP
  - also includes rabies exposure status based on a combination of clinic-based triage and contact
  tracing (Rajeev et al. 2018)
- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop

```{r data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
mada_communes <- readOGR("output/communes/communes_extracted.shp")

## catchment data
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Getting ctar district and commune
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)

```

```{r get data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
mada_communes <- readOGR("output/communes/communes_extracted.shp")

## catchment data
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Getting ctar district and commune
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                   proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
```

## Summary of analyses: Data
- Cleaning data
  - excluded contacts with suspect cases
  - estimated clinic level underreporting (i.e. incomplete submission of forms)
  - average number of annual bites corrected by clinic level underreporting at the district level
- Travel times
  - calculated weighted mean of travel times by human population summed to the commune level 
  - For each clinic, we defined the catchment area as all communes for which the clinic 
  was the closest ARMC.

## Figure 1: Catchment network and distribution of patient travel

```{r network fig}
## Network fig
## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments?

## Get color palette from other network fig!
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)

display_palette(ctar_metadata$color)
dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments)

## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines %>% 
  group_by(id_ctar) %>%
  summarise(n = sum(n)) %>%
  left_join(select(ctar_metadata, id_ctar, ctar_dist, LATITUDE, LONGITUDE, color)) -> ctar_points
ctar_points$size <- log(ctar_points$n + 0.1)*0.4
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co")) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, pch = 20, cex = dist_points$size, 
       col = ctar_points$color)
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, y1 = dist_lines$LATITUDE, lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  filter(exclude == 0) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop[catch_check$catch_match == 1], main = "", xlab = "Proportion patients reporting \n within catchment", col = "grey50", border = "white")

```

## Figure 2: Adjustment for reporting
```{r clinic level reporting}
######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

# ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
#   geom_tile(aes(fill = n)) +
#   scale_fill_continuous(type = "viridis", name = "Patients")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```

```{r district level exposures}
## districts that are non-reporter and data not available yet CTAR
mada_district@data %>%
select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$excl <- 0
ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) -> exps_cleaned

exps_cleaned %>%
  group_by(district, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(district) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_dist

exps_cleaned %>%
  group_by(ctar, date) %>% ## Need to rename CTAR so clear if it's CTAR in catchment vs. ctar of bites...
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(ctar) %>%
  summarise(bites = mean(bites)) -> exps_ctar

```

```{r ttimes + pop at commune level}
## Need: weighted ttimes, pop, ctar_in, exclude
comm_catchments %>%
  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) -> comm_covars
comm_covars <- filter(comm_covars, ttimes_weighted != "Inf")
comm_covars %>% 
  left_join(select(mada_district@data, mdg_dis_co, district)) -> comm_data
comm_data$exclude <- dist_catchments$exclude[match(comm_data$mdg_dis_co, dist_catchments$mdg_dis_co)]

## Data to include
nrow(comm_data) - sum(comm_data$exclude)
```

## Summary of analyses: Models of bites as a function of travel time, Moramanga District
- Moramanga models to the commune level:
  $$ mu_{j} = exp(\beta_{t}T_{j} + \beta_{0})\times pop_{j} $$
  $$ mu_{j} $$ = the mean number of bites in commune $j$ 
- We then estimate the likelihood of observing the bites at the commune level at the Moramanga clinic where bites are a poisson distribution around the mean $mu_{j}$

## Summary of analyses: Extending this to the district level
- Moramanga models to the commune level:
  $$ \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_{j} + \beta_{0})\times pop_{j} $$
  $$  \mu_{d} $$ is the mean number of bites in district which is the sum of bites at the commune level given **commune level travel times** $d$
- We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $mu_{d}$

```{r Moramanga data}
## Libraries
library(ggThemeAssist)
library(colourpicker)
library(dplyr)
library(stringr)
library(lubridate)
library(ISOweek)
library(ggplot2)
library(knitr)
library(prettymapr)
# libraries for gis
library(raster)
library(maptools)
library(maps)
library(GISTools)
library(rgdal)
library(sp)
library(rgdal)
library(gdistance)

source("R/utils.R")

## data from WM
triage <- read.csv("data/moramanga/triage_up_todate.csv", header = TRUE)
ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
pep <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "PEP")), header = TRUE)
hh <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Household")), header = TRUE)
human <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Human")), header = TRUE)
animal <- read.csv (paste0("data/WM/", list.files("data/WM/", pattern = "Animal")), header = TRUE)

## Read in new column names
match.colnames <- function (dataframe, lookup){
  colnames(dataframe) <- lookup$new.colnames[match(colnames(dataframe), 
                                                   lookup$og.colnames)]
  dataframe <- dataframe[ , !names(dataframe) %in% c("NULL")]
  return(dataframe)
}

triage <- match.colnames(triage, subset(col_lookup, form == "triage"))
ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
pep <- match.colnames(pep, subset(col_lookup, form == "pep"))
hh <- match.colnames(hh, subset(col_lookup, form == "hh"))
animal <- match.colnames(animal, subset(col_lookup, form == "animal"))
human <- match.colnames(human, subset(col_lookup, form == "human"))

# fix factor names
match.within <- function (vec_return, mismatch, match){
  for (i in 1:length(match)){
    if (is.na(mismatch[i])) next
    mismatched <- paste0 ("\\<", mismatch[i], "\\>")
    vec_return <- gsub(mismatched, match[i], vec_return, ignore.case = TRUE)
  }
  return (vec_return)
}

## Changing suspected to probable
animal$suspect.animal <- match.within(animal$suspect.animal, "Suspected", "Probable")
human$suspect.exposure <- match.within(human$suspect.exposure, "Suspected", "Probable")
triage$suspect.animal <- match.within(triage$suspect.animal, "Suspected", "Probable")

## get number of animals + humans bitten by biter.IDs here!
animals.bitten.byID <- table(animal$biter.ID)
humans.bitten.byID <- table(human$biter.ID)

## Set-up time series plotting (axes)
tfunc <- function (date, format_date="%Y-%m-%d", start="01-09-2016", format_start="%d-%m-%Y",
                   year1=2016, tstep="month", period=FALSE) {
  get.consec (date, format_date, start, format_start,
              year1, tstep, period) - 8
}

## all dates
triage %>% 
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> triage
pep$pep.date <- dmy(pep$pep.date)
ctar %>%
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
animal %>%
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> animal
human %>%
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> human

# triage
triage$commune <- sapply(strsplit(as.character(triage$exposure.location), "\\("), "[", 1)
triage$commune <- trimws(triage$commune, which = "right")
triage$district <- sapply(strsplit(as.character(triage$exposure.location), "\\, "), "[", 2)
triage$district <- gsub(" \\(District\\)", "", triage$district)

# ctar
ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
ctar$commune <- trimws(ctar$commune, which = "right")
ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
ctar$district <- gsub(" \\(District\\)", "", ctar$district)

## animal and human locations
animal <- merge(animal, hh, by = "reference.hh")
human <- merge(human, hh, by = "reference.hh")

## get apprx case dates
ctar$ctar.ID <- paste0(ctar$ctar.ID, "_", year(ctar$date.reported))
human$ctar.ID <- paste0(human$ctar.ID, "_", year(human$date.bitten))
human$reference.ctar <- ctar$reference.ctar[match(human$ctar.ID, ctar$ctar.ID)]
#human$date.bitten[!is.na(human$reference.ctar)] <- ctar$date.bitten[match(human$reference.ctar[!is.na(human$reference.ctar)], ctar$reference.ctar)]

human %>%
  group_by(biter.ID) %>% 
  summarise(sort(date.bitten)[1]) -> date.first.bite
animal$date.case <- date.first.bite$`sort(date.bitten)[1]`[match(animal$ID, date.first.bite$biter.ID)]
animal$date.case[is.na(animal$date.case)] <- animal$date.first.bite[is.na(animal$date.case)]
animal$date.case[is.na(animal$date.case)] <- animal$date.symptoms.started[is.na(animal$date.case)] # fixing animals which don't have a case date (i.e. non-biters, but ones that are still suspect!)
## what about ones without a date?? recheck!

ctar %>%
  group_by(reference.triage) %>% 
  summarise(sort(date.bitten)[1]) -> date.first.bite
triage$date.case <- date.first.bite$`sort(date.bitten)[1]`[match(triage$reference.triage, date.first.bite$reference.triage)]
triage$date.case[is.na(triage$date.case)] <-triage$date.reported[is.na(triage$date.case)]

## get all months 
triage$month <- tfunc(triage$date.reported)
ctar$month <- tfunc(ctar$date.reported)
human$month <- tfunc(human$date.bitten)
animal$month <- tfunc(animal$date.case)
pep$month <- tfunc(pep$pep.date)

## change all true/false to 0/1
triage[triage == FALSE] <- 0
ctar[ctar == FALSE] <- 0
animal[animal == FALSE] <- 0
human[human == FALSE] <- 0
pep[pep == FALSE] <- 0
pep[pep == TRUE] <- 1

## subset to study period (Sep 2016 - to date)
## checks for cases where month is NA!
triage <- subset(triage, month > 0 & month < 25)
ctar <- subset(ctar, month > 0 & month < 25)
human <- subset(human, month > 0 & month < 25)
animal <- subset(animal, month > 0 & month < 25)
pep <- subset(pep, month > 0 & month < 25)

## linking up
ctar$traced <- NA
ctar$traced <- match(ctar$ctar.ID, human$ctar.ID)
ctar$traced[ctar$traced > 0] <- 1
ctar$traced[is.na(ctar$traced)] <- 0
ctar$reference.human <- human$reference.human[match(ctar$ctar.ID, human$ctar.ID)]
triage$reference.human <- ctar$reference.human[match(triage$reference.triage, ctar$reference.triage)] ## here!
triage$suspect.animal[!is.na(triage$reference.human)] <- human$suspect.exposure[match(triage$reference.human[!is.na(triage$reference.human)], human$reference.human)]

ctar$suspect.exposure <- triage$suspect.animal[match(ctar$reference.triage, triage$reference.triage)] ## matched to triage!

nonbites <- c("CTAR_201708_000016", "CTAR_201708_000017", "CTAR_201708_000001", "CTAR_201708_000002",
              "CTAR_201708_000003", "CTAR_201709_000037")
ctar$traced[ctar$reference.ctar %in% nonbites] <- 1
triage <- triage[which(triage$reference.triage %in% ctar$reference.triage), ] ## only triage with corresponding exposures
ctar <- ctar[which(ctar$reference.triage %in% triage$reference.triage), ] ## only ctar with corresponding exposures

## triage data
triage$triaged <- NA
triage$triaged[triage$triage.status == "Successful"] <- 1
triage$triaged[is.na(triage$triaged)] <- 0
triage$traced <- ctar$traced[match(triage$reference.triage, ctar$reference.triage)]
triage$traced[which(triage$reference.triage %in% ctar$reference.triage[ctar$traced == 1])] <- 1
triage$traced[is.na(triage$traced)] <- 0
nonlinked <- c("TR_201709_000014", "TR_201802_001040", "TR_201802_001045", "TR_201802_001063",  "TR_201802_001107", "TR_201802_001219", "TR_201802_001253", "TR_201802_001332", "TR_201802_001361", "TR_201802_001372", "TR_201709_000031", "TR_201708_000042", "TR_201708_000010", "TR_201708_000001", "TR_201709_000020", "TR_201709_000019", "TR_201709_000035", "TR_201709_000023", "TR_201709_000027", "TR_201710_000025", "TR_201711_000008", "TR_201712_000020", "TR_201712_000027", "TR_201708_000002")
triage$traced[triage$reference.triage %in% nonlinked] <- 1
triage$followedup <- triage$traced + triage$triaged
triage$followedup[triage$followedup > 1] <- 1

## cleaning up statuses
ctar$suspect.exposure <- as.factor(ctar$suspect.exposure)
triage$suspect.animal <- as.factor(triage$suspect.animal)
levels(ctar$suspect.exposure) <- c(levels(ctar$suspect.exposure), "Contact with suspect case")
ctar$suspect.exposure[ctar$consult.type == "Contact with suspect case"] <- "Contact with suspect case"
levels(ctar$suspect.exposure)[levels(ctar$suspect.exposure) == ""] <- "Unknown"
levels(triage$suspect.animal)[levels(triage$suspect.animal) == ""] <- "Unknown"
ctar$followedup <- triage$followedup[match(ctar$reference.triage, triage$reference.triage)]

## folks that reported after tracing
ctar$date.traced <- human$date.traced[match(ctar$ctar.ID, human$ctar.ID)]
date.diff <- ctar$date.traced - ctar$date.reported
date.diff[is.na(date.diff)] <-0
after.tracing <- ctar[date.diff < 0,]
ctar <- ctar[date.diff >= 0, ]
nrow(after.tracing)

ctar$triaged <- triage$triaged[match(ctar$reference.triage, triage$reference.triage)]
## Bite data in Moramanga
ctar %>%
  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
                          by = c("commune" = "commune", "district" = "mtch_ds")) %>%
  filter(consult.type == "Animal bite/exposure") %>%
  group_by(mdg_cm_) %>%
  summarize(bites = n()) %>%
  left_join(comm_covars) %>%
  filter(!is.na(ttimes_weighted), district %in% c("Moramanga", "Anosibe-An'ala")) -> bites

```

## Figure 2: Model results Moramanga and District
```{r likelihood of observing moramanga data @ commune level}

get.likelihood.mora <- function(commune, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
}

get.likelihood <- function(commune, district, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}
```

```{r likelihood of observing district data given sum of communes in district}
bites$bites <- round(bites$bites/18*12)
mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = bites, hessian = TRUE)
B_ttimes_mora <- mora_pars$par[1]
B_0_mora <- mora_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(mora_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(mora_pars$hessian)
library(MASS)

ntest = 1000; testpar = mora_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, mora_pars$par, varcov)
}

B_ttimes_mora_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_mora_CI <- quantile(sim.par[,2],c(0.025,0.975))

mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_dist,
             commune = comm_data, hessian = TRUE)
B_ttimes_mada <- mada_pars$par[1]
B_0_mada <- mada_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(mada_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(mada_pars$hessian)
library(MASS)

ntest = 1000; testpar = mada_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, mada_pars$par, varcov)
}

B_ttimes_mada_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
par_df <- data.frame(values = c(B_ttimes_mada, B_ttimes_mora, B_0_mada, B_0_mora),  
                     par = c("B_ttimes", "B_ttimes", "intercept", "intercept"),
                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_0_mada_CI[2], B_0_mora_CI[2]),
                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  CI = B_0_mada_CI[1], B_0_mora_CI[1]),
                     model = c("Mada", "Mora", "Mada", "Mora"))

ggplot(data = par_df, aes(x = model, y = values)) +
  geom_point() +
  geom_linerange(aes(ymin = lower_CI, ymax = upper_CI)) +
  coord_flip()

ggplot(data = filter(par_df, par == "intercept"), aes(x = model, y = values)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI))

ttimes_plot <- seq(0, 15*60, by = 30)
pop_plot <- 1e5

## Mada
plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
       lty = 2, col = "lightblue")
points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
       lty = 2, col = "lightblue")

points(ttimes_plot/60, exp(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "purple")
points(ttimes_plot/60, exp(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
       lty = 2, col = "magenta")
points(ttimes_plot/60, exp(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
       lty = 2, col = "magenta")
legend("topright", c("Mada", "Mora", "CIs"), col = c("darkblue", "purple", "grey"), lty = c(1, 1, 2), 
       bty = "n")

## Data vs. predicted
comm_data %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
                   xlab = "Observed mean bites at district (Mada)", ylab = "Model predicted mean bites")
abline(a = 0, b = 1, col = "blue", lty = 2)

bites %>%
      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
plot(check$bites, check$bites_mu, bty = "n", pch = 20, 
      xlab = "Observed mean bites at commune (Moramanga)", ylab = "Model predicted mean bites")
abline(a = 0, b = 1, col = "blue", lty = 2)

```

```{burden by incidence}

## Mada
plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "n",
     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
       lty = 2, col = "lightblue")
points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
       lty = 2, col = "lightblue")
abline(h = pop_plot/25*0.01*0.39, col = "darkred") # low rabies incidence
abline(h = pop_plot/5*0.01*0.39, col = "red") # high rabies incidence

bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
R_100k_min <- pop_plot/25*0.01*0.39
R_100k_max <- pop_plot/5*0.01*0.39

plot(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, R_100k_min/bites_100k, 0.6))/R_100k_min, 
     type = "l", bty = "n",
     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, R_100k_max/bites_100k, 0.2))/R_100k_max,
     col = "red")

ests_lower <- predict(mod_lower, newdata = as.data.frame(ttimes_plot), 
                                type = "response")
ests_upper <- predict(mod_upper, newdata = as.data.frame(ttimes_plot), 
                                type = "response")

plot(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*ests_upper > R_100k_min, R_100k_min/bites_100k,
                        ests_upper))/R_100k_min, 
     type = "l", bty = "n",
     ylab = "Reporting", xlab = "Travel times (hrs)", col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*ests_lower > R_100k_max, R_100k_max/bites_100k,
                        ests_lower))/R_100k_max,
     col = "red")

```

## Summary of analyses: Models of proportion rabid as a function of travel time, Moramanga District
- We use a generalized linear model with a binomial link function where the response is the number of rabid cases triaged/traced (with the # trials as the total # of triaged/traced patients in the commune and the predictor is the travel times. 
  - the lower end of the number of rabid cases as the number of confirmed rabid and probable status
  - the higher end of the number of rabid cases includes those of unknown exposure status
```{r model of p_rabid}
ctar %>%
  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
                          by = c("commune" = "commune", "district" = "mtch_ds")) %>%
  filter(triaged == 1, consult.type != "Contact with suspect case") %>%
  dplyr::select(mdg_cm_, suspect.exposure) %>%
  gather(suspect.exposure, value, -mdg_cm_) %>%
  group_by(mdg_cm_, value) %>%
  tally %>%
  spread(value, n, fill = 0) %>%
  mutate(total = Confirmed + No + Probable + Unknown, 
         rabid_upper = Confirmed + Probable + Unknown,
         rabid_lower = Confirmed + Probable) %>% 
  left_join(comm_covars) %>%
  filter(!is.na(ttimes_weighted), district != "Antananarivo Renivohitra") -> prop_rabid

## Model of proportion rabid
prop_rabid %$% plot(ttimes_weighted, rabid_lower/total)
mod_lower <- glm(cbind(rabid_lower, total - rabid_lower) ~ ttimes_weighted, data = prop_rabid, 
           family = binomial)
prop_rabid %$% plot(rabid_lower/total, predict(mod_lower, type = "response"), 
                    ylim = c(0, 1), xlim = c(0, 1), bty = "n", xlab = "Observed proportion rabid",
                    ylab = "Model predicted proportion rabid", cex = log(total + 0.01)/2)
abline(a = 0, b = 1, lty = 2, col = "blue")
prop_rabid %$% plot(ttimes_weighted/60, predict(mod_lower, type = "response"), ylim = c(0, 1), 
                    bty = "n", xlab = "Travel times (hrs)", ylab = "Proportion rabid", 
                    pch = 20, col = "darkblue")
prop_rabid %$% points(ttimes_weighted/60, rabid_lower/total, col = "blue", cex = log(total + 0.01)/2)
legend("topright", inset = c(-0.3, -0.2), legend = c("Model predicted", "Observed"), 
       col = c("darkblue", "blue"), pch = c(20, 1), bty = "n", xpd = TRUE)

mod_upper <- glm(cbind(rabid_upper, total - rabid_upper) ~ ttimes_weighted, data = prop_rabid, 
           family = binomial)
prop_rabid %$% plot(rabid_upper/total, predict(mod_upper, type = "response"), 
                    ylim = c(0, 1), xlim = c(0, 1), bty = "n", xlab = "Observed proportion rabid",
                    ylab = "Model predicted proportion rabid", cex = log(total + 0.01/2))
abline(a = 0, b = 1, lty = 2, col = "blue")
prop_rabid %$% plot(ttimes_weighted/60, predict(mod_upper, type = "response"), ylim = c(0, 1), 
                    bty = "n", xlab = "Travel times (hrs)", ylab = "Proportion rabid", 
                    pch = 20, col = "darkblue")
prop_rabid %$% points(ttimes_weighted/60, rabid_upper/total, col = "blue", cex = log(total + 0.01)/2)

ttimes_weighted <- ttimes_plot
prop_rabid %$% plot(ttimes_plot/60, predict(mod_lower, newdata = as.data.frame(ttimes_plot), 
                                            type = "response"), 
                    ylim = c(0, 1), 
                    bty = "n", xlab = "Travel times (hrs)", ylab = "Proportion rabid", 
                    type = "l", col = "darkblue")
prop_rabid %$% points(ttimes_weighted/60, rabid_lower/total, col = "blue", cex = log(total + 0.01)/2)

prop_rabid %$% points(ttimes_plot/60, predict(mod_upper, newdata = as.data.frame(ttimes_plot),
                                              type = "response"), type = "l", col = "red")
prop_rabid %$% points(ttimes_weighted/60, rabid_upper/total, col = "red", cex = log(total + 0.01)/2)
legend("topright", inset = c(-0.3, -0.3), legend = c("Upper observed", "Lower observed"), 
       col = c("Red", "Blue"), pch = 1, bty = "n", xpd = TRUE)
legend("topright", inset = c(-0.25, -0.15), legend = c("Upper model", "Lower model"), 
       col = c("Red", "Blue"), lty = 1, bty = "n", xpd = TRUE)
```

## Figure 3: Proportion rabid models

## Estimating burden
- We used our model to predict average annual bite incidence to the commune level given our model:
- We calculated deaths as:
$$ 
      deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}
$$
- where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
            \frac{R_i}{B_i}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_{i} \times \rho_{max}$. $p_{rabid}$ is taken from Rajeev et al. 2018 on contact tracing data from the Moramanga District.

## Estimating burden
$R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

<br>

$$r \times D \times p_{exp}$$

<br>

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018).

- We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.

## Figure 4: Burden estimates

```{r functions for getting burden}
get.burden <- function(comm = comm_data, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, 
                       pop = comm_data$pop,
                       p_rabid = 0.5, rho_max = 0.9,
                       max_HDR = 25, min_HDR = 5, 
                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                       prob_death = 0.16, nsims = 1000) {
  store_deaths <- store_p_rabid <- matrix(NA, nrow(comm), nsims)

  for (i in 1:nsims) {
    mean_bites <- exp(B_ttimes*comm$ttimes_weighted + B_0)*comm$pop
    bites <- rpois(length(mean_bites), mean_bites)
    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
    rabid_exps <- rpois(length(mean_bites), mean_exps)
    p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid)
    deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), prob_death)
    store_deaths[, i] <- deaths
    store_p_rabid[, i] <- p_rabid_t
  }

  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.burden.flat <- function(covs = covars, pop = covars$pop, model = mod,
                       p_rabid = 0.5, rho = 0.85, 
                       prob_death = 0.16, nsims = 1000) {
  store_deaths <- matrix(NA, nrow(covs), nsims)

  for (i in 1:nsims) {
    mean_bites <- predict(model, type = "response", newdata = covs)
    bites <- rpois(length(mean_bites), mean_bites)
    unreported <- bites*p_rabid/rho - bites*p_rabid
    deaths <- rbinom(length(mean_bites), round(unreported), prob_death)
    store_deaths[, i] <- deaths
  }
  results <- get.meanCI(store_deaths)
  colnames(results) <- c("deaths_mean", "death_lowerCI", "deaths_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.meanCI <- function(matrix_by_row, nboots = 1000) {
    mean_rows <- apply(matrix_by_row, 1, mean)
    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
    lower <- unlist(CIs)[1, ] 
    upper <- unlist(CIs)[2, ]
    summ <- cbind(mean_rows, lower, upper)
    colnames(summ) <- c("mean", "lower", "upper")
    return(summ)
}

boots <- function(x, nboot = 1000){
  n = length(x)
  xbar = mean(x)
  tmpdata = sample(x,n*nboot, replace=TRUE)
  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
  bsmeans = colMeans(bootstrapsample)
  deltastar = bsmeans - xbar
  d = quantile(deltastar, c(0.05, 0.95), na.rm = TRUE)
  ## Get from quantiles
  ci = xbar - c(d[2], d[1])

  ## Sorted CIs
  sorteddeltastar = sort(deltastar)
  # # Look at the sorted results
  # Find the .05 and .95 critical values of deltastar
  d9alt = sorteddeltastar[5]
  d1alt = sorteddeltastar[950]
  # Find and print the 80% confidence interval for the mean
  ciAlt = xbar - c(d1alt,d9alt)
  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
  return(ci)
}

```

```{r}
results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)

plot(results_pr20$ttimes_weighted/60, results_pr20$deaths_mean, col = "darkred", pch = 20)
points(results_pr60$ttimes_weighted/60, results_pr60$deaths_mean, col = "red", pch = 20)
segments(x0 = results_pr60$ttimes_weighted/60, x1 = results_pr60$ttimes_weighted/60,
         y0 = results_pr60$deaths_mean, y1 = results_pr20$deaths_mean, col = "red")
```

Generally, estimated reporting decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 5A). Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under `r max(ttimes_plot[bites_100k*0.2 > R_100k_max])/60` given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under `r max(ttimes_plot[bites_100k*0.6 > R_100k_min])/60` given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).

```{r functions for getting burden}

boots <- function(x, nboot = 1000){
  n = length(x)
  xbar = mean(x)
  tmpdata = sample(x,n*nboot, replace=TRUE)
  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
  bsmeans = colMeans(bootstrapsample)
  deltastar = bsmeans - xbar
  d = quantile(deltastar, c(0.05, 0.95), na.rm = TRUE)
  ## Get from quantiles
  ci = xbar - c(d[2], d[1])

  ## Sorted CIs
  sorteddeltastar = sort(deltastar)
  # # Look at the sorted results
  # Find the .05 and .95 critical values of deltastar
  d9alt = sorteddeltastar[5]
  d1alt = sorteddeltastar[950]
  # Find and print the 80% confidence interval for the mean
  ciAlt = xbar - c(d1alt,d9alt)
  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
  return(ci)
}

get.meanCI <- function(matrix_by_row, nboots = 1000) {
    mean_rows <- apply(matrix_by_row, 1, mean)
    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
    lower <- unlist(CIs)[1, ] 
    upper <- unlist(CIs)[2, ]
    summ <- cbind(mean_rows, lower, upper)
    colnames(summ) <- c("mean", "lower", "upper")
    return(summ)
}

get.burden <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
                       param_ttimes = B_ttimes_mada,
                       param_intercept = B_0_mada,
                       p_rabid = 0.5, rho_max = 0.9,
                       max_HDR = 25, min_HDR = 5, 
                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                       prob_death = 0.16, nsims = 1000) {

  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)

  for (i in 1:nsims) {
    ## for testing
    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
    #                    param_ttimes = B_ttimes_mada;
    #                    param_intercept = B_0_mada;
    #                    p_rabid = 0.5; rho_max = 0.9;
    #                    max_HDR = 25; min_HDR = 5;
    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
    #                    prob_death = 0.16;

    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
    rabid_exps <- rpois(length(mean_exps), mean_exps)
    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
    reported_rabid <- round(bites*p_rabid_t)
    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
    store_deaths[, i] <- deaths
    store_p_rabid[, i] <- p_rabid_t
  }

  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.burden.flat <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
                       param_ttimes = B_ttimes_mada,
                       param_intercept = B_0_mada,
                       p_rabid = 0.5, rho = 0.9,
                       max_HDR = 25, min_HDR = 5, 
                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                       prob_death = 0.16, nsims = 1000) {

  store_deaths <- matrix(NA, length(ttimes), nsims)

  for (i in 1:nsims) {
    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
    reported_rabid <- rbinom(length(ttimes), size = bites, prob = 0.5)
    total_rabid <- round(reported_rabid/rho)
    deaths <- rbinom(length(mean_bites), total_rabid - reported_rabid, prob_death)
    store_deaths[, i] <- deaths
  }

  results <- get.meanCI(store_deaths)
  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI")
  results <- cbind(covars, results)
  return(results)
}

```

```{r}
results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
results_pr20$pr <- "0.20"
results_pr60$pr <- "0.60"
results <- rbind(results_pr20, results_pr60)

results %>% 
  group_by(mdg_dis_co, pr) %>% 
  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI), lowerCI = sum(deaths_lowerCI)) %>%
  left_join(dist_catchments) %>% 
  left_join(select(mada_district@data, mdg_dis_co, district)) %>%
  ggplot(., aes(y = reorder(district, ttimes_weighted), x = tot, color = pr)) +
  ggstance::geom_pointrangeh(aes(xmin = lowerCI, xmax = upperCI), fatten = 2) +
  ylab("") +
  xlab("Average annual deaths") +
  scale_color_manual(values = c("0.20" = alpha("red", 0.2), "0.60" = alpha("red", 0.6)), 
                     name = "Proportion rabid") -> p
p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))

```

When we estimate burden of deaths stochastically within this range of incidence and given our high and low estimates of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure 6, results presented aggregated at the district level). Overall, we estimate average annual deaths between `r round(sum(results_pr60$deaths_mean))` - `r round(sum(results_pr20$deaths_mean))`. **Also estimate deaths averted here!**.

### Burden sensitivity analyses
- Compare to assuming flat reporting + bite model
- $\rho_{max}$, change by urban vs. rural commune?
- If rabies incidence, $R$, scales with population size within our estimated range
- Impacts on # of deaths vs. change in scenarios (i.e. focusing on relative decreases in incidence of deaths based on expanded access rather than raw #s)

<br>

## Discussion

### Key findings

### Strengths and Limitations
- Not accounting for clinic functioning
- Not explicit data on reporting or rabies incidence!
- Other factors that drive reporting
- While we use a signficiant amount of assumptions, these are the same set of assumptions that underlie current rabies estimates, we're just improving them a bit... and ours are a likely a little bit more in line with reality @ a sub-national level?

### Broader context

### Conclusions

### 2019-01-29 accessPEP.Rmd
added in bits on pop sensitivity analyses etc.
will also add in jags function

@@
 ---
-
-## INTRODUCTION
-
-Access to healthcare is a significant driver of disease burden globally, and heterogeneities in access to care due to socioeconomic and geographic factors likely shape where and who are the most impacted by a disease, particularly at sub-national levels.
-However, these factors are rarely incorporated into estimates of burden due to limited data.
-
-Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
-However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
-Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
-The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
-the morebites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
-While at a national level, these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
-The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
-Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
-
-In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
-PEP is not available at any other public clinics or through the private sector.
-Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
-To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
-Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
-Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
-
-## METHODS
-
-### Bite patient data
-We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
-These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
-Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
-These data include the administrative district of the bite patient's address and the date of reporting.
-As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted, excluding any periods of time for which there were no forms submitted for 15 consecutive days.
-Clinic level reporting was calculated the proportion of days for which were included based on the 15 day threshold per year.
-As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
-2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
-Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this method using 27 months of bite patient data from the Moramanga ARMC for which we had details on the type of exposure.
-
-<br>
-
-### GIS Data
-We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
-2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
-We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
-For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
-Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
-2012) and also aggreagated to the commune level
-<br>
-
-### Model of reported bites as a function of travel time
-While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX, *need to think about how to show this*).
-In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
-Incidence at the commune level is then a function of travel times to the closest ARMC.
-Specifically, we modeled bites as follows:
-$$ 
-  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
-$$
-where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
-We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
-
-As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
-$$ 
-\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
-$$
-where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
-We only included communes which were designated to be within the catchment for the clinic.
-We compared our estimates of $\beta_t$ (i.e.
-the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
-
-### Model sensitivity/validation
-- Out-of-fit check for Tana + Fianarantsoa?
-- Compare to null models 
-- Look at sensitivity to inclusion criteria of days and exclusion of contacts
-
-<br>
-
-### Estimation of burden and reporting 
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
-
-We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) $:
-$$
-  \overline\rho = \frac{\mu \times p_{rabid}}{R}
-$$
-or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
-We look at the range of $p_{rabid}$ reported in Rajeev et al.
-2018 for data from the Moramanga District (0.2 - 0.6).
-where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
-$$
-           p_{rabid}=
-            \begin{cases}
-            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
-            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
-            \end{cases}
-$$
-<br>
-
-such that rabid reported bites (i.e.
-$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
-$R_i \times \rho_{max}$).
-$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
-the area with the minimum travel time).
-.
-
-<br>
-
-The rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
-2018) to generate $R$.
-We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
-
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
-$$
-deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
-$$
-where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
-We assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
-
-*Need to include bits about RIG in here somewhere*
-
-<br>
-
-### Estimating the impact of expanding PEP Access
-We use this framework to compare three scenarios of PEP provisioning in Madagascar:
-
-1. The baseline with the current clinic locations (n = 31)
-2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to all CSB IIs (*something to describe CSB IIs here*), that is primary hospitals (n = X)
-
-We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
-
-### Sensitivity analyses for burden estimates 
-- R (increasing or decreasing with pop) (min + max)
-- P-rabid min and max
-- Probability of death min and max
-- Look at change in raw # of deaths and also change in burden by scenarios
-
-<br>
-
-## RESULTS
 
@@
 ```
+
+## To do
+- Add in comparison to district model (+ demonstrate why estimating burden at district level does not result in realistic estimates of burden...because reporting for Moramanga District will be estimated at 0.98, same as for Antananarivo Renivohitra (i.e.
+Tana)
+- Sensitivity analyses for model: comparison to null models, with variation in reporting estimate
+- Sensitivity analyses for burden: comparison to null models (flat reporting, ttimes at district level) p_rabid (places with high ttimes), rho_max (places with high ttimes), R (scaling with pop); 
+- Calculate rho_max from Moramanga data
+- Calculate p_rabid from Moramanga data (not to include in model but as supplementary)
+
+## INTRODUCTION
+Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
+However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
+Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
+
+Here--discuss geographic variation in access to vaccination and care--how this shapes mortality for other diseases.
+
+The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
+the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
+
+In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
+PEP is not available at any other public clinics or through the private sector.
+In addition, there is limited control in dog populations and the disease is endemic throughout the country.
+Due to the spatially restricted nature of PEP and the endemic circulation of canine rabies, geographic access is likely to be a major driver of disease burden within the country.
+To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
+Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
+Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
+
+## METHODS
+
+### GIS Data
+We used the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
+2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
+We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
+2012) and also aggreagated to the commune level.
+
+<br>
 
@@
 
-### Estimating average bite incidence at the district level
+### Bite patient data
+We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
+These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
+These data include the administrative district of the bite patient's address and the date of reporting.
+
+As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
+2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
+Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
+For the national data for which a subset of patient forms were noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
+We further excluded known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
+
+```{r figure 3B, fig.cap= "Figure 3B"}
+patient_ts %>%
+  group_by(ctar) %>%
+  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
+
+ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
+  geom_jitter(alpha = 0.5, width = 0.25) +
+  coord_flip()
+
+to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
+to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
+ctar_data$exclude_bycomment <- 0
+ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
+ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
+patient_ts %>%
+  ungroup() %>%
+  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
+  right_join(ctar_data) -> ctar_check
+
+```
+
+``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
+## Testing with Moramanga data 
+sdev <- seq(2, 10, by = 1)
+check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
+for (i in 1:length(sdev)){
+  ctar %>%
+    group_by(date.reported) %>%
+    summarize(n = n()) %>%
+    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
+  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
+  check <- rbind(check, c(sdev[i], checksd[1:2]))
+}
+
+plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
+points(check$sd, check$prop_bites, col = "red", pch = 20)
+```
+
+<br>
+
 ```{r network fig}
@@
 
+For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
 For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).
-In addition, at the clinic level, the majority of bites reported to a given ARMC came from within the catchment (Figure 1C).
 
@@
 
+As there were substantial undersubmission of forms from clinics *...* need to describe the reason why we think this is temporal, we excluded any periods of time for which there were no forms submitted for 15 consecutive days.
 We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure 2A).
@@
 In all cases, we further excluded any years for which there was less than 25% reporting.
-
-``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
-## Testing with Moramanga data 
-sdev <- seq(2, 10, by = 1)
-check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
-for (i in 1:length(sdev)){
-  ctar %>%
-    group_by(date.reported) %>%
-    summarize(n = n()) %>%
-    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
-  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
-  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
-  check <- rbind(check, c(sdev[i], checksd[1:2]))
-}
-
-plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
-points(check$sd, check$prop_bites, col = "red", pch = 20)
-```
-
-```{r figure 3B, fig.cap= "Figure 3B"}
-patient_ts %>%
-  group_by(ctar) %>%
-  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
-
-ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
-  geom_jitter(alpha = 0.5, width = 0.25) +
-  coord_flip()
-
-to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
-to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$exclude_bycomment <- 0
-ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
-ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
-patient_ts %>%
-  ungroup() %>%
-  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
-  right_join(ctar_data) -> ctar_check
-
-```
-
-We attempted to exclude these by looking at the distribution of patients reporting per day to a given clini, excluding patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
-For the national data for which a subset of patient forms were noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
-We further excluded known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
 
@@
 
+### Model of reported bites as a function of travel time
+While the national bite patient data is available at the district level, travel times and therefore reporting likely vary significantly at the sub-district level.
+In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
+Incidence at the commune level is then a function of travel times to the closest ARMC.
+Specifically, we modeled bites as follows:
+
+$$ 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
+$$
+where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+
+As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
+
+$$ 
+\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+$$
+
+where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
+We only included communes which were designated to be within the catchment for the clinic.
+We compared our estimates of $\beta_t$ (i.e.
+the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
+
+Finally, we compared these models to a model of bites at the district level as the function of trave times averaged to the district level:
+$$ 
+  \mu_{d} = exp(\beta_{t}T_d + \beta_0)\times pop_d 
+$$
+where $pop_d$, $T_d$, and $mu_d$ are respectively human population size, weighted travel times, and mean number of bites in district $d$.
+As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models, as a test of whether the observed patterns could be predicted by bite incidence scaling with population size.
+We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.
+
+<br>
+
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
+
+We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
+$$
+  \overline\rho = \frac{\mu \times p_{rabid}}{R}
+$$
+or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
+We look at the range of $p_{rabid}$ reported in Rajeev et al.
+2018 for data from the Moramanga District (0.2 - 0.6).
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
+            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
+            \end{cases}
+$$
+<br>
+
+such that rabid reported bites (i.e.
+$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_i \times \rho_{max}$).
+$\rho_{max}$ is taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
+the area with the minimum travel time in the district, on average *xx* minutes).
+
+<br>
+
+To generate $R$, the rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018).
+We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
+
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
+We constrain $p_rabid$ as in Equation N, and we assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
+The probability of death in the absence of PEP is taken from cite GAVI/Joel paper.
+<br>
+
+### Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
+
+1. The baseline with the current clinic locations (n = 31)
+2. Expansion of ARMC to one clinic per district (n = 114)
+3. Expansion of ARMC to two clinics per district.
+
+We use data on the location of clinics provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+### Strategically expanding PEP access
+Given limited resources and functioning of clinics, we developed a framework to look at the incremental benefit of adding ARMC.
+Starting with the current locations, for each of the other N district hospitals, we ranked the clinics by what proportion of the population lived within X hours of a clinic.
+As there are interactions between adding clinics, we add the top ranked ARMC, and then re-run the analyses to look at the incremental benefit of adding the subsquent ARMC, we do this until we have added all 114 clinics.
+
+Similarly for the two clinics per district, within a district, we do the same for CSB IIs within the district, and add the clinic which reduces the proportion of the population living greater than N hours away for the catchment?
+
+### Sensitivity analyses for burden estimates and scenario analyses
+- R
+- Rho_max
+- P_rabid
+
+<br>
+
+## RESULTS
+
 ### Models of bites as a function of travel times
@@
 
+get.likelihood.dist <- function(district, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  district %>%
+      filter(exclude == 0) %>%
+      mutate(exp_bites = exp(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
+  return(-sum(dpois(round(district$exp_bites), lambda = round(district$bites), log = TRUE)))
+}
+
+### Moramanga model ###
 mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
@@
 
+### Madagascar model ###
 mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mada, district = exps_dist,
@@
 B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
-par_df <- data.frame(model = c("Mada", "Mora", "Mada", "Mora"),
-                     par = c("B_ttimes", "B_ttimes", "intercept", "intercept"),
-                     values = c(B_ttimes_mada, B_ttimes_mora, B_0_mada, B_0_mora),
+
+### District parameters
+dist_pars <- optim(par = c(1e-6, 0.1), get.likelihood.dist, district = exps_dist,
+              hessian = TRUE)
+B_ttimes_dist <- dist_pars$par[1]
+B_0_dist <- dist_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(dist_pars$hessian)
+
+ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
+}
+
+B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+par_df <- data.frame(model = c("Mada", "Mora", "District", "Mada", "Mora", "District"),
+                     par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
+                             "intercept"),
+                     values = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist, 
+                                B_0_mada, B_0_mora, B_0_dist),
                      upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
-                                  B_0_mada_CI[2], B_0_mora_CI[2]),
+                                  B_ttimes_dist_CI[2],
+                                  B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                      lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
-                                  CI = B_0_mada_CI[1], B_0_mora_CI[1]))
+                                  B_ttimes_dist_CI[1], 
+                                  B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]))
 kable(par_df, caption = "Table 1: summary of parameter estimates")
@@
        lty = 2, col = "pink")
-legend("topright", c("Mada", "Mora", "CIs"), col = c("darkblue", "red", "grey"), lty = c(1, 1, 2), 
+
+points(ttimes_plot/60, exp(B_ttimes_dist*ttimes_plot + B_0_dist)*pop_plot, type = "l", col = "purple")
+points(ttimes_plot/60, exp(B_ttimes_dist_CI[1]*ttimes_plot + B_0_dist_CI[1])*pop_plot, type = "l", 
+       lty = 2, col = "lavender")
+points(ttimes_plot/60, exp(B_ttimes_dist_CI[2]*ttimes_plot + B_0_dist_CI[2])*pop_plot, type = "l",
+       lty = 2, col = "lavender")
+
+legend("topright", c("Mada", "Mora", "District"), col = c("darkblue", "red", "purple"), 
+       lty = c(1, 1, 1), 
        bty = "n")
@@
 
+exps_dist %>%
+      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
+points(check$bites, check$bites_mu, col = alpha("blue", 0.5), pch = 20)
+
+library(bbmle)
+get.likelihood.bb <- function(commune, district, params, B_ttimes, B_0) {
+  # To test likelihood function
+  commune = comm_data
+  district = exps_dist
+  # params = c(1e-6, 0.1)
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
+m3 <- mle2(get.likelihood.bb, start = list(B_ttimes = 1e-6, B_0 = 0.1))
+
+```
+
+```{r pop sensitivity analsyis }
+## Pop only
+lm(exps_dist$bites ~ exps_dist$pop)
+
+get.likelihood.pop <- function(commune, district, B_pop, B_0) {
+  # To test likelihood function
+  commune = comm_data
+  district = exps_dist
+  # params = c(1e-6, 0.1)
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_pop*(pop/1000) + B_0)*pop) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
+
+### Pop parameters
+m3 <- mle2(get.likelihood.pop, start = list(B_pop = 1e-6, B_0 = 0.1))
+
+pop_pars <- optim(par = c(1e-6, 0.1), get.likelihood.pop, district = exps_dist,
+             commune = comm_data,
+              hessian = TRUE)
+B_pop_pop <- pop_pars$par[1]
+B_0_pop <- pop_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(pop_pars$hessian)
+
+ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
+}
+
+B_pop_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      mutate(bites = exp(B_pop_pop*(pop/1000) + B_0_pop)*pop) %>%
+      group_by(mdg_dis_co) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
+                   col = alpha("darkblue", 0.5),
+                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
+abline(a = 0, b = 1, col = "grey", lty = 2)
+sum_bites %$% cor(dist_bites, sum_bites)
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
+      group_by(mdg_dis_co) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% points(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("magenta", 0.5))
+abline(a = 0, b = 1, col = "grey", lty = 2)
+sum_bites %$% cor(dist_bites, sum_bites)
+
+get.likelihood.mora <- function(commune, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  commune %>%
+      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
+  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
+}
+
 ```
@@
 We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC.
-Both models produced reasonable fits to the data (Figure 5B,...), however, there was some variation in bite incidence which both models were not able to capture.
+All of the models produced reasonable fits to the data (Figure 5B,...), however, there was some variation in bite incidence which both models were not able to capture.
 
-### Model sensitivity analyses
 - compare to null models: flat incidence, incidence solely predicted by pop;
@@
 
+get.burden.dist <- function(covars = mada_district@data, 
+                            ttimes = mada_district$ttimes_weighted, 
+                            pop = mada_district$pop2015adj, 
+                            param_ttimes = B_ttimes_dist,
+                            param_intercept = B_0_dist,
+                            p_rabid = 0.5, rho_max = 0.9,
+                            max_HDR = 25, min_HDR = 5, 
+                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                            prob_death = 0.16, nsims = 1000) {
+
+  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)
+
+  for (i in 1:nsims) {
+    ## for testing
+    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
+    #                    param_ttimes = B_ttimes_mada;
+    #                    param_intercept = B_0_mada;
+    #                    p_rabid = 0.5; rho_max = 0.9;
+    #                    max_HDR = 25; min_HDR = 5;
+    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
+    #                    prob_death = 0.16;
+
+    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
+    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
+                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
+    rabid_exps <- rpois(length(mean_exps), mean_exps)
+    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
+                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
+    reported_rabid <- round(bites*p_rabid_t)
+    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
+    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
+    store_deaths[, i] <- deaths
+    store_p_rabid[, i] <- p_rabid_t
+  }
+
+  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
+  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
+                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
+  results <- cbind(covars, results)
+  return(results)
+}
+
 get.burden.flat <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
@@
 ```{r burden results, fig.cap = "Figure 7", cache = TRUE}
-results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
-results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
+results_pr20 <- get.burden.dist(p_rabid = 0.20, rho_max = 0.98)
+results_pr60 <- get.burden.dist(p_rabid = 0.60,rho_max = 0.98)
 results_pr20$pr <- "0.20"
@@
   group_by(mdg_dis_co, pr) %>% 
-  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI), lowerCI = sum(deaths_lowerCI)) %>%
+  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI),
+            lowerCI = sum(deaths_lowerCI)) %>%
   left_join(dist_catchments) %>% 
-  left_join(select(mada_district@data, mdg_dis_co, district)) %>%
-  ggplot(., aes(y = reorder(district, ttimes_weighted), x = tot, color = pr)) +
+  left_join(select(mada_district@data, mdg_dis_co, district)) -> results_grouped
+
+  ggplot(results_grouped, aes(y = reorder(district, ttimes_weighted), x = tot, 
+                              color = pr)) +
   ggstance::geom_pointrangeh(aes(xmin = lowerCI, xmax = upperCI), fatten = 2) +
@@
 
+ggplot(results_grouped, aes(y = tot, x = ttimes_weighted, color = pr)) +
+  geom_point()
+ggplot(results, aes(y = deaths_mean, x = ttimes_weighted, color = pr)) +
+  geom_point()
+results_grouped %>%
+  filter(pr == "0.60") %$%
+  plot(tot, results_pr60$deaths_mean, col = "pink")
+results_grouped %>%
+  filter(pr == "0.20") %$%
+  points(tot, results_pr60$deaths_mean, col = "red")
+abline(a = 0, b = 1, col = "grey", lty = 2)
+## color by ttimes!
+
+mada_district %>%
+  tidy(., region = "mdg_dis_co") %>%
+  left_join(results_pr60, by = c("id" = "district")) -> ggmada
+
+ggplot(gmmada, aes(long, lat, group = group, fill = deaths_mean)) +
+  geom_polygon()
 ```

### 2019-01-29 accessPEP.Rmd
edited scenario/incremental text

@@
 2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to two clinics per district.
+3. Expansion of ARMC to two clinics per district, choosing the clinic which minimizes the proportion of people living greater than 2 hours away from a clinic.
 
@@
 Given limited resources and functioning of clinics, we developed a framework to look at the incremental benefit of adding ARMC.
-Starting with the current locations, for each of the other N district hospitals, we ranked the clinics by what proportion of the population lived within X hours of a clinic.
-As there are interactions between adding clinics, we add the top ranked ARMC, and then re-run the analyses to look at the incremental benefit of adding the subsquent ARMC, we do this until we have added all 114 clinics.
-
-Similarly for the two clinics per district, within a district, we do the same for CSB IIs within the district, and add the clinic which reduces the proportion of the population living greater than N hours away for the catchment?
+Starting with the current locations (scenario 1), we added one clinic and recalculated the proportion of people living < x ttime hours away from a clinic for the country.
+We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding them sequentially.
+We calculate burden at each step and look at the incremental reduction in burden as each clinic is added.
 

### 2019-01-29 accessPEP.Rmd
reorganizing for meeting Jan 30th

--- 2019-01-09 accessPEP.Rmd
@@
 
-## INTRODUCTION
-
-Access to healthcare is a significant driver of disease burden globally, and heterogeneities in access to care due to socioeconomic and geographic factors likely shape where and who are the most impacted by a disease, particularly at sub-national levels.
-However, these factors are rarely incorporated into estimates of burden due to limited data.
-
-Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
-However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
-Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
-The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
-the morebites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
-While at a national level, these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
-The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
-Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
-
-In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
-PEP is not available at any other public clinics or through the private sector.
-Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
-To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
-Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
-Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
-
-## METHODS
-
-### Bite patient data
-We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
-These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
-Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
-These data include the administrative district of the bite patient's address and the date of reporting.
-As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted, excluding any periods of time for which there were no forms submitted for 15 consecutive days.
-Clinic level reporting was calculated the proportion of days for which were included based on the 15 day threshold per year.
-As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
-2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
-Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this method using 27 months of bite patient data from the Moramanga ARMC for which we had details on the type of exposure.
-
-<br>
-
-### GIS Data
-We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
-2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
-We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
-For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
-Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
-2012) and also aggreagated to the commune level
-<br>
-
-### Model of reported bites as a function of travel time
-While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX, *need to think about how to show this*).
-In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
-Incidence at the commune level is then a function of travel times to the closest ARMC.
-Specifically, we modeled bites as follows:
-$$ 
-  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
-$$
-where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
-We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
-
-As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
-$$ 
-\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
-$$
-where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
-We only included communes which were designated to be within the catchment for the clinic.
-We compared our estimates of $\beta_t$ (i.e.
-the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
-
-### Model sensitivity/validation
-- Out-of-fit check for Tana + Fianarantsoa?
-- Compare to null models 
-- Look at sensitivity to inclusion criteria of days and exclusion of contacts
-
-<br>
-
-### Estimation of burden and reporting 
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
-
-We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) $:
-$$
-  \overline\rho = \frac{\mu \times p_{rabid}}{R}
-$$
-or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
-We look at the range of $p_{rabid}$ reported in Rajeev et al.
-2018 for data from the Moramanga District (0.2 - 0.6).
-where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
-$$
-           p_{rabid}=
-            \begin{cases}
-            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
-            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
-            \end{cases}
-$$
-<br>
-
-such that rabid reported bites (i.e.
-$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
-$R_i \times \rho_{max}$).
-$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
-the area with the minimum travel time).
-.
-
-<br>
-
-The rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
-2018) to generate $R$.
-We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
-
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
-$$
-deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
-$$
-where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
-We assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
-
-*Need to include bits about RIG in here somewhere*
-
-<br>
-
-### Estimating the impact of expanding PEP Access
-We use this framework to compare three scenarios of PEP provisioning in Madagascar:
-
-1. The baseline with the current clinic locations (n = 31)
-2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to all CSB IIs (*something to describe CSB IIs here*), that is primary hospitals (n = X)
-
-We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
-
-### Sensitivity analyses for burden estimates 
-- R (increasing or decreasing with pop) (min + max)
-- P-rabid min and max
-- Probability of death min and max
-- Look at change in raw # of deaths and also change in burden by scenarios
-
-<br>
-
-## RESULTS
-
-```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+```{r set up, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
 rm(list = ls())
@@
 ```
+
+## INTRODUCTION
+
+Access to healthcare is a significant driver of disease burden globally, and heterogeneities in access to care due to socioeconomic and geographic factors likely shape where and who are the most impacted by a disease, particularly at sub-national levels.
+However, these factors are rarely incorporated into estimates of burden due to limited data.
+
+Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
+However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
+Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
+The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
+the morebites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
+While at a national level, these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
+The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
+Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
+
+In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
+PEP is not available at any other public clinics or through the private sector.
+Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
+To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
+Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
+Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
+
+<br>
+
+## METHODS
 
@@
 ```
+
+### GIS Data
+We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al.
+2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
+We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
+For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
+Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
+2012) and also aggreagated to the commune level
+
+<br>
+
+### Bite patient data
+We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
+These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
+Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database.
+These data include the administrative district of the bite patient's address and the date of reporting.
+As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted, excluding any periods of time for which there were no forms submitted for 15 consecutive days.
+Clinic level reporting was calculated the proportion of days for which were included based on the 15 day threshold per year.
+As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
+2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
+Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We validated this method using 27 months of bite patient data from the Moramanga ARMC for which we had details on the type of exposure.
 
@@
 
-We attempted to exclude these by looking at the distribution of patients reporting per day to a given clini, excluding patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic, excluding patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
 We validated this using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
@@
   xlab("") +
-  coord_flip()
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 
@@
 Our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
+
+<br>
+
+### Model of reported bites as a function of travel time
+While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX, *need to think about how to show this*).
+In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
+Incidence at the commune level is then a function of travel times to the closest ARMC.
+Specifically, we modeled bites as follows:
+$$ 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
+$$
+where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+
+As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
+$$ 
+\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+$$
+where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
+We only included communes which were designated to be within the catchment for the clinic.
+We compared our estimates of $\beta_t$ (i.e.
+the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
+
+<br>
+
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
+
+We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) $:
+$$
+  \overline\rho = \frac{\mu \times p_{rabid}}{R}
+$$
+or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
+We look at the range of $p_{rabid}$ reported in Rajeev et al.
+2018 for data from the Moramanga District (0.2 - 0.6).
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
+            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
+            \end{cases}
+$$
+<br>
+
+such that rabid reported bites (i.e.
+$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_i \times \rho_{max}$).
+$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
+the area with the minimum travel time).
+.
+
+<br>
+
+The rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018) to generate $R$.
+We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
+
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
+We assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
+
+*Need to include bits about RIG in here somewhere*
+
+<br>
+
+### Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
+
+1. The baseline with the current clinic locations (n = 31)
+2. Expansion of ARMC to one clinic per district (n = 114)
+3. Expansion of ARMC to all CSB IIs (*something to describe CSB IIs here*), that is primary hospitals (n = X)
+
+We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+### Sensitivity analyses for burden estimates 
+- R (increasing or decreasing with pop) (min + max)
+- P-rabid min and max
+- Probability of death min and max
+- Look at change in raw # of deaths and also change in burden by scenarios
+
+<br>
+
+## RESULTS
 

### 2019-01-31 accessPEP.Rmd
updated docs for 1/30 meeting with Jess and Katie

--- 2019-01-29 accessPEP.Rmd
@@
 output:
-  html_document: default
+  word_document:
+    reference_docx: format_docs/word_styles_reference_01.docx
   pdf_document:
@@
       in_header: mystyle.sty
-  word_document:
-    reference_docx: format_docs/word_styles_reference_01.docx
+  html_document: default
 ---
@@
 
-## To do
-- Add in comparison to district model (+ demonstrate why estimating burden at district level does not result in realistic estimates of burden...because reporting for Moramanga District will be estimated at 0.98, same as for Antananarivo Renivohitra (i.e.
-Tana)
-- Sensitivity analyses for model: comparison to null models, with variation in reporting estimate
-- Sensitivity analyses for burden: comparison to null models (flat reporting, ttimes at district level) p_rabid (places with high ttimes), rho_max (places with high ttimes), R (scaling with pop); 
-- Calculate rho_max from Moramanga data
-- Calculate p_rabid from Moramanga data (not to include in model but as supplementary)
-
 ## INTRODUCTION
 Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
-However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
+However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
 Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
@@
 
-The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+The majority of rabies burden studies use these data to estimate burden from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
 the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
-While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites.
+While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.
 The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
@@
 In addition, there is limited control in dog populations and the disease is endemic throughout the country.
-Due to the spatially restricted nature of PEP and the endemic circulation of canine rabies, geographic access is likely to be a major driver of disease burden within the country.
+Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
 To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
@@
 2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale.
-We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
+We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
+Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al.
 2012) and also aggreagated to the commune level.
@@
 ## Centers with no data ## Change this to be so that only ones with zero forms!
-no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
+no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy", "Tsiroanomandidy", 
+             "Taolagnaro", "Mandritsara", 
              "Antsiranana", "Marolambo", "Nosy be", "Sainte Marie", "Vangaindrano")
@@
 ### Bite patient data
-We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017.
+We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
 These were individual patient data forms that were submitted as frequently as monthly to annually by clinics.
+Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years.
+Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis.
+Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A).
 These data include the administrative district of the bite patient's address and the date of reporting.
-
-As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al.
-2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic.
-Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
-For the national data for which a subset of patient forms were noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
-We further excluded known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
-
-```{r figure 3B, fig.cap= "Figure 3B"}
-patient_ts %>%
-  group_by(ctar) %>%
-  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
-
-ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
-  geom_jitter(alpha = 0.5, width = 0.25) +
-  coord_flip()
-
-to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
-to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$exclude_bycomment <- 0
-ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
-ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
-patient_ts %>%
-  ungroup() %>%
-  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
-  right_join(ctar_data) -> ctar_check
-
-```
-
-``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
-## Testing with Moramanga data 
-sdev <- seq(2, 10, by = 1)
-check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
-for (i in 1:length(sdev)){
-  ctar %>%
-    group_by(date.reported) %>%
-    summarize(n = n()) %>%
-    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
-  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
-  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
-  check <- rbind(check, c(sdev[i], checksd[1:2]))
-}
-
-plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
-points(check$sd, check$prop_bites, col = "red", pch = 20)
-```
-
-<br>
+We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below district).
+
+<br>
+
+For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1).
+Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC.
+We excluded any districts from catchments for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).
 
@@
                    color, fill, ctar_dist)) -> dist_lines
-dist_lines$size <- log(dist_lines$n + 0.1)*0.2
+dist_lines$size <- log(dist_lines$n + 0.1)*0.25
 ```
@@
 ```{r figure 1A}
-pdf("figs/network.pdf")
+jpeg("figs/network.jpeg")
 plot(mada_district, col = mada_district$fill, border = "grey") 
@@
        col = dist_points$color)
+legend("topright", c("10", "50", "100", "200", "400", "800", "1600"), pch = 1, xjust = 1, yjust = 1,
+       pt.cex = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, 
+       lwd = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, bty = "n", y.intersp = 0.75,
+       title = "Number of \n reported bites", xpd = TRUE)
+legend("topleft", "A", bty = "n", inset = c(0.2, 0), text.font = 2)
 dev.off()
@@
 
-```{r network output, echo=FALSE, fig.cap="Figure 1C", results = "show", out.width = '100%'}
+```{r network output, echo=FALSE, fig.cap="Figure 1A", results = "show", out.width = '100%'}
 knitr::include_graphics("figs/network.pdf")
@@
 
-For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC.
-For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).
+As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2 shows the time series of form submission for each clinic and periods of time excluded by the 15 day threshold).
+Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold.
+To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
 
@@
 
-```{r figure 2B, fig.cap = "Figure 2B"}
-
+```{r figure 1B, fig.cap = "Figure 1B"}
 clinic_reporting %>%
@@
   ylab("Proportion of days included") +
+  coord_flip() +
+  labs(tag = "B")
+
+```
+
+```{r figure 3B, fig.cap= "Figure 3B"}
+patient_ts %>%
+  group_by(ctar) %>%
+  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
+
+ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
+  scale_color_discrete(name = "Exclude", labels = c("Yes", "No")) +
+  ylab("Number of patients per day") +
+  geom_jitter(alpha = 0.5, width = 0.25) +
   coord_flip()
 
-```
-
-As there were substantial undersubmission of forms from clinics *...* need to describe the reason why we think this is temporal, we excluded any periods of time for which there were no forms submitted for 15 consecutive days.
-We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure 2A).
-This estimate did vary based on our assumption of the threshold number of consecutive days (Figure 2B), so we looked at sensitivity of parameter estimates to changing this threshold (Supplementary Materials?).
-In all cases, we further excluded any years for which there was less than 25% reporting.
+to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
+                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
+to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
+ctar_data$exclude_bycomment <- 0
+ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
+ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
+patient_ts %>%
+  ungroup() %>%
+  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
+  right_join(ctar_data) -> ctar_check
+
+```
+
+``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
+## Testing with Moramanga data 
+sdev <- seq(2, 10, by = 1)
+check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
+for (i in 1:length(sdev)){
+  ctar %>%
+    group_by(date.reported) %>%
+    summarize(n = n()) %>%
+    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
+  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
+  check <- rbind(check, c(sdev[i], checksd[1:2]))
+}
+
+plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "n", ylim = c(0, 1), 
+     ylab = "Proportion excluded", xlab = "Number of standard deviations")
+points(check$sd, check$prop_bites, col = "red", pch = 20)
+legend("topright", c("Contacts", "Bites"), col = c("Blue", "Red"), pch = 20, bty = "n")
+```
+
+Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al.
+2018).
+Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
+We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
+For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
+We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
+We also compared this method to assuming that 20% of total bites at the district level were contacts.
+
+<br>
 
@@
 
-<br>
-
-Our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
+After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
+
+<br>
 
 ### Model of reported bites as a function of travel time
-While the national bite patient data is available at the district level, travel times and therefore reporting likely vary significantly at the sub-district level.
+While the national bite patient data is available at the district level, travel times and therefore reporting, likely vary significantly at the sub-district level.
 In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
@@
 
-Finally, we compared these models to a model of bites at the district level as the function of trave times averaged to the district level:
+Finally, we compared these models to a model of bites at the district level as the function of travel times averaged to the district level:
 $$ 
@@
 
-### Estimating the impact of expanding PEP Access
+### Estimating the impact of expanding PEP Access (TO DO)
 We use this framework to compare three scenarios of PEP provisioning in Madagascar:
@@
 
-### Strategically expanding PEP access
-Given limited resources and functioning of clinics, we developed a framework to look at the incremental benefit of adding ARMC.
-Starting with the current locations (scenario 1), we added one clinic and recalculated the proportion of people living < x ttime hours away from a clinic for the country.
-We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding them sequentially.
-We calculate burden at each step and look at the incremental reduction in burden as each clinic is added.
-
-### Sensitivity analyses for burden estimates and scenario analyses
-- R
-- Rho_max
-- P_rabid
+### Strategically expanding PEP access (TO DO)
+Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of additional ARMC.
+Starting with the current locations (Scenario 1), we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country.
+We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each interation sequentially.
+We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added.
+
+### Sensitivity analyses for burden estimates and scenario analyses (TO DO)
 
@@
 
-par_df <- data.frame(model = c("Mada", "Mora", "District", "Mada", "Mora", "District"),
+par_df <- data.frame(model = c("Mada communes", "Moramanga communes", "Mada districts", 
+                               "Mada communes", "Moramanga communes", "Mada districts"),
                      par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
@@
 
-legend("topright", c("Mada", "Mora", "District"), col = c("darkblue", "red", "purple"), 
+legend("topright", c("Mada communes", "Mada districts", "Moramanga communes"), 
+       col = c("darkblue", "purple", "red"), 
        lty = c(1, 1, 1), 
@@
 sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
-                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
+                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
+                   xlim = c(0, 5000))
 abline(a = 0, b = 1, col = "grey", lty = 2)
+
+exps_dist %>%
+      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
+points(check$bites, check$bites_mu, col = alpha("purple", 0.5), pch = 20)
 
@@
 points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))
-
-exps_dist %>%
-      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
-points(check$bites, check$bites_mu, col = alpha("blue", 0.5), pch = 20)
-
-library(bbmle)
-get.likelihood.bb <- function(commune, district, params, B_ttimes, B_0) {
-  # To test likelihood function
-  commune = comm_data
-  district = exps_dist
-  # params = c(1e-6, 0.1)
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-m3 <- mle2(get.likelihood.bb, start = list(B_ttimes = 1e-6, B_0 = 0.1))
-
-```
-
-```{r pop sensitivity analsyis }
-## Pop only
-lm(exps_dist$bites ~ exps_dist$pop)
-
-get.likelihood.pop <- function(commune, district, B_pop, B_0) {
-  # To test likelihood function
-  commune = comm_data
-  district = exps_dist
-  # params = c(1e-6, 0.1)
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_pop*(pop/1000) + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-
-### Pop parameters
-m3 <- mle2(get.likelihood.pop, start = list(B_pop = 1e-6, B_0 = 0.1))
-
-pop_pars <- optim(par = c(1e-6, 0.1), get.likelihood.pop, district = exps_dist,
-             commune = comm_data,
-              hessian = TRUE)
-B_pop_pop <- pop_pars$par[1]
-B_0_pop <- pop_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(pop_pars$hessian)
-
-ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
-}
-
-B_pop_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-comm_data %>%
-      filter(exclude == 0) %>%
-      mutate(bites = exp(B_pop_pop*(pop/1000) + B_0_pop)*pop) %>%
-      group_by(mdg_dis_co) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
-                   col = alpha("darkblue", 0.5),
-                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
-abline(a = 0, b = 1, col = "grey", lty = 2)
-sum_bites %$% cor(dist_bites, sum_bites)
-
-comm_data %>%
-      filter(exclude == 0) %>%
-      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
-      group_by(mdg_dis_co) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% points(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("magenta", 0.5))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-sum_bites %$% cor(dist_bites, sum_bites)
-
-get.likelihood.mora <- function(commune, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
-  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
-}
+legend("topright", inset = c(-0.1, -0.2),
+       c("Mada communes", "Mada districts", "Moramanga communes"), 
+       col = c(alpha("darkblue", 0.5), alpha("purple", 0.5), alpha("red", 0.5)), 
+       pch = 20, 
+       bty = "n", xpd = TRUE)
 
@@
 We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC.
-All of the models produced reasonable fits to the data (Figure 5B,...), however, there was some variation in bite incidence which both models were not able to capture.
-
-- compare to null models: flat incidence, incidence solely predicted by pop;
-- sensitivity of param estimates to reporting cut-offs (i.e. days excluded)
-- sensitivity of param estimates to exclusion of contacts
+All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the model.
+
+#### Model validation: TO DO
 
@@
         border = NA, col = alpha("red", 0.6))
-legend("topright", inset = c(-0.20, 0), c("Max rabies exposures", "Min rabies exposures"), lty = c(2, 1), 
+legend("topright", inset = c(-0.10, 0), c("Max rabies exposures", "Min rabies exposures"), lty = c(2, 1), 
        col = c("red", "darkred"),
@@
 
-```{r functions for getting burden}
-
-boots <- function(x, nboot = 1000){
-  n = length(x)
-  xbar = mean(x)
-  tmpdata = sample(x,n*nboot, replace=TRUE)
-  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
-  bsmeans = colMeans(bootstrapsample)
-  deltastar = bsmeans - xbar
-  d = quantile(deltastar, c(0.05, 0.95), na.rm = TRUE)
-  ## Get from quantiles
-  ci = xbar - c(d[2], d[1])
-
-  ## Sorted CIs
-  sorteddeltastar = sort(deltastar)
-  # # Look at the sorted results
-  # Find the .05 and .95 critical values of deltastar
-  d9alt = sorteddeltastar[5]
-  d1alt = sorteddeltastar[950]
-  # Find and print the 80% confidence interval for the mean
-  ciAlt = xbar - c(d1alt,d9alt)
-  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
-  return(ci)
-}
-
-get.meanCI <- function(matrix_by_row, nboots = 1000) {
-    mean_rows <- apply(matrix_by_row, 1, mean)
-    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
-    lower <- unlist(CIs)[1, ] 
-    upper <- unlist(CIs)[2, ]
-    summ <- cbind(mean_rows, lower, upper)
-    colnames(summ) <- c("mean", "lower", "upper")
-    return(summ)
-}
-
-get.burden <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
-                       param_ttimes = B_ttimes_mada,
-                       param_intercept = B_0_mada,
-                       p_rabid = 0.5, rho_max = 0.9,
-                       max_HDR = 25, min_HDR = 5, 
-                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                       prob_death = 0.16, nsims = 1000) {
-
-  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)
-
-  for (i in 1:nsims) {
-    ## for testing
-    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
-    #                    param_ttimes = B_ttimes_mada;
-    #                    param_intercept = B_0_mada;
-    #                    p_rabid = 0.5; rho_max = 0.9;
-    #                    max_HDR = 25; min_HDR = 5;
-    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
-    #                    prob_death = 0.16;
-
-    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
-    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
-                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
-    rabid_exps <- rpois(length(mean_exps), mean_exps)
-    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
-                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
-    reported_rabid <- round(bites*p_rabid_t)
-    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
-    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
-    store_deaths[, i] <- deaths
-    store_p_rabid[, i] <- p_rabid_t
-  }
-
-  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
-  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
-                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
-  results <- cbind(covars, results)
-  return(results)
-}
-
-get.burden.dist <- function(covars = mada_district@data, 
-                            ttimes = mada_district$ttimes_weighted, 
-                            pop = mada_district$pop2015adj, 
-                            param_ttimes = B_ttimes_dist,
-                            param_intercept = B_0_dist,
-                            p_rabid = 0.5, rho_max = 0.9,
-                            max_HDR = 25, min_HDR = 5, 
-                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                            prob_death = 0.16, nsims = 1000) {
-
-  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)
-
-  for (i in 1:nsims) {
-    ## for testing
-    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
-    #                    param_ttimes = B_ttimes_mada;
-    #                    param_intercept = B_0_mada;
-    #                    p_rabid = 0.5; rho_max = 0.9;
-    #                    max_HDR = 25; min_HDR = 5;
-    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
-    #                    prob_death = 0.16;
-
-    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
-    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
-                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
-    rabid_exps <- rpois(length(mean_exps), mean_exps)
-    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
-                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
-    reported_rabid <- round(bites*p_rabid_t)
-    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
-    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
-    store_deaths[, i] <- deaths
-    store_p_rabid[, i] <- p_rabid_t
-  }
-
-  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
-  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
-                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
-  results <- cbind(covars, results)
-  return(results)
-}
-
-get.burden.flat <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
-                       param_ttimes = B_ttimes_mada,
-                       param_intercept = B_0_mada,
-                       p_rabid = 0.5, rho = 0.9,
-                       max_HDR = 25, min_HDR = 5, 
-                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                       prob_death = 0.16, nsims = 1000) {
-
-  store_deaths <- matrix(NA, length(ttimes), nsims)
-
-  for (i in 1:nsims) {
-    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
-    reported_rabid <- rbinom(length(ttimes), size = bites, prob = 0.5)
-    total_rabid <- round(reported_rabid/rho)
-    deaths <- rbinom(length(mean_bites), total_rabid - reported_rabid, prob_death)
-    store_deaths[, i] <- deaths
-  }
-
-  results <- get.meanCI(store_deaths)
-  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI")
-  results <- cbind(covars, results)
-  return(results)
-}
-```
-
 ```{r burden results, fig.cap = "Figure 7", cache = TRUE}
-results_pr20 <- get.burden.dist(p_rabid = 0.20, rho_max = 0.98)
-results_pr60 <- get.burden.dist(p_rabid = 0.60,rho_max = 0.98)
+## Commune level
+source("R/get.burden.R")
+results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
+results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
 results_pr20$pr <- "0.20"
@@
 
-  ggplot(results_grouped, aes(y = reorder(district, ttimes_weighted), x = tot, 
+## District level
+results_pr20_dist <- get.burden.dist(p_rabid = 0.20, rho_max = 0.98)
+results_pr60_dist <- get.burden.dist(p_rabid = 0.60,rho_max = 0.98)
+results_pr20_dist$pr <- "0.20"
+results_pr60_dist$pr <- "0.60"
+results_dist <- rbind(results_pr20_dist, results_pr60_dist)
+
+results_dist %>% 
+  left_join(results_grouped, by = c("mdg_dis_co" = "mdg_dis_co", "pr" = "pr")) -> results_comp
+
+ggplot(results_comp, aes (x = deaths_mean, y = tot, color = ttimes_weighted.x)) +
+  geom_point() +
+  geom_abline(slope = 1, intercept = 0, color = "grey", linetype = "dashed") +
+  facet_wrap (~pr)
+
+results_dist %>% 
+  group_by(mdg_dis_co, pr) %>% 
+  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI),
+            lowerCI = sum(deaths_lowerCI)) %>%
+  left_join(dist_catchments) %>% 
+  left_join(select(mada_district@data, mdg_dis_co, district)) -> results_grouped_dist
+results_grouped_dist$scale <- "district"
+results_grouped$scale <- "commune"
+
+results_all <- rbind(results_grouped, results_grouped_dist)
+
+ggplot(results_all, aes(y = reorder(district, ttimes_weighted), x = tot, 
                               color = pr)) +
@@
   scale_color_manual(values = c("0.20" = alpha("red", 0.2), "0.60" = alpha("red", 0.6)), 
-                     name = "Proportion rabid") -> p
+                     name = "Proportion rabid") +
+  facet_wrap(~scale) -> p
 p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
 
-ggplot(results_grouped, aes(y = tot, x = ttimes_weighted, color = pr)) +
-  geom_point()
-ggplot(results, aes(y = deaths_mean, x = ttimes_weighted, color = pr)) +
-  geom_point()
-results_grouped %>%
-  filter(pr == "0.60") %$%
-  plot(tot, results_pr60$deaths_mean, col = "pink")
-results_grouped %>%
-  filter(pr == "0.20") %$%
-  points(tot, results_pr60$deaths_mean, col = "red")
-abline(a = 0, b = 1, col = "grey", lty = 2)
-## color by ttimes!
-
+## Maps of burden
+## district level
 mada_district %>%
   tidy(., region = "mdg_dis_co") %>%
-  left_join(results_pr60, by = c("id" = "district")) -> ggmada
-
-ggplot(gmmada, aes(long, lat, group = group, fill = deaths_mean)) +
-  geom_polygon()
+  left_join(results_pr60_dist, by = c("id" = "mdg_dis_co")) -> ggmada_dist
+
+ggplot(ggmada_dist, aes(long, lat, group = group, fill = deaths_mean)) +
+  geom_polygon() +
+  scale_fill_continuous(type = "viridis") -> burden_map
+burden_map
+ggsave("burden_map_district.pdf")
+
+## commune level
+mada_communes %>%
+  tidy(., region = "mdg_cm_") %>%
+  left_join(results_pr60, by = c("id" = "mdg_cm_")) -> ggmada_commune
+
+ggplot(ggmada_commune, aes(long, lat, group = group, fill = deaths_mean)) +
+  geom_polygon() +
+  scale_fill_continuous(type = "viridis") -> burden_map
+burden_map
+ggsave("burden_map_commune.pdf")
+
 ```
@@
 
-### Burden sensitivity analyses
-- Compare to assuming flat reporting + bite model?  Burden @ district level with bites @ district level?
-- $\rho_{max}$, change by urban vs. rural commune?
-- If rabies incidence, $R$, scales with population size within our estimated range
-- Impacts on # of deaths vs. change in scenarios (i.e. focusing on relative decreases in incidence of deaths based on expanded access rather than raw #s)
-
-<br>
-
-## Discussion
+When we compare our burden estimates at the district vs.
+the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden
+
+### Sensitivity analyses
+
+<br>
+
+## DISCUSSION
 

### 2019-02-11 accessPEP.Rmd
updating in order to merge branches

@@
 
+TO DO: Instead of reporting calculate as incidence of deaths here!!
 We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
@@
 
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows (TO DO: take neg bin around min and max for RI instead):
 $$
@@
 
-### Estimating the impact of expanding PEP Access (TO DO)
-We use this framework to compare three scenarios of PEP provisioning in Madagascar:
-
-1. The baseline with the current clinic locations (n = 31)
-2. Expansion of ARMC to one clinic per district (n = 114)
-3. Expansion of ARMC to two clinics per district, choosing the clinic which minimizes the proportion of people living greater than 2 hours away from a clinic.
-
-We use data on the location of clinics provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
-We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
-
 ### Strategically expanding PEP access (TO DO)
-Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of additional ARMC.
-Starting with the current locations (Scenario 1), we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country.
-We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each interation sequentially.
-We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added.
+Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC.
+Starting with the current locations, we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country.
+We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each iteration sequentially.
+We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added (taking the min/max approach).
 

### 2019-02-11 accessPEP.Rmd

@@
 ```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
-rm(list = ls())
-
 ## Libraries
@@
 ## INTRODUCTION
+
 Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
@@
 ```
+
+<br>
 
@@
 ### Sensitivity analyses for burden estimates and scenario analyses (TO DO)
+
+<br>
+
+## RESULTS
+
+<br>
+
+### Model of reported bites as a function of travel time
+While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX, *need to think about how to show this*).
+In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
+Incidence at the commune level is then a function of travel times to the closest ARMC.
+Specifically, we modeled bites as follows:
+$$ 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
+$$
+where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+
+As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
+$$ 
+\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+$$
+where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
+We only included communes which were designated to be within the catchment for the clinic.
+We compared our estimates of $\beta_t$ (i.e.
+the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
+
+<br>
+
+### Estimation of burden and reporting 
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
+
+We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) $:
+$$
+  \overline\rho = \frac{\mu \times p_{rabid}}{R}
+$$
+or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
+We look at the range of $p_{rabid}$ reported in Rajeev et al.
+2018 for data from the Moramanga District (0.2 - 0.6).
+where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
+$$
+           p_{rabid}=
+            \begin{cases}
+            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
+            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
+            \end{cases}
+$$
+<br>
+
+such that rabid reported bites (i.e.
+$p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e.
+$R_i \times \rho_{max}$).
+$\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
+the area with the minimum travel time).
+.
+
+<br>
+
+The rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al.
+2018) to generate $R$.
+We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
+As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
+
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
+$$
+deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+$$
+where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
+We assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
+
+*Need to include bits about RIG in here somewhere*
+
+<br>
+
+### Estimating the impact of expanding PEP Access
+We use this framework to compare three scenarios of PEP provisioning in Madagascar:
+
+1. The baseline with the current clinic locations (n = 31)
+2. Expansion of ARMC to one clinic per district (n = 114)
+3. Expansion of ARMC to all CSB IIs (*something to describe CSB IIs here*), that is primary hospitals (n = X)
+
+We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3.
+We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.
+
+### Sensitivity analyses for burden estimates 
+- R (increasing or decreasing with pop) (min + max)
+- P-rabid min and max
+- Probability of death min and max
+- Look at change in raw # of deaths and also change in burden by scenarios
 

### 2019-02-23 accessPEP.Rmd
with comparison of burden across scenario analysis and scale of estimation

--- 2019-02-11 accessPEP.Rmd
@@
       in_header: mystyle.sty
-  html_document: default
+  word_document:
+    reference_docx: format_docs/word_styles_reference_01.docx
 ---
@@
 
+<br>
+
 ### Bite patient data
@@
 Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A).
-These data include the administrative district of the bite patient's address and the date of reporting.
-We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below district).
+These data include details of the administrative district of the bite patient's address and the date of reporting.
+We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).
 
@@
 Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC.
-We excluded any districts from catchments for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).
+We excluded any catchment areas for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).
 
@@
 
-As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2 shows the time series of form submission for each clinic and periods of time excluded by the 15 day threshold).
+As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2).
 Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold.
@@
 Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B).
+We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4).
 For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
@@
 ### Model of reported bites as a function of travel time
-While the national bite patient data is available at the district level, travel times and therefore reporting, likely vary significantly at the sub-district level.
-In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level.
-Incidence at the commune level is then a function of travel times to the closest ARMC.
-Specifically, we modeled bites as follows:
+We modeled bites as a function of travel times as follows:
+$$ 
+  \mu = exp(\beta_{t}T + \beta_0)P 
+$$
+where $mu$ is the expected number of bites given travel times and population for a given location. In this way, we can model the impact of travel time on incidence as 
+$exp(\beta_{t}T + \beta_0)$ gives an estimate of incidence and multiplying by the population size gives the expected number of bites, so that we can fit this model to the number of bites in each location.
+
+We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+
+We fit this to our available data in three ways:
+1) the commune data available at the moramanga district, where we fit to $mu_j$
+2) The district data available nationally
+3) To compare estimates from 1 directly to the national data, the district data but where $mu_d$ is the sum of the travel time effect at the commune level, i.e.:
 
 $$ 
-  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
-$$
-where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$.
-We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
-
-As we had data available on reported bites at the commune level from the Moramanga ARMC (from Rajeev et al. 2018), we modeled observed commune bites in the same framework, but un-aggregated, where:
-
-$$ 
-\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
+  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
 $$
 
-where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$.
-We only included communes which were designated to be within the catchment for the clinic.
-We compared our estimates of $\beta_t$ (i.e.
-the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.
-
-Finally, we compared these models to a model of bites at the district level as the function of travel times averaged to the district level:
-$$ 
-  \mu_{d} = exp(\beta_{t}T_d + \beta_0)\times pop_d 
-$$
-where $pop_d$, $T_d$, and $mu_d$ are respectively human population size, weighted travel times, and mean number of bites in district $d$.
-As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models, as a test of whether the observed patterns could be predicted by bite incidence scaling with population size.
+ As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models (replacing $Tj$ with log-transformed population in $j$ to avoid squared terms), as a test of whether the observed patterns could be predicted by bite incidence scaling with population size alone.
 We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.
@@
 
-### Strategically expanding PEP access (TO DO)
+### Strategically expanding PEP access
 Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC.
 Starting with the current locations, we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country.
-We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each iteration sequentially.
-We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added (taking the min/max approach).
+We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially.
+We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added.
 
-### Sensitivity analyses for burden estimates and scenario analyses (TO DO)
+### Sensitivity analyses for burden estimates 
+- R (increasing or decreasing with pop) (min + max)
+- P-rabid min and max
+- Probability of death min and max
+- Look at change in raw # of deaths and also change in burden by scenarios
 
@@
 
+```{r scale example}
+## Comparing scales
+## Function for reporting
+get.reporting_burden <- function(pop, ttimes, B_ttimes, B_0, hdr = 25, incidence = 0.01, 
+                          exp_rate = 0.39, p_rabid = 0.6, rho_max = 0.98, p_death = 0.16, 
+                          type = "reporting") {
+  bites_100k <- exp(B_ttimes*ttimes + B_0)*pop
+  R_100k <- pop/hdr*incidence*exp_rate
+  if (type == "reporting"){
+    rho <- bites_100k*(ifelse(bites_100k*p_rabid > R_100k, rho_max*R_100k/bites_100k, p_rabid))/R_100k
+    return(rho)
+  } else {
+    unreported <- R_100k - bites_100k*(ifelse(bites_100k*p_rabid > R_100k, 
+                                              rho_max*R_100k/bites_100k, p_rabid))
+    deaths <- 0.16*unreported
+    return(deaths)
+  }
+}
+
+comm_data$reporting_min <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
+                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                                            hdr = 5, p_rabid = 0.2)
+
+comm_data$reporting_max <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
+                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                                            hdr = 25, p_rabid = 0.6)
+comm_data %>%
+  group_by(district) %>%
+  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
+  summarize(reporting_min = sum(reporting_min*prop_pop), 
+            reporting_max = sum(reporting_max*prop_pop)) %>%
+  mutate(reporting_med = (reporting_min + reporting_max)/2,
+         type = "summed from commune") -> summed_rep_dist
+
+mada_district@data %>%
+  mutate(reporting_min = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
+                                              p_rabid = 0.2), 
+         reporting_max = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
+                                              p_rabid = 0.6),
+         reporting_med = (reporting_min + reporting_max)/2,
+         type = "district level") %>%
+  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp
+
+rep_comp <- rbind(rep_comp, summed_rep_dist)
+rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]
+
+ggplot(rep_comp, aes(x = reporting_med, y = reorder(district, ttime), 
+                              color = type)) +
+  ggstance::geom_pointrangeh(aes(xmin = reporting_min, xmax = reporting_max), fatten = 2)
+
+bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
+R_100k_min <- pop_plot/25*0.01*0.39
+R_100k_max <- pop_plot/5*0.01*0.39
+
+plot(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
+     type = "l", bty = "n",
+     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+     col = "darkred")
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
+     col = "red", lty = 2)
+lines(ttimes_plot/60, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
+     col = "red", lty = 2)
+polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
+        border = NA, col = alpha("red", 0.2))
+polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
+        border = NA, col = alpha("red", 0.6))
+
+## Mora
+comm_data %>%
+  filter(district == "Moramanga") -> comm_level
+bites_100k <- exp(B_ttimes_mada*comm_level$ttimes_weighted + B_0_mada)*comm_level$pop
+R_100k_min <- comm_level$pop/25*0.01*0.39
+R_100k_max <- comm_level$pop/5*0.01*0.39
+
+## district val for Moramanga
+## need to merge names
+exps_dist %>%
+  filter(dist_name == "Moramanga") %>%
+  mutate(rep_max = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
+                                              p_rabid = 0.6),
+         rep_min = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
+                                              p_rabid = 0.2)) -> dist_level
+dist_level %$% points(ttimes_weighted_dist/60, rep_max, col = "red", pch = 20)
+dist_level %$% points(ttimes_weighted_dist/60, rep_min, col = "red", pch = 20)
+dist_level %$% abline(v = ttimes_weighted_dist/60, col = "red", lty = 2)
+
+```
+
 Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 6).
@@
 When we compare our burden estimates at the district vs.
-the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden
+the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden.
+
+### Expanding access to PEP
+
+```{r scenario analysis}
+## new ctar
+expanded <- read.csv("output/incremental_ARMC_nofilter.csv")
+dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+
+dist_burden_min <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, p_rabid = 0.6, type = "burden"))
+dist_burden_max <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, p_rabid = 0.2, type = "burden"))
+
+comm_burden_min <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 25, p_rabid = 0.6, type = "burden"))
+comm_burden_max <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 5, p_rabid = 0.2, type = "burden"))
+
+## Decrease in burden over times
+plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
+     ylab = "Proportion of deaths compared to baseline ")
+lines(colSums(dist_burden_min, na.rm = TRUE)/sum(dist_burden_min[, 1], na.rm = TRUE),
+      col = "purple")
+lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
+     type = "l", col = "blue", 
+     bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", ylab = "Number of deaths")
+lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")
+
+```
 
@@
 ### Strengths and Limitations
-- Not accounting for clinic functioning
-- Not explicit data on reporting or rabies incidence!
-- Other factors that drive reporting
-- While we use a signficiant amount of assumptions, these are the same set of assumptions that underlie current rabies estimates, we're just improving them a bit...
-and ours are a likely a little bit more in line with reality @ a sub-national level?
 
@@
 ### Conclusions
+
+## Acknowledgements
+
+## Supplementary Figures
+
+#### S1
+#### S2
+#### S3
+#### S4
+#### S5
+#### S6

### 2019-02-26 accessPEP.Rmd
working version of sensitivity analyses and model validation

@@
 ## Libraries
+library(data.table)
 library(knitr)
@@
 ```{r network output, echo=FALSE, fig.cap="Figure 1A", results = "show", out.width = '100%'}
-knitr::include_graphics("figs/network.pdf")
+knitr::include_graphics("figs/network.jpeg")
 ```
@@
 
-``` {r trying to exclude contacts, fig.cap = "Figure 3A"}
+```{r trying to exclude contacts, fig.cap="Figure 3A"}
 ## Testing with Moramanga data 
@@
 ```{r likelihood of observing moramanga data @ commune level, fig.cap = "Figure 5A"}
+library(boot)
+
+get.likelihood.bites <- function(bites, covar, pop, params) {
+  beta <- params[1]
+  intercept <- params[2]
+  exp_bites <- inv.logit(beta*covar + intercept)*pop
+  return(-sum(dpois(round(bites), lambda = exp_bites, log = TRUE)))
+}
 
@@
   commune %>%
-      mutate(bites_mu = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
-  return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$bites_mu, log = TRUE)))
+      mutate(bites_mu = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
+  return(-sum(dpois(round(sum_bites$), lambda = sum_bites$bites_mu, log = TRUE)))
 }
@@
       group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
+      mutate(bites = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) %>%
       summarize(sum_bites = sum(bites)) %>%
@@
 }
-
 get.likelihood.dist <- function(district, params) {
@@
       filter(exclude == 0) %>%
-      mutate(exp_bites = exp(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
+      mutate(exp_bites = inv.logit(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
   return(-sum(dpois(round(district$exp_bites), lambda = round(district$bites), log = TRUE)))
@@
 ## Mada
-plot(ttimes_plot/60, exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "l",
+par_df <- as.data.frame(list(model = c("Mada communes", "Moramanga communes", "Mada districts"),
+                     B_ttimes = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist),
+                     B_0 = c(B_0_mada, B_0_mora, B_0_dist),
+                     B_0_upper = c(B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
+                     B_0_lower = c(B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]),
+                     B_ttimes_upper = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
+                                  B_ttimes_dist_CI[2]),
+                     B_ttimes_lower = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
+                                  B_ttimes_dist_CI[1])))
+
+preds <- foreach(i = 1:nrow(par_df), .combine = rbind) %do% {
+  bites <- inv.logit(par_df$B_ttimes[i]*ttimes_plot + par_df$B_0[i])*pop_plot
+  upper <- inv.logit(par_df$B_ttimes_upper[i]*ttimes_plot + par_df$B_0_upper[i])*pop_plot
+  lower <- inv.logit(par_df$B_ttimes_lower[i]*ttimes_plot + par_df$B_0_lower[i])*pop_plot
+  return(as.data.frame(list(bites = bites, upper = upper, lower = lower, ttimes = ttimes_plot/60, 
+                            model = par_df$model[i])))
+}
+ggplot(data = preds, aes(x = ttimes, y = bites, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.5) +
+
+plot(ttimes_plot/60, inv.logit(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "l",
      ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
+points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
        lty = 2, col = "lightblue")
-points(ttimes_plot/60, exp(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
+points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
        lty = 2, col = "lightblue")
 
-points(ttimes_plot/60, exp(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "red")
-points(ttimes_plot/60, exp(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
+points(ttimes_plot/60, inv.logit(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "red")
+points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
        lty = 2, col = "pink")
-points(ttimes_plot/60, exp(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
+points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
        lty = 2, col = "pink")
 
-points(ttimes_plot/60, exp(B_ttimes_dist*ttimes_plot + B_0_dist)*pop_plot, type = "l", col = "purple")
-points(ttimes_plot/60, exp(B_ttimes_dist_CI[1]*ttimes_plot + B_0_dist_CI[1])*pop_plot, type = "l", 
+points(ttimes_plot/60, inv.logit(B_ttimes_dist*ttimes_plot + B_0_dist)*pop_plot, type = "l", col = "purple")
+points(ttimes_plot/60, inv(B_ttimes_dist_CI[1]*ttimes_plot + B_0_dist_CI[1])*pop_plot, type = "l", 
        lty = 2, col = "lavender")
@@
 
-#### Model validation: TO DO
+#### Model validation
+```{r model validation}
+## Models
+get.likelihood.other <- function(weight, bites, pop, params) {
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  exp_bites <- inv.logit(B_ttimes*weight + B_0)*pop
+  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
+}
+
+## Distance to CTAR
+ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
+district_coords <- coordinates(mada_district)
+library(geosphere)
+distance_mat <- distm(district_coords, ctar_coords)
+dist_min <- apply(distance_mat, 1, min)/1000
+exps_dist$dist_min <- dist_min[match(exps_dist$district, mada_district$mdg_dis_co)]
+
+## Unweighted travel times
+exps_dist$ttimes_unweighted <- mada_district$study_area[match(exps_dist$district,
+                                                              mada_district$mdg_dis_co)]
+
+### District parameters
+dist_pars <- optim(par = c(1e-5, -0.1), get.likelihood.other, 
+                   weight = exps_dist$dist_min, bites = exps_dist$bites, pop = exps_dist$pop,
+                   hessian = TRUE)
+B_ttimes_dist <- dist_pars$par[1]
+B_0_dist <- dist_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(dist_pars$hessian)
+
+ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
+}
+
+B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+plot(exps_dist$bites, inv.logit(B_ttimes_dist*exps_dist$dist_min+ B_0_dist)*exps_dist$pop)
+plot(exps_dist$ttimes_weighted_dist, exps_dist$bites)
+
+### Unweighted
+dist_pars <- optim(par = c(1e-3, 0.1), get.likelihood.other, 
+                   weight = exps_dist$ttimes_unweighted, bites = exps_dist$bites, 
+                   pop = exps_dist$pop,
+                   hessian = TRUE)
+B_ttimes_dist <- dist_pars$par[1]
+B_0_dist <- dist_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(dist_pars$hessian)
+
+ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
+}
+
+B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+plot(exps_dist$bites, exp(B_ttimes_dist*exps_dist$ttimes_unweighted+ B_0_dist))
+plot(exps_dist$ttimes_weighted_dist, exps_dist$dist_min)
+
+### Pop
+get.likelihood.pop <- function(bites, pop, params) {
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  exp_bites <- exp(B_ttimes*pop/1e5 + B_0)
+  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
+}
+pop_pars <- optim(par = c(1e-5, 0.1), get.likelihood.pop, 
+                   pop = exps_dist$pop, bites = exps_dist$bites, 
+                   hessian = TRUE)
+B_pop <- pop_pars$par[1]
+B_0_pop <- pop_pars$par[2]
+check <- glm(round(bites) ~ pop, data = exps_dist, family = "poisson")
+plot(exps_dist$bites, exp(B_pop*exps_dist$pop/1e5 + B_0_pop), pch = 20, col = alpha("red", 0.5))
+plot(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
+abline(a = 0, b = 1, col = "grey", lty = 2)
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(pop_pars$hessian)
+
+ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
+}
+
+B_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+## @ commune level
+get.likelihood.mada <- function(commune, district, params) {
+  # To test likelihood function
+  # commune = comm_data
+  # district = exps_dist
+  # params = c(1e-6, 0.1)
+  B_ttimes <- params[1]
+  B_0 <- params[2]
+  commune %>%
+      filter(exclude == 0) %>%
+      group_by(mdg_dis_co) %>%
+      mutate(bites = exp(B_ttimes*pop/1e5 + B_0)) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(district, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
+}
+comm_pars <- optim(par = c(1e-3, 0.1), get.likelihood.mada, 
+                   commune = comm_data, district = exps_dist, 
+                   hessian = TRUE)
+B_ttimes_comm <- comm_pars$par[1]
+B_0_comm <- comm_pars$par[2]
+
+## Getting standard errors
+# this is how you can get standard errors from optim, but inverting the hessian
+se <- sqrt(abs(diag(solve(comm_pars$hessian)))); se
+
+#simulate "standard errors"
+varcov <- solve(comm_pars$hessian)
+
+ntest = 1000; testpar = comm_pars$par; sim.par = matrix(NA, ntest, 2)
+for (k in 1:ntest) {
+  sim.par[k, ] = mvrnorm(1, comm_pars$par, varcov)
+}
+
+B_ttimes_comm_CI <- quantile(sim.par[,1],c(0.025,0.975))
+B_0_comm_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+comm_data %>%
+      filter(exclude == 0) %>%
+      mutate(bites = exp(B_ttimes_comm*pop/1e5 + B_0_comm)) %>%
+      group_by(mdg_dis_co) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
+                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
+                   xlim = c(0, 5000))
+comm_data %>%
+      filter(exclude == 0) %>%
+      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
+      group_by(mdg_dis_co) %>%
+      summarize(sum_bites = sum(bites)) %>%
+      left_join(select(exps_dist, district, dist_bites = bites), 
+                by = c("mdg_dis_co" = "district")) %>%
+      filter(!is.na(dist_bites)) -> sum_bites
+sum_bites %$% plot(dist_bites, sum_bites, pch = 20, col = alpha("red", 0.5))
+abline(a = 0, b = 1, col = "grey", lty = 2)
+
+exps_dist %>%
+      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check1
+points(check1$bites, check1$bites_mu, col = alpha("purple", 0.5), pch = 20)
+points(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
+abline(a = 0, b = 1, col = "grey", lty = 2)
+
+mora_bites %>%
+      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
+points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("blue", 0.5))
+check <- glm(round(bites) ~ pop, data = mora_bites, family = "poisson")
+plot(mora_bites$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
+abline(a = 0, b = 1, col = "grey", lty = 2)
+
+## Correlation plot
+
+## Table of fits/parameter results
+
+## Impact of contacts
+
+## Impact of RLE threshold
+
+```
 
@@
 ## Comparing scales
-## Function for reporting
-get.reporting_burden <- function(pop, ttimes, B_ttimes, B_0, hdr = 25, incidence = 0.01, 
-                          exp_rate = 0.39, p_rabid = 0.6, rho_max = 0.98, p_death = 0.16, 
-                          type = "reporting") {
-  bites_100k <- exp(B_ttimes*ttimes + B_0)*pop
-  R_100k <- pop/hdr*incidence*exp_rate
-  if (type == "reporting"){
-    rho <- bites_100k*(ifelse(bites_100k*p_rabid > R_100k, rho_max*R_100k/bites_100k, p_rabid))/R_100k
-    return(rho)
-  } else {
-    unreported <- R_100k - bites_100k*(ifelse(bites_100k*p_rabid > R_100k, 
-                                              rho_max*R_100k/bites_100k, p_rabid))
-    deaths <- 0.16*unreported
-    return(deaths)
-  }
-}
-
 comm_data$reporting_min <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
@@
 ```{r burden results, fig.cap = "Figure 7", cache = TRUE}
-## Commune level
+# Incremental analysis ------------------------------------------------------------------------
+## District travel times
+dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+
+dist_ttimes <- cbind(dist_catchments$ttimes_weighted, dist_ttimes)
+dist_mat <- matrix(NA, nrow(dist_ttimes), ncol(dist_ttimes))
+dist_mat[, 1] <- dist_ttimes[, 1]
+for (i in 2:ncol(dist_ttimes)){
+  dist_mat[, i] <- ifelse(dist_ttimes[, i] < dist_ttimes[, i - 1], dist_ttimes[, i], NA)
+}
+dist_mat <- as.data.table(dist_mat)
+dist_mat$mdg_dis_co <- as.character(mada_district$mdg_dis_co)
+dist_mat <- melt(dist_mat, id.vars = "mdg_dis_co", value.name = "ttimes", variable.name = "scenario")
+dist_mat$scenario <- as.numeric(gsub("V", "", dist_mat$scenario)) - 1
+
+dist_mat %>%
+  drop_na(ttimes) %>%
+  left_join(select(mada_district@data, mdg_dis_co, district, pop2015adj)) -> dist_mat
+
+## Commune travel times
+comm_ttimes <- cbind(comm_catchments$ttimes_weighted, comm_ttimes)
+comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
+comm_mat[, 1] <- comm_ttimes[, 1]
+for (i in 2:ncol(comm_ttimes)){
+  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], comm_ttimes[, i], NA)
+}
+comm_mat <- as.data.table(comm_mat)
+comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
+comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+
+comm_mat %>%
+  drop_na(ttimes) %>%
+  left_join(select(comm_data, mdg_cm_, pop)) -> comm_mat
+
+## Estimate burden
 source("R/get.burden.R")
-results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
-results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
-results_pr20$pr <- "0.20"
-results_pr60$pr <- "0.60"
-results <- rbind(results_pr20, results_pr60)
-
-results %>% 
-  group_by(mdg_dis_co, pr) %>% 
-  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI),
-            lowerCI = sum(deaths_lowerCI)) %>%
-  left_join(dist_catchments) %>% 
-  left_join(select(mada_district@data, mdg_dis_co, district)) -> results_grouped
-
-## District level
-results_pr20_dist <- get.burden.dist(p_rabid = 0.20, rho_max = 0.98)
-results_pr60_dist <- get.burden.dist(p_rabid = 0.60,rho_max = 0.98)
-results_pr20_dist$pr <- "0.20"
-results_pr60_dist$pr <- "0.60"
+system.time({
+  results_pr20_comm <- get.burden(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
+                           pop = comm_mat$pop, scenario = comm_mat$scenario,
+                           param_ttimes = B_ttimes_mada,
+                           param_intercept = B_0_mada,
+                           p_rabid = 0.2, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+
+results_pr60_comm <- get.burden(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
+                           pop = comm_mat$pop, scenario = comm_mat$scenario,
+                           param_ttimes = B_ttimes_mada,
+                           param_intercept = B_0_mada,
+                           p_rabid = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+})
+
+## Filling it in
+results_pr20_comm %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> results_pr20_comm
+
+results_pr60_comm %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> results_pr60_comm
+
+system.time({
+results_pr20_dist <- get.burden(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
+                           pop = dist_mat$pop, scenario = dist_mat$scenario,
+                           param_ttimes = B_ttimes_dist,
+                           param_intercept = B_0_dist,
+                           p_rabid = 0.2, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+
+results_pr60_dist <- get.burden(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
+                           pop = dist_mat$pop, scenario = dist_mat$scenario,
+                           param_ttimes = B_ttimes_dist,
+                           param_intercept = B_0_dist,
+                           p_rabid = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+})
+
+results_pr20_dist %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> results_pr20_dist
+
+results_pr60_dist %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> results_pr60_dist
+```
+
+``` {r plotting burden} 
+## Group commune results to district level
+results_comm <- rbind(results_pr20_comm, results_pr60_comm)
+colnames(results_comm) <- gsub("results.", "", colnames(results_comm))
+results_comm %>%
+  left_join(select(comm_data, mdg_dis_co, mdg_cm_), 
+            by = c("names" = "mdg_cm_")) %>%
+  group_by(scenario, mdg_dis_co, pr) %>%
+  summarize(names = first(mdg_dis_co), 
+            deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_comm_summarized
+results_comm_summarized$model <- "commune"
 results_dist <- rbind(results_pr20_dist, results_pr60_dist)
-
+colnames(results_dist) <- gsub("results.", "", colnames(results_dist))
 results_dist %>%
-  left_join(results_grouped, by = c("mdg_dis_co" = "mdg_dis_co", "pr" = "pr")) -> results_comp
-
-ggplot(results_comp, aes (x = deaths_mean, y = tot, color = ttimes_weighted.x)) +
-  geom_point() +
-  geom_abline(slope = 1, intercept = 0, color = "grey", linetype = "dashed") +
-  facet_wrap (~pr)
-
-results_dist %>% 
-  group_by(mdg_dis_co, pr) %>% 
-  summarize(tot = sum(deaths_mean), upperCI = sum(deaths_upperCI),
-            lowerCI = sum(deaths_lowerCI)) %>%
-  left_join(dist_catchments) %>% 
-  left_join(select(mada_district@data, mdg_dis_co, district)) -> results_grouped_dist
-results_grouped_dist$scale <- "district"
-results_grouped$scale <- "commune"
-
-results_all <- rbind(results_grouped, results_grouped_dist)
-
-ggplot(results_all, aes(y = reorder(district, ttimes_weighted), x = tot, 
-                              color = pr)) +
-  ggstance::geom_pointrangeh(aes(xmin = lowerCI, xmax = upperCI), fatten = 2) +
+  mutate(model = "district") %>%
+  select(scenario, names, pr, deaths_mean, deaths_lowerCI, deaths_upperCI, 
+         model) %>%
+  bind_rows(., results_comm_summarized) -> results_all
+results_all %>%
+  left_join(select(mada_district@data, district, mdg_dis_co, ttimes_weighted),
+            by = c("names" = "mdg_dis_co")) -> results_all
+
+## Plot national level burden
+results_all %>% 
+  group_by(scenario, pr, model) %>% 
+  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_natl
+
+## Plotting national level
+ggplot(data = results_natl, aes (x = scenario, y = deaths_mean,
+                                 color = interaction(model, pr))) +
+  geom_line() +
+  geom_ribbon(aes(ymin = deaths_upperCI, ymax = deaths_lowerCI, color = NULL, 
+                  fill = interaction(model, pr)), alpha = 0.5) +
+  scale_color_manual(values = c("district.0.2" = "blue", 
+                                "district.0.6" = "darkblue", 
+                                "commune.0.2" = "magenta", 
+                                "commune.0.6" = "purple"), 
+                     name = "model pars") +
+  scale_fill_manual(values = c("district.0.2" = "blue", 
+                                "district.0.6" = "darkblue", 
+                                "commune.0.2" = "magenta", 
+                                "commune.0.6" = "purple"), 
+                    guide = "none")
+
+## Plotting national level
+ggplot(data = results_all, aes(x = deaths_mean, y = reorder(district, ttimes_weighted), 
+                               color = scenario, shape = as.factor(pr))) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
+                             fatten = 2) +
+  facet_wrap(~model + as.factor(pr))
+
+## Baseline results
+results_all %>% filter(scenario == 0) -> results_all_base
+ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
+                              color = model, shape = as.factor(pr))) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
   ylab("") +
   xlab("Average annual deaths") +
-  scale_color_manual(values = c("0.20" = alpha("red", 0.2), "0.60" = alpha("red", 0.6)), 
-                     name = "Proportion rabid") +
-  facet_wrap(~scale) -> p
+  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
+                                "commune" = alpha("purple", 0.5)), 
+                     name = "Scale") +
+  facet_wrap(~ as.factor(pr)) -> p
 p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
 
-## Maps of burden
-## district level
-mada_district %>%
-  tidy(., region = "mdg_dis_co") %>%
-  left_join(results_pr60_dist, by = c("id" = "mdg_dis_co")) -> ggmada_dist
-
-ggplot(ggmada_dist, aes(long, lat, group = group, fill = deaths_mean)) +
-  geom_polygon() +
-  scale_fill_continuous(type = "viridis") -> burden_map
-burden_map
-ggsave("burden_map_district.pdf")
-
-## commune level
-mada_communes %>%
-  tidy(., region = "mdg_cm_") %>%
-  left_join(results_pr60, by = c("id" = "mdg_cm_")) -> ggmada_commune
-
-ggplot(ggmada_commune, aes(long, lat, group = group, fill = deaths_mean)) +
-  geom_polygon() +
-  scale_fill_continuous(type = "viridis") -> burden_map
-burden_map
-ggsave("burden_map_commune.pdf")
+```
+
+```{r doing it from the median, cache=TRUE}
+system.time({
+  results_comm_med <- get.burden_med(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
+                           pop = comm_mat$pop, scenario = comm_mat$scenario,
+                           param_ttimes = B_ttimes_mada,
+                           param_intercept = B_0_mada,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+})
+
+## Filling it in
+results_comm_med %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(results_comm_med), .direction = "down") -> results_comm_med
+
+system.time({
+results_dist_med <- get.burden_med(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
+                           pop = dist_mat$pop, scenario = dist_mat$scenario,
+                           param_ttimes = B_ttimes_dist,
+                           param_intercept = B_0_dist,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000)
+})
+
+results_dist_med %>%
+  complete(scenario, names) %>%
+  group_by(names) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> results_dist_med
+
+## Group commune results to district level
+colnames(results_comm_med) <- gsub("results.", "", colnames(results_comm_med))
+results_comm_med %>%
+  left_join(select(comm_data, mdg_dis_co, mdg_cm_), 
+            by = c("names" = "mdg_cm_")) %>%
+  group_by(scenario, mdg_dis_co) %>%
+  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) %>%
+  rename(names = mdg_dis_co) -> results_comm_summarized
+results_comm_summarized$model <- "commune"
+colnames(results_dist_med) <- gsub("results.", "", colnames(results_dist_med))
+results_dist_med %>%
+  mutate(model = "district") %>%
+  select(scenario, names, deaths_mean, deaths_lowerCI, deaths_upperCI, 
+         model) -> results_dist_med
+results_all <- bind_rows(results_dist_med, results_comm_summarized)
+results_all %>%
+  left_join(select(mada_district@data, district, mdg_dis_co, ttimes_weighted),
+            by = c("names" = "mdg_dis_co")) -> results_all
+
+## Plot national level burden
+results_all %>% 
+  group_by(scenario, model) %>% 
+  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_natl
+
+## Plotting national level
+ggplot(data = results_natl, aes (x = scenario, y = deaths_mean, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = deaths_upperCI, ymax = deaths_lowerCI, 
+                  color = NULL, fill = model),
+              alpha = 0.5) +
+  scale_color_manual(values = c("district" = "blue", 
+                                "commune" = "purple"), 
+                     guide = "none") +
+  scale_fill_manual(values = c("district" = "blue", 
+                                "commune" = "purple")) +
+  guides(color = guide_legend(), 
+         fill = guide_legend(override.aes = list(linetype = 1)))
+
+## Plotting national level
+ggplot(data = results_all, aes(x = deaths_mean, y = reorder(district, ttimes_weighted), 
+                               color = scenario)) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
+                             fatten = 2) +
+  facet_wrap(~model)
+
+## Baseline results
+results_all %>% filter(scenario == 0) -> results_all_base
+ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
+                              color = model)) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
+  ylab("") +
+  xlab("Average annual deaths") +
+  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
+                                "commune" = alpha("purple", 0.5)), 
+                     name = "Scale") -> p
+p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
 
@@
 When we estimate burden of deaths stochastically within this range of incidence and given our high and low estimates of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure 7, results presented aggregated at the district level).
-Overall, we estimate average annual deaths between `r round(sum(results_pr60$deaths_mean))` - `r round(sum(results_pr20$deaths_mean))`.
+Overall, we estimate average annual deaths between ....
 **Also estimate deaths averted here!**.
@@
 
-## Decrease in burden over times
+## Decrease in burden over times proportional
 plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
@@
 lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
-     type = "l", col = "blue", 
-     bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", ylab = "Number of deaths")
+     type = "l", col = "blue")
 lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")
 
+## Decrease in burden over times absolute
+plot(colSums(comm_burden_min, na.rm = TRUE), type = "l", col = "blue", bty = "n",
+     ylim = c(0, 1500), xlab = "# Additional ARMC",
+     ylab = "Number of deaths")
+lines(colSums(dist_burden_min, na.rm = TRUE), col = "purple")
+lines(colSums(comm_burden_max, na.rm = TRUE), type = "l", col = "blue")
+lines(colSums(dist_burden_max, na.rm = TRUE), col = "purple")
+
 ```
@@
 ### Sensitivity analyses
-
+```{r sensitivity analysis multivariate}
+hdr_vals <- c(25, 20, 15, 10, 5)
+p_rabid_vals <- c(0.2, 0.3, 0.4, 0.5, 0.6)
+rho_max <- c(0.80, 0.85, 0.9, 0.95, 0.9)
+library(foreach)
+
+dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
+dist_ttimes <- as.data.table(dist_ttimes)
+dist_ttimes$mdg_dis_co <- as.character(mada_district$mdg_dis_co)
+dist_ttimes <- melt(dist_ttimes, id.vars = "mdg_dis_co", value.name = "ttimes", variable.name = "scenario")
+dist_ttimes$scenario <- as.numeric(gsub("result.", "", dist_ttimes$scenario))
+dist_ttimes <- rbind(dist_ttimes, 
+                       as.data.frame(list(mdg_dis_co = dist_catchments$mdg_dis_co, scenario = 0,
+                                          ttimes = dist_catchments$ttimes_weighted)))
+dist_ttimes %>%
+  left_join(select(mada_district@data, mdg_dis_co, district, pop = pop2015adj)) -> dist_ttimes
+
+scenario_dist <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+  get.reporting_burden(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+                     ttimes = dist_ttimes$ttimes, scenario = dist_ttimes$scenario,
+                     B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
+                     hdr = hdr_vals[k], p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden")
+  }
+
+scenario_dist %>%
+  filter(scenario == 0) %>%
+  group_by(hdr, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ hdr)
+ggplot(data = filter(scenario_dist, scenario == 0), aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ hdr)
+
+scenario_dist %>%
+  group_by(scenario, hdr, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths)) -> deaths_scenario
+deaths_scenario %>%
+  left_join(select(deaths_multivar, hdr, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
+ggplot(data = deaths_scenario, aes(x = scenario, y = deaths, color = p_rabid)) +
+  geom_point() +
+  facet_grid(hdr ~ rho_max, scales = "free_y")
+
+```
+
+```{r scaling w/ pop}
+## Pop scaling
+pop <- mada_district$pop2015adj
+pop <- pop[order(pop, decreasing = FALSE)]
+inc100k_max <- 0.01*0.39/5
+inc100k_min <- 0.01*0.39/25
+pos_scale <- seq(0.01*0.39/25, 0.01*0.39/5, length.out = length(pop))
+neg_scale <- seq(0.01*0.39/5, 0.01*0.39/25, length.out = length(pop))
+pos <- lm(pos_scale ~ pop) ## use these and constrain
+neg <- lm(neg_scale ~ pop) ## use these and constrain
+
+scale_df <- rbind(as.data.frame(list(scale = "neg", 
+                                     sf = seq(neg$coefficients[2], 0, length.out = 5),
+                                     intercept = inc100k_max)),
+                  as.data.frame(list(scale = "pos", 
+                                     sf = seq(0, pos$coefficients[2], length.out = 5),
+                                     intercept = inc100k_min)))
+constrained_inc <- function(slope, intercept, pop, max, min){
+  inc <- slope*pop + intercept
+  inc[inc >= max] <- max
+  inc[inc <= min] <- min
+  return(inc)
+}
+scale_df %>%
+  merge(., select(mada_district@data, mdg_dis_co, ttimes_weighted, pop = pop2015adj)) -> scale_df_dist
+ggplot(data = scale_df_dist, aes(x = log(pop), 
+                            y = constrained_inc(sf, intercept, pop, inc100k_max, inc100k_min)*1e5, 
+                            color = sf)) +
+  geom_point() +
+  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey")
+
+## getting scaling factors 
+scenario_scaled <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+  get.reporting_burden(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+                     ttimes = dist_ttimes$ttimes, scenario = dist_ttimes$scenario,
+                     B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
+                     p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
+                     slope = scale_df$sf[k], intercept = scale_df$intercept[k])
+  }
+
+scenario_scaled %>%
+  filter(scenario == 0) %>%
+  group_by(slope, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ slope)
+ggplot(data = filter(scenario_scaled, scenario == 0),  
+       aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ slope)
+
+scenario_scaled %>%
+  group_by(scenario, slope, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths)) -> deaths_scenario
+deaths_scenario %>%
+  left_join(select(deaths_multivar, slope, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
+ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
+  geom_point() +
+  facet_grid(slope ~ rho_max, scales = "free_y")
+
+```
 <br>

### 2019-02-27 accessPEP.Rmd
all the messy model fitting code

@@
   summarize(bites = round(n()/n_months*12)) %>%
-  left_join(comm_data) %>%
+  right_join(filter(comm_data, CTAR == "Moramanga")) %>% ## only districts within catchment (@ comm level)
   filter(!is.na(ttimes_weighted), 
-         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
+         dist_catch == "Moramanga") %>%
+  replace_na(list(bites = 0)) -> mora_bites
 ```
@@
   summarise(bites = sum(bites, na.rm = TRUE)) %>%
-  left_join(covars) -> exps
+  right_join(covars) %>%
+  filter(exclude == 0) %>%
+  replace_na(list(bites = 0)) -> exps
 
@@
 
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows (TO DO: take neg bin around min and max for RI instead):
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows):
 $$
-deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
+deaths_i = (R_i - p_{rabid_i}B_i) \times p_{death}
 $$
 where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
-We constrain $p_rabid$ as in Equation N, and we assume that all rabies exposures reported to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
+We draw $p_rabid_i$ from a uniform distribution between 0.2 - 0.6, while constraining it as per Equation N.
+We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
 The probability of death in the absence of PEP is taken from cite GAVI/Joel paper.
@@
 Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC.
-Starting with the current locations, we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country.
+Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country.
 We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially.
@@
 ### Sensitivity analyses for burden estimates 
-- R (increasing or decreasing with pop) (min + max)
-- P-rabid min and max
-- Probability of death min and max
-- Look at change in raw # of deaths and also change in burden by scenarios
+We estimate burden deterministically across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden.
+Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden.
+We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden changes if rabies incidence were to scale positively or negatively with human population.
 
@@
 library(boot)
-
-get.likelihood.bites <- function(bites, covar, pop, params) {
-  beta <- params[1]
-  intercept <- params[2]
+likelihoods <- list(
+  get.likelihood.bites = function(bites, covar, pop, beta, intercept, ...) {
     exp_bites <- inv.logit(beta*covar + intercept)*pop
     return(-sum(dpois(round(bites), lambda = exp_bites, log = TRUE)))
-}
-
-get.likelihood.mora <- function(commune, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      mutate(bites_mu = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
-  return(-sum(dpois(round(sum_bites$), lambda = sum_bites$bites_mu, log = TRUE)))
-}
-
-get.likelihood.mada <- function(commune, district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-get.likelihood.dist <- function(district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  district %>%
-      filter(exclude == 0) %>%
-      mutate(exp_bites = inv.logit(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
-  return(-sum(dpois(round(district$exp_bites), lambda = round(district$bites), log = TRUE)))
-}
+  },
+  get.likelihood.sumbites = function(bites, aggregate_bites, covar_agg, aggregate_covar, 
+                                     pop_agg, beta, 
+                                     intercept, ...) {
+    exp_bites <- inv.logit(beta*covar_agg + intercept)*pop_agg
+    df <- as.data.frame(list(exp_bites = exp_bites, names = aggregate_covar))
+    df %>%
+      group_by(names) %>%
+      summarize(sum_bites = sum(exp_bites, na.rm = TRUE)) %>%
+      left_join(as.data.frame(list(bites = bites, names = aggregate_bites))) -> sum_bites
+    return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$sum_bites, log = TRUE)))
+  }, 
+  get.likelihood.pop = function(bites, pop, beta, intercept, trans...) {
+    exp_bites <- exp(beta*pop/trans + intercept)
+    return(-sum(dpois(round(bites), lambda = exp_bites, log = TRUE)))
+  },
+  get.likelihood.sumpop = function(bites, aggregate_bites, aggregate_covar, pop_agg, beta, 
+                                   intercept, trans, ...) {
+    exp_bites <- exp(beta*pop_agg/trans + intercept)
+    df <- as.data.frame(list(exp_bites = exp_bites, names = aggregate_covar))
+    df %>%
+      group_by(names) %>%
+      summarize(sum_bites = sum(exp_bites, na.rm = TRUE)) %>%
+      left_join(as.data.frame(list(bites = bites, names = aggregate_bites))) -> sum_bites
+    return(-sum(dpois(round(sum_bites$bites), lambda = sum_bites$sum_bites, log = TRUE)))
+  })
+
+mods_mada <- lapply(likelihoods, function(x) mle2(minuslogl = x, 
+                                     data = list(bites = exps_dist$bites, 
+                                                 aggregate_bites = exps_dist$district,
+                                                 aggregate_covar = comm_data$mdg_dis_co,
+                                                 covar_agg = comm_data$ttimes_weighted/60,
+                                                 covar = exps_dist$ttimes_weighted_dist/60, 
+                                                 pop = exps_dist$pop, 
+                                                 pop_agg = comm_data$pop, trans = 1e5), 
+       start = list(beta = 1e-5, intercept = 0.1)))
+prof_CI <- lapply(mods_mada, confint)
+
+mada_mod_predicts <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_mada), 
+     intercept = mapply(function(x) coef(x)["intercept"], mods_mada), 
+     beta_upper = mapply(function(x) x["beta", 2], prof_CI),
+     beta_lower = mapply(function(x) x["beta", 1], prof_CI),
+     intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
+     intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
+     likelihood = mapply(function(x) logLik(x), mods_mada),
+     data = "Mada"))
+
+mada_mod_results <- as.data.frame(list(params = c("beta", "intercept"), 
+                   estimates = t(mapply(function(x) coef(x), mods_mada)), 
+                   upper = t(mapply(function(x) x[, 2], prof_CI)),
+                   lower = t(mapply(function(x) x[, 1], prof_CI)), 
+                   data = "Mada"))
+
+## Predictions
+library(foreach)
+predicts <- list(
+  pred.ttimes = function(bites, covar, pop, beta, intercept, ...) {
+    exp_bites <- inv.logit(beta*covar + intercept)*pop
+    return(as.data.frame(list(observed = bites, 
+                              predicted = rpois(length(bites), exp_bites))))
+  },
+  pred.sumttimes = function(bites, aggregate_bites, covar_agg, aggregate_covar, 
+                                     pop_agg, beta, 
+                                     intercept, ...) {
+    exp_bites <- inv.logit(beta*covar_agg + intercept)*pop_agg
+    df <- as.data.frame(list(exp_bites = exp_bites, names = aggregate_covar))
+    df %>%
+      group_by(names) %>%
+      summarize(sum_bites = sum(exp_bites, na.rm = TRUE)) %>%
+      left_join(as.data.frame(list(bites = bites, names = aggregate_bites))) -> sum_bites
+    return(as.data.frame(list(observed = sum_bites$bites, 
+                              predicted = rpois(nrow(sum_bites), sum_bites$sum_bites))))
+  }, 
+  pred.pop = function(bites, pop, beta, intercept, trans, ...) {
+    exp_bites <- exp(beta*pop/trans + intercept)
+   return(as.data.frame(list(observed = bites, 
+                              predicted = rpois(length(bites), exp_bites))))
+  },
+  pred.sumpop = function(bites, aggregate_bites, aggregate_covar, pop_agg, beta, 
+                                   intercept, trans, ...) {
+    exp_bites <- exp(beta*pop_agg/trans + intercept)
+    df <- as.data.frame(list(exp_bites = exp_bites, names = aggregate_covar))
+    df %>%
+      group_by(names) %>%
+      summarize(sum_bites = sum(exp_bites, na.rm = TRUE)) %>%
+      left_join(as.data.frame(list(bites = bites, names = aggregate_bites))) -> sum_bites
+    return(as.data.frame(list(observed = sum_bites$bites, 
+                              predicted = rpois(nrow(sum_bites), sum_bites$sum_bites))))
+  })
+mada_mod_predicts$predict_fun <- predicts
+preds <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
+
+  check <- mada_mod_predicts$predict_fun[[i]]
+
+  pred <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
+          bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+          aggregate_covar = comm_data$mdg_dis_co, covar_agg = comm_data$ttimes_weighted/60,
+          covar = exps_dist$ttimes_weighted_dist/60, 
+          pop = exps_dist$pop, pop_agg = comm_data$pop, trans = 1e5)
+
+  pred$upper <- check(beta = mada_mod_predicts$beta_upper[i], 
+                      intercept = mada_mod_predicts$intercept_upper[i],
+                      bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+                      aggregate_covar = comm_data$mdg_dis_co, covar_agg = comm_data$ttimes_weighted/60,
+                      covar = exps_dist$ttimes_weighted_dist/60, 
+                      pop = exps_dist$pop, pop_agg = comm_data$pop, trans = 1e5)$predicted
+  pred$lower <- check(beta = mada_mod_predicts$beta_lower[i], 
+                      intercept = mada_mod_predicts$intercept_lower[i],
+                      bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+                      covar = exps_dist$ttimes_weighted_dist/60, 
+                      aggregate_covar = comm_data$mdg_dis_co, covar_agg = comm_data$ttimes_weighted/60,
+                      pop = exps_dist$pop, pop_agg = comm_data$pop, trans = 1e5)$predicted
+
+  pred$model <- rownames(mada_mod_predicts)[i]
+  pred$data <- mada_mod_predicts$data[i]
+  pred
+}
+
+expected <- list(
+  pred.ttimes = function(covar, pop, beta, intercept, ...) {
+    return(inv.logit(beta*covar + intercept)*pop)
+  },
+  pred.sumttimes = function(covar, pop, beta, 
+                                     intercept, ...) {
+    return(inv.logit(beta*covar + intercept)*pop)
+  }, 
+  pred.pop = function(pop_covar, beta, intercept, trans, ...) {
+   return(exp(beta*pop_covar/trans + intercept))
+  },
+  pred.sumpop = function(pop_covar, beta, intercept, trans, ...) {
+    return(exp(beta*pop_covar/trans + intercept))
+  })
+mada_mod_predicts$expect_fun <- expected
+ttimes_plot <- seq(0, 720, by = 30)
+pop_plot <- seq(100, 1e6, length.out = length(ttimes_plot))
+expect <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
+  check <- mada_mod_predicts$expect_fun[[i]]
+  exp <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
+               covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, 
+               trans = 1e5)
+  exp_upper <- check(beta = mada_mod_predicts$beta_upper[i], 
+                     intercept = mada_mod_predicts$intercept_upper[i],
+                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
+  exp_lower <- check(beta = mada_mod_predicts$beta_lower[i], 
+                     intercept = mada_mod_predicts$intercept_lower[i],
+                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
+  as.data.frame(list(ttimes = ttimes_plot, 
+                     pop = pop_plot, exp = exp, exp_upper = exp_upper, exp_lower = exp_lower, 
+                     model = rownames(mada_mod_predicts)[i], data = mada_mod_predicts$data[i]))
+}
+
+ggplot(data = preds, aes(x = observed, y = predicted, color = model)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey")
+ggplot(data = filter(expect, !grepl("pop", model, fixed = TRUE)), 
+       aes(x = ttimes, y = exp, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = exp_lower, ymax = exp_upper, fill = model), alpha = 0.5)
+
+## Mada pars
+library(bbmle)
+mada_ttimes_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-4, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, covar = exps_dist$ttimes_weighted_dist, 
+                              pop = exps_dist$pop))
+mada_sumttimes_fit <- mle2(get.likelihood.sumbites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+                              covar = comm_data$ttimes_weighted, 
+                              aggregate_covar = comm_data$mdg_dis_co,
+                              pop = comm_data$pop))
+## Pop
+mada_pop_fit <- mle2(get.likelihood.pop, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, pop = exps_dist$pop/1e5))
+mada_sumpop_fit <- mle2(get.likelihood.sumpop, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+                              aggregate_covar = comm_data$mdg_dis_co,
+                              pop = comm_data$pop/1e5))
+
+## Distance to CTAR
+ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
+district_coords <- coordinates(mada_district)
+library(geosphere)
+distance_mat <- distm(district_coords, ctar_coords)
+mada_district$dist_min <- apply(distance_mat, 1, min)/1000
+exps_dist$dist_min <- mada_district$dist_min[match(exps_dist$district, mada_district$mdg_dis_co)]
+mada_distance_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, covar = exps_dist$dist_min, 
+                              pop = exps_dist$pop))
+
+mada_communes@data %>%
+  mutate(long = coordinates(mada_communes)[, 1], lat = coordinates(mada_communes)[, 2]) %>%
+  select(commune, mdg_cm_, mtch_ds, long, lat) %>%
+  right_join(comm_data) %>%
+  mutate(distance = apply(distm(select(., long, lat), ctar_coords), 1, min)/1000) -> comm_data
+mada_sumdist_fit <- mle2(get.likelihood.sumbites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, aggregate_bites = exps_dist$district,
+                              covar = comm_data$distance, 
+                              aggregate_covar = comm_data$mdg_dis_co,
+                              pop = comm_data$pop))
+
+plot(exps_dist$bites, inv.logit(mada_distance_fit@coef["beta"]*exps_dist$dist_min +
+                                  mada_distance_fit@coef["intercept"])*exps_dist$pop)
+points(exps_dist$bites, inv.logit(mada_ttimes_fit@coef["beta"]*exps_dist$ttimes_weighted_dist +
+                                  mada_ttimes_fit@coef["intercept"])*exps_dist$pop, col = "blue")
+points(exps_dist$bites, exp(mada_pop_fit@coef["beta"]*exps_dist$pop/1e5 +
+                                  mada_pop_fit@coef["intercept"]), col = "red")
+abline(a = 0, b = 1, lty = 2)
+AICctab(mada_ttimes_fit, mada_distance_fit, mada_pop_fit, mada_sumttimes_fit, mada_sumpop_fit,
+        mada_sumdist_fit, nobs = nrow(exps_dist))
+
+exps_dist %>%
+      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
+points(check$bites, check$bites_mu, col = alpha("purple", 0.5), pch = 20)
 
 ### Moramanga model ###
-mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
-B_ttimes_mora <- mora_pars$par[1]
-B_0_mora <- mora_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(mora_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(mora_pars$hessian)
-
-ntest = 1000; testpar = mora_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, mora_pars$par, varcov)
-}
-
-B_ttimes_mora_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_mora_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-### Madagascar model ###
-mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mada, district = exps_dist,
-             commune = comm_data, hessian = TRUE)
-B_ttimes_mada <- mada_pars$par[1]
-B_0_mada <- mada_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(mada_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(mada_pars$hessian)
-
-ntest = 1000; testpar = mada_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, mada_pars$par, varcov)
-}
-
-B_ttimes_mada_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-### District parameters
-dist_pars <- optim(par = c(1e-6, 0.1), get.likelihood.dist, district = exps_dist,
-              hessian = TRUE)
-B_ttimes_dist <- dist_pars$par[1]
-B_0_dist <- dist_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(dist_pars$hessian)
-
-ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
-}
-
-B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
+mora_coords <- cbind(ctar_metadata$LONGITUDE[ctar_metadata$CTAR == "Moramanga"], 
+                     ctar_metadata$LATITUDE[ctar_metadata$CTAR == "Moramanga"])
+library(geosphere)
+mada_communes@data %>%
+  mutate(long = coordinates(mada_communes)[, 1], lat = coordinates(mada_communes)[, 2]) %>%
+  select(commune, mdg_cm_, mtch_ds, long, lat) %>%
+  right_join(mora_bites) %>%
+  mutate(distance = distm(select(mora_bites, long, lat), mora_coords)[, 1]/1000) -> mora_bites
+mora_distance_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = mora_bites$bites, covar = mora_bites$distance, 
+                              pop = mora_bites$pop))
+mora_pop_fit <- mle2(get.likelihood.pop, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = mora_bites$bites, pop = mora_bites$pop/1e5))
+mora_ttimes_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = mora_bites$bites, covar = mora_bites$ttimes_weighted/60, 
+                              pop = mora_bites$pop))
+AICctab(mora_ttimes_fit, mora_distance_fit, mora_pop_fit, nobs = nrow(mora_bites))
+
+mada_check_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = exps_dist$bites, covar = exps_dist$pop/1e5, 
+                              pop = exps_dist$pop))
+mora_check_fit <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1), 
+                  data = list(bites = mora_bites$bites, covar = mora_bites$pop/1e5, 
+                              pop = mora_bites$pop))
 
@@
   geom_line() +
-  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.5) +
-
-plot(ttimes_plot/60, inv.logit(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "l",
-     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
-points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-
-points(ttimes_plot/60, inv.logit(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "red")
-points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "pink")
-points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
-       lty = 2, col = "pink")
-
-points(ttimes_plot/60, inv.logit(B_ttimes_dist*ttimes_plot + B_0_dist)*pop_plot, type = "l", col = "purple")
-points(ttimes_plot/60, inv(B_ttimes_dist_CI[1]*ttimes_plot + B_0_dist_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lavender")
-points(ttimes_plot/60, exp(B_ttimes_dist_CI[2]*ttimes_plot + B_0_dist_CI[2])*pop_plot, type = "l",
-       lty = 2, col = "lavender")
-
-legend("topright", c("Mada communes", "Mada districts", "Moramanga communes"), 
-       col = c("darkblue", "purple", "red"), 
-       lty = c(1, 1, 1), 
-       bty = "n")
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.5)
 ```

### 2019-03-11 accessPEP.Rmd
cleaning up getting data and running models

--- 2019-02-26 accessPEP.Rmd
@@
 ```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
+## Clear memory
+rm(list = ls())
 ## Libraries
@@
 library(reshape2)
+library(geosphere)
+library(foreach)
+library(bbmle)
+library(boot)
+library(geosphere)
+
+## Files
 source("R/utils.R")
-
+source("R/models.R")
+source("R/data_functions.R")
 select <- dplyr::select
@@
 ## INTRODUCTION
-
-Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis.
+Inequities in access to care are a major driver of disease burden globally.
+Often, the populations at greatest risk of a given disease are the most underserved.
+Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful considerations of how best to optimize the allocation of limited resources.
+
+Canine mediated rabies is estimated to cause approximately 60,000 human deaths anually.
+These deaths can be prevented through prompt administration of post-exposure prophylaxis.
 However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
+GAVI bit here...potential to greatly reduce disease burden.
+
 Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
-
-Here--discuss geographic variation in access to vaccination and care--how this shapes mortality for other diseases.
-
-The majority of rabies burden studies use these data to estimate burden from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e.
+The majority of rabies burden studies thus far have integrated these data on reported bites into a probability decision tree framework to estimate rabies burden, often with the key assumption that overall reported bite incidence (i.e.
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
@@
 While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.
-The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level.
-Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.
 
@@
 Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
-Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.
+Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.
 
@@
 ## Need: weighted ttimes, pop, ctar_in, exclude
+ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
 comm_catchments %>%
@@
                    color, fill, ctar_dist)) -> comm_data
+mada_communes@data %>%
+  mutate(long = coordinates(mada_communes)[, 1], 
+         lat = coordinates(mada_communes)[, 2]) %>%
+  select(commune, mdg_cm_, mtch_ds, long, lat) %>%
+  right_join(comm_data) %>%
+  mutate(distance = apply(distm(select(., long, lat), ctar_coords), 1, min)/1000) -> comm_data
 comm_data <- filter(comm_data, ttimes_weighted != "Inf")
 
-# ## Data to include
-# nrow(dist_catchments) - sum(dist_catchments$exclude)
-# nrow(comm_data) - sum(comm_data$exclude)
-
-#### Moramanga data! ###
-## ctar for # bites and also checking days to rule out contacts
-col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
-ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
-ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))
-
-## Set-up time series
-tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
-                   year1 = 2016, tstep = "month", period = FALSE) {
-  get.consec (date, format_date, start, format_start,
-              year1, tstep, period) - 8 ## starts in Sep
-}
-
-## all dates
-ctar %>%
-  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
-ctar$month <- tfunc(ctar$date.reported)
-## subset to study period (Sep 2016 - to date)
-## checks for cases where month is NA!
-ctar <- subset(ctar, month > 0 & month < 28)
-
-# ctar district and commune
-ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
-ctar$commune <- trimws(ctar$commune, which = "right")
-ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
-ctar$district <- gsub(" \\(District\\)", "", ctar$district)
-
-## change all true/false to 0/1
-ctar[ctar == FALSE] <- 0
-
-## Bite data in Moramanga
-n_months <- max(ctar$month, na.rm = TRUE)
-
-ctar %>%
-  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
-            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
-  filter(consult.type == "Animal bite/exposure") %>%
-  group_by(mdg_cm_) %>%
-  summarize(bites = round(n()/n_months*12)) %>%
-  left_join(comm_data) %>%
-  filter(!is.na(ttimes_weighted), 
-         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
+## Getting district data
+mada_district@data %>% 
+  select(mdg_dis_co, district, pop = pop2015un, ttimes_weighted, ctar_in_dist, 
+         dist_catch = ctar_dist, exclude, color, fill) -> dist_data
+
+exps_dist <- get.district.data(ctar_data = ctar_data, ctar_metadata = ctar_metadata, 
+                               mada_district = mada_district, catchments, 
+                               consec_days = 15, 
+                               contact_cutoff = 3, reporting_thresh = 0.25, 
+                               start = ymd("2014-01-01"), end = ymd("2017-12-31"))
+exps_mora <- get.morabites(mada_communes = mada_communes)
+
 ```
@@
 
-```{r network fig}
-## Network fig
-## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments
-
-## get exposure matrix = first set of points
-## district exposures
-ctar_data %>% 
-  drop_na(id_ctar, district) %>%
-  group_by(district) %>% # group by district
-  summarize(n = n()) -> exps_dist # get count of exposures
-## ctar exposures
-ctar_data %>% 
-  drop_na(id_ctar) %>%
-  group_by(id_ctar) %>% # group by ctar
-  summarize(n = n()) -> exps_ctar
-## ctar by district exposure matrix
-ctar_data %>% 
-  drop_na(id_ctar, district) %>%
-  group_by(district, id_ctar) %>% # group by district and ctar
-  summarize(n = n()) -> exp_mat# get count of exposures
-
-## To plot district bites
-dist_points <- cbind(select(mada_district@data,
-                            mdg_dis_co), coordinates(mada_district))
-names(dist_points)[2:3] <- c("dist_long", "dist_lat")
-dist_points %>%
- left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
-           by = c("mdg_dis_co" = "ctar_dist")) %>%
- mutate(dist_long = coalesce(LONGITUDE, dist_long), 
-        dist_lat = coalesce(LATITUDE, dist_lat)) -> dist_points
-
-exps_dist %>%
-  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
-             by = c("district" = "mdg_dis_co")) %>%
-  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
-dist_points$size <- log(dist_points$n + 0.1)*0.25
-
-## To plot ctar points (only those not excluded)
-exps_ctar %>%
- left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE, 
-                  exclude, color, fill)) %>%
- filter(exclude == 0) -> ctar_points
-
-## To plot lines between ctar + districts
-exp_mat %>% 
-  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
-  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
-                   color, fill, ctar_dist)) -> dist_lines
-dist_lines$size <- log(dist_lines$n + 0.1)*0.25
-```
-
-```{r figure 1A}
-jpeg("figs/network.jpeg")
-plot(mada_district, col = mada_district$fill, border = "grey") 
-segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
-         y1 = dist_lines$LATITUDE,
-         lwd = dist_lines$size*2, col = dist_lines$color)
-points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
-       cex = log(ctar_points$n + 0.1)*0.25)
-points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
-       col = dist_points$color)
-legend("topright", c("10", "50", "100", "200", "400", "800", "1600"), pch = 1, xjust = 1, yjust = 1,
-       pt.cex = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, 
-       lwd = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, bty = "n", y.intersp = 0.75,
-       title = "Number of \n reported bites", xpd = TRUE)
-legend("topleft", "A", bty = "n", inset = c(0.2, 0), text.font = 2)
-dev.off()
-```
-
-```{r network output, echo=FALSE, fig.cap="Figure 1A", results = "show", out.width = '100%'}
-knitr::include_graphics("figs/network.jpeg")
-```
-
-```{r figure 1b, fig.cap = "Figure 1B"}
-exp_mat %>% 
-  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
-  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
-            by = c("district" = "mdg_dis_co")) %>%
-  group_by(district) %>% # group by ctar
-  mutate(prop = n/sum(n),
-         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
-  filter(catch_match == 1) -> catch_dist
-hist(catch_dist$prop, main = "", 
-     xlab = "Proportion of patients \n reporting to closest ARMC",
-     ylab = "Number of districts", col = "grey50", border = "white", xlim = c(0, 1))
-
-exp_mat %>% 
-  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
-  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
-            by = c("district" = "mdg_dis_co")) %>%
-  group_by(ctar_obs) %>% # group by ctar
-  mutate(prop = n/sum(n),
-         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
-  filter(catch_match == 1) %>%
-  summarize(prop = sum(prop)) -> catch_ctar
-```
-
-```{r figure 1C, fig.cap = "Figure 1C"}
-hist(catch_ctar$prop, main = "", 
-     xlab = "Proportion of patients reporting from within catchment",
-     ylab = "Number of ARMC", col = "grey50", border = "white", xlim = c(0, 1))
-```
-
 As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2).
@@
 To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
-
-```{r clinic level reporting, fig.cap = "Figure 2A"}
-######### Patient time series
-ctar_data %>%
-  select(id_ctar, date_de_consultation) %>%
-  drop_na(id_ctar) %>%
-  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
-  gather(dose, date_de_consultation, -id_ctar) %>%
-  group_by(id_ctar, date_de_consultation) %>%
-  summarise(n = n()) -> patient_ts
-
-patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
-patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
-patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]
-
-start_date <- ymd("2014-01-01")
-end_date <- ymd("2017-12-31")
-
-## getting reporting
-ts <- as_tibble(seq(start_date, end_date, by = "day"))
-
-patient_ts %>%
-  right_join(ts, by = c("date_de_consultation" = "value")) %>%
-  select(id_ctar, date_de_consultation, n) %>%
-  spread(id_ctar, n) %>%
-  replace(., is.na(.), 0) -> doses_wide
-dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])
-
-get.days <- function(dmat = dose_mat, threshold = 10) {
-  date_mat <- matrix(NA, nrow(dmat), ncol(dmat))
-  for (j in 1:ncol(date_mat)){
-    rle(dmat[ , j]) %>%
-      unclass() %>%
-      as.data.frame() %>%
-      mutate(end = cumsum(lengths),
-             start = c(1, lag(end)[-1] + 1)) %>%
-      filter(values == 0, lengths >= threshold) -> rles
-  if(nrow(rles) > 0){
-    for (i in 1:nrow(rles)){
-        date_mat[rles$start[i]:rles$end[i], j] <- 0
-      }
-    }
-  }
-
-  date_mat <- replace(date_mat, is.na(date_mat), 1)
-  date_mat %>% 
-    as_tibble() %>%
-    group_by(date = year(doses_wide$date_de_consultation)) %>%
-    summarise_all(funs(sum(.)/n())) -> clinic_reporting
-  colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
-
-  clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
-  clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
-                                                  ctar_metadata$id_ctar)]
-  clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
-  clinic_reporting$threshold <- threshold
-  return(list(clinic_reporting, date_mat))
-}
-
-date_mat <- as.data.frame(get.days(threshold = 15)[[2]])
-date_mat$date_de_consultation <- doses_wide$date_de_consultation
-date_mat <-melt(date_mat, id = "date_de_consultation")
-levels(date_mat$variable) <- names(doses_wide)[2:23]
-names(date_mat) <- c("date_de_consultation", "id_ctar", "exclude")
-
-patient_ts %>%
-  ungroup() %>%
-  mutate(id_ctar = as.factor(id_ctar)) %>%
-  right_join(date_mat) %>%
-  mutate(n = replace_na(n, 0), 
-         n = ifelse(exclude == 1, n, NA)) -> doses_long
-doses_long$ctar <- ctar_metadata$CTAR[match(doses_long$id_ctar, ctar_metadata$id_ctar)]
-
-ggplot(data = doses_long, aes(date_de_consultation, ctar)) + 
-   geom_tile(aes(fill = n)) +
-   scale_fill_gradientn(colours = c("white", "purple", "black"), values = c(0, 0.1, 1),
-                        na.value = alpha("lightgrey", 1)) +
-   xlim(start_date, end_date)
-
-clinic_reporting <- get.days(threshold = 15)[[1]]
-
-```
-
-```{r figure 1B, fig.cap = "Figure 1B"}
-clinic_reporting %>%
-  mutate(prop15 = prop, prop5 = get.days(threshold = 5)[[1]]$prop,
-         prop30 = get.days(threshold = 30)[[1]]$prop) %>%
-  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
-  geom_hline(yintercept = 0.25, color = "darkgrey") +
-  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
-               stat = "identity") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
-  xlab("") +
-  ylab("Proportion of days included") +
-  coord_flip() +
-  labs(tag = "B")
-
-```
-
-```{r figure 3B, fig.cap= "Figure 3B"}
-patient_ts %>%
-  group_by(ctar) %>%
-  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts
-
-ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
-  scale_color_discrete(name = "Exclude", labels = c("Yes", "No")) +
-  ylab("Number of patients per day") +
-  geom_jitter(alpha = 0.5, width = 0.25) +
-  coord_flip()
-
-to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
-                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
-to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
-ctar_data$exclude_bycomment <- 0
-ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
-ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
-patient_ts %>%
-  ungroup() %>%
-  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
-  right_join(ctar_data) -> ctar_check
-
-```
-
-```{r trying to exclude contacts, fig.cap="Figure 3A"}
-## Testing with Moramanga data 
-sdev <- seq(2, 10, by = 1)
-check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
-for (i in 1:length(sdev)){
-  ctar %>%
-    group_by(date.reported) %>%
-    summarize(n = n()) %>%
-    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
-  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
-  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
-  check <- rbind(check, c(sdev[i], checksd[1:2]))
-}
-
-plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "n", ylim = c(0, 1), 
-     ylab = "Proportion excluded", xlab = "Number of standard deviations")
-points(check$sd, check$prop_bites, col = "red", pch = 20)
-legend("topright", c("Contacts", "Bites"), col = c("Blue", "Red"), pch = 20, bty = "n")
-```
 
@@
 <br>
-
-```{r district level exposures, fig.cap = "Figure 4"}
-## districts that are non-reporter and data not available yet CTAR
-mada_district@data %>%
-select(district = mdg_dis_co, dist_name = district, pop = pop2015adj, ttimes_weighted_dist = ttimes_weighted, 
-       dist_catch = CTAR, exclude, color) -> covars
-
-ctar_check %>% 
-  filter(exclude_bydate == 0, exclude_bycomment == 0) %>% # exclude contacts
-  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
-  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
-  summarize(n = n()) %>% # get count of exposures
-  left_join(covars) %>% # add pop + ttimes + reporting
-  left_join(clinic_reporting) %>%   # add reporting by clinic + year
-  filter(exclude == 0, prop > 0.25) %>% ## filter out excluded catchments and ones with < 25% reporting
-  group_by(district, year) %>%
-  mutate(bites = n/prop) %>%
-  summarise(bites = sum(bites, na.rm = TRUE)) %>%
-  left_join(covars) -> exps
-
-col_vals <- dist_catchments$color
-names(col_vals) <- dist_catchments$mdg_dis_co
-
-ggplot(exps, aes(x = reorder(dist_name, ttimes_weighted_dist), y = bites/pop*1e5, color = district)) +
-  geom_boxplot() +
-  theme(legend.position = "none") +
-  scale_color_manual(values = col_vals) + 
-  ylab("Annual bites per 100k") +
-  xlab("") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
-
-exps %>%
-  group_by(district) %>%
-  summarise(bites = mean(bites), n = n()) %>%
-  left_join(covars) -> exps_dist
-```
 
@@
 ### Models of bites as a function of travel times
-```{r likelihood of observing moramanga data @ commune level, fig.cap = "Figure 5A"}
-library(boot)
-
-get.likelihood.bites <- function(bites, covar, pop, params) {
-  beta <- params[1]
-  intercept <- params[2]
-  exp_bites <- inv.logit(beta*covar + intercept)*pop
-  return(-sum(dpois(round(bites), lambda = exp_bites, log = TRUE)))
-}
-
-get.likelihood.mora <- function(commune, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      mutate(bites_mu = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
-  return(-sum(dpois(round(sum_bites$), lambda = sum_bites$bites_mu, log = TRUE)))
-}
-
-get.likelihood.mada <- function(commune, district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-get.likelihood.dist <- function(district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  district %>%
-      filter(exclude == 0) %>%
-      mutate(exp_bites = inv.logit(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
-  return(-sum(dpois(round(district$exp_bites), lambda = round(district$bites), log = TRUE)))
-}
-
-### Moramanga model ###
-mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
-B_ttimes_mora <- mora_pars$par[1]
-B_0_mora <- mora_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(mora_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(mora_pars$hessian)
-
-ntest = 1000; testpar = mora_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, mora_pars$par, varcov)
-}
-
-B_ttimes_mora_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_mora_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-### Madagascar model ###
-mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mada, district = exps_dist,
-             commune = comm_data, hessian = TRUE)
-B_ttimes_mada <- mada_pars$par[1]
-B_0_mada <- mada_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(mada_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(mada_pars$hessian)
-
-ntest = 1000; testpar = mada_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, mada_pars$par, varcov)
-}
-
-B_ttimes_mada_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-### District parameters
-dist_pars <- optim(par = c(1e-6, 0.1), get.likelihood.dist, district = exps_dist,
-              hessian = TRUE)
-B_ttimes_dist <- dist_pars$par[1]
-B_0_dist <- dist_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(dist_pars$hessian)
-
-ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
-}
-
-B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
+
+```{r models}
+model.bites <- function(bites, names_bites, covar, pop, names_covar, sum = TRUE, 
+                        pop_predict = "addPop", covar_name = "ttimes_weighted",
+                        beta, beta_pop, intercept, trans, run = "optim",...)
+
+## Exclude for comm data!
+comm_covars <- filter(comm_data, exclude == 0)
+
+pop_predict <- c("addPop", "onlyPop", "none")
+pop_vars <- list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop)
+covars <- list(distance_district = exps_dist$distance, distance_commune = comm_covars$distance, 
+               ttimes_district = exps_dist$ttimes_weighted_dist/60, ttimes_commune = comm_covars$ttimes_weighted/60)
+sum_it <- c(FALSE, TRUE, FALSE, TRUE)
+run_type <- "optim"
+
+mods_mada <- 
+  foreach(i = 1:length(pop_predict), .combine = c) %:%
+  foreach(j = 1:length(sum_it), .combine = c) %:%
+  foreach(k = 1:length(covars), .combine = c) %do% {
+  i = 1; j = 1; k = 1;
+
+  if(sum_it[j] == TRUE) {
+    names_covar <- comm_covars$mdg_dis_co
+  } else {
+    names_covar = ""
+  }
+  data_est <- list(bites = exps_dist$bites, run = run_type, covar = covars[[k]],
+                   names_bites = exps_dist$district,
+                   covar_name = names(covars)[k], names_covar = names_covar,
+                   trans = 1e5)
+  start_params <- list(beta = 1e-5, intercept = 0.1)
+
+  if(pop_predict[i] == "addPop") {
+    ## need beta_pop
+    start_params <- c(start_params, beta_pop = 1e-5)
+  }
+
+  if(grepl("commune", names(covars)[k], fixed = TRUE)) {
+    data_est <- c(data_est, list(pop = pop_vars$pop_commune))
+  } else {
+    data_est <- c(data_est, list(pop = pop_vars$pop_dist))
+  }
+
+  check <- mle2(model.bites, start = start_params, data = data_est)
+
+  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
+  check
+}
+
+mods_moramanga
+mods_pop_mada_moramanga
+
+## all mods = combine
+
+prof_CI <- lapply(mods, confint)
+
+mada_mod_predicts <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods), 
+     intercept = mapply(function(x) coef(x)["intercept"], mods), 
+     beta_upper = mapply(function(x) x["beta", 2], prof_CI),
+     beta_lower = mapply(function(x) x["beta", 1], prof_CI),
+     intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
+     intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
+     likelihood = mapply(function(x) logLik(x), mods),
+     model = mapply(function(x) attributes(x)$name, mods)))
+
+mada_mod_results <- as.data.frame(list(params = c("beta", "intercept"), 
+                   estimates = t(mapply(function(x) coef(x), mods)), 
+                   upper = t(mapply(function(x) x[, 2], prof_CI)),
+                   lower = t(mapply(function(x) x[, 1], prof_CI)), 
+                   data = "Mada"))
+
+## Predictions
+library(foreach)
+
+mods <- 
+  foreach(l = 1:nrow(mada_), .combine = c) %do% {
+      if(pop_predict[k] == TRUE) covar_name = "pop" else covar_name = names(covar_mada)[l]
+      if (scale[i] == "Mada") {
+        if(sum[j] == TRUE) {
+          sum_name <- "summed"
+          check <- predict.bites(bites = exps_dist$bites, 
+                           names_bites = exps_dist$district, 
+                           pop = exps_dist$pop,
+                           pop_sub = comm_data$pop, 
+                           names_pop = comm_data$mdg_dis_co, 
+                           names_covar = comm_data$mdg_dis_co, trans = 1e5,
+                           covar_sub = covar_mada_sub[[l]],
+                           sum = sum[j], pop_predict = pop_predict[k])
+        } else {
+          sum_name <- ""
+          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
+               data = c(data_mada, list(covar = covar_mada[[l]]),
+                           sum = sum[j], pop_predict = pop_predict[k]))
+        }
+      }   
+      if(scale[i] == "Mora") {
+        if(sum[j] == TRUE) {
+          check <- NULL
+        } else {
+          sum_name <- ""
+          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
+             data = c(data_mora, list(covar = covar_mora[[l]]),
+                           sum = sum[j], pop_predict = pop_predict[k]))
+        }
+      }
+  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
+  check
+}
+
+ttimes_plot <- seq(0, 720, by = 30)
+pop_plot <- seq(100, 1e6, length.out = length(ttimes_plot))
+expect <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
+  check <- mada_mod_predicts$expect_fun[[i]]
+  exp <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
+               covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, 
+               trans = 1e5)
+  exp_upper <- check(beta = mada_mod_predicts$beta_upper[i], 
+                     intercept = mada_mod_predicts$intercept_upper[i],
+                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
+  exp_lower <- check(beta = mada_mod_predicts$beta_lower[i], 
+                     intercept = mada_mod_predicts$intercept_lower[i],
+                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
+  as.data.frame(list(ttimes = ttimes_plot, 
+                     pop = pop_plot, exp = exp, exp_upper = exp_upper, exp_lower = exp_lower, 
+                     model = rownames(mada_mod_predicts)[i], data = mada_mod_predicts$data[i]))
+}
+
+ggplot(data = preds, aes(x = observed, y = predicted, color = model)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey")
+ggplot(data = filter(expect, !grepl("pop", model, fixed = TRUE)), 
+       aes(x = ttimes, y = exp, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = exp_lower, ymax = exp_upper, fill = model), alpha = 0.5)
+
+AICctab(mada_ttimes_fit, mada_distance_fit, mada_pop_fit, mada_sumttimes_fit, mada_sumpop_fit,
+        mada_sumdist_fit, nobs = nrow(exps_dist))
 
@@
                                   B_ttimes_dist_CI[1])))
-
-preds <- foreach(i = 1:nrow(par_df), .combine = rbind) %do% {
-  bites <- inv.logit(par_df$B_ttimes[i]*ttimes_plot + par_df$B_0[i])*pop_plot
-  upper <- inv.logit(par_df$B_ttimes_upper[i]*ttimes_plot + par_df$B_0_upper[i])*pop_plot
-  lower <- inv.logit(par_df$B_ttimes_lower[i]*ttimes_plot + par_df$B_0_lower[i])*pop_plot
-  return(as.data.frame(list(bites = bites, upper = upper, lower = lower, ttimes = ttimes_plot/60, 
-                            model = par_df$model[i])))
-}
-ggplot(data = preds, aes(x = ttimes, y = bites, color = model)) +
-  geom_line() +
-  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.5) +
-
-plot(ttimes_plot/60, inv.logit(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot, type = "l", bty = "l",
-     ylab = "Mean bites per 100k", xlab = "Travel times (hrs)", col = "darkblue")
-points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[1]*ttimes_plot + B_0_mada_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-points(ttimes_plot/60, inv.logit(B_ttimes_mada_CI[2]*ttimes_plot + B_0_mada_CI[2])*pop_plot, type = "l", 
-       lty = 2, col = "lightblue")
-
-points(ttimes_plot/60, inv.logit(B_ttimes_mora*ttimes_plot + B_0_mora)*pop_plot, type = "l", col = "red")
-points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[1]*ttimes_plot + B_0_mora_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "pink")
-points(ttimes_plot/60, inv.logit(B_ttimes_mora_CI[2]*ttimes_plot + B_0_mora_CI[2])*pop_plot, type = "l",
-       lty = 2, col = "pink")
-
-points(ttimes_plot/60, inv.logit(B_ttimes_dist*ttimes_plot + B_0_dist)*pop_plot, type = "l", col = "purple")
-points(ttimes_plot/60, inv(B_ttimes_dist_CI[1]*ttimes_plot + B_0_dist_CI[1])*pop_plot, type = "l", 
-       lty = 2, col = "lavender")
-points(ttimes_plot/60, exp(B_ttimes_dist_CI[2]*ttimes_plot + B_0_dist_CI[2])*pop_plot, type = "l",
-       lty = 2, col = "lavender")
-
-legend("topright", c("Mada communes", "Mada districts", "Moramanga communes"), 
-       col = c("darkblue", "purple", "red"), 
-       lty = c(1, 1, 1), 
-       bty = "n")
 ```
@@
 district_coords <- coordinates(mada_district)
-library(geosphere)
 distance_mat <- distm(district_coords, ctar_coords)
@@
 
-## Supplementary Figures
-
-#### S1
-#### S2
-#### S3
-#### S4
-#### S5
-#### S6
+## Supplementary Materials
+### S1. RLE thresholds for reporting
+### S2. Removing contacts from the data set
+### S3. Comparing models of reported bites
+### S4. Sensitivity of model parameter estimates to contact and RLE cut-offs
+### S5. Sensitivity of burden estimates to paramater values (multivariate for R, p_rabid, rho_max)
+### S6. Sensitity of scenario analyses to parameter values (multivariate for R, p_rabid, rho_max)
+
+**Remember that the only other variable is p_death which should just shift these estimates up or down**

### 2019-03-14 accessPEP.Rmd
model comparison (corrected bug in district model)
+ sensitivity analyses etc.

@@
 source("R/data_functions.R")
+source("R/get.burden.R")
 select <- dplyr::select
@@
 Often, the populations at greatest risk of a given disease are the most underserved.
-Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful considerations of how best to optimize the allocation of limited resources.
+Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to optimize the allocation of limited resources.
 
@@
 However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*).
-GAVI bit here...potential to greatly reduce disease burden.
-
 Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.
@@
 the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.
-While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.
+At the sub-national level where access to PEP is highly heterogenous, these assumptions likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.
+
+In 2018, Gavi added PEP to their vaccine investment portfolio, and soon countries will be able to apply for funding to expand access to PEP.
+Expanding access strategically by identifying places with limited access...
 
@@
 Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
-To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
+To get spatial estimates of disease burden in this context, we assume that bite incidence reflects access to PEP rather than than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
 Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
@@
 Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis.
-Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A).
+Overall, we had data from *`r 31 - length(no_data)`* clinics across the country (Fig 1A).
 These data include details of the administrative district of the bite patient's address and the date of reporting.
-We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).
+We also had *27* months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).
 
@@
 We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4).
-For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
-We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.
+For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])**100, 2)"* of known contacts.
+We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)**100, 2)*% of patient records from the national data.
 We also compared this method to assuming that 20% of total bites at the district level were contacts.
@@
 
-After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
+After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from *r length(unique(exps_dist$dist_catch))* catchments (Figure 4).
 
@@
 
-```{r models}
-model.bites <- function(bites, names_bites, covar, pop, names_covar, sum = TRUE, 
-                        pop_predict = "addPop", covar_name = "ttimes_weighted",
-                        beta, beta_pop, intercept, trans, run = "optim",...)
-
+```{r models ttimes, cache = TRUE}
 ## Exclude for comm data!
@@
 
-pop_predict <- c("addPop", "onlyPop", "none")
-pop_vars <- list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop)
-covars <- list(distance_district = exps_dist$distance, distance_commune = comm_covars$distance, 
-               ttimes_district = exps_dist$ttimes_weighted_dist/60, ttimes_commune = comm_covars$ttimes_weighted/60)
-sum_it <- c(FALSE, TRUE, FALSE, TRUE)
-run_type <- "optim"
-
-mods_mada <- 
-  foreach(i = 1:length(pop_predict), .combine = c) %:%
-  foreach(j = 1:length(sum_it), .combine = c) %:%
-  foreach(k = 1:length(covars), .combine = c) %do% {
-  i = 1; j = 1; k = 1;
-
-  if(sum_it[j] == TRUE) {
-    names_covar <- comm_covars$mdg_dis_co
-  } else {
-    names_covar = ""
-  }
-  data_est <- list(bites = exps_dist$bites, run = run_type, covar = covars[[k]],
-                   names_bites = exps_dist$district,
-                   covar_name = names(covars)[k], names_covar = names_covar,
-                   trans = 1e5)
-  start_params <- list(beta = 1e-5, intercept = 0.1)
-
-  if(pop_predict[i] == "addPop") {
-    ## need beta_pop
-    start_params <- c(start_params, beta_pop = 1e-5)
-  }
-
-  if(grepl("commune", names(covars)[k], fixed = TRUE)) {
-    data_est <- c(data_est, list(pop = pop_vars$pop_commune))
-  } else {
-    data_est <- c(data_est, list(pop = pop_vars$pop_dist))
-  }
-
-  check <- mle2(model.bites, start = start_params, data = data_est)
-
-  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
-  check
-}
-
-mods_moramanga
-mods_pop_mada_moramanga
-
-## all mods = combine
-
-prof_CI <- lapply(mods, confint)
-
-mada_mod_predicts <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods), 
-     intercept = mapply(function(x) coef(x)["intercept"], mods), 
+### Running estimation
+mada_mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                             pop_predict = c("flatPop"),
+                             pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
+                             covars = list(ttimes_district = exps_dist$ttimes_weighted_dist/60,
+                                           ttimes_commune = comm_covars$ttimes_weighted/60), 
+                             sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
+mora_mods_covars <- run.mods(bites = exps_mora$bites, names_bites = exps_mora$commune,
+                             pop_predict = c("flatPop"),
+                             pop_vars = list(pop_dist = exps_mora$pop),
+                             covars = list(ttimes_mora = exps_mora$ttimes_weighted/60), 
+                             sum_it = c(FALSE, FALSE), run_type = "optim", label = "Mora")
+mada_mods <- c(mada_mods_covars, mora_mods_covars)
+
+## profile CIs
+prof_CI <- lapply(mada_mods, confint)
+bite_mods <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mada_mods),
+                                intercept = mapply(function(x) coef(x)["intercept"], mada_mods),
                                 beta_upper = mapply(function(x) x["beta", 2], prof_CI),
                                 beta_lower = mapply(function(x) x["beta", 1], prof_CI),
+                                intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
                                 intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
-     intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
-     likelihood = mapply(function(x) logLik(x), mods),
-     model = mapply(function(x) attributes(x)$name, mods)))
-
-mada_mod_results <- as.data.frame(list(params = c("beta", "intercept"), 
-                   estimates = t(mapply(function(x) coef(x), mods)), 
-                   upper = t(mapply(function(x) x[, 2], prof_CI)),
-                   lower = t(mapply(function(x) x[, 1], prof_CI)), 
-                   data = "Mada"))
-
-## Predictions
-library(foreach)
-
-mods <- 
-  foreach(l = 1:nrow(mada_), .combine = c) %do% {
-      if(pop_predict[k] == TRUE) covar_name = "pop" else covar_name = names(covar_mada)[l]
-      if (scale[i] == "Mada") {
-        if(sum[j] == TRUE) {
-          sum_name <- "summed"
-          check <- predict.bites(bites = exps_dist$bites, 
-                           names_bites = exps_dist$district, 
-                           pop = exps_dist$pop,
-                           pop_sub = comm_data$pop, 
-                           names_pop = comm_data$mdg_dis_co, 
-                           names_covar = comm_data$mdg_dis_co, trans = 1e5,
-                           covar_sub = covar_mada_sub[[l]],
-                           sum = sum[j], pop_predict = pop_predict[k])
-        } else {
-          sum_name <- ""
-          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
-               data = c(data_mada, list(covar = covar_mada[[l]]),
-                           sum = sum[j], pop_predict = pop_predict[k]))
-        }
-      }   
-      if(scale[i] == "Mora") {
-        if(sum[j] == TRUE) {
-          check <- NULL
-        } else {
-          sum_name <- ""
-          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
-             data = c(data_mora, list(covar = covar_mora[[l]]),
-                           sum = sum[j], pop_predict = pop_predict[k]))
-        }
-      }
-  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
-  check
-}
-
-ttimes_plot <- seq(0, 720, by = 30)
-pop_plot <- seq(100, 1e6, length.out = length(ttimes_plot))
-expect <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
-  check <- mada_mod_predicts$expect_fun[[i]]
-  exp <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
-               covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, 
-               trans = 1e5)
-  exp_upper <- check(beta = mada_mod_predicts$beta_upper[i], 
-                     intercept = mada_mod_predicts$intercept_upper[i],
-                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
-  exp_lower <- check(beta = mada_mod_predicts$beta_lower[i], 
-                     intercept = mada_mod_predicts$intercept_lower[i],
-                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
-  as.data.frame(list(ttimes = ttimes_plot, 
-                     pop = pop_plot, exp = exp, exp_upper = exp_upper, exp_lower = exp_lower, 
-                     model = rownames(mada_mod_predicts)[i], data = mada_mod_predicts$data[i]))
-}
-
-ggplot(data = preds, aes(x = observed, y = predicted, color = model)) +
-  geom_point() +
-  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey")
-ggplot(data = filter(expect, !grepl("pop", model, fixed = TRUE)), 
-       aes(x = ttimes, y = exp, color = model)) +
+                                likelihood = mapply(function(x) logLik(x), mada_mods),
+                                model = mapply(function(x) attributes(x)$label, mada_mods)))
+## Predictions: fixed
+ttimes_plot <- seq(0, 25, by = 0.5)
+distance_plot <- seq(0, 200, by = 2.5)
+pop_plot <- seq(1000, 1e6, length.out = 30)
+
+library(iterators)
+preds_fixed <- run.predicts.fixed(covar_df = bite_mods, pop_fixed = 1e5, pop_seq = pop_plot)
+preds_data <- run.predicts.data(covar_df = bite_mods, run_type = "predict.data")
+preds_raw <- run.predicts.data(covar_df = bite_mods, run_type = "predict.raw")
+
+ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
   geom_line() + 
-  geom_ribbon(aes(ymin = exp_lower, ymax = exp_upper, fill = model), alpha = 0.5)
-
-AICctab(mada_ttimes_fit, mada_distance_fit, mada_pop_fit, mada_sumttimes_fit, mada_sumpop_fit,
-        mada_sumdist_fit, nobs = nrow(exps_dist))
-
-par_df <- data.frame(model = c("Mada communes", "Moramanga communes", "Mada districts", 
-                               "Mada communes", "Moramanga communes", "Mada districts"),
-                     par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
-                             "intercept"),
-                     values = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist, 
-                                B_0_mada, B_0_mora, B_0_dist),
-                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
-                                  B_ttimes_dist_CI[2],
-                                  B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
-                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
-                                  B_ttimes_dist_CI[1], 
-                                  B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]))
-kable(par_df, caption = "Table 1: summary of parameter estimates")
-ttimes_plot <- seq(0, 15*60, by = 30)
-pop_plot <- 1e5
-
-## Mada
-par_df <- as.data.frame(list(model = c("Mada communes", "Moramanga communes", "Mada districts"),
-                     B_ttimes = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist),
-                     B_0 = c(B_0_mada, B_0_mora, B_0_dist),
-                     B_0_upper = c(B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
-                     B_0_lower = c(B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]),
-                     B_ttimes_upper = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
-                                  B_ttimes_dist_CI[2]),
-                     B_ttimes_lower = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
-                                  B_ttimes_dist_CI[1])))
-```
-
-```{r model table, results = "show"}
-kable(par_df, caption = "Table 1: summary of parameter estimates")
-```
-
-```{r figure 5B, fig.cap="Figure 5B"}
-## Data vs. predicted
-comm_data %>%
-      filter(exclude == 0) %>%
-      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
-      group_by(mdg_dis_co) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
-                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
-                   xlim = c(0, 5000))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-
-exps_dist %>%
-      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
-points(check$bites, check$bites_mu, col = alpha("purple", 0.5), pch = 20)
-
-mora_bites %>%
-      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
-points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))
-legend("topright", inset = c(-0.1, -0.2),
-       c("Mada communes", "Mada districts", "Moramanga communes"), 
-       col = c(alpha("darkblue", 0.5), alpha("purple", 0.5), alpha("red", 0.5)), 
-       pch = 20, 
-       bty = "n", xpd = TRUE)
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
+  scale_color_manual(values = c("navy", "blue", "purple")) +
+  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
+  xlab("Travel time (hrs)") +
+  ylab("Reported bites per \n 100,000 persons")
+
+ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
+  expand_limits(y = c(0, 4000)) +
+  scale_color_manual(values = c("navy", "blue", "purple"))
 
@@
 
-#### Model validation
-```{r model validation}
-## Models
-get.likelihood.other <- function(weight, bites, pop, params) {
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  exp_bites <- inv.logit(B_ttimes*weight + B_0)*pop
-  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
-}
-
-## Distance to CTAR
-ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
-district_coords <- coordinates(mada_district)
-distance_mat <- distm(district_coords, ctar_coords)
-dist_min <- apply(distance_mat, 1, min)/1000
-exps_dist$dist_min <- dist_min[match(exps_dist$district, mada_district$mdg_dis_co)]
-
-## Unweighted travel times
-exps_dist$ttimes_unweighted <- mada_district$study_area[match(exps_dist$district,
-                                                              mada_district$mdg_dis_co)]
-
-### District parameters
-dist_pars <- optim(par = c(1e-5, -0.1), get.likelihood.other, 
-                   weight = exps_dist$dist_min, bites = exps_dist$bites, pop = exps_dist$pop,
-                   hessian = TRUE)
-B_ttimes_dist <- dist_pars$par[1]
-B_0_dist <- dist_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(dist_pars$hessian)
-
-ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
-}
-
-B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-plot(exps_dist$bites, inv.logit(B_ttimes_dist*exps_dist$dist_min+ B_0_dist)*exps_dist$pop)
-plot(exps_dist$ttimes_weighted_dist, exps_dist$bites)
-
-### Unweighted
-dist_pars <- optim(par = c(1e-3, 0.1), get.likelihood.other, 
-                   weight = exps_dist$ttimes_unweighted, bites = exps_dist$bites, 
-                   pop = exps_dist$pop,
-                   hessian = TRUE)
-B_ttimes_dist <- dist_pars$par[1]
-B_0_dist <- dist_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(dist_pars$hessian)
-
-ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
-}
-
-B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-plot(exps_dist$bites, exp(B_ttimes_dist*exps_dist$ttimes_unweighted+ B_0_dist))
-plot(exps_dist$ttimes_weighted_dist, exps_dist$dist_min)
-
-### Pop
-get.likelihood.pop <- function(bites, pop, params) {
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  exp_bites <- exp(B_ttimes*pop/1e5 + B_0)
-  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
-}
-pop_pars <- optim(par = c(1e-5, 0.1), get.likelihood.pop, 
-                   pop = exps_dist$pop, bites = exps_dist$bites, 
-                   hessian = TRUE)
-B_pop <- pop_pars$par[1]
-B_0_pop <- pop_pars$par[2]
-check <- glm(round(bites) ~ pop, data = exps_dist, family = "poisson")
-plot(exps_dist$bites, exp(B_pop*exps_dist$pop/1e5 + B_0_pop), pch = 20, col = alpha("red", 0.5))
-plot(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(pop_pars$hessian)
-
-ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
-}
-
-B_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-## @ commune level
-get.likelihood.mada <- function(commune, district, params) {
-  # To test likelihood function
-  # commune = comm_data
-  # district = exps_dist
-  # params = c(1e-6, 0.1)
-  B_ttimes <- params[1]
-  B_0 <- params[2]
-  commune %>%
-      filter(exclude == 0) %>%
-      group_by(mdg_dis_co) %>%
-      mutate(bites = exp(B_ttimes*pop/1e5 + B_0)) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(district, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
-}
-comm_pars <- optim(par = c(1e-3, 0.1), get.likelihood.mada, 
-                   commune = comm_data, district = exps_dist, 
-                   hessian = TRUE)
-B_ttimes_comm <- comm_pars$par[1]
-B_0_comm <- comm_pars$par[2]
-
-## Getting standard errors
-# this is how you can get standard errors from optim, but inverting the hessian
-se <- sqrt(abs(diag(solve(comm_pars$hessian)))); se
-
-#simulate "standard errors"
-varcov <- solve(comm_pars$hessian)
-
-ntest = 1000; testpar = comm_pars$par; sim.par = matrix(NA, ntest, 2)
-for (k in 1:ntest) {
-  sim.par[k, ] = mvrnorm(1, comm_pars$par, varcov)
-}
-
-B_ttimes_comm_CI <- quantile(sim.par[,1],c(0.025,0.975))
-B_0_comm_CI <- quantile(sim.par[,2],c(0.025,0.975))
-
-comm_data %>%
-      filter(exclude == 0) %>%
-      mutate(bites = exp(B_ttimes_comm*pop/1e5 + B_0_comm)) %>%
-      group_by(mdg_dis_co) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
-                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
-                   xlim = c(0, 5000))
-comm_data %>%
-      filter(exclude == 0) %>%
-      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
-      group_by(mdg_dis_co) %>%
-      summarize(sum_bites = sum(bites)) %>%
-      left_join(select(exps_dist, district, dist_bites = bites), 
-                by = c("mdg_dis_co" = "district")) %>%
-      filter(!is.na(dist_bites)) -> sum_bites
-sum_bites %$% plot(dist_bites, sum_bites, pch = 20, col = alpha("red", 0.5))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-
-exps_dist %>%
-      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check1
-points(check1$bites, check1$bites_mu, col = alpha("purple", 0.5), pch = 20)
-points(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-
-mora_bites %>%
-      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
-points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("blue", 0.5))
-check <- glm(round(bites) ~ pop, data = mora_bites, family = "poisson")
-plot(mora_bites$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
-abline(a = 0, b = 1, col = "grey", lty = 2)
-
-## Correlation plot
-
-## Table of fits/parameter results
-
-## Impact of contacts
-
-## Impact of RLE threshold
-
-```
-
 ### Estimation of burden and reporting
-
-```{r reporting by incidence, fig.cap="Figure 6"}
-## Mada
-bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
-R_100k_min <- pop_plot/25*0.01*0.39
-R_100k_max <- pop_plot/5*0.01*0.39
-
-plot(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-     type = "l", bty = "n",
-     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-     col = "darkred")
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
-     col = "red", lty = 2)
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
-     col = "red", lty = 2)
-polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
-        border = NA, col = alpha("red", 0.2))
-polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
-        border = NA, col = alpha("red", 0.6))
-legend("topright", inset = c(-0.10, 0), c("Max rabies exposures", "Min rabies exposures"), lty = c(2, 1), 
-       col = c("red", "darkred"),
-       bty = "n", xpd = TRUE)
-legend("topright", inset = c(0.01, 0.2), c("p_rabid = 0.2", "p_rabid = 0.6"), border = NA, 
-       fill = c(alpha("red", 0.2), alpha("red", 0.6)), bty = "n", xpd = TRUE)
-abline(v = max(ttimes_plot[bites_100k*0.6 > R_100k_min])/60, col = "grey", lty = 3)
-abline(v = max(ttimes_plot[bites_100k*0.2 > R_100k_max])/60, col = "grey", lty = 3)
-```
-
-```{r scale example}
-## Comparing scales
-comm_data$reporting_min <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
-                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
-                                            hdr = 5, p_rabid = 0.2)
-
-comm_data$reporting_max <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
-                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
-                                            hdr = 25, p_rabid = 0.6)
-comm_data %>%
-  group_by(district) %>%
-  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
-  summarize(reporting_min = sum(reporting_min*prop_pop), 
-            reporting_max = sum(reporting_max*prop_pop)) %>%
-  mutate(reporting_med = (reporting_min + reporting_max)/2,
-         type = "summed from commune") -> summed_rep_dist
-
-mada_district@data %>%
-  mutate(reporting_min = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
-                                              p_rabid = 0.2), 
-         reporting_max = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
-                                              p_rabid = 0.6),
-         reporting_med = (reporting_min + reporting_max)/2,
-         type = "district level") %>%
-  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp
-
-rep_comp <- rbind(rep_comp, summed_rep_dist)
-rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]
-
-ggplot(rep_comp, aes(x = reporting_med, y = reorder(district, ttime), 
-                              color = type)) +
-  ggstance::geom_pointrangeh(aes(xmin = reporting_min, xmax = reporting_max), fatten = 2)
-
-bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
-R_100k_min <- pop_plot/25*0.01*0.39
-R_100k_max <- pop_plot/5*0.01*0.39
-
-plot(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-     type = "l", bty = "n",
-     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-     col = "darkred")
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
-     col = "red", lty = 2)
-lines(ttimes_plot/60, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
-     col = "red", lty = 2)
-polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
-        border = NA, col = alpha("red", 0.2))
-polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
-        border = NA, col = alpha("red", 0.6))
-
-## Mora
-comm_data %>%
-  filter(district == "Moramanga") -> comm_level
-bites_100k <- exp(B_ttimes_mada*comm_level$ttimes_weighted + B_0_mada)*comm_level$pop
-R_100k_min <- comm_level$pop/25*0.01*0.39
-R_100k_max <- comm_level$pop/5*0.01*0.39
-
-## district val for Moramanga
-## need to merge names
-exps_dist %>%
-  filter(dist_name == "Moramanga") %>%
-  mutate(rep_max = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
-                                              p_rabid = 0.6),
-         rep_min = get.reporting_burden(pop = pop, ttimes = ttimes_weighted_dist, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
-                                              p_rabid = 0.2)) -> dist_level
-dist_level %$% points(ttimes_weighted_dist/60, rep_max, col = "red", pch = 20)
-dist_level %$% points(ttimes_weighted_dist/60, rep_min, col = "red", pch = 20)
-dist_level %$% abline(v = ttimes_weighted_dist/60, col = "red", lty = 2)
-
-```
 
@@
 comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
+B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]
+
+B_ttimes_dist <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_district"]
+B_0_dist <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_district"]
 
@@
 
-## Estimate burden
-source("R/get.burden.R")
 system.time({
-  results_pr20_comm <- get.burden(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
-                           pop = comm_mat$pop, scenario = comm_mat$scenario,
-                           param_ttimes = B_ttimes_mada,
-                           param_intercept = B_0_mada,
-                           p_rabid = 0.2, rho_max = 0.98,
-                           max_HDR = 25, min_HDR = 5, 
-                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 1000)
-
-results_pr60_comm <- get.burden(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
-                           pop = comm_mat$pop, scenario = comm_mat$scenario,
-                           param_ttimes = B_ttimes_mada,
-                           param_intercept = B_0_mada,
-                           p_rabid = 0.6, rho_max = 0.98,
-                           max_HDR = 25, min_HDR = 5, 
-                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 1000)
-})
-
-## Filling it in
-results_pr20_comm %>%
-  complete(scenario, names) %>%
-  group_by(names) %>%
-  arrange(scenario) %>%
-  fill(3:ncol(.), .direction = "down") -> results_pr20_comm
-
-results_pr60_comm %>%
-  complete(scenario, names) %>%
-  group_by(names) %>%
-  arrange(scenario) %>%
-  fill(3:ncol(.), .direction = "down") -> results_pr60_comm
-
-system.time({
-results_pr20_dist <- get.burden(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
-                           pop = dist_mat$pop, scenario = dist_mat$scenario,
-                           param_ttimes = B_ttimes_dist,
-                           param_intercept = B_0_dist,
-                           p_rabid = 0.2, rho_max = 0.98,
-                           max_HDR = 25, min_HDR = 5, 
-                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 1000)
-
-results_pr60_dist <- get.burden(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
-                           pop = dist_mat$pop, scenario = dist_mat$scenario,
-                           param_ttimes = B_ttimes_dist,
-                           param_intercept = B_0_dist,
-                           p_rabid = 0.6, rho_max = 0.98,
-                           max_HDR = 25, min_HDR = 5, 
-                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 1000)
-})
-
-results_pr20_dist %>%
-  complete(scenario, names) %>%
-  group_by(names) %>%
-  arrange(scenario) %>%
-  fill(3:ncol(.), .direction = "down") -> results_pr20_dist
-
-results_pr60_dist %>%
-  complete(scenario, names) %>%
-  group_by(names) %>%
-  arrange(scenario) %>%
-  fill(3:ncol(.), .direction = "down") -> results_pr60_dist
-```
-
-``` {r plotting burden} 
-## Group commune results to district level
-results_comm <- rbind(results_pr20_comm, results_pr60_comm)
-colnames(results_comm) <- gsub("results.", "", colnames(results_comm))
-results_comm %>%
-  left_join(select(comm_data, mdg_dis_co, mdg_cm_), 
-            by = c("names" = "mdg_cm_")) %>%
-  group_by(scenario, mdg_dis_co, pr) %>%
-  summarize(names = first(mdg_dis_co), 
-            deaths_mean = sum(deaths_mean, na.rm = TRUE), 
-            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
-            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_comm_summarized
-results_comm_summarized$model <- "commune"
-results_dist <- rbind(results_pr20_dist, results_pr60_dist)
-colnames(results_dist) <- gsub("results.", "", colnames(results_dist))
-results_dist %>%
-  mutate(model = "district") %>%
-  select(scenario, names, pr, deaths_mean, deaths_lowerCI, deaths_upperCI, 
-         model) %>%
-  bind_rows(., results_comm_summarized) -> results_all
-results_all %>%
-  left_join(select(mada_district@data, district, mdg_dis_co, ttimes_weighted),
-            by = c("names" = "mdg_dis_co")) -> results_all
-
-## Plot national level burden
-results_all %>% 
-  group_by(scenario, pr, model) %>% 
-  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
-            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
-            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_natl
-
-## Plotting national level
-ggplot(data = results_natl, aes (x = scenario, y = deaths_mean,
-                                 color = interaction(model, pr))) +
-  geom_line() +
-  geom_ribbon(aes(ymin = deaths_upperCI, ymax = deaths_lowerCI, color = NULL, 
-                  fill = interaction(model, pr)), alpha = 0.5) +
-  scale_color_manual(values = c("district.0.2" = "blue", 
-                                "district.0.6" = "darkblue", 
-                                "commune.0.2" = "magenta", 
-                                "commune.0.6" = "purple"), 
-                     name = "model pars") +
-  scale_fill_manual(values = c("district.0.2" = "blue", 
-                                "district.0.6" = "darkblue", 
-                                "commune.0.2" = "magenta", 
-                                "commune.0.6" = "purple"), 
-                    guide = "none")
-
-## Plotting national level
-ggplot(data = results_all, aes(x = deaths_mean, y = reorder(district, ttimes_weighted), 
-                               color = scenario, shape = as.factor(pr))) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
-                             fatten = 2) +
-  facet_wrap(~model + as.factor(pr))
-
-## Baseline results
-results_all %>% filter(scenario == 0) -> results_all_base
-ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
-                              color = model, shape = as.factor(pr))) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
-  ylab("") +
-  xlab("Average annual deaths") +
-  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
-                                "commune" = alpha("purple", 0.5)), 
-                     name = "Scale") +
-  facet_wrap(~ as.factor(pr)) -> p
-p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
-
-```
-
-```{r doing it from the median, cache=TRUE}
-system.time({
-  results_comm_med <- get.burden_med(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes,
+  results_comm_med <- get.burden.stochastic(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes/60,
                            pop = comm_mat$pop, scenario = comm_mat$scenario,
@@
 
-system.time({
-results_dist_med <- get.burden_med(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes,
+results_dist_med <- get.burden.stochastic(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes/60,
                            pop = dist_mat$pop, scenario = dist_mat$scenario,
@@
                            prob_death = 0.16, nsims = 1000)
-})
 
@@
 
-```{r scenario analysis}
-## new ctar
-expanded <- read.csv("output/incremental_ARMC_nofilter.csv")
-dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
-comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
-
-dist_burden_min <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, p_rabid = 0.6, type = "burden"))
-dist_burden_max <- apply(dist_ttimes, 2, function(x) get.reporting_burden(pop = mada_district$pop2015adj, ttimes = x, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, p_rabid = 0.2, type = "burden"))
-
-comm_burden_min <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 25, p_rabid = 0.6, type = "burden"))
-comm_burden_max <- apply(comm_ttimes, 2, function(x) get.reporting_burden(pop = mada_communes$MDG__201, ttimes = x, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 5, p_rabid = 0.2, type = "burden"))
-
-## Decrease in burden over times proportional
-plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
-     ylab = "Proportion of deaths compared to baseline ")
-lines(colSums(dist_burden_min, na.rm = TRUE)/sum(dist_burden_min[, 1], na.rm = TRUE),
-      col = "purple")
-lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
-     type = "l", col = "blue")
-lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")
-
-## Decrease in burden over times absolute
-plot(colSums(comm_burden_min, na.rm = TRUE), type = "l", col = "blue", bty = "n",
-     ylim = c(0, 1500), xlab = "# Additional ARMC",
-     ylab = "Number of deaths")
-lines(colSums(dist_burden_min, na.rm = TRUE), col = "purple")
-lines(colSums(comm_burden_max, na.rm = TRUE), type = "l", col = "blue")
-lines(colSums(dist_burden_max, na.rm = TRUE), col = "purple")
-
+<br>
+
+## DISCUSSION
+
+### Key findings
+
+### Strengths and Limitations
+
+### Broader context
+
+### Conclusions
+
+## Acknowledgements
+
+## Supplementary Materials
+### S1. RLE thresholds for reporting
+
+### S2. Removing contacts from the data set
+
+### S3. Comparing models of reported bites
+
+```{r models, cache = TRUE}
+## Exclude for comm data!
+comm_covars <- filter(comm_data, exclude == 0)
+
+## Correlation of pop + ttimes + distance for Mada districts, Mada communes (highlight Moramanga by color, and report separate corr metric) (color by excluded vs.
+included data and shape = Moramanga for commune level)
+library(GGally)
+ggpairs(data = select(exps_dist, pop, ttimes_weighted_dist, distance))
+ggpairs(data = select(comm_covars, pop, ttimes_weighted, distance))
+
+ggpairs(data = select(exps_mora, pop, ttimes_weighted, distance))
+
+### Running estimation
+mada_mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                             pop_predict = c("addPop", "flatPop"),
+                             pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
+                             covars = list(distance_district = exps_dist$distance, 
+                                           distance_commune = comm_covars$distance, 
+                                           ttimes_district = exps_dist$ttimes_weighted_dist/60,
+                                           ttimes_commune = comm_covars$ttimes_weighted/60), 
+                             sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
+mada_mods_pop <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                          pop_predict = "onlyPop", 
+                          pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
+                          covars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop), # dummy covars
+                          sum_it = c(FALSE, TRUE), run_type = "optim", label = "Mada")
+mora_mods_covars <- run.mods(bites = exps_mora$bites, names_bites = exps_mora$commune,
+                             pop_predict = c("addPop", "flatPop"),
+                             pop_vars = list(pop_dist = exps_mora$pop),
+                             covars = list(distance_mora = exps_mora$distance, 
+                                           ttimes_mora = exps_mora$ttimes_weighted/60), 
+                             sum_it = c(FALSE, FALSE), run_type = "optim", label = "Mora")
+mora_mods_pop <- run.mods(bites = exps_mora$bites, names_bites = exps_mora$commune,
+                          pop_predict = "onlyPop", 
+                          pop_vars = list(pop_commune = exps_mora$pop),
+                          covars = list(pop_commune = exps_mora$pop), # dummy covar
+                          sum_it = FALSE, run_type = "optim", label = "Mora")
+mada_mods <- c(mada_mods_covars, mada_mods_pop)
+
+kable(AICctab(mada_mods, nobs = nrow(exps_dist), mnames = mapply(function(x) attributes(x)$label, mada_mods), weights = TRUE))
+
+mora_mods <- c(mora_mods_covars, mora_mods_pop)
+kable(AICctab(mora_mods, nobs = nrow(exps_mora), mnames = mapply(function(x) attributes(x)$label, mora_mods), weights = TRUE))
+
+mods_all <- c(mada_mods, mora_mods)
+
+## profile CIs
+mods_covars <- Filter(function(x) !grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
+prof_CI_covars <- lapply(mods_covars, confint)
+mods_pop <- Filter(function(x) grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
+prof_CI_pop <- lapply(mods_pop, confint)
+
+bite_mods_covar_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_covars),
+                                intercept = mapply(function(x) coef(x)["intercept"], mods_covars),
+                                beta_upper = mapply(function(x) x["beta", 2], prof_CI_covars),
+                                beta_lower = mapply(function(x) x["beta", 1], prof_CI_covars),
+                                intercept_lower = mapply(function(x) x["intercept", 1], prof_CI_covars),
+                                intercept_upper = mapply(function(x) x["intercept", 2], prof_CI_covars),
+                                likelihood = mapply(function(x) logLik(x), mods_covars),
+                                model = mapply(function(x) attributes(x)$label, mods_covars)))
+
+bite_mods_pop_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_pop),
+                                beta_pop = mapply(function(x) coef(x)["beta_pop"], mods_pop),
+                                intercept = mapply(function(x) coef(x)["intercept"], mods_pop),
+                                beta_upper = mapply(function(x) x["beta", 2], prof_CI_pop),
+                                beta_lower = mapply(function(x) x["beta", 1], prof_CI_pop),
+                                beta_pop_upper = mapply(function(x) x["beta_pop", 2], prof_CI_pop),
+                                beta_pop_lower = mapply(function(x) x["beta_pop", 1], prof_CI_pop),
+                                intercept_lower = mapply(function(x) x["intercept", 1], prof_CI_pop),
+                                intercept_upper = mapply(function(x) x["intercept", 2], prof_CI_pop),
+                                likelihood = mapply(function(x) logLik(x), mods_pop),
+                                model = mapply(function(x) attributes(x)$label, mods_pop)))
+check <- bind_rows(bite_mods_covar_df, bite_mods_pop_df)
+
+## Predictions: fixed
+ttimes_plot <- seq(0, 25, by = 0.5)
+distance_plot <- seq(0, 200, by = 2.5)
+pop_plot <- seq(1000, 1e6, length.out = 30)
+
+library(iterators)
+preds_fixed <- run.predicts.fixed(covar_df = check, pop_fixed = 1e5, pop_seq = pop_plot)
+preds_data <- run.predicts.data(covar_df = check, run_type = "predict.data")
+preds_raw <- run.predicts.data(covar_df = check, run_type = "predict.raw")
+
+ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
+  geom_point() +
+  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
+  facet_grid(covar_name ~ pop_predict, scales = "free_x") +
+  expand_limits(y = 0)
 ```
 
-### Sensitivity analyses
-```{r sensitivity analysis multivariate}
+### S4. Sensitivity of model parameter estimates to contact and RLE cut-offs
+
+```{r rles and cut-offs, cache = TRUE}
+consec_days <- c(5, 15, 30)
+contact_cutoff <- c(0, 3, 10)
+combine <- function(x, ...) {  
+  mapply(rbind, x, ..., SIMPLIFY = FALSE)
+}
+mods <- 
+    foreach(i = 1:length(consec_days), .combine = combine, .multicombine = TRUE) %:%
+    foreach(k = 1:length(contact_cutoff), .combine = combine, .multicombine = TRUE) %do% {
+    exps_dist <- get.district.data(ctar_data = ctar_data, ctar_metadata = ctar_metadata, 
+                               mada_district = mada_district, catchments, 
+                               consec_days = consec_days[i], 
+                               contact_cutoff = contact_cutoff[k], reporting_thresh = 0.25, 
+                               start = ymd("2014-01-01"), end = ymd("2017-12-31"))
+
+    mada_mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                             pop_predict = c("addPop", "flatPop"),
+                             pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
+                             covars = list(distance_district = exps_dist$distance, 
+                                           distance_commune = comm_covars$distance, 
+                                           ttimes_district = exps_dist$ttimes_weighted_dist/60,
+                                           ttimes_commune = comm_covars$ttimes_weighted/60), 
+                             sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
+    mada_mods_pop <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                              pop_predict = "onlyPop", 
+                              pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
+                              covars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop), # dummy covars
+                              sum_it = c(FALSE, TRUE), run_type = "optim", label = "Mada")
+
+    mods_all <- c(mada_mods_covars, mada_mods_pop)
+
+    ## profile CIs
+    mods_covars <- Filter(function(x) !grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
+    prof_CI_covars <- lapply(mods_covars, confint)
+    mods_pop <- Filter(function(x) grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
+    prof_CI_pop <- lapply(mods_pop, confint)
+    bite_mods_covar_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_covars),
+                                    intercept = mapply(function(x) coef(x)["intercept"], mods_covars),
+                                    beta_upper = mapply(function(x) x["beta", 2], prof_CI_covars),
+                                    beta_lower = mapply(function(x) x["beta", 1], prof_CI_covars),
+                                    intercept_lower = mapply(function(x) x["intercept", 1], prof_CI_covars),
+                                    intercept_upper = mapply(function(x) x["intercept", 2], prof_CI_covars),
+                                    likelihood = mapply(function(x) logLik(x), mods_covars),
+                                    model = mapply(function(x) attributes(x)$label, mods_covars)))
+
+    bite_mods_pop_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_pop),
+                                    beta_pop = mapply(function(x) coef(x)["beta_pop"], mods_pop),
+                                    intercept = mapply(function(x) coef(x)["intercept"], mods_pop),
+                                    beta_upper = mapply(function(x) x["beta", 2], prof_CI_pop),
+                                    beta_lower = mapply(function(x) x["beta", 1], prof_CI_pop),
+                                    beta_pop_upper = mapply(function(x) x["beta_pop", 2], prof_CI_pop),
+                                    beta_pop_lower = mapply(function(x) x["beta_pop", 1], prof_CI_pop),
+                                    intercept_lower = mapply(function(x) x["intercept", 1], prof_CI_pop),
+                                    intercept_upper = mapply(function(x) x["intercept", 2], prof_CI_pop),
+                                    likelihood = mapply(function(x) logLik(x), mods_pop),
+                                    model = mapply(function(x) attributes(x)$label, mods_pop)))
+    check <- bind_rows(bite_mods_covar_df, bite_mods_pop_df)
+
+    check$consec_days <- consec_days[i]
+    check$contact_cutoff <- contact_cutoff[k]
+    # mod_attrs <- str_split(check$model, " ")[[1]]
+    # mod_pop <- mod_attrs[2]
+    # covar_name <- mod_attrs[3]
+    # scale <- ifelse(mod_attrs[1] == "Mora", "Moramanga", 
+    #                 ifelse(grepl("district", covar_name, fixed = TRUE), "District", 
+    #                        "Commune summed"))
+    # covar_name <- str_split(covar_name, "_")[[1]][1]
+    preds_fixed <- run.predicts.fixed(covar_df = check, pop_fixed = 1e5, pop_seq = pop_plot)
+    preds_fixed$consec_days <- consec_days[i]
+    preds_fixed$contact_cutoff <- contact_cutoff[k]
+    preds_data <- run.predicts.data(covar_df = bite_mods, run_type = "predict.data")
+    preds_data$consec_days <- consec_days[i]
+    preds_data$contact_cutoff <- contact_cutoff[k]
+    list(check, preds_fixed, preds_data)
+  }
+
+mods_df <- mods[[1]]
+preds_fixed_df <- mods[[2]]
+preds_data_df <- mods[[3]]
+
+mod_attrs <- str_split(mods_df$model, " ")
+mods_df$structure <-mapply(`[[`, mod_attrs, 2)
+covar_name <- mapply(`[[`, mod_attrs, 3)
+mods_df$scale <- ifelse(lapply(mod_attrs, `[[`, 1) == "Mora", "Moramanga", 
+                ifelse(grepl("district", covar_name, fixed = TRUE), "District", 
+                       "Commune summed"))
+mods_df$covar <- mapply(`[[`, str_split(covar_name, "_"), 1)
+
+ggplot(data = mods_df, aes(x = structure, y = beta)) +
+  geom_boxplot(aes(color = covar)) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(data = mods_df, aes(x = structure, y = intercept)) +
+  geom_boxplot(aes(color = covar)) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(preds_fixed_df, aes(x = covar, y = predicted, color = interaction(consec_days, contact_cutoff), 
+                           linetype = scale)) + 
+  geom_line() + 
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
+  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
+  xlab("Travel time (hrs)") +
+  ylab("Reported bites per \n 100,000 persons") +
+  facet_wrap(scale ~ covar_name, scales = "free")
+
+ggplot(preds_data_df, aes(x = observed, y = predicted,
+                       color = scale)) + 
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
+  scale_color_manual(values = c("blue", "lightblue", "pink")) +
+  facet_grid(contact_cutoff ~ consec_days, scales = "free")
+
+```
+
+### S5. Sensitivity of burden estimates to paramater values (multivariate for R, p_rabid, rho_max)
+
+```{r sensitivity analysis multivariate, cache = TRUE}
 hdr_vals <- c(25, 20, 15, 10, 5)
 p_rabid_vals <- c(0.2, 0.3, 0.4, 0.5, 0.6)
-rho_max <- c(0.80, 0.85, 0.9, 0.95, 0.9)
+rho_max_vals <- c(0.80, 0.85, 0.9, 0.95, 0.9)
 library(foreach)
@@
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
-  get.reporting_burden(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
-                     ttimes = dist_ttimes$ttimes, scenario = dist_ttimes$scenario,
+ get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+                     ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
                      B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
                      hdr = hdr_vals[k], p_rabid = p_rabid_vals[i], 
-                     rho_max = rho_max_vals[j], type = "burden")
+                     rho_max = rho_max_vals[j], type = "burden", scale = FALSE)
   }
@@
 
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+comm_ttimes <- as.data.table(comm_ttimes)
+comm_ttimes$mdg_cm_ <- as.character(mada_communes$mdg_cm_)
+comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
+comm_ttimes <- rbind(comm_ttimes, 
+                       as.data.frame(list(mdg_cm_ = comm_catchments$mdg_cm_, scenario = 0,
+                                          ttimes = comm_catchments$ttimes_weighted)))
+comm_ttimes %>%
+  left_join(select(mada_communes@data, mdg_cm_, commune, pop = MDG_201, mtch_ds)) -> comm_ttimes
+
+scenario_comm <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+ get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
+                     ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
+                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada, 
+                     hdr = hdr_vals[k], p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden", scale = FALSE)
+  }
+
+scenario_comm %>%
+  filter(scenario == 0) %>%
+  group_by(hdr, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ hdr)
+ggplot(data = filter(scenario_comm, scenario == 0), aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ hdr)
+
+scenario_comm %>%
+  group_by(scenario, hdr, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths)) -> deaths_scenario
+deaths_scenario %>%
+  left_join(select(deaths_multivar, hdr, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
+ggplot(data = deaths_scenario, aes(x = scenario, y = deaths, color = p_rabid)) +
+  geom_point() +
+  facet_grid(hdr ~ rho_max, scales = "free_y")
+
 ```
 
-```{r scaling w/ pop}
+### S6. Sensitity of scenario analyses to parameter values (multivariate for R, p_rabid, rho_max)
+
+```{r scaling w/ pop, cache = TRUE}
 ## Pop scaling
@@
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
-  get.reporting_burden(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
-                     ttimes = dist_ttimes$ttimes, scenario = dist_ttimes$scenario,
+ get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+                     ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
                      B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
@@
 
+## Commune
+scenario_scaled <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+ get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
+                     ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
+                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
+                     slope = scale_df$sf[k], intercept = scale_df$intercept[k])
+  }
+
+scenario_scaled %>%
+  filter(scenario == 0) %>%
+  group_by(slope, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ slope)
+ggplot(data = filter(scenario_scaled, scenario == 0),  
+       aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ slope)
+
+scenario_scaled %>%
+  group_by(scenario, slope, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths)) -> deaths_scenario
+deaths_scenario %>%
+  left_join(select(deaths_multivar, slope, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
+ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
+  geom_point() +
+  facet_grid(slope ~ rho_max, scales = "free_y")
+
 ```
-<br>
-
-## DISCUSSION
-
-### Key findings
-
-### Strengths and Limitations
-
-### Broader context
-
-### Conclusions
-
-## Acknowledgements
-
-## Supplementary Materials
-### S1. RLE thresholds for reporting
-### S2. Removing contacts from the data set
-### S3. Comparing models of reported bites
-### S4. Sensitivity of model parameter estimates to contact and RLE cut-offs
-### S5. Sensitivity of burden estimates to paramater values (multivariate for R, p_rabid, rho_max)
-### S6. Sensitity of scenario analyses to parameter values (multivariate for R, p_rabid, rho_max)
+
+### Scale example
+
+```{r scale example}
+## Comparing scales
+comm_data$reporting_min <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                                            hdr = 5, p_rabid = 0.2, type = "reporting", scale = FALSE)
+
+comm_data$reporting_max <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                                            hdr = 25, p_rabid = 0.6,  type = "reporting", scale = FALSE)
+comm_data %>%
+  group_by(district) %>%
+  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
+  summarize(reporting_min = sum(reporting_min*prop_pop), 
+            reporting_max = sum(reporting_max*prop_pop)) %>%
+  mutate(reporting_med = (reporting_min + reporting_max)/2,
+         type = "summed from commune") -> summed_rep_dist
+
+mada_district@data %>%
+  mutate(reporting_min = get.burden.fixed(pop = pop2015un, ttimes = ttimes_weighted/60, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
+                                              p_rabid = 0.2, type = "reporting", scale = FALSE), 
+         reporting_max =get.burden.fixed(pop = pop2015un, ttimes = ttimes_weighted/60, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
+                                              p_rabid = 0.6,type = "reporting", scale = FALSE),
+         reporting_med = (reporting_min + reporting_max)/2,
+         type = "district level") %>%
+  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp
+
+pop_plot = 1e5
+rep_comp <- rbind(rep_comp, summed_rep_dist)
+rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]
+
+ggplot(rep_comp, aes(x = reporting_med, y = reorder(district, ttime), 
+                              color = type)) +
+  ggstance::geom_pointrangeh(aes(xmin = reporting_min, xmax = reporting_max), fatten = 2)
+
+bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
+R_100k_min <- pop_plot/25*0.01*0.39
+R_100k_max <- pop_plot/5*0.01*0.39
+
+plot(ttimes_plot, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
+     type = "l", bty = "n",
+     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
+lines(ttimes_plot, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+     col = "darkred")
+lines(ttimes_plot, 
+     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
+     col = "red", lty = 2)
+lines(ttimes_plot, 
+     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
+     col = "red", lty = 2)
+polygon(x = c(ttimes_plot, rev(ttimes_plot)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
+        border = NA, col = alpha("red", 0.2))
+polygon(x = c(ttimes_plot, rev(ttimes_plot)), 
+        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
+             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
+        border = NA, col = alpha("red", 0.6))
+
+## Mora
+comm_data %>%
+  filter(district == "Moramanga") -> comm_level
+bites_100k <- exp(B_ttimes_mada*comm_level$ttimes_weighted + B_0_mada)*comm_level$pop
+R_100k_min <- comm_level$pop/25*0.01*0.39
+R_100k_max <- comm_level$pop/5*0.01*0.39
+
+## district val for Moramanga
+## need to merge names
+exps_dist %>%
+  filter(dist_name == "Moramanga") %>%
+  mutate(rep_max =get.burden.fixed(pop = pop, ttimes = ttimes_weighted_dist, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
+                                              p_rabid = 0.6),
+         rep_min =get.burden.fixed(pop = pop, ttimes = ttimes_weighted_dist, 
+                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
+                                              p_rabid = 0.2)) -> dist_level
+dist_level %$% plot(ttimes_weighted_dist/60, rep_max, col = "red", pch = 20)
+dist_level %$% points(ttimes_weighted_dist/60, rep_min, col = "red", pch = 20)
+dist_level %$% abline(v = ttimes_weighted_dist/60, col = "red", lty = 2)
+
+```
+
+```{r scenario analysis}
+# ## new ctar
+# expanded <- read.csv("output/incremental_ARMC_nofilter.csv")
+# dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
+# comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+# 
+# 
+# scenario_scaled <- 
+#   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+#   foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
+#   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+#   get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+#                      ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
+#                      B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
+#                      p_rabid = p_rabid_vals[i], 
+#                      rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
+#                      slope = scale_df$sf[k], intercept = scale_df$intercept[k])
+#   }
+# dist_burden_min <- do.call("rbind", apply(dist_ttimes, 2, function(x) get.burden.fixed(pop = mada_district$pop2015adj, ttimes = x/60, names = mada_district$mdg_dis_co, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, p_rabid = 0.6, type = "burden")))
+# dist_burden_max <- do.call("rbind", apply(dist_ttimes, 2, function(x) get.burden.fixed(pop = mada_district$pop2015adj, ttimes = x/60, names = mada_district$mdg_dis_co, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, p_rabid = 0.2, type = "burden")))
+# 
+# comm_burden_min <- do.call("rbind", apply(comm_ttimes, 2, function(x) get.burden.fixed(pop = mada_communes$MDG__201, ttimes = x/60, names = mada_communes$mtch_ds, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 25, p_rabid = 0.6, type = "burden")))
+# comm_burden_max <- do.call("rbind", apply(comm_ttimes, 2, function(x) get.burden.fixed(pop = mada_communes$MDG__201, ttimes = x/60, names = mada_communes$mtch_ds, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 5, p_rabid = 0.2, type = "burden")))
+# 
+# ## Decrease in burden over times proportional
+# comm_burden_min$scale <- "commune"
+# comm_burden_max$scale <- "commune"
+# dist_burden_max$scale <- "district"
+# dist_burden_min$scale <- "district"
+# burden_range <- bind_rows(comm_burden_max, comm_burden_min, dist_burden_max, dist_burden_min)
+# burden_range %>%
+#   group_by(scenario, scale, hdr) %>%
+#   summarize(burden_total = sum(deaths)) %>%
+#   group_by(scale, hdr, names) %>%
+#   mutate(burden_base = deaths[scenario == 0]) -> burden_all
+#   
+# ggplot(data = comm_burden_min, aes(x = scenario, ))
+# 
+# plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
+#      ylab = "Proportion of deaths compared to baseline ")
+# lines(colSums(dist_burden_min, na.rm = TRUE)/sum(dist_burden_min[, 1], na.rm = TRUE),
+#       col = "purple")
+# lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
+#      type = "l", col = "blue")
+# lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")
+# 
+# ## Decrease in burden over times absolute
+# plot(colSums(comm_burden_min, na.rm = TRUE), type = "l", col = "blue", bty = "n",
+#      ylim = c(0, 1500), xlab = "# Additional ARMC",
+#      ylab = "Number of deaths")
+# lines(colSums(dist_burden_min, na.rm = TRUE), col = "purple")
+# lines(colSums(comm_burden_max, na.rm = TRUE), type = "l", col = "blue")
+# lines(colSums(dist_burden_max, na.rm = TRUE), col = "purple")
+
+```
 

### 2019-03-14 accessPEP.Rmd
with fixed sensitivity estimates

@@
 
-In 2018, Gavi added PEP to their vaccine investment portfolio, and soon countries will be able to apply for funding to expand access to PEP.
-Expanding access strategically by identifying places with limited access...
-
-In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country.
+In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country.
 PEP is not available at any other public clinics or through the private sector.
@@
 Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country.
-To get spatial estimates of disease burden in this context, we assume that bite incidence reflects access to PEP rather than than differences in rabies incidence, using travel times to clinics as a predictor of reported bites.
+To get spatial estimates of disease burden in this context, we use data on reported bite incidence at the sub-national level (district) and assume that bite incidence reflects access to PEP rather than than differences in rabies incidence.
+We fit the data to travel times to the closest ARMC, a metric of geographic access.
 Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.
-Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.
+
+ In 2018, Gavi added PEP to their vaccine investment portfolio, and by 2020, Gavi-eligible countries will be able to apply for funding to expand access to PEP, which has the potential to drastically reduce burden of rabies.
+We explore the impact of this potential intervention using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.
 
@@
 
+## Getting other data
+rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
+                          end_date = ymd("2017-12-31"), 
+                          threshold = 10)
+ctar_contacts <- get.contacts(ctar_data, times_sd = 3)
+ctar_contacts <- ctar_contacts[["ctar_data"]]
 ```
@@
 Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC.
-We excluded any catchment areas for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).
+We excluded any catchment areas for which the clinic did not submit any forms (catchments = `r length(no_data)`, districts = `r nrow(exps_dist) - nrow(mada_district)`, in grey in Fig 1A).
 
 As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2).
-Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold.
+Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so in subsequent analyses we looked at the sensitivity of model parameter estimates to changing this threshold.
 To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
 
-Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al.
+Previous work in the Moramanga District showed that low risk contacts, i.e.
+touching or feeding a suspected animal or human rabies case, make up approximately 20% of patients reporting to ARMC (Rajeev et al.
 2018).
+These low risk exposures do not contribute to burden and therefore we attempted to exclude them from our analysis.
 Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
 We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4).
-For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])**100, 2)"* of known contacts.
-We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)**100, 2)*% of patient records from the national data.
-We also compared this method to assuming that 20% of total bites at the district level were contacts.
+For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_contacts %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_contacts %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
+We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_contacts %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_contacts)*100, 2)` % of patient records from the national data.
 
@@
 
-After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from *r length(unique(exps_dist$dist_catch))* catchments (Figure 4).
+After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).
 
@@
 ### Model of reported bites as a function of travel time
-We modeled bites as a function of travel times as follows:
+We modeled incidence of bites as a function of travel times in hours as follows:
 $$ 
@@
 $$
-where $mu$ is the expected number of bites given travel times and population for a given location. In this way, we can model the impact of travel time on incidence as 
-$exp(\beta_{t}T + \beta_0)$ gives an estimate of incidence and multiplying by the population size gives the expected number of bites, so that we can fit this model to the number of bites in each location.
-
+where $mu$ is the expected number of bites given travel times and population for a given location.
 We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
@@
 We fit this to our available data in three ways:
-1) the commune data available at the moramanga district, where we fit to $mu_j$
-2) The district data available nationally
-3) To compare estimates from 1 directly to the national data, the district data but where $mu_d$ is the sum of the travel time effect at the commune level, i.e.:
+1) Commune level bite incidence from the Moramanga District
+2) District level bite incidence as calculated above from the national database
+3) District level bite incidence, but with commune level travel times ($T_j$) so that:
 
 $$ 
-  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
+  \mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
 $$
 
- As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models (replacing $Tj$ with log-transformed population in $j$ to avoid squared terms), as a test of whether the observed patterns could be predicted by bite incidence scaling with population size alone.
-We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.
+As travel times are correlated with human population size (Figure S4), we also compared how well bites were predicted by human population size.
+We used a generalized linear model with a possion response-now predicting the number of bites rather than bite incidence` with either population size alone or in combination with travel times, so that incidence could scale with population size.
+We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access, replacing travel times in hours ($T$) with distance from the nearest ARMC in km ($D$).
+Distance to the nearest ARMC was taken as the minimum distance from the centroid of the commune or district to any ARMC.
 
@@
 
-TO DO: Instead of reporting calculate as incidence of deaths here!!
 We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
@@
 $\rho_{max}$ is taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e.
-the area with the minimum travel time in the district, on average *xx* minutes).
+the area with the minimum travel time in the district, on average `r round(exps_mora$ttimes_weighted[exps_mora$commune == "Moramanga"], 2)` minutes).
 
@@
 2018).
-We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence.
-As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).
-
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows):
+We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence, which gives a range of approximately `r 1e5/25*0.01*0.39` - `r 1e5/5*0.01*0.39` exposures per 100,000 persons.
+As there is little data on dog populations in Madagascar, this range encompasses both a wide range of observed HDRs and exposure rates across Africa (cite!).
+
+To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
 $$
@@
 
+We also look at how this may impact vaccine demand.
+Specifically, we fit the observed daily distribution of patients receiving PEP at each clinic to a negative binomial distribution, with both the dispersion parameter and mean as a function of the annual average number of reported bites.
+We use this to simulate clinic throughput and vials used as clinics are added, assuming routine vial sharing practices used currently in Madagascar (i.e.
+each vial is split between two patients, and vials are discarded at the end of each day).
+We project the potential increase in vial demand and change in the ratio of number of vials per death averted.
+
 ### Sensitivity analyses for burden estimates 
@@
 Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden.
-We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden changes if rabies incidence were to scale positively or negatively with human population.
+We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population.
 
@@
 
-ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
+library(patchwork)
+a <- ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
   geom_line() + 
@@
   scale_color_manual(values = c("navy", "blue", "purple")) +
-  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
+  scale_fill_manual(values = c("blue", "lightblue", "pink"), name = "Model scale") +
   xlab("Travel time (hrs)") +
-  ylab("Reported bites per \n 100,000 persons")
-
-ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
+  ylab("Reported bites per \n 100,000 persons") +
+  labs(tag = "A")
+
+b <- ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
   geom_point(alpha = 0.5) +
@@
   expand_limits(y = c(0, 4000)) +
-  scale_color_manual(values = c("navy", "blue", "purple"))
-
+  scale_color_manual(values = c("navy", "blue", "purple"), name = "Model scale") +
+  xlab("Mean annual bites observed") +
+  ylab("Mean annual bites predicted") +
+  labs(tag = "B")
+a / b
+
+kable(bite_mods)
 ```
@@
 
+The best fitting model did not include population as a predictor for the district data...write up from supplement!
+
 ### Estimation of burden and reporting 
-
-Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 6).
+```{r}
+B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
+B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]
+
+B_ttimes_dist <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_district"]
+B_0_dist <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_district"]
+
+ttimes <- seq(0, 1200, by = 60)/60
+p_rabid_vals <- c(0.2, 0.6)
+hdr_vals <- c(5, 25)
+beta_vals <- list(District = B_ttimes_dist, Commune = B_ttimes_mada)
+int_vals <- list(District = B_0_dist, Commune = B_0_mada)
+
+rep_df <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
+  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
+  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
+    rep <- as.data.frame(list(p_rabid = p_rabid_vals[i], ttimes = ttimes, hdr = hdr_vals[k], 
+                                         reporting = get.burden.fixed(pop = 1e5, ttimes = ttimes,
+                                                          B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
+                                            hdr = hdr_vals[k], p_rabid = p_rabid_vals[i],
+                                            type = "reporting", scale = FALSE), scale = names(beta_vals)[j]))
+  }
+
+rep_min_max <- 
+  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
+    rep_max <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
+                                            hdr = 25, p_rabid = 0.6,
+                                            type = "reporting", scale = FALSE)
+
+    rep_min <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
+                                            hdr = 5, p_rabid = 0.2,
+                                            type = "reporting", scale = FALSE)
+    rep <- as.data.frame(list(ttimes = ttimes, rep_max = rep_max, rep_min = rep_min, 
+                              scale = names(beta_vals)[j]))
+  }
+
+ggplot(rep_df, aes(x = ttimes, y = reporting, color = interaction(p_rabid, hdr), linetype = as.factor(hdr))) +
+  geom_line() +
+  facet_wrap(~ scale) +
+  scale_color_manual(values = c("red", "red", "blue", "blue"))
+
+ggplot(rep_min_max, aes(x = ttimes, y = rep_min, fill = scale)) +
+  geom_line(aes(y = rep_min, color = scale), linetype = 2) +
+  geom_line(aes(y = rep_max, color = scale), linetype = 1) +
+  geom_ribbon(aes(ymax = rep_max, ymin = rep_min), alpha = 0.5) +
+  scale_fill_manual(values = c("navy", "blue")) +
+  scale_color_manual(values = c("navy", "blue"))
+
+```
+
+Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure N).
 Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5.5 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).
+We found that the commune model of bites estimated a lower range of reporting probabilities that the district model (Figure SX), so we compared estimates of these two models in subsequent analyses.
 
@@
 comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
-B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
-B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]
-
-B_ttimes_dist <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_district"]
-B_0_dist <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_district"]
 
@@
             deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
-            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) %>%
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE),
+            averted_mean = sum(averted_mean, na.rm = TRUE), 
+            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
+            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE)) %>%
   rename(names = mdg_dis_co) -> results_comm_summarized
@@
   select(scenario, names, deaths_mean, deaths_lowerCI, deaths_upperCI, 
+         averted_mean, averted_lowerCI, averted_upperCI,
          model) -> results_dist_med
@@
             deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
-            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_natl
-
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE), 
+            averted_mean = sum(averted_mean, na.rm = TRUE), 
+            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
+            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE)) -> results_natl
+
+## Baseline results
+results_all %>% filter(scenario == 0) -> results_all_base
+ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
+                              color = model)) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
+  ylab("") +
+  xlab("Average annual deaths") +
+  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
+                                "commune" = alpha("purple", 0.5)), 
+                     name = "Scale") -> p
+p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+
+## Deaths averted
+ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = averted_mean, 
+                              color = model)) +
+  ggstance::geom_pointrangeh(aes(xmin = averted_lowerCI, xmax = averted_upperCI), fatten = 2) +
+  ylab("") +
+  xlab("Average annual deaths") +
+  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
+                                "commune" = alpha("purple", 0.5)), 
+                     name = "Scale") -> p
+p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+
+kable(results_all_base %>%
+  group_by(model) %>%
+  summarize(averted = sum(averted_mean), 
+            deaths = sum(deaths_mean)) -> summary_burden)
+
+```
+
+When we estimate burden of deaths stochastically within this range of incidence and of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure XA, results presented aggregated at the district level).
+Overall, we estimate average annual deaths between `r min(summary_burden$deaths)` - `r max(summary_burden$deaths)` and deaths_averted between `r min(summary_burden$averted)` - `r max(summary_burden$averted)`.
+
+Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust...write up from supplement!
+
+### Expanding access to PEP
+
+``` {r scenario}
 ## Plotting national level
@@
 
-## Baseline results
-results_all %>% filter(scenario == 0) -> results_all_base
-ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
-                              color = model)) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
-  ylab("") +
-  xlab("Average annual deaths") +
-  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
-                                "commune" = alpha("purple", 0.5)), 
-                     name = "Scale") -> p
-p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+## Deaths averted
+ggplot(data = results_natl, aes (x = scenario, y = averted_mean, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = averted_upperCI, ymax = averted_lowerCI, 
+                  color = NULL, fill = model),
+              alpha = 0.5) +
+  scale_color_manual(values = c("district" = "blue", "commune" = "purple"),  guide = "none") +
+  scale_fill_manual(values = c("district" = "blue", "commune" = "purple")) +
+  guides(color = guide_legend(), fill = guide_legend(override.aes = list(linetype = 1)))
+
+ggplot(data = results_all, aes(x = averted_mean, y = reorder(district, ttimes_weighted), 
+                               color = scenario)) +
+  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
+                             fatten = 2) +
+  facet_wrap(~model)
 
@@
 
-When we estimate burden of deaths stochastically within this range of incidence and given our high and low estimates of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure 7, results presented aggregated at the district level).
-Overall, we estimate average annual deaths between ....
-**Also estimate deaths averted here!**.
-
-When we compare our burden estimates at the district vs.
-the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden.
-
-### Expanding access to PEP
-
-<br>
+We find that both expanding access to PEP reduces burden of rabies death, but this effect decays as ARMC are added, saturating after about approximately 200 clinics (or 2 ARMC per district).
+This effect is more pronouned for the commune level model than for the district model.
+Sensitivity analyses...write up from supplement!
+
+### Estimating vaccine demand with increased access
+
+```{r vial analysis}
+# ## new ctar
+rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
+                          end_date = ymd("2017-12-31"), 
+                          threshold = 10)
+date_mat <- rep_data[["date_mat"]]
+dose_mat <- rep_data[["doses_wide"]]
+colnames(date_mat) <- colnames(dose_mat)[2:(ncol(dose_mat)-1)]
+doses <- melt(dose_mat, id.vars = "date_de_consultation", value.name = "patients", variable.name = "id_ctar")
+
+date_mat %>% 
+  as_tibble() %>%
+  mutate(date = dose_mat$date_de_consultation) -> date_mat
+exclude <- melt(date_mat, id.vars = "date", value.name = "exclude", variable.name = "id_ctar")
+doses %>%
+  left_join(exclude, by = c("date_de_consultation" = "date", "id_ctar" = "id_ctar")) -> doses
+doses %>%
+  filter(exclude == 1) %>%
+  group_by(id_ctar) %>%
+  mutate(sum = sum(patients, na.rm = TRUE), mean = mean(patients, na.rm = TRUE)) %>%
+  filter(sum > 10) -> doses
+
+ggplot(data = doses, aes(x = date_de_consultation, y = id_ctar)) +
+  geom_tile(aes(fill = patients))
+
+library(ggridges)
+ggplot(doses, aes(x = patients, y = id_ctar)) +
+  geom_density_ridges()
+
+doses %>%
+  split(.$id_ctar) %>% # from base R
+  map(~ fitdistr(.$patients, "negative binomial")) -> dist_ts
+
+coeff_df <- as.data.frame(list(mu = mapply(function(x) coef(x)["mu"], dist_ts),
+                                k = mapply(function(x) coef(x)["size"], dist_ts),
+                                mu_sd = mapply(function(x) x$sd["mu"], dist_ts),
+                                k_sd = mapply(function(x) x$sd["size"], dist_ts)))
+coeff_df$id_ctar <- names(dist_ts)
+ggplot(data = coeff_df, aes(x = mu, y = k)) + geom_point()
+
+exps_dist %>%
+  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
+  mutate(id_ctar = as.character(id_ctar)) %>%
+  group_by(id_ctar) %>%
+  summarize(bites = sum(bites)) %>%
+  left_join(coeff_df, by = c("id_ctar" = "id_ctar")) -> coeff_df
+ggplot(data = coeff_df, aes(x = bites, y = mu)) + geom_point()
+ggplot(data = coeff_df, aes(x = bites, y = k)) + geom_point()
+
+exps_dist %>%
+  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
+  mutate(id_ctar = as.character(id_ctar)) %>%
+  group_by(id_ctar) %>%
+  summarize(bites = sum(bites)) %>%
+  left_join(doses, by = c("id_ctar" = "id_ctar")) -> doses
+
+get.throughput <- function(patient_ts, bites, B_k, B_0_k, B_mu, B_0_mu) {
+  mu <- exp(B_mu*bites/100 + B_0_mu) 
+  k <- exp(B_k*bites/100 + B_0_k)
+  return(-sum(dnbinom(patient_ts, size = k, mu = mu, log = TRUE)))
+}
+
+fit <- mle2(get.throughput, start = list(B_0_mu = 0.1, B_k = 0.1, B_mu = 0.1, B_0_k = 0.1),
+             data = list(patient_ts = doses$patients, bites = doses$bites))
+
+## Get the observed + predicted distribution for each set of coeffs
+check <- foreach(i = iter(coeff_df, by = "row"), .combine = rbind) %do% {
+  check <- rnbinom(n = 365, mu = exp(0.029*i$bites/100 + -0.307),
+                  size = exp(0.022*i$bites/100 + -0.39))
+  as.data.frame(list(patients = check, type = "predicted", id_ctar = i$id_ctar))
+}
+
+doses %>%
+  select(id_ctar, patients) %>%
+  mutate(type = "observed") %>%
+  bind_rows(check) -> compare_dists
+
+compare_dists %>%
+  group_by(type, id_ctar, patients) %>%
+  summarize(count = n()) %>%
+  group_by(type, id_ctar) %>% # next get # of days total (should always be 365 for predicted)
+  mutate(total = sum(count), prop = count/total) -> compare_dists
+compare_dists %>%
+  mutate(prop_plot = ifelse(type == "predicted", -prop, prop)) -> compare_dists
+
+ggplot(compare_dists, aes(x = patients, y = prop_plot, fill = type)) +
+  geom_col(color = NA, alpha = 0.5, position = "identity") +
+  scale_fill_manual(values = c("red", "blue")) +
+  facet_wrap(~id_ctar, scales = "free") +
+  xlab("# of Patients") +
+  ylab("Proportion of days")
+
+rep_bites = seq(0, 15000, by = 100)
+predicts <- as.data.frame(list(rep_bites = rep_bites,
+                               predicted_mu = exp(0.029*rep_bites/100 + -0.307),
+                               predicted_k = exp(0.022*rep_bites/100 + -0.39)))
+ggplot(predicts, aes(x = rep_bites, y = predicted_mu)) +
+  geom_line()
+ggplot(predicts, aes(x = rep_bites, y = predicted_k)) +
+  geom_line()
+
+```
+
+We fit the distributions of patients reporting per day to the estimated average annual bites reported to each clinic.
+We found that as reported bites increase, the average number of patients reporting per day increases linearly, and that the heterogeneity in this increases with lower reported bites to a clinic.
+
+We use this to estimate how vaccine demand will change as access to PEP is expanded...
 
@@
 comm_ttimes %>%
-  left_join(select(mada_communes@data, mdg_cm_, commune, pop = MDG_201, mtch_ds)) -> comm_ttimes
+  left_join(select(mada_communes@data, mdg_cm_, commune, pop = MDG__201, mtch_ds)) -> comm_ttimes
 
@@
   group_by(hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+  summarize(deaths = sum(deaths, na.rm = TRUE), scenario = 0) -> deaths_multivar
+
 ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
@@
 
+hdr_name <- c(`5` = paste0(1e5/5*0.01*0.39, " per \n 100k"),
+                    `10` = paste0(1e5/10*0.01*0.39, " per \n 100k"),
+                    `15` = paste0(1e5/15*0.01*0.39, " per \n 100k"),
+                    `20` = paste0(1e5/20*0.01*0.39, " per \n 100k"),
+                    `25` = paste0(1e5/25*0.01*0.39, " per \n 100k"))
+
 scenario_comm %>%
   group_by(scenario, hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths)) -> deaths_scenario
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_scenario
 deaths_scenario %>%
   left_join(select(deaths_multivar, hdr, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
-ggplot(data = deaths_scenario, aes(x = scenario, y = deaths, color = p_rabid)) +
-  geom_point() +
-  facet_grid(hdr ~ rho_max, scales = "free_y")
+ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
+  geom_point() +
+  facet_grid(hdr ~ rho_max, scales = "free_y", labeller = labeller(hdr = as_labeller(hdr_name)))
 
@@
 ## Pop scaling
+p_rabid_vals <- c(0.2, 0.3, 0.4, 0.5, 0.6)
+rho_max_vals <- c(0.80, 0.85, 0.9, 0.95, 0.9)
+
 pop <- mada_district$pop2015adj
@@
 
-scale_df <- rbind(as.data.frame(list(scale = "neg", 
+scale_df_dist <- rbind(as.data.frame(list(scale = "neg", 
                                      sf = seq(neg$coefficients[2], 0, length.out = 5),
@@
 }
-scale_df %>%
-  merge(., select(mada_district@data, mdg_dis_co, ttimes_weighted, pop = pop2015adj)) -> scale_df_dist
-ggplot(data = scale_df_dist, aes(x = log(pop), 
+
+scale_df_dist %>%
+  merge(., select(mada_district@data, mdg_dis_co, ttimes_weighted, pop = pop2015adj)) -> scale_df_check
+ggplot(data = scale_df_check, aes(x = log(pop), 
                             y = constrained_inc(sf, intercept, pop, inc100k_max, inc100k_min)*1e5, 
@@
   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
-  foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df_dist), .combine = rbind) %:%
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
- get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
+  check <- get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
                      ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
@@
                      rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
-                     slope = scale_df$sf[k], intercept = scale_df$intercept[k])
+                     slope = scale_df_dist$sf[k], intercept = scale_df_dist$intercept[k])
+  check$scale <- scale_df_dist$scale[k]
+  check
   }
@@
 scenario_scaled %>%
+  mutate(slope = case_when(slope == 0 & scale =="neg" ~ -10, 
+                           slope == 0 & scale =="pos" ~ 10,
+                           slope > 0 | slope < 0 ~ round(slope*1e10))) -> scenario_scaled_dist
+scenario_scaled_dist %>%
   filter(scenario == 0) %>%
@@
   summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+
+slope_name <- c(`-9` = "----", `-7` = "---", `-4` = "--", `-2` = "--",
+                    `-10` =  "max", `9` = "++++", `7` = "+++", `4` = "++", `2` = "+",`10` =  "min")
+
 ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
   geom_point() +
-  facet_grid(~ slope)
-ggplot(data = filter(scenario_scaled, scenario == 0),  
+  facet_grid(~ slope, labeller = labeller(slope = as_labeller(slope_name)))
+ggplot(data = filter(scenario_scaled_dist, scenario == 0),  
        aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
   geom_point() +
-  facet_grid(rho_max ~ slope)
-
-scenario_scaled %>%
+  facet_grid(rho_max ~ slope, labeller = labeller(slope = as_labeller(slope_name)))
+
+scenario_scaled_dist %>%
   group_by(scenario, slope, p_rabid, rho_max) %>%
@@
 ## Commune
+pop <- mada_communes$MDG__201
+pop <- pop[order(pop, decreasing = FALSE)]
+inc100k_max <- 0.01*0.39/5
+inc100k_min <- 0.01*0.39/25
+pos_scale <- seq(0.01*0.39/25, 0.01*0.39/5, length.out = length(pop))
+neg_scale <- seq(0.01*0.39/5, 0.01*0.39/25, length.out = length(pop))
+pos <- lm(pos_scale ~ pop) ## use these and constrain
+neg <- lm(neg_scale ~ pop) ## use these and constrain
+
+scale_df_comm <- rbind(as.data.frame(list(scale = "neg", 
+                                     sf = seq(neg$coefficients[2], 0, length.out = 5),
+                                     intercept = inc100k_max)),
+                  as.data.frame(list(scale = "pos", 
+                                     sf = seq(0, pos$coefficients[2], length.out = 5),
+                                     intercept = inc100k_min)))
+constrained_inc <- function(slope, intercept, pop, max, min){
+  inc <- slope*pop + intercept
+  inc[inc >= max] <- max
+  inc[inc <= min] <- min
+  return(inc)
+}
+
+scale_df_comm <- rbind(as.data.frame(list(scale = "neg", 
+                                     sf = seq(neg$coefficients[2], 0, length.out = 5),
+                                     intercept = inc100k_max)),
+                  as.data.frame(list(scale = "pos", 
+                                     sf = seq(0, pos$coefficients[2], length.out = 5),
+                                     intercept = inc100k_min)))
+constrained_inc <- function(slope, intercept, pop, max, min){
+  inc <- slope*pop + intercept
+  inc[inc >= max] <- max
+  inc[inc <= min] <- min
+  return(inc)
+}
+scale_df_comm %>%
+  merge(., select(mada_communes@data, mdg_cm_, ttms_ll, pop = MDG__201)) -> scale_df_check_comm
+ggplot(data = scale_df_check_comm, aes(x = log(pop), 
+                            y = constrained_inc(sf, intercept, pop, inc100k_max, inc100k_min)*1e5, 
+                            color = sf)) +
+  geom_point() +
+  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey")
+
 scenario_scaled <- 
   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
-  foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df_comm), .combine = rbind) %:%
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
+  check <- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
                      ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
@@
                      rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
-                     slope = scale_df$sf[k], intercept = scale_df$intercept[k])
+                     slope = scale_df_comm$sf[k], intercept = scale_df_comm$intercept[k])
+  check$scale <- scale_df_comm$scale[k]
+  check
   }
@@
 scenario_scaled %>%
+  mutate(slope = case_when(slope == 0 & scale =="neg" ~ -10, 
+                           slope == 0 & scale =="pos" ~ 10,
+                           slope > 0 | slope < 0 ~ round(slope*1e10))) -> scenario_scaled_comm
+
+scenario_scaled_comm %>%
   filter(scenario == 0) %>%
@@
   summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+
+slope_name <- c(`-9` = "----", `-7` = "---", `-4` = "--", `-2` = "--",
+                    `-10` =  "max", `9` = "++++", `7` = "+++", `4` = "++", `2` = "+",`10` =  "min")
+
 ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
   geom_point() +
-  facet_grid(~ slope)
-ggplot(data = filter(scenario_scaled, scenario == 0),  
+  facet_grid(~ slope, labeller = labeller(slope = as_labeller(slope_name)))
+ggplot(data = filter(scenario_scaled_comm, scenario == 0),  
        aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
   geom_point() +
-  facet_grid(rho_max ~ slope)
-
-scenario_scaled %>%
+  facet_grid(rho_max ~ slope, labeller = labeller(slope = as_labeller(slope_name)))
+
+scenario_scaled_comm %>%
   group_by(scenario, slope, p_rabid, rho_max) %>%
@@
   geom_point() +
-  facet_grid(slope ~ rho_max, scales = "free_y")
+  facet_grid(slope ~ rho_max, scales = "free_y", labeller = labeller(slope = as_labeller(slope_name)))
 
@@
 
-**Remember that the only other variable is p_death which should just shift these estimates up or down**
+```{r simulate vials used}
+## Get catchment
+# comm_catchments <- read.csv("output/temp_ARMC.csv", row.names = 1)
+# comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
+# comm_ttimes <- cbind(mada_communes@data$ttms_ll, comm_ttimes)
+# comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
+# comm_mat[, 1] <- comm_ttimes[, 1]
+# 
+# for (i in 2:ncol(comm_ttimes)){
+#   comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], comm_ttimes[, i], NA)
+# }
+# 
+# comm_mat <- as.data.table(comm_mat)
+# comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
+# comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+# comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+# 
+# comm_catchments$scenario <- 1:nrow(comm_catchments)
+# comm_mat %>%
+#   drop_na(ttimes) %>%
+#   left_join(select(comm_data, mdg_cm_, pop)) -> comm_mat
+# comm_mat %>%
+#   left_join(comm_catchments, by = c("scenario" = "scenario")) -> comm_mat
+# comm_mat$CTAR <- as.character(comm_mat$CTAR)
+# comm_data$CTAR <- as.character(comm_data$CTAR)
+# comm_mat %>%
+#   left_join(select(comm_data, mdg_cm_, catch0 = CTAR), by = c("mdg_cm_" = "mdg_cm_")) %>%
+#   mutate(catchment = ifelse(scenario == 0 | scenario == 1, catch0, CTAR)) -> comm_mat
+
+## getting new catchments-> comm_mat ## getting new catchments
+
+# pop <- mada_communes$MDG__201[!is.na(comm_ttimes[1, ])]
+# comm_ttimes <- comm_ttimes[!is.na(comm_ttimes[1, ]), ]
+# comm_ttimes <- as.matrix(comm_ttimes)
+# bites_predicted <- inv.logit(B_ttimes_mada*comm_ttimes/60 + B_0_mada)*pop
+# mu_predicted <- 
+# k_predicted <- 
+# 
+# function(bites, mu, k, times){
+#   for (t in 1:times) {
+#   ## draw bites
+#   
+#   ## do RLEs to get distribution of ts
+#   
+#   ## get # vials used
+#   }
+# }
+
+```

### 2019-03-18 accessPEP.Rmd
with mapping of added pts

@@
 library(geosphere)
+library(iterators)
+library(patchwork)
 
@@
                  "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
+ctar_metadata$fill <- ctar_metadata$color
 ctar_metadata$color[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
-ctar_metadata$fill <- ctar_metadata$color
-ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.4)
+# ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.4)
 # display_palette(ctar_metadata$color)
@@
 
-## Getting district data
+## Getting district covars
+district_coords <- coordinates(mada_district)
+distance_mat <- distm(district_coords, ctar_coords)
+mada_district$dist_min <- apply(distance_mat, 1, min)/1000
 mada_district@data %>%
-  select(mdg_dis_co, district, pop = pop2015un, ttimes_weighted, ctar_in_dist, 
-         dist_catch = ctar_dist, exclude, color, fill) -> dist_data
-
-exps_dist <- get.district.data(ctar_data = ctar_data, ctar_metadata = ctar_metadata, 
-                               mada_district = mada_district, catchments, 
+  select(district = mdg_dis_co, dist_name = district, pop = pop2015adj, 
+         ttimes_weighted_dist = ttimes_weighted, 
+         dist_catch = CTAR, distance = dist_min, exclude, color) -> dist_data
+
+exps <- get.district.data(ctar_data = ctar_data, covars = dist_data, ctar_metadata, 
                                consec_days = 15, 
@@
                                start = ymd("2014-01-01"), end = ymd("2017-12-31"))
+exps %>%
+  group_by(district) %>%
+  summarise(bites = mean(bites), n = n()) %>%
+  left_join(dist_data) -> exps_dist
+
 exps_mora <- get.morabites(mada_communes = mada_communes)
-
-## Getting other data
-rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
-                          end_date = ymd("2017-12-31"), 
-                          threshold = 10)
+exps_mora <- exps_mora[["mora_bites"]]
+## Other data
 ctar_contacts <- get.contacts(ctar_data, times_sd = 3)
 ctar_contacts <- ctar_contacts[["ctar_data"]]
+
 ```
@@
 Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis.
-Overall, we had data from *`r 31 - length(no_data)`* clinics across the country (Fig 1A).
+Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A).
 These data include details of the administrative district of the bite patient's address and the date of reporting.
-We also had *27* months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).
+We also had 27 months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).
 
 <br>
+
+```{r raw data, cache = TRUE}
+## Network fig (make with ggplot!)
+## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments
+
+## get exposure matrix = first set of points
+## district exposures
+ctar_data %>% 
+  drop_na(id_ctar, district) %>%
+  group_by(district) %>% # group by district
+  summarize(n = n()) -> exps_district # get count of exposures
+## ctar exposures
+ctar_data %>% 
+  drop_na(id_ctar) %>%
+  group_by(id_ctar) %>% # group by ctar
+  summarize(n = n()) -> exps_ctar
+## ctar by district exposure matrix
+ctar_data %>% 
+  drop_na(id_ctar, district) %>%
+  group_by(district, id_ctar) %>% # group by district and ctar
+  summarize(n = n()) -> exp_mat# get count of exposures
+
+## To plot district bites
+dist_points <- cbind(select(mada_district@data,
+                            mdg_dis_co), coordinates(mada_district))
+names(dist_points)[2:3] <- c("dist_long", "dist_lat")
+dist_points %>%
+  left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
+            by = c("mdg_dis_co" = "ctar_dist")) %>%
+  mutate(dist_long = coalesce(LONGITUDE, dist_long), 
+         dist_lat = coalesce(LATITUDE, dist_lat)) -> dist_points
+
+exps_district %>%
+  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
+             by = c("district" = "mdg_dis_co")) %>%
+  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
+dist_points$size <- log(dist_points$n + 0.1)*0.25
+
+## To plot ctar points (only those not excluded)
+exps_ctar %>%
+  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE, 
+                   exclude, color, fill)) %>%
+  filter(exclude == 0) -> ctar_points
+
+## To plot lines between ctar + districts
+exp_mat %>% 
+  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
+  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
+                   color, fill, ctar_dist)) -> dist_lines
+dist_lines$size <- log(dist_lines$n + 0.1)*0.25
+
+gg_district <- fortify(mada_district, region = "mdg_dis_co")
+gg_district %>% 
+  left_join(mada_district@data, by = c("id" = "mdg_dis_co")) -> gg_district
+
+cols <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
+                         "#99d8c9", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                         "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
+                         "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
+                         "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2", "#D3D3D3")
+names(cols) <- as.character(cols)
+
+sizes <- log(c(100, 200, 400, 800, 1600) + 0.1)
+p <- ggplot() +
+  geom_polygon(data = gg_district, aes(x = long, y = lat, group = group, fill = color), color = "grey50", alpha = 0.5) +
+  geom_point(data = ctar_points, aes(x = LONGITUDE, y = LATITUDE, size = log(n + 0.1), color = color), shape = 1) +
+  geom_point(data = dist_points, aes(x = dist_long, y = dist_lat, color = fill, size = log(n + 0.1)*0.75)) +
+  geom_segment(data = dist_lines, aes(x = dist_long, y = dist_lat, xend = LONGITUDE, yend = LATITUDE, 
+                                      color = color, size = size*1.5), alpha = 0.75, show.legend = FALSE) +
+  scale_fill_manual(values = cols, guide = "none") +
+  scale_color_manual(values = cols, guide = "none") +
+  scale_size_identity("Reported bites", labels = as.character(c(100, 200, 400, 800, 1600)),
+                      breaks = sizes, guide = "legend") +
+  guides(size = guide_legend(override.aes = list(color = "grey50"))) +
+  labs(tag = "A") +
+  xlab("") +
+  ylab("") +
+  theme(axis.text.x = element_blank(), axis.text.y = element_blank())
+
+### Part 2
+col_vals <- dist_catchments$color
+names(col_vals) <- dist_catchments$mdg_dis_co
+
+c <- ggplot(exps, aes(x = reorder(dist_name, -ttimes_weighted_dist), y = bites/pop*1e5, color = district)) +
+  geom_boxplot() +
+  theme(legend.position = "none") +
+  scale_color_manual(values = col_vals) + 
+  ylab("Annual bites per 100k") +
+  xlab("Districts ordered by \n travel times (high to low)") +
+  coord_flip() +
+  theme(axis.text.y = element_blank()) +
+  labs(tag = "B")
+
+p + plot_spacer() + c + plot_layout(nrow = 1, widths = c(1, 0.1, 1))
+
+```
 
@@
 $$
-  \overline\rho = \frac{\mu \times p_{rabid}}{R}
+  \overline\rho = \frac{\mu p_{rabid}}{R}
 $$
@@
             \begin{cases}
-            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
-            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
+            x, & \text{if}\ \frac{R\rho_{max}}{B} > x \\
+            \frac{R\rho_{max}}{B}, & \text{otherwise}
             \end{cases}
@@
 Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden.
-We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population.
+We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR, see SN) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population.
 
@@
                                 model = mapply(function(x) attributes(x)$label, mada_mods)))
+kable(bite_mods)
 ## Predictions: fixed
@@
 
-library(iterators)
 preds_fixed <- run.predicts.fixed(covar_df = bite_mods, pop_fixed = 1e5, pop_seq = pop_plot)
@@
 
-library(patchwork)
 a <- ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
@@
   geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
-  scale_color_manual(values = c("navy", "blue", "purple")) +
+  scale_color_manual(values = c("navy", "blue", "purple"), guide = "none") +
   scale_fill_manual(values = c("blue", "lightblue", "pink"), name = "Model scale") +
@@
 All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the model.
-
-The best fitting model did not include population as a predictor for the district data...write up from supplement!
 
@@
 
-ggplot(rep_min_max, aes(x = ttimes, y = rep_min, fill = scale)) +
+c <- ggplot(rep_min_max, aes(x = ttimes, y = rep_min, fill = scale)) +
   geom_line(aes(y = rep_min, color = scale), linetype = 2) +
@@
   geom_ribbon(aes(ymax = rep_max, ymin = rep_min), alpha = 0.5) +
-  scale_fill_manual(values = c("navy", "blue")) +
-  scale_color_manual(values = c("navy", "blue"))
+  scale_fill_manual(values = c("navy", "purple")) +
+  scale_color_manual(values = c("navy", "purple")) +
+  xlab("Reporting") +
+  ylab("Travel times (in hours)") +
+  labs(tag = "A")
 
@@
 
-system.time({
 results_comm_med <- get.burden.stochastic(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes/60,
@@
 results_all %>% filter(scenario == 0) -> results_all_base
-ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = deaths_mean, 
+ggplot(filter(results_all_base, !is.na(district)), 
+       aes(y = deaths_mean, x = reorder(district, ttimes_weighted), 
                               color = model)) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), fatten = 2) +
-  ylab("") +
-  xlab("Average annual deaths") +
+  geom_pointrange(aes(ymin = deaths_lowerCI, ymax = deaths_upperCI), fatten = 2) +
+  xlab("Districts ordered by travel times \n (low to high)") +
+  ylab("Average annual deaths") +
   scale_color_manual(values = c("district" = alpha("blue", 0.5), 
                                 "commune" = alpha("purple", 0.5)), 
-                     name = "Scale") -> p
-p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+                     name = "Scale") +
+  theme(axis.text.x = element_blank()) +
+  labs(tag = "B") -> d
+
+c / d
 
 ## Deaths averted
-ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = averted_mean, 
+ggplot(filter(results_all_base, !is.na(district)), 
+       aes(y = reorder(district, ttimes_weighted), x = averted_mean, 
                               color = model)) +
@@
   ylab("") +
-  xlab("Average annual deaths") +
+  xlab("Average annual deaths averted") +
   scale_color_manual(values = c("district" = alpha("blue", 0.5), 
                                 "commune" = alpha("purple", 0.5)), 
-                     name = "Scale") -> p
-p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+                     name = "Scale") +
+  theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
 
@@
 
-``` {r scenario}
+``` {r scenario, cache = TRUE}
+##
+ctar_pts <- read.csv("output/incremental_ARMC_nofilter.csv")
+ctar_pts$order <- 1:nrow(ctar_pts)
+ggplot() + geom_polygon(data = gg_district, aes(x = long, y = lat, group = group), fill = "grey",
+                         color = "grey50", alpha = 0.5) +
+  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "black", shape = 1) +
+  geom_point(data = ctar_pts, aes(x = Y_COORD, y = X_COORD, color = order, alpha = order))
+
 ## Plotting national level
@@
   guides(color = guide_legend(), 
-         fill = guide_legend(override.aes = list(linetype = 1)))
+         fill = guide_legend(override.aes = list(linetype = 1))) +
+  xlab("# of Additional ARMC") +
+  ylab("Mean annual deaths \n due to rabies") +
+  labs(tag = "A") -> e
 
@@
 ## Deaths averted
-ggplot(data = results_natl, aes (x = scenario, y = averted_mean, color = model)) +
+ggplot(data = results_natl, aes (x = scenario, y = averted_mean*0.16, color = model)) +
   geom_line() +
@@
   scale_color_manual(values = c("district" = "blue", "commune" = "purple"),  guide = "none") +
-  scale_fill_manual(values = c("district" = "blue", "commune" = "purple")) +
-  guides(color = guide_legend(), fill = guide_legend(override.aes = list(linetype = 1)))
+  scale_fill_manual(values = c("district" = "blue", "commune" = "purple"), guide = "none") +
+  xlab("# of Additional ARMC") +
+  ylab("Mean annual \n deaths averted") +
+  labs(tag = "B") -> f
+
+e / f
 
@@
 
-```{r vial analysis}
+```{r vial analysis, cache = TRUE}
 # ## new ctar
@@
                           threshold = 10)
-date_mat <- rep_data[["date_mat"]]
+date_mat <- rep_data[["check_mat"]]
 dose_mat <- rep_data[["doses_wide"]]
@@
   xlab("# of Patients") +
-  ylab("Proportion of days")
+  ylab("Proportion of days") -> a
 
@@
                                predicted_k = exp(0.022*rep_bites/100 + -0.39)))
-ggplot(predicts, aes(x = rep_bites, y = predicted_mu)) +
-  geom_line()
-ggplot(predicts, aes(x = rep_bites, y = predicted_k)) +
-  geom_line()
+ggplot(predicts, aes(x = rep_bites)) +
+  geom_line(aes(y = predicted_mu), color = "red") +
+  geom_line(aes(y = predicted_k), color = "blue") +
+  xlab("Annual reported bites to a clinic") +
+  ylab("Predicted parameter") -> b
+
+a / b
 
@@
 
-### Key findings
-
-### Strengths and Limitations
-
-### Broader context
-
-### Conclusions
+Main findings
+- [ ] Travel times are a good predictor of bites, suggesting that this could shape burden significantly
+- [ ] We develop a framework to look at the potential impact of this access on burden, to get more realistic sub-national and national estimates
+- [ ] Scenario analysis shows that expanding access could significantly reduce burden
+- [ ] However due to broad remoteness, expanding access alone will not eliminate human rabies deaths
+
+Limitations
+- [ ] Not thinking about clinic functioning
+- [ ] Reporting issues and what it means (i.e. we're assuming that clinics are not reporting, but they may not be provisioning at all, etc.) 
+- [ ] Not accounting for other factors which drive access/reporting (i.e. socioeconomics and awareness although these are likely correlated and that has implications for the scenario analysis)
+- [ ] Just another set of assumptions and a set that needs validation--what data could do this?
+- [ ] Contacts issue and what it means for the results
+- [ ] Approach only works in endemic setting, and where reporting = receiving (i.e. in a free setting), although for GAVI countries this is likely the case
+
+Discussion: broader context
+- [ ] Compare to previous estimates of burden (Moramanga, GAVI paper, etc.) for Mada
+- [ ] Compare range of estimates of human rabies exposure incidence and deaths to estimates from other endemic settings
+- [ ] P_rabid estimates compared to other endemic settings (look at Moramanga data in Supplementary materials?) Differences in reporting of rabid vs. non-rabid bites?
+- [ ] Availability of PEP in other countries is also similarly or even more limited
+- [ ] Other studies that have looked at access and mortality/vacc
+
+Discussion: conclusions
+- [ ] Access likely shapes burden in a big way in Mada
+- [ ] Expanding access will therefore likely decrease burden in a big way, but there are limitations due to infrastructure, etc.
+- [ ] Spatial differences in reporting also has implications for surveillance (and could parallel dog vaccination heterogeneity) and this needs to be explored *hint*hint*
+- [ ] Our results depend on strong assumptions, so key data on dog pops, rabies, reporting, etc. are needed to validate this, surveillance @ clinic level!
+- [ ] Regardless expanding access alone will likely not prevent all deaths, you need dog vacc!
 
@@
 ## Supplementary Materials
-### S1. RLE thresholds for reporting
-
-### S2. Removing contacts from the data set
+
+### S1. Estimates of reporting
+```{r}
+## Getting other data
+rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
+                          end_date = ymd("2017-12-31"), 
+                          threshold = 15)
+clinic_reporting <- rep_data[["clinic_reporting"]]
+patient_ts <- rep_data[["patient_ts"]]
+date_mat <- rep_data[["date_mat"]]
+dose_mat <- rep_data[["dose_mat"]]
+doses_wide <- rep_data[["doses_wide"]]
+
+patient_ts %>%
+  ungroup() %>%
+  mutate(id_ctar = as.factor(id_ctar)) %>%
+  right_join(date_mat) %>%
+  mutate(n = replace_na(n, 0), 
+         n = ifelse(exclude == 1, n, NA)) -> doses_long
+doses_long$ctar <- ctar_metadata$CTAR[match(doses_long$id_ctar, ctar_metadata$id_ctar)]
+doses_long %>%
+  mutate(year = year(date_de_consultation)) %>%
+  left_join(select(clinic_reporting, ctar, prop, year = date_de_consultation)) -> doses_long
+
+p <- ggplot(data = doses_long, aes(x = date_de_consultation, y = reorder(ctar, prop))) + 
+  geom_tile(aes(fill = n)) +
+  scale_fill_gradientn(colours = c("white", "purple", "black"), values = c(0, 0.1, 1),
+                       na.value = alpha("lightgrey", 1), name = "Number of \n patients") +
+  xlim(ymd("2014-01-01"), ymd("2017-12-31")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  xlab("Year") +
+  ylab("ARMC") +
+  labs(tag = "A")
+
+col_vals <- ctar_metadata$fill
+names(col_vals) <- ctar_metadata$CTAR
+
+n <- clinic_reporting %>%
+  mutate(prop15 = prop, prop5 = get.days(doses_wide = doses_wide, threshold = 5)[[1]]$prop,
+         prop30 = get.days(doses_wide = doses_wide, threshold = 30)[[1]]$prop) %>%
+  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
+  geom_hline(yintercept = 0.25, color = "darkgrey", alpha = 0.5) +
+  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
+               stat = "identity") +
+  scale_color_manual(values = col_vals) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
+        legend.position = "none") +
+  xlab("") +
+  ylab("Proportion of days included") +
+  coord_flip() +
+  labs(tag = "B")
+
+p + n
+
+```
+
+### S2. Accounting for low-risk contacts reporting to clinics
+
+```{r excluding contacts}
+ctar_contacts <- get.contacts(ctar_data, times_sd = 3)
+patient_ts <- ctar_contacts[["patient_ts"]]
+patient_ts <- patient_ts[patient_ts$ctar != "Tsiroanomandidy", ]
+A <- ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
+  scale_color_manual(name = "Exclude", values = c("navy", "red"), labels = c("No", "Yes")) +
+  ylab("Number of patients per day") +
+  xlab("ARMC") +
+  geom_jitter(alpha = 0.5, width = 0.25) +
+  coord_flip() +
+  labs(tag = "A")
+
+## Testing with Moramanga data 
+ctar <- get.morabites(mada_communes = mada_communes)
+ctar <- ctar[["ctar"]]
+sdev <- seq(2, 10, by = 0.1)
+check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
+for (i in 1:length(sdev)){
+  ctar %>%
+    group_by(date.reported) %>%
+    summarize(n = n()) %>%
+    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
+  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
+  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
+  check <- rbind(check, c(sdev[i], checksd[1:2]))
+}
+check %>% drop_na() %>% gather(prop_bites, prop_contacts, -sd) -> check
+B <- ggplot(check, aes(x = sd, y = prop_contacts, color = prop_bites)) +
+  geom_line() +
+  geom_line() +
+  scale_color_manual(name = "Type of patient", values = c("blue", "red"), 
+                     labels = c("Bites/scratches", "Low-risk contacts")) + 
+  xlab("Number of standard \n deviations above the mean") +
+  ylab("Proportion excluded") +
+  labs(tag = "B")
+
+A + B
+
+```
 
@@
 ggpairs(data = select(comm_covars, pop, ttimes_weighted, distance))
-
-ggpairs(data = select(exps_mora, pop, ttimes_weighted, distance))
 
@@
 
-kable(AICctab(mada_mods, nobs = nrow(exps_dist), mnames = mapply(function(x) attributes(x)$label, mada_mods), weights = TRUE))
+print(AICctab(mada_mods, nobs = nrow(exps_dist), mnames = mapply(function(x) attributes(x)$label, mada_mods), weights = TRUE))
 
 mora_mods <- c(mora_mods_covars, mora_mods_pop)
-kable(AICctab(mora_mods, nobs = nrow(exps_mora), mnames = mapply(function(x) attributes(x)$label, mora_mods), weights = TRUE))
+print(AICctab(mora_mods, nobs = nrow(exps_mora), mnames = mapply(function(x) attributes(x)$label, mora_mods), weights = TRUE))
 
@@
 check <- bind_rows(bite_mods_covar_df, bite_mods_pop_df)
+kable(check)
 
@@
 
-### S4. Sensitivity of model parameter estimates to contact and RLE cut-offs
+### S4. Sensitivity of parameter estimation to different reporting estimates and exclusion of contacts
 
@@
     foreach(k = 1:length(contact_cutoff), .combine = combine, .multicombine = TRUE) %do% {
-    exps_dist <- get.district.data(ctar_data = ctar_data, ctar_metadata = ctar_metadata, 
-                               mada_district = mada_district, catchments, 
+    exps_dist <- get.district.data(ctar_data = ctar_data, covars = dist_data, ctar_metadata, 
                                consec_days = consec_days[i], 
@@
 
-### S5. Sensitivity of burden estimates to paramater values (multivariate for R, p_rabid, rho_max)
-
-```{r sensitivity analysis multivariate, cache = TRUE}
-hdr_vals <- c(25, 20, 15, 10, 5)
-p_rabid_vals <- c(0.2, 0.3, 0.4, 0.5, 0.6)
-rho_max_vals <- c(0.80, 0.85, 0.9, 0.95, 0.9)
-library(foreach)
-
+### S5. Sensitity analyses for burden estimates
+
+#### District
+```{r scaling w/ pop, cache = TRUE}
+## Dist ttimes
 dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
@@
 
-scenario_dist <- 
-  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
-  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
-  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
- get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
-                     ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
-                     B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
-                     hdr = hdr_vals[k], p_rabid = p_rabid_vals[i], 
-                     rho_max = rho_max_vals[j], type = "burden", scale = FALSE)
-  }
-
-scenario_dist %>%
-  filter(scenario == 0) %>%
-  group_by(hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
-ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
-  geom_point() +
-  facet_grid(~ hdr)
-ggplot(data = filter(scenario_dist, scenario == 0), aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
-  geom_point() +
-  facet_grid(rho_max ~ hdr)
-
-scenario_dist %>%
-  group_by(scenario, hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths)) -> deaths_scenario
-deaths_scenario %>%
-  left_join(select(deaths_multivar, hdr, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
-ggplot(data = deaths_scenario, aes(x = scenario, y = deaths, color = p_rabid)) +
-  geom_point() +
-  facet_grid(hdr ~ rho_max, scales = "free_y")
-
-comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
-comm_ttimes <- as.data.table(comm_ttimes)
-comm_ttimes$mdg_cm_ <- as.character(mada_communes$mdg_cm_)
-comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
-comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
-comm_ttimes <- rbind(comm_ttimes, 
-                       as.data.frame(list(mdg_cm_ = comm_catchments$mdg_cm_, scenario = 0,
-                                          ttimes = comm_catchments$ttimes_weighted)))
-comm_ttimes %>%
-  left_join(select(mada_communes@data, mdg_cm_, commune, pop = MDG__201, mtch_ds)) -> comm_ttimes
-
-scenario_comm <- 
-  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
-  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
-  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
-                     ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
-                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada, 
-                     hdr = hdr_vals[k], p_rabid = p_rabid_vals[i], 
-                     rho_max = rho_max_vals[j], type = "burden", scale = FALSE)
-  }
-
-scenario_comm %>%
-  filter(scenario == 0) %>%
-  group_by(hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE), scenario = 0) -> deaths_multivar
-
-ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
-  geom_point() +
-  facet_grid(~ hdr)
-ggplot(data = filter(scenario_comm, scenario == 0), aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
-  geom_point() +
-  facet_grid(rho_max ~ hdr)
-
-hdr_name <- c(`5` = paste0(1e5/5*0.01*0.39, " per \n 100k"),
-                    `10` = paste0(1e5/10*0.01*0.39, " per \n 100k"),
-                    `15` = paste0(1e5/15*0.01*0.39, " per \n 100k"),
-                    `20` = paste0(1e5/20*0.01*0.39, " per \n 100k"),
-                    `25` = paste0(1e5/25*0.01*0.39, " per \n 100k"))
-
-scenario_comm %>%
-  group_by(scenario, hdr, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_scenario
-deaths_scenario %>%
-  left_join(select(deaths_multivar, hdr, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
-ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
-  geom_point() +
-  facet_grid(hdr ~ rho_max, scales = "free_y", labeller = labeller(hdr = as_labeller(hdr_name)))
-
-```
-
-### S6. Sensitity of scenario analyses to parameter values (multivariate for R, p_rabid, rho_max)
-
-```{r scaling w/ pop, cache = TRUE}
 ## Pop scaling
@@
   geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey")
-
 ## getting scaling factors 
@@
 
+```
+
+#### Commune
+
+```{r commune scaling, cache = TRUE}
 ## Commune
+
+## Commune ttimes
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
+comm_ttimes <- as.data.table(comm_ttimes)
+comm_ttimes$mdg_cm_ <- as.character(mada_communes$mdg_cm_)
+comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
+comm_ttimes <- rbind(comm_ttimes, 
+                       as.data.frame(list(mdg_cm_ = comm_catchments$mdg_cm_, scenario = 0,
+                                          ttimes = comm_catchments$ttimes_weighted)))
+comm_ttimes %>%
+  left_join(select(comm_data, mdg_cm_, mdg_dis_co, pop)) -> comm_ttimes
+
 pop <- mada_communes$MDG__201
@@
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
-  check <- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
+  check <- get.burden.fixed(names = comm_ttimes$mdg_dis_co, pop = comm_ttimes$pop, 
                      ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
@@
                      slope = scale_df_comm$sf[k], intercept = scale_df_comm$intercept[k])
+  check %>% 
+    group_by(names, scenario, p_rabid, rho_max, slope) %>%
+    summarize(ttimes = mean(ttimes, na.rm = TRUE), pop = sum(pop, na.rm = TRUE),
+              deaths = sum(deaths, na.rm = TRUE)) -> check
   check$scale <- scale_df_comm$scale[k]
@@
 
-### Scale example
+#### Difference between scale examples
 

### 2019-03-19 accessPEP.Rmd
some preliminary code for vial estimates

@@
 knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
-                      results = "hide",
                       include = TRUE)
@@
 for (i in 2:ncol(dist_ttimes)){
+  i = 3
   dist_mat[, i] <- ifelse(dist_ttimes[, i] < dist_ttimes[, i - 1], dist_ttimes[, i], NA)
@@
                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 1000)
+                           prob_death = 0.16, nsims = 2)
 
@@
 ``` {r scenario, cache = TRUE}
-##
-ctar_pts <- read.csv("output/incremental_ARMC_nofilter.csv")
-ctar_pts$order <- 1:nrow(ctar_pts)
-ggplot() + geom_polygon(data = gg_district, aes(x = long, y = lat, group = group), fill = "grey",
-                         color = "grey50", alpha = 0.5) +
-  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "black", shape = 1) +
-  geom_point(data = ctar_pts, aes(x = Y_COORD, y = X_COORD, color = order, alpha = order))
-
 ## Plotting national level
@@
 
-ggplot(data = results_all, aes(x = averted_mean, y = reorder(district, ttimes_weighted), 
-                               color = scenario)) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
-                             fatten = 2) +
-  facet_wrap(~model)
-
 ```
@@
 We use this to estimate how vaccine demand will change as access to PEP is expanded...
+```{r simulate vials used}
+## Get catchment
+catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
+mada_communes@data %>%
+  left_join(comm_data) -> comm_catches
+comm_ttimes <- cbind(comm_ttimes[, 1], comm_ttimes)
+names(comm_ttimes)[1] <- "result.0"
+comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
+comm_mat[, 1] <- as.character(comm_catches$CTAR)
+
+for (i in 2:ncol(comm_ttimes)){
+  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
+                          comm_mat[, 1])
+}
+
+comm_mat <- as.data.table(comm_mat)
+comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
+comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
+comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+comm_ttimes <- as.data.table(comm_ttimes)
+comm_ttimes$mdg_cm_ <- mada_communes$mdg_cm_
+comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
+
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
+  drop_na(catchment) -> comm_master
+
+sim.throughput <- function(bites, nsims = 100) {
+  get.vials <- function(x) {
+    day0 <- round(runif(rpois(1, x), min = 1, max = 365))
+    days <- c(day0, day0 + 3, day0 + 7)
+    vials <- sum(ceiling(table(days)/2))
+    return(vials)
+  }
+
+  check <- foreach(i = 1:nsims, .combine = cbind) %dopar% {
+    vials <- sapply(bites, get.vials, simplify = TRUE)
+  }
+
+  mean_vials <- rowMeans(check, na.rm = TRUE)
+  sd_vials <- apply(check, 1, sd, na.rm = TRUE)
+  return(as.data.frame(list(bites = bites, mean_vials = mean_vials, sd_vials = sd_vials)))
+}
+
+## local cluster
+library(doParallel)
+# Calculate the number of cores
+no_cores <- detectCores() - 1
+# Initiate cluster
+cl <- makeCluster(no_cores)
+registerDoParallel(cl)
+system.time({throughput <- sim.throughput(bites = comm_master$bites, nsims = 10)})
+stopCluster(cl) ## for doParallel
+
+comm_throughput <- bind_cols(comm_master, throughput)
+
+```
+
+```{r throughput district}
+## Get catchment
+catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
+dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
+dist_ttimes <- cbind(dist_ttimes[, 1], dist_ttimes)
+names(dist_ttimes)[1] <- "result.0"
+dist_mat <- matrix(NA, nrow(dist_ttimes), ncol(dist_ttimes))
+dist_mat[, 1] <- as.character(mada_district$CTAR)
+
+for (i in 2:ncol(dist_ttimes)){
+  dist_mat[, i] <- ifelse(dist_ttimes[, i] < dist_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
+                          dist_mat[, 1])
+}
+
+dist_mat <- as.data.table(dist_mat)
+dist_mat$mdg_dis_co <- mada_district$mdg_dis_co
+dist_mat <- melt(dist_mat, id.vars = "mdg_dis_co", value.name = "catchment", variable.name = "scenario")
+dist_mat$scenario <- as.numeric(gsub("V", "", dist_mat$scenario)) - 1
+dist_ttimes <- as.data.table(dist_ttimes)
+dist_ttimes$mdg_dis_co <- mada_district$mdg_dis_co
+dist_ttimes <- melt(dist_ttimes, id.vars = "mdg_dis_co", value.name = "ttimes", variable.name = "scenario")
+dist_ttimes$scenario <- as.numeric(gsub("result.", "", dist_ttimes$scenario))
+
+dist_mat %>%
+  left_join(dist_ttimes) %>%
+  left_join(select(mada_district@data, pop = pop2015adj, mdg_dis_co)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
+  drop_na(catchment) -> dist_master
+
+## local cluster
+library(doParallel)
+# Calculate the number of cores
+no_cores <- detectCores() - 1
+# Initiate cluster
+cl <- makeCluster(no_cores)
+registerDoParallel(cl)
+system.time({throughput <- sim.throughput(bites = dist_master$bites, nsims = 10)})
+stopCluster(cl) ## for doParallel
+
+dist_throughput <- bind_cols(dist_master, throughput)
+
+```
 
@@
 ### S5. Sensitity analyses for burden estimates
-
 #### District
@@
 scale_df_dist %>%
-  merge(., select(mada_district@data, mdg_dis_co, ttimes_weighted, pop = pop2015adj)) -> scale_df_check
-ggplot(data = scale_df_check, aes(x = log(pop), 
+  merge(., select(mada_district@data, mdg_dis_co, ttimes_weighted, pop = pop2015adj)) -> scale_df_check_dist
+
+A_dist <- ggplot(data = scale_df_check_dist, aes(x = log(pop), 
                             y = constrained_inc(sf, intercept, pop, inc100k_max, inc100k_min)*1e5, 
@@
   geom_point() +
-  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey")
-## getting scaling factors 
-scenario_scaled <- 
+  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey") +
+  scale_color_gradient2(low = "red", mid = "grey50", high = "blue", midpoint = 0, name = "Scaling Factor") +
+  ylab("Rabies exposures per \n 100k persons") +
+  xlab("Human population size (log)")
+
+system.time({
+scenario_natl_dist <- 
   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
@@
                      ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
-                     B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
+                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
+                     slope = scale_df_dist$sf[k], intercept = scale_df_dist$intercept[k])
+  check$scale <- scale_df_dist$scale[k]
+  check %>% 
+    group_by(p_rabid, rho_max, slope, scale, scenario) %>% 
+    summarize(deaths = sum(deaths, na.rm = TRUE)) -> check
+  check
+  }
+})
+
+dist_base <- filter(dist_ttimes, scenario == 0)
+
+system.time({
+scenario_base_dist <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df_dist), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+  check <- get.burden.fixed(names = dist_base$mdg_dis_co, pop = dist_base$pop, 
+                     ttimes = dist_base$ttimes/60, scenario = dist_base$scenario,
+                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
                      p_rabid = p_rabid_vals[i], 
@@
   }
-
-scenario_scaled %>%
-  mutate(slope = case_when(slope == 0 & scale =="neg" ~ -10, 
+})
+
+scenario_natl_dist %>%
+  mutate(slope_order = case_when(slope == 0 & scale =="neg" ~ -10, 
                            slope == 0 & scale =="pos" ~ 10,
-                           slope > 0 | slope < 0 ~ round(slope*1e10))) -> scenario_scaled_dist
+                           slope > 0 | slope < 0 ~ round(slope*1e10)), 
+         slope_order = round(1/slope_order, 2), 
+         slope_order = ifelse(slope_order == -0.1, -Inf, 
+                                 ifelse(slope_order == 0.1, Inf, slope_order))) -> scenario_scaled_dist
+scenario_base_dist %>%
+   mutate(slope_order = case_when(slope == 0 & scale =="neg" ~ -10, 
+                           slope == 0 & scale =="pos" ~ 10,
+                           slope > 0 | slope < 0 ~ round(slope*1e10)), 
+         slope_order = round(1/slope_order, 2), 
+         slope_order = ifelse(slope_order == -0.1, -Inf, 
+                                 ifelse(slope_order == 0.1, Inf, slope_order))) -> scenario_base_dist
+scenario_base_dist %>%
+  group_by(slope_order, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+
+slope_name_dist <- c(`-0.11` = "----", `-0.14` = "---", `-0.25` = "--", `-0.5` = "-",
+                `-Inf` =  "max", `0.11` = "++++", `0.14` = "+++", `0.25` = "++", 
+                `0.5` = "+",`Inf` =  "min")
+
+B_dist <- ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ slope_order, labeller = labeller(slope_order = as_labeller(slope_name_dist))) +
+  xlab("Prop. of reported \n bites assumed \n to be rabid") +
+  ylab("Burden of rabies deaths \n at national level") +
+  scale_x_continuous(breaks = c(0.2, 0.6)) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  scale_color_gradient(name = "Maximum \n reporting")
+
+C_dist <- ggplot(data = scenario_base_dist,  
+       aes(x = ttimes, y = deaths/pop*1e5, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ slope_order, labeller = labeller(slope_order = as_labeller(slope_name_dist))) +
+  scale_color_gradient(name = "Proportion of reported \n bites assumed \n to be rabid") + 
+  ylab("Rabies deaths \n per 100k persons") +
+  xlab("Travel times (hours)") +
+  scale_x_continuous(breaks = c(0, 6, 12)) + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
 scenario_scaled_dist %>%
   filter(scenario == 0) %>%
-  group_by(slope, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
-
-slope_name <- c(`-9` = "----", `-7` = "---", `-4` = "--", `-2` = "--",
-                    `-10` =  "max", `9` = "++++", `7` = "+++", `4` = "++", `2` = "+",`10` =  "min")
-
-ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  mutate(deaths_base = deaths) %>%
+  select(-deaths, -scenario) %>%
+  right_join(scenario_scaled_dist) -> deaths_scenario_dist
+
+D_dist <- ggplot(data = deaths_scenario_dist, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
   geom_point() +
-  facet_grid(~ slope, labeller = labeller(slope = as_labeller(slope_name)))
-ggplot(data = filter(scenario_scaled_dist, scenario == 0),  
-       aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
-  geom_point() +
-  facet_grid(rho_max ~ slope, labeller = labeller(slope = as_labeller(slope_name)))
-
-scenario_scaled_dist %>%
-  group_by(scenario, slope, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths)) -> deaths_scenario
-deaths_scenario %>%
-  left_join(select(deaths_multivar, slope, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
-ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
-  geom_point() +
-  facet_grid(slope ~ rho_max, scales = "free_y")
+  facet_grid(rho_max ~ slope_order, 
+             labeller = labeller(slope_order = as_labeller(slope_name_dist))) +
+  scale_color_gradient(name = "Proportion of reported \n bites assumed \n to be rabid") + 
+  scale_x_continuous(breaks = c(0, 250, 500)) +
+  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) + 
+  xlab("# Additional ARMC") +
+  ylab("Proportion of baseline burden") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
 ## Commune
-
 ## Commune ttimes
@@
                                      intercept = inc100k_min)))
+
 constrained_inc <- function(slope, intercept, pop, max, min){
@@
 
-scale_df_comm <- rbind(as.data.frame(list(scale = "neg", 
-                                     sf = seq(neg$coefficients[2], 0, length.out = 5),
-                                     intercept = inc100k_max)),
-                  as.data.frame(list(scale = "pos", 
-                                     sf = seq(0, pos$coefficients[2], length.out = 5),
-                                     intercept = inc100k_min)))
-constrained_inc <- function(slope, intercept, pop, max, min){
-  inc <- slope*pop + intercept
-  inc[inc >= max] <- max
-  inc[inc <= min] <- min
-  return(inc)
-}
 scale_df_comm %>%
   merge(., select(mada_communes@data, mdg_cm_, ttms_ll, pop = MDG__201)) -> scale_df_check_comm
-ggplot(data = scale_df_check_comm, aes(x = log(pop), 
+
+scale_df_check_comm$admin <- "commune"
+scale_df_check_dist$admin <- "district"
+scale_df_check <- bind_rows(scale_df_check_dist, scale_df_check_comm)
+
+A_comm <- ggplot(data = scale_df_check_comm, aes(x = log(pop), 
                             y = constrained_inc(sf, intercept, pop, inc100k_max, inc100k_min)*1e5, 
@@
   geom_point() +
-  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey")
-
-scenario_scaled <- 
+  geom_hline(yintercept = c(inc100k_max*1e5, inc100k_min*1e5), linetype = 2, color = "grey") +
+  scale_color_gradient2(low = "red", mid = "grey50", high = "blue", midpoint = 0, name = "Scaling Factor", 
+                        guide = "none") +
+  ylab("Rabies exposures per \n 100k persons") +
+  xlab("Human population size (log)")
+
+system.time({
+scenario_natl_comm <- 
   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
@@
   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
-  check <- get.burden.fixed(names = comm_ttimes$mdg_dis_co, pop = comm_ttimes$pop, 
+  check <- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
                      ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
@@
                      slope = scale_df_comm$sf[k], intercept = scale_df_comm$intercept[k])
+  check$scale <- scale_df_comm$scale[k]
   check %>% 
-    group_by(names, scenario, p_rabid, rho_max, slope) %>%
-    summarize(ttimes = mean(ttimes, na.rm = TRUE), pop = sum(pop, na.rm = TRUE),
-              deaths = sum(deaths, na.rm = TRUE)) -> check
+    group_by(p_rabid, rho_max, slope, scale, scenario) %>% 
+    summarize(deaths = sum(deaths, na.rm = TRUE)) -> check
+  check
+  }
+})
+
+comm_base <- filter(comm_ttimes, scenario == 0)
+
+system.time({
+scenario_base_comm <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
+  foreach(k = 1:nrow(scale_df_comm), .combine = rbind) %:%
+  foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
+  check <- get.burden.fixed(names = comm_base$mdg_cm_, pop = comm_base$pop, 
+                     ttimes = comm_base$ttimes/60, scenario = comm_base$scenario,
+                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     p_rabid = p_rabid_vals[i], 
+                     rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
+                     slope = scale_df_comm$sf[k], intercept = scale_df_comm$intercept[k])
   check$scale <- scale_df_comm$scale[k]
@@
   }
-
-scenario_scaled %>%
-  mutate(slope = case_when(slope == 0 & scale =="neg" ~ -10, 
+})
+
+scenario_natl_comm %>%
+  mutate(slope_order = case_when(slope == 0 & scale =="neg" ~ -10, 
                            slope == 0 & scale =="pos" ~ 10,
-                           slope > 0 | slope < 0 ~ round(slope*1e10))) -> scenario_scaled_comm
+                           slope > 0 | slope < 0 ~ round(slope*1e9, 2)), 
+         slope_order = round(1/slope_order, 2), 
+         slope_order = ifelse(slope_order == -0.1, -Inf, 
+                                 ifelse(slope_order == 0.1, Inf, slope_order))) -> scenario_scaled_comm
+scenario_base_comm %>%
+   mutate(slope_order = case_when(slope == 0 & scale =="neg" ~ -10, 
+                           slope == 0 & scale =="pos" ~ 10,
+                           slope > 0 | slope < 0 ~ round(slope*1e9, 2)), 
+         slope_order = round(1/slope_order, 2), 
+         slope_order = ifelse(slope_order == -0.1, -Inf, 
+                                 ifelse(slope_order == 0.1, Inf, slope_order))) -> scenario_base_comm
+scenario_base_comm %>%
+  group_by(slope_order, p_rabid, rho_max) %>%
+  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
+
+slope_name_comm <- c(`-0.19` = "----", `-0.25` = "---", `-0.37` = "--", `-0.75` = "-",
+                `-Inf` =  "max", `0.19` = "++++", `0.25` = "+++", `0.37` = "++", 
+                `0.75` = "+",`Inf` =  "min")
+
+B_comm <- ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  geom_point() +
+  facet_grid(~ slope_order, labeller = labeller(slope_order = as_labeller(slope_name_comm))) +
+  xlab("Prop. of reported \n bites assumed \n to be rabid") +
+  ylab("Burden of rabies deaths \n at national level") +
+  scale_x_continuous(breaks = c(0.2, 0.6)) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  scale_color_gradient(name = "Maximum \n reporting", guide = "none")
+
+C_comm <- ggplot(data = scenario_base_comm,  
+       aes(x = ttimes, y = deaths/pop*1e5, color = p_rabid)) +
+  geom_point() +
+  facet_grid(rho_max ~ slope_order, labeller = labeller(slope_order = as_labeller(slope_name_comm))) +
+  scale_color_gradient(name = "Proportion of reported \n bites assumed \n to be rabid", guide = "none") + 
+  ylab("Rabies deaths \n per 100k persons") +
+  xlab("Travel times (hours)") +
+  scale_x_continuous(breaks = c(0, 6, 12)) + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
   filter(scenario == 0) %>%
-  group_by(slope, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths, na.rm = TRUE)) -> deaths_multivar
-
-slope_name <- c(`-9` = "----", `-7` = "---", `-4` = "--", `-2` = "--",
-                    `-10` =  "max", `9` = "++++", `7` = "+++", `4` = "++", `2` = "+",`10` =  "min")
-
-ggplot(data = deaths_multivar, aes(x = p_rabid, y = deaths, color = rho_max)) +
+  mutate(deaths_base = deaths) %>%
+  select(-deaths, -scenario) %>%
+  right_join(scenario_scaled_comm) -> deaths_scenario_comm
+
+D_comm <- ggplot(data = deaths_scenario_comm, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
   geom_point() +
-  facet_grid(~ slope, labeller = labeller(slope = as_labeller(slope_name)))
-ggplot(data = filter(scenario_scaled_comm, scenario == 0),  
-       aes(x = ttimes, y = deaths/pop, color = p_rabid)) +
-  geom_point() +
-  facet_grid(rho_max ~ slope, labeller = labeller(slope = as_labeller(slope_name)))
-
-scenario_scaled_comm %>%
-  group_by(scenario, slope, p_rabid, rho_max) %>%
-  summarize(deaths = sum(deaths)) -> deaths_scenario
-deaths_scenario %>%
-  left_join(select(deaths_multivar, slope, p_rabid, rho_max, deaths_base = deaths)) -> deaths_scenario
-ggplot(data = deaths_scenario, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
-  geom_point() +
-  facet_grid(slope ~ rho_max, scales = "free_y", labeller = labeller(slope = as_labeller(slope_name)))
+  facet_grid(rho_max ~ slope_order, 
+             labeller = labeller(slope_order = as_labeller(slope_name_comm))) +
+  scale_color_gradient(name = "Proportion of reported \n bites assumed \n to be rabid", guide = "none") + 
+  scale_x_continuous(breaks = c(0, 250, 500)) +
+  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) + 
+  xlab("# Additional ARMC") +
+  ylab("Proportion of baseline burden") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# S5.2 <- (D + A)/(B + C)
+# ggsave("figs/S5.2.jpeg", device = "jpeg", plot = S5.2, width = 10, height = 10)
+
+```
+
+```{r scaling factor}
+A_dist / A_comm
+```
+
+```{r differences with ttimes}
+C_dist / C_comm
+```
+
+```{r national burden baseline}
+B_dist / B_comm
+
+```
+
+```{r national burden scenario}
+D_dist / D_comm
 
@@
 
-```{r scenario analysis}
-# ## new ctar
-# expanded <- read.csv("output/incremental_ARMC_nofilter.csv")
-# dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
-# comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
-# 
-# 
-# scenario_scaled <- 
-#   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:%
-#   foreach(k = 1:nrow(scale_df), .combine = rbind) %:%
-#   foreach(j = 1:length(rho_max_vals), .combine = rbind) %do% {
-#   get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
-#                      ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
-#                      B_ttimes = B_ttimes_dist, B_0 = B_0_dist, 
-#                      p_rabid = p_rabid_vals[i], 
-#                      rho_max = rho_max_vals[j], type = "burden", scale = TRUE, 
-#                      slope = scale_df$sf[k], intercept = scale_df$intercept[k])
-#   }
-# dist_burden_min <- do.call("rbind", apply(dist_ttimes, 2, function(x) get.burden.fixed(pop = mada_district$pop2015adj, ttimes = x/60, names = mada_district$mdg_dis_co, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, p_rabid = 0.6, type = "burden")))
-# dist_burden_max <- do.call("rbind", apply(dist_ttimes, 2, function(x) get.burden.fixed(pop = mada_district$pop2015adj, ttimes = x/60, names = mada_district$mdg_dis_co, B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, p_rabid = 0.2, type = "burden")))
-# 
-# comm_burden_min <- do.call("rbind", apply(comm_ttimes, 2, function(x) get.burden.fixed(pop = mada_communes$MDG__201, ttimes = x/60, names = mada_communes$mtch_ds, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 25, p_rabid = 0.6, type = "burden")))
-# comm_burden_max <- do.call("rbind", apply(comm_ttimes, 2, function(x) get.burden.fixed(pop = mada_communes$MDG__201, ttimes = x/60, names = mada_communes$mtch_ds, B_ttimes = B_ttimes_mada, B_0 = B_0_mada, hdr = 5, p_rabid = 0.2, type = "burden")))
-# 
-# ## Decrease in burden over times proportional
-# comm_burden_min$scale <- "commune"
-# comm_burden_max$scale <- "commune"
-# dist_burden_max$scale <- "district"
-# dist_burden_min$scale <- "district"
-# burden_range <- bind_rows(comm_burden_max, comm_burden_min, dist_burden_max, dist_burden_min)
-# burden_range %>%
-#   group_by(scenario, scale, hdr) %>%
-#   summarize(burden_total = sum(deaths)) %>%
-#   group_by(scale, hdr, names) %>%
-#   mutate(burden_base = deaths[scenario == 0]) -> burden_all
-#   
-# ggplot(data = comm_burden_min, aes(x = scenario, ))
-# 
-# plot(colSums(comm_burden_min, na.rm = TRUE)/sum(comm_burden_min[,1], na.rm = TRUE), type = "l", col = "blue", bty = "n", ylim = c(0, 1), xlab = "# Additional ARMC", 
-#      ylab = "Proportion of deaths compared to baseline ")
-# lines(colSums(dist_burden_min, na.rm = TRUE)/sum(dist_burden_min[, 1], na.rm = TRUE),
-#       col = "purple")
-# lines(colSums(comm_burden_max, na.rm = TRUE)/sum(comm_burden_max[, 1], na.rm = TRUE), 
-#      type = "l", col = "blue")
-# lines(colSums(dist_burden_max, na.rm = TRUE)/sum(dist_burden_max[, 1], na.rm = TRUE), col = "purple")
-# 
-# ## Decrease in burden over times absolute
-# plot(colSums(comm_burden_min, na.rm = TRUE), type = "l", col = "blue", bty = "n",
-#      ylim = c(0, 1500), xlab = "# Additional ARMC",
-#      ylab = "Number of deaths")
-# lines(colSums(dist_burden_min, na.rm = TRUE), col = "purple")
-# lines(colSums(comm_burden_max, na.rm = TRUE), type = "l", col = "blue")
-# lines(colSums(dist_burden_max, na.rm = TRUE), col = "purple")
-
-```
-
 ```{r simulate vials used}
 ## Get catchment
-# comm_catchments <- read.csv("output/temp_ARMC.csv", row.names = 1)
-# comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
-# comm_ttimes <- cbind(mada_communes@data$ttms_ll, comm_ttimes)
-# comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
-# comm_mat[, 1] <- comm_ttimes[, 1]
-# 
-# for (i in 2:ncol(comm_ttimes)){
-#   comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], comm_ttimes[, i], NA)
-# }
-# 
-# comm_mat <- as.data.table(comm_mat)
-# comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
-# comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
-# comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
-# 
-# comm_catchments$scenario <- 1:nrow(comm_catchments)
+catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
+mada_communes@data %>%
+  left_join(comm_data) -> comm_catches
+comm_ttimes <- cbind(comm_ttimes[, 1], comm_ttimes)
+names(comm_ttimes)[1] <- "result.0"
+comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
+comm_mat[, 1] <- as.character(comm_catches$CTAR)
+
+for (i in 2:ncol(comm_ttimes)){
+  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
+                          comm_mat[, 1])
+}
+
+comm_mat <- as.data.table(comm_mat)
+comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
+comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
+comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+comm_ttimes <- as.data.table(comm_ttimes)
+comm_ttimes$mdg_cm_ <- mada_communes$mdg_cm_
+comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
+
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(bites = sum(exp_bites, na.rm = TRUE)) -> comm_master
+
+sim.throughput <- function(B_bites, B_0, )
+
+# catch_scenario$scenario <- 1:nrow(catch_scenario)
 # comm_mat %>%
@@
 # comm_mat %>%
-#   left_join(comm_catchments, by = c("scenario" = "scenario")) -> comm_mat
+#   left_join(catch_scenario, by = c("scenario" = "scenario")) -> comm_mat
 # comm_mat$CTAR <- as.character(comm_mat$CTAR)

### 2019-03-20 accessPEP.Rmd
working on figure outputs

@@
 exps_mora <- exps_mora[["mora_bites"]]
+
 ## Other data
@@
 
-```{r raw data, cache = TRUE}
-## Network fig (make with ggplot!)
-## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments
-
+```{r fig M1 data, cache = TRUE, fig.cap= ""}
 ## get exposure matrix = first set of points
@@
   labs(tag = "A") +
-  xlab("") +
-  ylab("") +
+  xlab("Longitude") +
+  ylab("Latitude") +
   theme(axis.text.x = element_blank(), axis.text.y = element_blank())
@@
 
-p + plot_spacer() + c + plot_layout(nrow = 1, widths = c(1, 0.1, 1))
-
-```
-
-For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1).
+figM1 <- p + c
+
+figM1
+
+ggsave("figs/M1.jpeg", figM1, device = "jpeg", width = 10, height = 8)
+
+```
+
+For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A).
 Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC.
@@
 
-As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2).
-Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so in subsequent analyses we looked at the sensitivity of model parameter estimates to changing this threshold.
+As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2.1A).
+Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S2.1B), so in subsequent analyses we looked at the sensitivity of model parameter estimates to changing this threshold.
 To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
@@
 These low risk exposures do not contribute to burden and therefore we attempted to exclude them from our analysis.
-Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A).
-We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4).
+Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure S2.1A).
+We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S2.1B).
 For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_contacts %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_contacts %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts.
@@
 We modeled incidence of bites as a function of travel times in hours as follows:
-$$ 
+
+\begin{equation}
   \mu = exp(\beta_{t}T + \beta_0)P 
-$$
-where $mu$ is the expected number of bites given travel times and population for a given location.
-We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
+\begin{equation}
+
+where $mu$ is the expected number of bites given travel times and the human population size for a given location.
+We then estimate the likelihood of observing our data where observed bites are a poisson distribution around the mean $\mu$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.
 
 We fit this to our available data in three ways:
-1) Commune level bite incidence from the Moramanga District
-2) District level bite incidence as calculated above from the national database
-3) District level bite incidence, but with commune level travel times ($T_j$) so that:
-
-$$ 
+1) Average annual reported bites at the commune level estimated from 27 months of data from the Moramanga District
+2) Average annual reported bites at the district level (calculated as above from the national database)
+3) Average annual reported bites at the district level, but with commune level travel times ($T_j$) so that:
+
+\begin{equation}
   \mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
-$$
-
-As travel times are correlated with human population size (Figure S4), we also compared how well bites were predicted by human population size.
-We used a generalized linear model with a possion response-now predicting the number of bites rather than bite incidence` with either population size alone or in combination with travel times, so that incidence could scale with population size.
+\begin{equation}
+
 We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access, replacing travel times in hours ($T$) with distance from the nearest ARMC in km ($D$).
 Distance to the nearest ARMC was taken as the minimum distance from the centroid of the commune or district to any ARMC.
+As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone and in combination with our metrics of access.
+For the models with population size, we used a model framework with either population size alone or population size and a access metric (either travel times or distance) as predictors of bites (for example, for a model with population size and distance as covariates: $\mu = exp(\beta_{t}D + \beta_pP + \beta_0)$).
 
@@
 ### Estimation of burden and reporting 
-We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
+We used our model to predict average annual bite incidence for all 114 districts in Madagascar.
+We estimate average reporting of rabid exposures and deaths due to rabies given this and assumptions about rabies exposures.
 
 We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
-$$
+
+\begin{equation}
   \overline\rho = \frac{\mu p_{rabid}}{R}
-$$
+\begin{equation}
+
 or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$.
@@
 where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
-$$
+
+\begin{equation}
            p_{rabid}=
@@
             \end{cases}
-$$
+\begin{equation}
 <br>
@@
 We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence, which gives a range of approximately `r 1e5/25*0.01*0.39` - `r 1e5/5*0.01*0.39` exposures per 100,000 persons.
-As there is little data on dog populations in Madagascar, this range encompasses both a wide range of observed HDRs and exposure rates across Africa (cite!).
-
-To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
-$$
+As there is little data on dog populations in Madagascar, this range encompasses both a wide range of observed HDRs and exposure rates across Africa (*cite*).
+
+To estimate the average number of deaths for each administrative unit, we extended the above framework into a stochastic framework as follows:
+
+\begin{equation}
 deaths_i = (R_i - p_{rabid_i}B_i) \times p_{death}
-$$
+\begin{equation}
+
 where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model.
-We draw $p_rabid_i$ from a uniform distribution between 0.2 - 0.6, while constraining it as per Equation N.
+We draw $p_rabid_i$ from a uniform distribution between 0.2 - 0.6, while constraining it as per Equation 4.
 We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
-The probability of death in the absence of PEP is taken from cite GAVI/Joel paper.
+The probability of death in the absence of PEP is taken from Changalucha et al.
+2018 (*cite*).
 <br>
@@
 ### Strategically expanding PEP access
-Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional ARMC.
+Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of expanding PEP provisioning to additional clinics.
 Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country.
@@
 
-We also look at how this may impact vaccine demand.
-Specifically, we fit the observed daily distribution of patients receiving PEP at each clinic to a negative binomial distribution, with both the dispersion parameter and mean as a function of the annual average number of reported bites.
-We use this to simulate clinic throughput and vials used as clinics are added, assuming routine vial sharing practices used currently in Madagascar (i.e.
-each vial is split between two patients, and vials are discarded at the end of each day).
-We project the potential increase in vial demand and change in the ratio of number of vials per death averted.
+We also look at how expansion to additional clinics may impact vaccine demand.
+Specifically, we use our model predicted estimates of annual bites for each administrative unit and then aggregate these to the clinic catchment (assuming that all patients report to their closest clinic).
+For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assuming perfect compliance (i.e.
+all patients reporting subsequently on day 3 and day 7 for their second and third doses) generate dates of subsequent doses.
+We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e.
+each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged TRC regime (2 injection sites on day 0, 3, and 7).
+We take the mean of 100 simulations of throughput for each catchment and scenario.
 
@@
 Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden.
-We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR, see SN) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population.
-
+We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see section S5) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
 <br>
@@
 ### Models of bites as a function of travel times
-
-```{r models ttimes, cache = TRUE}
+```{r fig M2 models of ttimes, cache = TRUE}
 ## Exclude for comm data!
@@
                                 model = mapply(function(x) attributes(x)$label, mada_mods)))
-kable(bite_mods)
 ## Predictions: fixed
@@
   geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
-  scale_color_manual(values = c("navy", "blue", "purple"), guide = "none") +
-  scale_fill_manual(values = c("blue", "lightblue", "pink"), name = "Model scale") +
+  scale_color_manual(values = c("navy", "purple", "red"), name = "Model scale") +
+  scale_fill_manual(values = c("blue", "lavender", "pink"), guide = "none") +
   xlab("Travel time (hrs)") +
@@
   expand_limits(y = c(0, 4000)) +
-  scale_color_manual(values = c("navy", "blue", "purple"), name = "Model scale") +
+  scale_color_manual(values = c("navy", "purple", "red"), name = "Model scale") +
   xlab("Mean annual bites observed") +
@@
   labs(tag = "B")
-a / b
+
+figM2 <- a / b
+figM2
+
+ggsave("figs/M2.jpeg", figM2, device = "jpeg", width = 7, height = 5)
 
 kable(bite_mods)
+
 ```
@@
 
-We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC.
-All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the model.
+We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 2A), with reported bite incidence decreasing with travel times to the ARMC.
+The model of bites as a function of travel times at the commune level summed to the district level resulted in a fit closer to the model fit to the Moramanga data at the commune level than the district model.
+All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the models, generally underestimating bite incidence in some districts.
 
 ### Estimation of burden and reporting 
-```{r}
+```{r fig M3 reporting}
 B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
@@
 
-ggplot(rep_df, aes(x = ttimes, y = reporting, color = interaction(p_rabid, hdr), linetype = as.factor(hdr))) +
-  geom_line() +
-  facet_wrap(~ scale) +
-  scale_color_manual(values = c("red", "red", "blue", "blue"))
+# ggplot(rep_df, aes(x = ttimes, y = reporting, color = interaction(p_rabid, hdr), linetype = as.factor(hdr))) +
+#   geom_line() +
+#   facet_wrap(~ scale) +
+#   scale_color_manual(values = c("red", "red", "blue", "blue"))
 
@@
   geom_ribbon(aes(ymax = rep_max, ymin = rep_min), alpha = 0.5) +
-  scale_fill_manual(values = c("navy", "purple")) +
-  scale_color_manual(values = c("navy", "purple")) +
-  xlab("Reporting") +
-  ylab("Travel times (in hours)") +
+  scale_fill_manual(values = c("purple", "navy")) +
+  scale_color_manual(values = c("purple", "navy")) +
+  ylab("Reporting") +
+  xlab("Travel times (in hours)") +
   labs(tag = "A")
 
-```
-
-Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure N).
-Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5.5 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).
-We found that the commune model of bites estimated a lower range of reporting probabilities that the district model (Figure SX), so we compared estimates of these two models in subsequent analyses.
-
-```{r burden results, fig.cap = "Figure 7", cache = TRUE}
-# Incremental analysis ------------------------------------------------------------------------
+## Comparing scales
+p_rabid_vals <- seq(0.2, 0.6, by = 0.05)
+hdr_vals <- seq(5, 25, by = 5)
+
+rep_df <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
+    foreach(j = 1:length(hdr_vals), .combine = rbind) %do% {
+    reporting_comm <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+    comm <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = reporting_comm, hdr = hdr_vals[j], 
+                               pop = comm_data$pop, ttimes = comm_data$ttimes_weighted,
+                               p_rabid = p_rabid_vals[i], 
+                               scale = "Commune", params = "Commune model"))
+
+    reporting_dist <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
+                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+
+     dist <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist, 
+                                pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+                                hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                params = "District model"))
+     check <- bind_rows(comm, dist)
+  }
+
+rep_df %>%
+  group_by(names, hdr, p_rabid, scale, params) %>%
+  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
+  summarize(reporting = sum(reporting*prop_pop), ttimes = mean(ttimes, na.rm = TRUE)) -> reporting
+
+d <- ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = scale)) +
+  geom_boxplot(outlier.shape = NA) +
+  scale_color_manual(values = c(alpha("blue", 0.5), alpha("purple", 0.5)), name = "Parameters") +
+  xlab("Districts ordered by travel times \n (low to high)") +
+  ylab("Reporting of rabid exposures") +
+  theme(axis.text.x = element_blank()) +
+  labs(tag = "B")
+
+c / d
+
+```
+
+Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 3).
+Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5 - 10 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).
+We found that the commune model estimated a lower range of reporting probabilities than the district model (Figure 3, see section S6), so we compared estimates of burden given these two models in subsequent analyses.
+
+```{r fig M4 burden, cache = TRUE}
 ## District travel times
@@
 for (i in 2:ncol(dist_ttimes)){
-  i = 3
   dist_mat[, i] <- ifelse(dist_ttimes[, i] < dist_ttimes[, i - 1], dist_ttimes[, i], NA)
@@
   rename(names = mdg_dis_co) -> results_comm_summarized
-results_comm_summarized$model <- "commune"
+results_comm_summarized$model <- "Commune"
 colnames(results_dist_med) <- gsub("results.", "", colnames(results_dist_med))
 results_dist_med %>%
-  mutate(model = "district") %>%
+  mutate(model = "District") %>%
   select(scenario, names, deaths_mean, deaths_lowerCI, deaths_upperCI, 
@@
 results_all %>% filter(scenario == 0) -> results_all_base
+results_all_base %>%
+  left_join(dist_data, by = c("names" = "district")) -> results_all_base
 ggplot(filter(results_all_base, !is.na(district)), 
-       aes(y = deaths_mean, x = reorder(district, ttimes_weighted), 
+       aes(y = deaths_mean/pop*1e5, x = reorder(district, ttimes_weighted), 
                               color = model)) +
@@
   xlab("Districts ordered by travel times \n (low to high)") +
-  ylab("Average annual deaths") +
-  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
-                                "commune" = alpha("purple", 0.5)), 
+  ylab("Annual deaths per 100k persons") +
+  coord_flip() +
+  scale_color_manual(values = c("district" = alpha("purple", 0.5), 
+                                "commune" = alpha("navy", 0.5)), 
                      name = "Scale") +
-  theme(axis.text.x = element_blank()) +
-  labs(tag = "B") -> d
-
-c / d
-
-## Deaths averted
-ggplot(filter(results_all_base, !is.na(district)), 
-       aes(y = reorder(district, ttimes_weighted), x = averted_mean, 
-                              color = model)) +
-  ggstance::geom_pointrangeh(aes(xmin = averted_lowerCI, xmax = averted_upperCI), fatten = 2) +
-  ylab("") +
-  xlab("Average annual deaths averted") +
-  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
-                                "commune" = alpha("purple", 0.5)), 
-                     name = "Scale") +
-  theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))
+  theme(axis.text.y = element_blank()) +
+  labs(tag = "A") -> d
+
+gg_district %>%
+  left_join(results_all_base, by = c("id" = "names")) -> gg_district_burden
+
+p <- ggplot() +
+  geom_polygon(data = gg_district_burden, aes(x = long, y = lat, group = group, 
+                                              fill = deaths_mean/pop2015adj*1e5), 
+               color = "grey50") +
+  geom_point(data = ctar_points, aes(x = LONGITUDE, y = LATITUDE), color = "blue") + 
+  scale_fill_gradient(low = "white", high = "red", name = "Annual deaths per 100k \n persons") +
+  labs(tag = "B") +
+  facet_wrap(~model)
+
+burden <- d / p
+
+burden
+
+ggsave("figs/burden_maps.jpeg", burden, device = "jpeg", width = 10, height = 10)
 
@@
 
-When we estimate burden of deaths stochastically within this range of incidence and of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure XA, results presented aggregated at the district level).
 Overall, we estimate average annual deaths between `r min(summary_burden$deaths)` - `r max(summary_burden$deaths)` and deaths_averted between `r min(summary_burden$averted)` - `r max(summary_burden$averted)`.
-
-Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust...write up from supplement!
+In general, incidence of rabies deaths increases with travel times, but this effect is more pronounced for the district model compared to the commune model.
+
+Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust (see section S5).
+The district model is more sensitive to assumptions of $p_rabid$ and $\rho_max$ (Fig S5.1).
 
@@
 
-``` {r scenario, cache = TRUE}
+We find that expanding access to PEP reduces burden of rabies death, but this effect decays as ARMC are added, saturating after about approximately 200 clinics (or 2 ARMC per district).
+This effect is more pronouned for the commune level model than for the district model.
+These patterns are also generally robust to our assumptions, although the decrease in deaths saturates quickly for the district model as $p_rabid$ and $rho_max$ decrease?
+
+```{r simulate vials used, cache = TRUE}
+## Get catchment
+catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
+comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
+mada_communes@data %>%
+  left_join(comm_data) -> comm_catches
+comm_ttimes <- cbind(comm_ttimes[, 1], comm_ttimes)
+names(comm_ttimes)[1] <- "result.0"
+comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
+comm_mat[, 1] <- as.character(comm_catches$CTAR)
+
+for (i in 2:ncol(comm_ttimes)){
+  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
+                          comm_mat[, 1])
+}
+
+comm_mat <- as.data.table(comm_mat)
+comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
+comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
+comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+comm_ttimes <- as.data.table(comm_ttimes)
+comm_ttimes$mdg_cm_ <- mada_communes$mdg_cm_
+comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
+comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
+
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites_comm = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop, 
+         exp_bites_dist = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE), 
+            bites_comm = sum(exp_bites_comm, na.rm = TRUE),
+            bites_dist = sum(exp_bites_dist, na.rm = TRUE)) %>%
+  drop_na(catchment) -> comm_master
+
+sim.throughput <- function(bites, nsims = 100) {
+  get.vials <- function(x) {
+    day0 <- round(runif(rpois(1, x), min = 1, max = 365))
+    days <- c(day0, day0 + 3, day0 + 7)
+    vials <- sum(ceiling(table(days)/2))
+    return(vials)
+  }
+
+  check <- foreach(i = 1:nsims, .combine = cbind) %dopar% {
+    vials <- sapply(bites, get.vials, simplify = TRUE)
+  }
+
+  mean_vials <- rowMeans(check, na.rm = TRUE)
+  sd_vials <- apply(check, 1, sd, na.rm = TRUE)
+  ci_upper <- mean_vials + 1.96*(sd_vials/sqrt(nsims))
+  ci_lower <- mean_vials - 1.96*(sd_vials/sqrt(nsims))
+
+  return(as.data.frame(list(bites = bites, mean_vials = mean_vials, ci_upper = ci_upper,
+                            ci_lower = ci_lower)))
+}
+
+## local cluster
+library(doParallel)
+# Calculate the number of cores
+no_cores <- detectCores() - 1
+# Initiate cluster
+cl <- makeCluster(no_cores)
+registerDoParallel(cl)
+system.time({comm <- sim.throughput(bites = comm_master$bites_comm, nsims = 3)})
+system.time({dist <- sim.throughput(bites = comm_master$bites_dist, nsims = 3)})
+comm$model <- "commune"
+dist$model <- "district"
+
+stopCluster(cl) ## for doParallel
+
+throughput_comm <- bind_cols(comm_master, select(comm, -bites))
+throughput_dist <- bind_cols(comm_master, select(dist, -bites))
+throughput <- bind_rows(throughput_comm, throughput_dist)
+throughput %>%
+  group_by(scenario, model) %>%
+  summarize(mean_vials = sum(mean_vials, na.rm = TRUE), 
+            upper_vials = sum(ci_upper, na.rm = TRUE), 
+            lower_vials = sum(ci_lower, na.rm = TRUE), 
+            bites_comm = sum(bites_comm, na.rm = TRUE), 
+            bites_dist = sum(bites_dist, na.rm = TRUE)) -> throughput_scenario
+
+```
+
+```{r fig M5 scenario deaths and vials, cache = TRUE}
 ## Plotting national level
@@
   xlab("# of Additional ARMC") +
-  ylab("Mean annual deaths \n due to rabies") +
+  ylab("Annual deaths \n due to rabies nationally") +
   labs(tag = "A") -> e
 
-## Plotting national level
-ggplot(data = results_all, aes(x = deaths_mean, y = reorder(district, ttimes_weighted), 
-                               color = scenario)) +
-  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
-                             fatten = 2) +
-  facet_wrap(~model)
-
-## Deaths averted
-ggplot(data = results_natl, aes (x = scenario, y = averted_mean*0.16, color = model)) +
+ggplot(data = throughput_scenario, aes (x = scenario, y = mean_vials, color = model)) +
   geom_line() +
-  geom_ribbon(aes(ymin = averted_upperCI, ymax = averted_lowerCI, 
+  geom_ribbon(aes(ymin = lower_vials, ymax = upper_vials, 
                   color = NULL, fill = model),
               alpha = 0.5) +
-  scale_color_manual(values = c("district" = "blue", "commune" = "purple"),  guide = "none") +
-  scale_fill_manual(values = c("district" = "blue", "commune" = "purple"), guide = "none") +
+  scale_color_manual(values = c("district" = "blue", 
+                                "commune" = "purple"), 
+                     guide = "none") +
+  scale_fill_manual(values = c("district" = "blue", 
+                                "commune" = "purple")) +
+  guides(color = guide_legend(), 
+         fill = guide_legend(override.aes = list(linetype = 1))) +
   xlab("# of Additional ARMC") +
-  ylab("Mean annual \n deaths averted") +
+  ylab("Mean estimate of vials needed") +
+  labs(tag = "B")
+
+throughput_scenario %>% left_join(results_natl) -> results_natl_all
+
+ggplot(data = results_natl_all, aes (x = scenario, y = mean_vials/averted_mean, color = model)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lower_vials/averted_mean, ymax = upper_vials/averted_mean, 
+                  color = NULL, fill = model),
+              alpha = 0.5) +
+  scale_color_manual(values = c("district" = "blue", 
+                                "commune" = "purple"), 
+                     guide = "none") +
+  scale_fill_manual(values = c("district" = "blue", 
+                                "commune" = "purple")) +
+  guides(color = guide_legend(), 
+         fill = guide_legend(override.aes = list(linetype = 1))) +
+  xlab("# of Additional ARMC") +
+  ylab("Estimate of vials \n per death averted") +
   labs(tag = "B") -> f
@@
 e / f
-
-```
-
-We find that both expanding access to PEP reduces burden of rabies death, but this effect decays as ARMC are added, saturating after about approximately 200 clinics (or 2 ARMC per district).
-This effect is more pronouned for the commune level model than for the district model.
-Sensitivity analyses...write up from supplement!
-
-### Estimating vaccine demand with increased access
-
-```{r vial analysis, cache = TRUE}
-# ## new ctar
-rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
-                          end_date = ymd("2017-12-31"), 
-                          threshold = 10)
-date_mat <- rep_data[["check_mat"]]
-dose_mat <- rep_data[["doses_wide"]]
-colnames(date_mat) <- colnames(dose_mat)[2:(ncol(dose_mat)-1)]
-doses <- melt(dose_mat, id.vars = "date_de_consultation", value.name = "patients", variable.name = "id_ctar")
-
-date_mat %>% 
-  as_tibble() %>%
-  mutate(date = dose_mat$date_de_consultation) -> date_mat
-exclude <- melt(date_mat, id.vars = "date", value.name = "exclude", variable.name = "id_ctar")
-doses %>%
-  left_join(exclude, by = c("date_de_consultation" = "date", "id_ctar" = "id_ctar")) -> doses
-doses %>%
-  filter(exclude == 1) %>%
-  group_by(id_ctar) %>%
-  mutate(sum = sum(patients, na.rm = TRUE), mean = mean(patients, na.rm = TRUE)) %>%
-  filter(sum > 10) -> doses
-
-ggplot(data = doses, aes(x = date_de_consultation, y = id_ctar)) +
-  geom_tile(aes(fill = patients))
-
-library(ggridges)
-ggplot(doses, aes(x = patients, y = id_ctar)) +
-  geom_density_ridges()
-
-doses %>%
-  split(.$id_ctar) %>% # from base R
-  map(~ fitdistr(.$patients, "negative binomial")) -> dist_ts
-
-coeff_df <- as.data.frame(list(mu = mapply(function(x) coef(x)["mu"], dist_ts),
-                                k = mapply(function(x) coef(x)["size"], dist_ts),
-                                mu_sd = mapply(function(x) x$sd["mu"], dist_ts),
-                                k_sd = mapply(function(x) x$sd["size"], dist_ts)))
-coeff_df$id_ctar <- names(dist_ts)
-ggplot(data = coeff_df, aes(x = mu, y = k)) + geom_point()
-
-exps_dist %>%
-  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
-  mutate(id_ctar = as.character(id_ctar)) %>%
-  group_by(id_ctar) %>%
-  summarize(bites = sum(bites)) %>%
-  left_join(coeff_df, by = c("id_ctar" = "id_ctar")) -> coeff_df
-ggplot(data = coeff_df, aes(x = bites, y = mu)) + geom_point()
-ggplot(data = coeff_df, aes(x = bites, y = k)) + geom_point()
-
-exps_dist %>%
-  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
-  mutate(id_ctar = as.character(id_ctar)) %>%
-  group_by(id_ctar) %>%
-  summarize(bites = sum(bites)) %>%
-  left_join(doses, by = c("id_ctar" = "id_ctar")) -> doses
-
-get.throughput <- function(patient_ts, bites, B_k, B_0_k, B_mu, B_0_mu) {
-  mu <- exp(B_mu*bites/100 + B_0_mu) 
-  k <- exp(B_k*bites/100 + B_0_k)
-  return(-sum(dnbinom(patient_ts, size = k, mu = mu, log = TRUE)))
-}
-
-fit <- mle2(get.throughput, start = list(B_0_mu = 0.1, B_k = 0.1, B_mu = 0.1, B_0_k = 0.1),
-             data = list(patient_ts = doses$patients, bites = doses$bites))
-
-## Get the observed + predicted distribution for each set of coeffs
-check <- foreach(i = iter(coeff_df, by = "row"), .combine = rbind) %do% {
-  check <- rnbinom(n = 365, mu = exp(0.029*i$bites/100 + -0.307),
-                  size = exp(0.022*i$bites/100 + -0.39))
-  as.data.frame(list(patients = check, type = "predicted", id_ctar = i$id_ctar))
-}
-
-doses %>%
-  select(id_ctar, patients) %>%
-  mutate(type = "observed") %>%
-  bind_rows(check) -> compare_dists
-
-compare_dists %>%
-  group_by(type, id_ctar, patients) %>%
-  summarize(count = n()) %>%
-  group_by(type, id_ctar) %>% # next get # of days total (should always be 365 for predicted)
-  mutate(total = sum(count), prop = count/total) -> compare_dists
-compare_dists %>%
-  mutate(prop_plot = ifelse(type == "predicted", -prop, prop)) -> compare_dists
-
-ggplot(compare_dists, aes(x = patients, y = prop_plot, fill = type)) +
-  geom_col(color = NA, alpha = 0.5, position = "identity") +
-  scale_fill_manual(values = c("red", "blue")) +
-  facet_wrap(~id_ctar, scales = "free") +
-  xlab("# of Patients") +
-  ylab("Proportion of days") -> a
-
-rep_bites = seq(0, 15000, by = 100)
-predicts <- as.data.frame(list(rep_bites = rep_bites,
-                               predicted_mu = exp(0.029*rep_bites/100 + -0.307),
-                               predicted_k = exp(0.022*rep_bites/100 + -0.39)))
-ggplot(predicts, aes(x = rep_bites)) +
-  geom_line(aes(y = predicted_mu), color = "red") +
-  geom_line(aes(y = predicted_k), color = "blue") +
-  xlab("Annual reported bites to a clinic") +
-  ylab("Predicted parameter") -> b
-
-a / b
-
-```
-
-We fit the distributions of patients reporting per day to the estimated average annual bites reported to each clinic.
-We found that as reported bites increase, the average number of patients reporting per day increases linearly, and that the heterogeneity in this increases with lower reported bites to a clinic.
-
-We use this to estimate how vaccine demand will change as access to PEP is expanded...
-```{r simulate vials used}
-## Get catchment
-catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
-comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
-mada_communes@data %>%
-  left_join(comm_data) -> comm_catches
-comm_ttimes <- cbind(comm_ttimes[, 1], comm_ttimes)
-names(comm_ttimes)[1] <- "result.0"
-comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
-comm_mat[, 1] <- as.character(comm_catches$CTAR)
-
-for (i in 2:ncol(comm_ttimes)){
-  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
-                          comm_mat[, 1])
-}
-
-comm_mat <- as.data.table(comm_mat)
-comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
-comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
-comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
-comm_ttimes <- as.data.table(comm_ttimes)
-comm_ttimes$mdg_cm_ <- mada_communes$mdg_cm_
-comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
-comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
 
@@
   group_by(scenario, catchment) %>%
-  summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
-  drop_na(catchment) -> comm_master
-
-sim.throughput <- function(bites, nsims = 100) {
-  get.vials <- function(x) {
-    day0 <- round(runif(rpois(1, x), min = 1, max = 365))
-    days <- c(day0, day0 + 3, day0 + 7)
-    vials <- sum(ceiling(table(days)/2))
-    return(vials)
-  }
-
-  check <- foreach(i = 1:nsims, .combine = cbind) %dopar% {
-    vials <- sapply(bites, get.vials, simplify = TRUE)
-  }
-
-  mean_vials <- rowMeans(check, na.rm = TRUE)
-  sd_vials <- apply(check, 1, sd, na.rm = TRUE)
-  return(as.data.frame(list(bites = bites, mean_vials = mean_vials, sd_vials = sd_vials)))
-}
-
-## local cluster
-library(doParallel)
-# Calculate the number of cores
-no_cores <- detectCores() - 1
-# Initiate cluster
-cl <- makeCluster(no_cores)
-registerDoParallel(cl)
-system.time({throughput <- sim.throughput(bites = comm_master$bites, nsims = 10)})
-stopCluster(cl) ## for doParallel
-
-comm_throughput <- bind_cols(comm_master, throughput)
-
-```
-
-```{r throughput district}
+  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE), 
+            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
+  mutate(model = "commune") %>%
+  drop_na(catchment) -> comm_cmod
+pop_natl <- sum(mada_district$pop2015adj)
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE), 
+            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
+  mutate(model = "district") %>%
+  drop_na(catchment) -> comm_dmod
+bites_exp <- bind_rows(comm_cmod, comm_dmod)
+bites_exp %>%
+  group_by(scenario, model) %>%
+  summarize(bites = sum(bites)/pop_natl*1e5) %>%
+  ggplot(., aes(x = scenario, y = bites, color = model)) +
+  geom_line() +
+  ylab("Reported bites \n per 100k persons") +
+  xlab("# Additional ARMC")
+
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) -> comm_exp
+
+## Do district one
 ## Get catchment
@@
 dist_ttimes$scenario <- as.numeric(gsub("result.", "", dist_ttimes$scenario))
+# 
+# dist_mat %>%
+#   left_join(dist_ttimes) %>%
+#   left_join(select(mada_district@data, pop = pop2015adj, mdg_dis_co)) %>%
+#   mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+#   group_by(scenario, catchment) %>%
+#   summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
+#   drop_na(catchment) -> dist_master
 
@@
   left_join(select(mada_district@data, pop = pop2015adj, mdg_dis_co)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
-  drop_na(catchment) -> dist_master
-
-## local cluster
-library(doParallel)
-# Calculate the number of cores
-no_cores <- detectCores() - 1
-# Initiate cluster
-cl <- makeCluster(no_cores)
-registerDoParallel(cl)
-system.time({throughput <- sim.throughput(bites = dist_master$bites, nsims = 10)})
-stopCluster(cl) ## for doParallel
-
-dist_throughput <- bind_cols(dist_master, throughput)
+  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) -> dist_exp
+ggplot(data = filter(dist_exp, scenario %in% c(0, 1, 10, 50, 100, 200)), 
+       aes(x = ttimes/60, y = as.factor(scenario))) +
+  ggridges::geom_density_ridges()
+dist_exp$scale <- "District"
+comm_exp$scale <- "Commune"
+
+check_exps <- bind_rows(select(comm_exp, scenario, ttimes, pop, scale, exp_bites), 
+                        select(dist_exp, scenario, ttimes, pop, scale, exp_bites))
+ggplot(data = filter(check_exps, scenario %in% c(0, 1, 10, 50, 100, 200)), 
+       aes(x = ttimes/60, y = as.factor(scenario), fill = scale)) +
+  ggridges::geom_density_ridges(alpha = 0.5)
+
+ggplot(data = filter(check_exps, scenario %in% c(0, 1, 10, 50, 100, 200)), 
+       aes(x = exp_bites/pop*1e5, y = as.factor(scenario), fill = scale)) +
+  ggridges::geom_density_ridges(alpha = 0.5)
 
@@
 
-ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
-  geom_point() +
+mod_names <- c("flatPop" = "Flat incidence", "addPop" = "With pop", 
+               "onlyPop" = "Only pop")
+covar_names <- c("distance" = "Distance (km)", "ttimes" = "Travel times (hrs)", "pop" = "Population")
+
+ggplot(filter(preds_data, pop_predict != "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
+  geom_point(alpha = 0.5) +
   geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
-  facet_grid(covar_name ~ pop_predict, scales = "free_x") +
-  expand_limits(y = 0)
+  scale_color_manual(values = c("blue", "purple", "red"), name = "Scale") +
+  facet_grid(pop_predict ~ covar_name, scales = "free_x", 
+             labeller = labeller(pop_predict = as_labeller(mod_names), 
+                                 covar_name = as_labeller(covar_names)), drop = TRUE) +
+  expand_limits(y = 0) +
+  xlab("Observed bites") +
+  ylab("Predicted bites")
+
+ggplot(filter(preds_data, pop_predict == "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
+  scale_color_manual(values = c("blue", "purple", "red"), name = "Scale") +
+  facet_grid(pop_predict ~ covar_name, scales = "free_x", 
+             labeller = labeller(pop_predict = as_labeller(mod_names), 
+                                 covar_name = as_labeller(covar_names)), drop = TRUE) +
+  expand_limits(y = 0) +
+  xlab("Observed bites") +
+  ylab("Predicted bites")
+
 ```
@@
 ```{r rles and cut-offs, cache = TRUE}
-consec_days <- c(5, 15, 30)
-contact_cutoff <- c(0, 3, 10)
+consec_days <- c(5, 15, 30, 60)
+contact_cutoff <- c(1, 3, 10, 100)
+
 combine <- function(x, ...) {  
@@
 }
+
 mods <- 
@@
     foreach(k = 1:length(contact_cutoff), .combine = combine, .multicombine = TRUE) %do% {
-    exps_dist <- get.district.data(ctar_data = ctar_data, covars = dist_data, ctar_metadata, 
+
+    exps <- get.district.data(ctar_data = ctar_data, covars = dist_data, ctar_metadata, 
                                consec_days = consec_days[i], 
@@
                                start = ymd("2014-01-01"), end = ymd("2017-12-31"))
-
-    mada_mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
-                             pop_predict = c("addPop", "flatPop"),
+    exps %>%
+      group_by(district) %>%
+      summarise(bites = mean(bites), n = n()) %>%
+      left_join(dist_data) -> exps_dist
+
+    mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
+                             pop_predict = c("flatPop"),
                              pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
-                             covars = list(distance_district = exps_dist$distance, 
-                                           distance_commune = comm_covars$distance, 
-                                           ttimes_district = exps_dist$ttimes_weighted_dist/60,
+                             covars = list(ttimes_district = exps_dist$ttimes_weighted_dist/60,
                                            ttimes_commune = comm_covars$ttimes_weighted/60), 
                              sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
-    mada_mods_pop <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
-                              pop_predict = "onlyPop", 
-                              pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
-                              covars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop), # dummy covars
-                              sum_it = c(FALSE, TRUE), run_type = "optim", label = "Mada")
-
-    mods_all <- c(mada_mods_covars, mada_mods_pop)
 
     ## profile CIs
-    mods_covars <- Filter(function(x) !grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
     prof_CI_covars <- lapply(mods_covars, confint)
-    mods_pop <- Filter(function(x) grepl("addPop", attributes(x)$label, fixed = TRUE), mods_all)
-    prof_CI_pop <- lapply(mods_pop, confint)
-    bite_mods_covar_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_covars),
+    check <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_covars),
                                     intercept = mapply(function(x) coef(x)["intercept"], mods_covars),
@@
                                     model = mapply(function(x) attributes(x)$label, mods_covars)))
-
-    bite_mods_pop_df <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods_pop),
-                                    beta_pop = mapply(function(x) coef(x)["beta_pop"], mods_pop),
-                                    intercept = mapply(function(x) coef(x)["intercept"], mods_pop),
-                                    beta_upper = mapply(function(x) x["beta", 2], prof_CI_pop),
-                                    beta_lower = mapply(function(x) x["beta", 1], prof_CI_pop),
-                                    beta_pop_upper = mapply(function(x) x["beta_pop", 2], prof_CI_pop),
-                                    beta_pop_lower = mapply(function(x) x["beta_pop", 1], prof_CI_pop),
-                                    intercept_lower = mapply(function(x) x["intercept", 1], prof_CI_pop),
-                                    intercept_upper = mapply(function(x) x["intercept", 2], prof_CI_pop),
-                                    likelihood = mapply(function(x) logLik(x), mods_pop),
-                                    model = mapply(function(x) attributes(x)$label, mods_pop)))
-    check <- bind_rows(bite_mods_covar_df, bite_mods_pop_df)
 
@@
     preds_fixed$contact_cutoff <- contact_cutoff[k]
-    preds_data <- run.predicts.data(covar_df = bite_mods, run_type = "predict.data")
+    preds_data <- run.predicts.data(covar_df = check, run_type = "predict.data")
     preds_data$consec_days <- consec_days[i]
@@
 
-ggplot(data = mods_df, aes(x = structure, y = beta)) +
-  geom_boxplot(aes(color = covar)) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-ggplot(data = mods_df, aes(x = structure, y = intercept)) +
-  geom_boxplot(aes(color = covar)) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-ggplot(preds_fixed_df, aes(x = covar, y = predicted, color = interaction(consec_days, contact_cutoff), 
-                           linetype = scale)) + 
+ggplot(preds_fixed_df, aes(x = covar, y = predicted, group = interaction(consec_days, scale, contact_cutoff), 
+                           color = consec_days, linetype = scale), alpha = 0.2) + 
   geom_line() + 
-  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
-  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
+  scale_color_gradient(low = "pink", high = "red", name = "Reporting cut-off") +
+  scale_linetype(name = "Model") +
   xlab("Travel time (hrs)") +
   ylab("Reported bites per \n 100,000 persons") +
-  facet_wrap(scale ~ covar_name, scales = "free")
+  facet_grid( ~ contact_cutoff)
 
@@
   geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
-  scale_color_manual(values = c("blue", "lightblue", "pink")) +
-  facet_grid(contact_cutoff ~ consec_days, scales = "free")
+  scale_color_manual(values = c("blue", "purple", "red"), name = "Scale") +
+  facet_grid(contact_cutoff ~ consec_days, scales = "free") +
+  xlab("Observed bites") +
+  ylab("Predicted bites")
 
@@
 ## Comparing scales
-comm_data$reporting_min <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+p_rabid_vals <- seq(0.2, 0.6, by = 0.05)
+hdr_vals <- seq(5, 25, by = 5)
+
+rep_df <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
+    foreach(j = 1:length(hdr_vals), .combine = rbind) %do% {
+    reporting_comm <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
                                             B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
-                                            hdr = 5, p_rabid = 0.2, type = "reporting", scale = FALSE)
-
-comm_data$reporting_max <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+    comm <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = reporting_comm, hdr = hdr_vals[j], 
+                               pop = comm_data$pop, ttimes = comm_data$ttimes_weighted,
+                               p_rabid = p_rabid_vals[i], 
+                               scale = "Commune", params = "Commune model"))
+
+    reporting_comm_dist <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
+                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+    comm_dist <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = reporting_comm_dist, 
+                                    hdr = hdr_vals[j], pop = comm_data$pop, ttimes = comm_data$ttimes_weighted,
+                                    p_rabid = p_rabid_vals[i], scale = "Commune", params = "District model"))
+
+    reporting_dist <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
+                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+
+     dist <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist, 
+                                pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+                                hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                params = "District model"))
+
+     reporting_dist_comm <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
                                             B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
-                                            hdr = 25, p_rabid = 0.6,  type = "reporting", scale = FALSE)
-comm_data %>%
-  group_by(district) %>%
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+
+     dist_comm <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist_comm, 
+                                     pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+                                     hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                     params = "Commune model"))
+
+     check <- bind_rows(comm, dist, comm_dist, dist_comm)
+  }
+
+rep_df %>%
+  group_by(names, hdr, p_rabid, scale, params) %>%
   mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
-  summarize(reporting_min = sum(reporting_min*prop_pop), 
-            reporting_max = sum(reporting_max*prop_pop)) %>%
-  mutate(reporting_med = (reporting_min + reporting_max)/2,
-         type = "summed from commune") -> summed_rep_dist
-
-mada_district@data %>%
-  mutate(reporting_min = get.burden.fixed(pop = pop2015un, ttimes = ttimes_weighted/60, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
-                                              p_rabid = 0.2, type = "reporting", scale = FALSE), 
-         reporting_max =get.burden.fixed(pop = pop2015un, ttimes = ttimes_weighted/60, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
-                                              p_rabid = 0.6,type = "reporting", scale = FALSE),
-         reporting_med = (reporting_min + reporting_max)/2,
-         type = "district level") %>%
-  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp
-
-pop_plot = 1e5
-rep_comp <- rbind(rep_comp, summed_rep_dist)
-rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]
-
-ggplot(rep_comp, aes(x = reporting_med, y = reorder(district, ttime), 
-                              color = type)) +
-  ggstance::geom_pointrangeh(aes(xmin = reporting_min, xmax = reporting_max), fatten = 2)
-
-bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
-R_100k_min <- pop_plot/25*0.01*0.39
-R_100k_max <- pop_plot/5*0.01*0.39
-
-plot(ttimes_plot, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-     type = "l", bty = "n",
-     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
-lines(ttimes_plot, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-     col = "darkred")
-lines(ttimes_plot, 
-     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
-     col = "red", lty = 2)
-lines(ttimes_plot, 
-     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
-     col = "red", lty = 2)
-polygon(x = c(ttimes_plot, rev(ttimes_plot)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
-        border = NA, col = alpha("red", 0.2))
-polygon(x = c(ttimes_plot, rev(ttimes_plot)), 
-        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
-             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
-        border = NA, col = alpha("red", 0.6))
-
-## Mora
-comm_data %>%
-  filter(district == "Moramanga") -> comm_level
-bites_100k <- exp(B_ttimes_mada*comm_level$ttimes_weighted + B_0_mada)*comm_level$pop
-R_100k_min <- comm_level$pop/25*0.01*0.39
-R_100k_max <- comm_level$pop/5*0.01*0.39
-
-## district val for Moramanga
-## need to merge names
-exps_dist %>%
-  filter(dist_name == "Moramanga") %>%
-  mutate(rep_max =get.burden.fixed(pop = pop, ttimes = ttimes_weighted_dist, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
-                                              p_rabid = 0.6),
-         rep_min =get.burden.fixed(pop = pop, ttimes = ttimes_weighted_dist, 
-                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
-                                              p_rabid = 0.2)) -> dist_level
-dist_level %$% plot(ttimes_weighted_dist/60, rep_max, col = "red", pch = 20)
-dist_level %$% points(ttimes_weighted_dist/60, rep_min, col = "red", pch = 20)
-dist_level %$% abline(v = ttimes_weighted_dist/60, col = "red", lty = 2)
-
-```
-
-```{r simulate vials used}
-## Get catchment
-catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
-comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
-mada_communes@data %>%
-  left_join(comm_data) -> comm_catches
-comm_ttimes <- cbind(comm_ttimes[, 1], comm_ttimes)
-names(comm_ttimes)[1] <- "result.0"
-comm_mat <- matrix(NA, nrow(comm_ttimes), ncol(comm_ttimes))
-comm_mat[, 1] <- as.character(comm_catches$CTAR)
-
-for (i in 2:ncol(comm_ttimes)){
-  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
-                          comm_mat[, 1])
-}
-
-comm_mat <- as.data.table(comm_mat)
-comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
-comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
-comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
-comm_ttimes <- as.data.table(comm_ttimes)
-comm_ttimes$mdg_cm_ <- mada_communes$mdg_cm_
-comm_ttimes <- melt(comm_ttimes, id.vars = "mdg_cm_", value.name = "ttimes", variable.name = "scenario")
-comm_ttimes$scenario <- as.numeric(gsub("result.", "", comm_ttimes$scenario))
-
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(bites = sum(exp_bites, na.rm = TRUE)) -> comm_master
-
-sim.throughput <- function(B_bites, B_0, )
-
-# catch_scenario$scenario <- 1:nrow(catch_scenario)
-# comm_mat %>%
-#   drop_na(ttimes) %>%
-#   left_join(select(comm_data, mdg_cm_, pop)) -> comm_mat
-# comm_mat %>%
-#   left_join(catch_scenario, by = c("scenario" = "scenario")) -> comm_mat
-# comm_mat$CTAR <- as.character(comm_mat$CTAR)
-# comm_data$CTAR <- as.character(comm_data$CTAR)
-# comm_mat %>%
-#   left_join(select(comm_data, mdg_cm_, catch0 = CTAR), by = c("mdg_cm_" = "mdg_cm_")) %>%
-#   mutate(catchment = ifelse(scenario == 0 | scenario == 1, catch0, CTAR)) -> comm_mat
-
-## getting new catchments-> comm_mat ## getting new catchments
-
-# pop <- mada_communes$MDG__201[!is.na(comm_ttimes[1, ])]
-# comm_ttimes <- comm_ttimes[!is.na(comm_ttimes[1, ]), ]
-# comm_ttimes <- as.matrix(comm_ttimes)
-# bites_predicted <- inv.logit(B_ttimes_mada*comm_ttimes/60 + B_0_mada)*pop
-# mu_predicted <- 
-# k_predicted <- 
-# 
-# function(bites, mu, k, times){
-#   for (t in 1:times) {
-#   ## draw bites
-#   
-#   ## do RLEs to get distribution of ts
-#   
-#   ## get # vials used
-#   }
-# }
-
-```
+  summarize(reporting = sum(reporting*prop_pop), ttimes = mean(ttimes, na.rm = TRUE)) -> reporting
+
+ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = params)) +
+  geom_boxplot(outlier.shape = NA) +
+  scale_color_manual(values = c(alpha("blue", 0.5), alpha("purple", 0.5)), name = "Parameters") +
+  xlab("Districts ordered by travel times \n (low to high)") +
+  ylab("Reporting of rabid exposures") +
+  facet_wrap(~ scale) +
+  coord_flip() +
+  theme(axis.text.y = element_blank())
+
+```

### 2019-03-20 accessPEP.Rmd
with section S7 on vials and bite incidence per catchment

@@
 
+We compare the estimates of reporting generated from the commune model vs.
+the district model by summing the commune level reporting estimates weighted by the proportion of the district population they are applied to such that:
+
+\begin{equation}
+           \rho_d = \sum \limits_{j=1}\frac{rho_j*P_j}{P_d}
+\begin{equation}
+
 To estimate the average number of deaths for each administrative unit, we extended the above framework into a stochastic framework as follows:
@@
 
-c / d
+figM3 <- c / d 
+figM3 
+ggsave("figs/M3.jpeg", figM3, device = "jpeg", width = 7, height = 7)
 
@@
 
-```{r fig M4 burden, cache = TRUE}
+```{r run burden and scenario, cache = TRUE}
 ## District travel times
@@
             by = c("names" = "mdg_dis_co")) -> results_all 
-
+```
+
+``` {r fig M4 baseline burden, cache = TRUE}
 ## Plot national level burden
@@
 results_all_base %>%
-  left_join(dist_data, by = c("names" = "district")) -> results_all_base
-ggplot(filter(results_all_base, !is.na(district)), 
+  left_join(dist_data, by = c("names" = "district")) %>%
+  drop_na(names) -> results_all_base
+ggplot(results_all_base, 
        aes(y = deaths_mean/pop*1e5, x = reorder(district, ttimes_weighted), 
                               color = model)) +
-  geom_pointrange(aes(ymin = deaths_lowerCI, ymax = deaths_upperCI), fatten = 2) +
+  geom_pointrange(aes(ymin = deaths_lowerCI/pop*1e5, ymax = deaths_upperCI/pop*1e5), fatten = 2) +
   xlab("Districts ordered by travel times \n (low to high)") +
@@
   coord_flip() +
-  scale_color_manual(values = c("district" = alpha("purple", 0.5), 
-                                "commune" = alpha("navy", 0.5)), 
+  scale_color_manual(values = c(alpha("navy", 0.5), alpha("purple", 0.5)), 
                      name = "Scale") +
@@
 
-burden <- d / p
-
-burden
-
-ggsave("figs/burden_maps.jpeg", burden, device = "jpeg", width = 10, height = 10)
+figM4 <- d / p
+
+figM4
+
+ggsave("figs/M4.jpeg", figM4, device = "jpeg", width = 10, height = 10)
 
@@
 These patterns are also generally robust to our assumptions, although the decrease in deaths saturates quickly for the district model as $p_rabid$ and $rho_max$ decrease?
+Vials...
 
@@
 system.time({dist <- sim.throughput(bites = comm_master$bites_dist, nsims = 3)})
-comm$model <- "commune"
-dist$model <- "district"
+comm$model <- "Commune"
+dist$model <- "District"
 
@@
               alpha = 0.5) +
-  scale_color_manual(values = c("district" = "blue", 
-                                "commune" = "purple"), 
+  scale_color_manual(values = c("navy", "purple"), 
                      guide = "none") +
-  scale_fill_manual(values = c("district" = "blue", 
-                                "commune" = "purple")) +
+  scale_fill_manual(values = c("navy", "purple")) +
   guides(color = guide_legend(), 
@@
 
-ggplot(data = throughput_scenario, aes (x = scenario, y = mean_vials, color = model)) +
-  geom_line() +
-  geom_ribbon(aes(ymin = lower_vials, ymax = upper_vials, 
-                  color = NULL, fill = model),
-              alpha = 0.5) +
-  scale_color_manual(values = c("district" = "blue", 
-                                "commune" = "purple"), 
-                     guide = "none") +
-  scale_fill_manual(values = c("district" = "blue", 
-                                "commune" = "purple")) +
-  guides(color = guide_legend(), 
-         fill = guide_legend(override.aes = list(linetype = 1))) +
-  xlab("# of Additional ARMC") +
-  ylab("Mean estimate of vials needed") +
-  labs(tag = "B")
-
 throughput_scenario %>% left_join(results_natl) -> results_natl_all
@@
 
-e / f
-
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE), 
-            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
-  mutate(model = "commune") %>%
-  drop_na(catchment) -> comm_cmod
-pop_natl <- sum(mada_district$pop2015adj)
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE), 
-            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
-  mutate(model = "district") %>%
-  drop_na(catchment) -> comm_dmod
-bites_exp <- bind_rows(comm_cmod, comm_dmod)
-bites_exp %>%
-  group_by(scenario, model) %>%
-  summarize(bites = sum(bites)/pop_natl*1e5) %>%
-  ggplot(., aes(x = scenario, y = bites, color = model)) +
-  geom_line() +
-  ylab("Reported bites \n per 100k persons") +
-  xlab("# Additional ARMC")
-
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) -> comm_exp
-
-## Do district one
-## Get catchment
-catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
-dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
-dist_ttimes <- cbind(dist_ttimes[, 1], dist_ttimes)
-names(dist_ttimes)[1] <- "result.0"
-dist_mat <- matrix(NA, nrow(dist_ttimes), ncol(dist_ttimes))
-dist_mat[, 1] <- as.character(mada_district$CTAR)
-
-for (i in 2:ncol(dist_ttimes)){
-  dist_mat[, i] <- ifelse(dist_ttimes[, i] < dist_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
-                          dist_mat[, 1])
-}
-
-dist_mat <- as.data.table(dist_mat)
-dist_mat$mdg_dis_co <- mada_district$mdg_dis_co
-dist_mat <- melt(dist_mat, id.vars = "mdg_dis_co", value.name = "catchment", variable.name = "scenario")
-dist_mat$scenario <- as.numeric(gsub("V", "", dist_mat$scenario)) - 1
-dist_ttimes <- as.data.table(dist_ttimes)
-dist_ttimes$mdg_dis_co <- mada_district$mdg_dis_co
-dist_ttimes <- melt(dist_ttimes, id.vars = "mdg_dis_co", value.name = "ttimes", variable.name = "scenario")
-dist_ttimes$scenario <- as.numeric(gsub("result.", "", dist_ttimes$scenario))
-# 
-# dist_mat %>%
-#   left_join(dist_ttimes) %>%
-#   left_join(select(mada_district@data, pop = pop2015adj, mdg_dis_co)) %>%
-#   mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
-#   group_by(scenario, catchment) %>%
-#   summarize(bites = sum(exp_bites, na.rm = TRUE)) %>%
-#   drop_na(catchment) -> dist_master
-
-dist_mat %>%
-  left_join(dist_ttimes) %>%
-  left_join(select(mada_district@data, pop = pop2015adj, mdg_dis_co)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) -> dist_exp
-ggplot(data = filter(dist_exp, scenario %in% c(0, 1, 10, 50, 100, 200)), 
-       aes(x = ttimes/60, y = as.factor(scenario))) +
-  ggridges::geom_density_ridges()
-dist_exp$scale <- "District"
-comm_exp$scale <- "Commune"
-
-check_exps <- bind_rows(select(comm_exp, scenario, ttimes, pop, scale, exp_bites), 
-                        select(dist_exp, scenario, ttimes, pop, scale, exp_bites))
-ggplot(data = filter(check_exps, scenario %in% c(0, 1, 10, 50, 100, 200)), 
-       aes(x = ttimes/60, y = as.factor(scenario), fill = scale)) +
-  ggridges::geom_density_ridges(alpha = 0.5)
-
-ggplot(data = filter(check_exps, scenario %in% c(0, 1, 10, 50, 100, 200)), 
-       aes(x = exp_bites/pop*1e5, y = as.factor(scenario), fill = scale)) +
-  ggridges::geom_density_ridges(alpha = 0.5)
+figM5 <- e / f
+ggsave("figs/M5.jpeg", figM5, device = "jpeg")
 
@@
 ### S1. Estimates of reporting
-```{r}
+```{r fig S1.1 reporting cut-offs}
 ## Getting other data
@@
 
-p + n
+figS1.1 <- p + n
+figS1.1
+ggsave("figs/figS1.1.jpeg", figS1.1, device = "jpeg")
 
@@
 
-```{r excluding contacts}
+```{r fig S2.1 excluding contacts}
 ctar_contacts <- get.contacts(ctar_data, times_sd = 3)
@@
 
-A + B
-
+figS2.1 <- A + B
+ggsave("figs/S2.1.jpeg", figS2.1, device = "jpeg")
 ```
@@
 
-```{r models, cache = TRUE}
+```{r figs S3.1 and S3. 2comparing models, cache = TRUE}
 ## Exclude for comm data!
@@
 
-ggplot(filter(preds_data, pop_predict != "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
+A <- ggplot(filter(preds_data, pop_predict != "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
   geom_point(alpha = 0.5) +
@@
 
-ggplot(filter(preds_data, pop_predict == "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
+B <- ggplot(filter(preds_data, pop_predict == "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
   geom_point(alpha = 0.5) +
@@
 
+A
+B
+ggsave("figs/figS3.1.jpeg", A, device = "jpeg")
+ggsave("figs/figS3.2.jpeg", B, device = "jpeg")
+
 ```
@@
 
-```{r rles and cut-offs, cache = TRUE}
+```{r fig S4.1 and S4.2 rles and cut-offs, cache = TRUE}
 consec_days <- c(5, 15, 30, 60)
@@
 
-ggplot(preds_fixed_df, aes(x = covar, y = predicted, group = interaction(consec_days, scale, contact_cutoff), 
+A <- ggplot(preds_fixed_df, aes(x = covar, y = predicted, group = interaction(consec_days, scale, contact_cutoff), 
                            color = consec_days, linetype = scale), alpha = 0.2) + 
@@
 
-ggplot(preds_data_df, aes(x = observed, y = predicted,
+B <- ggplot(preds_data_df, aes(x = observed, y = predicted,
                        color = scale)) + 
@@
   ylab("Predicted bites")
+A
+B
+ggsave("figs/figS4.1.jpeg", A, device = "jpeg")
+ggsave("figs/figS4.2.jpeg", B, device = "jpeg")
 
@@
 ### S5. Sensitity analyses for burden estimates
-#### District
-```{r scaling w/ pop, cache = TRUE}
+```{r district scaling w/ pop, cache = TRUE}
 ## Dist ttimes
@@
 
-#### Commune
-
 ```{r commune scaling, cache = TRUE}
@@
 
-```{r scaling factor}
-A_dist / A_comm
-```
-
-```{r differences with ttimes}
-C_dist / C_comm
-```
-
-```{r national burden baseline}
-B_dist / B_comm
-
-```
-
-```{r national burden scenario}
-D_dist / D_comm
-
-```
-
-#### Difference between scale examples
-
-```{r scale example}
+```{r fig S5.1 scaling factor}
+figS5.1 <- A_dist / A_comm
+figS5.1
+ggsave("figs/figS5.1.jpeg", figS5.1, device = "jpeg")
+
+```
+
+```{r fig S5.2 differences with ttimes}
+figS5.2 <- C_dist / C_comm
+figS5.2
+ggsave("figs/figS5.2.jpeg", figS5.2, device = "jpeg")
+
+```
+
+```{r fig S5.3 national burden baseline}
+figS5.3 <-B_dist / B_comm
+figS5.3
+ggsave("figs/figS5.3.jpeg", figS5.3, device = "jpeg")
+```
+
+```{r fig S5.4 national burden scenario}
+figS5.4 <- D_dist / D_comm
+figS5.4
+ggsave("figs/figS5.4.jpeg", figS5.4, device = "jpeg")
+```
+
+### S6: Difference between reporting estimates for each model
+
+```{r fig S6.1 scale example}
 ## Comparing scales
@@
 
-ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = params)) +
+figS6.1 <- ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = params)) +
   geom_boxplot(outlier.shape = NA) +
@@
 
-```
+figS6.1
+ggsave("figs/figS6.1.jpeg", figS6.1, device = "jpeg")
+
+```
+
+### S7: Vial estimates and total reported bites with PEP expansion
+```{r fig S7.1 bites and vials}
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE),
+            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
+  mutate(model = "Commune") %>%
+  drop_na(catchment) -> comm_cmod
+comm_mat %>%
+  left_join(comm_ttimes) %>%
+  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
+  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+  group_by(scenario, catchment) %>%
+  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE),
+            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
+  mutate(model = "District") %>%
+  drop_na(catchment) -> comm_dmod
+bites_exp <- bind_rows(comm_cmod, comm_dmod)
+pop_natl <- sum(mada_district$pop2015adj)
+
+ggplot(data = filter(bites_exp, scenario %in% c(0, 1, 10, 50, 100, 200)),
+       aes(x = bites/pop*1e5, y = as.factor(scenario), fill = model)) +
+  ggridges::geom_density_ridges(alpha = 0.5) +
+  scale_fill_manual(values = c("navy", "purple"), name = "Model") +
+  ylab("# Additional ARMC") +
+  xlab("Bites per 100k persons")
+
+```

### 2019-04-02 accessPEP.Rmd
fixed vial analyses and running distance models for burden analyses, etc.

@@
 library(patchwork)
+library(doParallel)
 
@@
 knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
-                      include = TRUE)
+                      include = TRUE, cache.lazy = FALSE)
 ```
@@
 
-figM1
+## figM1
 
@@
 Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC.
-We excluded any catchment areas for which the clinic did not submit any forms (catchments = `r length(no_data)`, districts = `r nrow(exps_dist) - nrow(mada_district)`, in grey in Fig 1A).
+We excluded any catchment areas for which the clinic did not submit any forms (catchments = `r length(no_data)`, districts = `r nrow(mada_district) - nrow(exps_dist)`, in grey in Fig 1A).
 
@@
 We modeled incidence of bites as a function of travel times in hours as follows:
-
-\begin{equation}
+$$
   \mu = exp(\beta_{t}T + \beta_0)P 
-\begin{equation}
+$$
 
@@
 
-\begin{equation}
+$$
   \mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
-\begin{equation}
+$$
 
@@
 
-\begin{equation}
+$$
   \overline\rho = \frac{\mu p_{rabid}}{R}
-\begin{equation}
+$$
 
@@
 
-\begin{equation}
+$$
            p_{rabid}=
@@
             \end{cases}
-\begin{equation}
+$$
 <br>
@@
 
-\begin{equation}
+$$
            \rho_d = \sum \limits_{j=1}\frac{rho_j*P_j}{P_d}
-\begin{equation}
+$$
 
@@
 
-\begin{equation}
+$$
 deaths_i = (R_i - p_{rabid_i}B_i) \times p_{death}
-\begin{equation}
+$$
 
@@
 We estimate burden deterministically across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden.
-Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden.
+Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of $p_{rabid}$ between 0.2 - 0.6 (as per Moramanga), and the range of values of $rho_{max}$ (0.85 - 0.99) to get at maximum and minimum estimates of burden.
 We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see section S5) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
@@
 
-### Models of bites as a function of travel times
+### Models of bites as a function of access
 ```{r fig M2 models of ttimes, cache = TRUE}
@@
                              pop_predict = c("flatPop"),
-                             pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
-                             covars = list(ttimes_district = exps_dist$ttimes_weighted_dist/60,
+                             pop_vars = list(pop_dist = exps_dist$pop, 
+                                             pop_commune = comm_covars$pop),
+                             covars = list(distance_district = exps_dist$distance, 
+                                           distance_commune = comm_covars$distance,
+                                           ttimes_district = exps_dist$ttimes_weighted_dist/60,
                                            ttimes_commune = comm_covars$ttimes_weighted/60), 
-                             sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
+                             sum_it = c(FALSE, TRUE, FALSE, TRUE), 
+                             run_type = "optim", label = "Mada")
+
 mora_mods_covars <- run.mods(bites = exps_mora$bites, names_bites = exps_mora$commune,
@@
                              pop_vars = list(pop_dist = exps_mora$pop),
-                             covars = list(ttimes_mora = exps_mora$ttimes_weighted/60), 
+                             covars = list(distance_mora = exps_mora$distance,
+                                           ttimes_mora = exps_mora$ttimes_weighted/60), 
                              sum_it = c(FALSE, FALSE), run_type = "optim", label = "Mora")
@@
 preds_raw <- run.predicts.data(covar_df = bite_mods, run_type = "predict.raw")
-
+covar_labels <- c("distance" = "Distance (km)", "ttimes" = "Travel times (hrs)")
 a <- ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
   geom_line() + 
-  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
+  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.45) +
   scale_color_manual(values = c("navy", "purple", "red"), name = "Model scale") +
-  scale_fill_manual(values = c("blue", "lavender", "pink"), guide = "none") +
-  xlab("Travel time (hrs)") +
+  scale_fill_manual(values = c("blue", "purple", "pink"), guide = "none") +
+  xlab("Access metric") +
   ylab("Reported bites per \n 100,000 persons") +
-  labs(tag = "A")
+  labs(tag = "A") + 
+  facet_wrap(~covar_name, scales = "free_x", 
+             labeller = labeller(covar_name = as_labeller(covar_labels)))
 
@@
   ylab("Mean annual bites predicted") +
+  facet_wrap(~covar_name, scales = "free_x", 
+             labeller = labeller(covar_name = as_labeller(covar_labels))) +
   labs(tag = "B")
@@
 figM2 <- a / b
-figM2
+## figM2
 
@@
 kable(bite_mods)
+bite_mods %>%
+  mutate(model_scale = 
+           case_when(grepl("district", model, fixed = TRUE) ~ "District",
+                     grepl("commune", model, fixed = TRUE) ~ "Commune", 
+                     grepl("mora", model, fixed = TRUE) ~ "Moramanga"),
+         model_covar = 
+           case_when(grepl("distance", model, fixed = TRUE) ~ "Distance (km)",
+                     grepl("ttimes", model, fixed = TRUE) ~ "Travel time (hrs)")) -> bite_mods
+
+beta_vals <- paste0(round(bite_mods$beta, 3), 
+               paste0(" (", round(bite_mods$beta_lower, 3), " - ", round(bite_mods$beta_upper, 3), ")"))
+intercept_vals <- paste0(round(bite_mods$intercept, 3), 
+               paste0(" (", round(bite_mods$intercept_lower, 3), " - ", round(bite_mods$intercept_upper, 3), ")"))
+likelihood <- round(bite_mods$likelihood, 2)
+kable(cbind(bite_mods$model_covar, bite_mods$model_scale, beta_vals, intercept_vals, likelihood), 
+      col.names = c("Covariate", "Model", "Access effect", "Intercept", "Likelihood"))
 
@@
 
-We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 2A), with reported bite incidence decreasing with travel times to the ARMC.
-The model of bites as a function of travel times at the commune level summed to the district level resulted in a fit closer to the model fit to the Moramanga data at the commune level than the district model.
-All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the models, generally underestimating bite incidence in some districts.
+We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 2A), with reported bite incidence decreasing with increased distance or travel times to the ARMC.
+The models of bites as a function of access at the commune level summed to the district level resulted in a fit closer to the model fit to the Moramanga data at the commune level than the district model.
+All of the models produced reasonable fits to the data (Figure 2B), however there were multiple districts for the models underestimated observed bites from the national data (points below the dashed grey 1:1 line in Figure 2B), although the Commune model seemed to better capture observed bites than the District model in these cases.
 
@@
 ```{r fig M3 reporting}
+## Params from distance models
+B_distance_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop distance_commune"]
+B_0_distance_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop distance_commune"]
+B_distance_dist <- bite_mods$beta[bite_mods$model == "Mada flatPop distance_district"]
+B_0_distance_dist <- bite_mods$intercept[bite_mods$model == "Mada flatPop distance_district"]
+
+## Params from ttime models
 B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
 B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]
-
 B_ttimes_dist <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_district"]
@@
 
-rep_df <- 
+rep_df_ttimes <- 
   foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
@@
   foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
-    rep <- as.data.frame(list(p_rabid = p_rabid_vals[i], ttimes = ttimes, hdr = hdr_vals[k], 
-                                         reporting = get.burden.fixed(pop = 1e5, ttimes = ttimes,
-                                                          B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
-                                            hdr = hdr_vals[k], p_rabid = p_rabid_vals[i],
-                                            type = "reporting", scale = FALSE), scale = names(beta_vals)[j]))
-  }
-
-rep_min_max <- 
+    rep <- as.data.frame(list(p_rabid = p_rabid_vals[i], covar = ttimes, hdr = hdr_vals[k],
+                              reporting = get.burden.fixed(pop = 1e5, covar = ttimes,
+                                                          B_covar = beta_vals[[j]], 
+                                                          B_0 = int_vals[[j]],
+                                                          hdr = hdr_vals[k], 
+                                                          p_rabid = p_rabid_vals[i],
+                                                          type = "reporting", scale = FALSE), 
+                              model = names(beta_vals)[j]))
+  }
+rep_df_ttimes$covar_name <- "Travel times (hrs)"
+rep_min_max_ttimes <- 
   foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
-    rep_max <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
+    rep_max <- get.burden.fixed(pop = 1e5, covar = ttimes, 
+                                B_covar = beta_vals[[j]], B_0 = int_vals[[j]],
                                             hdr = 25, p_rabid = 0.6,
@@
 
-    rep_min <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
-                                            hdr = 5, p_rabid = 0.2,
+    rep_min <- get.burden.fixed(pop = 1e5, covar = ttimes, B_covar= beta_vals[[j]], 
+                                B_0 = int_vals[[j]], hdr = 5, p_rabid = 0.2, 
                                 type = "reporting", scale = FALSE)
-    rep <- as.data.frame(list(ttimes = ttimes, rep_max = rep_max, rep_min = rep_min, 
-                              scale = names(beta_vals)[j]))
-  }
-
-# ggplot(rep_df, aes(x = ttimes, y = reporting, color = interaction(p_rabid, hdr), linetype = as.factor(hdr))) +
-#   geom_line() +
-#   facet_wrap(~ scale) +
-#   scale_color_manual(values = c("red", "red", "blue", "blue"))
-
-c <- ggplot(rep_min_max, aes(x = ttimes, y = rep_min, fill = scale)) +
-  geom_line(aes(y = rep_min, color = scale), linetype = 2) +
-  geom_line(aes(y = rep_max, color = scale), linetype = 1) +
+    rep <- as.data.frame(list(covar = ttimes, rep_max = rep_max, rep_min = rep_min, 
+                              model = names(beta_vals)[j]))
+  }
+rep_min_max_ttimes$covar_name <- "Travel times (hrs)"
+
+beta_vals <- list(District = B_distance_dist, Commune = B_distance_mada)
+int_vals <- list(District = B_0_distance_dist, Commune = B_0_distance_mada)
+distance <- seq(0, 200, by = 5)
+
+rep_df_distance <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
+  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
+  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
+    rep <- as.data.frame(list(p_rabid = p_rabid_vals[i], covar = distance, hdr = hdr_vals[k],
+                              reporting = get.burden.fixed(pop = 1e5, covar = distance,
+                                                          B_covar = beta_vals[[j]], 
+                                                          B_0 = int_vals[[j]],
+                                                          hdr = hdr_vals[k], 
+                                                          p_rabid = p_rabid_vals[i],
+                                                          type = "reporting", scale = FALSE), 
+                              model = names(beta_vals)[j]))
+  }
+rep_df_distance$covar_name <- "Distance  (hrs)"
+
+rep_min_max_distance <- 
+  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
+    rep_max <- get.burden.fixed(pop = 1e5, covar = distance, 
+                                B_covar = beta_vals[[j]], B_0 = int_vals[[j]],
+                                            hdr = 25, p_rabid = 0.6,
+                                            type = "reporting", scale = FALSE)
+    rep_min <- get.burden.fixed(pop = 1e5, covar = distance, B_covar = beta_vals[[j]], 
+                                B_0 = int_vals[[j]], hdr = 5, p_rabid = 0.2, 
+                                type = "reporting", scale = FALSE)
+    rep <- as.data.frame(list(covar = distance, rep_max = rep_max, rep_min = rep_min, 
+                              model = names(beta_vals)[j]))
+  }
+rep_min_max_distance$covar_name <- "Distance (km)"
+
+rep_min_max <- bind_rows(rep_min_max_ttimes, rep_min_max_distance)
+c <- ggplot(rep_min_max, aes(x = covar, y = rep_min, fill = model)) +
+  geom_line(aes(y = rep_min, color = model), linetype = 2) +
+  geom_line(aes(y = rep_max, color = model), linetype = 1) +
   geom_ribbon(aes(ymax = rep_max, ymin = rep_min), alpha = 0.5) +
@@
   ylab("Reporting") +
-  xlab("Travel times (in hours)") +
+  facet_wrap(~covar_name, scales = "free_x") +
+  xlab("Access metric") +
   labs(tag = "A")
@@
     foreach(j = 1:length(hdr_vals), .combine = rbind) %do% {
-    reporting_comm <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
-                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+    comm_ttimes <- get.burden.fixed(pop = comm_data$pop, 
+                                        covar = comm_data$ttimes_weighted/60, 
+                                        B_covar = B_ttimes_mada, B_0 = B_0_mada,
                                         hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
                                         type = "reporting", scale = FALSE)
-    comm <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = reporting_comm, hdr = hdr_vals[j], 
-                               pop = comm_data$pop, ttimes = comm_data$ttimes_weighted,
+    comm_ttimes <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = comm_ttimes, 
+                               hdr = hdr_vals[j], 
+                               pop = comm_data$pop, covar = comm_data$ttimes_weighted/60,
                                p_rabid = p_rabid_vals[i], 
-                               scale = "Commune", params = "Commune model"))
-
-    reporting_dist <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
-                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+                               scale = "Commune", params = "Commune model", 
+                               covar_name = "Travel times (hrs)"))
+    dist_ttimes <- get.burden.fixed(pop = dist_data$pop, 
+                                       covar = dist_data$ttimes_weighted_dist/60, 
+                                       B_covar = B_ttimes_dist, B_0 = B_0_dist,
                                        hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
                                        type = "reporting", scale = FALSE)
-
-     dist <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist, 
-                                pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+    dist_ttimes <- as.data.frame(list(names = dist_data$district, reporting = dist_ttimes, 
+                                pop = dist_data$pop, covar = dist_data$ttimes_weighted_dist,
                                 hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
-                                params = "District model"))
-     check <- bind_rows(comm, dist)
+                                params = "District model", covar_name = "Travel times (hrs)"))
+
+    comm_distance <- get.burden.fixed(pop = comm_data$pop, 
+                                        covar = comm_data$distance, 
+                                        B_covar = B_distance_mada, B_0 = B_0_distance_mada,
+                                        hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                        type = "reporting", scale = FALSE)
+    comm_distance <- as.data.frame(list(names = comm_data$mdg_dis_co, reporting = comm_distance, 
+                               hdr = hdr_vals[j], 
+                               pop = comm_data$pop, covar = comm_data$distance,
+                               p_rabid = p_rabid_vals[i], 
+                               scale = "Commune", params = "Commune model", 
+                               covar_name = "Distance (km)"))
+    dist_distance <- get.burden.fixed(pop = dist_data$pop, 
+                                       covar = dist_data$distance, 
+                                       B_covar = B_distance_dist, B_0 = B_0_distance_dist,
+                                       hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                       type = "reporting", scale = FALSE)
+    dist_distance <- as.data.frame(list(names = dist_data$district, reporting = dist_distance, 
+                                pop = dist_data$pop, covar = dist_data$distance,
+                                hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                params = "District model", covar_name = "Distance (km)"))
+    check <- bind_rows(comm_ttimes, dist_ttimes, comm_distance, dist_distance)
   }
@@
 rep_df %>%
-  group_by(names, hdr, p_rabid, scale, params) %>%
+  group_by(names, hdr, p_rabid, scale, params, covar_name) %>%
   mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
-  summarize(reporting = sum(reporting*prop_pop), ttimes = mean(ttimes, na.rm = TRUE)) -> reporting
-
-d <- ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = scale)) +
+  summarize(reporting = sum(reporting*prop_pop), covar = mean(covar, na.rm = TRUE)) -> reporting
+
+d <- ggplot(reporting, aes(x = reorder_within(names, by = covar, within = covar_name), 
+                           y = reporting, color = scale)) +
   geom_boxplot(outlier.shape = NA) +
   scale_color_manual(values = c(alpha("blue", 0.5), alpha("purple", 0.5)), name = "Parameters") +
-  xlab("Districts ordered by travel times \n (low to high)") +
+  xlab("Districts ordered by access \n (high to low)") +
   ylab("Reporting of rabid exposures") +
+  scale_x_reordered() +
   theme(axis.text.x = element_blank()) +
+  facet_wrap(~covar_name, scales = "free_x") +
   labs(tag = "B")
@@
 figM3 <- c / d 
-figM3 
+## figM3 
 ggsave("figs/M3.jpeg", figM3, device = "jpeg", width = 7, height = 7)
@@
 
-```{r run burden and scenario, cache = TRUE}
+```{r run baseline burden, cache = TRUE}
+# Calculate the number of cores
+no_cores <- 3
+
+distance_dist <- get.burden.stochastic(names = dist_data$district, covar = dist_data$distance,
+                           pop = dist_data$pop, scenario = 0,
+                           param_covar = B_distance_dist,
+                           param_intercept = B_0_distance_dist,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
+
+distance_dist$model <- "District"
+distance_dist$covar_name <- "Distance (km)"
+
+distance_comm <- get.burden.stochastic(names = comm_data$mdg_dis_co, covar = comm_data$distance,
+                           pop = comm_data$pop, scenario = 0,
+                           param_covar= B_distance_mada,
+                           param_intercept = B_0_distance_mada,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
+distance_comm$model <- "Commune"
+distance_comm$covar_name <- "Distance (km)"
+
+### District ttimes
+ttimes_dist <- get.burden.stochastic(names = dist_data$district, 
+                                     covar = dist_data$ttimes_weighted_dist/60,
+                           pop = dist_data$pop, scenario = 0,
+                           param_covar = B_ttimes_dist,
+                           param_intercept = B_0_dist,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
+ttimes_dist$model <- "District"
+ttimes_dist$covar_name <- "Travel times (hrs)"
+
+ttimes_comm <- get.burden.stochastic(names = comm_data$mdg_dis_co, 
+                                       covar = comm_data$ttimes_weighted/60, 
+                           pop = comm_data$pop, scenario = 0,
+                           param_covar= B_ttimes_mada,
+                           param_intercept = B_0_mada,
+                           p_rab_min = 0.2, p_rab_max = 0.6, rho_max = 0.98,
+                           max_HDR = 25, min_HDR = 5, 
+                           dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
+ttimes_comm$model <- "Commune"
+ttimes_comm$covar_name <- "Travel times (hrs)"
+
+results_base <- bind_rows(ttimes_comm, ttimes_dist, distance_comm, distance_dist)
+results_base %>%
+  group_by(scenario, names, model, covar_name) %>%
+  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE),
+            averted_mean = sum(averted_mean, na.rm = TRUE), 
+            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
+            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE), 
+            covar = mean(covar, na.rm = TRUE)) -> results_base_summarized
+results_base_summarized %>%
+  left_join(dist_data, by = c("names" = "district")) %>%
+  drop_na(names) -> results_all_base
+
+ggplot(results_all_base, 
+       aes(y = deaths_mean/pop*1e5, x = reorder_within(names, by = covar, within = covar_name), 
+                              color = model)) +
+  geom_pointrange(aes(ymin = deaths_lowerCI/pop*1e5, ymax = deaths_upperCI/pop*1e5), fatten = 2) +
+  xlab("Districts ordered by access \n (high to low)") +
+  ylab("Annual deaths per 100k persons") +
+  coord_flip() +
+  scale_color_manual(values = c(alpha("navy", 0.5), alpha("purple", 0.5)), 
+                     name = "Scale") +
+  scale_x_reordered() +
+  theme(axis.text.y = element_blank()) +
+  facet_wrap(~covar_name, scales = "free_y") +
+  labs(tag = "A") -> d
+
+gg_district %>%
+  left_join(results_all_base, by = c("id" = "names")) -> gg_district_burden
+
+p <- ggplot() +
+  geom_polygon(data = gg_district_burden, aes(x = long, y = lat, group = group, 
+                                              fill = deaths_mean/pop*1e5), 
+               color = "grey50") +
+  geom_point(data = ctar_points, aes(x = LONGITUDE, y = LATITUDE), color = "blue") + 
+  scale_fill_gradient(low = "white", high = "red", name = "Annual deaths per 100k \n persons") +
+  labs(tag = "B") +
+  facet_grid(covar_name~model)
+
+figM4 <- d / p + plot_layout(ncol = 1, height = c(1, 3))
+
+## figM4
+
+ggsave("figs/M4.jpeg", figM4, device = "jpeg", width = 10, height = 15)
+
+kable(results_all_base %>%
+  group_by(model, covar_name) %>%
+  summarize(averted = sum(averted_mean), 
+            averted_upper = sum(averted_upperCI),
+            averted_lower = sum(averted_lowerCI), 
+            deaths = sum(deaths_mean), 
+            deaths_upper = sum(deaths_upperCI),
+            deaths_lower = sum(deaths_lowerCI),
+            death_vals = paste0(deaths, 
+                                paste0(" (", deaths_upper, " - ", deaths_lower, ")")),
+            averted_vals = paste0(averted, 
+                                paste0(" (", averted_upper, " - ", averted_lower, ")"))) %>%
+            select(`Access metric` = covar_name, Model = model, 
+                   `Burden of deaths` = death_vals, `Deaths averted by PEP` = averted_vals) ->
+  summary_burden, digits = 2)
+
+```
+
+Overall, we estimate average annual deaths between `r round(min(summary_burden$deaths))` - `r round(max(summary_burden$deaths))` and deaths_averted between `r round(min(summary_burden$averted))` - `r round(max(summary_burden$averted))`.
+In general, incidence of rabies deaths increases with travel times, but this effect is more pronounced for the district model compared to the commune model.
+Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust for both models (see figure S5.2).
+Comparing the two models is, the district model generally predicts lower reported bite incidence overall and less steep declines in reporting with higher travel times.
+
+### Expanding access to PEP
+
+```{r run expanded scenario, cache = TRUE}
 ## District travel times
@@
 dist_mat[, 1] <- dist_ttimes[, 1]
+
 for (i in 2:ncol(dist_ttimes)){
@@
 }
+
 dist_mat <- as.data.table(dist_mat)
@@
 
-results_comm_med <- get.burden.stochastic(names = comm_mat$mdg_cm_, ttimes = comm_mat$ttimes/60,
+results_comm_med <- get.burden.stochastic(names = comm_mat$mdg_cm_, covar = comm_mat$ttimes/60,
                            pop = comm_mat$pop, scenario = comm_mat$scenario,
-                           param_ttimes = B_ttimes_mada,
+                           param_covar = B_ttimes_mada,
                            param_intercept = B_0_mada,
@@
                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 2)
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
 
@@
 results_comm_med %>%
-  complete(scenario, names) %>%
+  complete(scenario = 0:max(comm_mat$scenario, na.rm = TRUE), names) %>%
   group_by(names) %>%
@@
 
-results_dist_med <- get.burden.stochastic(names = dist_mat$mdg_dis_co, ttimes = dist_mat$ttimes/60,
+## District
+results_dist_med <- get.burden.stochastic(names = dist_mat$mdg_dis_co, covar = dist_mat$ttimes/60,
                            pop = dist_mat$pop, scenario = dist_mat$scenario,
-                           param_ttimes = B_ttimes_dist,
+                           param_covar= B_ttimes_dist,
                            param_intercept = B_0_dist,
@@
                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
-                           prob_death = 0.16, nsims = 2)
+                           prob_death = 0.16, nsims = 1000, cores = no_cores)
 
 results_dist_med %>%
-  complete(scenario, names) %>%
+  complete(scenario = 0:max(comm_mat$scenario, na.rm = TRUE), names) %>%
   group_by(names) %>%
@@
             by = c("names" = "mdg_dis_co")) -> results_all
-```
-
-``` {r fig M4 baseline burden, cache = TRUE}
+
 ## Plot national level burden
@@
 
-## Baseline results
-results_all %>% filter(scenario == 0) -> results_all_base
-results_all_base %>%
-  left_join(dist_data, by = c("names" = "district")) %>%
-  drop_na(names) -> results_all_base
-ggplot(results_all_base, 
-       aes(y = deaths_mean/pop*1e5, x = reorder(district, ttimes_weighted), 
-                              color = model)) +
-  geom_pointrange(aes(ymin = deaths_lowerCI/pop*1e5, ymax = deaths_upperCI/pop*1e5), fatten = 2) +
-  xlab("Districts ordered by travel times \n (low to high)") +
-  ylab("Annual deaths per 100k persons") +
-  coord_flip() +
-  scale_color_manual(values = c(alpha("navy", 0.5), alpha("purple", 0.5)), 
-                     name = "Scale") +
-  theme(axis.text.y = element_blank()) +
-  labs(tag = "A") -> d
-
-gg_district %>%
-  left_join(results_all_base, by = c("id" = "names")) -> gg_district_burden
-
-p <- ggplot() +
-  geom_polygon(data = gg_district_burden, aes(x = long, y = lat, group = group, 
-                                              fill = deaths_mean/pop2015adj*1e5), 
-               color = "grey50") +
-  geom_point(data = ctar_points, aes(x = LONGITUDE, y = LATITUDE), color = "blue") + 
-  scale_fill_gradient(low = "white", high = "red", name = "Annual deaths per 100k \n persons") +
-  labs(tag = "B") +
-  facet_wrap(~model)
-
-figM4 <- d / p
-
-figM4
-
-ggsave("figs/M4.jpeg", figM4, device = "jpeg", width = 10, height = 10)
-
-kable(results_all_base %>%
-  group_by(model) %>%
-  summarize(averted = sum(averted_mean), 
-            deaths = sum(deaths_mean)) -> summary_burden)
-
-```
-
-Overall, we estimate average annual deaths between `r min(summary_burden$deaths)` - `r max(summary_burden$deaths)` and deaths_averted between `r min(summary_burden$averted)` - `r max(summary_burden$averted)`.
-In general, incidence of rabies deaths increases with travel times, but this effect is more pronounced for the district model compared to the commune model.
-
-Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust (see section S5).
-The district model is more sensitive to assumptions of $p_rabid$ and $\rho_max$ (Fig S5.1).
-
-### Expanding access to PEP
+```
 
@@
 This effect is more pronouned for the commune level model than for the district model.
-These patterns are also generally robust to our assumptions, although the decrease in deaths saturates quickly for the district model as $p_rabid$ and $rho_max$ decrease?
-Vials...
+Vaccine demand also increases, and more vials are needed per averted death as PEP is expanded, with vials needed outpacing the increase in deaths averted.
+These patterns are also generally robust to our model assumptions (see figure S5.3 and S5.4).
 
@@
 catch_scenario <- read.csv("output/temp_ARMC.csv", row.names = 1)
+catch_scenario <- catch_scenario[32:nrow(catch_scenario), ]
 comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)[, 1:231]
@@
 for (i in 2:ncol(comm_ttimes)){
-  comm_mat[, i] <- ifelse(comm_ttimes[, i] < comm_ttimes[, i - 1], as.character(catch_scenario$CTAR[i - 1]), 
-                          comm_mat[, 1])
+  comm_mat[, i] <- ifelse(comm_ttimes[, i] < 
+                            comm_ttimes[, i - 1],
+                          as.character(catch_scenario$CTAR[i - 1]),
+                          NA)
 }
@@
 comm_mat$mdg_cm_ <- mada_communes$mdg_cm_
-comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", variable.name = "scenario")
+comm_mat <- melt(comm_mat, id.vars = "mdg_cm_", value.name = "catchment", 
+                 variable.name = "scenario")
 comm_mat$scenario <- as.numeric(gsub("V", "", comm_mat$scenario)) - 1
+comm_mat %>%
+  drop_na(catchment) %>%
+  complete(scenario = 0:max(comm_mat$scenario, na.rm = TRUE), mdg_cm_) %>%
+  group_by(mdg_cm_) %>%
+  arrange(scenario) %>%
+  fill(3:ncol(.), .direction = "down") -> comm_mat
+
 comm_ttimes <- as.data.table(comm_ttimes)
@@
   left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites_comm = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop, 
-         exp_bites_dist = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
+  mutate(exp_bites_comm = exp(B_ttimes_mada*ttimes/60 + B_0_mada)*pop, 
+         exp_bites_dist = exp(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
   group_by(scenario, catchment) %>%
@@
             bites_comm = sum(exp_bites_comm, na.rm = TRUE),
-            bites_dist = sum(exp_bites_dist, na.rm = TRUE)) %>%
+            bites_dist = sum(exp_bites_dist, na.rm = TRUE), 
+            communes_in_catch = n()) %>%
   drop_na(catchment) -> comm_master
 
-sim.throughput <- function(bites, nsims = 100) {
+# comm_master %>%
+#   group_by(scenario) %>%
+#   summarize(catches = n()) %$% plot(scenario, catches)
+# 
+# ggplot(comm_master, aes(x = as.factor(scenario), y = communes_in_catch)) + geom_boxplot()
+
+sim.throughput <- function(bites, nsims = 100, cores = 3) {
   get.vials <- function(x) {
@@
 
+  # Initiate cluster
+  cl <- makeCluster(cores)
+  registerDoParallel(cl)
+
   check <- foreach(i = 1:nsims, .combine = cbind) %dopar% {
@@
   }
+
+  stopCluster(cl) ## for doParallel
 
@@
 # Calculate the number of cores
-no_cores <- detectCores() - 1
-# Initiate cluster
-cl <- makeCluster(no_cores)
-registerDoParallel(cl)
-system.time({comm <- sim.throughput(bites = comm_master$bites_comm, nsims = 3)})
-system.time({dist <- sim.throughput(bites = comm_master$bites_dist, nsims = 3)})
-comm$model <- "Commune"
-dist$model <- "District"
-
-stopCluster(cl) ## for doParallel
-
-throughput_comm <- bind_cols(comm_master, select(comm, -bites))
-throughput_dist <- bind_cols(comm_master, select(dist, -bites))
+no_cores <- 3
+system.time({comm_tp <- sim.throughput(bites = comm_master$bites_comm,
+                                       nsims = 100, cores = no_cores)})
+system.time({dist_tp <- sim.throughput(bites = comm_master$bites_dist, nsims = 100, 
+                                       cores = no_cores)})
+
+comm_tp$model <- "Commune"
+dist_tp$model <- "District"
+
+throughput_comm <- bind_cols(comm_master, select(comm_tp, -bites))
+throughput_dist <- bind_cols(comm_master, select(dist_tp, -bites))
 throughput <- bind_rows(throughput_comm, throughput_dist)
@@
 figM5 <- e / f
+## figM5
 ggsave("figs/M5.jpeg", figM5, device = "jpeg")
@@
         legend.position = "none") +
-  xlab("") +
+  xlab("ARMC") +
   ylab("Proportion of days included") +
@@
 
-figS1.1 <- p + n
-figS1.1
-ggsave("figs/figS1.1.jpeg", figS1.1, device = "jpeg")
+figS1.1 <- p / n
+## figS1.1
+ggsave("figs/S1.1.jpeg", figS1.1, device = "jpeg", width = 7, height = 7)
 
@@
 
-figS2.1 <- A + B
-ggsave("figs/S2.1.jpeg", figS2.1, device = "jpeg")
+figS2.1 <- A / B
+## figS2.1
+ggsave("figs/S2.1.jpeg", figS2.1, device = "jpeg", height = 7, width = 7)
+
 ```
@@
 
-```{r figs S3.1 and S3. 2comparing models, cache = TRUE}
+```{r figs S3.1 and S3.2 comparing models, cache = TRUE}
 ## Exclude for comm data!
 comm_covars <- filter(comm_data, exclude == 0)
-
-## Correlation of pop + ttimes + distance for Mada districts, Mada communes (highlight Moramanga by color, and report separate corr metric) (color by excluded vs.
-included data and shape = Moramanga for commune level)
-library(GGally)
-ggpairs(data = select(exps_dist, pop, ttimes_weighted_dist, distance))
-ggpairs(data = select(comm_covars, pop, ttimes_weighted, distance))
 
@@
 
-print(AICctab(mada_mods, nobs = nrow(exps_dist), mnames = mapply(function(x) attributes(x)$label, mada_mods), weights = TRUE))
+mada_mod_table <- AICctab(mada_mods, nobs = nrow(exps_dist), mnames = mapply(function(x) attributes(x)$label, mada_mods), weights = TRUE)
+
+delta_aic <- mada_mod_table[["dAICc"]]
+model_names <- attributes(mada_mod_table)$row.names
+kable(cbind(model_names, delta_aic))
 
 mora_mods <- c(mora_mods_covars, mora_mods_pop)
-print(AICctab(mora_mods, nobs = nrow(exps_mora), mnames = mapply(function(x) attributes(x)$label, mora_mods), weights = TRUE))
+mora_mod_table <- AICctab(mora_mods, nobs = nrow(exps_mora), mnames = mapply(function(x) attributes(x)$label, mora_mods), weights = TRUE)
+delta_aic <- mora_mod_table[["dAICc"]]
+model_names <- attributes(mora_mod_table)$row.names
+kable(cbind(model_names, delta_aic))
 
@@
 
+## Correlation of pop + ttimes + distance for Mada districts, Mada communes (highlight Moramanga by color, and report separate corr metric) (color by excluded vs.
+included data and shape = Moramanga for commune level)
+library(GGally)
+figS3.1 <- ggpairs(data = select(dist_data, Population = pop, `Travel times (hrs)` = ttimes_weighted_dist, 
+                      `Distance (km)` = distance)) + labs(tag = "A")
+## figS3.1
+ggsave("figs/S3.1.jpeg", figS3.1, device = "jpeg")
+figS3.2 <- ggpairs(data = select(comm_covars, Population = pop, `Travel times (hrs)` = ttimes_weighted, 
+                      `Distance (km)` = distance)) + labs(tag = "B")
+## figS3.2
+ggsave("figs/S3.2.jpeg", figS3.2, device = "jpeg")
+
 A <- ggplot(filter(preds_data, pop_predict != "onlyPop"), aes(x = observed, y = predicted, color = scale)) + 
@@
   xlab("Observed bites") +
-  ylab("Predicted bites")
+  ylab("Predicted bites") +
+  labs(tag = "A")
 
@@
   geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
-  scale_color_manual(values = c("blue", "purple", "red"), name = "Scale") +
-  facet_grid(pop_predict ~ covar_name, scales = "free_x", 
+  scale_color_manual(values = c("blue", "purple", "red"), guide = "none") +
+  facet_grid(~pop_predict, scales = "free_x", 
              labeller = labeller(pop_predict = as_labeller(mod_names), 
@@
   xlab("Observed bites") +
-  ylab("Predicted bites")
-
-A
-B
-ggsave("figs/figS3.1.jpeg", A, device = "jpeg")
-ggsave("figs/figS3.2.jpeg", B, device = "jpeg")
+  ylab("Predicted bites") +
+  labs(tag = "B")
+
+figS3.3 <- A / (B + plot_spacer()) + plot_layout(ncol = 1, height = c(3, 1))
+## figS3.3
+ggsave("figs/S3.3.jpeg", figS3.3, device = "jpeg")
 
@@
   ylab("Predicted bites")
-A
-B
-ggsave("figs/figS4.1.jpeg", A, device = "jpeg")
-ggsave("figs/figS4.2.jpeg", B, device = "jpeg")
+## A
+## B
+ggsave("figs/S4.1.jpeg", A, device = "jpeg")
+ggsave("figs/S4.2.jpeg", B, device = "jpeg")
 
@@
   check <- get.burden.fixed(names = dist_ttimes$mdg_dis_co, pop = dist_ttimes$pop, 
-                     ttimes = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
-                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     covar = dist_ttimes$ttimes/60, scenario = dist_ttimes$scenario,
+                     B_covar = B_ttimes_mada, B_0 = B_0_mada,
                      p_rabid = p_rabid_vals[i], 
@@
   check <- get.burden.fixed(names = dist_base$mdg_dis_co, pop = dist_base$pop, 
-                     ttimes = dist_base$ttimes/60, scenario = dist_base$scenario,
-                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     covar = dist_base$ttimes/60, scenario = dist_base$scenario,
+                     B_covar = B_ttimes_mada, B_0 = B_0_mada,
                      p_rabid = p_rabid_vals[i], 
@@
 C_dist <- ggplot(data = scenario_base_dist,  
-       aes(x = ttimes, y = deaths/pop*1e5, color = p_rabid)) +
+       aes(x = covar, y = deaths/pop*1e5, color = p_rabid)) +
   geom_point() +
@@
   check <- get.burden.fixed(names = comm_ttimes$mdg_cm_, pop = comm_ttimes$pop, 
-                     ttimes = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
-                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     covar = comm_ttimes$ttimes/60, scenario = comm_ttimes$scenario,
+                     B_covar = B_ttimes_mada, B_0 = B_0_mada,
                      p_rabid = p_rabid_vals[i], 
@@
   check <- get.burden.fixed(names = comm_base$mdg_cm_, pop = comm_base$pop, 
-                     ttimes = comm_base$ttimes/60, scenario = comm_base$scenario,
-                     B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+                     covar = comm_base$ttimes/60, scenario = comm_base$scenario,
+                     B_covar = B_ttimes_mada, B_0 = B_0_mada,
                      p_rabid = p_rabid_vals[i], 
@@
 C_comm <- ggplot(data = scenario_base_comm,  
-       aes(x = ttimes, y = deaths/pop*1e5, color = p_rabid)) +
+       aes(x = covar, y = deaths/pop*1e5, color = p_rabid)) +
   geom_point() +
@@
 
-D_comm <- ggplot(data = deaths_scenario_comm, aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
+D_comm <- ggplot(data = deaths_scenario_comm, 
+                 aes(x = scenario, y = deaths/deaths_base, color = p_rabid)) +
   geom_point() +
@@
 ```{r fig S5.1 scaling factor}
-figS5.1 <- A_dist / A_comm
-figS5.1
-ggsave("figs/figS5.1.jpeg", figS5.1, device = "jpeg")
+figS5.1 <- (A_dist + labs(tag = "A", title = "District")) / (A_comm + labs(tag = "B", title = "Commune"))
+## figS5.1
+ggsave("figs/S5.1.jpeg", figS5.1, device = "jpeg", height = 7, width = 7)
 
@@
 ```{r fig S5.2 differences with ttimes}
-figS5.2 <- C_dist / C_comm
-figS5.2
-ggsave("figs/figS5.2.jpeg", figS5.2, device = "jpeg")
+figS5.2 <- (C_dist + labs(tag = "A", title = "District")) / (C_comm + labs(tag = "B", title = "Commune"))
+## figS5.2
+ggsave("figs/S5.2.jpeg", figS5.2, device = "jpeg", height = 12, width = 12)
 
@@
 ```{r fig S5.3 national burden baseline}
-figS5.3 <-B_dist / B_comm
-figS5.3
-ggsave("figs/figS5.3.jpeg", figS5.3, device = "jpeg")
+figS5.3 <- (B_dist + labs(tag = "A", title = "District")) / (B_comm + labs(tag = "B", title = "Commune"))
+## figS5.3
+ggsave("figs/S5.3.jpeg", figS5.3, device = "jpeg", height = 10, width = 10)
 ```
@@
 ```{r fig S5.4 national burden scenario}
-figS5.4 <- D_dist / D_comm
-figS5.4
-ggsave("figs/figS5.4.jpeg", figS5.4, device = "jpeg")
+figS5.4 <- (D_dist + labs(tag = "A", title = "District")) / (D_comm + labs(tag = "B", title = "Commune"))
+## figS5.4
+ggsave("figs/S5.4.jpeg", figS5.4, device = "jpeg", height = 12, width = 12)
 ```
@@
     foreach(j = 1:length(hdr_vals), .combine = rbind) %do% {
-    reporting_comm <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
-                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+    reporting_comm <- get.burden.fixed(pop = comm_data$pop, covar = comm_data$ttimes_weighted/60, 
+                                            B_covar = B_ttimes_mada, B_0 = B_0_mada,
                                             hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
@@
 
-    reporting_comm_dist <- get.burden.fixed(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted/60, 
-                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+    reporting_comm_dist <- get.burden.fixed(pop = comm_data$pop, covar = comm_data$ttimes_weighted/60, 
+                                            B_covar = B_ttimes_dist, B_0 = B_0_dist,
                                             hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
@@
 
-    reporting_dist <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
-                                            B_ttimes = B_ttimes_dist, B_0 = B_0_dist,
+    reporting_dist <- get.burden.fixed(pop = dist_data$pop, 
+                                       covar = dist_data$ttimes_weighted_dist/60, 
+                                            B_covar = B_ttimes_dist, B_0 = B_0_dist,
                                             hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
@@
 
-     reporting_dist_comm <- get.burden.fixed(pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist/60, 
-                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
+     reporting_dist_comm <- get.burden.fixed(pop = dist_data$pop, covar = dist_data$ttimes_weighted_dist/60, 
+                                            B_covar = B_ttimes_mada, B_0 = B_0_mada,
                                             hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
@@
 
-figS6.1
-ggsave("figs/figS6.1.jpeg", figS6.1, device = "jpeg")
-
-```
-
-### S7: Vial estimates and total reported bites with PEP expansion
-```{r fig S7.1 bites and vials}
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_mada*ttimes/60 + B_0_mada)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE),
-            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
-  mutate(model = "Commune") %>%
-  drop_na(catchment) -> comm_cmod
-comm_mat %>%
-  left_join(comm_ttimes) %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_)) %>%
-  mutate(exp_bites = inv.logit(B_ttimes_dist*ttimes/60 + B_0_dist)*pop) %>%
-  group_by(scenario, catchment) %>%
-  summarize(pop = sum(pop, na.rm = TRUE), ttimes = mean(ttimes, na.rm = TRUE),
-            bites = sum(exp_bites, na.rm = TRUE), n_comms = n()) %>%
-  mutate(model = "District") %>%
-  drop_na(catchment) -> comm_dmod
-bites_exp <- bind_rows(comm_cmod, comm_dmod)
-pop_natl <- sum(mada_district$pop2015adj)
-
-ggplot(data = filter(bites_exp, scenario %in% c(0, 1, 10, 50, 100, 200)),
-       aes(x = bites/pop*1e5, y = as.factor(scenario), fill = model)) +
-  ggridges::geom_density_ridges(alpha = 0.5) +
-  scale_fill_manual(values = c("navy", "purple"), name = "Model") +
-  ylab("# Additional ARMC") +
-  xlab("Bites per 100k persons")
-
-```
+## figS6.1
+ggsave("figs/S6.1.jpeg", figS6.1, device = "jpeg")
+
+```
+
+``` {r comparing to previous results}
+exps <- get.district.data(ctar_data = ctar_data, covars = dist_data, ctar_metadata, 
+                               consec_days = 15, 
+                               contact_cutoff = 3, reporting_thresh = 0.25, 
+                               start = ymd("2014-01-01"), end = ymd("2017-12-31"))
+exps %>%
+  group_by(district) %>%
+  summarise(bites = mean(bites), n = n()) %>%
+  left_join(dist_data) %>%
+  filter(bites != 0) -> exps_dist
+
+get.likelihood.dist <- function(bites, pop, ttimes, params) {
+  beta <- params[1]
+  intercept <- params[2]
+  exp_bites <- inv.logit(beta*ttimes + intercept)*pop
+  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
+}
+
+dist_pars <- optim(par = c(1e-6, 0.1), get.likelihood.dist, bites = exps_dist$bites, 
+                   pop = exps_dist$pop, ttimes = exps_dist$ttimes_weighted_dist/60,
+              hessian = TRUE)
+B_ttimes_dist_wrong <- dist_pars$par[1]
+B_0_dist_wrong <- dist_pars$par[2]
+
+B_0_dist_wrong
+B_ttimes_dist_wrong
+
+rep_df <- 
+  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
+    foreach(j = 1:length(hdr_vals), .combine = rbind) %do% {
+
+    reporting_dist <- get.burden.fixed(pop = dist_data$pop, 
+                                       covar = dist_data$ttimes_weighted_dist/60, 
+                                            B_covar = B_ttimes_dist, B_0 = B_0_dist,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+
+     dist <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist, 
+                                pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+                                hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                params = "District model"))
+
+     reporting_dist_wrong <- get.burden.fixed(pop = dist_data$pop, 
+                                              covar = dist_data$ttimes_weighted_dist/60, 
+                                            B_covar = B_ttimes_dist_wrong, B_0 = B_0_dist_wrong,
+                                            hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], 
+                                      type = "reporting", scale = FALSE)
+
+     dist_wrong <- as.data.frame(list(names = dist_data$district, reporting = reporting_dist_wrong, 
+                                     pop = dist_data$pop, ttimes = dist_data$ttimes_weighted_dist,
+                                     hdr = hdr_vals[j], p_rabid = p_rabid_vals[i], scale = "District",
+                                     params = "Wrong model"))
+
+     check <- bind_rows(dist, dist_wrong)
+  }
+
+rep_df %>%
+  group_by(names, hdr, p_rabid, scale, params) %>%
+  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
+  summarize(reporting = sum(reporting*prop_pop), ttimes = mean(ttimes, na.rm = TRUE)) -> reporting
+
+ggplot(reporting, aes(x = reorder(names, ttimes), y = reporting, color = params)) +
+  geom_boxplot(outlier.shape = NA) +
+  scale_color_manual(values = c(alpha("purple", 0.5), alpha("red", 0.5)), name = "Parameters") +
+  xlab("Districts ordered by travel times \n (low to high)") +
+  ylab("Reporting of rabid exposures") +
+  facet_wrap(~ scale) +
+  coord_flip() +
+  theme(axis.text.y = element_blank())
+
+```

### 2019-04-04 accessPEP.Rmd
fixed rho max vals in se analyses + figM1

@@
   left_join(results_all_base, by = c("id" = "names")) -> gg_district_burden
-
+```
+
+```{r burden figure}
 p <- ggplot() +
@@
                color = "grey50") +
-  geom_point(data = ctar_points, aes(x = LONGITUDE, y = LATITUDE), color = "blue") + 
+  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "blue") + 
   scale_fill_gradient(low = "white", high = "red", name = "Annual deaths per 100k \n persons") +
@@
 ### S5. Sensitity analyses for burden estimates
-```{r district scaling w/ pop, cache = TRUE}
+```{r district scaling w pop, cache = TRUE}
 ## Dist ttimes
@@
 p_rabid_vals <- c(0.2, 0.3, 0.4, 0.5, 0.6)
-rho_max_vals <- c(0.80, 0.85, 0.9, 0.95, 0.9)
+rho_max_vals <- c(0.80, 0.85, 0.9, 0.95, 0.99)
 

### 2019-05-02 accessPEP.Rmd
plotting ttimes

@@
                       include = TRUE, cache.lazy = FALSE)
+
+## Travel times
+ttimes <- raster("output/ttimes_masked.tif")
+gg_SD <- fortify(SD_shape, region = "VILLCODE")
+
+##### Getting raster of coverage
+library(rasterVis)
+library(patchwork)
+library(animation)
+ctar_metadata <- read.csv("data/ctar_metadata.csv")
+gplot(ttimes) +
+  geom_raster(aes(fill = value/60)) +
+  scale_fill_gradient(low = "white", high = "red", na.value = "white", 
+                      name = "Travel times \n (hours)") +
+  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), 
+             color = "blue", alpha = 0.5) +
+  theme_void() +
+  coord_equal()
+
+## Travel times
+ttimes_best <- raster("output/ttimes_best.tif")
+gps_locs <- read.csv(file = "data/ctar_metadata.csv", stringsAsFactors = FALSE)[,c(1, 3, 4)]
+names(gps_locs) <- c ("CTAR", "X_COORD", "Y_COORD")
+
+## candidate points
+csbs <- read.csv("data/csbs.csv", stringsAsFactors = FALSE)
+csbs %>% 
+  dplyr::filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
+  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
+
+point_mat <- rbind(gps_locs, csbs)
+##### Getting raster of coverage
+library(rasterVis)
+library(patchwork)
+library(animation)
+gplot(ttimes_best) +
+  geom_raster(aes(fill = value/60)) +
+  scale_fill_gradient(low = "white", high = "red", na.value = "white", 
+                      name = "Travel times \n (hours)") +
+  geom_point(data = point_mat, aes(x = Y_COORD, y = X_COORD), color = "blue", alpha = 0.25) +
+  theme_void() +
+  coord_equal()
+
+pop10 <- raster("output/pop10.tif")
+prop_pop <- pop10/sum(values(pop10), na.rm = TRUE)
+
+## baseline proportion
+threshold = 3 # threshold # of hours to calculate the proportion of pop at
+sum(prop_pop[ttimes >= 60*threshold], na.rm = TRUE)
+sum(prop_pop[ttimes_best >= 60*5], na.rm = TRUE)
+
 ```

### 2019-09-11 manuscript/accessPEP.Rmd
moved manuscript files to sep directory
also changed rmd so that directory defaults to the .Rproj directory

--- 2019-05-02 accessPEP.Rmd
@@
     includes:
-      in_header: mystyle.sty
+      in_header: format/mystyle.sty
   word_document:
-    reference_docx: format_docs/word_styles_reference_01.docx
+    reference_docx: format/word_styles_reference_01.docx
 ---
@@
 ttimes <- raster("output/ttimes_masked.tif")
-gg_SD <- fortify(SD_shape, region = "VILLCODE")
 
@@
 sum(prop_pop[ttimes >= 60*threshold], na.rm = TRUE)
-sum(prop_pop[ttimes_best >= 60*5], na.rm = TRUE)
+sum(prop_pop[ttimes_best >= 60*3], na.rm = TRUE)
 
@@
 
+### Figure from 538
+distance_summ <- bind_rows(distance_dist, distance_comm)
+distance_summ %>%
+  group_by(scenario, names, model, covar_name) %>%
+  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
+            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
+            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE),
+            averted_mean = sum(averted_mean, na.rm = TRUE), 
+            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
+            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE), 
+            covar = mean(covar, na.rm = TRUE))   %>%
+  left_join(dist_data, by = c("names" = "district")) %>%
+  drop_na(names) -> distance_summ
+
+(sum(distance_comm$deaths_mean)/sum(distance_comm$pop))*1e5
+
+ggplot(data = distance_summ, 
+       aes(y = deaths_mean/pop*1e5, x = reorder_within(names, by = covar, within = covar_name))) +
+  geom_point(data = distance_comm, 
+             aes(y = deaths_mean/pop*1e5, x = reorder_within(names, by = covar, within = covar_name)), 
+             shape = 16, color = "#3a9679", alpha = 0.5) + 
+  geom_point(shape = 22, aes(fill = model), color = "black", alpha = 1, size = 2.5) +
+  xlab("Districts ordered by access \n (high to low)") +
+  ylab("Annual deaths per 100k persons") +
+  geom_hline(yintercept = 1, color = "black", linetype = 2) + 
+  scale_fill_manual(values = c(alpha("#3a9679", 1), alpha("#e4c666", 1)), 
+                     name = "Scale") +
+  scale_x_reordered() +
+  coord_flip() +
+  theme(axis.text.y = element_blank(), 
+        panel.grid.major.y = element_line(color = alpha("grey", 0.25)), 
+        panel.grid.major.x = element_blank(), 
+        panel.grid.minor.x = element_blank(), 
+        panel.background = element_rect(fill = "white")) +
+  # facet_wrap(~covar_name, scales = "free_y") +
+  labs(tag = "A")
+
 ### District ttimes
@@
   coord_flip() +
-  scale_color_manual(values = c(alpha("navy", 0.5), alpha("purple", 0.5)), 
+  scale_color_manual(values = c(alpha("#3a9679", 1), alpha("#e4c666", 1)), 
                      name = "Scale") +

### 2019-10-28 manuscript/summary_Oct24th.Rmd
prelim scripts added

---
title: "summary_Oct23"
author: "Malavika Rajeev"
date: "10/23/2019"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)

##' Libraries
library(rgdal)
library(raster)
library(malariaAtlas)
library(gdistance)
library(doRNG)
library(foreach)
library(dplyr)
library(geosphere)
select <- dplyr::select

## calling in functions
source("R/functions/access_functions.R")

## data
source("R/02_bitedata/03_estimate_biteinc.R") # either source this or output bite data
mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
mada_districts <- readOGR("data/processed/shapefiles/mada_districts.shp")
pop1x1<- raster("data/processed/rasters/worldpop2015adj_mada_1x1km.tif")
ctar_metadata <- read.csv("data/raw/ctar_metadata.csv")
friction_mada_masked <- raster("data/processed/rasters/friction_mada_masked.tif")

```

## The garden of forking paths...
Currently stuck proliferating options.

### Covariate options {.tabset}
```{r maps + rasters, echo = FALSE}
## Baseline surfaces
point_mat <- select(ctar_metadata, Y_COORD = LONGITUDE, X_COORD = LATITUDE)
# ttimes <- get.access(friction = friction_mada_masked, shapefile = mada_districts, 
#                      coords = point_mat, trans_matrix_exists = TRUE, 
#                      filename_trans = "data/processed/rasters/trans_gc_masked.rds", metric = "ttimes")
# writeRaster(ttimes, "data/processed/rasters/ttimes_baseline.tif")
# distance <- get.access(friction = friction_mada_masked, shapefile = mada_districts, 
#                       coords = point_mat, trans_matrix_exists = TRUE, 
#                        filename_trans = "data/processed/rasters/trans_gc_masked.rds", metric = "distance")
# writeRaster(distance, "data/processed/rasters/distance_baseline.tif")

ttimes <- raster("data/processed/rasters/ttimes_baseline.tif")
distance <- raster("data/processed/rasters/distance_baseline.tif")

library(rasterVis)
library(patchwork)

gplot(ttimes) +
  geom_raster(aes(fill = value/60)) +
  scale_fill_gradient(low = "white", high = "red", na.value = "white", 
                      name = "Travel times \n (hours)") +
  geom_point(data = point_mat, aes(x = Y_COORD, y = X_COORD), color = "blue", alpha = 0.25) +
  theme_void() +
  coord_equal()

gplot(distance) +
  geom_raster(aes(fill = value)) +
  scale_fill_gradient(low = "white", high = "red", na.value = "white", 
                      name = "Distance \n (km)") +
  geom_point(data = point_mat, aes(x = Y_COORD, y = X_COORD), color = "blue", alpha = 0.25) +
  theme_void() +
  coord_equal()

mada_districts@data %>%
  select(distcode, ctch_dsct, dist_cent) %>%
  mutate(from_long = coordinates(mada_districts)[, 1], 
         from_lat = coordinates(mada_districts)[, 2]) -> centroids_district
ctar_points <- select(ctar_metadata, CTAR, to_long = LONGITUDE, to_lat = LATITUDE)
segments_plot <- left_join(centroids_district, ctar_points, by = c("ctch_dsct" = "CTAR"))

gg_districts <- fortify(mada_districts, region = "distcode")

catch_cols <- as.character(ctar_metadata$color)
names(catch_cols) <- ctar_metadata$CTAR
```

```{r fig.width = 5, fig.height = 8}
ggplot() +
  geom_polygon(data = gg_districts, aes(x = long, y = lat, group = group), color = "grey50", 
               fill = NA) +
  geom_segment(data = segments_plot, aes(x = to_long, xend = from_long, y = to_lat,
                                          yend = from_lat, color = ctch_dsct)) +
  geom_point(data = segments_plot, aes(x = to_long, y = to_lat, color = ctch_dsct), size = 2) +
  geom_point(data = segments_plot, aes(x = from_long, y = from_lat, color = ctch_dsct), size = 2, 
             alpha = 0.5) +
  scale_color_manual(values = catch_cols, guide = "none") +
  theme_void() +
  labs(title = "Distance by centroid (km)")

```

### Resulting data
```{r}
allbites <- bind_rows(bites_by_distcent, bites_by_distwtd, bites_by_ttimes)

ggplot(allbites, aes(x = covar, y = avg_bites/pop*1e5, 
                            color = catchment)) +
  geom_point() +
  theme(legend.position = "none") +
  scale_color_manual(values = catch_cols, guide = "none") + 
  ylab("Annual bites per 100k") +
  xlab("Districts ordered by \n travel times (high to low)") +
  facet_wrap(~covar_name, scales = "free")

```

### Modeling options

#### Districts with & without CTAR
```{r ctar_bump, echo = FALSE}
ggplot(allbites, aes(x = as.factor(ctar_in_district), y = avg_bites/pop*1e5)) +
  geom_violin() +
  facet_wrap(~covar_name)
```

#### A Random Effect by Catchment
Models with a random intercept by catchment. May help to capture variation between clinics 
not explained by travel times alone. And provide more realistic estimates of vials.

```{r echo = FALSE}
ggplot(allbites, aes(x = as.factor(catchment), y = avg_bites/pop*1e5)) +
  geom_boxplot() +
  facet_wrap(~covar_name) + 
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
```

## Model Results
### Convergence fig
```{r convergence}
model_ests <- read.csv("output/bitemod_results.csv")
## convergence plots
model_ests %>%
  select(data_source, covar_name, data_source, scale, pop_predict, ctar_bump, 
           summed, intercept, val = psrf_upper) %>%
  mutate(type = "psrf_upper") -> psrf
model_ests %>%
  select(data_source, covar_name, data_source, scale, pop_predict, ctar_bump, 
         summed, intercept, val = mpsrf) %>%
  mutate(type = "mpsrf") -> mpsrf
convergence <- bind_rows(mpsrf, psrf)
ggplot(convergence, aes(x = type, y = val, color = interaction(data_source, intercept))) + 
  geom_boxplot() +
  ylim(c(0.90, 1.2)) +
  facet_grid(pop_predict ~ covar_name, scales = "free_x", drop = TRUE) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey")

```

### DIC fig
```{r}
## Estimates
model_ests %>%
  group_by(data_source, covar_name, data_source, scale, pop_predict, ctar_bump, 
           summed, intercept) %>%
  summarize(dic = mean(dic)) -> dic_ests
dic_ests %>%
  group_by(data_source, covar_name) %>%
  arrange(dic, .by_group = TRUE) -> dic_ranks
ggplot(data = dic_ranks, aes(x = pop_predict, y = dic, color = interaction(scale, intercept),
                             shape = ctar_bump))+
  geom_point() +
  facet_grid(data_source ~ covar_name, scales = "free") +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
```

### Estimates
```{r}
model_ests %>%
  select(params, Mean, covar_name, pop_predict, intercept, ctar_bump, summed, data_source,
         scale) %>%
  spread(key = params, value = Mean, fill = 0) -> model_means

## Estimates of access
ggplot(data = filter(model_means, pop_predict != "onlyPop"), 
       aes(x = interaction(scale, data_source), y = beta_access, 
           color = intercept, shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 4) +
  scale_color_manual(values = c("darkred", "#cc7722", "#004b49"), name = "Scale") +
  facet_grid(covar_name ~ pop_predict, scales = "free", drop = TRUE)  +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))

```

### Predictions to Data
```{r}
preds_grouped <- read.csv("output/preds/preds_grouped_all.csv")

ggplot(data = filter(preds_grouped, intercept == "fixed"), 
       aes(x = log(avg_bites + 0.1), y = log(bites_mean + 0.1), 
                                 color = interaction(scale, data_source), shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  scale_color_manual(values = c("darkred", "#cc7722", "#004b49"), name = "Scale") +
  facet_grid(pop_predict ~ covar_name, scales = "free_x", drop = TRUE) +
  expand_limits(y = 0) +
  xlab("log(Observed bites)") +
  ylab("log(Predicted bites)") +
  labs(tag = "A")

ggplot(data = filter(preds_grouped, intercept == "random"), 
       aes(x = log(avg_bites + 0.1), y = log(bites_mean + 0.1), 
           color = interaction(scale, data_source), shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  scale_color_manual(values = c("darkred", "#cc7722", "#004b49"), name = "Scale") +
  facet_grid(pop_predict ~ covar_name, scales = "free_x", drop = TRUE) +
  expand_limits(y = 0) +
  xlab("log(Observed bites)") +
  ylab("log(Predicted bites)") +
  labs(tag = "B") 
```

### Predictions out-of-fit

#### Moramanga estimates on Mada commune and districts

```{r}
outfit_mora <- read.csv("output/preds/outfit_mora.csv")

ggplot(data = filter(outfit_mora, intercept == "fixed"), 
       aes(x = log(avg_bites + 0.1), y = log(bites_mean + 0.1), 
           color = scale, shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  scale_color_manual(values = c("darkred", "#cc7722"), name = "Scale") +
  facet_grid(pop_predict ~ covar_name, scales = "free_x", drop = TRUE) +
  expand_limits(y = 0) +
  xlab("log(Observed bites)") +
  ylab("log(Predicted bites)") +
  labs(tag = "A")

ggplot(data = filter(outfit_mora, intercept == "random"), 
       aes(x = log(avg_bites + 0.1), y = log(bites_mean + 0.1), 
           color = scale, shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  scale_color_manual(values = c("darkred", "#cc7722"), name = "Scale") +
  facet_grid(pop_predict ~ covar_name, scales = "free_x", drop = TRUE) +
  expand_limits(y = 0) +
  xlab("log(Observed bites)") +
  ylab("log(Predicted bites)") +
  labs(tag = "B")

```

#### Mada commune and districts estimates on Moramanga data
```{r}
outfit_grouped <- read.csv("output/preds/outfit_grouped_mada.csv")

ggplot(data = outfit_grouped, 
       aes(x = log(avg_bites + 0.1), y = log(bites_mean + 0.1), 
           color = scale, shape = ctar_bump)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  scale_color_manual(values = c("darkred", "#cc7722", "#004b49"), name = "Scale") +
  facet_grid(pop_predict ~ covar_name, scales = "free", drop = TRUE) +
  expand_limits(y = 0) +
  xlab("log(Observed bites)") +
  ylab("log(Predicted bites)")

```

## Burden estimates

### 2019-10-28 accessPEP.Rmd
deleting ttimes plot code

--- 2019-04-04 accessPEP.Rmd
@@
                       include = TRUE, cache.lazy = FALSE)
+
+##### Getting raster of coverage
+ctar_metadata <- read.csv("data/ctar_metadata.csv")
+
 ```

### 2019-10-29 accessPEP.Rmd
fixed bite incidence estimates
there was a bug in the way left_joins were happening with duplicate records for each year:/

@@
 
-##### Getting raster of coverage
-ctar_metadata <- read.csv("data/ctar_metadata.csv")
-
 ```
@@
 ## read in data
-ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
-ctar_metadata <- read.csv("data/ctar_metadata.csv")
-mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
-mada_communes <- readOGR("output/communes/communes_extracted.shp")
-
-## catchment data
-dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
-names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")
-comm_catchments <- read.csv("output/catchments_commune_unmasked.csv", row.names = 1)
-names(comm_catchments) <- c("mdg_cm_", "CTAR", "ttimes_weighted")
+ctar_data <- read.csv("data/raw/bitedata/peripheral/SaisieRage_DATA_2018-09-21_1755.csv")
+ctar_data$district <- gsub("D", "", ctar_data$district, fixed = TRUE)
+ctar_metadata <- read.csv("data/raw/ctar_metadata.csv")
+mada_district <- readOGR("data/processed/shapefiles/mada_districts.shp")
+mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
 
@@
 ## Getting ctar district and commune
-ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
-                                                           mada_district$district)]
+ctar_metadata$ctar_dist <- mada_district$distcode[match(ctar_metadata$District, 
+                                                           mada_district$ADM2_EN)]
 pts <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                    proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
-ctar_metadata$ctar_comm <- over(pts, mada_communes)$mdg_cm_
-dist_catchments$ctar_in_dist <- ifelse(dist_catchments$mdg_dis_co %in% ctar_metadata$ctar_dist, 1, 0)
-comm_catchments$ctar_in_comm <- ifelse(comm_catchments$mdg_cm_ %in% ctar_metadata$ctar_comm, 1, 0)
+ctar_metadata$ctar_comm <- over(pts, mada_communes)$ADM3_PCODE
 
@@
 
-dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
-                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
-mada_district@data <- left_join(mada_district@data, dist_catchments) ## need this?
-
 ## Getting commune covars
-## Need: weighted ttimes, pop, ctar_in, exclude
+## Need: weighted ttimes, pop, exclude
 ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
-comm_catchments %>%
-  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) %>%
-  left_join(select(mada_district@data, mdg_dis_co, district, ctar_in_dist, dist_catch = CTAR, exclude,
-                   color, fill, ctar_dist)) -> comm_data
+mada_communes@data %>%
+  select(ttimes_weighted = ttms_wtd, pop, mdg_cm_ = ADM3_PCODE, distcode = ADM2_PCODE) %>%
+  left_join(select(mada_district@data, distcode, dist_catch = ctch_ttwtd)) -> comm_data
+
 mada_communes@data %>%
@@
 ## Getting district covars
-district_coords <- coordinates(mada_district)
-distance_mat <- distm(district_coords, ctar_coords)
-mada_district$dist_min <- apply(distance_mat, 1, min)/1000
 mada_district@data %>%
-  select(district = mdg_dis_co, dist_name = district, pop = pop2015adj, 
-         ttimes_weighted_dist = ttimes_weighted, 
-         dist_catch = CTAR, distance = dist_min, exclude, color) -> dist_data
+  select(district = distcode, dist_name = ADM2_PCODE, pop, 
+         ttimes_weighted_dist = ttms_wtd, 
+         dist_catch = ctch_ttwtd, distance = dist_cent) -> dist_data
+dist_data$exclude <- ctar_metadata$exclude[match(dist_data$dist_catch, ctar_metadata$CTAR)]
 

### 2019-12-04 supplementary.Rmd
supplementary materials rmd

---
title: "Supplementary Materials"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: format/mystyle.sty
  word_document:
    reference_docx: format/word_styles_reference_01.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## S1. Estimating travel times to the nearest ARMC

```{r ttimes}
```

## S2. Correcting data for underreporting and excluding Category I exposures

### Underreporting of forms
Figure S1.1 correcting
```{r reporting}
```
Figure S1.2 vial ests

### Cat I exposures
Figure S1.3 contacts
```{r cat I}
```
Figure S1.4 ttime differences

## S3. Candidate models of bite incidence

## S4. Sensitivity of these models to reporting/contact cut_offs

## S5. Incrementally adding clinics (where they were added, shifts in ttimes + catchment pops, shifts in burden/reporting etc. --answer q here: why are there still abt 100 deaths)

## S6. Sensitivity of burden estimates to parameter assumptions (baseline + incremental)

### 2019-12-05 manuscript/supplementary.Rmd
trying to figure out md docs on github

---
title: "Supplementary Materials"
author: ""
date: 
output:
  html_document: 
    keep_md: TRUE
  pdf_document:
    includes:
      in_header: format/mystyle.sty
  word_document:
    reference_docx: format/word_styles_reference_01.docx
---

```{r setup, include=FALSE}
library(knitr)
library(bookdown)
opts_chunk$set(echo = FALSE)
```

## S1. Estimating travel times to the nearest ARMC 
We used two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project (REF) at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop raster of population originally at an N resolution. We then extracted the mean of travel times weighted by the population in that grid cell for each district or commune to get administrative level estimates of travel times to the nearest ARMC. Figure S1.1 shows the raster inputs as well as the resulting estimates of travel times at the admin level.

```{r fig S1.1, fig.width=10, fig.height=10, fig.fullwidth=TRUE, fig.cap=cap}
library(here)
include_graphics(here("figs/S1.1.jpeg"))
cap <-"checking!"

```

## S2. Correcting data for underreporting and excluding Category I exposures

### Underreporting of forms
Figure S1.1 correcting
```{r reporting}
```
Figure S1.2 vial ests

### Cat I exposures
Figure S1.3 contacts
```{r cat I}
```
Figure S1.4 ttime differences

## S3. Candidate models of bite incidence

## S4. Sensitivity of these models to reporting/contact cut_offs

## S5. Incrementally adding clinics (where they were added, shifts in ttimes + catchment pops, shifts in burden/reporting etc. --answer q here: why are there still abt 100 deaths)

## S6. Sensitivity of burden estimates to parameter assumptions (baseline + incremental)

### 2019-12-05 R/figures/SM1_base_ttimes.R

@@ -41,10 +41,13 @@ names(ttime_cols) <- ttime_labs
 
 figS1.A <- gplot(base_times) + 
   geom_raster(aes(fill = cut(value/60, breaks = ttime_breaks, labels = ttime_labs))) + 
-  scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)") +
-  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "darkgrey", shape = 4,
+  scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)",
+                    drop = FALSE) +
+  geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "darkgrey", 
+             shape = 4,
              stroke = 2) +
   theme_void() +
+  theme(text = element_text(size = 14)) +
   labs(tag = "A")
 
 figS1.B <- gplot(pop) + 
@@ -52,26 +55,31 @@ figS1.B <- gplot(pop) +
   scale_fill_distiller(type = "seq", palette = "GnBu", direction = 1, na.value = "white",
                    name = expression("Log(population) \n per 1km"^2)) +
   theme_void() +
+  theme(text = element_text(size = 14)) +
   labs(tag = "B")
 
 figS1.C <- ggplot() +
   geom_polygon(data = gg_districts, aes(x = long, y = lat, group = group, 
                                         fill = cut(ttimes_wtd/60, breaks = ttime_breaks, 
                                                    labels = ttime_labs))) + 
-  scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)") +
+  scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)",
+                    drop = FALSE) +
   geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "darkgrey", shape = 4,
              stroke = 2) +
   theme_void() +
+  theme(text = element_text(size = 14)) +
   labs(tag = "C")
 
 figS1.D <- ggplot() +
   geom_polygon(data = gg_communes, aes(x = long, y = lat, group = group, 
                                        fill = cut(ttimes_wtd/60, breaks = ttime_breaks,
                                                   labels = ttime_labs))) + 
[…]

### 2019-12-05 manuscript/supplementary.Rmd
getting rid of extra files

@@
 output:
-  html_document: 
-    keep_md: TRUE
+  github_document: 
+  html_document: default
   pdf_document:
@@
 We then extracted the mean of travel times weighted by the population in that grid cell for each district or commune to get administrative level estimates of travel times to the nearest ARMC.
-Figure S1.1 shows the raster inputs as well as the resulting estimates of travel times at the admin level.
+Figure S1 shows the raster inputs as well as the resulting estimates of travel times at the admin level.
 

### 2019-12-05 manuscript/supplementary.Rmd
.gitignore fix

--- 2019-12-05 manuscript/supplementary.Rmd
@@
 output:
+  word_document:
+    reference_docx: format/word_styles_reference_01.docx
   html_document:
-    keep_md: TRUE
+  github_document:
   pdf_document:
@@
       in_header: format/mystyle.sty
-  word_document:
-    reference_docx: format/word_styles_reference_01.docx
+bibliography: ["../manuscript/references.bib"]
 ---
@@
 ## S1. Estimating travel times to the nearest ARMC 
-We used two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project (REF) at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop raster of population originally at an N resolution.
-We then extracted the mean of travel times weighted by the population in that grid cell for each district or commune to get administrative level estimates of travel times to the nearest ARMC.
-Figure S1.1 shows the raster inputs as well as the resulting estimates of travel times at the admin level.
+To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop ([worldpop.org](https://www.worldpop.org)) raster of population originally at an ~100m^2 resolution.
+We extracted the mean of travel times (Fig S1.1A) weighted by the population in that grid cell (Fig S1.1B) for each district or commune.
+Figure S1.1 shows the raster inputs as well as the resulting estimates of travel times at the district (Fig S1.1D) and commune (Fig S1.1D) level.
 
-```{r fig S1.1, fig.width=10, fig.height=10, fig.fullwidth=TRUE, fig.cap=cap}
-library(here)
-include_graphics(here("figs/S1.1.jpeg"))
-cap <-"checking!"
-
+```{r fig S1.1, fig.fullwidth=TRUE}
+include_graphics("../figs/S1.1.jpeg")
 ```
+_**Figure S1.1.
+A)** Travel time estimates at an approximately 1x1 km scale **B)** Population per 1x1 km grid cell.
+Mean travel times weighted by population size extracted for each **C)** District and **D)** Commune._
 
@@
 
-### Underreporting of forms
+### Estimating underreporting of forms
 Figure S1.1 correcting
@@
 
-### Cat I exposures
+### Excluding Category I exposures
 Figure S1.3 contacts
@@
 ## S6. Sensitivity of burden estimates to parameter assumptions (baseline + incremental)
+
+## References

### 2019-12-31 docs/accessPEP.Rmd
wip main manuscript file

---
title: |
 and potential for expanded access to prevent human rabies deaths in Madagascar
 How accesss to care shapes disease burden: the current impact of post-exposure prophylaxis 
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa:
      institute: [IPMepi]
      correspondence: false
  - Reziky:
      institute: [IPMepi]
      correspondence: false
  - Ravo:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::gitbook:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib"]
---

<!-- To Do: -->
<!-- 1) Go through methods + results -->
<!--   a) update figs -->
<!--   b) update references -->
<!--   c) add in stats -->
<!-- 2) Work through comments in discussion per KH -->
<!-- 3) Insert citations for discussion -->
<!-- 4) Work through introductions -->
<!-- 5) Think through doc workflow (how to split docs & refs) -->

```{r setup, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)

opts_chunk$set(echo = FALSE)
out_main_fig = TRUE

## data
ctar_metadata <- fread("../data/raw/ctar_metadata.csv")
district_bites <- fread("../output/bites/district_bites.csv")
mora_bites <- fread("../output/bites/mora_bites.csv")
mada_communes <- readOGR("../data/processed/shapefiles/mada_communes.shp")
mada_districts <- readOGR("../data/processed/shapefiles/mada_districts.shp")

```
\newpage

# ABSTRACT

\newpage

# INTRODUCTION
[see figure 1 A](#fig1A)

# METHODS
## Spatial covariate data {#methods-spatial}
To calculate travel times, we used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale. We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see section S1](#S1)). Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, @linardPopulationDistributionSettlement2012) and aggregated to the commune and district levels.

## Bite patient data {#methods-bitedata}
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually. These data include details of the administrative district of the bite patient’s address and the date of reporting.

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1.2). Therefore, we defined the catchment for each clinic as all districts for which the clinic was the closest ARMC for the majority of the population in that district. We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 1A).

```{r fig 1, fig.fullwidth=TRUE}
include_graphics("../figs/main/M1.jpeg")
```

*[__Figure 1.__]{#fig1} __A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data. Color corresponds to the clinic catchment. Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients. Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. __B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district. Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic. __C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis). Colors correspond to the clinic catchment, 
shape indicates the dataset, and the size of the point indicates the number of observations (i.e. the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*

For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted. We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2.1A and B). To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level. We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics (Fig S2.1C, [further details in section S2](#S2)).

```{r contacts stats}
contacts_mora <- fread("../output/sensitivity/contacts_mora.csv")
contacts_natl <- fread("../output/sensitivity/contacts_natl.csv")
prop_excluded_natl <- (sum(contacts_natl[estimated_cat1 == 1]$no_patients) + sum(contacts_natl[estimated_cat1 == 0 & !is.na(no_known)]$no_known))/sum(contacts_natl$no_patients)

```

Previous work in the Moramanga District showed that Category I exposures, i.e. touching or feeding a suspected animal or human rabies case, make up approximately 20% of patients reporting to ARMCs [@rajeevHealthcareUtilizationProvisioning2018]. These low risk exposures do not contribute to the mortality burden of rabies. However, we could not directly exclude them as details on exposure type was missing for the majority of the patient forms. Detailed data from Moramanga indicates that these contacts consistently present as clustered cases and the majority report on a single day. Accordingly, we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure S2.2A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) excluded approximately `r round(contacts_mora[prop == "contacts" & sd == 3.00]$excluded*100, 2)`% of Category I exposures, with only `r round(contacts_mora[prop == "bites" & sd == 3.00]$excluded*100, 2)`% of Category II and III patients excluded (Figure S2.2B). We further excluded patients which were explicitly noted to be Categort I exposures but were not identified based on clustering of presentation dates, resulting in the exclusion of approximately `r round(prop_excluded_natl*100, 2)`% of patient records from the national data. After excluding patients likely to reflect Category I exposures as described, we estimated average bite incidence as the mean of the yearly bites reported/the proportion of bite patient data which were submitted.

The 31 months of data available from the Moramanga district resolved to the commune level (the administrative level below the district, Figure 1B) provide a useful comparison to the national scale data. These data were completely reported with details on the type of exposure documented. We calculated average annual bite incidence to the commune level by excluding Category I exposures and taking the monthly average of reported bites and multiplying by 12. Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C).

## Model of reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled incidence of bites as a function of travel time ($T$) as follows: 
$$ \mu = exp(\beta_{t}T \ + \ \beta_0)P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC, and the human population size, $P$ (an offset which scales the predicted number of bites), for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j\ + \ \beta_0)P_j$$

As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel time. For the models with population size, we used a model framework with either population size alone ($\mu = exp(\beta_pP + \beta_0)$) or population size and travel times ($\mu = exp(\beta_{t}T + \beta_pP + \beta_0)$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercpet).

To test our models predictions further, we used parameter estimates from each model to generate predictions for data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data).

## Estimating burden of rabies deaths {#methods-burden}
We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@consortiumPotentialImpactImproved2019]. To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs and human exposure incidence across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating]. $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

$B_i$, the number of total reported bites, is drawn from a poisson distribution with the mean predicted number of bites predicted from the models of incidence as a function of travel times. To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. We draw $p_{rabid}$ from a uniform distribution ranging from 0.2 - 0.6 based on investigations of bite patients reporting in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018]. So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
where $\rho_{max}$ is the maximum reporting. We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018]. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. 
We subtract $R_i$ from $E_i$ to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated as:
$$ D_i = binom(U_i,\ p_{death}) $$
Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_i = binom(R_i, \ p_{death})$). Table 1 lists all parameter values and sources used in the decision tree. \

Table:**Table 1.** Parameters used in the decision tree to estimate human rabies deaths at the administrative level.\

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@consortiumPotentialImpactImproved2019] |
| $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$. | [@consortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
| $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

## Strategically expanding PEP access {#methods-incremental}

```{r ncsbs}
## candidate points
csbs <- read.csv("../data/raw/csbs.csv", stringsAsFactors = FALSE)
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

We developed a framework to rank clinics by how much they could improve access for people currently underserved (< 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics. We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric weighted by the reduction in travel times, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also estimate travel times given that all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We also estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment. As clinics are added, multiple clinics may serve a single district, but in general, the majority of the population within a given commune are closest to a single clinic (Fig S5.3B). Therefore, we assign the catchment clinic as the one which is the closest for the majority of people living in a commune. For the district model, we disaggregate the reported bites to the commune level (assuming uniform incidence within the district). For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 100 simulations of throughput for each catchment and set of clinics.

## Sensitivity analyses for burden {#methods-sensitivity}
We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplement section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.

# RESULTS

## Travel times as a predictor of bites {#results-bitemods}
All of the candidate models which included travel times produced reasonable predictions when compared to the observed data (Figure SY). Population size alone was the poorest fit to the data as estimated by DIC, and all models with population size as a covariate did not generate realistic predictions to the data or when used to predict out of fit. For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Figure 3B, Figure SX). In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Figure 3A). As these models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence and burden. 
_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons A) the estimated relationship between travel times in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis). The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept. The envelopes shows the 95% CI given the random intercept for the commune and district models. The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept.  B) Log(observed) vs. log(predicted) mean annual reported bites for the three different models (added a small positive offset of 0.1 to account for log of zeros)._

## Estimation of baseline burden {#results-burden}
Overall, we estimate between 306 - 373 average annual deaths in Madagascar due to rabies depending on the model used (Table 2). We also estimate that under the current system of 31 ARMCs in Madagascar, approximately 483 - 548 deaths due to rabies are prevented through provisioning of PEP each year. In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally at the district level, but also within districts (Figure 4).

_**Figure 4.** Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons. A) for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district. Points are colored by and districts are ordered by travel times. The dashed lines show the average national incidence of human rabies deaths for the commune (red) and district (blue) models. Incidence mapped to the B) commune level and C) district level from the respective models; grey x’s show locations of current CTAR._

table:*__Table 2.__ Model predictions of mean annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% CIs in parentheses.*

## Incremental reduction of burden with expanded PEP {#results-incremental}
We use our travel time models to look at the benefits of expanding PEP incrementally. We find that expanding access to PEP reduces rabies deaths, but benefits accrue less rapidly  as ARMC are added, saturating after approximately 200 clinics. This effect is more pronounced for the commune level model than for the district model. Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded, with vials needed outpacing the increase in deaths averted. **Report actual # of vials needed as PEP is expanded.** Even with maximal PEP access, our model still predicts > 100 rabies deaths per annum, as average reporting stays at x% (see section S6 to see how things shift!)

figure:_**Figure 5.** A) Decrease in deaths due to rabies and B) increase in vials needed per death averted as additional ARMC are added at the national level based on the two models of reported bites. Lines are the mean of 1000 simulations with envelopes representing 95% CIs. The “...” on the axes indicates a break in the axes, and the last point is the burden and vial predictions when all CSBS in the country provision PEP (max access)._

## Sensitivity analyses 
There is significant uncertainty in the predicted numbers of rabies deaths, but the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (see Supplementary Materials, section S5). Our estimate of the absolute number of deaths is most sensitive to our assumptions of rabies exposure incidence in the population.

# DISCUSSION
## Main findings

## Strengths and Limitations

## Broader Context

## Conclusion

<!-- # REFERENCES -->

<!-- ::: {#refs_main} -->
<!-- ::: -->

# SUPPLEMENTARY MATERIALS AND METHODS

## S1. Estimating travel times to the nearest ARMC {#S1}
[*jump back to main methods section*](#methods-spatial)  
To estimate travel times to the nearest clinic at the scale of administrative units, we generated two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop ([worldpop.org](https://www.worldpop.org)) raster of population originally at an ~100m^2 resolution. We extracted the mean of travel times (Fig S1.1A) weighted by the population in that grid cell (Fig S1.1B) for each district or commune. Figure S1.1 shows the rasters we generated as well as the resulting extracted estimates of travel times at the district (Fig S1.1D) and commune (Fig S1.1D) level. Raw inputs are availble through World Pop and MAP[@abbasRabiesControlInitiative2011].

```{r fig S1.1, fig.fullwidth=TRUE}
include_graphics("../figs/supplementary/S1.1.jpeg")
```
_**Figure S1.1. A)** Travel time estimates at an approximately 1x1 km scale **B)** Population per 1x1 km grid cell. Mean travel times weighted by population size extracted for each **C)** District and **D)** Commune._

We assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit. Figure S1.2A shows that for most districts and communes a single clinic served the majority of the population by this metric. In addition, our data show that the majority of bite patients do report to the closest clinic as assigned by our travel time metric (Fig S1.2B).

```{r fig S1.2, fig.fullwidth=TRUE}
include_graphics("../figs/supplementary/S1.2.jpeg")
```
_**Figure S1.1. A)** Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned. **B)** The proportion of bites reported to each clinic which originated from a district within the assigned catchment._

## S2. Correcting data for underreporting and excluding Category I exposures {#S2}

### Estimating under-submission of forms
To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted (Figure S2.1A). We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period. We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.

Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared + any additional wasteage from unused doses discarded at the end of the day. While these simplifying assumptions do not capture variation in adherence, completion, delays, or errors in administration, PEP completion is overall high in Madagascar and in other settings where PEP is provided at no direct costs to patients, so our estimates at 3 doses per patient vs. 4 are likely encompassing much of this variation at the annual level.

We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial underreporting (Figure S2.1B). Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned.

Figure S1.1 correcting
```{r reporting}
```
Figure S1.2 vial ests

### Excluding Category I exposures
We also excluded Category I exposures because they do not contribute to rabies risk and more importantly follow a different travel time distribution. Because they are usually actively referred to clinics (Moramanga data fig). They also present clustered on the same dates (Moramanga data fig). So we excluded by this clustering.

Figure S1.3 contacts
```{r cat I}
```
Figure S1.4 ttime differences

Because there was variation in estimates of reporting, and in our resulting incidence estimates we looked at how this impacted estimates of travel time effects (Supplementary Materials S4).

## S3. Candidate models of bite incidence
We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations). We used priors centered around zero (*norm(0, 10-6)*) except for the precision parameter for which we used an uniform distribution (*unif(0, 100)*).\

The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment. We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book). All models and posterior estimates of parameters converged (Fig S3.1, both psrf and mpsrf approach 1). Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not significantly improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1). However, when used to predict data that were not used to fit the models (i.e. model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop, onlyPop) generated reasonable predictions to the data (Figure S3.3).\

```{r table 1}
model_ests <- read.csv("../output/mods/estimates.csv")
model_ests %>%
  select(params, Mean, pop_predict, intercept, data_source, scale) %>%
  spread(key = params, value = Mean, fill = 0) -> model_means
model_ests %>%
  group_by(data_source, scale, pop_predict, intercept) %>%
  summarize(dic = mean(dic)) -> dic_ests
dic_ests %>%
  group_by(data_source) %>%
  arrange(dic, .by_group = TRUE) %>%
  mutate(dic = round(dic, 2)) -> dic_ranks
knitr::kable(dic_ranks, 
             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"), 
             caption = "Table S3.1. DIC Estimates for all models. Models with the lowest DIC for each dataset and scale of covariate are in bold.  For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model. For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).")
```
<br/>

```{r psrf}
include_graphics("../figs/supplementary/S3.1.jpeg")
```
_**Figure S3.1. Potential scale reduction factors (psrf) for all models.** The x-axis is the type of  psrf, either multivariate which assesses the convergence of the model, or the individual which assesses the convergence for each parameter estimated from the model. Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model). Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\  
<br/>

```{r s3.2, fig.fullwidth=TRUE}

include_graphics("../figs/supplementary/S3.2.jpeg")

```
_**Figure S3.2. Prediction to data used to fit each model.** Log of the observed bites against the predicted bites using the mean of the posterior distributions for each parameter. Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model). Colors show which data set  was used for fitting and the scale of the model (Moramanga = Moramanga data, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\

<br/>

```{r s3.3, fig.fullwidth=TRUE}
include_graphics("../figs/supplementary/S3.3.jpeg")
```
_**Figure S3.3. Out of fit predictions to data.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data not used to fit the model. A) Estimates from commune and district model fitted to the national data predicting the Moramanga data. B) Estimates from models fitted to the Moramanga data predicting the national data at the commune and district scale. Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model). Colors show which data set was used for fitting (Commune = National data with covariates at the commune level, District = National data with covariates at the district level, A) and the scale of the model at which covariates were applied (Commune = covariates at the commune scale, District = covariates at the District scale)._\  
<br/>

## S4. Sensitivity of these models to reporting/contact cut_offs

For all resulting data sets used to fit models, the commune scale model with population as an offset and a random catchment effect had the lowest DIC.

```{r Table S4.1}
## DIC Tables
## DIC Tables
# model_se %>%
#   group_by(scale, contacts, reporting, intercept, pop_predict) %>%
#   summarize(dic = dic[1]) %>%
#   arrange(reporting, contacts, dic) %>%
#   mutate(colnames = interaction(reporting, contacts),
#          colnames = case_when(colnames == "15.3" ~ "Corrected for underreporting & Cat I excluded", 
#                               colnames == "Inf.3" ~ "Cat I excluded only", 
#                               colnames == "15.Inf" ~ "Corrected for underreporting only", 
#                               colnames == "Inf.Inf" ~ "None (raw data)")) %>%
#   ungroup() %>%
#   select(-contacts, - reporting) %>%
#   pivot_wider(names_from = colnames, values_from = dic) -> dic_table
# 
# knitr::kable(dic_ranks, 
#              col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"), 
#              caption = "Table S4.1. Comparing DIC ranks for models using datasets with different data processing. For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model. For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).")

```

## S5. Incrementally adding clinics (where they were added, shifts in ttimes + catchment pops, shifts in burden/reporting etc. --answer q here: why are there still abt 100 deaths)

## S6. Sensitivity of burden estimates to parameter assumptions (baseline + incremental)

<!-- ## Supplementary References -->

<!-- ::: {#refs_supp} -->
<!-- ::: -->

# REFERENCES

### 2019-12-31 docs/manuscript.Rmd
fixing title

--- 2019-12-31 docs/accessPEP.Rmd
@@
 title: |
+ How accesss to care shapes disease burden: the current impact of post-exposure prophylaxis 
  and potential for expanded access to prevent human rabies deaths in Madagascar
- How accesss to care shapes disease burden: the current impact of post-exposure prophylaxis 
 author:

### 2020-02-18 docs/manuscript.Rmd
updating organization of man

@@
 ---
-title: |
- How accesss to care shapes disease burden: the current impact of post-exposure prophylaxis 
- and potential for expanded access to prevent human rabies deaths in Madagascar
+title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
 author:
@@
       correspondence: false
-  - Anjasoa:
+  - Anjasoa Randrianarijaona:
       institute: [IPMepi]
       correspondence: false
-  - Reziky:
+  - Reziky Tiandraza Mangahasimbola:
       institute: [IPMepi]
       correspondence: false
-  - Ravo:
+  - Ravo Ramiandrasoa:
       institute: [IPMvacc]
@@
     pandoc_args:
-      - --csl=format/plos-neglected-tropical-diseases.csl
       - --filter=pandoc-crossref
@@
       - --reference-doc=format/Reference_Document.docx
- bookdown::gitbook:
+ bookdown::html_document2:
     toc: true
@@
 bibliography: ["bib/refs_all.bib"]
+# bibliography_main: ["bib/refs_main.bib"]
+# bibliography_supp: ["bib/refs_supp.bib"]
 ---
 
-<!-- To Do: -->
-<!-- 1) Go through methods + results -->
-<!--   a) update figs -->
-<!--   b) update references -->
-<!--   c) add in stats -->
-<!-- 2) Work through comments in discussion per KH -->
-<!-- 3) Insert citations for discussion -->
-<!-- 4) Work through introductions -->
-<!-- 5) Think through doc workflow (how to split docs & refs) -->
+##### TO DO: fix affiliations
 
-```{r setup, include=FALSE}
+```{r setupmain, include=FALSE}
 ## Libraries
@@
 library(rgdal)
+library(glue)
 
-opts_chunk$set(echo = FALSE)
+opts_chunk$set(echo = FALSE, warning = FALSE)
 out_main_fig = TRUE
@@
 # ABSTRACT
+Access to care shapes burden of diseases globally.
+For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries.
+Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners.
+In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths.
+Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP.
+We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country.
+We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden.
+We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic).
+Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.
+These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.
 
@@
 # INTRODUCTION
-[see figure 1 A](#fig1A)
+
+Inequities in access to care are a major driver of disease burden globally [@world2018world].
+Often, the populations at greatest risk of a given disease are the most underserved [@peters2008poverty].
+Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to allocate limited resources [@mainaSpatialDatabaseHealth2019].
+
+Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans .
+While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018].
+However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
+
+In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero].
+Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS].
+From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.
+Yet it is unclear how these vaccines should be distributed to prevent deaths most effectively.
+
+A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
+The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
+the more bites reported in a location, the higher the incidence of *rabies* exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018].
+If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.
+
+In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country [@rajeevHealthcareUtilizationProvisioning2018].
+PEP is not available at any other public clinics or through the private sector.
+In addition, there is limited control of rabies in dog populations and the disease appears to be endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011].
+Due to the spatially restricted nature of PEP distribution points and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
+Previously, we  estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2018].
+Here, we revise estimates of burden by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
 
 # METHODS
-## Spatial covariate data {#methods-spatial}
-To calculate travel times, we used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale.
-We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see section S1](#S1)).
-Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, @linardPopulationDistributionSettlement2012) and aggregated to the commune and district levels.
 
-## Bite patient data {#methods-bitedata}
+## Estimating travel time to closest ARMC {#methods-spatial}
+We used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale.
+We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see supplementary section S1](#S1)).
+To asses the accuracy of this method, we compared this estimate to driving time data collected by IPM during field missions and self-reported travel times from a subset of patients reporting to the Moramanga ARMC.
+We also compared unweighted average travel times and distance  Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, [cite older WP paper here!]) and aggregated to the commune and district levels.
+We compared these population estimates to recently released estimates from the 2018 census of Madagascar, and a recent census of three communes in the Moramanga District (see SZ).
+
+## Estimating average annual bite incidence {#methods-bitedata}
 We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
@@
 
-For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1.2).
-Therefore, we defined the catchment for each clinic as all districts for which the clinic was the closest ARMC for the majority of the population in that district.
-We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 1A).
-
-```{r fig 1, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M1.jpeg")
-```
-
-*[__Figure 1.__]{#fig1} __A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data.
-Color corresponds to the clinic catchment.
-Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients.
-Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
-__B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district.
-Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic.
-__C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis).
-Colors correspond to the clinic catchment,
-shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
-the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*
-
 For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted.
-We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2.1A and B).
+We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days ([Figure S2.1A and B](#figS2.1)).
 To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
-We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics (Fig S2.1C, [further details in section S2](#S2)).
-
-```{r contacts stats}
-contacts_mora <- fread("../output/sensitivity/contacts_mora.csv")
-contacts_natl <- fread("../output/sensitivity/contacts_natl.csv")
-prop_excluded_natl <- (sum(contacts_natl[estimated_cat1 == 1]$no_patients) + sum(contacts_natl[estimated_cat1 == 0 & !is.na(no_known)]$no_known))/sum(contacts_natl$no_patients)
-
-```
-
-Previous work in the Moramanga District showed that Category I exposures, i.e.
-touching or feeding a suspected animal or human rabies case, make up approximately 20% of patients reporting to ARMCs [@rajeevHealthcareUtilizationProvisioning2018].
-These low risk exposures do not contribute to the mortality burden of rabies.
-However, we could not directly exclude them as details on exposure type was missing for the majority of the patient forms.
-Detailed data from Moramanga indicates that these contacts consistently present as clustered cases and the majority report on a single day.
-Accordingly, we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure S2.2A).
-We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) excluded approximately `r round(contacts_mora[prop == "contacts" & sd == 3.00]$excluded*100, 2)`% of Category I exposures, with only `r round(contacts_mora[prop == "bites" & sd == 3.00]$excluded*100, 2)`% of Category II and III patients excluded (Figure S2.2B).
-We further excluded patients which were explicitly noted to be Categort I exposures but were not identified based on clustering of presentation dates, resulting in the exclusion of approximately `r round(prop_excluded_natl*100, 2)`% of patient records from the national data.
-After excluding patients likely to reflect Category I exposures as described, we estimated average bite incidence as the mean of the yearly bites reported/the proportion of bite patient data which were submitted.
 
@@
 These data were completely reported with details on the type of exposure documented.
-We calculated average annual bite incidence to the commune level by excluding Category I exposures and taking the monthly average of reported bites and multiplying by 12.
-Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C).
+We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.
 
-## Model of reported bites as a function of travel time to closest ARMC {#methods-bitemods}
+## Modeling reported bite incidence as a function of travel time to closest ARMC {#methods-bitemods}
 We modeled incidence of bites as a function of travel time ($T$) as follows: 
@@
 
-## Estimating burden of rabies deaths {#methods-burden}
+Finally, to check how correcting for underreporting affected our modeling results, we fitted these same models to data uncorrected for reporting (see [supplementary section S4](#S4)).
+
+## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}
+
+<!-- TO DO: simplify discussion here but bring in some figs, etc to Supplementary section to explain assumptions! -->
+
 We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2).
-We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@consortiumPotentialImpactImproved2019].
+We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$).
 We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons.
-As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs and human exposure incidence across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating].
+As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] .
 $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.
 
+```{r fig 2, fig.fullwidth=TRUE}
+include_graphics("../figs/main/M2.jpeg")
+```
 *__Figure 2.__Decision tree for burden estimation.
@@
 |:----------------|:----------------:|:----------------------------------|:---------:|
-| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@consortiumPotentialImpactImproved2019] |
+| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@whorabiesmodelingconsortiumPotentialImpactImproved2019] |
 | $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$.
-| [@consortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
+| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
 | $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
@@
 
-## Strategically expanding PEP access {#methods-incremental}
+## Estimating the impact of expanding PEP provisioning to additional clinics {#methods-incremental}
 
@@
 
-We developed a framework to rank clinics by how much they could improve access for people currently underserved (< 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
+We developed a framework to rank clinics by how much they could improve access for people currently underserved (> 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
 We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.
@@
 In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons).
-We also estimate travel times given that all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning.
+We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning.
+Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5).
 Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.
 
-We also estimate how expansion to additional clinics may impact vaccine demand.
+We estimate how expansion to additional clinics may impact vaccine demand.
 Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment.
@@
 wastage due to injection error).
-We take the mean of 100 simulations of throughput for each catchment and set of clinics.
+We take the mean of 1000 simulations of throughput for each catchment and set of clinics.
 
-## Sensitivity analyses for burden {#methods-sensitivity}
+## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
 We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden.
 Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden.
-We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplement section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
+We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplementary section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
+
+## Data availability
+All analyses were done in R and using these packages (list packages with commas and citations).
+All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo.
+The raw bite patient data at the national level is maintained by IPM.
+This database was last queried on *date here* for these analyses.
+Anonymized raw data from this database are available upon request from IPM.
+Anonymized raw data collected from the Moramanga District are shared in the archived repository.
+
+## Ethics statement
+Data collection from the Moramanga District was approved by the Princeton University IRB (# 7801) and the Madagascar Ministry of Public Health Ethics Committee (# 105-MSANP/CE).
+Oral informed consent was obtained from all interviewed participants.
+Data collected from bite patients at the national level are maintained jointly by the Ministry of Health and IPM as a routine part of PEP provisioning.
 
@@
 
-## Travel times as a predictor of bites {#results-bitemods}
-All of the candidate models which included travel times produced reasonable predictions when compared to the observed data (Figure SY).
-Population size alone was the poorest fit to the data as estimated by DIC, and all models with population size as a covariate did not generate realistic predictions to the data or when used to predict out of fit.
-For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Figure 3B, Figure SX).
-In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Figure 3A).
-As these models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence and burden.
-_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons A) the estimated relationship between travel times in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
+## Estimating travel times to and catchment populations of 
+Travel times to the closest ARMC varies greatly across the country. We compared these estimates based
+on the MAP friction surface to self-reported estimates of hours traveled to the Moramanga clinic for a subset of patients and driving times across the country from IPM.
+We found that the travel times estimated from the friction surface underestimate true driving times and patient reported travel times to the Moramanga ARMC.
+Importantly, the friction surface does not account for road condition, traffic, or seasonality in travel times.
+It also assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence.
+From the self reported travel patterns, we know that people use mutliple modes of transport and walking is common adding to the variation (SX).
+Despite these limitations, the estimated travel times correlate strongly with the reported and groundtruthed driving times.
+Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (SY).
+We use the estimated weighted mean of travel times by population as a proxy for access at the commune/district level in subsequent analyses.
+
+For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1.2).
+Therefore, we defined the catchment for each clinic as all districts for which the clinic was the closest ARMC for the majority of the population in that district.
+We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 1A).
+
+```{r fig 1, fig.fullwidth=TRUE}
+include_graphics("../figs/main/M1.jpeg")
+```
+*__Figure 1.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data.
+Color corresponds to the clinic catchment.
+Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients.
+Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
+__B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district.
+Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic.
+__C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis).
+Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
+the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*
+
+## Estimating average annual bite incidence at the district and commune level
+We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics ([Fig S2.1C](#figS2.1)). See [supplementary section S2](#S2) for more details.
+
+Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C).
+Scale discussion.
+Outlier discussion.
+
+## Modeling reported bite incidence as a function of travel times {#results-bitemods}
+All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2).
+Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
+Sensitivity to underreproting.
+For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2).
+In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3).
+These predictions were robust to how we corrected for underreporting ([supplementary section S4](#S4)).
+As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence.
+
+```{r fig 3, fig.fullwidth=TRUE}
+include_graphics("../figs/main/M3.jpeg")
+```
+_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons.
+The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
 The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept.
@@
 The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept.
-B) Log(observed) vs.
-log(predicted) mean annual reported bites for the three different models (added a small positive offset of 0.1 to account for log of zeros)._
+The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model)._
 
-## Estimation of baseline burden {#results-burden}
-Overall, we estimate between 306 - 373 average annual deaths in Madagascar due to rabies depending on the model used (Table 2).
-We also estimate that under the current system of 31 ARMCs in Madagascar, approximately 483 - 548 deaths due to rabies are prevented through provisioning of PEP each year.
+## Estimating the burden of human rabies deaths subnationally {#results-burden}
+
+```{r burden results}
+natl_burden <- read.csv("../output/preds/natl_burden.csv")
+```
+
+Overall, we estimate between `r round(sort(natl_burden$deaths_mean)[1])` - `r round(sort(natl_burden$deaths_mean)[nrow(natl_burden)])` average annual deaths in Madagascar due to rabies depending on the model used (Table 2).
+We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_burden$averted_mean)[1])` - `r round(sort(natl_burden$averted_mean)[nrow(natl_burden)])` deaths due to rabies are prevented through provisioning of PEP each year.
 In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally at the district level, but also within districts (Figure 4).
 
-_**Figure 4.** Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
-A) for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district.
+*__Table 2.__ Model predictions of mean annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% CIs in parentheses.*
+
+```{r Table 2}
+natl_burden %>%
+  mutate_at(vars(starts_with("bites")), function(x, p) x/p*1e5, p = .$pop) %>%
+  mutate_at(vars(bites_mean:averted_lower), round) %>%
+  mutate(bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
+         deaths = glue("{deaths_mean} ({deaths_lower} - {deaths_upper})"),
+         averted = glue("{averted_mean} ({averted_lower} - {averted_upper})")) %>%
+  select(scale, bites_100k, deaths, averted) -> table1
+
+kable(table1, digits = -0, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
+                                        "Deaths averted"))           
+```
+
+```{r fig 4, fig.fullwidth=TRUE}
+include_graphics("../figs/main/M4.jpeg")
+```
+*__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
+__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district.
 Points are colored by and districts are ordered by travel times.
-The dashed lines show the average national incidence of human rabies deaths for the commune (red) and district (blue) models.
-Incidence mapped to the B) commune level and C) district level from the respective models; grey x’s show locations of current CTAR._
+The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
+Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current CTAR.*
 
-table:*__Table 2.__ Model predictions of mean annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% CIs in parentheses.*
+## The impact of expanding PEP access to additional clinics {#results-incremental}
+ We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 100 {inline} clinics reducing burden by almost 50% {inline}.
+However, benefits accrue less rapidly as ARMC are added, saturating after approximately 200 {inline} additional clinics.
+As clinics are added, reported bite incidence saturates, and instead there are shifts in which clinics bites are reported to ([see supplementary section S5](#S5)).
+Even at this maximal PEP access, our model still predicts > 100 {inline} rabies deaths per annum, as average reporting of rabid exposures still remains at approximately 85 % ([Fig S5.4](#figS5.4)).
+This is not due to saturation of reporting (see distribution of reporting fig in S5), but instead to to broad patterns of remoteness across Mada.
+Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded.
+Our model predictes an increase from 40,000 {inline} vials under current clinic provisioning, to ~50,000 {inline} vials with 200 {inline} clinics, ~60,000 {inline} and approximately 90,000 {inline} vials if all clinics were to provision PEP.
 
-## Incremental reduction of burden with expanded PEP {#results-incremental}
-We use our travel time models to look at the benefits of expanding PEP incrementally.
-We find that expanding access to PEP reduces rabies deaths, but benefits accrue less rapidly  as ARMC are added, saturating after approximately 200 clinics.
-This effect is more pronounced for the commune level model than for the district model.
-Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded, with vials needed outpacing the increase in deaths averted.
-**Report actual # of vials needed as PEP is expanded.** Even with maximal PEP access, our model still predicts > 100 rabies deaths per annum, as average reporting stays at x% (see section S6 to see how things shift!)
+<!-- TO DO: State what happens to throughput + vial sensitivity! -->
 
-figure:_**Figure 5.** A) Decrease in deaths due to rabies and B) increase in vials needed per death averted as additional ARMC are added at the national level based on the two models of reported bites.
+```{r fig 5, fig.fullwidth=TRUE}
+include_graphics("../figs/main/M5.jpeg")
+```
+
+*__Figure 5.
+A)__ Decrease in deaths due to rabies, __B)__ increase in vials needed per death averted, and __C)__ increase in total # of vials as additional ARMC are added at the national level based on the two models of reported bites.
 Lines are the mean of 1000 simulations with envelopes representing 95% CIs.
-The “...” on the axes indicates a break in the axes, and the last point is the burden and vial predictions when all CSBS in the country provision PEP (max access)._
+The “...” on the axes indicates a break in the axes, and the last point is the burden and vial predictions when all CSBS in the country provision PEP (maximum access).*
 
-## Sensitivity analyses 
-There is significant uncertainty in the predicted numbers of rabies deaths, but the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (see Supplementary Materials, section S5).
-Our estimate of the absolute number of deaths is most sensitive to our assumptions of rabies exposure incidence in the population.
+## Sensitivity of burden estimates to decision tree parameters {#results-sensitivity}
+There is significant uncertainty in the predicted numbers of rabies deaths, but the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (see [supplementary section S6](#S6)).
+Our estimate of the absolute number of deaths and the predicted impact of access is most sensitive to assumptions of underlying rabies exposure incidence in the population.
+Other parameters don't contribute nearly as much uncertainty.
+In some ways is it better to present this here?
+
+<!-- TO DO: Univariate sensitivity analyses -->
 
 # DISCUSSION
+
 ## Main findings
+Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar.
+Using an adapted decision tree framework and assuming endemic rabies incidence, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
+Even given only 31 clinics, we estimate that current provisioning averts 50% {inline} of deaths that would occur due to rabies in the absence of PEP.
+We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by 50 {inline}.
+However, our model predicts that these effects saturate, and that even given maximum expansion of PEP (i.e.
+to all major health centers (n = 1676) in the country), over 100 deaths {inline} may still occur nationally.
+In addition, vial demand outpaces additional deaths averted, with more vials needed per averted death as clinics are added.
 
-## Strengths and Limitations
+## Limitations
+Our modeling framework makes strong assumptions about baseline rabies exposure incidence, assuming a range of 11 - 55 exposures per 100,000 persons in Madagascar.
+We base these estimates on a reasonable range of human:dog ratios [@knobelReevaluatingBurdenRabies2005; @ratsitorahinaDogEcologyDemography2009; @sambo2018estimating] and dog rabies incidence level [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+A previous study in the Moramanga District in Madagascar estimated a range of 42 – 110 exposures per 100,000 persons, however this was over a 16 month period with frequent rabies cases and multiple human deaths [@rajeevHealthcareUtilizationProvisioning2018].
+As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we sample uniformly over this range and over the proportion of reported bites estimated to be rabid (0.2 - 0.6 from the Moramanga data) to generate a mean estimate of burden.
+Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same.
+While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies incidence.
+Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.
 
-## Broader Context
+The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
+Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.
+<!-- TO DO: SOMETHING ABT SHORTAGES HERE AND BITS FROM CLAIRE'S MSC -->
 
-## Conclusion
+We make several simplifying assumptions regarding PEP provisioning and efficacy.
+First, we assume perfect compliance of patients and efficacy of PEP.
+Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018].
+We also assume that patients report to their closest clinic, but in our data approximately 10% of patients reported to other clinics.
+As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level.
+Current practices of provisioning PEP reactively to Category I exposures.
+Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped [@rajeevHealthcareUtilizationProvisioning2018].
 
-<!-- # REFERENCES -->
+We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
+Importantly, in the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel.
+In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that interact to determine an individual’s access to PEP.
+While we did not explicitly include variation in clinic functioning, we did include a catchment level random intercept, to account for spatial differences in reported bite incidence not due to travel times.
+This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e.
+differences in clinic functioning such as vaccine availability) and at the catchment level (i.e.
+differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
 
-<!-- ::: {#refs_main} -->
-<!-- ::: -->
+## Broader context
+```{r}
+burden_preds <- fread("../output/preds/complete/burden_filled.csv")
+## Grouped to district
+burden_preds$distcode <- mada_communes$distcode[burden_preds$names]
+burden_preds$commcode <- mada_communes$commcode[burden_preds$names]
+
+burden_preds %>%
+  filter(distcode == "MG33314", scenario == 0) %>%
+  group_by(scale) %>%
+  summarise(death_inc = sum(deaths_mean)/sum(pop)*1e5,
+            reporting = mean(reporting_mean)) -> mora_data
+
+```
+While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings (1 - 5 subnationally, on average 1.5, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009].
+In particular, the estimate of incidence of deaths for the Moramanga District fell within the range of incidence estimated from a contact tracing study in this district (1.35 {inline} vs.
+1-3 deaths {inline} per 100,000 persons and reporting 72 vs 85 {inline}, [@rajeevHealthcareUtilizationProvisioning2018]).
+At the national level, they are in line with estimates of burden based on a more standard decision tree approach [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+
+As bites due to other reasons than rabies can be more common than genuine rabies exposures [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], reported bite data may be a poor indicator of rabies transmission, particularly in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
+Surveillance instead...Integrated bite case management and clinic-based surveillance approaches have been established as feasible tools that could accomplish this [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017], and ARMC could act as sentinels for rabies surveillance across the country.
+Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen surveillance.
+
+Our work is in line with previous studies demonstrating the impacts of geographic access to care on health outcomes [@aleganaTreatmentseekingBehaviourLow2017; @nikolayEvaluatingHospitalBasedSurveillance2017; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @Hegde2019dt; @metcalfTransportNetworksInequities2015].
+While many of these studies directly estimated reporting using community based and cross-sectional survey data [@aleganaTreatmentseekingBehaviourLow2017; @nikolayEvaluatingHospitalBasedSurveillance2017], we use model derived estimates of reporting and assumptions of baseline disease incidence to estimate burden, similar to methods developed for other pathogens to estimate burden in data sparse settings [@macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @Hegde2019dt].
+For canine rabies, estimating burden of disease directly is challenging [@scottAddressingDisconnectEstimated2017; @taylorDifficultiesEstimatingHuman2017] and direct empirical estimates of rabies burden have previously relied on exhaustive contact tracing studies [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008] or large-scale mortality surveys [@sudarshanAssessingBurdenHuman2007].
+Recent studies have estimated access to health-seeking behavior and PEP completion and adherence, but not directly linked these estimates to burden [@tarantolaRabiesPostexposureProphylaxis2017; @zaidiGeographicVariationAccess2013].
+Our framework for incorporating estimates of access into previously developed burden estimation methods (i.e.
+decision trees) could help to improve and refine existing burden estimates and explore impacts of expanding access, particularly at the sub-national level.
+
+In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
+Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons {inline}, compared to Tanzania where catchment sizes are about half of that.
+Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons.
+Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine upon demand.
+However, once the more Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013].
+However, Gavi investment can remove this cost barrier.
+If this results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
+Given existing challenges in data collection and monitoring of how and to whom PEP is provisioned, strong monitoring and evaluation systems will be necessary to ensure these vaccines are being used judiciously and equitably.
+
+Ultimately even given expanded access, achieving 'the last mile', i.e.
+preventing deaths due to rabies in the most remote populations will require a disproportionate level of resources and may not be achievable through PEP alone [@delriovilasTribulationsLastMile2017].
+IBCM programs and risked based assessments are one way to tackle the issue of managing PEP resources effectively even as demand outpaces reductions in burden [@rysavaPathRabiesElimination2018].
+Combining PEP provisioning with mass dog vaccination will be key to interrrupting transmission, and eliminating deaths due to rabies.
+These same issues of access, however apply to both dog vaccination programs and surveillance systems, and integrating spatial heterogeneities in vaccination and detection will be critical to understand how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015].
+
+## Conclusion {inline these stats}
+Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths.
+Based on our modeling framework, a more equitable target may be aiming to minimize the population living within 3-5 hours travel time of a clinic, but due to broad patterns of remoteness across Madagascar, even when all primary health facilities are providing PEP, approximately 10 % {inline} of people live > 3 hours away from a clinic.
+Expanding PEP to clinics that target currently underserved populations may be the best strategy for preventing human rabies deaths.
+If expanded only to one clinic per district (N = 114), an additional 40% of deaths could be averted.
+Expansion of PEP can greatly reduce risk and inequities in access to this life-saving vaccine, but ultimately, interrupting transmission of rabies through mass dog vaccination programs will be the most equitable way to prevent, and eventually eliminate, rabies deaths.
+
+## Acknowledgements
+
+\newpage
+
+## References
+::: {#refs_main}
+:::
+
+\newpage
 
@@
 
-## S1. Estimating travel times to the nearest ARMC {#S1}
-[*jump back to main methods section*](#methods-spatial)  
+## S1. Estimating travel times to the closest ARMC {#S1}
+[*jump back to main methods section: Spatial covariate data*](#methods-spatial)  
 To estimate travel times to the nearest clinic at the scale of administrative units, we generated two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop ([worldpop.org](https://www.worldpop.org)) raster of population originally at an ~100m^2 resolution.
-We extracted the mean of travel times (Fig S1.1A) weighted by the population in that grid cell (Fig S1.1B) for each district or commune.
-Figure S1.1 shows the rasters we generated as well as the resulting extracted estimates of travel times at the district (Fig S1.1D) and commune (Fig S1.1D) level.
-Raw inputs are availble through World Pop and MAP[@abbasRabiesControlInitiative2011].
+We extracted the mean of travel times (Fig S1.1A) weighted by the population in that grid cell (Fig S1.1B) for each district (Fig S1.1C) or commune (Fig S1.1D).
+
+(euclidean distances for the driving data with GPS point locations and distance from the centroid of commune where the patient resided to the location of the ARMC for the self reported travel times)
 
@@
 ```
-_**Figure S1.1.
-A)** Travel time estimates at an approximately 1x1 km scale **B)** Population per 1x1 km grid cell.
-Mean travel times weighted by population size extracted for each **C)** District and **D)** Commune._
+*__[Figure S1.1.]{#figS1.1} A)__ Travel time estimates at an approximately 1x1 km scale __B)__ Population at same ~1x1 km scale.
+Mean travel times weighted by population size extracted for each __C)__ District and __D)__ Commune.*
 
 We assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit.
-Figure S1.2A shows that for most districts and communes a single clinic served the majority of the population by this metric.
+Most districts and communes a single clinic served the majority of the population by this metric (Fig S1.2A).
 In addition, our data show that the majority of bite patients do report to the closest clinic as assigned by our travel time metric (Fig S1.2B).
@@
 ```
-_**Figure S1.1.
-A)** Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
-**B)** The proportion of bites reported to each clinic which originated from a district within the assigned catchment._
+*__[Figure S1.2.]{#figS1.1} A)__ Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
+__B)__ The proportion of bites reported to each clinic which originated from a district within the assigned catchment.*
 
-## S2. Correcting data for underreporting and excluding Category I exposures {#S2}
+## S2. Estimating average annual bite incidence at the district and commune level
+[*jump back to main methods section: Bite patient data*](#methods-bitedata)
 
-### Estimating under-submission of forms
-To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted (Figure S2.1A).
+To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1).
 We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period.
@@
 
-We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial underreporting (Figure S2.1B).
+We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial underreporting ([Figure S2.1B](#figS2.1)).
 Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned.
 
-Figure S1.1 correcting
-```{r reporting}
+```{r fig 2.1, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S2.1.jpeg")
 ```
-Figure S1.2 vial ests
+*__[Figure S2.1.]{#figS2.1}A)__ The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
+__B)__ Reporting estimates for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the  criteria of 15 consecutive days of non-reporting/365.
+__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013-2016 for each clinic correcting for underreporting (squares) using the 15 day cut-off show in A and B, vs.
+not correcting for underreporting (circles).
+We did not have data on vials provisioned for IPM.*
 
-### Excluding Category I exposures
-We also excluded Category I exposures because they do not contribute to rabies risk and more importantly follow a different travel time distribution.
-Because they are usually actively referred to clinics (Moramanga data fig).
-They also present clustered on the same dates (Moramanga data fig).
-So we excluded by this clustering.
+## S3. Modeling reported bite incidence as a function of travel times
+<!--  TO DO = COMBINE SECTIONS S3/S4 -->
+### Candidate models of bite incidence {#S3}
 
-Figure S1.3 contacts
-```{r cat I}
-```
-Figure S1.4 ttime differences
+[*jump back to main results section: Travel times as a predictor of bites*](#results-bitemods)
 
-Because there was variation in estimates of reporting, and in our resulting incidence estimates we looked at how this impacted estimates of travel time effects (Supplementary Materials S4).
-
-## S3. Candidate models of bite incidence
 We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations).
@@
 All models and posterior estimates of parameters converged (Fig S3.1, both psrf and mpsrf approach 1).
-Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not significantly improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
+Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
 However, when used to predict data that were not used to fit the models (i.e.
-model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop, onlyPop) generated reasonable predictions to the data (Figure S3.3).\
+model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop) generated reasonable predictions to the data (Figure S3.3).\
 
+*__Table S3.1.__ DIC Estimates for all models.
+Models with the lowest DIC for each dataset and scale of covariate are in bold.
+For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
+For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
 ```{r table 1}
@@
 knitr::kable(dic_ranks, 
-             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"), 
-             caption = "Table S3.1.
-DIC Estimates for all models.
-Models with the lowest DIC for each dataset and scale of covariate are in bold.
-For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
-For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).")
+             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"))
 ```
@@
 
-```{r psrf}
+```{r fig S3.1, fig.fullwidth=TRUE}
 include_graphics("../figs/supplementary/S3.1.jpeg")
@@
 Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
+Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
 <br/>
 
-```{r s3.2, fig.fullwidth=TRUE}
-
+```{r fig S3.2, fig.fullwidth=TRUE}
 include_graphics("../figs/supplementary/S3.2.jpeg")
-
 ```
@@
 Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set  was used for fitting and the scale of the model (Moramanga = Moramanga data, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
+Colors show which data set  was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
+
+<!-- Example of how to link to full figure online -->
+<!-- [link to full fig](https://raw.githubusercontent.com/mrajeev08/ModelingChapter/master/figs/fig1.jpeg) -->
 
@@
 
-```{r s3.3, fig.fullwidth=TRUE}
+```{r fig S3.3, fig.fullwidth=TRUE}
 include_graphics("../figs/supplementary/S3.3.jpeg")
@@
 Out of fit predictions to data.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data not used to fit the model.
-A) Estimates from commune and district model fitted to the national data predicting the Moramanga data.
-B) Estimates from models fitted to the Moramanga data predicting the national data at the commune and district scale.
-Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set was used for fitting (Commune = National data with covariates at the commune level, District = National data with covariates at the district level, A) and the scale of the model at which covariates were applied (Commune = covariates at the commune scale, District = covariates at the District scale)._\
+First two columns are the predictions from the commune and district model fitted to the national data for the Moramanga data with fixed and random intercepts.
+The third column are predictions from models fitted to the Moramanga data predicting the national data at the commune and district scale (only fixed intercept models).
+Rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
+Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._
+
 <br/>
 
-## S4. Sensitivity of these models to reporting/contact cut_offs
+### Sensitivity of models to correction for underreporting {#S4}
 
-For all resulting data sets used to fit models, the commune scale model with population as an offset and a random catchment effect had the lowest DIC.
+<!-- TO DO: add one for lower cut-off (i.e. 5 days consec without reporting) -->
+[*jump back to main results section: Models*](#results-bitemods)
+
+We also looked at the sensitivity of our model estimates to how we corrected our data for underreporting. Specifically we compared results from models fitted x, y, x.
+
+We found that the model with population size as an offset (flatPop) still generated the closest predictions to the data (Table S4.1 for DIC comparisons and Fig S4.1 for predictions to the data).
+These models with a random intercept also resulted in qualitatively similar predictions of the impact of travel times on reported bite incidence, although correcting for underreporting did impact the intercept (modulating the overall incidence estimates, but not the relationship with travel times, Fig S4.2).
+
+*__Table S4.1.
+Comparing DIC ranks for models using datasets with different data correction.
+For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
+For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
 
@@
 ## DIC Tables
-## DIC Tables
-# model_se %>%
-#   group_by(scale, contacts, reporting, intercept, pop_predict) %>%
-#   summarize(dic = dic[1]) %>%
-#   arrange(reporting, contacts, dic) %>%
-#   mutate(colnames = interaction(reporting, contacts),
-#          colnames = case_when(colnames == "15.3" ~ "Corrected for underreporting & Cat I excluded", 
-#                               colnames == "Inf.3" ~ "Cat I excluded only", 
-#                               colnames == "15.Inf" ~ "Corrected for underreporting only", 
-#                               colnames == "Inf.Inf" ~ "None (raw data)")) %>%
-#   ungroup() %>%
-#   select(-contacts, - reporting) %>%
-#   pivot_wider(names_from = colnames, values_from = dic) -> dic_table
-# 
-# knitr::kable(dic_ranks, 
-#              col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"), 
-#              caption = "Table S4.1.
-Comparing DIC ranks for models using datasets with different data processing.
-For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
-For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).")
+model_se <- fread("../output/sensitivity/model_se.csv")
+model_se %>%
+  group_by(scale, contacts, reporting, intercept, pop_predict) %>%
+  summarize(dic = dic[1]) %>%
+  arrange(reporting, contacts, dic) %>%
+  mutate(colnames = interaction(reporting, contacts),
+         colnames = case_when(colnames == "15.3" ~ "Corrected for underreporting & Cat I excluded",
+                              colnames == "Inf.3" ~ "Cat I excluded only",
+                              colnames == "15.Inf" ~ "Corrected for underreporting only",
+                              colnames == "Inf.Inf" ~ "None (raw data)")) %>%
+  ungroup() %>%
+  select(-contacts, - reporting) %>%
+  mutate(dic = round(dic, 2)) %>%
+  pivot_wider(names_from = colnames, values_from = dic) -> dic_table
 
+knitr::kable(dic_table,
+             col.names = c("Scale", "Intercept type", "Pop effect", "Corrected for underreporting & Cat I excluded", "Corrected for underreporting only", "Cat I excluded only", "None (raw data)"))
 ```
 
-## S5. Incrementally adding clinics (where they were added, shifts in ttimes + catchment pops, shifts in burden/reporting etc. --answer q here: why are there still abt 100 deaths)
+```{r fig S4.1, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S4.1.jpeg")
+```
+_**Figure S4.1 Predictions to the national data with different corrections.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data corrected for both underreporting and with Category I exposures excluded, only corrected for underreporting, with only Category I exposures excluded, and the raw data with no corrections.
+Rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model) and the intercept type (random intercept by catchment or fixed intercept).
+Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._
 
-## S6. Sensitivity of burden estimates to parameter assumptions (baseline + incremental)
+```{r fig S4.2, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S4.2.jpeg")
+```
+_**Figure S4.2.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with a random catchment intercept and with population as an offset.
+Panels show predictions from models fitted to (1) data corrected for both underreporting and with Category I exposures excluded, (2) only corrected for underreporting, (3) with only Category I exposures excluded, and the raw (4) data with no corrections (raw data).
+The lines are the estimates using the mean of the posterior distributions of the parameters and the envelopes shows the 95% CI given the random intercept.
+The colors are the two candidate models with travel times at the commune and district level fitted to the national data with a random intercept by catchment.
+The points show the data used to fit the models._
 
-<!-- ## Supplementary References -->
+## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
+[*jump back to main methods section: Incremental reduction of burden*](#methods-incremental)  
+[*jump back to main results section: Incremental reduction of burden*](#results-incremental)
 
-<!-- ::: {#refs_supp} -->
-<!-- ::: -->
+```{r scenario stats}
+## candidate points
+csbs <- read.csv("../data/raw/csbs.csv", stringsAsFactors = FALSE)
+csbs %>% 
+  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
+  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
+catches_incremental <- fread("../output/stats/pop_by_catch.csv")
+```
 
-# REFERENCES
+We algorithimically added clinics by their relative reductions in travel times for people living > 3 hours away from a clinic.
+Fig S5.1 shows the location and at what step clinics were added.
+`r max(catches_incremental$scenario[catches_incremental$scenario != max(catches_incremental$scenario)])` clinics were added before there were no clinics which reduced travel times for more than 0.01% of people who also lived > 3 hours away from a clinic.
+We also estimated travel times given that all additional CSB IIs provisioned PEP (an additional `r nrow(csbs) - 31` clinics).
+Fig S5.2 shows the changes in travel times at the grid cell and administrative unit scale as clinics are added.
+As clinics were added, more than one clinic could serve a single district (the proportion of the pop fig S5.3).
+Even at the maximum level of access, 10% of people still live more than 3 hours from a clinic hours (S5.4).
+Fig S.55 shows how average travel times, bite incidence, reporting of rabid exposures, and incidence of deaths at the commune and district scale decline as clinics are added.
+Fig S56 shows the shifts in these distribution as clinics are added.
+At the clinic level, Fig S5.7 shows the shifts in the distribution of catchment populations, annual vial demand, and average daily throughput as clinics are added (S5.8).
+
+```{r fig S5.1, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.1.jpeg")
+```
+_**Figure S5.1.** Maps of where and at what step (i.e.
+commune/district where the Nth clinic was added) at the **A)** district and **B)** commune level.
+X's indicate the location of existing ARMC (N = 31), polygons in white are locations where an additional clinic was not added based on the algorithm criteria (reducing travel times for > 0.01 % of people living > 3 hours away from any clinic)._
+
+```{r fig S5.2, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.2.tiff")
+```
+_**Figure S5.2.** Maps of how travel times change as clinics are added at the **A)** ~1 x 1 km grid cell **B)** commune and **C)** district scales.
+The columns are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country).
+Grey pixels show the location of ARMC at each step.
+Commune and district values are the average grid cell travel times weighted by the population in each cell._
+
+```{r fig S5.3, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.3.jpeg")
+```
+_**Figure S5.3.** As clinics are added, shifts in distribution of the proportion of population in a given catchment at the commune and district scales (colors).
+Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
+
+```{r fig S5.4, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.4.jpeg")
+```
+_**Figure S5.4.** As clinics are added, the proportion of the population binned by travel times in hours.
+Panel labels are the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
+
+```{r fig S5.5, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.5.jpeg")
+```
+_**Figure S5.5.** As clinics are added, declines in commune and district averages of **A)** travel times (hrs), **B)** bite incidence per 100,000 persons, **C)** reporting of rabid exposures, and **D)** incidence of deaths per 100,000 persons._
+
+```{r fig S5.6, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.6.jpeg")
+```
+_**Figure S5.5.** As clinics are added, shifts in the distribution of **A)** travel times (hrs), **B)** bite incidence per 100,000 persons, **C)** reporting of rabid exposures, and **D)** incidence of deaths per 100,000 persons at the commune and district scales (colors).
+Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
+
+```{r fig S5.7, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S5.7.jpeg")
+```
+_**Figure S5.7.** As clinics are added, declines in clinic average **A)** catchment population size, **B)** daily throughput, and **C)** annual vial demand given estimates of bite incidence for the two models (commune and district, colors)._
+
+## S6. Sensitivity of burden estimates to parameter assumptions {#S6}
+
+[*jump back to main methods section: Incremental reduction of burden*](#methods-sensitivity)  
+[*or to main results section: Incremental reduction of burden*](#results-sensitivity)
+
+## SN. Assumptions of underlying rabies incidence 
+<!-- TO DO: HERE OR EARLIER BIT? -->
+
+We explore the sensitivity of our burden estimates and estimates of the impacts of PEP expansion to assumptions of rabies exposure incidence.
+Specifically, we estimate burden fixed at the minimum (11.1 exposures per 100k persons) and maximum (55.7 exposures per 100k persons)  exposure incidence within our estimated range.
+We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1).
+Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e.
+if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases.
+Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e.
+if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.
+
+```{r fig S6.1, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S6.1.jpeg")
+```
+_**Figure S6.1.**Range of constrained scaling factors generate for district and commune population size.
+The black dashed lines represent the minimum and maximum estimated rabies exposure incidence (minimum = 11.1 exposures/100k persons, maximum = 55.7 exposures/100k persons).
+We generate constrained scaling factors with human population size so that rabies exposure incidence in each location scales positively or negatively across the range of observed population sizes (panels)._
+
+Overall, our sensitivity analyses show that while there is significant uncertainty in the absolute numbers of deaths due to rabies, patterns with increasing risk of death with reduced access and the potential for expanding access to PEP to reduce these deaths appear to be robust (Figs S6.2, S6.3).
+In general, for a given incidence level the higher our assumption of the proportion of reported bites which are rabid, the lower the estimated burden.
+The lower our estimate of maximum reporting, the higher overall burden of rabies, and as this parameter decreases, the relationship between travel times and rabies deaths are dampened.
+Scaling with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths.
+Finally, raw numbers of deaths greatly depend on what we assume the rabies exposure incidence is in the population.
+
+```{r fig S6.2, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S6.2.jpeg")
+```
+*__Figure S6.2.__Predicted relationship between travel times and rabies death incidence per 100k persons for the district model vs.
+commune models (colors) with columns arranged by the scaling factor (minimum incidence of 11.1 rabies exposures/100,000 persons, the maximum negative scaling with population size, the maximum positive scaling with population size, and maximum incidence of 55.7 rabies exposures/100,000 persons, see Fig S6.1 for the scaling factors).
+Rows show burden estimates given the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]).
+Colors represent the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]); shapes show the maximum and minimum assumption of $\rho_{max}$ = [0.8 , 0.95]).*
+
+```{r fig S6.3, fig.fullwidth=TRUE}
+include_graphics("../figs/supplementary/S6.3.jpeg")
+```
+*__Figure S6.3.__ Estimates of the impact of expanding access to PEP incrementally on  burden at the national level for the district model vs.
+commune models (colors) with columns arranged by the scaling factor (minimum incidence of 11.1 rabies exposures/100,000 persons, the maximum negative scaling with population size, the maximum positive scaling with population size, and maximum incidence of 55.7 rabies exposures/100,000 persons--see Fig S6.1 for the scaling factors).
+Rows show burden estimates given the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]).
+Colors represent the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]); shapes show the maximum and minimum assumption of $\rho_{max}$ = [0.8 , 0.95].*
+
+## Supplementary references
+::: {#refs_supp}
+:::

### 2020-02-20 R/03_bitemodels/02_get_modpreds.R
poisson models with overdispersion

@@ -23,7 +23,7 @@ mora_bites <- fread("output/bites/mora_bites.csv")
 
 
 # Estimates
-model_ests <- read.csv("output/mods/estimates.csv")
+model_ests <- read.csv("output/mods/estimates_poisOD.csv")
 model_ests %>%
   select(params, Mean, pop_predict, intercept, summed, data_source,
          scale) %>%
@@ -60,7 +60,7 @@ preds_mada <-
     bite_mat <- predict.bites(ttimes = covar_df$ttimes, pop = covar_df$pop, 
                               catch = covar_df$pop, names = covar_df$distcode,
                               beta_ttimes = i$beta_ttimes, beta_0 = i$beta_0, 
-                              beta_pop = i$beta_pop, sigma_0 = i$sigma_0, 
+                              beta_pop = i$beta_pop, sigma_0 = i$sigma_0, sigma_e = i$sigma_e,
                               known_alphas = known_alphas, pop_predict = i$pop_predict, 
                               intercept = i$intercept, trans = 1e5, known_catch = TRUE, 
                               nsims = 1000)
@@ -71,7 +71,7 @@ preds_mada <-
                catch = covar_df$catch, mean_bites = mean, sd_bites = sd, 
                pop_predict = i$pop_predict, intercept = i$intercept, scale = i$scale, 
                data_source = i$data_source)
-          }
+}
 
 preds_mada %>%
   filter(data_source != "Moramanga") %>%
@@ -86,8 +86,8 @@ preds_mada %>%
   left_join(mora_bites,
             by = c("names" = "commcode")) -> preds_mora_grouped
 preds_grouped <- bind_rows(preds_mora_grouped, preds_mada_grouped)
-write.csv(preds_grouped, "output/preds/bites/fitted_grouped_all.csv", row.names = FALSE)
-write.csv(preds_mada, "output/preds/bites/fitted_ungrouped_all.csv", row.names = FALSE)
+write.csv(preds_grouped, "output/preds/bites/fitted_grouped_all_OD.csv", row.names = FALSE)
+write.csv(preds_mada, "output/preds/bites/fitted_ungrouped_all_OD.csv", row.names = FALSE)
 
 # Out of fit ----------------------------------------------------------------------------------
 # Use commune and district models to predict Moramanga data
@@ -109,7 +109,7 @@ outfit_mora <-
     bite_mat <- predict.bites(ttimes = mora_bites$ttimes_wtd/60, pop = mora_bites$pop, 
[…]

### 2020-02-20 R/figures/M1_bitemaps.R

@@ -133,7 +133,7 @@ mada_map <- ggplot() +
   geom_polygon(data = gg_district, 
                aes(x = long, y = lat, group = group, fill = ctar), 
                color = "white", alpha = 0.5) +
-  geom_bezier2(data = filter(ctar_todist_bez, count > 4, distcode != dist_ctar), 
+  geom_bezier2(data = filter(ctar_todist_bez, count > 10, distcode != dist_ctar), 
                aes(x = long, y = lat, group = fct_reorder(group, total_lines, .desc = TRUE), 
                    color = ctar, size = ifelse(index == 1, 0.05, size_lines(count))), n = 1000, 
                alpha = 0.8) +
@@ -145,14 +145,85 @@ mada_map <- ggplot() +
              shape = 21, color = "white", alpha = 1) +
   geom_point(data = filter(ctar_all, exclude == 1), aes(x = long_cent, y = lat_cent), 
              shape = 1, stroke = 1, color = "black") +
+  geom_bezier2(data = filter(leg_pts, group != length(long_from) + 1), 
+               aes(x = long, y = lat, group = group, 
+                   size = ifelse(index == 1, 0.05, size_lines(sizes))), n = 1000, color = "grey50",
+               alpha = 0.8) +
+  geom_point(data = filter(leg_pts, type != "mid"), 
+             aes(x = long, y = lat, size = size_pts(sizes)), fill = "grey50",
+             shape = 21) + 
+  geom_text(data = filter(leg_pts, type != "mid"), aes(x = long, y = lat, label = sizes),
+            hjust = 0, vjust = 0, angle = 45, nudge_y = 0.2) +
+  geom_text(data = filter(leg_pts, long == min(long) | long == max(long)), 
+            aes(x = long, y = lat, label = c("District", "ARMC")), hjust = c(0, 1),
+            nudge_x = c(0.5, -0.5)) +
   scale_color_manual(values = catch_fills, guide = "none") +
   scale_fill_manual(values = catch_cols, guide = "none") +
-  scale_size_identity("Reported bites", 
-                      guide = "none") +
+  scale_size_identity("Reported bites", guide = "none") +
   labs(tag = "A") +
   theme_minimal_hgrid() +
   theme_map() +
-  coord_quickmap()
+  coord_quickmap(clip = "off")
+
+
+get.bezleg <- function(bbox, n_pts, size_vec, min_size, offset_long, offset_lat) {
+  
[…]

### 2020-02-20 R/figures/SM3_bitemods.R

@@ -17,7 +17,7 @@ library(cowplot)
 select <- dplyr::select
 
 # Estimates
-model_ests <- read.csv("output/mods/estimates.csv")
+model_ests <- read.csv("output/mods/estimates_poisOD.csv")
 model_ests %>%
   select(params, Mean, pop_predict, intercept, data_source, scale) %>%
   spread(key = params, value = Mean, fill = 0) -> model_means
@@ -44,7 +44,6 @@ psrfs <- ggplot(convergence, aes(x = type, y = val, color = interaction(data_sou
   geom_boxplot() +
   scale_color_manual(values = model_cols, name = "Scale", 
                      labels = scale_labs) +
-  ylim(c(0.98, 1.02)) +
   facet_grid(pop_predict ~ intercept, scales = "free_x", drop = TRUE) +
   geom_hline(yintercept = 1, linetype = 2, color = "grey") +
   scale_x_discrete(labels= c("psrf_upper" = "Indiviudal covariate", "mpsrf" = "Multivariate")) +
@@ -56,7 +55,7 @@ psrfs <- ggplot(convergence, aes(x = type, y = val, color = interaction(data_sou
 ggsave("figs/supplementary/psrfs.jpeg", psrfs, device = "jpeg", height = 10, width = 8)
 
 # Fitted predictions
-preds_grouped <- read.csv("output/preds/bites/fitted_grouped_all.csv")
+preds_grouped <- read.csv("output/preds/bites/fitted_grouped_all_OD.csv")
 preds_grouped$mod_intercept <- preds_grouped$intercept
 fitted_preds <- ggplot(data = filter(preds_grouped), 
        aes(x = log(avg_bites + 0.1), y = log(mean_bites + 0.1), 
@@ -75,8 +74,8 @@ fitted_preds <- ggplot(data = filter(preds_grouped),
 ggsave("figs/supplementary/fitted_preds.jpeg", fitted_preds, device = "jpeg", height = 10, width = 8)
 
 # Out of fit predictions
-outfit_mora <- read.csv("output/preds/bites/outfit_mora.csv")
-outfit_mada <- read.csv("output/preds/bites/outfit_grouped_mada.csv")
+outfit_mora <- read.csv("output/preds/bites/outfit_mora_OD.csv")
+outfit_mada <- read.csv("output/preds/bites/outfit_grouped_mada_OD.csv")
 
 # Trying nested facet labels
[…]

### 2020-02-20 R/functions/candidate_mods/fixed_intercept.R

@@ -6,14 +6,21 @@ if (summed == TRUE) {
     # Priors
     beta_0 ~ dnorm(0, 10^-6)
     beta_ttimes ~ dnorm(0, 10^-6)
-
+    sigma_e ~ dunif(0, 100) # SD overdispersion parameter
+    tau_e <- pow(sigma_e, -2)
+    
+    for(j in 1:ncovars){
+        epsilon[j] ~ dnorm(0, tau_e)
+    }
+    
     # Likelihood
     for (i in 1:ncovars) {
-      exp_bites[i] <- exp(beta_0 + beta_ttimes*ttimes[i])*pop[i] # Expectation
+      nbites[i] <- exp_bites[i]*pop[i]
+      log(exp_bites[i]) <- beta_0 + beta_ttimes*ttimes[i] + epsilon[i]
     }
     
     for(k in 1:nlocs) {
-      sum_bites[k] <- sum(exp_bites[start[k]:end[k]])
+      sum_bites[k] <- sum(nbites[start[k]:end[k]])
       bites[k] ~ dpois(sum_bites[k]) # The actual (random) responses
     }
   }"
@@ -30,11 +37,18 @@ if (summed == FALSE) {
     # Priors
     beta_0 ~ dnorm(0, 10^-6)
     beta_ttimes ~ dnorm(0, 10^-6)
+    sigma_e ~ dunif(0, 100) # SD overdispersion parameter
+    tau_e <- pow(sigma_e, -2)
+    
+    for(j in 1:nlocs){
+        epsilon[j] ~ dnorm(0, tau_e)
+    }
 
     # Likelihood
     for (i in 1:nlocs) {
-        exp_bites[i] <- exp(beta_0 + beta_ttimes*ttimes[i])*pop[i] # Expectation
-        bites[i] ~ dpois(exp_bites[i]) # The actual (random) responses
+      bites[i] ~ dpois(nbites[i])
+      nbites[i] <- exp_bites[i]*pop[i]
+      log(exp_bites[i]) <- beta_0 + beta_ttimes*ttimes[i] + epsilon[i]
     }
   }"
   
@@ -45,10 +59,11 @@ if (summed == FALSE) {
 }
 
 ## params
-params <- c("beta_0", "beta_ttimes")
+params <- c("beta_0", "beta_ttimes", "sigma_e")
 ## inits
 inits <- list(beta_0 = rnorm(1, 0, 1e-6), 
[…]

### 2020-02-20 R/functions/candidate_mods/random_intercept.R

@@ -8,18 +8,25 @@ if (summed == TRUE) {
     tau_0 <- pow(sigma_0, -2)
     beta_0 ~ dnorm(0, 10^-6)
     beta_ttimes ~ dnorm(0, 10^-6) 
+    sigma_e ~ dunif(0, 100) # SD overdispersion parameter
+    tau_e <- pow(sigma_e, -2)
 
     for (i in 1:ncatches) {
         alpha[i] ~ dnorm(beta_0, tau_0) # Random intercepts
     }
- 
+    
+    for(j in 1:ncovars){
+        epsilon[j] ~ dnorm(0, tau_e)
+    }
+    
     # Likelihood
     for (i in 1:ncovars) {
-      exp_bites[i] <- exp(alpha[catch[i]] + beta_ttimes*ttimes[i])*pop[i]
+      nbites[i] <- exp_bites[i]*pop[i]
+      log(exp_bites[i]) <- alpha[catch[i]] + beta_ttimes*ttimes[i] + epsilon[i]
     }
     
     for(k in 1:nlocs) {
-      sum_bites[k] <- sum(exp_bites[start[k]:end[k]])
+      sum_bites[k] <- sum(nbites[start[k]:end[k]])
       bites[k] ~ dpois(sum_bites[k]) # The actual (random) responses
     }
 
@@ -40,15 +47,22 @@ if(summed == FALSE) {
     sigma_0 ~ dunif(0, 100) # SD hyperparameter for random intercepts
     tau_0 <- 1/(sigma_0*sigma_0)
     beta_ttimes ~ dnorm(0, 10^-6) 
-    
+    sigma_e ~ dunif(0, 100) # SD overdispersion parameter
+    tau_e <- pow(sigma_e, -2)
+
     for (i in 1:ncatches) {
         alpha[i] ~ dnorm(beta_0, tau_0) # Random intercepts
+    }
+    
+    for(j in 1:nlocs){
+        epsilon[j] ~ dnorm(0, tau_e)
     }
 
     # Likelihood
     for (i in 1:nlocs) {
-        exp_bites[i] <- exp(alpha[catch[i]] + beta_ttimes*ttimes[i])*pop[i] 
-        bites[i] ~ dpois(exp_bites[i]) # The actual (random) responses
+        bites[i] ~ dpois(nbites[i]) # The actual (random) responses
+        nbites[i] <- exp_bites[i]*pop[i]
[…]

### 2020-02-20 R/functions/estimate_pars.R

@@ -7,7 +7,8 @@
 estimate.pars <- function(bites, ttimes, pop, ncovars, nlocs, catch, ncatches, start, end,
                           pop_predict = "addPop", intercept = "random",
                           summed = TRUE, data_source = "national", scale = "Commune",
-                          trans = 1e5, chains = 3, adapt = 500, iter = 10000, thinning = 5, 
+                          trans = 1e5, chains = 3, adapt = 500, burn = 100, 
+                          iter = 10000, thinning = 5, 
                           dic = TRUE, save = FALSE) {
   
   if(summed == TRUE) sumit = "summed" else sumit = "nosum"
@@ -19,7 +20,7 @@ estimate.pars <- function(bites, ttimes, pop, ncovars, nlocs, catch, ncatches, s
   
   jags_mod <- jags.model(textConnection(model), data = data, inits = inits, 
                          n.chains = chains, n.adapt = adapt)
-  
+  update(jags_mod, burn)
   samps <- coda.samples(jags_mod, params, n.iter = iter, thin = thinning)
 
   if(dic == TRUE) {

### 2020-03-13 docs/manuscript.Rmd
poisson models with overdispersion
updating sections on travel times & bite incidence
- also cleaning up bezier & plotting scripts
- simplified shapefiles for easier plotting and lower file sizes (yay!)

@@
 library(glue)
+library(here)
 
@@
 ## data
-ctar_metadata <- fread("../data/raw/ctar_metadata.csv")
-district_bites <- fread("../output/bites/district_bites.csv")
-mora_bites <- fread("../output/bites/mora_bites.csv")
-mada_communes <- readOGR("../data/processed/shapefiles/mada_communes.shp")
-mada_districts <- readOGR("../data/processed/shapefiles/mada_districts.shp")
+ctar_metadata <- fread(here("data/raw/ctar_metadata.csv"))
+district_bites <- fread(here("output/bites/district_bites.csv"))
+mora_bites <- fread(here("output/bites/mora_bites.csv"))
+mada_communes <- readOGR(here("data/processed/shapefiles/mada_communes.shp"))
+mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))
 
@@
 These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.
+
+# AUTHOR SUMMARY
 
@@
 ## Estimating travel time to closest ARMC {#methods-spatial}
-We used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale.
-We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see supplementary section S1](#S1)).
-To asses the accuracy of this method, we compared this estimate to driving time data collected by IPM during field missions and self-reported travel times from a subset of patients reporting to the Moramanga ARMC.
-We also compared unweighted average travel times and distance  Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, [cite older WP paper here!]) and aggregated to the commune and district levels.
-We compared these population estimates to recently released estimates from the 2018 census of Madagascar, and a recent census of three communes in the Moramanga District (see SZ).
+To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).
+
+Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale.
+We then compared three different metrics of access:
+
+1) Travel time
+2) Weighted travel time
+3) Distance
+
+extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell.
+We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC.
+We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.
+
+We compared these travel time and distance estimates to driving time data collected by IPM during field missions and patient reported travel times from a subset of patients reporting to the Moramanga ARMC (see Fig S1.2 for raw data).
 
 ## Estimating average annual bite incidence {#methods-bitedata}
-We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
-These were individual patient data forms that were submitted from clinics as frequently as monthly to annually.
-These data include details of the administrative district of the bite patient’s address and the date of reporting.
-
-For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted.
-We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days ([Figure S2.1A and B](#figS2.1)).
-To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.
-
-The 31 months of data available from the Moramanga district resolved to the commune level (the administrative level below the district, Figure 1B) provide a useful comparison to the national scale data.
-These data were completely reported with details on the type of exposure documented.
+We used two datasets on bite patients reporting to ARMC:
+1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
+These were individual patient data forms that were submitted from clinics as frequently as monthly to annually that included data on the date of reporting and were resolved to the district level (i.e.
+the district where the patient resided)
+2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level (the administrative level below the district).
 We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.
 
-## Modeling reported bite incidence as a function of travel time to closest ARMC {#methods-bitemods}
-We modeled incidence of bites as a function of travel time ($T$) as follows: 
-$$ \mu = exp(\beta_{t}T \ + \ \beta_0)P$$
-where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC, and the human population size, $P$ (an offset which scales the predicted number of bites), for a given source location (district or commune).
-We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian framework via MCMC using the R package ‘rjags’.
+For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
+To correct for this, we estimated clinic level undersubmission as the proportion of days on which forms were submitted in a given year, excluding any periods for which there were no forms submitted for 15 consecutive days.
+We then divided the total number of bites reported to each district at each clinic in a given year by the undersubmission rate for that clinic in that year.
+Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates.
+The data from the Moramanga district were completely reported across the 31 month period.
+
+At both the commune and district scale, we assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit.
+For the national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.
+
+## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
+We modeled the number of reported bites as a function of travel time ($T$) as follows: 
+$$ \mu = e^{(\beta_{t}T \ + \ \beta_0 \ )}P$$
+where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T$) and the human population size ($P$) (an offset which scales the predicted number of bites) for a given source location (district or commune).
+We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’.
 In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:
 
-$$\mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j\ + \ \beta_0)P_j$$
+$$\mu = \sum \limits_{j=1}^je^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$
 
-As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel time.
-For the models with population size, we used a model framework with either population size alone ($\mu = exp(\beta_pP + \beta_0)$) or population size and travel times ($\mu = exp(\beta_{t}T + \beta_pP + \beta_0)$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.
+As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel times.
+For the models with population size, we used a model framework with either population size alone ($\mu = e^{(\beta_pP + \beta_0 )}$) or population size and travel times ($\mu = e^{(\beta_{t}T + \beta_pP + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.
 
@@
 
-Finally, to check how correcting for underreporting affected our modeling results, we fitted these same models to data uncorrected for reporting (see [supplementary section S4](#S4)).
+Since we're assuming for poisson errors: we tested for overdispersion.
+
+Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to uncorrected data.
 
@@
 ```{r fig 2, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M2.jpeg")
+include_graphics(here("figs/main/M2.jpeg"))
 ```
@@
 ## candidate points
-csbs <- read.csv("../data/raw/csbs.csv", stringsAsFactors = FALSE)
+csbs <- read.csv(here("data/raw/csbs.csv", stringsAsFactors = FALSE))
 csbs %>% 
@@
 
-## Estimating travel times to and catchment populations of 
-Travel times to the closest ARMC varies greatly across the country. We compared these estimates based
-on the MAP friction surface to self-reported estimates of hours traveled to the Moramanga clinic for a subset of patients and driving times across the country from IPM.
-We found that the travel times estimated from the friction surface underestimate true driving times and patient reported travel times to the Moramanga ARMC.
-Importantly, the friction surface does not account for road condition, traffic, or seasonality in travel times.
-It also assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence.
-From the self reported travel patterns, we know that people use mutliple modes of transport and walking is common adding to the variation (SX).
-Despite these limitations, the estimated travel times correlate strongly with the reported and groundtruthed driving times.
-Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (SY).
-We use the estimated weighted mean of travel times by population as a proxy for access at the commune/district level in subsequent analyses.
+## Estimating travel times to the closest ARMC
+```{r ttime stats}
+ttime_stats <- read.csv(here("output/stats/prop_pop_ttimes.csv"))
+# get prop of pop > 3 hrs away at baseline
+ttime_stats %>%
+  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
+  sum(prop_pop) -> prop_over3_base
+ttime_stats %>%
+  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
+  sum(prop_pop) -> prop_over6_base
+```
 
-For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1.2).
-Therefore, we defined the catchment for each clinic as all districts for which the clinic was the closest ARMC for the majority of the population in that district.
-We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 1A).
+Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population living greater than 3 hrs away from a clinic (Fig 3B).
+We found that travel times estimated from the friction surface generally underestimate true driving times and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2).
+The friction surface assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence (i.e.
+if there is a road, then people will travel at vehicle rather than walking speeds).
+However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3).
+Despite these discrepancies, travel times correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4).
+Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1).
+Therefore we use this metric as a proxy for access at the commune/district level in subsequent analyses.
 
-```{r fig 1, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M1.jpeg")
+```{r fig M2, fig.fullwidth=TRUE}
+include_graphics(here("figs/main/M3_ttimes.jpeg"))
 ```
-*__Figure 1.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data.
-Color corresponds to the clinic catchment.
-Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients.
-Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
-__B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district.
-Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic.
-__C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis).
-Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
-the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*
+*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar.
+__B)__ Distribution of the population across travel times.
+__C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level.
+For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune.
+The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown.
+The dashed black line is the 1:1 line.*
 
 ## Estimating average annual bite incidence at the district and commune level
-We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics ([Fig S2.1C](#figS2.1)). See [supplementary section S2](#S2) for more details.
+We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population.
+While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B).
+In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A) and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).
 
-Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C).
-Scale discussion.
-Outlier discussion.
+```{r fig M3 catchmaps}
+include_graphics(here("figs/main/M3_bitemaps.jpeg"))
+```
+*__Figure 3.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each ARMC for which we have data.
+Color corresponds to the clinic catchment.
+Circles with white outline are the total number of bites reported for that district.
+Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded.
+Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
+Out-of-catchment reporting is indicated where points and line colors are mismatched.
+__B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the area plotted.
+The black outline shows the district of Moramanga, and the red polygons shows all the communes included in the Moramanga ARMC catchment.
+Circle with black outline is the location of the Moramanga ARMC with the size of points proportional to the total number of bites reported to the clinic, as per A.*
+
+Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1).
+However, we found that corrected estimates of reproted bites generated predictions of vial demand close to what was provisioned over the same time period ([Figure S2.2A and B](#figS2.2)).
+We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 3A).
+After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).
+
+```{r fig M4 biteinc}
+include_graphics(here("figs/main/M4_biteinc.jpeg"))
+```
+
+*__Figure 4.__ The estimated average annual bite incidence from the national and Moramanga data plotted at the __A)__ district scale (both datasets) and at the __B)__ commune scale (Moramanga dataset).
+Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
+the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).
+The point lines indicate the range of estimated bite incidence for each district.*
+
+```{r}
+outliers <- read.csv(here("output/stats/outliers.csv"))
+head(outliers)
+```
+
+Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship.
+There were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
+While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher incidence at low travel times at the commune level compared to the district.
 
 ## Modeling reported bite incidence as a function of travel times {#results-bitemods}
+
 All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2).
 Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
-Sensitivity to underreproting.
+Sensitivity to underreporting.
 For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2).
 In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3).
-These predictions were robust to how we corrected for underreporting ([supplementary section S4](#S4)).
+These predictions were robust to how we corrected for incomplete submission of data by clinics ([supplementary section S4](#S4)).
 As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence.
+(OD COMP--use the Moramanga model with OD & the District model with OD)
 
 ```{r fig 3, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M3.jpeg")
+include_graphics(here("figs/main/M3.jpeg"))
 ```
@@
 ```{r burden results}
-natl_burden <- read.csv("../output/preds/natl_burden.csv")
+natl_burden <- read.csv(here("output/preds/natl_burden.csv"))
 ```
@@
 ```{r fig 4, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M4.jpeg")
+include_graphics(here("figs/main/M4.jpeg"))
 ```
@@
 ## The impact of expanding PEP access to additional clinics {#results-incremental}
+
  We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 100 {inline} clinics reducing burden by almost 50% {inline}.
@@
 ```{r fig 5, fig.fullwidth=TRUE}
-include_graphics("../figs/main/M5.jpeg")
+include_graphics(here("figs/main/M5.jpeg"))
 ```
@@
 ## Limitations
+We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
+In the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel.
+In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.
+
+Importantly,the travel time estimates from the MAP friction surface underestimate true travel times reported by patients.
+This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018].
+In Madagascar, the presence of paved roads does not necessarily reflect their quality and the modes of transport used to travel these roads.
+Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune.
+Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access.
+
+<!-- TO DO: UPDATE TO REFLECT CHANGES IN OD MODS ETC. -->
+
+We did include a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times.
+This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e.
+differences in clinic functioning such as vaccine availability) and at the catchment level (i.e.
+differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
+
+Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior.
+We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning.
+
+<!-- TO DO: CLAIRE'S MSC ON OTHER REASONS + CLINIC FUNCTIONING -->
+<!-- TO DO: OUTLIERS...SCALE? -->
+
+<!-- TO DO: UPDATE BASED ON CHANGES TO BURDEN MODEL ETC. -->
 Our modeling framework makes strong assumptions about baseline rabies exposure incidence, assuming a range of 11 - 55 exposures per 100,000 persons in Madagascar.
@@
 Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.
-<!-- TO DO: SOMETHING ABT SHORTAGES HERE AND BITS FROM CLAIRE'S MSC -->
 
@@
 As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level.
+
+<!-- TO DO: CAT 1 -->
+
 Current practices of provisioning PEP reactively to Category I exposures.
@@
 
-We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
-Importantly, in the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel.
-In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that interact to determine an individual’s access to PEP.
-While we did not explicitly include variation in clinic functioning, we did include a catchment level random intercept, to account for spatial differences in reported bite incidence not due to travel times.
-This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e.
-differences in clinic functioning such as vaccine availability) and at the catchment level (i.e.
-differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
+<!-- TO DO: CASE STUDIES -->
+
+## Box 1: case studies of exceptions to the travel time effects
+
+### Anosibe A'nala
+In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours.
+Despite this, we see a high incidence of reported exposures.
+We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were close to our touched a person that died from suspected cases.
+An xx% of them were suspected cases.
+And no cases were confirmed.
+The case of AAA points to three key issues: 1) High burden of rabies in this district, we're probably seeing the tip of the iceberg.
+2) people when aware of the risk do travel (in many cases walking 3 days to attend the clinic) and when surveillance efforts target cat I around deaths too.
+Potential for awareness and outreach to go far.
+3) This is also where our travel time estimates fail us.
+Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks.
+This is also highly seasonal, during the rainy season it's almost impossible to travel other than by foot.
+And most people use multiple modes of transport depending on cost and availability (i.e.
+boat, on foot, by camion).
+
+### Antananarivo (too identifiable?)
+In 2016, A middle aged taxi driver died of rabies in suburban Antananarivo.
+Bitten by an unknown dog.
+Promptly reported to a clinic where he was reffered to IPM.
+His daughter, a prospective medical student, and other family members urged him to seek the vaccine but he did not in the end seek care.
+He developed symptoms 2 weeks later and died within a few days.
+The case was confirmed by the NRL.
+This case points to the other side of the health-seeking behavior coin from AAA.
+This man had the resources, was aware of the risk, and was reffered by the hospital (and sought care immediately).
+But ultimately, it's difficult to predict human behavior.
 
@@
 ```{r}
-burden_preds <- fread("../output/preds/complete/burden_filled.csv")
+burden_preds <- fread(here("output/preds/complete/burden_filled.csv"))
 ## Grouped to district
@@
 
+<!-- TO DO: WHAT'S THE PT I'M TRYING TO MAKE HERE? -->
 As bites due to other reasons than rabies can be more common than genuine rabies exposures [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], reported bite data may be a poor indicator of rabies transmission, particularly in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
@@
 
+<!-- TO DO -->
+citing giorbhail paper!
+What are the key pts!
+
 ## Conclusion {inline these stats}
@@
 [*jump back to main methods section: Spatial covariate data*](#methods-spatial)
-To estimate travel times to the nearest clinic at the scale of administrative units, we generated two raster inputs: 1) the travel time estimates generated using the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale and the locations of existing ARMC and 2) the population estimates which we resampled from the World Pop ([worldpop.org](https://www.worldpop.org)) raster of population originally at an ~100m^2 resolution.
-We extracted the mean of travel times (Fig S1.1A) weighted by the population in that grid cell (Fig S1.1B) for each district (Fig S1.1C) or commune (Fig S1.1D).
-
-(euclidean distances for the driving data with GPS point locations and distance from the centroid of commune where the patient resided to the location of the ARMC for the self reported travel times)
 
 ```{r fig S1.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S1.1.jpeg")
+include_graphics(here("figs/supplementary/S1.1_inputs.jpeg"))
 ```
-*__[Figure S1.1.]{#figS1.1} A)__ Travel time estimates at an approximately 1x1 km scale __B)__ Population at same ~1x1 km scale.
-Mean travel times weighted by population size extracted for each __C)__ District and __D)__ Commune.*
-
-We assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit.
-Most districts and communes a single clinic served the majority of the population by this metric (Fig S1.2A).
-In addition, our data show that the majority of bite patients do report to the closest clinic as assigned by our travel time metric (Fig S1.2B).
+*__[Figure S1.1.]{#figS1.1}__ Raster inputs to estimate travel times to the closest ARMC for Madagascar __A)__ Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale __B)__ Population estimates resampled to the same friction surface.*
 
 ```{r fig S1.2, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S1.2.jpeg")
+include_graphics(here("figs/supplementary/S1.2_raw.jpeg"))
 ```
-*__[Figure S1.2.]{#figS1.1} A)__ Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
-__B)__ The proportion of bites reported to each clinic which originated from a district within the assigned catchment.*
+*__[Figure S1.2.]{#figS1.2}__ Source of true travel time estimates __A)__ Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
+__B)__ Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
+Paths are bezier curves from origin to destination, but do not show actual paths driven.*
+
+```{r fig S1.3, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S1.3_modes.jpeg"))
+```
+*__[Figure S1.3.]{#figS1.3}__ Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
+
+```{r fig S1.4, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S1.4_lms.jpeg"))
+```
+*__[Figure S1.4.]{#figS1.4}__ Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) predicted by __A)__ Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) __B)__ Travel time estimates and C) Travel time estimates weighted by population (for commune means only).*
+
+*__[Table S1.1.]{#tabS1.1} R^2 values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
+```{r table S1.1}
+S1.1_table <- read.csv(here("output/stats/access_met_r2.csv"), row.names = 1)
+kable(S1.1_table, col.names = c("Predictor", "Driving times", 
+                                "Commune mean of patient reported travel times"))
+```
+
+```{r fig S1.5, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S1.5_outputs.jpeg"))
+```
+*__[Figure S1.5.]{#figS1.5}__ Estimates of mean travel times weighted by population at the __A)__ District __B)__ Commune scale.*
 
@@
 [*jump back to main methods section: Bite patient data*](#methods-bitedata)  
+CATCHMENT
+Most districts and communes a single clinic served the majority of the population by this metric (Fig S2.1B).
+In addition, our data show that the majority of bite patients do report to the closest clinic as assigned by our travel time metric (Fig S2.1C).
+
+```{r fig S2.1, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S2.1_catchments.jpeg"))
+```
+*__[Figure S2.1]{#figS2.1} __A)__ Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or which a district (polygon outline).
+Admin units where the fill and border colors do not match show places where assigned catchments differ at the district vs commune scale.
+__B)__ Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
+__C)__ The proportion of bites reported to each clinic which originated from a district within the assigned catchment.*
 
@@
 
-We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial underreporting ([Figure S2.1B](#figS2.1)).
+We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)).
 Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned.
 
-```{r fig 2.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S2.1.jpeg")
+```{r fig 2.2, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S2.2_vialcomps.jpeg"))
 ```
-*__[Figure S2.1.]{#figS2.1}A)__ The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
-__B)__ Reporting estimates for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the  criteria of 15 consecutive days of non-reporting/365.
-__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013-2016 for each clinic correcting for underreporting (squares) using the 15 day cut-off show in A and B, vs.
-not correcting for underreporting (circles).
+*__[Figure S2.2.]{#figS2.2}A)__ The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
+__B)__ Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the  criteria of 15 consecutive days of non-submission/365.
+__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013-2016 for each clinic correcting for undersubmission (squares) using the 15 day cut-off show in A and B, vs.
+not correcting for undersubmission (circles).
 We did not have data on vials provisioned for IPM.*
@@
 We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).
-All models and posterior estimates of parameters converged (Fig S3.1, both psrf and mpsrf approach 1).
-Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
+All models and posterior estimates of parameters converged ({inline stat for max psrf & mpsrf}) Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
 However, when used to predict data that were not used to fit the models (i.e.
@@
 ```{r table 1}
-model_ests <- read.csv("../output/mods/estimates.csv")
+model_ests <- read.csv(here("output/mods/estimates.csv"))
 model_ests %>%
@@
 
-```{r fig S3.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S3.1.jpeg")
-```
-_**Figure S3.1.
-Potential scale reduction factors (psrf) for all models.** The x-axis is the type of psrf, either multivariate which assesses the convergence of the model, or the individual which assesses the convergence for each parameter estimated from the model.
-Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
-<br/>
-
 ```{r fig S3.2, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S3.2.jpeg")
+include_graphics(here("figs/supplementary/S3.2.jpeg"))
 ```
@@
 ```{r fig S3.3, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S3.3.jpeg")
+include_graphics(here("figs/supplementary/S3.3.jpeg"))
 ```
@@
 
-### Sensitivity of models to correction for underreporting {#S4}
+### Estimating overdispersion of poisson errors
 
-<!-- TO DO: add one for lower cut-off (i.e. 5 days consec without reporting) -->
+### Sensitivity of models to correction for incomplete submission of patient forms 
+<!-- TO DO: only need to show this for main models (i.e. ttimes models) -->
+<!-- TO DO: add one for lower cut-off (i.e. 7 days consec without reporting) -->
 [*jump back to main results section: Models*](#results-bitemods)
 
-We also looked at the sensitivity of our model estimates to how we corrected our data for underreporting. Specifically we compared results from models fitted x, y, x.
+We also looked at the sensitivity of our model estimates to how we corrected our data for undersubmission. Specifically we compared results from models fitted x, y, x.
 
 We found that the model with population size as an offset (flatPop) still generated the closest predictions to the data (Table S4.1 for DIC comparisons and Fig S4.1 for predictions to the data).
-These models with a random intercept also resulted in qualitatively similar predictions of the impact of travel times on reported bite incidence, although correcting for underreporting did impact the intercept (modulating the overall incidence estimates, but not the relationship with travel times, Fig S4.2).
-
-*__Table S4.1.
-Comparing DIC ranks for models using datasets with different data correction.
-For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
-For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
-
-```{r Table S4.1}
-## DIC Tables
-model_se <- fread("../output/sensitivity/model_se.csv")
-model_se %>%
-  group_by(scale, contacts, reporting, intercept, pop_predict) %>%
-  summarize(dic = dic[1]) %>%
-  arrange(reporting, contacts, dic) %>%
-  mutate(colnames = interaction(reporting, contacts),
-         colnames = case_when(colnames == "15.3" ~ "Corrected for underreporting & Cat I excluded",
-                              colnames == "Inf.3" ~ "Cat I excluded only",
-                              colnames == "15.Inf" ~ "Corrected for underreporting only",
-                              colnames == "Inf.Inf" ~ "None (raw data)")) %>%
-  ungroup() %>%
-  select(-contacts, - reporting) %>%
-  mutate(dic = round(dic, 2)) %>%
-  pivot_wider(names_from = colnames, values_from = dic) -> dic_table
-
-knitr::kable(dic_table,
-             col.names = c("Scale", "Intercept type", "Pop effect", "Corrected for underreporting & Cat I excluded", "Corrected for underreporting only", "Cat I excluded only", "None (raw data)"))
-```
-
-```{r fig S4.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S4.1.jpeg")
-```
-_**Figure S4.1 Predictions to the national data with different corrections.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data corrected for both underreporting and with Category I exposures excluded, only corrected for underreporting, with only Category I exposures excluded, and the raw data with no corrections.
-Rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model) and the intercept type (random intercept by catchment or fixed intercept).
-Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._
+These models with a random intercept also resulted in qualitatively similar predictions of the impact of travel times on reported bite incidence, although correcting for undersubmission did impact the intercept (modulating the overall incidence estimates, but not the relationship with travel times, Fig S4.2).
 
 ```{r fig S4.2, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S4.2.jpeg")
+include_graphics(here("figs/supplementary/S4.2.jpeg"))
 ```
 _**Figure S4.2.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with a random catchment intercept and with population as an offset.
-Panels show predictions from models fitted to (1) data corrected for both underreporting and with Category I exposures excluded, (2) only corrected for underreporting, (3) with only Category I exposures excluded, and the raw (4) data with no corrections (raw data).
+Panels show predictions from models fitted to (1) data corrected for both undersubmission and with Category I exposures excluded, (2) only corrected for undersubmission, (3) with only Category I exposures excluded, and the raw (4) data with no corrections (raw data).
 The lines are the estimates using the mean of the posterior distributions of the parameters and the envelopes shows the 95% CI given the random intercept.
@@
   dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
-catches_incremental <- fread("../output/stats/pop_by_catch.csv")
+catches_incremental <- fread(here("output/stats/pop_by_catch.csv"))
 ```
 
-We algorithimically added clinics by their relative reductions in travel times for people living > 3 hours away from a clinic.
+We iterativel added clinics by their relative reductions in travel times for people living > 3 hours away from a clinic.
 Fig S5.1 shows the location and at what step clinics were added.
@@
 
+<!-- TO DO: add catchmaps -->
+
 ```{r fig S5.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.1.jpeg")
+include_graphics(here("figs/supplementary/S5.1.jpeg"))
 ```
@@
 ```{r fig S5.2, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.2.tiff")
+include_graphics(here("figs/supplementary/S5.2.tiff"))
 ```
@@
 ```{r fig S5.3, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.3.jpeg")
+include_graphics(here("figs/supplementary/S5.3.jpeg"))
 ```
@@
 ```{r fig S5.4, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.4.jpeg")
+include_graphics(here("figs/supplementary/S5.4.jpeg"))
 ```
@@
 ```{r fig S5.5, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.5.jpeg")
+include_graphics(here("figs/supplementary/S5.5.jpeg"))
 ```
@@
 ```{r fig S5.6, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.6.jpeg")
+include_graphics(here("figs/supplementary/S5.6.jpeg"))
 ```
@@
 ```{r fig S5.7, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S5.7.jpeg")
+include_graphics(here("figs/supplementary/S5.7.jpeg"))
 ```
@@
 ```{r fig S6.1, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S6.1.jpeg")
+include_graphics(here("figs/supplementary/S6.1.jpeg"))
 ```
@@
 ```{r fig S6.2, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S6.2.jpeg")
+include_graphics(here("figs/supplementary/S6.2.jpeg"))
 ```
@@
 ```{r fig S6.3, fig.fullwidth=TRUE}
-include_graphics("../figs/supplementary/S6.3.jpeg")
+include_graphics(here("figs/supplementary/S6.3.jpeg"))
 ```

### 2020-03-13 R/01_gis/05_make_shapefiles.R

@@ -9,6 +9,7 @@ library(dplyr)
 library(raster)
 library(rgdal)
 library(data.table)
+library(rmapshaper)
 source("R/functions/out.session.R")
 
 # Shapefiles
@@ -71,6 +72,19 @@ writeOGR(mada_communes, dsn = "data/processed/shapefiles", layer = "mada_commune
          driver = "ESRI Shapefile", overwrite_layer = TRUE)
 writeOGR(mada_districts, dsn = "data/processed/shapefiles", layer = "mada_districts", 
          driver = "ESRI Shapefile", overwrite_layer = TRUE)
+
+# Simplify shapefiles for easier plotting ---------------------------------------------------------
+mada_districts <- rmapshaper::ms_simplify(mada_districts)
+format(object.size(mada_districts), units = "Mb")
+
+mada_communes <- rmapshaper::ms_simplify(mada_communes)
+format(object.size(mada_communes), units = "Mb")
+
+# convert to geojson using command line gdal
+writeOGR(mada_districts, "data/processed/shapefiles", layer = "mada_districts_simple", 
+         driver = "ESRI Shapefile", overwrite = TRUE)
+writeOGR(mada_communes, "data/processed/shapefiles", layer = "mada_communes_simple", 
+         driver = "ESRI Shapefile", overwrite = TRUE)
 
 # Saving session info
 out.session(path = "R/01_gis/05_make_shapefiles.R", filename = "output/log_local.csv")

### 2020-03-13 R/01_gis/06_groundtruth_ttimes.R

@@ -26,7 +26,6 @@ pop1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
 friction_masked <- raster("data/processed/rasters/friction_mada_masked.tif")
 
 # Groundtruthing ------------------------------------------------------------------------------
-
 # Mora self reported ttimes
 mora <- read.csv("data/processed/bitedata/moramanga_ttimes.csv")
 mada_communes$catchment <- ctar_metadata$CTAR[mada_communes$catchment]

### 2020-03-13 R/figures/M3_ttimes.R

@@ -22,7 +22,6 @@ base_times <- raster("output/ttimes/base_ttimes.tif")
 pop1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
 friction_masked <- raster("data/processed/rasters/friction_mada_masked.tif")
 
-
 # Figure: raw ttimes + prop pop  ----------------------------------------------------------------
 ttime_cols <- c('#fff7f3', '#fde0dd', '#fcc5c0', '#fa9fb5', '#f768a1', '#dd3497', '#ae017e', 
                 '#7a0177', '#49006a')
@@ -31,19 +30,13 @@ ttime_labs <- c("< 1", "1 - 2", "2 - 3", "3 - 4", "4 - 6",  "6 - 8", "8 - 10", "
 names(ttime_cols) <- ttime_labs
 
 base_df <- as.data.frame(base_times, xy = TRUE)
-base_df$pop <- getValues(pop1x1)
-  
+
 ttimes_A <- ggplot() + 
   geom_raster(data = base_df, aes(x, y, 
                                   fill = cut(base_ttimes/60, breaks = ttime_breaks, 
                                              labels = ttime_labs))) + 
   scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)",
                     drop = FALSE, na.value = "black") +
-  scale_alpha_continuous(range = c(0.1, 1), 
-                         trans = "log", breaks = c(10, 1000, 1e4, 1e5), na.value = 1,
-                         guide = guide_legend(title = "log(pop)", override.aes = list(fill = '#dd3497',
-                                                                                      linetype = 1,
-                                                                                      color = '#dd3497'))) +
   geom_point(data = ctar_metadata, aes(x = LONGITUDE, y = LATITUDE), color = "darkgrey", 
              shape = 4,
              stroke = 2) +
@@ -66,21 +59,12 @@ prop_B <- ggplot(data = prop_pop_ttimes, aes(x = cut_times, y = prop_pop,
                                               fill = cut_times)) +
   geom_col(color = "grey") +
   scale_fill_manual(values = ttime_cols, na.translate = FALSE, name = "Travel times \n (hrs)",
-                    drop = FALSE, guide = "none") +
[…]

### 2020-03-13 R/figures/M4_biteinc.R

@@ -1,0 +1,93 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Making maps for figure 1 of ctar and patient locations
+#' Details: network style figure for vizualizing where patients are reporting to in each 
+#' district
+# ------------------------------------------------------------------------------------------------ #
+
+# Libraries
+library(patchwork)
+library(tidyverse)
+library(rgdal)
+library(data.table)
+library(cowplot)
+select <- dplyr::select
+
+# Read in raw data (not processed)
+district_bites <- fread("output/bites/district_bites.csv")
+mora_bites <- fread("output/bites/mora_bites.csv")
+ctar_metadata <- fread("data/raw/ctar_metadata.csv")
+mada_communes <- readOGR("data/processed/shapefiles/mada_communes_simple.shp")
+mada_districts <- readOGR("data/processed/shapefiles/mada_districts_simple.shp")
+
+# colors by catchment
+catch_cols <- c("#FCC56F","#004D43", "#7A4900", "#CBCDD2", "#006b3c", "#EFF2F1",
+                "#21abcd", "#B79762", "#FFDBE5", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
+                "#F2300F", "#FF2F80","#61615A", "#4A3B53", "#6B7900", "#00C2A0", "#FFAA92",
+                "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
+catch_fills <- catch_cols
+names(catch_cols) <- ctar_metadata$CTAR
+names(catch_fills) <- ctar_metadata$CTAR
+
+# District bite incidence estimates -----------------------------------------------------------
+district_bites %>%
+  select(group_name = distcode, pop, catchment, ttimes_wtd, avg_bites, min_bites, max_bites, 
+         sd_bites, 
+         nobs) %>%
+  mutate(dataset = "National", 
+         nobs = ifelse(is.na(nobs), 1, nobs)) -> bites_plot
+
+mora_bites %>%
+  select(group_name = distcode, pop, catchment, ttimes_wtd, avg_bites) %>%
+  group_by(group_name, catchment) %>%
+  summarize(avg_bites = sum(avg_bites)) %>%
+  filter(avg_bites > 10) %>%
[…]

### 2020-03-13 R/figures/SM1_basetimes.R

@@ -22,7 +22,6 @@ mada_districts <- readOGR("data/processed/shapefiles/mada_districts_simple.shp")
 
 # Input rasters
 friction <- raster("data/processed/rasters/friction_mada_masked.tif")
-base_times <- raster("output/ttimes/base_ttimes.tif")
 pop1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
 
 # Shapefiles to dataframes
@@ -43,7 +42,7 @@ friction <- ggplot() +
              shape = 4,
              stroke = 2) +
   scale_fill_distiller(palette = "Purples", na.value = "white", direction = -1,
-                       name = "Friction value \n (travel speed in minutes/meter)") +
+                       name = "Friction value \n (travel speed \n minutes per meter)") +
   theme_map() +
   labs(tag = "A") +
   coord_quickmap()
@@ -51,72 +50,22 @@ friction <- ggplot() +
 # population
 pop_cols <- c('#f0f9e8','#ccebc5','#a8ddb5','#7bccc4','#4eb3d3','#2b8cbe','#08589e')
 pop_breaks <- c(1e-6, 1, 5, 10, 100, 1e4, 1e5, 1e6)
-pop_labs <- c("< 1", "1-5", "5 - 10", "10 - 100", "100 - 1000", "1000 - 10000", "10000 +")
+pop_labs <- c("< 1", "1 - 5", "5 - 10", "10 - 100", "100 - 1000", "1000 - 10000", "10000 +")
 names(pop_cols) <- pop_labs
 
 pop <- ggplot() + 
   geom_raster(data = pop_df, aes(x, y, fill = cut(wp_2015_1x1, breaks = pop_breaks, 
                                                   labels = pop_labs))) + 
-  scale_fill_manual(values = pop_cols, na.translate = FALSE, name = "Pop per km^2",
+  scale_fill_manual(values = pop_cols, na.translate = FALSE, 
+                    name = bquote('Pop per'~km^2),
                     drop = FALSE, na.value = "white") +
   theme_map() +
   theme(text = element_text(size = 14)) +
   labs(tag = "B") +
   coord_quickmap()
 
-inputs <- friction | pop
-ggsave("figs/supplementary/inputs.jpeg", inputs, device = "jpeg", height = 8, width = 12)
-
-# Plotting GIS outputs (ttimes @ grid/district/commune) ------------------------------------------
[…]

### 2020-03-13 R/figures/SM2_bitedata.R

@@ -1,38 +1,137 @@
-####################################################################################################
-##' Looking at reporting cut offs and translating to vial estimates 
-##' Details:  
-##' Author: Malavika Rajeev 
-####################################################################################################
-rm(list = ls())
+# ------------------------------------------------------------------------------------------------ #
+#' Correcting for undersubmission of forms & comparing to predicted vial ests
+#' Looking at prop of bites reporting within catchment
+# ------------------------------------------------------------------------------------------------ #
 
-## Libraries
+# Libraries
 library(tidyverse)
+library(rgdal)
+library(rgeos)
 library(data.table)
 library(lubridate)
 library(patchwork)
 library(cowplot)
 select <- dplyr::select
-source("R/functions/utils.R")
+source("R/functions/out.session.R")
 
-## Read in metadata
+# Read in metadata
 ctar_metadata <- fread("data/raw/ctar_metadata.csv")
-mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
+mada_communes <- readOGR("data/processed/shapefiles/mada_communes_simple.shp")
+mada_districts <- readOGR("data/processed/shapefiles/mada_districts_simple.shp")
+national <- fread("data/processed/bitedata/national.csv")
 
-##' Plotting sensitivity to rle cut-offs
-##' ------------------------------------------------------------------------------------------------
-## throughput and reporting
+# Catchment maps (commune = fill & district = color ) ------------------------------------------
+# colors
+catch_cols <- c("#FCC56F","#004D43", "#7A4900", "#CBCDD2", "#006b3c", "#EFF2F1",
+                "#21abcd", "#B79762", "#FFDBE5", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
+                "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
+                "#F2300F", "#FF2F80","#61615A", "#4A3B53", "#6B7900", "#00C2A0", "#FFAA92",
[…]

### 2020-03-13 R/functions/bezier.R

@@ -1,22 +1,69 @@
-####################################################################################################
-##' Bezier curve control points 
-##' Details: adapted from Dudas et al. Curonia 
-##' https://github.com/evogytis/baltic/blob/master/curonia.ipynb
-##' Author: Malavika Rajeev 
-####################################################################################################
-get.bezier.pts <- function(origin, destination, frac, transform) {
+#' Title
+#' Description
+#' Details
+#' @param Paramters
+#' @return Returned
+#' @section Dependencies:
+#'     List dependencies here, i.e. packages and other functions
+#' Bezier curve control points 
+#' Details: adapted from Dudas et al. Curonia 
+#' https://github.com/evogytis/baltic/blob/master/curonia.ipynb
+#' Output = data.frame with to/mid/from
+get.bezier.pts <- function(from, to, frac = 0.8, transform = function(x) sqrt(1/x)*0.5) {
   
-  sign <- ifelse(origin[, "x"] > destination[, "x"], -1, 1)
-  slope <- (destination[, "y"] - origin[, "y"]) / (destination[, "x"] - origin[, "x"])
-  distance <- sqrt((destination[, "y"] - origin[, "y"])^2 + (destination[, "x"] - origin[, "x"])^2) 
+  sign <- ifelse(from[, "long"] > to[, "long"], -1, 1)
+  slope <- (to[, "lat"] - from[, "lat"]) / (to[, "long"] - from[, "long"])
+  distance <- sqrt((to[, "lat"] - from[, "lat"])^2 + (to[, "long"] - from[, "long"])^2) 
   height <- transform(distance)
   
-  hdist <- sqrt(height^2 + (distance*frac)^2) ## distance between desired height and point along line
+  hdist <- sqrt(height^2 + (distance*frac)^2) # distance between desired height and point along line
+
+  from <-  data.frame(long = from[, "long"], lat = from[, "lat"], index = 1, group = 1:nrow(from))
+  ctrl <- data.frame(long = from[, "long"] + hdist*cos(atan(height/distance/frac) + atan(slope))*sign,
+                     lat = from[, "lat"] + hdist*sin(atan(height/distance/frac) + atan(slope))*sign, 
+                     index = 2, group = 1:nrow(from)) # the magic control point
[…]

### 2020-04-09 docs/manuscript.Rmd
adding processing scripts for packages etc

@@
 library(here)
+library(magrittr)
 
 opts_chunk$set(echo = FALSE, warning = FALSE)
-out_main_fig = TRUE
 
@@
 mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))
+
+trans_perc <- function(x) {
+  round(x, 2)*100
+}
 
@@
 To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).
-
 Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale.
-We then compared three different metrics of access:
-
-1) Travel time
-2) Weighted travel time
-3) Distance
-
-extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell.
+We then extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell.
 We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC.
@@
 For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
-To correct for this, we estimated clinic level undersubmission as the proportion of days on which forms were submitted in a given year, excluding any periods for which there were no forms submitted for 15 consecutive days.
-We then divided the total number of bites reported to each district at each clinic in a given year by the undersubmission rate for that clinic in that year.
-Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates.
-The data from the Moramanga district were completely reported across the 31 month period.
+To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1).
+We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period.
+We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.
+
+Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported.
+For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs.
+4 doses total.
+During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit.
+Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared + any additional wasteage from unused doses discarded at the end of the day.
+While these simplifying assumptions do not capture variation in adherence, completion, delays, or errors in administration, PEP completion is overall high in Madagascar and in other settings where PEP is provided at no direct costs to patients, so our estimates at 3 doses per patient vs.
+4 are likely encompassing much of this variation at the annual level.
+
+We then divided the total number of bites reported to each clinic from each clinic in a given year by the undersubmission rate for eachclinic.
+Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.
+The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.
 
@@
 ## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
-We modeled the number of reported bites as a function of travel time ($T$) as follows: 
-$$ \mu = e^{(\beta_{t}T \ + \ \beta_0 \ )}P$$
-where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T$) and the human population size ($P$) (an offset which scales the predicted number of bites) for a given source location (district or commune).
+We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
+$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
+$$ y_i = Poisson(\mu_i) $$
+where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the predicted number of bites) for a given source location (district or commune).
 We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’.
@@
 
-$$\mu = \sum \limits_{j=1}^je^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$
+$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$
 
-As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel times.
-For the models with population size, we used a model framework with either population size alone ($\mu = e^{(\beta_pP + \beta_0 )}$) or population size and travel times ($\mu = e^{(\beta_{t}T + \beta_pP + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.
+As travel times are correlated with human population size (Figure S3.1), we also compared how well bites were predicted by human population size alone, and in combination with travel times.
+For the models with population size, we used a model framework with either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.
 
-For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercpet).
+For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercept).
+We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by comparing the ratio of the sum of squared residuals to the expected $\chi^2_{n - k}$ distribution given a poisson error structure [@gelman2006data].
+
+We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations).
+We used priors centered around zero (*norm(0, 10-6)*) except for the precision parameter for which we used an uniform distribution (*unif(0, 100)*).\
+
+The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.
+We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).
 
@@
 we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data).
-
-Since we're assuming for poisson errors: we tested for overdispersion.
-
-Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to uncorrected data.
+Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to the raw data uncorrected for submission (i.e.
+assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs.
+15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).
 
@@
 
-<!-- TO DO: simplify discussion here but bring in some figs, etc to Supplementary section to explain assumptions! -->
+We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2).
+As with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], to model uncertainty in parameter estimates we used triangular distributions for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures.
+The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).
 
-We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2).
-We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Specifically: annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination .
 To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$).
@@
 
-```{r fig 2, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M2.jpeg"))
+```{r fig 1, fig.fullwidth=TRUE}
+include_graphics(here("figs/main/M1_sankey.tiff"))
 ```
@@
 $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$).
-$E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text).
 $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$).
@@
 those who would have died given exposure, but instead received PEP.
-All parameter values and sources are listed in Table 1.*
+Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*
 
-$B_i$, the number of total reported bites, is drawn from a poisson distribution with the mean predicted number of bites predicted from the models of incidence as a function of travel times.
-To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures.
-We draw $p_{rabid}$ from a uniform distribution ranging from 0.2 - 0.6 based on investigations of bite patients reporting in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018].
+$B_i$, the number of total reported bites, is the mean predicted number of bites from the models of bite incidence as a function of travel times.
+To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6.
 So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
@@
 We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
-We subtract $R_i$ from $E_i$ to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated as:
-$$ D_i = binom(U_i,\ p_{death}) $$
-Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_i = binom(R_i, \ p_{death})$).
+We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$).
+Deaths due to rabies in each administrative unit ($D_i$) are then estimated by mutliplying $D_i$ times $p_{death}$.
+Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_ip_{death}$).
 Table 1 lists all parameter values and sources used in the decision tree.
@@
 |:----------------|:----------------:|:----------------------------------|:---------:|
-| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@whorabiesmodelingconsortiumPotentialImpactImproved2019] |
-| $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$.
-| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
-| $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
+| $E_i$ |	Triangular(a = 15.6 b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see S4|
+| $p_{rabid}$	| Triangular(a = 0.2 b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
 | $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
@@
 ## candidate points
-csbs <- read.csv(here("data/raw/csbs.csv", stringsAsFactors = FALSE))
+csbs <- read.csv(here("data/raw/csbs.csv"), stringsAsFactors = FALSE)
 csbs %>% 
@@
 Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country.
-We added the clinic which minimized this metric weighted by the reduction in travel times, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially.
+We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially.
 In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons).
@@
 We estimate how expansion to additional clinics may impact vaccine demand.
-Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment.
-As clinics are added, multiple clinics may serve a single district, but in general, the majority of the population within a given commune are closest to a single clinic (Fig S5.3B).
-Therefore, we assign the catchment clinic as the one which is the closest for the majority of people living in a commune.
-For the district model, we disaggregate the reported bites to the commune level (assuming uniform incidence within the district).
-For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
+Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level.
+As clinics are added, multiple clinics may serve a single district or commune (Fig TO CHECK S5.3B).
+Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest (Fig CHECK maps).
+For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
 all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses.
@@
 ## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
-We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden.
-Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden.
-We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplementary section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
+To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and models of burden.
+Additionally, we also examine the systematic variation in human rabies exposure incidence with human population (a potential proxy for changes in the dog population, see section S6) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
+We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1).
+Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e.
+if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases.
+Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e.
+if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.
 
-## Data availability
+## Data and code 
 All analyses were done in R and using these packages (list packages with commas and citations).
@@
 
-## Estimating travel times to the closest ARMC
+## Estimating travel times to the closest ARMC {#results-ttimes}
 ```{r ttime stats}
@@
 ```{r fig M2, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M3_ttimes.jpeg"))
+include_graphics(here("figs/main/M2_ttimes.jpeg"))
 ```
@@
 
-## Estimating average annual bite incidence at the district and commune level
+## Estimating average annual bite incidence at the district and commune level {#results-biteinc}
 We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population.
 While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B).
-In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A) and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).
+In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A), and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).
 
@@
 Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1).
-However, we found that corrected estimates of reproted bites generated predictions of vial demand close to what was provisioned over the same time period ([Figure S2.2A and B](#figS2.2)).
-We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 3A).
+We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)), and correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned ([Figure S2.2A and B](#figS2.2)).
+We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A).
 After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).
@@
 outliers <- read.csv(here("output/stats/outliers.csv"))
-head(outliers)
 ```
@@
 Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship.
-There were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
-While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher incidence at low travel times at the commune level compared to the district.
+For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
+While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher reported incidence at low travel times at the commune level compared to the district.
 
 ## Modeling reported bite incidence as a function of travel times {#results-bitemods}
+All models and posterior estimates of parameters converged ({inline stat for max psrf & mpsrf}).
+Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
+However, when used to predict data that were not used to fit the models (i.e.
+model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop) generated reasonable predictions to the data (Figure S3.3).\
 
 All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2).
-Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
-Sensitivity to underreporting.
+Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
 For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2).
-In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3).
-These predictions were robust to how we corrected for incomplete submission of data by clinics ([supplementary section S4](#S4)).
-As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence.
-(OD COMP--use the Moramanga model with OD & the District model with OD)
+These predictions were robust to how we corrected for incomplete submission of data by clinics (Fig S3.5).
+The models without the random effect were significantly overdispersed, but the catchment random effect still captured additional variation beyond this overdispersion (Fig S3.4).
+The parameter estimates from the different models were similar, with the model fit to the Moramanga data showing the highest impact of travel times and the district model the lowest (Fig 5B-D), however the increased bite incidence observed at low travel times in the Moramanga data was not captured by the commune model (Fig 5A).
+To capture the range of estimates of travel time impact, we compared the district model fitted to the national data with the commune scale model fitted to the Moramanga data in subsequent analyses.
 
-```{r fig 3, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M3.jpeg"))
+```{r fig 5, fig.fullwidth=TRUE}
+include_graphics(here("figs/main/M5_mods.jpeg"))
 ```
-_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons.
-The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
-The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept.
-The envelopes shows the 95% CI given the random intercept for the commune and district models.
-The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept.
-The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model)._
+_**Figure 5.** Travel times as a predictor of reported bite incidence per 100,000 persons.
+A) The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
+The lines are the estimates using the posterior estimates of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and adjusted for overdispersion.
+The envelopes shows the 95% prediction interval.
+The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model).
+B - D) The posterior distribution of parameters from the respective models for the B) model intercept, C) travel time effect, and D) the catchment random effect (for national data only)._
 
@@
 We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_burden$averted_mean)[1])` - `r round(sort(natl_burden$averted_mean)[nrow(natl_burden)])` deaths due to rabies are prevented through provisioning of PEP each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally at the district level, but also within districts (Figure 4).
+In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally when burden is modeled at the district level, with even stronger patterns of increase in burden with travel times when modeling burden at the commune level with estimates from the Moramanga data (Figure 4).
+The estimates of burden from the model fitted to the Moramanga data result in estimates of burden that are lower at the national level, but higher relatively in places with high travel times {fix!}.
+However, there is considerable uncertainty in these estimates, in particular given the district model of bite incidence.
 
-*__Table 2.__ Model predictions of mean annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% CIs in parentheses.*
+*__Table 2.__ Model predictions of average annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% prediction interval in parentheses.*
 
@@
   mutate_at(vars(bites_mean:averted_lower), round) %>%
-  mutate(bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
+  mutate(scale = case_when(scale == "Commune" ~ "Commune (Moramanga data)",
+                           scale == "District" ~ "District (National data)"), 
+         bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
          deaths = glue("{deaths_mean} ({deaths_lower} - {deaths_upper})"),
@@
 ```{r fig 4, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M4.jpeg"))
+include_graphics(here("figs/main/M6_burden_base.jpeg"))
 ```
 *__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
-__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model for all communes in a district.
+__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model fit to the national data and the squares show the incidence for the commune model fit to the Moramanga data for all communes in a district.
 Points are colored by and districts are ordered by travel times.
@@
 ## The impact of expanding PEP access to additional clinics {#results-incremental}
+```{r}
+natl_preds_all <- read.csv(here("output/stats/natl_preds_all.csv"))
+natl_preds_all %>%
+  group_by(scale) %>%
+  arrange(scenario) %>%
+  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
+trends <- read.csv(here("output/stats/trend_preds.csv"))
 
- We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 100 {inline} clinics reducing burden by almost 50% {inline}.
-However, benefits accrue less rapidly as ARMC are added, saturating after approximately 200 {inline} additional clinics.
-As clinics are added, reported bite incidence saturates, and instead there are shifts in which clinics bites are reported to ([see supplementary section S5](#S5)).
-Even at this maximal PEP access, our model still predicts > 100 {inline} rabies deaths per annum, as average reporting of rabid exposures still remains at approximately 85 % ([Fig S5.4](#figS5.4)).
-This is not due to saturation of reporting (see distribution of reporting fig in S5), but instead to to broad patterns of remoteness across Mada.
-Vaccine demand also increases, and more vials are needed per averted death as PEP access is expanded.
-Our model predictes an increase from 40,000 {inline} vials under current clinic provisioning, to ~50,000 {inline} vials with 200 {inline} clinics, ~60,000 {inline} and approximately 90,000 {inline} vials if all clinics were to provision PEP.
+```
+
+We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 250 clinics reducing burden by almost `r round(min(1 - props$deaths_mean[props$scenario == 250])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 250])*100, 2)`% (Fig M7.A).
+However, benefits accrue less rapidly as ARMC are added: with approximately 700 clinics, we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == 700])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 700])*100, 2)`, while even when all CSB IIs in the country provision PEP, the most deaths are reduced is between `r round(min(1 - props$deaths_mean[props$scenario == max(props$scenario)])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == max(props$scenario)])*100, 2)`%.
+This saturation is a result of the landscape of clinics across the country, as most CSBs are in areas with already low travel times (Fig S5.X).
+Therefore, as clinics are added, reported bite incidence saturates (Fig S5.Z), and instead there are shifts in which clinics bites are reported to (Fig S5.Y).
+Even at this maximal PEP access, our model still predicts `r round(min(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures still remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.N).
+
+Vial demand also increases, and more vials are needed per averted death as PEP access is expanded.
+Our model predictes an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
+visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` {inline} vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
+Most bites become reported within a commune--why does throughput decrease!
+
+Similar to our baseline estimates of burden, there was significant uncertainty in these estimates, in particular for the district model fit to the national data.
 
@@
 ```{r fig 5, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M5.jpeg"))
+include_graphics(here("figs/main/M7_addARMC.jpeg"))
 ```
@@
 *__Figure 5.
-A)__ Decrease in deaths due to rabies, __B)__ increase in vials needed per death averted, and __C)__ increase in total # of vials as additional ARMC are added at the national level based on the two models of reported bites.
-Lines are the mean of 1000 simulations with envelopes representing 95% CIs.
+A)__ Decrease in deaths due to rabies, __B)__ increase in total # of vials as additional ARMC are added at the national level , and __C)__ increase in vials needed per death averted based on the two models of reported bites.
+Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
 The “...” on the axes indicates a break in the axes, and the last point is the burden and vial predictions when all CSBS in the country provision PEP (maximum access).*
 
-## Sensitivity of burden estimates to decision tree parameters {#results-sensitivity}
-There is significant uncertainty in the predicted numbers of rabies deaths, but the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (see [supplementary section S6](#S6)).
-Our estimate of the absolute number of deaths and the predicted impact of access is most sensitive to assumptions of underlying rabies exposure incidence in the population.
-Other parameters don't contribute nearly as much uncertainty.
-In some ways is it better to present this here?
+## Sensitivity of estimates to bite incidence and decision tree model parameters {#results-sensitivity}
 
-<!-- TO DO: Univariate sensitivity analyses -->
+Overall, our sensitivity analyses show that while there is significant uncertainty in the absolute numbers of deaths due to rabies, patterns with increasing risk of death with reduced access and the potential for expanding access to PEP to reduce these deaths appear to be robust (Figs S6.2, S6.3).
+In general, for a given incidence level the higher our assumption of the proportion of reported bites which are rabid, the lower the estimated burden.
+The lower our estimate of maximum reporting, the higher overall burden of rabies, and as this parameter decreases, the relationship between travel times and rabies deaths are dampened.
+Scaling with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths.
+Finally, raw numbers of deaths greatly depend on what we assume the rabies exposure incidence is in the population.
+
+To quantify which parameters contribute to the uncertainty observed in our estimates, we performed a univariate sensitivity analyses, varying parameters in both the model of bite incidence (taking the lower and upper 95% confidence intervals of the posterior estimates, Table 1) and the decision tree model (fixing parameters at the ranges used in the parameter distributions, Table 2).
+These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden (Fig S6.Y).
+For the impacts of expanded access, what we assume is the maximum reporting of rabies exposures given minimal travel times ($rho_max$) is an important factor for the commune model (Fig S6.X), because as clinics are added, more locations are shifted to this threshold (Fig S5.X).
+For the district model, $p_{rabid}$ and $E_i$ are also important in determining the magnitude of reduction of burden.
+Despite the large absolute ranges of parameter estimates, the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (Fig S6.N).
+
+Uncertainty in bite model parameters contribute less to variation of estimates of burden.
+For the estimates of vial demand, uncertainty around the intercept of the model (i.e.
+the baseline reported bite incidence) has the greatest impact, rather than the travel time effect or the catchment effect (Fig S6.4).
+
+```{r se}
+se_pars <- read.csv(here("output/sensitivity/se_pars.csv"))
+
+se_pars %>%
+  pivot_longer(exp_min:beta_0) %>%
+  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
+                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
+  group_by(scale, par = name) %>%
+  summarize(upper = max(value), lower = min(value)) %>%
+  mutate(range = glue("{lower} \U2012 {upper}")) -> se_table_all
+
+se_table_all %>%
+  filter(grepl("beta|sigma", par)) %>%
+  pivot_wider(par, values_from = range, names_from = scale) %>%
+  as.data.table(.) -> bites_se
+se_table_all %>%
+  filter(!grepl("beta|sigma|min", par), scale != "Commune") %>%
+  as.data.table(.) -> dtree_se
+```
+
+Table:**Table 3.** Range of parameters used in models of bite incidence for the univariate sensitivity analyses.\
+
+| **Parameter**   | **Description**        | **Commune model (Moramanga data)** | **District model (National data)** | 
+|:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
+| $\beta_t$ | Travel time effect	| `r bites_se$Commune[par == "beta_t"]` | `r bites_se$Commune[par == "beta_t"]`
+| $\beta_0$ | Model intercept	| `r bites_se$Commune[par == "beta_0"]` | `r bites_se$Commune[par == "beta_0"]`
+| $\sigma_0$ | Standard deviation of intercept by catchment	| -- | `r bites_se$Commune[par == "sigma_0"]`
+
+Table:**Table 4.** Range of parameters used in decision tree model for the univariate sensitivity analyses.\
+
+| **Parameter**   | **Description**        | **Range**	                 | **Notes**|
+|:----------------|:----------------------:|:----------------------------|:---------:|
+| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se$range[par == "exp_max"]`  	
+| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se$range[par == "p_rab_max"]` 	
+| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se$range[par == "rho_max"]`	 
+| $p_{death}$ | The probability of death given a rabies exposure^[1] | `r dtree_se$range[par == "p_death"]`	
+[1] 95% CI of estimate from @changaluchaNeedImproveAccess2018|
 
 # DISCUSSION  
+```{r}
+natl_preds_all %>%
+  filter(scenario == 0) %>%
+  mutate(prop_averted = averted_mean/(deaths_mean + averted_mean)) %>%
+  select(prop_averted) -> prop_averted
+```
 
@@
 Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar.
-Using an adapted decision tree framework and assuming endemic rabies incidence, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
-Even given only 31 clinics, we estimate that current provisioning averts 50% {inline} of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by 50 {inline}.
+Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
+Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
+We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.
 However, our model predicts that these effects saturate, and that even given maximum expansion of PEP (i.e.
-to all major health centers (n = 1676) in the country), over 100 deaths {inline} may still occur nationally.
-In addition, vial demand outpaces additional deaths averted, with more vials needed per averted death as clinics are added.
+to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country and the predictions of maximum reported bite incidence even given low travel times.
+In addition, vial demand slightly outpaces additional deaths averted, with more vials needed per averted death as clinics are added.
+However, after a certain number of clinics are added, rather than large increases in reported incidence and thus vial demand, we estimate shifts in the clinic locations where bites are reported to.
 
@@
 This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018].
-In Madagascar, the presence of paved roads does not necessarily reflect their quality and the modes of transport used to travel these roads.
+In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
 Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune.
@@
 
-<!-- TO DO: UPDATE TO REFLECT CHANGES IN OD MODS ETC. -->
-
-We did include a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times.
-This random component of the model did improve fit to the data, suggesting that variation at the clinic level (i.e.
+We included a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times.
+This random component of the model improved fit to the data, suggesting that variation at the clinic level (i.e.
 differences in clinic functioning such as vaccine availability) and at the catchment level (i.e.
 differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
+In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access.
+The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times, and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times.
+However, we lacked data from other catchments at the commune scale, so it is difficult to say how broadly these patterns apply.
+The district model fitted to the national data therefore likely shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.
 
-Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior.
-We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning.
-
-<!-- TO DO: CLAIRE'S MSC ON OTHER REASONS + CLINIC FUNCTIONING -->
-<!-- TO DO: OUTLIERS...SCALE? -->
-
-<!-- TO DO: UPDATE BASED ON CHANGES TO BURDEN MODEL ETC. -->
-Our modeling framework makes strong assumptions about baseline rabies exposure incidence, assuming a range of 11 - 55 exposures per 100,000 persons in Madagascar.
-We base these estimates on a reasonable range of human:dog ratios [@knobelReevaluatingBurdenRabies2005; @ratsitorahinaDogEcologyDemography2009; @sambo2018estimating] and dog rabies incidence level [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-A previous study in the Moramanga District in Madagascar estimated a range of 42 – 110 exposures per 100,000 persons, however this was over a 16 month period with frequent rabies cases and multiple human deaths [@rajeevHealthcareUtilizationProvisioning2018].
-As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we sample uniformly over this range and over the proportion of reported bites estimated to be rabid (0.2 - 0.6 from the Moramanga data) to generate a mean estimate of burden.
+Our modeling framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden.
 Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same.
-While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies incidence.
+While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies exposure incidence.
 Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.
-
-The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
-Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.
 
@@
 
-<!-- TO DO: CAT 1 -->
+Claire's MsC + Helene's comments here!
+Other ways to designate clinics
+Shortages!
+
+The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
+Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.
 
@@
 
-<!-- TO DO: CASE STUDIES -->
+Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior.
+We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.
 
@@
 ### Antananarivo (too identifiable?)
-In 2016, A middle aged taxi driver died of rabies in suburban Antananarivo.
+A middle aged taxi driver died of rabies in suburban Antananarivo.
 Bitten by an unknown dog.
@@
 ## Broader context
-```{r}
-burden_preds <- fread(here("output/preds/complete/burden_filled.csv"))
-## Grouped to district
-burden_preds$distcode <- mada_communes$distcode[burden_preds$names]
-burden_preds$commcode <- mada_communes$commcode[burden_preds$names]
-
-burden_preds %>%
-  filter(distcode == "MG33314", scenario == 0) %>%
-  group_by(scale) %>%
-  summarise(death_inc = sum(deaths_mean)/sum(pop)*1e5,
-            reporting = mean(reporting_mean)) -> mora_data
-
-```
 While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings (1 - 5 subnationally, on average 1.5, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009].
@@
 
-# SUPPLEMENTARY MATERIALS AND METHODS
+# SUPPLEMENTARY FIGURES & TABLES
 
 ## S1. Estimating travel times to the closest ARMC {#S1}
-[*jump back to main methods section: Spatial covariate data*](#methods-spatial)
+[*jump back to main results section: estimating travel times*](#results-ttimes)
 
@@
 ## S2. Estimating average annual bite incidence at the district and commune level
-[*jump back to main methods section: Bite patient data*](#methods-bitedata)  
-CATCHMENT
-Most districts and communes a single clinic served the majority of the population by this metric (Fig S2.1B).
-In addition, our data show that the majority of bite patients do report to the closest clinic as assigned by our travel time metric (Fig S2.1C).
+[*jump back to main results section: Bite patient data*](#results-biteinc)
 
@@
 __C)__ The proportion of bites reported to each clinic which originated from a district within the assigned catchment.*
-
-To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1).
-We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period.
-We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.
-
-Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported.
-For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs.
-4 doses total.
-During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit.
-Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared + any additional wasteage from unused doses discarded at the end of the day.
-While these simplifying assumptions do not capture variation in adherence, completion, delays, or errors in administration, PEP completion is overall high in Madagascar and in other settings where PEP is provided at no direct costs to patients, so our estimates at 3 doses per patient vs.
-4 are likely encompassing much of this variation at the annual level.
-
-We find that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)).
-Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned.
 
@@
 
-## S3. Modeling reported bite incidence as a function of travel times
-<!--  TO DO = COMBINE SECTIONS S3/S4 -->
-### Candidate models of bite incidence {#S3}
+## S3. Modeling reported bite incidence as a function of travel times {#S3}
 
 [*jump back to main results section: Travel times as a predictor of bites*](#results-bitemods)
-
-We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations).
-We used priors centered around zero (*norm(0, 10-6)*) except for the precision parameter for which we used an uniform distribution (*unif(0, 100)*).\
-
-The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.
-We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).
-All models and posterior estimates of parameters converged ({inline stat for max psrf & mpsrf}) Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
-However, when used to predict data that were not used to fit the models (i.e.
-model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop) generated reasonable predictions to the data (Figure S3.3).\
 
@@
 ```{r fig S3.2, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S3.2.jpeg"))
+include_graphics(here("figs/supplementary/S3.1_corrs.jpeg"))
+```
+_**Figure S3.1.** Correlation between travel time in hours and population size of administrative unit the district and commune scale._
+
+```{r fig S3.2, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.2_fitted.jpeg"))
 ```
 _**Figure S3.2.
-Prediction to data used to fit each model.** Log of the observed bites against the predicted bites using the mean of the posterior distributions for each parameter.
+Prediction to data used to fit each model.** Log of the observed bites against the log of predicted bites using the mean of the posterior distributions for each parameter.
 Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
 Colors show which data set  was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
-
-<!-- Example of how to link to full figure online -->
-<!-- [link to full fig](https://raw.githubusercontent.com/mrajeev08/ModelingChapter/master/figs/fig1.jpeg) -->
 
@@
 ```{r fig S3.3, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S3.3.jpeg"))
+include_graphics(here("figs/supplementary/S3.3_outfit.jpeg"))
 ```
@@
 Out of fit predictions to data.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data not used to fit the model.
-First two columns are the predictions from the commune and district model fitted to the national data for the Moramanga data with fixed and random intercepts.
-The third column are predictions from models fitted to the Moramanga data predicting the national data at the commune and district scale (only fixed intercept models).
-Rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._
+The first two columns are the predictions from the commune and district model fitted to the national data for the Moramanga data with fixed and random intercepts.
+The third column are predictions from models fitted to the Moramanga data for the national data at the commune and district scale (only fixed intercept models).
+Rows are the type of model structure with respect to the population covariate and colors show which data set was used for fitting as per Figure S3.2._
 
@@
 
-### Estimating overdispersion of poisson errors
+```{r fig S3.4, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.4_ODcheck.jpeg"))
+```
+_**Figure S3.4.** Overdispersion....
 
-### Sensitivity of models to correction for incomplete submission of patient forms 
-<!-- TO DO: only need to show this for main models (i.e. ttimes models) -->
-<!-- TO DO: add one for lower cut-off (i.e. 7 days consec without reporting) -->
-[*jump back to main results section: Models*](#results-bitemods)
-
-We also looked at the sensitivity of our model estimates to how we corrected our data for undersubmission. Specifically we compared results from models fitted x, y, x.
-
-We found that the model with population size as an offset (flatPop) still generated the closest predictions to the data (Table S4.1 for DIC comparisons and Fig S4.1 for predictions to the data).
-These models with a random intercept also resulted in qualitatively similar predictions of the impact of travel times on reported bite incidence, although correcting for undersubmission did impact the intercept (modulating the overall incidence estimates, but not the relationship with travel times, Fig S4.2).
-
-```{r fig S4.2, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S4.2.jpeg"))
+```{r fig S3.5, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.5_cutoffs.jpeg"))
 ```
-_**Figure S4.2.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with a random catchment intercept and with population as an offset.
-Panels show predictions from models fitted to (1) data corrected for both undersubmission and with Category I exposures excluded, (2) only corrected for undersubmission, (3) with only Category I exposures excluded, and the raw (4) data with no corrections (raw data).
+_**Figure S3.5.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with a random catchment intercept and with population as an offset.
+Panels show predictions from models fitted to (1) data corrected for both undersubmission by correcting for periods of at least 7 days where zero patient forms where submiited, (2) correcting for periods of 15 days where zero patient forms where submiited, and (3) with the raw data not correcting for undersubmission.
 The lines are the estimates using the mean of the posterior distributions of the parameters and the envelopes shows the 95% CI given the random intercept.
@@
 
+## S4. Range of rabies exposure incidence in people
+
 ## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
@@
 [*jump back to main results section: Incremental reduction of burden*](#results-incremental)
-
-```{r scenario stats}
-## candidate points
-csbs <- read.csv(here("data/raw/csbs.csv", stringsAsFactors = FALSE))
-csbs %>% 
-  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
-  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
-catches_incremental <- fread(here("output/stats/pop_by_catch.csv"))
-```
-
-We iterativel added clinics by their relative reductions in travel times for people living > 3 hours away from a clinic.
-Fig S5.1 shows the location and at what step clinics were added.
-`r max(catches_incremental$scenario[catches_incremental$scenario != max(catches_incremental$scenario)])` clinics were added before there were no clinics which reduced travel times for more than 0.01% of people who also lived > 3 hours away from a clinic.
-We also estimated travel times given that all additional CSB IIs provisioned PEP (an additional `r nrow(csbs) - 31` clinics).
-Fig S5.2 shows the changes in travel times at the grid cell and administrative unit scale as clinics are added.
-As clinics were added, more than one clinic could serve a single district (the proportion of the pop fig S5.3).
-Even at the maximum level of access, 10% of people still live more than 3 hours from a clinic hours (S5.4).
-Fig S.55 shows how average travel times, bite incidence, reporting of rabid exposures, and incidence of deaths at the commune and district scale decline as clinics are added.
-Fig S56 shows the shifts in these distribution as clinics are added.
-At the clinic level, Fig S5.7 shows the shifts in the distribution of catchment populations, annual vial demand, and average daily throughput as clinics are added (S5.8).
 
@@
 
-## SN. Assumptions of underlying rabies incidence 
-<!-- TO DO: HERE OR EARLIER BIT? -->
-
-We explore the sensitivity of our burden estimates and estimates of the impacts of PEP expansion to assumptions of rabies exposure incidence.
-Specifically, we estimate burden fixed at the minimum (11.1 exposures per 100k persons) and maximum (55.7 exposures per 100k persons)  exposure incidence within our estimated range.
-We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1).
-Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e.
-if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases.
-Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e.
-if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.
-
 ```{r fig S6.1, fig.fullwidth=TRUE}
@@
 We generate constrained scaling factors with human population size so that rabies exposure incidence in each location scales positively or negatively across the range of observed population sizes (panels)._
-
-Overall, our sensitivity analyses show that while there is significant uncertainty in the absolute numbers of deaths due to rabies, patterns with increasing risk of death with reduced access and the potential for expanding access to PEP to reduce these deaths appear to be robust (Figs S6.2, S6.3).
-In general, for a given incidence level the higher our assumption of the proportion of reported bites which are rabid, the lower the estimated burden.
-The lower our estimate of maximum reporting, the higher overall burden of rabies, and as this parameter decreases, the relationship between travel times and rabies deaths are dampened.
-Scaling with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
-more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths.
-Finally, raw numbers of deaths greatly depend on what we assume the rabies exposure incidence is in the population.
 

### 2020-04-09 R/01_gis/01_process_rasters.R

@@ -8,9 +8,11 @@
 rm(list=ls())
 library(malariaAtlas) # for friction surface
 library(raster) # for reading in rasters
-library(rgdal) # for reading in shapefiles
+require(rgdal) # for reading in shapefiles
 library(gdistance) # for making transition object
 source("R/functions/out.session.R")
+
+here::here()
 
 # Shapefile for masking to (from OCHA)
 mada_districts <- readOGR("data/raw/shapefiles/districts/mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp")

### 2020-04-09 R/01_gis/pop_comp.R

@@ -1,0 +1,39 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Pop comparison of resampled World Pop 2015 estimates (older version from Linard et al. 2012) 
+# ------------------------------------------------------------------------------------------------ #
+
+# Setup
+library(raster)
+library(rgdal)
+library(data.table)
+# Data
+pop1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
+census_2018 <- read.csv("data/raw/census2018.csv")
+mada_districts <- readOGR("data/processed/shapefiles/mada_districts.shp")
+mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
+
+# Extract pops to region and check ------------------------------------------------------------
+mada_districts$reg_code <- substr(mada_districts$distcode, 1, 4)
+length(unique(mada_districts$regcode)) == length(unique(census_2018$Region)) 
+nrow(census_2018) == length(unique(census_2018$Region)) 
+
+reg_code <- getValues(rasterize(mada_districts, pop1x1))[!is.na(getValues(pop1x1))] # row of shp
+reg_code <- mada_districts$reg_code[reg_code]
+
+# Comparing total 
+sum(getValues(pop1x1), na.rm = TRUE) == sum(pop_by_region$pop_comp) # okay (no missing pops)
+
+# data table and agg by region code
+pop_by_region <- data.table(reg_code = reg_code, pop = getValues(pop1x1)[!is.na(getValues(pop1x1))])
+pop_by_region <- pop_by_region[, .(pop_comp = sum(pop, na.rm = TRUE)), by = reg_code] 
+pop_by_region <- as.data.table(census_2018)[pop_by_region, on = "reg_code"]
+write.csv(pop_by_region, "output/stats/pop_by_region.csv", row.names = FALSE)
+
+# Moramanga comparison (Ratovoson et al. 2019) ------------------------------------------------
+mora_pop_comp <- data.table(commune = c("Ambohibary", "Moramanga", "Ampasimpotsy Gara"), 
+                            commcode = c("MG33314030", "MG33314010", "MG33314050"), 
+                            pop_MHURAM = c(40767, 37634/2, 37634/2), 
+                            pop_est = mada_communes$pop[mada_communes$commcode %in% c("MG33314030", "MG33314010", "MG33314050")],
+                            notes = c("", rep("split between Ambohibary & Ampasimpotsy", 2)))
[…]

### 2020-04-09 R/03_bitemodels/01_run_bitemods.R

@@ -89,8 +89,8 @@ warnings()
 write.csv(mods_all, "output/mods/estimates.csv", row.names = FALSE)
 
 # Parse these from bash for where to put things
-sync_to <- "~/Documents/Projects/MadaAccess/output/mods/"
-sync_from <- "mrajeev@della.princeton.edu:~/MadaAccess/output/mods/"
+syncto <- "~/Documents/Projects/MadaAccess/output/mods/"
+syncfrom <- "mrajeev@della.princeton.edu:~/MadaAccess/output/mods/"
 
 # Close out 
 file_path <- "R/03_bitemodels/01_run_bitemods.R"

### 2020-04-09 R/04_addclinics/01_ttimes_candidates.R

@@ -8,7 +8,7 @@
 
 # set up cluster on single node with do Parallel
 library(doParallel)
-cl <- makeCluster(cores)
+cl <- makeCluster(18)
 registerDoParallel(cl)
 Sys.time()
 

### 2020-04-09 R/05_predictions/02_vials_incremental.R

@@ -47,7 +47,8 @@ model_ests %>%
 model_sds <- bind_rows(filter(model_sds_unadj, data_source == "National", scale == "District",
                               intercept == "random"), filter(model_sds_adj, 
                                                              data_source == "Moramanga"))
-# Vials given commune/district ttimes -----------------------------------------------------------------
+
+# Vials given commune/district ttimes -------------------------------------------------------
 
 # make df with the lookup + scale (reverse vec so big ones at end don't slow things down)
 lookup <- expand.grid(loop = rev(scenario_loop), scale = c("Commune", "District"))
@@ -87,7 +88,7 @@ foreach(j = iter(lookup, by = "row"), .combine = multicomb,
                                     beta_pop = 0, beta_pop_sd = sd$beta_pop, 
                                     sigma_0 = mean$sigma_0, known_alphas = NA, 
                                     pop_predict = mean$pop_predict, intercept = mean$intercept, 
-                                    trans = 1e5, known_catch = FALSE, nsims = 1000, type = "inc",
+                                    trans = 1e5, known_catch = FALSE, nsims = 10, type = "inc",
                                     dist = OD_dist)
           
           bites <- data.table(commcode = comm$commcode, scenario = comm$scenario,
@@ -95,21 +96,37 @@ foreach(j = iter(lookup, by = "row"), .combine = multicomb,
           
           check <- comm_all[bites, on = c("commcode", "scenario")]
           
-          cols <- names(check[, .SD, .SDcols = result.1:result.1000])
+          cols <- names(check[, .SD, .SDcols = grepl("result", names(check), fixed = TRUE)])
           
           check[, (cols) := lapply(.SD, function(x) x*prop_pop_catch), .SDcols = cols]
           bites_by_catch <- check[, lapply(.SD, sum, na.rm = TRUE), .SDcols = cols, 
                                   by = c("catchment", "scenario")]
           
+
+          
           # Get mean proportion of bites from each district
-          check_prop <- comm_all[bites, on = c("commcode", "scenario")]
-          catchment <- check_prop$catchment
+          check_prop <- comm_all[bites_by_catch, on = c("catchment", "scenario")]
+          
+          # make sure both check prop & check are in same order
[…]

### 2020-04-09 R/figures/M1_dtree.R

@@ -1,43 +1,11 @@
 # ------------------------------------------------------------------------------------------------ #
-#' Sankey diagram for decision tree                                                                                        
+#' Sankey diagram for decision tree   
 # ------------------------------------------------------------------------------------------------ #
 
 # set up
 library(ggplot2)
-library(cowplot)
-
-df <- data.frame(x = c(15, 42, 76), y = c(0, 2/(76 - 15), 0))
-ggplot(data = df, aes(x = x, y = y)) + 
-  geom_polygon(fill = "orchid", color = "NA", alpha = 1) +
-  geom_vline(xintercept = 42, color = "grey", linetype = 2) +
-  scale_x_continuous(breaks = df$x) +
-  theme_minimal_hgrid()  +
-  theme(axis.text.y = element_blank(), axis.ticks.x = element_blank(),
-        panel.grid.major.y = element_blank()) +
-  labs(x = "Rabies exposure incidence per 100k persons", y = "")
-
-df <- data.frame(x = c(0.2, 0.4, 0.6), y = c(0, 2/(0.6 - 0.4), 0))
-ggplot(data = df, aes(x = x, y = y)) + 
-  geom_polygon(fill = "darkblue", color = "NA", alpha = 1) +
-  geom_vline(xintercept = 0.4, color = "grey", linetype = 2) +
-  scale_x_continuous(breaks = df$x) +
-  theme_minimal_hgrid()  +
-  theme(axis.text.y = element_blank(), axis.ticks.x = element_blank(),
-        panel.grid.major.y = element_blank()) +
-  labs(x = "Probability of reported bite being true rabies exposure", y = "")
-
-df <- data.frame(x = c(0.13, 0.16, 0.20), y = c(0, 2/(0.20 - 0.13), 0))
-ggplot(data = df, aes(x = x, y = y)) + 
-  geom_polygon(fill = "darkred", color = "NA", alpha = 1) +
-  geom_vline(xintercept = 0.16, color = "grey", linetype = 2) +
-  scale_x_continuous(breaks = df$x) +
-  theme_minimal_hgrid()  +
-  theme(axis.text.y = element_blank(), axis.ticks.x = element_blank(),
-        panel.grid.major.y = element_blank()) +
-  labs(x = "Probability of death in the absence of PEP", y = "")
-
-
[…]

### 2020-04-09 R/figures/M2_ttimes.R

@@ -8,11 +8,9 @@ library(rgdal)
 library(raster)
 library(patchwork)
 library(cowplot)
-library(foreach)
 library(data.table)
 select <- dplyr::select
 source('R/functions/out.session.R')
-source('R/functions/ttime_functions.R')
 
 # data
 ctar_metadata <- read.csv("data/raw/ctar_metadata.csv")

### 2020-04-09 R/figures/M7_incremental.R

@@ -31,7 +31,7 @@ natl_preds$scenario[natl_preds$scenario == max(natl_preds$scenario)] <- max_tota
 # Colors
 # labs & cols
 scale_levs <- c("Commune", "District")
-scale_labs <- c("Moramanga", "District")
+scale_labs <- c("Commune (Moramanga data)", "District (National data)")
 model_cols <- c("#F2300F", "#35274A")
 names(scale_labs) <- scale_levs 
 names(model_cols) <- scale_levs
@@ -50,7 +50,7 @@ burden <- ggplot(data = filter(natl_preds, scenario != max_total),
   scale_x_continuous(breaks = c(0, 100, 300, 500, max_added, max_total), 
                      labels = c("baseline", 100, 300, 500, max_added, "max (1648)")) +
   annotate(geom = "text", x = max_added + 100, y = -30, label = "......") +
-  labs(x = "", y = "Mean deaths \n nationally",
+  labs(x = "", y = "Average annual deaths",
        tag = "A") +
   coord_cartesian(clip = "off", ylim = c(0, max(natl_preds$deaths_upper))) +
   theme_minimal_hgrid() +
@@ -59,12 +59,12 @@ burden <- ggplot(data = filter(natl_preds, scenario != max_total),
 
 # Vials per death
 vials_per_death <- ggplot(data = filter(natl_preds_all, scenario != max_total), 
-                   aes(x = scenario, y = vials_mean/deaths_mean, color = scale, fill = scale)) +
+                   aes(x = scenario, y = vials_mean/averted_mean, color = scale, fill = scale)) +
   geom_line(size = 1) +
-  geom_ribbon(aes(ymax = vials_upper/deaths_upper, ymin = vials_lower/deaths_lower), color = NA, alpha = 0.35) +
+  geom_ribbon(aes(ymax = vials_upper/averted_upper, ymin = vials_lower/averted_lower), color = NA, alpha = 0.35) +
   geom_pointrange(data = filter(natl_preds_all, scenario == max_total),
-                  aes(x = scenario, y = vials_mean/deaths_mean,
-                      ymax = vials_upper/deaths_upper, ymin = vials_lower/deaths_lower, 
+                  aes(x = scenario, y = vials_mean/averted_mean,
+                      ymax = vials_upper/averted_upper, ymin = vials_lower/averted_lower, 
                       color = scale),
                   position = position_dodge(width = 50)) +
[…]

### 2020-04-09 R/figures/SM1_ttimes.R

@@ -138,7 +138,6 @@ gtruth %>%
             ttimes_reported = mean(ttimes_reported, na.rm = TRUE),
             nobs = n()) -> mode_sum
 gtruth_mora <- filter(gtruth, type != "point", !is.na(mode))
-ggplot(data = mode_sum, aes(x = mode, y = nobs, fill = ttimes_reported)) + geom_col()
 
 S1.3_modes <- ggplot(data = gtruth_mora, aes(x = mode, y = ttimes_reported)) + 
   geom_boxplot(outlier.color = NA) +

### 2020-04-09 R/figures/SM3_bitemods.R

@@ -8,14 +8,8 @@ source("R/functions/out.session.R")
 source("R/functions/nested_facets.R")
 
 # libraries
-library(foreach)
-library(iterators)
 library(tidyverse)
-library(glue)
-library(rgdal)
-library(patchwork)
 library(cowplot)
-library(ggmcmc)
 library(data.table)
 select <- dplyr::select
 

### 2020-04-09 R/figures/SM4_exprange.R

@@ -1,0 +1,55 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Looking at reasonable range of exposure incidence
+#' Details: Pulling in district and commune estimates of travel times as clinics are added 
+# ------------------------------------------------------------------------------------------------ #
+
+# Set up ----------------------------------------------------------------------------------
+library(data.table)
+library(tidyverse)
+library(cowplot)
+source("R/functions/predict_functions.R")
+source("R/functions/out.session.R")
+select <- dplyr::select
+
+# From Claire's MSC
+range_people_ph <- c(4.13, 5.654)
+range_dogs_ph <- c(0.24, 0.26)
+# Range of hdrs then
+mora_hdrs <- range_people_ph/range_dogs_ph
+
+# incidence from hdrs of Moramanga
+inc_from_hdr(hdr = 5)
+inc_from_hdr(hdr = 25)
+
+# hdrs from exposure incidence estimated
+hdr_from_inc(inc = 76)
+hdr_from_inc(inc = 110)
+
+# How to show relationship between these things ...
+var_df <- expand.grid(hdr_val = seq(2, 40, by = 1), p_exp =  c(0.2, 0.38, 0.6), 
+            dog_inc =  seq(0.005, 0.025, by = 0.001))
+var_df %>%
+  mutate(exp_inc = inc_from_hdr(hdr = hdr_val, p_exp = p_exp, dog_inc = dog_inc)) -> inc_exps
+inc_labs <- as_labeller(c(`0.38`= "p[exp] == 0.38", `0.2` = "p[exp] == 0.2", `0.6` = "p[exp] == 0.6"), 
+                          label_parsed)
+
+S4.1_exprange <- ggplot(data = inc_exps, aes(x = hdr_val, y = dog_inc, 
+                            fill = cut(inc_exps$exp_inc, breaks = c(1, 15, 42, 76, 110, Inf),
+                                       labels = c("1 - 15", "15 - 42", "42 - 76", "76 - 110", "110 +")), 
+                            color = ifelse(exp_inc > 42 & exp_inc < 110, "red", NA))) +
+  geom_tile() +
+  geom_hline(yintercept = c(0.01, 0.02)) +
+  geom_vline(xintercept = c(8, 20)) +
+  geom_vline(xintercept = round(mora_hdrs), linetype = 2, color = "grey50") +
[…]

### 2020-04-09 R/install.R

@@ -1,0 +1,20 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Install needed packages                                                                                       
+# ------------------------------------------------------------------------------------------------ #
+
+# Get list of needed packages
+pkgs_used <- read.csv("output/logs/pkgs_used.csv", stringsAsFactors = FALSE)
+installed <- sapply(pkgs_used$pkgs, require, character.only = TRUE, quietly = TRUE)
+to_install <- pkgs_used$pkgs[!installed]
+
+# Install needed packages
+install.packages(to_install, repos = "http://cloud.r-project.org")
+
+# Comparing installed versions 
+sapply(pkgs_used$pkgs, require, character.only = TRUE, quietly = TRUE) # run this again to load pkgs
+version_comp <- unlist(lapply(pkgs_used$pkgs, function(x) as.character(packageVersion(x))))
+version_installed <- gsub("-", ".", as.character(pkgs_used$version))
+
+# If any installed are older, update these
+to_update <- pkgs_used$pkgs[version_comp < version_installed]
+update.packages(to_update, repos = "http://cloud.r-project.org")

### 2020-04-09 R/process_session.R

@@ -1,0 +1,69 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Processing out session & getting packages
+# ------------------------------------------------------------------------------------------------ #
+
+library(tidyverse)
+library(knitr)
+library(lubridate)
+library(RCurl)
+library(glue)
+
+# Do out.session here so any packages used in this script get included too!
+source("R/functions/out.session.R")
+out.session("R/process_session.R", filename = "output/log_local")
+
+# Find all .R & .Rmd files 
+files <- list.files(recursive = TRUE)
+files <- files[grep(".R$|.Rmd$", files)]
+files <- files[grep("archive", files, invert = TRUE)] # excluding these that are archived
+
+# Pull in local log
+log_local <- read.csv("output/log_local.csv", stringsAsFactors = FALSE)
+log_local <- as.data.frame(apply(log_local, 2, function (x) as.character(x)), 
+                           stringsAsFactors = FALSE)
+
+# Pull down log from cluster
+# Make sure you're vpn'd in before doing this
+# system("rsync -rvlzt --update mrajeev@della.princeton.edu:~/MadaAccess/log_cluster.csv ~/Documents/Projects/MadaAccess/output/")
+log_cluster <- read.csv("output/log_cluster.csv", stringsAsFactors = FALSE)
+log_cluster <- as.data.frame(apply(log_cluster, 2, function (x) as.character(x)),
+                             stringsAsFactors = FALSE)
+log_cluster %>%
+  filter(path %in% files) %>%
+  mutate(timestamp = parse_date_time(timestamp, "Ymd HMS")) %>%
+  group_by(path) %>%
+  filter(timestamp == max(timestamp)) -> log_cluster
+
+# Process the logs to get the package list
+log_local %>%
+  filter(path %in% files) %>%
+  mutate(timestamp = parse_date_time(timestamp, "Ymd HMS")) %>%
+  group_by(path) %>%
+  filter(timestamp == max(timestamp)) %>%
+  bind_rows(log_cluster) -> log_scripts
+
+# write log out of when scripts last ran
+write.csv(log_scripts, "output/logs/log_last_ran.csv", row.names = FALSE)
+
+log_scripts %>%
+  pivot_longer(-(ran:running), names_to = "pkgs", values_to = "version") %>%
+  ungroup() %>%
[…]

### 2020-04-19 docs/manuscript.Rmd
WIP miscellaneous edits
- adding in out sessions to some figs
- sub commands to others
- work in progress!

@@
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
-bibliography: ["bib/refs_all.bib"]
-# bibliography_main: ["bib/refs_main.bib"]
-# bibliography_supp: ["bib/refs_supp.bib"]
+bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
+# bibliography_main: ["bib/refs_all.bib"]
+# bibliography_supp: ["bib/refs_pkgs.bib"]
 ---
@@
 ```{r fig 1, fig.fullwidth=TRUE}
-include_graphics(here("figs/main/M1_sankey.tiff"))
+include_graphics(here("figs/main/M1_sankey.jpeg"))
 ```
@@
 ## Data and code 
-All analyses were done in R and using these packages (list packages with commas and citations).
+All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}.
 All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo.
 The raw bite patient data at the national level is maintained by IPM.
-This database was last queried on *date here* for these analyses.
-Anonymized raw data from this database are available upon request from IPM.
-Anonymized raw data collected from the Moramanga District are shared in the archived repository.
+This database was last queried on September 19, 2019 for these analyses.
+Anonymized raw data from this database are available from IPM following insitutional data transfer protocols.
+Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.
 
@@
 
+Notes from Claire's MSc
+22 ARMC ask for fees (consultation fees to fees for guaze etc for the form. ranging from 2000 Ar - 10000 Ar in some cases, 0.50 - 3,00 USD)
+60% would vaccinate cat I  exposures
+3 ARMC were out of stock 
+282000 Euros a year on 23800 + 9500 vacc vials, 2000 RIG vials.
+
 ## Main findings
@@
 We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.
-However, our model predicts that these effects saturate, and that even given maximum expansion of PEP (i.e.
-to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country and the predictions of maximum reported bite incidence even given low travel times.
-In addition, vial demand slightly outpaces additional deaths averted, with more vials needed per averted death as clinics are added.
-However, after a certain number of clinics are added, rather than large increases in reported incidence and thus vial demand, we estimate shifts in the clinic locations where bites are reported to.
+However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP (i.e.
+to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country.
+After a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
+Vial demand reflects these patterns as well, with more vials needed per death averted as clinics are added, due to the demand for PEP from patients not genuinely exposed to a rabid animal.
 
 ## Limitations
+The travel time estimates from the MAP friction surface underestimate true travel times reported by patients.
+This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018].
+In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
+Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune.
+Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access, or by collecting data on patient reported travel times.
+While patient reported travel times may not be precise, they will likely be more reflective of the range of travel realities that people experience.
+
+The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
+Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems.
+In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access.
+The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times.
+However, we lacked data from other catchments at the commune scale, and the commune model we developed to fit to the district scale data was not able to capture these same patterns from the national data.
+The district model fitted to the national data shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.
+
 We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
@@
 In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.
-
-Importantly,the travel time estimates from the MAP friction surface underestimate true travel times reported by patients.
-This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018].
-In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
-Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune.
-Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access.
-
 We included a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times.
@@
 differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
-In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access.
-The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times, and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times.
-However, we lacked data from other catchments at the commune scale, so it is difficult to say how broadly these patterns apply.
-The district model fitted to the national data therefore likely shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.
+A recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [cite Leblanc 2019].
+Outlier discussion?
 
-Our modeling framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Our decision tree framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden.
@@
 
-We make several simplifying assumptions regarding PEP provisioning and efficacy.
+We also make several simplifying assumptions regarding PEP compliance and efficacy.
 First, we assume perfect compliance of patients and efficacy of PEP.
 Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018].
-We also assume that patients report to their closest clinic, but in our data approximately 10% of patients reported to other clinics.
-As we may not be assigning reported bites to the right clinic, we may potentially be biasing our estimates of vial sharing and throughput at the clinic level.
+We also assume that all ARMC reliably provision PEP, however Leblanc 2019 found that at the time of the KAP survey, three clinics were out of vaccine stock which likely contributes to burden.
+We do not account for differential risk for severely exposed patients (i.e.
+Category III exposures) that do not receive rabies Immunoglobulin (RIG).
+RIG is only available at IPM in Antananarivo, and while exposed persons may be referred to seek RIG, the majority of Category III exposures that seek care are unlikely to have received RIG.
+Unfortunately, we were not able to link patient records between peripheral ARMC and IPM to directly estimate this.
+Recent studies have shown that even in the absence of RIG, PEP is highly effective at preventing rabies deaths, and only adds a marginal benefit in terms of reducing burden but at very high cost.
+Therefore, it is both unlikely that availability of RIG contributes significantly to rabies burden in Madagascar, or that expansion of RIG provisioning would be a feasible option given the costs.
 
-Claire's MsC + Helene's comments here!
-Other ways to designate clinics
-Shortages!
-
-The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
-Improving data submission and quality from clinics across the country will be important to validate these results, but also to monitor how well clinics are functioning and to develop better provisioning systems.
-
-Current practices of provisioning PEP reactively to Category I exposures.
-Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped [@rajeevHealthcareUtilizationProvisioning2018].
-
-Ultimately, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to care, and human behavior.
+Ideally, many of these assumptions could either be validated by or replaced with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
+However, even with these data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict rabies burden.
 We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.
 
-## Box 1: case studies of exceptions to the travel time effects
+## Box 1: case studies of the challenges of predicting health seeking behavior
 
 ### Anosibe A'nala
-In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours.
-Despite this, we see a high incidence of reported exposures.
-We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were close to our touched a person that died from suspected cases.
-An xx% of them were suspected cases.
-And no cases were confirmed.
-The case of AAA points to three key issues: 1) High burden of rabies in this district, we're probably seeing the tip of the iceberg.
-2) people when aware of the risk do travel (in many cases walking 3 days to attend the clinic) and when surveillance efforts target cat I around deaths too.
-Potential for awareness and outreach to go far.
-3) This is also where our travel time estimates fail us.
+In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
+Despite these barriers, we see a high incidence of reported exposures (~~ xx per 100,000 persons).
+We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were in close proximity or had touched a person that had died from suspected rabies.
+Approximately xx% of non Category exposures were classified as a probable rabies case based on clinical criteria.
+No cases were confirmed diagnostically however.
+The data from Anosibe A'nala points to three key insights:
+1) This remote community does suffer from rabies exposure and is likely experiencing a high burden of death that we are only seeing the tip of the iceberg for. 
+2) When people are aware of the risk of death due to rabies, they are willing to travel a long time (in many cases walking 3 days to attend the clinic) to seek the vaccine.
+In addition, surveillance responses, i.e.
+when there is a suspected cases, there is often referral of these people from far away, can work and if these investigations are triggered on exposure rather than death then they could greatly help to prevent deaths.
+There is great potential for community outreach to go far.
+3) The travel times estimated from the friction surface are much lower than those observed.
 Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks.
@@
 boat, on foot, by camion).
+So these metrics may miss key areas that are underserved and warrants improvement.
 
-### Antananarivo (too identifiable?)
-A middle aged taxi driver died of rabies in suburban Antananarivo.
+### Antananarivo 
+A middle aged taxi driver died of rabies in suburban Antananarivo in 2016.
 Bitten by an unknown dog.
@@
 He developed symptoms 2 weeks later and died within a few days.
-The case was confirmed by the NRL.
+The case was confirmed by the Naional Rabies Reference Laboratory at IPM.
 This case points to the other side of the health-seeking behavior coin from AAA.
@@
 But ultimately, it's difficult to predict human behavior.
+Ultimately it will be difficult to acheive 100% coverage of this intervention even in the best of settings.
+Why we don't see people dying in other places is because of the combo of dog vacc + PEP vacc.
 
 ## Broader context
-While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings (1 - 5 subnationally, on average 1.5, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009].
-In particular, the estimate of incidence of deaths for the Moramanga District fell within the range of incidence estimated from a contact tracing study in this district (1.35 {inline} vs.
-1-3 deaths {inline} per 100,000 persons and reporting 72 vs 85 {inline}, [@rajeevHealthcareUtilizationProvisioning2018]).
-At the national level, they are in line with estimates of burden based on a more standard decision tree approach [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-
-<!-- TO DO: WHAT'S THE PT I'M TRYING TO MAKE HERE? -->
-As bites due to other reasons than rabies can be more common than genuine rabies exposures [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], reported bite data may be a poor indicator of rabies transmission, particularly in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
-Surveillance instead...Integrated bite case management and clinic-based surveillance approaches have been established as feasible tools that could accomplish this [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017], and ARMC could act as sentinels for rabies surveillance across the country.
-Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen surveillance.
 
@@
 
+The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa, and largely is in line with previous model estimates of bite incidence based on national statistics (i.e.
+human development index & proportion of the population that is urban, see @@whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
+Ultimately, reported bite incidence is driven by many factors unrelated to rabies including the underlying size of dog populations, barriers to access to PEP (i.e.
+both financial costs and geographic access), and awareness of rabies risk.
+In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are due to rabies.
+The approach we used here is likely less useful in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
+in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
+
+While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies (1 - 7 subnationally, on average 2.7 - 3.1, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Our estimates of rabies burden fell within the higher end of the range of previous estimates of burden for Madagascar [@rajeevHealthcareUtilizationProvisioning2018].
+While we assumed a lower range of incidence of underlying rabies exposure, accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning.
+However, our results covered wide 95% prediction intervals, and improved clinic level data, as well as surveillance data on rabies exposures will be necessary to further refine these absolute estimates.
+
+One approach to obtaining this data is integrated bite case management and clinic-based surveillance, methods that relies on follow-up of bite patients reporting to determine rabies exposure status, recommend treatment, and identify other potential persons exposed [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017; @rajeevHealthcareUtilizationProvisioning2018].
+If these methods were implemented, ARMC could act as sentinels for rabies surveillance across the country, with bite patients reporting to clinics trigger investigations of suspect rabies cases.
+Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen clinic based surveillance efforts.
+
+Our estimates of vial demand depend on countrywide adoption of the new abridged intradermal regimen recommended by WHO [@tarantolaIntradermalRabiesPostexposure2018].
+While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019 only 15% of staff interviewed were aware of this regimen [Leblanc 2019].
+In addition, the majority of staff were also not aware of the WHO classifications of exposure categories, and vaccination of Category I exposures, i.e.
+those not recommended PEP, is a common practice.
+In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018].
+More judicious PEP provisioning could reduce vial demand.
+Current practices of provisioning PEP reactively to Category I exposures.
+Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped.
+
+We predict that as clinics are expanded, overall throughput (i.e.
+the number of patients reporting daily to a clinic) will decrease.
+This may make provisioning PEP more challenging, as vial demand may become less predictable annually, and result in stockouts or vaccine wastage.
+Establishing decentralized provisioning mechanisms, for example through combining provisioning of PEP with routine childhood vaccines, or novel methods for vaccine delivery (i.e.
+drone delivery of emergency medicine, [@boeras2019use]) may be potential ways to mitigate these challenges.
+Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine from IPM upon demand.
+However, once the more expensive Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013].
+This suggests that wider spread vaccine provisioning is feasible if cost barriers are removed.
+
+A Gavi investment can remove this cost barrier.
 In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
-Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons {inline}, compared to Tanzania where catchment sizes are about half of that.
-Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons.
-Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine upon demand.
-However, once the more Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013].
-However, Gavi investment can remove this cost barrier.
-If this results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
-Given existing challenges in data collection and monitoring of how and to whom PEP is provisioned, strong monitoring and evaluation systems will be necessary to ensure these vaccines are being used judiciously and equitably.
+Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons {inline}.
+And while PEP is free, KAP bit?
+In fact, 22 clinics still charge patients (on average ~ 2000 Ar but anywhere from 500 - 10000 Ar for costs other than the vaccine) so this could be a factor too.
+Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons, which would require approximately 200 additional ARMC in Madagascar.
+If GAVI investment results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
+Importantly, our estimates of expansion suggest that it remians cost-effectivehowever these remain reasonably cost-effective.
+Just given vial costs alone, the cost would be between xxx and yyy per patient (similar to GAVI).
+Given a GAVI investment, if the government could shift to improving provisioning systems.
+Cost effectiveness point here (not per DALY but per death averted.)
 
 Ultimately even given expanded access, achieving 'the last mile', i.e.
-preventing deaths due to rabies in the most remote populations will require a disproportionate level of resources and may not be achievable through PEP alone [@delriovilasTribulationsLastMile2017].
+preventing deaths due to rabies in the most remote populations will require a disproportionate level of resources and our results suggest that this will not be achievable through PEP alone [@delriovilasTribulationsLastMile2017].
 IBCM programs and risked based assessments are one way to tackle the issue of managing PEP resources effectively even as demand outpaces reductions in burden [@rysavaPathRabiesElimination2018].
@@
 These same issues of access, however apply to both dog vaccination programs and surveillance systems, and integrating spatial heterogeneities in vaccination and detection will be critical to understand how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015].
-
-<!-- TO DO -->
-citing giorbhail paper!
-What are the key pts!
+Wentworth human right bit...
 
 ## Conclusion {inline these stats}
+Adding clinics alone wont be enough. Need to improve awareness & clinic functioning. 
+Conclusions = You should expand access.
+Focusing on expanding access to a set of clinics that can reliable provision PEP might be the best bet (rather than rolling out to all clinics).
+Developing novel methods of distributing PEP/decentralized way, i.e.
+drone delivery.
+Data is really important at all levels!
+Clinic functioning, health seeking behavior, and also acting as sentinels for rabies surveillance.
+But you won't get rid of deaths completely this way.
+
 Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths.
@@
 
+```{r fig S3.1, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.1_corrs.jpeg"))
+```
+_**Figure S3.1.** Correlation between travel time in hours and population size of administrative unit the district and commune scale._
+
+<br/>
+
 *__Table S3.1.__ DIC Estimates for all models.
@@
 <br/>
-
-```{r fig S3.2, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S3.1_corrs.jpeg"))
-```
-_**Figure S3.1.** Correlation between travel time in hours and population size of administrative unit the district and commune scale._
 
@@
 
+```{r fig S4.1, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S4.1_exprange.jpeg"))
+```
+
+_**Figure S4.1.** Estimated exposures per 100,000 persons using parameters ranges from the literature and data from the Moramanga District.
+The vertical bars show the range of human to dog ratios (HDRs) observed in the literature across African countries.
+The dashed lines show the HDRs estimated from the Moramanga district from a recent study.
+The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Mormanga District.
+For the range of exposure incidence used to estimate burden in this study, we used a minimum consistent with higher hdr and a maximum consistent with an upper hdr and set the midpoint to be the lower end of our estimates from Moramanga so that it makes sense with previous estimates & assumptions.
+However in the se analysis we looked at the upper range of this as well....._
+
 ## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
@@
 ```{r fig S5.1, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.1.jpeg"))
+include_graphics(here("figs/supplementary/S5.1_ARMCadded.jpeg"))
 ```
-_**Figure S5.1.** Maps of where and at what step (i.e.
-commune/district where the Nth clinic was added) at the **A)** district and **B)** commune level.
-X's indicate the location of existing ARMC (N = 31), polygons in white are locations where an additional clinic was not added based on the algorithm criteria (reducing travel times for > 0.01 % of people living > 3 hours away from any clinic)._
+_**Figure S5.1.** Map of the location and at what step each clinic was added.
+The circles are each of the CSB II across the country sized by the proportion of the population for which they reduced travel times (excluding those already living within 3 hours of an ARMC) and the colors showing at what step each clinic was added (i.e.
+1 = the first clinic added).
+The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the remaining CSB II in the country (which reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked)._
 
 ```{r fig S5.2, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.2.tiff"))
+include_graphics(here("figs/supplementary/S5.2_map_shifts.tiff"))
 ```
 _**Figure S5.2.** Maps of how travel times change as clinics are added at the **A)** ~1 x 1 km grid cell **B)** commune and **C)** district scales.
-The columns are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country).
+The columns are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 723 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country).
 Grey pixels show the location of ARMC at each step.
@@
 ```{r fig S5.3, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.3.jpeg"))
-```
-_**Figure S5.3.** As clinics are added, shifts in distribution of the proportion of population in a given catchment at the commune and district scales (colors).
-Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
-
-```{r fig S5.4, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.4.jpeg"))
+include_graphics(here("figs/supplementary/S5.3_prop_ttimes.jpeg"))
 ```
@@
 
+```{r fig S5.4, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.4_prop_catch.jpeg"))
+```
+_**Figure S5.3.** As clinics are added, shifts in distribution of the proportion of population in a given catchment  for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data).
+Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
+
 ```{r fig S5.5, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.5.jpeg"))
+include_graphics(here("figs/supplementary/S5.5_ttimes_shifts.jpeg"))
 ```
-_**Figure S5.5.** As clinics are added, declines in commune and district averages of **A)** travel times (hrs), **B)** bite incidence per 100,000 persons, **C)** reporting of rabid exposures, and **D)** incidence of deaths per 100,000 persons._
+_**Figure S5.5.** Shifts in estimated **A)** travel times (hrs), **B)** bite incidence per 100,000 persons.
+The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading) for communes and districts  for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
 
 ```{r fig S5.6, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.6.jpeg"))
+include_graphics(here("figs/supplementary/S5.6 _reporting_shifts.jpeg"))
 ```
-_**Figure S5.5.** As clinics are added, shifts in the distribution of **A)** travel times (hrs), **B)** bite incidence per 100,000 persons, **C)** reporting of rabid exposures, and **D)** incidence of deaths per 100,000 persons at the commune and district scales (colors).
-Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
+_**Figure S5.6.** Shifts in estimated **A)** Reporting, **B)** death incidence per 100,000 persons.
+The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading) for communes and districts for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
 
 ```{r fig S5.7, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.7.jpeg"))
+include_graphics(here("figs/supplementary/S5.7_clinic_shifts.jpeg"))
 ```
-_**Figure S5.7.** As clinics are added, declines in clinic average **A)** catchment population size, **B)** daily throughput, and **C)** annual vial demand given estimates of bite incidence for the two models (commune and district, colors)._
+_**Figure S5.7.** Shifts in clinic average **A)** catchment population size, **B)** and annual vial demand, and **C)**  daily throughput given estimates of bite incidence for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data).
+The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading).
+Catchment population sizes are the same for each model._
+
+```{r fig S5.8, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.8_comm_map.jpeg"))
+```
+_**Figure S5.8.** Where bites are reported to as clinics are added for the Commune scale model fitted to the Moramanga data.
+The color of the points show at what step the clinic was added (with the grey points in the last panel the CSB II in the country that reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked), the size shows the average number of bites reported to that clinic annually, and the lines show from which commune bites are reported from.
+The shading of the polygons show the reported bite incidence at the commune level._
+
+```{r fig S5.9, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.9_dist_map.jpeg"))
+```
+_**Figure S5.9.** Where bites are reported to as clinics are added for the District scale model fitted to the National data.
+The color of the points show at what step the clinic was added (with the grey points in the last panel the CSB II in the country that reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked) and the lines show from which district bites are reported from.
+The shading of the polygons show the reported bite incidence at the district level._
+
+```{r fig S5.10, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.10_prop_in_admin.jpeg"))
+```
+_**Figure S5.9.** Shifts in the distribution of the proportion of bites reported to a clinic that came from within the same A) commune or B) district as the ARMC was located for the two models (Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
 
@@
 ```{r fig S6.1, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S6.1.jpeg"))
+include_graphics(here("figs/supplementary/S6.1_base_se.jpeg"))
 ```
-_**Figure S6.1.**Range of constrained scaling factors generate for district and commune population size.
-The black dashed lines represent the minimum and maximum estimated rabies exposure incidence (minimum = 11.1 exposures/100k persons, maximum = 55.7 exposures/100k persons).
-We generate constrained scaling factors with human population size so that rabies exposure incidence in each location scales positively or negatively across the range of observed population sizes (panels)._
+_**Figure S6.1.** ._
 
 ```{r fig S6.2, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S6.2.jpeg"))
+include_graphics(here("figs/supplementary/S6.2_addARMC_se.jpeg"))
 ```
-*__Figure S6.2.__Predicted relationship between travel times and rabies death incidence per 100k persons for the district model vs.
+*__Figure S6.2.__ Predicted relationship between travel times and rabies death incidence per 100k persons for the district model vs.
 commune models (colors) with columns arranged by the scaling factor (minimum incidence of 11.1 rabies exposures/100,000 persons, the maximum negative scaling with population size, the maximum positive scaling with population size, and maximum incidence of 55.7 rabies exposures/100,000 persons, see Fig S6.1 for the scaling factors).
@@
 ```{r fig S6.3, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S6.3.jpeg"))
+include_graphics(here("figs/supplementary/S6.3_vial_se.jpeg"))
 ```
@@
 
-## Supplementary references
-::: {#refs_supp}
+```{r fig S6.4, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S6.4_scaling_rels.jpeg"))
+```
+*__Figure S6.4.__ Range of constrained scaling factors generated for district and commune population size.
+The black dashed lines represent the minimum and maximum estimated rabies exposure incidence (minimum = 11.1 exposures/100k persons, maximum = 55.7 exposures/100k persons).
+We generate constrained scaling factors with human population size so that rabies exposure incidence in each location scales positively or negatively across the range of observed population sizes (panels)*
+
+```{r fig S6.5, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S6.5_scaling_se.jpeg"))
+```
+*__Figure S6.5.__ *
+
+## Supplementary Reference (R packages used) {#srefs}
+
+::: {#refs_R}
 :::

### 2020-04-19 R/06_sensitivity/03_scaling_se.R

@@ -23,8 +23,6 @@ incidence_max <- 110/1e5
 incidence_min <- 15/1e5
 pos_scale <- seq(incidence_min, incidence_max, length.out = length(pop))
 neg_scale <- seq(incidence_max, incidence_min, length.out = length(pop))
-# pos <- lm(pos_scale ~ pop) # use these and constrain
-# neg <- lm(neg_scale ~ pop) # use these and constrain
 pos <- lm(pos_scale ~ 0 + pop, offset=rep(10/1e5, length(pop)))
 neg <- lm(neg_scale ~ 0 + pop, offset=rep(110/1e5, length(pop)))
 neg_comm <- data.table(scaling = "neg", sfactor = neg$coefficients[1], intercept = 110/1e5,
@@ -37,8 +35,6 @@ pop <- mada_districts$pop - min(mada_districts$pop)
 pop <- pop[order(pop, decreasing = FALSE)]
 pos_scale <- seq(incidence_min, incidence_max, length.out = length(pop))
 neg_scale <- seq(incidence_max, incidence_min, length.out = length(pop))
-# pos <- lm(pos_scale ~ pop) ## use these and constrain
-# neg <- lm(neg_scale ~ pop) ## use these and constrain
 pos <- lm(pos_scale ~ 0 + pop, offset=rep(10/1e5, length(pop)))
 neg <- lm(neg_scale ~ 0 + pop, offset=rep(110/1e5, length(pop)))
 neg_dist <- data.table(scaling = "neg", sfactor = neg$coefficients[1], intercept = 110/1e5, 
@@ -70,15 +66,9 @@ model_sds <- bind_rows(filter(model_sds_unadj, data_source == "National", scale 
                                                              data_source == "Moramanga"))
 
 # Burden predictions ---------------------------------------------------------------------
+
 # Filter to only do ones for which travel times have changed
-commune_master <- fread("output/ttimes/commune_maxcatch.gz")
-setorder(commune_master, commcode, scenario)
-commune_master[, diff_comm := ttimes_wtd - shift(ttimes_wtd, 1), by = commcode]
-commune_master[, diff_dist := ttimes_wtd_dist - shift(ttimes_wtd_dist, 1), by = commcode]
-comm_run <- commune_master[diff_comm != 0 | scenario %in% c(0, 1648)]
-dist_run <- commune_master[diff_dist != 0 | scenario %in% c(0, 1648)]
-rm(commune_master) # cleaning up memory!
[…]

### 2020-04-19 R/figures/SM1_ttimes.R

@@ -242,4 +242,4 @@ S1.5_outputs <- (ttimes_district | ttimes_commune) + plot_layout(guides = "colle
 ggsave("figs/supplementary/S1.5_outputs.jpeg", S1.5_outputs, device = "jpeg", height = 10, width = 12)
 
 #' Saving session info
-out.session(path = "R/figures/SM1_basetimes.R", filename = "sessionInfo.csv")
+out.session(path = "R/figures/SM1_basetimes.R", filename = "output/log_local.csv")

### 2020-04-19 R/figures/SM2_bitedata.R

@@ -175,5 +175,5 @@ figS2.2_vialcomps <-
 ggsave("figs/supplementary/S2.2_vialcomps.jpeg", figS2.2_vialcomps, device = "jpeg", height = 8, width = 10)
 
 #' Saving session info
-out.session(path = "R/figures/SM2_bitedata.R", filename = "sessionInfo.csv")
+out.session(path = "R/figures/SM2_bitedata.R", filename = "output/log_local.csv")
 

### 2020-05-12 docs/manuscript.Rmd
fixing how clinics are added
- + various figure fixes
- + updates to man & bib
- + functions to ttimes & fixed out session so that it gives you the max memory used in Mbs

@@
       correspondence: false
+  - Anjasoa Randrianarijaona:
+      institute: [IPMepi]
+      correspondence: false
+  - Reziky Tiandraza Mangahasimbola:
+      institute: [IPMepi]
+      correspondence: false
+  - Fleur Hierink:
+      institute: [Geneva]
+      correspondence: false
+  - Ravo Ramiandrasoa:
+      institute: [IPMvacc]
+      correspondence: false
+  - Jean-Michel Heraud:
+      institute: [IPMviro]
+      correspondence: false
   - Soa Fy Andriamandimby:
@@
       correspondence: false
-  - Anjasoa Randrianarijaona:
-      institute: [IPMepi]
-      correspondence: false
-  - Reziky Tiandraza Mangahasimbola:
-      institute: [IPMepi]
-      correspondence: false
-  - Ravo Ramiandrasoa:
-      institute: [IPMvacc]
-      correspondence: false
   - Laurence Baril:
-      institute: [IC]
+      institute: [IPCambodia]
       correspondence: false
@@
   - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
+  - Geneva: University of Geneva
 date: ""
@@
 
-opts_chunk$set(echo = FALSE, warning = FALSE)
+opts_chunk$set(echo = FALSE, warning = FALSE, include = FALSE)
 
@@
 Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP.
+
+Go in order.
+
 We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country.
 We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden.
+
 We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic).
 Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.
+
+However our estimates are dependent on blah blah blah.
+
 These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.
@@
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
-the more bites reported in a location, the higher the incidence of *rabies* exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018].
+the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018].
 If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.
@@
 To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).
-Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale.
-We then extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell.
+
+Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest ARMC (i.e.
+a clinic provisioning PEP) for the country of Madagascar at an approximately 1 x 1 km scale.
+We then extracted the mean of travel times at the district and commune (the administrative unit below the district) level, as well as the mean of travel times weighted by the population in that grid cell.
 We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC.
@@
 We used two datasets on bite patients reporting to ARMC:
+
 1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
-These were individual patient data forms that were submitted from clinics as frequently as monthly to annually that included data on the date of reporting and were resolved to the district level (i.e.
-the district where the patient resided)
-2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level (the administrative level below the district).
+These were individual patient data forms that were submitted from clinics as frequently as monthly to annually, which included data on the date of reporting and were resolved to the district level (i.e.
+the district where the patient resided).
+2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level.
 We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.
 
-For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
+The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.
+However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
 To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1).
-We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period.
-We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.
-
+We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic.
+Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.
+
+We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period.
 Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported.
@@
 During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit.
-Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared + any additional wasteage from unused doses discarded at the end of the day.
-While these simplifying assumptions do not capture variation in adherence, completion, delays, or errors in administration, PEP completion is overall high in Madagascar and in other settings where PEP is provided at no direct costs to patients, so our estimates at 3 doses per patient vs.
-4 are likely encompassing much of this variation at the annual level.
-
-We then divided the total number of bites reported to each clinic from each clinic in a given year by the undersubmission rate for eachclinic.
-Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.
-The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.
+Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day.
+We did not have the date or month when vials were provisioned, so we compared the totals across this four year period.
 
@@
 $$ y_i = Poisson(\mu_i) $$
-where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the predicted number of bites) for a given source location (district or commune).
-We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’.
+where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to expected number of bites) for a given source location (district or commune).
+We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level.
 In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:
 
-$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$
+$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0_j)}P_j$$
 
@@
 
-For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercept).
-We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by comparing the ratio of the sum of squared residuals to the expected $\chi^2_{n - k}$ distribution given a poisson error structure [@gelman2006data].
-
-We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations).
-We used priors centered around zero (*norm(0, 10-6)*) except for the precision parameter for which we used an uniform distribution (*unif(0, 100)*).\
-
+For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma_0)$) where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept).
 The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.
-We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).
-
-To test our models predictions further, we used parameter estimates from each model to generate predictions for data that were not used to fit the model (i.e.
+
+We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
+
+$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j + \sigma_e)}P_i$$
+where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters and for a given model.
+
+We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’.
+We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10-6)$).
+For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$).
+We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model [@gelman2006data].
+
+To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to the data.
+In addition, we used the models to predict the data that were not used to fit the model (i.e.
 we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data).
-Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to the raw data uncorrected for submission (i.e.
+Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e.
 assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs.
@@
 ## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}
-
-We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2).
-As with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], to model uncertainty in parameter estimates we used triangular distributions for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures.
+We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1).
+To model uncertainty in parameter estimates we used triangular distributions, as with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures.
 The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).
 
-Specifically: annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination .
-To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$).
-We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons.
-As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] .
-$E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.
+For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons).
+The range is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons.
+These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019].
+We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate).
+As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [CITE!!].
+See figure S4 for further details.
 
@@
 |:----------------|:----------------:|:----------------------------------|:---------:|
-| $E_i$ |	Triangular(a = 15.6 b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see S4|
-| $p_{rabid}$	| Triangular(a = 0.2 b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
+| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see Fig S4.1|
+| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
 | $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
@@
 ## candidate points
-csbs <- read.csv(here("data/raw/csbs.csv"), stringsAsFactors = FALSE)
-csbs %>% 
-  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
-  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
+csbs <- read.csv(here("data/processed/clinics/csb2.csv"), stringsAsFactors = FALSE)
+clin_per_comm <- read.csv(here("data/processed/clinics/clinic_per_comm.csv"), 
+                          stringsAsFactors = FALSE)
 ```
@@
 In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons).
-We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning.
-Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5).
 Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.
+
+We also explore three additional scenarios: 
+1) Each district has one clinic (the CSB II in the highest population density grid cell compared to other potential clinics) provision PEP (N = 114 clinics). 
+2) Each commune has one clinic (the CSB II, or when there are no CSB II in a commune the health post or dispensary, in the highest density grid cell compared to other potential clinics) provision PEP (N = `r nrow(clinic_per_comm)`.
+Note that not all communes (N = `r length(mada_communes)`) have a clinic).
+3) All CSB II in the country provision PEP (N = `r nrow(csbs)` clinics).
 
@@
 Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level.
-As clinics are added, multiple clinics may serve a single district or commune (Fig TO CHECK S5.3B).
-Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest (Fig CHECK maps).
+As clinics are added, multiple clinics may serve a single district or commune (S5.4).
+Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest.
 For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
 all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses.
-We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e.
-each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]).
-We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e.
-wastage due to injection error).
+We use these dates to estimate vial usage given routine vial sharing practices in Madagascar as described previously, but under the assumption of adoption of the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]).
 We take the mean of 1000 simulations of throughput for each catchment and set of clinics.
 
-## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
-To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and models of burden.
-Additionally, we also examine the systematic variation in human rabies exposure incidence with human population (a potential proxy for changes in the dog population, see section S6) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.
-We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1).
+## Sensitivity analysis {#methods-sensitivity}
+To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and the decision tree parameters.
+For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1).
+We did the same for parameter estimates from the models of reported bite incidence, but rather fixed these estimates at the upper and lower 95% quantiles of their posterior distributions (Table S6.2).
+We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden.
 Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e.
@@
 if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.
+We use scaling factors which scale incidence either positively or negatively with the population sizes observed at the district and commune levels, but are still constrained within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4).
 
@@
 All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}.
-All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo.
-The raw bite patient data at the national level is maintained by IPM.
+All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at this github repository.
+The raw bite patient data at the national level are maintained in a database by IPM.
 This database was last queried on September 19, 2019 for these analyses.
@@
 We found that travel times estimated from the friction surface generally underestimate true driving times and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2).
-The friction surface assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence (i.e.
-if there is a road, then people will travel at vehicle rather than walking speeds).
+The friction surface assumes that the fastest available mode of transport is used to travel each grid cell (i.e.
+the presence of a road in a given grid cell indicates that all travel through that grid cell will be at vehicle rather than walking speeds).
 However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3).
@@
 ```
-*__Figure 3.A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each ARMC for which we have data.
+*__Figure 3.A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
 Color corresponds to the clinic catchment.
-Circles with white outline are the total number of bites reported for that district.
+Circles with a white outline are the total number of bites reported for that administrative unit (plotted as the centroid).
 Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded.
-Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
 Out-of-catchment reporting is indicated where points and line colors are mismatched.
-__B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the area plotted.
-The black outline shows the district of Moramanga, and the red polygons shows all the communes included in the Moramanga ARMC catchment.
-Circle with black outline is the location of the Moramanga ARMC with the size of points proportional to the total number of bites reported to the clinic, as per A.*
+For panel __A)__ districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
+For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), and all communes included in it's catchment (red polygons) and other communes where bites were reported to in grey.*
 
@@
 For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
-While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher reported incidence at low travel times at the commune level compared to the district.
+While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district fell within the ranges observed from the national dataset.
 
 ## Modeling reported bite incidence as a function of travel times {#results-bitemods}
-All models and posterior estimates of parameters converged ({inline stat for max psrf & mpsrf}).
-Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1).
-However, when used to predict data that were not used to fit the models (i.e.
-model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop) generated reasonable predictions to the data (Figure S3.3).\
-
+```{r}
+convergence <- read.csv(here("output/stats/convergence.csv"))
+```
+
+All models and posterior estimates of parameters converged (Gelman-Rubin diagnostic < `r round(max(convergence$val[convergence$OD != TRUE]), 3)`, @gelman2006data).
 All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2).
 Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
-For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2).
-These predictions were robust to how we corrected for incomplete submission of data by clinics (Fig S3.5).
-The models without the random effect were significantly overdispersed, but the catchment random effect still captured additional variation beyond this overdispersion (Fig S3.4).
-The parameter estimates from the different models were similar, with the model fit to the Moramanga data showing the highest impact of travel times and the district model the lowest (Fig 5B-D), however the increased bite incidence observed at low travel times in the Moramanga data was not captured by the commune model (Fig 5A).
-To capture the range of estimates of travel time impact, we compared the district model fitted to the national data with the commune scale model fitted to the Moramanga data in subsequent analyses.
+
+For the national data, including a random catchment effect improved predictions of the observed data (Fig S3.2 & Fig S3.3), and catchment level effects were strongly identified in the posterior estimates.
+However, after accounting for overdispersion of the data, these catchment effects were not clearly identifiable and the posterior estimates of the catchment intercepts were broadly similar (Fig S3.6), and the models resulted in similar predictions to the data (Fig S3.7), indicating that the catchment effect could not be differentiated from random variation in the data.
+Similarly, while the commune model fit to the Moramanga data generated stronger estimates of the travel time effect (Fig 5B), after accounting for overdispersion of the data, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
+Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models (Fig S3.8) and generated similar predictions of the range of bite incidence across travel times (Fig S3.9).
+
+As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates from the commune and district models fit to the national data, as they encompass the range of travel time effects observed in our datasets.
 
@@
 A) The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
-The lines are the estimates using the posterior estimates of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and adjusted for overdispersion.
-The envelopes shows the 95% prediction interval.
-The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model).
-B - D) The posterior distribution of parameters from the respective models for the B) model intercept, C) travel time effect, and D) the catchment random effect (for national data only)._
+The lines are the mean estimates and the envelopes are the 95% prediction intervals for predictions generated by drawing 1000 independent samples from the parameter posterior distributions for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with an overdispersion parameter ($\sigma_e$) and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and unadjusted for overdispersion.
+The envelopes show the 95% prediction interval.
+The points show the data used to fit the models: National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model.
+The posterior distribution of parameters from the respective models for the B) model intercept, C) travel time effect, and D) the overdisperson parameter (for national data only)._
 
@@
 ```{r burden results}
-natl_burden <- read.csv(here("output/preds/natl_burden.csv"))
-```
-
-Overall, we estimate between `r round(sort(natl_burden$deaths_mean)[1])` - `r round(sort(natl_burden$deaths_mean)[nrow(natl_burden)])` average annual deaths in Madagascar due to rabies depending on the model used (Table 2).
-We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_burden$averted_mean)[1])` - `r round(sort(natl_burden$averted_mean)[nrow(natl_burden)])` deaths due to rabies are prevented through provisioning of PEP each year.
+natl_preds <- fread(here("output/preds/burden_natl.gz"))
+natl_base <- natl_preds[scenario == 0]
+```
+
+Overall, we estimate between `r round(sort(natl_base$deaths_mean)[1])` - `r round(sort(natl_base$deaths_mean)[nrow(natl_base)])` average annual deaths in Madagascar due to rabies depending on the model used (Table 2).
+We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_base$averted_mean)[1])` - `r round(sort(natl_base$averted_mean)[nrow(natl_base)])` deaths due to rabies are prevented through provisioning of PEP each year.
 In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally when burden is modeled at the district level, with even stronger patterns of increase in burden with travel times when modeling burden at the commune level with estimates from the Moramanga data (Figure 4).
@@
 
-*__Table 2.__ Model predictions of average annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% prediction interval in parentheses.*
+*__Table 2.__ Model predictions of average annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% prediction interval of national estimate in parentheses.*
 
 ```{r Table 2}
-natl_burden %>%
-  mutate_at(vars(starts_with("bites")), function(x, p) x/p*1e5, p = .$pop) %>%
+pop <- sum(mada_districts$pop)
+natl_base %>%
+  mutate_at(vars(starts_with("bites")), function(x) x/pop*1e5) %>%
   mutate_at(vars(bites_mean:averted_lower), round) %>%
-  mutate(scale = case_when(scale == "Commune" ~ "Commune (Moramanga data)",
-                           scale == "District" ~ "District (National data)"), 
-         bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
+  mutate(bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
          deaths = glue("{deaths_mean} ({deaths_lower} - {deaths_upper})"),
@@
 *__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
-__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model fit to the national data and the squares show the incidence for the commune model fit to the Moramanga data for all communes in a district.
+__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model fit to the National data for all communes in a district.
 Points are colored by and districts are ordered by travel times.
 The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
-Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current CTAR.*
+Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current ARMC*
 
@@
 ```{r}
-natl_preds_all <- read.csv(here("output/stats/natl_preds_all.csv"))
-natl_preds_all %>%
-  group_by(scale) %>%
-  arrange(scenario) %>%
-  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
+max_added <- max(as.numeric(natl_preds$scenario[!(natl_preds$scenario %in% c("max", "armc_per_dist", "armc_per_comm"))]))
+natl_preds %>%
+  mutate(scenario_num = case_when(!(scenario %in% c("max", "armc_per_dist", "armc_per_comm")) ~ as.numeric(scenario),
+                                  scenario == "max" ~ max_added + 200, scenario == "armc_per_dist" ~ 114, 
+                                  scenario == "armc_per_comm" ~ max_added + 150)) %>%
+  group_by(scale, data_source) %>%
+  arrange(scenario_num) %>%
+  mutate_at(vars(-c(group_cols(), scenario, scenario_num)), function(x) x/x[1]) -> props
 trends <- read.csv(here("output/stats/trend_preds.csv"))
@@
 
-We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 250 clinics reducing burden by almost `r round(min(1 - props$deaths_mean[props$scenario == 250])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 250])*100, 2)`% (Fig M7.A).
-However, benefits accrue less rapidly as ARMC are added: with approximately 700 clinics, we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == 700])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 700])*100, 2)`, while even when all CSB IIs in the country provision PEP, the most deaths are reduced is between `r round(min(1 - props$deaths_mean[props$scenario == max(props$scenario)])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == max(props$scenario)])*100, 2)`%.
-This saturation is a result of the landscape of clinics across the country, as most CSBs are in areas with already low travel times (Fig S5.X).
-Therefore, as clinics are added, reported bite incidence saturates (Fig S5.Z), and instead there are shifts in which clinics bites are reported to (Fig S5.Y).
-Even at this maximal PEP access, our model still predicts `r round(min(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures still remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.N).
-
-Vial demand also increases, and more vials are needed per averted death as PEP access is expanded.
+We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 250 clinics reducing burden by `r round(min(1 - props$deaths_mean[props$scenario == 250])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 250])*100, 2)`% (Fig 7A).
+However, benefits accrue less rapidly as ARMC are added: with approximately 700 clinics, we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == 700])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 700])*100, 2)`.
+This saturation is a result of the landscape of clinics across the country, as additional clinics reduce clinics for fewer people as more clinics are added (Fig S5.1).
+Even at maximal PEP access, our model still predicts `r round(min(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.6).
+
+Vial demand also increases (Fig 7B), and more vials are needed per averted death as PEP access is expanded (Fig 7C).
 Our model predictes an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
-visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` {inline} vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
-Most bites become reported within a commune--why does throughput decrease!
-
-Similar to our baseline estimates of burden, there was significant uncertainty in these estimates, in particular for the district model fit to the national data.
-
-<!-- TO DO: State what happens to throughput + vial sensitivity! -->
+visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
+Importantly, clinic throughput decreases, as well, with clinics seeing fewer patients per day (S5.7), as reported bite incidence saturates (Fig S5.5), and instead there are shifts in which clinics bites are reported to.
+
+We also look at two additional scenarios: one CSB II provisions PEP for each district (ARMC per district, N = 114 clinics) and one where one clinic (either a CSB II, CSB I, or dispensary) provisions PEP to each commune (ARMC per commune, N = `r nrow(clin_per_comm)` as all communes have one clinic).
+These expansion scenarios result in similar estimates of impacts on reducing human rabies deaths and increasing vial demand compared to adding the same number of clinics through our stepwise approach.
 
@@
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The “...” on the axes indicates a break in the axes, and the last point is the burden and vial predictions when all CSBS in the country provision PEP (maximum access).*
+The points show three additional scenarios: 1) one ARMC per district (N = 114 clinics total, + 83 clinics), 2) one ARMC per commune where atleast one is present (N = `r nrow(clin_per_comm)` total, + `r nrow(clin_per_comm) - 31` clinics), and 3) all CSB II in the country provision PEP (N = `r nrow(csbs) + 31` clinics total, + `r nrow(csbs)`)*
 
 ## Sensitivity of estimates to bite incidence and decision tree model parameters {#results-sensitivity}
-
-Overall, our sensitivity analyses show that while there is significant uncertainty in the absolute numbers of deaths due to rabies, patterns with increasing risk of death with reduced access and the potential for expanding access to PEP to reduce these deaths appear to be robust (Figs S6.2, S6.3).
-In general, for a given incidence level the higher our assumption of the proportion of reported bites which are rabid, the lower the estimated burden.
-The lower our estimate of maximum reporting, the higher overall burden of rabies, and as this parameter decreases, the relationship between travel times and rabies deaths are dampened.
-Scaling with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
-more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths.
-Finally, raw numbers of deaths greatly depend on what we assume the rabies exposure incidence is in the population.
-
-To quantify which parameters contribute to the uncertainty observed in our estimates, we performed a univariate sensitivity analyses, varying parameters in both the model of bite incidence (taking the lower and upper 95% confidence intervals of the posterior estimates, Table 1) and the decision tree model (fixing parameters at the ranges used in the parameter distributions, Table 2).
-These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden (Fig S6.Y).
-For the impacts of expanded access, what we assume is the maximum reporting of rabies exposures given minimal travel times ($rho_max$) is an important factor for the commune model (Fig S6.X), because as clinics are added, more locations are shifted to this threshold (Fig S5.X).
-For the district model, $p_{rabid}$ and $E_i$ are also important in determining the magnitude of reduction of burden.
-Despite the large absolute ranges of parameter estimates, the broad patterns of higher risk with increased travel times are robust for both our estimates of burden and the impacts of expanding access to PEP, even given systematic bias in our assumptions of incidence (Fig S6.N).
-
+To quantify which parameters contribute to the uncertainty observed in our estimates, we performed a univariate sensitivity analyses, varying parameters in both the model of bite incidence (Table S6.1) and the decision tree model (Table S6.2).
+These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden and for determining the impact of expanding access (Fig S6.2).
 Uncertainty in bite model parameters contribute less to variation of estimates of burden.
 For the estimates of vial demand, uncertainty around the intercept of the model (i.e.
-the baseline reported bite incidence) has the greatest impact, rather than the travel time effect or the catchment effect (Fig S6.4).
-
-```{r se}
-se_pars <- read.csv(here("output/sensitivity/se_pars.csv"))
-
-se_pars %>%
-  pivot_longer(exp_min:beta_0) %>%
-  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
-                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
-  group_by(scale, par = name) %>%
-  summarize(upper = max(value), lower = min(value)) %>%
-  mutate(range = glue("{lower} \U2012 {upper}")) -> se_table_all
-
-se_table_all %>%
-  filter(grepl("beta|sigma", par)) %>%
-  pivot_wider(par, values_from = range, names_from = scale) %>%
-  as.data.table(.) -> bites_se
-se_table_all %>%
-  filter(!grepl("beta|sigma|min", par), scale != "Commune") %>%
-  as.data.table(.) -> dtree_se
-```
-
-Table:**Table 3.** Range of parameters used in models of bite incidence for the univariate sensitivity analyses.\
-
-| **Parameter**   | **Description**        | **Commune model (Moramanga data)** | **District model (National data)** | 
-|:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
-| $\beta_t$ | Travel time effect	| `r bites_se$Commune[par == "beta_t"]` | `r bites_se$Commune[par == "beta_t"]`
-| $\beta_0$ | Model intercept	| `r bites_se$Commune[par == "beta_0"]` | `r bites_se$Commune[par == "beta_0"]`
-| $\sigma_0$ | Standard deviation of intercept by catchment	| -- | `r bites_se$Commune[par == "sigma_0"]`
-
-Table:**Table 4.** Range of parameters used in decision tree model for the univariate sensitivity analyses.\
-
-| **Parameter**   | **Description**        | **Range**	                 | **Notes**|
-|:----------------|:----------------------:|:----------------------------|:---------:|
-| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se$range[par == "exp_max"]`  	
-| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se$range[par == "p_rab_max"]` 	
-| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se$range[par == "rho_max"]`	 
-| $p_{death}$ | The probability of death given a rabies exposure^[1] | `r dtree_se$range[par == "p_death"]`	
-[1] 95% CI of estimate from @changaluchaNeedImproveAccess2018|
+the baseline reported bite incidence) has the greatest impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
+Finally, scaling of incidence with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths (Fig S6.5A).
+However, the impact of adding clinics remains broadly the same (Fig S6.5B).
+Overall, while uncertainty in the underlying rabies exposure incidence results in significant variation in the absolute number of rabies burden, our results of the current impact of travel times on burden and of the potential for expanded access to reduce burden are qualitatively similar across parameter estimates.
 
@@
 ```{r}
-natl_preds_all %>%
+natl_preds %>%
   filter(scenario == 0) %>%
@@
 ```
-
-Notes from Claire's MSc
-22 ARMC ask for fees (consultation fees to fees for guaze etc for the form. ranging from 2000 Ar - 10000 Ar in some cases, 0.50 - 3,00 USD)
-60% would vaccinate cat I  exposures
-3 ARMC were out of stock 
-282000 Euros a year on 23800 + 9500 vacc vials, 2000 RIG vials.
 
@@
 Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to greatly reduce deaths due to rabies, with expansion to an additional 200 {inline} clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.
-However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP (i.e.
-to all major health centers (n = 1676) in the country), between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally, due to the distribution of clinic locations across the country.
-After a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
-Vial demand reflects these patterns as well, with more vials needed per death averted as clinics are added, due to the demand for PEP from patients not genuinely exposed to a rabid animal.
+We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to an additional 250 clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.
+
+However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all major health centers (n = 1676) in the country, between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally.
+
+This result is driven by three key factors.
+First, given the baseline levels of reported bite incidence and our assumptions of underlying rabies exposure, we estimate that burden in areas with low travel times could still potentially be high {inline estimate of reporting/incidence}.
+Second, given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}.
+Finally, the travel time effect, while consistently non-zero and negative, does not account much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {inline overdispersion of sigma_e}.
+
+We find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
+Vial demand reflects these patterns as well, however, vials needed per death averted match predictions from a recent cost effectiveness analysis of PEP expansion, suggesting that expanding PEP access will remain highly cost effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 
 ## Limitations
-The travel time estimates from the MAP friction surface underestimate true travel times reported by patients.
-This likely has to do with the quality of data on transport speeds in Madagascar, which is taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018].
+The travel time estimates from the MAP friction surface underestimate true travel times reported by patients, likely due to discrepancies between assigned transport speeds and modes and the reality of travel in Madagascar.
+These speeds are taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018], and also assumes that given multiple types of transport in a cell, the fastest available mode of transport is used.
 In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
@@
 Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access, or by collecting data on patient reported travel times.
-While patient reported travel times may not be precise, they will likely be more reflective of the range of travel realities that people experience.
-
-The national data we used were underreported and incomplete, but we attempt to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
-Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems.
-In addition, we found that the scale of data used to fit these models is likely an important factor in estimating the impact of spatial access.
-The commune level data from the Moramanga District showed a steeper decline in reported bite incidence with travel times and higher estimates of bite incidence at lower travel times, resulting in lower burden estimates overall, but a greater concentration of burden in areas with high travel times.
-However, we lacked data from other catchments at the commune scale, and the commune model we developed to fit to the district scale data was not able to capture these same patterns from the national data.
-The district model fitted to the national data shows the more conservative estimate of the impact of current and expanded PEP access on human rabies deaths.
+While patient reported travel times may also lack precision due to recall and estimation error, they will likely be more reflective of the range of travel realities that people experience.
 
@@
 In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.
-We included a catchment level random intercept to account for spatial differences in reported bite incidence not due to travel times.
-This random component of the model improved fit to the data, suggesting that variation at the clinic level (i.e.
-differences in clinic functioning such as vaccine availability) and at the catchment level (i.e.
+In fact, a recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [@leblancclaireRabiesMadagascarThreepronged2019].
+OUTLIERS TO TIE INTO OD & other factors
+
+We included an overdispersion parameter in our model to quantify variation not due to travel times, and for all models, the data were significantly overdispersed, suggesting that variation at the clinic level (i.e.
+differences in clinic functioning such as vaccine availability) and at the administrative level (i.e.
 differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
-A recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [cite Leblanc 2019].
-Outlier discussion?
+While clinic specific effects were identifiable in models without overdispersion, these random effects were not identifiable when accounting for random variance in the data.
+Importantly, the national data we used were underreported and incomplete, but we attempted to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
+Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems.
+
+SCALE SCALE We were able to use multiple sources of data to compare estimates of bite incidence across scales.
+We found that we could effectively disaggregate burden to the commune scale using our travel time metrics.
+The Moramanga data demonstrates that sub-district patterns of access are important to account for when estimating potential areas with low access.
+However, we lacked data from other catchments at the commune scale to further confirm and compare finer scale patterns.
 
 Our decision tree framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden.
+As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden at the national level.
 Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same.
@@
 We also make several simplifying assumptions regarding PEP compliance and efficacy.
-First, we assume perfect compliance of patients and efficacy of PEP.
+We assume perfect compliance of patients and efficacy of PEP.
 Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018].
-We also assume that all ARMC reliably provision PEP, however Leblanc 2019 found that at the time of the KAP survey, three clinics were out of vaccine stock which likely contributes to burden.
 We do not account for differential risk for severely exposed patients (i.e.
@@
 RIG is only available at IPM in Antananarivo, and while exposed persons may be referred to seek RIG, the majority of Category III exposures that seek care are unlikely to have received RIG.
-Unfortunately, we were not able to link patient records between peripheral ARMC and IPM to directly estimate this.
-Recent studies have shown that even in the absence of RIG, PEP is highly effective at preventing rabies deaths, and only adds a marginal benefit in terms of reducing burden but at very high cost.
-Therefore, it is both unlikely that availability of RIG contributes significantly to rabies burden in Madagascar, or that expansion of RIG provisioning would be a feasible option given the costs.
-
-Ideally, many of these assumptions could either be validated by or replaced with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
-However, even with these data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict rabies burden.
+We were not able to link patient records between peripheral ARMC and IPM to directly estimate this.
+However, recent studies have shown that even in the absence of RIG, PEP is highly effective at preventing rabies deaths, and only adds a marginal benefit in terms of reducing burden but at very high cost.
+It is both unlikely that availability of RIG contributes significantly to rabies burden in Madagascar, or that expansion of RIG provisioning would be a feasible option given the costs.
+
+We also assume that all ARMC reliably provision PEP, but this is unlikely to be the case, as Leblanc 2019 found that at the time of their KAP survey, three clinics were experiencing vaccine stock-outs, which likely contributes to burden.
+PART ABT THEM CHARGING PATIENTS FOR OTHER FEES HERE (UNCLEAR WHETHER THEY ARE IN THE DATA).
+22 ARMC ask for fees (consultation fees to fees for guaze etc for the form.
+ranging from 2000 Ar - 10000 Ar in some cases, 0.50 - 3,00 USD).
+And while PEP is free, KAP bit?
+In fact, 22 clinics still charge patients (on average ~ 2000 Ar but anywhere from 500 - 10000 Ar for costs other than the vaccine) so this could be a factor too.
+PRovisioning and clinic functioning will be as important as expanding to everywhere!
+
+Ideally, many of these assumptions could either be confirmed or replaced with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
+Even with these additional data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict the impact of PEP spatially.
 We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.
@@
 ## Box 1: case studies of the challenges of predicting health seeking behavior
+```{r}
+mora_bites %>%
+  filter(district == "Anosibe-An'ala") %>%
+  summarize(pop = sum(pop, na.rm = TRUE), bites = sum(avg_bites, na.rm = TRUE), 
+            inc = bites/pop*1e5) -> aa_bites
+mora_raw <- read.csv(here("data/processed/bitedata/moramanga.csv"))
+mora_raw %>%
+  filter(distcode == "MG33317") %>%
+  summarize(prop_cat1 = sum(known_cat1)/n()) -> prop_cat1_AA
+admin_preds <- fread(here("output/preds/burden_base.gz")) 
+admin_preds %>%
+  mutate(distcode = substr(names, 1, 7)) %>%
+  filter(distcode == "MG33317") %>%
+  group_by(scale) %>%
+  summarize(inc = sum(deaths_mean)/sum(pop)*1e5, bite_inc = sum(bites_mean)/sum(pop)*1e5) -> aa_preds
+```
 
 ### Anosibe A'nala
-In the district of Anosibe A'nala (population of ~ xx people) south of the Moramanga District, travel times to the nearest ARMC (the Moramanga ARMC) are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
-Despite these barriers, we see a high incidence of reported exposures (~~ xx per 100,000 persons).
-We interviewed xx% of these patients when they reported to the clinic, and about xx% {inline} of persons interviewed were loosely defined Category I exposures, members who were in close proximity or had touched a person that had died from suspected rabies.
-Approximately xx% of non Category exposures were classified as a probable rabies case based on clinical criteria.
-No cases were confirmed diagnostically however.
+In the district of Anosibe A'nala (population of ~ `r round(mada_districts$pop[mada_districts$district = "Anosibe-An'ala"])` people) south of the Moramanga District, travel times to the nearest ARMC are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
+Despite these barriers, we see a moderate incidence of reported bites (~ `r round(aa_bites$inc)` per 100,000 persons).
+Importantly `r trans_perc(prop_cat1_AA$prop_cat1)`% of these patients were loosely defined Category I exposures, i.e.
+people who had been in close proximity or had touched a person that had died from suspected rabies.
+Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 5 deaths in this district over a `r max(mora_bites$month)` month period.
+Only one animal case was confirmed diagnostically however.
 The data from Anosibe A'nala points to three key insights:
+
 1) This remote community does suffer from rabies exposure and is likely experiencing a high burden of death that we are only seeing the tip of the iceberg for.
+
 2) When people are aware of the risk of death due to rabies, they are willing to travel a long time (in many cases walking 3 days to attend the clinic) to seek the vaccine.
@@
 There is great potential for community outreach to go far.
-3) The travel times estimated from the friction surface are much lower than those observed.
+
+3) What do we estimate as the death incidence?
+When would we add a clinic?
+The travel times estimated from the friction surface are much lower than those observed.
 Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks.
@@
 ### Antananarivo 
-A middle aged taxi driver died of rabies in suburban Antananarivo in 2016.
-Bitten by an unknown dog.
-Promptly reported to a clinic where he was reffered to IPM.
-His daughter, a prospective medical student, and other family members urged him to seek the vaccine but he did not in the end seek care.
+In 2016, a taxi driver in his 60s died of rabies in suburban Antananarivo.
+He was bitten by an unknown dog returning from work one evening, and the next day he promptly reported to a clinic where he was reffered to the ARMC at IPM.
+His daughter, a prospective medical student, and other family members urged him to seek the vaccine, but he did not believe his risk was high and so in the end did not seek PEP.
 He developed symptoms 2 weeks later and died within a few days.
 The case was confirmed by the Naional Rabies Reference Laboratory at IPM.
-This case points to the other side of the health-seeking behavior coin from AAA.
-This man had the resources, was aware of the risk, and was reffered by the hospital (and sought care immediately).
-But ultimately, it's difficult to predict human behavior.
-Ultimately it will be difficult to acheive 100% coverage of this intervention even in the best of settings.
-Why we don't see people dying in other places is because of the combo of dog vacc + PEP vacc.
+This case points to the other side of the health-seeking behavior coin from Anosibe A'nala.
+Even given prompt reporting and referral, and socioeconomic indicators that would suggest a high probability of seeking PEP, this person did not seek PEP.
+However, the case does point to the fact that community education on rabies risk, as well as availability of PEP at peripheral clinics even in areas with already high access could remove further barriers to PEP and prevent additional deaths.
 
@@
 
-The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa, and largely is in line with previous model estimates of bite incidence based on national statistics (i.e.
-human development index & proportion of the population that is urban, see @@whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
+The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa, and largely is in line with previous model estimates of bite incidence based on national statistics.
+USE SPECIFIC ESTIMATES HERE (i.e.
+human development index & proportion of the population that is urban, see @whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
 Ultimately, reported bite incidence is driven by many factors unrelated to rabies including the underlying size of dog populations, barriers to access to PEP (i.e.
 both financial costs and geographic access), and awareness of rabies risk.
-In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are due to rabies.
-The approach we used here is likely less useful in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
+In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are genuine rabies exposures.
+As we did not have location specific estimates of this parameter, we used a range estimated from a previous study in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018], however this approach is likely to be less applicable in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
 in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
 
-While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies (1 - 7 subnationally, on average 2.7 - 3.1, see [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
-Our estimates of rabies burden fell within the higher end of the range of previous estimates of burden for Madagascar [@rajeevHealthcareUtilizationProvisioning2018].
-While we assumed a lower range of incidence of underlying rabies exposure, accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning.
-However, our results covered wide 95% prediction intervals, and improved clinic level data, as well as surveillance data on rabies exposures will be necessary to further refine these absolute estimates.
-
-One approach to obtaining this data is integrated bite case management and clinic-based surveillance, methods that relies on follow-up of bite patients reporting to determine rabies exposure status, recommend treatment, and identify other potential persons exposed [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017; @rajeevHealthcareUtilizationProvisioning2018].
-If these methods were implemented, ARMC could act as sentinels for rabies surveillance across the country, with bite patients reporting to clinics trigger investigations of suspect rabies cases.
+While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies STATE RANGES HERE this range estimates in our study vs this range estimated in these studies [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
+We estimate a higher burden of rabies deaths here STATE that previously for Madagascar STATE [@rajeevHealthcareUtilizationProvisioning2018].
+While in this study we assumed a lower range of incidence of underlying rabies exposure, previously we assumed a fixed reporting of 85% of rabies exposures across the country.
+Accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning.
+Improved clinic level data, as well as surveillance data on rabies exposures will be necessary to further refine these estimates, particularly given sensitivity of these estimates to assumptions of rabies exposure incidence.
+
+One approach to obtaining this data is integrated bite case management which relies on follow-up of bite patients to determine rabies exposure status, recommend treatment, and identify other potential persons exposed [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017; @rajeevHealthcareUtilizationProvisioning2018].
+If these methods were implemented, ARMC could act as sentinels for rabies surveillance across the country, with bite patients reporting to clinics triggering investigations of suspect rabies cases.
 Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen clinic based surveillance efforts.
@@
 Our estimates of vial demand depend on countrywide adoption of the new abridged intradermal regimen recommended by WHO [@tarantolaIntradermalRabiesPostexposure2018].
-While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019 only 15% of staff interviewed were aware of this regimen [Leblanc 2019].
+While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019 only 15% of staff interviewed were aware of this regimen [@leblancclaireRabiesMadagascarThreepronged2019].
 In addition, the majority of staff were also not aware of the WHO classifications of exposure categories, and vaccination of Category I exposures, i.e.
 those not recommended PEP, is a common practice.
-In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018].
-More judicious PEP provisioning could reduce vial demand.
-Current practices of provisioning PEP reactively to Category I exposures.
-Based on data from the Moramanga District, vial demand could be up to 20% lower if these practices were stopped.
+In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018], and more judicious PEP provisioning could reduce vial demand, particularly if expansion of PEP results in increases in reported bite incidence beyond what we predict.
 
@@
 the number of patients reporting daily to a clinic) will decrease.
-This may make provisioning PEP more challenging, as vial demand may become less predictable annually, and result in stockouts or vaccine wastage.
+This may make provisioning PEP more challenging, as vial demand may become less predictable annually, resulting in stockouts or vaccine wastage.
 Establishing decentralized provisioning mechanisms, for example through combining provisioning of PEP with routine childhood vaccines, or novel methods for vaccine delivery (i.e.
 drone delivery of emergency medicine, [@boeras2019use]) may be potential ways to mitigate these challenges.
+IMPROVED vacc technology!
 Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine from IPM upon demand.
 However, once the more expensive Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013].
-This suggests that wider spread vaccine provisioning is feasible if cost barriers are removed.
+This suggests that wider spread vaccine provisioning is feasible given health infrastructure in Madagascar if cost barriers are removed.
 
@@
 In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
-Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons {inline}.
-And while PEP is free, KAP bit?
-In fact, 22 clinics still charge patients (on average ~ 2000 Ar but anywhere from 500 - 10000 Ar for costs other than the vaccine) so this could be a factor too.
+Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons.
 Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons, which would require approximately 200 additional ARMC in Madagascar.
 If GAVI investment results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
-Importantly, our estimates of expansion suggest that it remians cost-effectivehowever these remain reasonably cost-effective.
-Just given vial costs alone, the cost would be between xxx and yyy per patient (similar to GAVI).
-Given a GAVI investment, if the government could shift to improving provisioning systems.
-Cost effectiveness point here (not per DALY but per death averted.)
+Importantly, given our estimates of vial demand and deaths averted and accounting for direct vaccine costs only, PEP expansion remains highly cost-effective (< 500 USD per death averted given average vaccine vial costs, @whorabiesmodelingconsortiumPotentialImpactImproved2019).
 
@@
 These same issues of access, however apply to both dog vaccination programs and surveillance systems, and integrating spatial heterogeneities in vaccination and detection will be critical to understand how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015].
-Wentworth human right bit...
 
 ## Conclusion {inline these stats}
-Adding clinics alone wont be enough. Need to improve awareness & clinic functioning. 
-Conclusions = You should expand access.
-Focusing on expanding access to a set of clinics that can reliable provision PEP might be the best bet (rather than rolling out to all clinics).
-Developing novel methods of distributing PEP/decentralized way, i.e.
-drone delivery.
-Data is really important at all levels!
-Clinic functioning, health seeking behavior, and also acting as sentinels for rabies surveillance.
-But you won't get rid of deaths completely this way.
-
 Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths.
-Based on our modeling framework, a more equitable target may be aiming to minimize the population living within 3-5 hours travel time of a clinic, but due to broad patterns of remoteness across Madagascar, even when all primary health facilities are providing PEP, approximately 10 % {inline} of people live > 3 hours away from a clinic.
-Expanding PEP to clinics that target currently underserved populations may be the best strategy for preventing human rabies deaths.
-If expanded only to one clinic per district (N = 114), an additional 40% of deaths could be averted.
-Expansion of PEP can greatly reduce risk and inequities in access to this life-saving vaccine, but ultimately, interrupting transmission of rabies through mass dog vaccination programs will be the most equitable way to prevent, and eventually eliminate, rabies deaths.
+In the absence of data on rabies incidence and risk of exposure, targeting expansion of PEP at areas currently underserved may be the best way to reduce rabies burden and provide equitable access to PEP.
+Specifically, choosing clinics serving populations that under the status quo would have to travel greater than 3 hours to seek PEP, or alternatively aiming for clinics to serve a catchment size of approximately 100,000 persons, are two ways that PEP access could be expanded strategically.
+However, our results also suggest that increasing outreach and awareness of rabies risk, improving clinic functioning and provisioning practices, and removing socioeconomic barriers for persons seeking PEP will be critical to ensure the efficacy of PEP expansion.
+Improved surveillance of animal rabies will also be needed to better understand the landscape of rabies burden across countries and to identify populations at the greatest risk for rabies deaths.
+Expansion of PEP can greatly reduce risk and inequities in access to this life-saving vaccine, and a GAVI investment would allow countries to more equitably provide PEP to populations that currently face immense barriers to seeking this intervention ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies).
+In addition, as PEP alone cannot prevent all rabies deaths, it could facilitate investment in mass dog vaccination to interrupt transmission, and eventually eliminate deaths due to tnis devastating disease.
 
@@
 __B)__ Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the  criteria of 15 consecutive days of non-submission/365.
-__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013-2016 for each clinic correcting for undersubmission (squares) using the 15 day cut-off show in A and B, vs.
+__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013 -  2017 for each clinic correcting for undersubmission (squares) using the 15 day cut-off show in A and B, vs.
 not correcting for undersubmission (circles).
@@
 ```
-_**Figure S3.1.** Correlation between travel time in hours and population size of administrative unit the district and commune scale._
+_**Figure S3.1.** Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale._
 
@@
 For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
-```{r table 1}
+```{r table S3.1}
 model_ests <- read.csv(here("output/mods/estimates.csv"))
 model_ests %>%
+  filter(OD == FALSE) %>%
   select(params, Mean, pop_predict, intercept, data_source, scale) %>%
@@
 _**Figure S3.2.
-Prediction to data used to fit each model.** Log of the observed bites against the log of predicted bites using the mean of the posterior distributions for each parameter.
+Prediction to data used to fit each model.** Log of the observed bites against the log of predicted bites generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
 Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
@@
 _**Figure S3.3.
-Out of fit predictions to data.** Log of the observed bites against the log of predicted bites generated from the mean of the posterior distributions from each model for data not used to fit the model.
+Out of fit predictions to data.** Log of the observed bites against the log of predicted bites for data not used to fit the model.
+Predictions were generated by sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
 The first two columns are the predictions from the commune and district model fitted to the national data for the Moramanga data with fixed and random intercepts.
@@
 ```{r fig S3.4, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S3.4_ODcheck.jpeg"))
-```
-_**Figure S3.4.** Overdispersion....
+include_graphics(here("figs/supplementary/S3.4_posts_all.jpeg"))
+```
+_**Figure S3.4.** Posterior estimates of parameters from models with travel time and population as an offset (flatPop in figures S3.2 & S3.3), comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter.
+For the Moramanga model, as data came from a single catchment, models with a random catchment effect ($\sigma_0$) were not fitted._
 
 ```{r fig S3.5, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S3.5_cutoffs.jpeg"))
-```
-_**Figure S3.5.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with a random catchment intercept and with population as an offset.
-Panels show predictions from models fitted to (1) data corrected for both undersubmission by correcting for periods of at least 7 days where zero patient forms where submiited, (2) correcting for periods of 15 days where zero patient forms where submiited, and (3) with the raw data not correcting for undersubmission.
-The lines are the estimates using the mean of the posterior distributions of the parameters and the envelopes shows the 95% CI given the random intercept.
-The colors are the two candidate models with travel times at the commune and district level fitted to the national data with a random intercept by catchment.
+include_graphics(here("figs/supplementary/S3.5_preds_OD.jpeg"))
+```
+_**Figure S3.5.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
+Rows are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and columns are whether the model estimated an overdispersion parameter.
+The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset._\
+
+```{r fig S3.6, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.6_posts_intercepts.jpeg"))
+```
+_**Figure S3.6.** Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter._\
+
+```{r fig S3.7, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.7_preds_ODcomp.jpeg"))
+```
+_**Figure S3.7.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons for random intercept model without overdispersion vs.
+fixed intercept model with overdispersion, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
+The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset._\
+
+```{r fig S3.8, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.8_posts_rep_se.jpeg"))
+```
+_**Figure S3.8.** Posterior estimates for models with population as an offset and an overdispersion parameter.
+Columns show estimates from models fitted to the national dataset (1) corrected for both undersubmission by correcting for periods of at least 7 days where zero patient forms where submiited, (2) correcting for periods of 15 days where zero patient forms where submiited, and (3) with the raw data not correcting for undersubmission.
+._\
+
+```{r fig S3.9, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S3.9_preds_rep_se.jpeg"))
+```
+
+_**Figure S3.5.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with population as an offset and an overdispersion parameter.
+Panels show predictions from models fitted to the national dataset 1) correcting for periods of at least 7 days where zero patient forms where submiited (a less stringent cutoff resulting in lower estimates of the proportion of forms submitted and thus higher estimates of reported bite incidence), (2) correcting for periods of 15 days where zero patient forms where submitted, as presented in the main analysis, and (3) with the raw data not correcting for undersubmission (resulting in lower estimates of reported bite incidence).
+Predictions were generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
 The points show the data used to fit the models._
@@
 
-_**Figure S4.1.** Estimated exposures per 100,000 persons using parameters ranges from the literature and data from the Moramanga District.
-The vertical bars show the range of human to dog ratios (HDRs) observed in the literature across African countries.
-The dashed lines show the HDRs estimated from the Moramanga district from a recent study.
-The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Mormanga District.
-For the range of exposure incidence used to estimate burden in this study, we used a minimum consistent with higher hdr and a maximum consistent with an upper hdr and set the midpoint to be the lower end of our estimates from Moramanga so that it makes sense with previous estimates & assumptions.
-However in the se analysis we looked at the upper range of this as well....._
+_**Figure S4.1.** Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis), assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
+The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (the red horizontal line).
+The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@leblancclaireRabiesMadagascarThreepronged2019].
+The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2018]._
 
@@
 
-<!-- TO DO: add catchmaps -->
-
-```{r fig S5.1, fig.fullwidth=TRUE}
+```{r fig S5.1 add clinics, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.1_ARMCadded.jpeg"))
@@
 
-```{r fig S5.2, fig.fullwidth=TRUE}
+```{r fig S5.2 ttime map shifts, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.2_map_shifts.tiff"))
@@
 
-```{r fig S5.3, fig.fullwidth=TRUE}
+```{r fig S5.3 prop pop ttimes, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.3_prop_ttimes.jpeg"))
@@
 
-```{r fig S5.4, fig.fullwidth=TRUE}
+```{r fig S5.4 prop pop catchment, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.4_prop_catch.jpeg"))
@@
 
-```{r fig S5.5, fig.fullwidth=TRUE}
+```{r fig S5.5 travel times(a) bite inc(b), fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.5_ttimes_shifts.jpeg"))
@@
 
-```{r fig S5.6, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.6 _reporting_shifts.jpeg"))
+```{r fig S5.6 reporting(a) death inc(b), fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.6_reporting_shifts.jpeg"))
 ```
@@
 
-```{r fig S5.7, fig.fullwidth=TRUE}
+```{r fig S5.7 catch(a) vials(b) throughput(c), fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S5.7_clinic_shifts.jpeg"))
@@
 
-```{r fig S5.8, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.8_comm_map.jpeg"))
-```
-_**Figure S5.8.** Where bites are reported to as clinics are added for the Commune scale model fitted to the Moramanga data.
-The color of the points show at what step the clinic was added (with the grey points in the last panel the CSB II in the country that reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked), the size shows the average number of bites reported to that clinic annually, and the lines show from which commune bites are reported from.
-The shading of the polygons show the reported bite incidence at the commune level._
-
-```{r fig S5.9, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.9_dist_map.jpeg"))
-```
-_**Figure S5.9.** Where bites are reported to as clinics are added for the District scale model fitted to the National data.
-The color of the points show at what step the clinic was added (with the grey points in the last panel the CSB II in the country that reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked) and the lines show from which district bites are reported from.
-The shading of the polygons show the reported bite incidence at the district level._
-
-```{r fig S5.10, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.10_prop_in_admin.jpeg"))
-```
-_**Figure S5.9.** Shifts in the distribution of the proportion of bites reported to a clinic that came from within the same A) commune or B) district as the ARMC was located for the two models (Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
+```{r fig S5.8 prop same admin, fig.fullwidth=TRUE}
+include_graphics(here("figs/supplementary/S5.8_prop_in_admin.jpeg"))
+```
+_**Figure S5.8.** Shifts in the distribution of the proportion of bites reported to a clinic that came from within the same A) commune or B) district as the ARMC was located for the two models (Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
 
@@
 
-```{r fig S6.1, fig.fullwidth=TRUE}
+```{r se tables}
+se_pars <- read.csv(here("output/sensitivity/se_pars.csv"))
+
+se_pars %>%
+  pivot_longer(exp_min:beta_0) %>%
+  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
+                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
+  group_by(scale, par = name) %>%
+  summarize(upper = max(value), lower = min(value)) %>%
+  mutate(range = glue("{lower} \U2012 {upper}")) -> se_table_all
+
+se_table_all %>%
+  filter(grepl("beta|sigma", par)) %>%
+  pivot_wider(par, values_from = range, names_from = scale) %>%
+  as.data.table(.) -> bites_se
+se_table_all %>%
+  filter(!grepl("beta|sigma|min", par), scale != "Commune") %>%
+  as.data.table(.) -> dtree_se
+```
+
+Table:**Table S6.1.** Range of parameters used in models of bite incidence for the univariate sensitivity analyses.\
+
+| **Parameter**   | **Description**        | **Commune model (Moramanga data)** | **District model (National data)** | 
+|:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
+| $\beta_t$ | Travel time effect	| `r bites_se$Commune[par == "beta_t"]` | `r bites_se$Commune[par == "beta_t"]`
+| $\beta_0$ | Model intercept	| `r bites_se$Commune[par == "beta_0"]` | `r bites_se$Commune[par == "beta_0"]`
+| $\sigma_0$ | Standard deviation of intercept by catchment	| -- | `r bites_se$Commune[par == "sigma_0"]`
+
+Table:**Table S6.2.** Range of parameters used in decision tree model for the univariate sensitivity analyses.\
+
+| **Parameter**   | **Description**        | **Range**	                 | **Notes**|
+|:----------------|:----------------------:|:----------------------------|:---------:|
+| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se$range[par == "exp_max"]`  	
+| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se$range[par == "p_rab_max"]` 	
+| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se$range[par == "rho_max"]`	 
+| $p_{death}$ | The probability of death given a rabies exposure^[1] | `r dtree_se$range[par == "p_death"]`	
+[1] 95% CI of estimate from @changaluchaNeedImproveAccess2018|
+
+```{r fig S6.1 base se, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S6.1_base_se.jpeg"))
 ```
-_**Figure S6.1.** ._
-
-```{r fig S6.2, fig.fullwidth=TRUE}
+_**Figure S6.1.** A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B)  estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter._
+
+```{r fig S6.2 add se, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S6.2_addARMC_se.jpeg"))
 ```
-*__Figure S6.2.__ Predicted relationship between travel times and rabies death incidence per 100k persons for the district model vs.
-commune models (colors) with columns arranged by the scaling factor (minimum incidence of 11.1 rabies exposures/100,000 persons, the maximum negative scaling with population size, the maximum positive scaling with population size, and maximum incidence of 55.7 rabies exposures/100,000 persons, see Fig S6.1 for the scaling factors).
-Rows show burden estimates given the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]).
-Colors represent the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]); shapes show the maximum and minimum assumption of $\rho_{max}$ = [0.8 , 0.95]).*
-
-```{r fig S6.3, fig.fullwidth=TRUE}
+_**Figure S6.2.** A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter._
+
+```{r fig S6.3 vial se, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S6.3_vial_se.jpeg"))
 ```
-*__Figure S6.3.__ Estimates of the impact of expanding access to PEP incrementally on  burden at the national level for the district model vs.
-commune models (colors) with columns arranged by the scaling factor (minimum incidence of 11.1 rabies exposures/100,000 persons, the maximum negative scaling with population size, the maximum positive scaling with population size, and maximum incidence of 55.7 rabies exposures/100,000 persons--see Fig S6.1 for the scaling factors).
-Rows show burden estimates given the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]).
-Colors represent the maximum and minimum assumption of the proportion of reported bites that are assumed to be rabid ($p_{rabid}$ = [0.2, 0.6]); shapes show the maximum and minimum assumption of $\rho_{max}$ = [0.8 , 0.95].*
-
-```{r fig S6.4, fig.fullwidth=TRUE}
+_**Figure S6.3.** A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter._
+
+```{r fig S6.4 scaling factors, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S6.4_scaling_rels.jpeg"))
 ```
-*__Figure S6.4.__ Range of constrained scaling factors generated for district and commune population size.
-The black dashed lines represent the minimum and maximum estimated rabies exposure incidence (minimum = 11.1 exposures/100k persons, maximum = 55.7 exposures/100k persons).
-We generate constrained scaling factors with human population size so that rabies exposure incidence in each location scales positively or negatively across the range of observed population sizes (panels)*
-
-```{r fig S6.5, fig.fullwidth=TRUE}
+*__Figure S6.4.__ Range of constrained scaling factors generated for district and commune population size so that underlying rabies exposures either A) decreases with increasing population size or B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
+Lines show the expected relationship, with points showing where administrative units fall along this curve, and maps show how this results in variation in assumed exposure incidence spatially at the commune and district level.*
+
+```{r fig S6.5 scaling se, fig.fullwidth=TRUE}
 include_graphics(here("figs/supplementary/S6.5_scaling_se.jpeg"))
 ```
-*__Figure S6.5.__ *
+*__Figure S6.5.__ Predicted relationship between A) travel times and rabies death incidence per 100k persons and B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
+commune models (colors) with columns comparing the baseline (presented in main analyses) and assumptions of rabies exposure incidence increasing or decreasing with human population size (shapes).
+The points show the mean estimates from 1000 simulations and the lineranges show the 95% prediction intervals.*
 

### 2020-05-12 R/01_gis/06_process_clinic_data.R

@@ -18,42 +18,53 @@ mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
 ctar_metadata <- read.csv("data/processed/clinics/ctar_metadata.csv")
 csbs <- read.csv("data/raw/csbs.csv", stringsAsFactors = FALSE)
 
-# filter out ctar from csb IIs
-csbs %>% 
-  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
-  select(lat = ycoor, long = xcoor) -> csb2
-
-dist_mat <- distm(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
-                  cbind(csb2$long, csb2$lat))
-csb2 <- csb2[-apply(dist_mat, 1, which.min), ]
-csb2$clinic_id <- 1:nrow(csb2) + 31 # id for clinics
-write.csv(csb2, "data/processed/clinics/csb2.csv", row.names = FALSE)
-
 # get clinic commcodes & distcodes
-csb2_coords <- SpatialPoints(cbind(csb2$long, csb2$lat), 
-                     proj4string = 
-                       CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
-csb2$commcode <- over(csb2_coords, mada_communes)$commcode
-csb2$distcode <- over(csb2_coords, mada_districts)$distcode
+csbs_coords <- SpatialPoints(cbind(csbs$xcoor, csbs$ycoor), 
+                             proj4string = 
+                               CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+csbs$commcode <- over(csbs_coords, mada_communes)$commcode
+csbs$commune <- over(csbs_coords, mada_communes)$commune
+csbs$distcode <- over(csbs_coords, mada_districts)$distcode
+csbs$district <- over(csbs_coords, mada_communes)$district
 
+# Process exisiting ctar data -------------------------------------------------------------
 ctar_coords <- SpatialPoints(cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE), 
                              proj4string = 
                                CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
-ctar_metadata$commcode <- over(ctar_coords, mada_communes)$commcode
[…]

### 2020-05-12 R/04_addclinics/01_ttimes_candidates.R

@@ -1,9 +1,7 @@
 # ------------------------------------------------------------------------------------------------ #
 #' Getting travel times for each grid cell for each clinic
-#' Code should be run in parallel and final dataset requires a lot of memory! ~ 10 GB
-#' Split up clinics or work with fewer clinics if there are memory limitations
-#'   with three cores, it takes approximately 120 minutes on MacOS with 16 GB 1867 MHz DDR3 and
-#'   2.9 GHz Intel Core i5
+#' Code should be run in parallel and data is split to make easier to work with/store
+#' Adjust the chunksize option based on memory limitations
 # ------------------------------------------------------------------------------------------------ #
 
 #sub_cmd=-t 12 -n 31 -sp "./R/04_addclinics/01_ttimes_candidates.R" -jn candidates -wt 2m -n@
@@ -24,7 +22,6 @@ library(gdistance)
 library(iterators)
 library(data.table)
 library(glue)
-library(fst)
 
 # Source scripts
 source("R/functions/out.session.R")
@@ -37,43 +34,41 @@ ctar_metadata <- read.csv("data/processed/clinics/ctar_metadata.csv")
 friction_masked <- raster("data/processed/rasters/friction_mada_masked.tif")
 
 # candidate points
-# filter out existing csbs and also write out to file!
-csbs <- read.csv("data/processed/clinics/csb2.csv")
+csbs <- fread("data/processed/clinics/csb2.csv")
+clin_per_comm <- fread("data/processed/clinics/clinic_per_comm.csv")
+clin_per_comm <- clin_per_comm[!(clinic_id %in% csbs$clinic_id) & clinic_id > 31][, pop_dens := NULL]
+csbs <- rbind(csbs, clin_per_comm)
+setorder(csbs, clinic_id) # make sure they're in the right order
 point_mat_candidates <- as.matrix(select(csbs, long, lat))
 rownames(point_mat_candidates) <- csbs$clinic_id
-baseline_df <- fread("output/ttimes/baseline_grid.gz")
-out_dir <- "/scratch/gpfs/mrajeev/output/ttimes/candidates/"
+
+out_dir <- "output/ttimes/candidates/"
 
 if(!dir.exists(out_dir)) dir.create(out_dir)
 
-# Depends on # of files you want--lets aim for 16
-# Alternatively do the chunks in parallel and you get 30 files (which one is faster/has manageable file sizes?)
[…]

### 2020-05-12 R/04_addclinics/02_ttimes_added.R

@@ -7,13 +7,12 @@
 #'   it takes approximately 10 hours
 # ------------------------------------------------------------------------------------------------ #
 
-#sub_cmd=-sn -t 12 -n 18 -mem 4500 -sp "./R/04_addclinics/02_ttimes_added.R" -jn addclinics -wt 5m -n@
+#sub_cmd=-sn -t 12 -n 15 -sp "./R/04_addclinics/02_ttimes_added.R" -jn addclinics -wt 5m -n@
   
 # set up cluster on single node with do Parallel
-library(doParallel) 
-cl <- makeCluster(18)
+library(doParallel)
+cl <- makeCluster(15)
 registerDoParallel(cl)
-getDoParWorkers()
 start <- Sys.time()
 
 # Libraries
@@ -21,22 +20,25 @@ library(foreach)
 library(tidyverse)
 library(iterators)
 library(data.table)
+library(raster)
 
 # Source
 source("R/functions/out.session.R")
 source("R/functions/ttime_functions.R")
 
-# Pull in candidates
-cand_mat <- fread("/scratch/gpfs/mrajeev/output/ttimes/candidate_matrix.gz") # this is a huge file!
+# baseline rasters 
+base_df <- fread("output/ttimes/base_df.gz")
+
+# Pull in candidates & figure out which files they're in
 csb2 <- fread("data/processed/clinics/csb2.csv")
+brick_dt <- get.bricks(brick_dir = "output/ttimes/candidates")
+csb2 <- brick_dt[csb2, on = "clinic_id"]
+csb2[, band := (clinic_id - min + 1)]
 
-## Baseline df
-base_df <- fread("output/ttimes/baseline_grid.gz")
-
-## Do the candidates
-system.time ({
-  add.armc(base_df = base_df, clinic_names = csb2$clinic_id, clinic_catchmat = cand_mat, 
-           max_clinics = ncol(cand_mat), thresh_ttimes = 3*60, thresh_prop = 1e-4, 
+# Do the candidates
+system.time({
+  add.armc(base_df = base_df, cand_df = csb2, max_clinics = nrow(csb2), 
+           thresh_ttimes = 3*60, thresh_prop = 1e-4, 
            dir_name = "/scratch/gpfs/mrajeev/output/ttimes/addclinics_")
 })
 
@@ -47,5 +49,5 @@ syncfrom <- "mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/output/ttimes/add
 # Close out
[…]

### 2020-05-12 R/figures/M7_incremental.R

@@ -21,7 +21,7 @@ max_added <- max(as.numeric(natl_preds_seq$scenario))
 max_total <- max_added + 200
 natl_preds_sep %>%
   mutate(scenario_num = case_when(scenario == "max" ~ max_added + 200, 
-                              scenario == "armc_per_dist" ~ 114,
+                              scenario == "armc_per_dist" ~ 83,
                               scenario == "armc_per_comm" ~ max_added + 150),
          scenario = fct_relevel(scenario, "armc_per_dist", "armc_per_comm", "max")) -> natl_preds_sep
 

### 2020-05-12 R/figures/SM3_bitemods.R

@@ -230,7 +230,8 @@ all_samps_se %>%
   mutate(Parameter = fct_recode(Parameter, `beta[t]` = "beta_ttimes", `beta[0]` = "beta_0",
                                 `sigma[e]` = "sigma_e"), 
          rep_cutoff = gsub("[^0-9]", "", all_samps_se$filenames), 
-         rep_cutoff = ifelse(rep_cutoff == "", Inf, rep_cutoff)) -> all_samps_se
+         rep_cutoff = fct_relevel(fct_recode(rep_cutoff, `Uncorrected` = "", `15 days` = "15",
+                                 `7 days` = "7"), "7 days", "15 days", "Uncorrected")) -> all_samps_se
 
 ggplot(data = filter(all_samps_se, !(grepl("alpha", Parameter))), aes(x = scale, y = value, 
            fill = scale)) +
@@ -239,8 +240,7 @@ ggplot(data = filter(all_samps_se, !(grepl("alpha", Parameter))), aes(x = scale,
   scale_x_discrete(labels = NULL) +
   labs(x = "", y = "Posterior estimates") +
   facet_grid(Parameter ~ rep_cutoff, scales = "free_y", 
-             labeller = labeller(rep_cutoff = cutoff_labs,
-                                 Parameter = label_parsed)) +
+             labeller = labeller(Parameter = label_parsed)) +
   theme_minimal_hgrid() +
   theme(panel.background = element_rect(color = "NA", size = 1.2, fill = "grey92"),
         panel.grid = element_line(color = "white", size = 0.5)) -> posts_rep_se

### 2020-05-12 R/figures/SM5_shifts.R

@@ -3,7 +3,7 @@
 #' Pulling in district and commune estimates of travel times as clinics are added 
 # ------------------------------------------------------------------------------------------------ #
 
-start <- start()
+start <- Sys.time()
 
 # Libraries and packages
 library(data.table)
@@ -83,6 +83,14 @@ point_mat_all %>%
                               !is.na(prop_pop) ~ prop_pop),
          step_added = case_when(is.na(step_added) & clinic_id < 32 ~ 0,
                                 !is.na(step_added) ~ as.numeric(step_added))) -> point_mat_all
+pts <- SpatialPoints(cbind(point_mat_all$long, point_mat_all$lat), 
+                     proj4string = 
+                       CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+point_mat_all$commcode_ctar <- over(pts, mada_communes)$commcode
+point_mat_all$commname_ctar <- over(pts, mada_communes)$commune
+point_mat_all$distcode_ctar <- over(pts, mada_communes)$distcode
+point_mat_all$distcode_ctar <- over(pts, mada_communes)$district
+write.csv(point_mat_all, "output/stats/clinics_added.csv")
 
 gg_commune <- fortify(mada_communes, region = "commcode")
 

### 2020-05-12 R/figures/SM6_burden_se.R

@@ -249,7 +249,8 @@ vial_se_B <- ggplot(data = filter(vials_se, name == "vials_mean",
                                   "armc_per_dist" = "1 per district\n (+ 83)",
                                   "armc_per_comm" = "1 per commune\n (+ 1375)"), 
                        name = "Additional\n scenarios") + 
-  facet_grid(vary ~ scale, scales = "free_x") +
+  facet_grid(vary ~ scale, scales = "free_x", 
+             labeller = labeller(scale = c("Commune" = "", "District" = ""), vary = label_parsed)) +
   labs(x = "# Additional ARMC", y = "Increase in vial demand (national)",
        tag = "B") +
   coord_cartesian(clip = "off") +

### 2020-05-12 R/functions/out.session.R

@@ -5,21 +5,25 @@
 ## Function for writing out session info at end of each script run
 out.session <- function(path, filename = "log.csv", start = NULL) {
   
+  
   library(data.table) ## using data table to deal with easy parsing / binding
   out <- sessionInfo()
   versions <- lapply(out$otherPkgs, function(x) x["Version"])
   versions <- as.data.table(t(unlist(versions)))
   names(versions) <- gsub(".Version", "", names(versions))
   
+  out_gc <- out
+  max_mem_mb <- sum(gc()[, ncol(gc())])
+  
   if (!is.null(start)) {
-    jobtime <- as.numeric(Sys.time() - start, units = "mins")
+    jobtime_mins <- as.numeric(Sys.time() - start, units = "mins")
   } else {
-    jobtime <- NA
+    jobtime_mins <- NULL
   }
   
   toappend <- data.table(ran = timestamp(), 
                          timestamp = format(Sys.time(), "%Y%m%d_%H%M%S"),
-                         jobtime = jobtime, 
+                         jobtime_mins, max_mem_mb,
                          path, R.version = out$R.version$version.string,
                          platform = out$platform, running = out$running, versions)
   

### 2020-05-12 R/functions/ttime_functions.R

@@ -80,84 +80,104 @@ get.ttimes <- function(friction, shapefile, coords, trans_matrix_exists = TRUE,
 #' @section Dependencies:
 #'     Packages: data.table, foreach
 
-add.armc <- function(base_df, clinic_names, clinic_catchmat, 
-                     max_clinics = ncol(clinic_catchmat), thresh_ttimes, 
+
+# Pass through base df, otherwise you need all the vectors!
+
+add.armc <- function(base_df, cand_df, max_clinics, thresh_ttimes, 
                      thresh_prop, dir_name, overwrite = TRUE) {
   
-  prop.lessthan <- function(x, prop_pop, base_ttimes, threshold) {
-    # Sum of the proportion of the population for which ttimes decreased with addition of clinic
-    # BUT only for those people living above threshold ttimes
-    sum(prop_pop[which(x < base_ttimes & x > threshold)], na.rm = TRUE)
+  # Pull in references to files (that way not as big!)
+  cand_list <- vector("list", nrow(cand_df))
+  foreach(j = 1:nrow(cand_df), .packages = "raster") %do% {
+    cand_list[[j]] <- raster(cand_df$candfile[j], band = cand_df$band[j])
   }
   
-  # Add clinics incrementally
   for (i in 1:max_clinics) {
     
-    print(ncol(clinic_catchmat))
-    
-    if (ncol(clinic_catchmat) > 0) {
+    if (nrow(cand_df) > 0) {
+      
       print(i)
-
-      sum.prop <-
-        foreach(vals = iter(clinic_catchmat, by = "col"),
-                .combine = c) %dopar% {
-                  prop.lessthan(vals, prop_pop = base_df$prop_pop, 
-                                        base_ttimes = base_df$ttimes, threshold = thresh_ttimes)
-                }
+      
+      if(i == 27) browser()
+      
+      foreach(j = 1:nrow(cand_df), .combine = c, .packages = "raster") %dopar% {
+                  cand_ttimes <- values(cand_list[[j]])
+                  prop <- sum(base_df$prop_pop[which(cand_ttimes < base_df$ttimes & 
+                                               base_df$ttimes > thresh_ttimes)], na.rm = TRUE)
+      } -> sum_prop
       
       # In case all admin units go below the threshold: stop adding
-      clinic_id <- clinic_names[which.max(sum.prop)]
[…]

### 2020-05-12 bash/bitemods.sh

@@ -1,24 +1,0 @@
-#!/bin/bash
-ssh -T mrajeev@della.princeton.edu <<HERE
-    cd  MadaAccess  # change to repo
-    jid=\$(sbatch bash/bitemods.slurm | cut -c 21-)
-    echo "Here's the job id: \$jid"
-    jstat=\$(sacct -j "\$jid" -u mrajeev | head -n 3)
-    echo "Here's the job stat: \$jstat"
-    until grep -q "COMPLETED\|FAILED\|CANCELLED" <<< \$jstat  # if completed or failed
-    do
-        echo waiting   # updating
-        jstat=\$(sacct -j "\$jid" -u mrajeev | head -n 3)
-        echo "Here's the job stat: \$jstat"
-        sleep 2m # time to sleep for (base it on how long the job should take)
-    done
-    if grep -q "FAILED\CANCELLED" <<< \$jstat
-    then
-        echo "Failed or cancelled"
-        exit
-    else
-        logout
-    fi
-HERE
-        sleep 1m    # sleep again as sometimes takes a while to write output
-        rsync -rLvzt mrajeev@della.princeton.edu:~/MadaAccess/output/mods/ ~/Documents/Projects/MadaAccess/output/mods/ 

### 2020-05-24 docs/manuscript.Rmd
fixing scripts & adding raincloud esque plots

@@
 ```{r setupmain, include=FALSE}
-## Libraries
+# Libraries
 library(knitr)
@@
 
-Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest ARMC (i.e.
-a clinic provisioning PEP) for the country of Madagascar at an approximately 1 x 1 km scale.
-We then extracted the mean of travel times at the district and commune (the administrative unit below the district) level, as well as the mean of travel times weighted by the population in that grid cell.
-We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC.
+Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP at an approximately 1 x 1 km scale.
+We then extracted the mean travel times as well as the mean of travel times weighted by population for the district and commune (the administrative unit below the district) level.
+We also compared these metrics to euclidean distance, taking the minimum distance from the administrative unit centroid to any ARMC.
 We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.
@@
 Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day.
-We did not have the date or month when vials were provisioned, so we compared the totals across this four year period.
+We did not have the date or month when vials were provisioned, so we compared estimated demand to the total vials provisioned across this four year period.
 
@@
 $$ y_i = Poisson(\mu_i) $$
-where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to expected number of bites) for a given source location (district or commune).
+where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
 We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level.
-In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:
+In order to more directly compare estimates between these two datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$) so that:
 
@@
 
-We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
+We tested whether the catchment random effect was capturing overdispersion in the data (i.e.
+variance greater than the mean--the expectation of variance given a poisson distribution) rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
 
@@
 We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’.
-We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10-6)$).
+We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10^{-6})$).
 For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$).
@@
 Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e.
-assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs.
+assuming that data was completely reported resulting in lower estimatates of bite incidence) and with a lower cut-off for identifying periods of undersubmission (7 days vs.
 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).
@@
 For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons).
-The range is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons.
+The range for $E_i$ is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons.
 These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019].
 We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate).
-As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [CITE!!].
+As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
 See figure S4 for further details.
@@
 
-$B_i$, the number of total reported bites, is the mean predicted number of bites from the models of bite incidence as a function of travel times.
+$B_i$, the number of total reported bites, is the predicted number of bites from the models of bite incidence as a function of travel times.
 To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6.
@@
 We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
+
 We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$).
@@
 ## candidate points
-csbs <- read.csv(here("data/processed/clinics/csb2.csv"), stringsAsFactors = FALSE)
-clin_per_comm <- read.csv(here("data/processed/clinics/clinic_per_comm.csv"), 
-                          stringsAsFactors = FALSE)
-```
-
-We developed a framework to rank clinics by how much they could improve access for people currently underserved (> 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
-We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.
-
-Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country.
-We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially.
-In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons).
-Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.
-
-We also explore three additional scenarios: 
-1) Each district has one clinic (the CSB II in the highest population density grid cell compared to other potential clinics) provision PEP (N = 114 clinics). 
-2) Each commune has one clinic (the CSB II, or when there are no CSB II in a commune the health post or dispensary, in the highest density grid cell compared to other potential clinics) provision PEP (N = `r nrow(clinic_per_comm)`.
-Note that not all communes (N = `r length(mada_communes)`) have a clinic).
-3) All CSB II in the country provision PEP (N = `r nrow(csbs)` clinics).
-
-We estimate how expansion to additional clinics may impact vaccine demand.
+csbs <- fread(here("data/processed/clinics/csb2.csv"))
+clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
+clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
+missing_comms <- fread(here("output/stats/missing_comms.csv"))
+comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
+csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
+
+```
+
+We developed a framework to rank clinics by how much they could improve access for people currently underserved and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
+We simulate expansion first to each district (N = `r length(mada_districts)`)and then to each commune in the country for all communes with a clinic.
+We select the CSB II (Centre de Sante Niveau II, major health centers with capacity for delivering vaccines) in the highest density grid cell of the administrative unit as candidates for expansion.
+At the commune level, `r nrow(missing_comms) - comms_no_clins`  communes did not have a CSB II, so we chose the CSB I (Centre de Sante Niveau I, secondary health posts) in the highest density grid cell.
+`r comms_no_clins` communes lacked any CSB I or II.
+
+We first tested three different metrics for ranking clinics:
+
+ 1. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced when a clinic is added.
+ 2. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced weighted by the  magnitude of the change in travel times when a clinic is added.
+ 3. The mean reduction in travel times for people living > 3 hours away from an existing ARMC when a clinic is added.
+
+For all three metrics, once any clinics added only reduced travel times below that threshold (i.e.
+> 3 hours away) for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons), we reduce the threshold by 25%.
+
+For each of these metrics, we sequentially add the clinic with the maximal metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially until each district had a clinic.
+At each step, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.
+We compared the reductions in burden using the decision tree framework described previously for these three strategies, and also to 10 different scenarios where clinics were added randomly.
+We then use the ranking metric that reduced burden the most effectively as clinics were added to expand iteratively to all communes with a clinic.
+Finally, we explore a final scenario where all additional CSB II in the country provision PEP in addition to the clinics added to each district and commune for a total of `r nrow(csbs_max)`.
+
+We also estimate how expansion to additional clinics may impact vaccine demand.
 Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level.
-As clinics are added, multiple clinics may serve a single district or commune (S5.4).
+As clinics are added, multiple clinics may serve a single district or commune.
 Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest.
@@
 For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1).
-We did the same for parameter estimates from the models of reported bite incidence, but rather fixed these estimates at the upper and lower 95% quantiles of their posterior distributions (Table S6.2).
+For parameters from the models of reported bite incidence, we fixed the estimate at the upper and lower 95% quantiles of their posterior distributions (Table S6.2).
 We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden.
@@
 Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e.
-if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.
-We use scaling factors which scale incidence either positively or negatively with the population sizes observed at the district and commune levels, but are still constrained within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4).
+if dog ownership is more common in less populated, rural areas), we might expect rabies exposure incidence within our range to scale negatively with human population size.
+We use scaling factors to scale incidence either positively or negatively with the population sizes observed at the district and commune levels, while still constraining them within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4) and simulated baseline burden, as well as the impact of expanding PEP access.
 
@@
 All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}.
-All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at this github repository.
+All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at [this github repository](https://github.com/mrajeev08/MadaAccess).
 The raw bite patient data at the national level are maintained in a database by IPM.
@@
 Vial demand also increases (Fig 7B), and more vials are needed per averted death as PEP access is expanded (Fig 7C).
-Our model predictes an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
+Our model predicts an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
 visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
 Importantly, clinic throughput decreases, as well, with clinics seeing fewer patients per day (S5.7), as reported bite incidence saturates (Fig S5.5), and instead there are shifts in which clinics bites are reported to.
-
-We also look at two additional scenarios: one CSB II provisions PEP for each district (ARMC per district, N = 114 clinics) and one where one clinic (either a CSB II, CSB I, or dispensary) provisions PEP to each commune (ARMC per commune, N = `r nrow(clin_per_comm)` as all communes have one clinic).
-These expansion scenarios result in similar estimates of impacts on reducing human rabies deaths and increasing vial demand compared to adding the same number of clinics through our stepwise approach.
 
@@
 Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to an additional 250 clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.
+We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic provisioning PEP per district reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`.
 
 However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all major health centers (n = 1676) in the country, between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally.
-
-This result is driven by three key factors.
-First, given the baseline levels of reported bite incidence and our assumptions of underlying rabies exposure, we estimate that burden in areas with low travel times could still potentially be high {inline estimate of reporting/incidence}.
-Second, given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}.
-Finally, the travel time effect, while consistently non-zero and negative, does not account much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {inline overdispersion of sigma_e}.
-
-We find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
+This result is driven by two key factors.
+Given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}.
+Finally, the travel time effect, while consistently non-zero and negative, does not account for much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {expected range of incidence at low travel times}.
+
+Because of these factors, we find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
 Vial demand reflects these patterns as well, however, vials needed per death averted match predictions from a recent cost effectiveness analysis of PEP expansion, suggesting that expanding PEP access will remain highly cost effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+While the qualitative patterns hold, the actualy # of deaths is highly sensitive to our assumptions of rabies exposure incidence.
 

### 2020-05-24 R/01_gis/07_groundtruth_ttimes.R

@@ -5,6 +5,9 @@
 #'  3) IPM driving times (from Helene and Luciano)                                                
 # ------------------------------------------------------------------------------------------------ #
 
+source('R/functions/out.session.R')
+start <- Sys.time()
+
 # Set-up
 library(tidyverse)
 library(rgdal)
@@ -14,14 +17,13 @@ library(data.table)
 library(doParallel)
 library(foreach)
 select <- dplyr::select
-source('R/functions/out.session.R')
 source('R/functions/ttime_functions.R')
 
 # data
 ctar_metadata <- read.csv("data/processed/clinics/ctar_metadata.csv")
 mada_communes <- readOGR("data/processed/shapefiles/mada_communes.shp")
 mada_districts <- readOGR("data/processed/shapefiles/mada_districts.shp")
-base_times <- raster("output/ttimes/baseline_ttimes.tif")
+base_times <- raster("output/ttimes/base/ttimes.tif")
 pop1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
 friction_masked <- raster("data/processed/rasters/friction_mada_masked.tif")
 
@@ -44,8 +46,8 @@ mora %>%
                           Bicycle == 1 ~ "Bicycle", 
                           Pus == 1 ~ "Pus-pus",
                           Other == 1 ~ "Other"),
-         to_long = ctar_metadata$LONGITUDE[ctar_metadata$CTAR == "Moramanga"], 
-         to_lat  =  ctar_metadata$LATITUDE[ctar_metadata$CTAR == "Moramanga"], 
+         to_long = ctar_metadata$long[ctar_metadata$CTAR == "Moramanga"], 
+         to_lat  =  ctar_metadata$lat[ctar_metadata$CTAR == "Moramanga"], 
          ttimes_reported = hours, 
          type = "commune_wtd") %>%
   select(-(car:known_cat1), -ttimes_wtd) -> gtruth_mora 
@@ -94,4 +96,4 @@ gtruth <- bind_rows(gtruth_IPM, gtruth_mora)
 write.csv(gtruth, "output/ttimes/gtruth_ttimes.csv", row.names = FALSE)
 
 # Save session info ---------------------------------------------------------------------------
-out.session(path = "R/01_gis/06_groundtruth.R", filename = "output/log_local.csv")
+out.session(path = "R/01_gis/06_groundtruth.R", filename = "output/log_local.csv", start = start)

### 2020-05-24 R/04_addclinics/04_postprocess.R

@@ -38,13 +38,15 @@ max_stats <- update.base(base_df = base_df, cand_df = csbs_max, nsplit = 5)
 
 max_dist <- aggregate.admin(base_df = max_stats, admin = "distcode", 
                             scenario = nrow(csbs_max))
-max_dist_maxcatch <- get.maxcatch(max_dist, admin = "distcode")
+max_dist_maxcatch <- max_dist[, .SD[prop_pop_catch == max(prop_pop_catch, na.rm = TRUE)], 
+                              by = .(distcode, scenario)]
 fwrite(max_dist, "output/ttimes/addclinics/district_allcatch_max.gz")
 fwrite(max_dist_maxcatch, "output/ttimes/addclinics/district_maxcatch_max.gz")
 
 max_comm <- aggregate.admin(base_df = max_stats, admin = "commcode", 
                             scenario = nrow(csbs_max))
-max_comm_maxcatch <- get.maxcatch(max_comm, admin = "commcode")
+max_comm_maxcatch <- max_comm[, .SD[prop_pop_catch == max(prop_pop_catch, na.rm = TRUE)], 
+                              by = .(commcode, scenario)]
 fwrite(max_comm, "output/ttimes/addclinics/commune_allcatch_max.gz")
 fwrite(max_comm_maxcatch, "output/ttimes/addclinics/commune_maxcatch_max.gz")
 

### 2020-05-24 R/05_predictions/01_preds_incremental.R

@@ -1,6 +1,6 @@
 # ------------------------------------------------------------------------------------------------ #
 #' Getting incremental estimates of burden  
-#' Details: Pulling in district and commune estimates of travel times as clinics are added 
+#' Pulling in district and commune estimates of travel times as clinics are added 
 # ------------------------------------------------------------------------------------------------ #
 
 #sub_cmd=-t 12 -n 30 -mem 4000 -sp "./R/05_predictions/01_preds_incremental.R" -jn preds -wt 5m -n@
@@ -26,11 +26,9 @@ source("R/functions/predict_functions.R")
 source("R/functions/batch_functions.R")
 select <- dplyr::select
 
-# Set-up these two vectors for reading in data using cmd line args
-scenario_loop <- unique(fread("output/ttimes/commune_maxcatch.csv")$lookup)
-colnames <- colnames(fread("output/ttimes/commune_maxcatch.csv"))
-
 # Preds given commune/district ttimes -------------------------------------------------------
+scenario_loop <- unique(fread("output/ttimes/addclinics/commpreds_max.csv")$lookup)
+
 # make df with the lookup + mod pars (reverse vec so big ones at end don't slow things down)
 lookup <- expand.grid(loop = rev(scenario_loop), scale = c("Commune", "District"), 
                       pop_predict = "flatPop", intercept = "fixed", data_source = "National", 
@@ -38,9 +36,9 @@ lookup <- expand.grid(loop = rev(scenario_loop), scale = c("Commune", "District"
                       exp_min = 15/1e5, exp_max = 76/1e5, p_death = 0.16)
 
 all_preds <- run_scenarios(lookup = lookup, pred_type = c("vials", "burden"), 
-                           par_type = "posterior", scaled = FALSE, directory = "output/ttimes/",
-                           colnames_max = colnames(fread("output/ttimes/commune_maxcatch.csv")), 
-                           colnames_all = colnames(fread("output/ttimes/commune_allcatch.csv")),
+                           par_type = "posterior", scaled = FALSE, directory = "output/ttimes/addclinics",
+                           colnames_max = colnames(fread("output/ttimes/addclinics/commpreds_max.csv")), 
+                           colnames_all = colnames(fread("output/ttimes/addclinics/commpreds_all.csv")),
                            colnames_j = c("scale", "data_source", "OD"), 
                            admin_to_keep = lookup$loop, # keep all admin ones for burden
                            multicomb = function(x, ...) { mapply(rbind, x, ..., SIMPLIFY = FALSE)}, 

### 2020-05-24 R/05_predictions/03_preds_se.R

@@ -24,7 +24,7 @@ source("R/functions/batch_functions.R")
 select <- dplyr::select
 
 # Set-up these two vectors for reading in data using cmd line args
-scenario_loop <- unique(fread("output/ttimes/commune_maxcatch.csv")$lookup)
+scenario_loop <- unique(fread("output/ttimes/addclinics/commpreds_max.csv")$lookup)
 se_pars <- fread("output/sensitivity/se_pars.csv")
 
 # reverse so bigger dfs don't slow everything down
@@ -32,13 +32,14 @@ lookup <- expand_grid(loop = rev(scenario_loop), se_pars)
 
 # Do burden -----------------------------------------------------------------
 all_preds <- run_scenarios(lookup = lookup, pred_type = "burden", 
-                           par_type = "point_est", scaled = FALSE, directory = "output/ttimes/",
-                           colnames_max = colnames(fread("output/ttimes/commune_maxcatch.csv")), 
-                           colnames_all = colnames(fread("output/ttimes/commune_allcatch.csv")),
+                           par_type = "point_est", scaled = FALSE,
+                           directory = "output/ttimes/addclinics",
+                           colnames_max = colnames(fread("output/ttimes/addclinics/commpreds_max.csv")), 
+                           colnames_all = colnames(fread("output/ttimes/addclinics/commpreds_all.csv")),
                            colnames_j = c("scale", "data_source", "vary", "direction"), 
                            admin_to_keep = "scenario_0", 
                            multicomb = function(x, ...) { mapply(rbind, x, ..., SIMPLIFY = FALSE)}, 
-                           rng_seed = 2342, sims = 1000)
+                           rng_seed = 23492, sims = 1000)
 
 # Write out data
 fwrite(all_preds$admin_preds, "output/sensitivity/burden_baseline_se.gz") 
@@ -49,10 +50,10 @@ lookup <- filter(lookup, grepl("beta|sigma", vary))
 vial_se_pars <- fread("output/sensitivity/se_pars.csv")[grep("beta|sigma", vary)]
 
 all_preds <- run_scenarios(lookup = lookup, pred_type = "vials", 
-                           par_type = "point_est", scaled = FALSE, catch_keep = FALSE,
-                           directory = "output/ttimes/",
-                           colnames_max = colnames(fread("output/ttimes/commune_maxcatch.csv")), 
[…]

### 2020-05-24 R/05_predictions/04_scaling_se.R

@@ -57,7 +57,7 @@ scaling_df <- rbind(neg_comm, pos_comm, neg_dist, pos_dist)
 write.csv(scaling_df, "output/sensitivity/scaling.csv", row.names = FALSE)
 
 # Scaled preds --------------------------------------------------------------------------------
-scenario_loop <- unique(fread("output/ttimes/commune_maxcatch.csv")$lookup)
+scenario_loop <- unique(fread("output/ttimes/addclinics/commpreds_max.csv")$lookup)
 
 lookup <- expand_grid(loop = rev(scenario_loop), scaling_df, pop_predict = "flatPop", 
                       data_source = "National", intercept = "fixed", OD = TRUE,
@@ -65,9 +65,9 @@ lookup <- expand_grid(loop = rev(scenario_loop), scaling_df, pop_predict = "flat
                       exp_min = 15/1e5, exp_max = 76/1e5, p_death = 0.16)
 
 all_preds <- run_scenarios(lookup = lookup, pred_type = "burden", 
-                           par_type = "posterior", scaled = TRUE, directory = "output/ttimes/",
-                           colnames_max = colnames(fread("output/ttimes/commune_maxcatch.csv")), 
-                           colnames_all = colnames(fread("output/ttimes/commune_allcatch.csv")),
+                           par_type = "posterior", scaled = TRUE, directory = "output/ttimes/addclinics",
+                           colnames_max = colnames(fread("output/ttimes/addclinics/commpreds_max.csv")), 
+                           colnames_all = colnames(fread("output/ttimes/addclinics/commpreds_all.csv")),
                            colnames_j = c("scale", "data_source", "scaling"), 
                            admin_to_keep = "scenario_0", 
                            multicomb = function(x, ...) { mapply(rbind, x, ..., SIMPLIFY = FALSE)}, 
@@ -83,7 +83,7 @@ out.session(path = file_path, filename = "log_cluster.csv", start = start)
 
 # Parse these from bash for where to put things
 syncto <- "~/Documents/Projects/MadaAccess/output/sensitivity/"
-syncfrom <- "mrajeev@della.princeton.edu:~/MadaAccess/output/sensitivity/burden*"
+syncfrom <- "mrajeev@della.princeton.edu:~/MadaAccess/output/sensitivity/*"
 
 closeCluster(cl)
 mpi.quit()

### 2020-05-26 docs/manuscript.Rmd
cleaning up repo

@@
   - Laurence Baril:
-      institute: [IPCambodia]
+      institute: [IC]
       correspondence: false
@@
   - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
+  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
+  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
+  - OMS: World Health Organization, Madagascar
+  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
+  - Geneva: University of Geneva
+  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
   - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
   - IC: Institut Pasteur de Cambodge
-  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
-  - OMS: OMS
-  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
   - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
-  - Geneva: University of Geneva
 date: ""
 link-citations: no
+nocite: '@*'
 output:
@@
     pandoc_args:
+      - --csl=format/plos-neglected-tropical-diseases.csl
       - --filter=pandoc-crossref
@@
     toc_float: true
+    number_sections: no
     pandoc_args:
@@
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
-bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
-# bibliography_main: ["bib/refs_all.bib"]
-# bibliography_supp: ["bib/refs_pkgs.bib"]
+bibliography_main: ["bib/refs_all.bib"]
+bibliography_supp: ["bib/refs_pkgs.bib"]
 ---
-
-##### TO DO: fix affiliations
 
@@
 library(magrittr)
-
-opts_chunk$set(echo = FALSE, warning = FALSE, include = FALSE)
+opts_chunk$set(echo = FALSE, warning = FALSE, out.width="80%")
+
+trans_perc <- function(x) {
+  round(x, 2)*100
+}
 
 ## data
-ctar_metadata <- fread(here("data/raw/ctar_metadata.csv"))
+ctar_metadata <- fread(here("data/processed/clinics/ctar_metadata.csv"))
 district_bites <- fread(here("output/bites/district_bites.csv"))
@@
 
-trans_perc <- function(x) {
-  round(x, 2)*100
-}
-
-```
+## All preds (for abstract stats etc)
+natl_preds <- fread(here("output/preds/natl_preds.gz"))
+natl_base <- natl_preds[scenario == 0]
+
+natl_preds %>%
+  filter(scenario == 0) %>%
+  mutate(prop_averted = averted_mean/(deaths_mean + averted_mean)) %>%
+  select(prop_averted) -> prop_averted
+
+admin_base <- fread(here("output/preds/admin_preds_base.gz"))
+
+## candidate points
+csbs <- fread(here("data/processed/clinics/csb2.csv"))
+clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
+clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
+missing_comms <- fread(here("output/stats/missing_comms.csv"))
+comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
+csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
+
+max_added <- max(natl_preds$scenario)
+natl_preds %>%
+  group_by(scale, data_source) %>%
+  arrange(scenario) %>%
+  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
+trends <- read.csv(here("output/stats/trends_adminpreds.csv"))
+
+one_per_dist <- nrow(clin_per_dist)
+one_per_comm <- sort(natl_preds$scenario, decreasing = TRUE)[2]
+
+no_figs <- FALSE
+if(no_figs) include_graphics <- function(x, ...) { return(invisible(NULL)) }
+
+```
+
 \newpage
@@
 # ABSTRACT
-Access to care shapes burden of diseases globally.
+
+## Background
 For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries.
-Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners.
-In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths.
-Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP.
-
-Go in order.
-
-We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country.
-We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden.
-
-We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic).
-Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.
-
-However our estimates are dependent on blah blah blah.
-
-These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.
+A target to eliminate deaths due to canine rabies by the year 2030 has been set forth by global health partners, and in 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly expand PEP availability and reduce rabies burden.
+Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited spatially (N = 31 Anti-Rabies Medical Center or ARMC across the country).
+
+## Methodology & Principal Findings
+We use travel times to the closest ARMC as a proxy for access to PEP.
+We find that travel times are a strong predictor of reported bite incidence across the country.
+We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes burden sub-nationally.
+We estimate an annual burden of `r round(sort(natl_base$deaths_mean)[1])` (95% CI: `r round(sort(natl_base$deaths_lower)[nrow(natl_base)])` - `r round(sort(natl_base$deaths_upper)[nrow(natl_base)])`), with current PEP provisioning averting ~ `r glue("{trans_perc(min(prop_averted))}%")` percent of deaths.
+Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`.
+However, our results suggest that even with all major health centers in the country provisioning PEP (N = `r max(natl_preds$scenario)` additional clinics), we could still see substantial mortality due to rabies.
+Our estimates are most sensitive to assumptions of underlying rabies exposure incidence, but the qualitative patterns of the impacts of travel times and expanded access on burden are robust to these assumptions.
+
+## Conclusions & Significance
+PEP is effective at preventing deaths and in the absence of rabies data, targeting expansion to populations currently underserved may be the most equitable way to provision PEP.
+Data on rabies incidence will be neccessary to refine these estimates and to better target interventions.
+While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed & improved clinic functioning & surveillance systems.
+A GAVI investment can propel countries forward on this path combining these life saving interventions.
 
 # AUTHOR SUMMARY 
+Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low and middle income countries where people have limited access to both human (post-exposure prophylaxis or PEP) and dog rabies vaccine.
+Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by the year 2030, and recently, GAVI, a organization that provides funding for immunization programs to eligible countries, added human rabies vaccine to it's funding agenda.
+In this study, we model estimates of reported bite incidence as a function of travel times and estimates of rabies exposure incidence to extrapolate the burden of human rabies given the current availability of PEP across the country.
+We find that PEP currently averts a signficant number of deaths, but that burden remains high, particularly in areas with high travel times to a clinic.
+Expanding PEP availability to more clinics could significantly reduce the current burden of rabies deaths, but our results suggest that expansion alone will not eliminate deaths due to rabies under current levels of reporting.
+Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other countries like it, forwards on the path to elimination.
 
@@
 Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans .
+Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans.
 While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018].
@@
 From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.
-Yet it is unclear how these vaccines should be distributed to prevent deaths most effectively.
-
-A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
-The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e.
+
+A primary challenge in expanding access effectively is the lack of data on rabies exposures and deathss in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
+Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in the most vulnerable and remote communities [@taylor2017difficulties].
+Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites and a decision tree framework to extrapolate deaths, with the key assumption that overall reported bite incidence (i.e.
 both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
@@
 
-## Estimating travel time to closest ARMC {#methods-spatial}
-To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).
+## Estimating travel time to closest ARMC {#methods-ttimes}
+To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B.
 
@@
 However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
-To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1).
+To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1)).
 We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic.
@@
 
-```{r fig 1, fig.fullwidth=TRUE}
+```{r fig 1}
 include_graphics(here("figs/main/M1_sankey.jpeg"))
 ```
-*__Figure 2.__Decision tree for burden estimation.
+
+*__Figure 1__.
+Decision tree for burden estimation.
 For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$).
@@
 ## Estimating the impact of expanding PEP provisioning to additional clinics {#methods-incremental}
-
-```{r ncsbs}
-## candidate points
-csbs <- fread(here("data/processed/clinics/csb2.csv"))
-clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
-clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
-missing_comms <- fread(here("output/stats/missing_comms.csv"))
-comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
-csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
-
-```
 
@@
 
-## Data and code 
-All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}.
-All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at [this github repository](https://github.com/mrajeev08/MadaAccess).
-The raw bite patient data at the national level are maintained in a database by IPM.
-This database was last queried on September 19, 2019 for these analyses.
-Anonymized raw data from this database are available from IPM following insitutional data transfer protocols.
-Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.
+## Data and analyses 
+All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (section S7).
+All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at github.com/mrajeev08/MadaAccess.
+The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC.
+These databases were last queried on September 19, 2019 for these analyses.
+Anonymized raw bite patient data are available from IPM following insitutional data transfer protocols.
+Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (wisemonkeyfoundation.org) on the same date and are shared in the archived repository.
 
@@
 
-Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population living greater than 3 hrs away from a clinic (Fig 3B).
-We found that travel times estimated from the friction surface generally underestimate true driving times and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2).
+Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live greater than 3 hrs away from a clinic (Fig 3B).
+However, we found that these estimates from the friction surface generally underestimate true driving times across the country and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2).
 The friction surface assumes that the fastest available mode of transport is used to travel each grid cell (i.e.
@@
 However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3).
-Despite these discrepancies, travel times correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4).
+Despite these discrepancies, travel time estimates from the friction surface correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4).
 Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1).
@@
 
-```{r fig M2, fig.fullwidth=TRUE}
+```{r fig M2}
 include_graphics(here("figs/main/M2_ttimes.jpeg"))
 ```
-*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar.
+
+*__Figure 2A)__ Travel times to the closest ARMC estimated at an ~ 1 km^2^ scale for Madagascar.
 __B)__ Distribution of the population across travel times.
-__C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level.
+__C)__ Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level.
 For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune.
-The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown.
+The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
 The dashed black line is the 1:1 line.*
@@
 ```
-*__Figure 3.A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
+
+*__Figure 3A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
 Color corresponds to the clinic catchment.
@@
 For panel __A)__ districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
-For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), and all communes included in it's catchment (red polygons) and other communes where bites were reported to in grey.*
+For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), all communes included in it's catchment (red polygons), and other communes where bites were reported to in grey.*
 
 Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1).
-We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)), and correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned ([Figure S2.2A and B](#figS2.2)).
+We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms (Figure S2.1B).
+Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned for the majority of clinics (Figure S2.2A and B).
 We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A).
@@
 For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
-While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district fell within the ranges observed from the national dataset.
+While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.
 
@@
 Similarly, while the commune model fit to the Moramanga data generated stronger estimates of the travel time effect (Fig 5B), after accounting for overdispersion of the data, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
-Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models (Fig S3.8) and generated similar predictions of the range of bite incidence across travel times (Fig S3.9).
-
-As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates from the commune and district models fit to the national data, as they encompass the range of travel time effects observed in our datasets.
-
-```{r fig 5, fig.fullwidth=TRUE}
+Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models and performed similarly in predicting the data, and instead the correction applied impacted the magnitude of the bite incidence estimates (Fig S3.8 and Fig S3.9).
+
+As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates models fit to the national data, as they encompass the range of travel time effects observed in our datasets.
+In addition, our out of fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at this scale, therefore we use both the district and commune model to disaggregate burden to the finest scale possible given our modeling framework.
+
+```{r fig 5}
 include_graphics(here("figs/main/M5_mods.jpeg"))
 ```
+
 _**Figure 5.** Travel times as a predictor of reported bite incidence per 100,000 persons.
@@
 The envelopes show the 95% prediction interval.
-The points show the data used to fit the models: National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model.
+The points show the data used to fit the models: National data (circles) at the district level used to fit the District and Commune models, and Moramanga data (triangles) at the commune level used to fit the Moramanga model.
 The posterior distribution of parameters from the respective models for the B) model intercept, C) travel time effect, and D) the overdisperson parameter (for national data only)._
@@
 ## Estimating the burden of human rabies deaths subnationally {#results-burden}
-
-```{r burden results}
-natl_preds <- fread(here("output/preds/burden_natl.gz"))
-natl_base <- natl_preds[scenario == 0]
-```
 
@@
 We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_base$averted_mean)[1])` - `r round(sort(natl_base$averted_mean)[nrow(natl_base)])` deaths due to rabies are prevented through provisioning of PEP each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally when burden is modeled at the district level, with even stronger patterns of increase in burden with travel times when modeling burden at the commune level with estimates from the Moramanga data (Figure 4).
-The estimates of burden from the model fitted to the Moramanga data result in estimates of burden that are lower at the national level, but higher relatively in places with high travel times {fix!}.
+In general, the incidence of rabies deaths increases with travel times to clinics, and  there is significant variation sub-nationally when burden is modeled at the district level, with even stronger patterns of increase in burden with travel times when modeling burden at the commune level (Figure 4).
 However, there is considerable uncertainty in these estimates, in particular given the district model of bite incidence.
@@
 
-```{r fig 4, fig.fullwidth=TRUE}
+```{r fig 6}
 include_graphics(here("figs/main/M6_burden_base.jpeg"))
 ```
+
 *__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
@@
 ## The impact of expanding PEP access to additional clinics {#results-incremental}
-```{r}
-max_added <- max(as.numeric(natl_preds$scenario[!(natl_preds$scenario %in% c("max", "armc_per_dist", "armc_per_comm"))]))
-natl_preds %>%
-  mutate(scenario_num = case_when(!(scenario %in% c("max", "armc_per_dist", "armc_per_comm")) ~ as.numeric(scenario),
-                                  scenario == "max" ~ max_added + 200, scenario == "armc_per_dist" ~ 114, 
-                                  scenario == "armc_per_comm" ~ max_added + 150)) %>%
-  group_by(scale, data_source) %>%
-  arrange(scenario_num) %>%
-  mutate_at(vars(-c(group_cols(), scenario, scenario_num)), function(x) x/x[1]) -> props
-trends <- read.csv(here("output/stats/trend_preds.csv"))
-
-```
-
-We found that expanding PEP to additional clinics reduces rabies deaths, with strategic expansion of access to an additional 250 clinics reducing burden by `r round(min(1 - props$deaths_mean[props$scenario == 250])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 250])*100, 2)`% (Fig 7A).
-However, benefits accrue less rapidly as ARMC are added: with approximately 700 clinics, we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == 700])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == 700])*100, 2)`.
-This saturation is a result of the landscape of clinics across the country, as additional clinics reduce clinics for fewer people as more clinics are added (Fig S5.1).
-Even at maximal PEP access, our model still predicts `r round(min(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.6).
-
-Vial demand also increases (Fig 7B), and more vials are needed per averted death as PEP access is expanded (Fig 7C).
-Our model predicts an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
-visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
-Importantly, clinic throughput decreases, as well, with clinics seeing fewer patients per day (S5.7), as reported bite incidence saturates (Fig S5.5), and instead there are shifts in which clinics bites are reported to.
-
-```{r fig 5, fig.fullwidth=TRUE}
+For a subset of clinics (n = 83, up to one clinic per district), we compared the impact of three different metrics for ranking clinics on reducing annual burden of deaths.
+We found that metrics that targeted expansion at clinics based on the proportion of the population they reduced travel times for (above a threshold of three hours) performed the best (Fig S5.1).
+Therefore, we used this metric for simulating expansion to the larger set of clinics (N = `r max_added` clinics).
+
+We found that strategic expansion of access to these additional `r nrow(clin_per_dist)` clinics (i.e.
+1 clinic per district) reduced burden by `r round(min(1 - props$deaths_mean[props$scenario == one_per_dist])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == one_per_dist])*100, 2)`% (Fig 7A).
+With one clinic per commune (where a clinic was available, N = `r one_per_comm`), we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == one_per_comm])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == one_per_comm])*100, 2)`.
+However, reductions in burden saturate as more clinics are added (Fig S5.2).
+Even when all CSB II provision PEP in the country, our model still predicts `r round(min(1 - natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` - `r round(max(1 - natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.5).
+This is largely due to the fact that as more clinics are added, reported bite incidence saturates (Fig S5.4), and instead there are shifts in which clinic bites are reported to (S5.7 & S5.8).
+
+Vial demand also outpaces reductions in burden (Fig 7B), with more vials  needed per averted death as PEP access is expanded (Fig 7C).
+Our model predicts an increase from `r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == 0]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
+visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == 250]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
+Importantly, clinic catchment populations and throughput decrease, as well, with clinics serving fewer patients per day (S5.6).
+
+```{r fig 7}
 include_graphics(here("figs/main/M7_addARMC.jpeg"))
@@
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The points show three additional scenarios: 1) one ARMC per district (N = 114 clinics total, + 83 clinics), 2) one ARMC per commune where atleast one is present (N = `r nrow(clin_per_comm)` total, + `r nrow(clin_per_comm) - 31` clinics), and 3) all CSB II in the country provision PEP (N = `r nrow(csbs) + 31` clinics total, + `r nrow(csbs)`)*
+The points show when all additional csb (N = `r max_added`) clinics have been added).
+*
 
@@
 To quantify which parameters contribute to the uncertainty observed in our estimates, we performed a univariate sensitivity analyses, varying parameters in both the model of bite incidence (Table S6.1) and the decision tree model (Table S6.2).
-These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden and for determining the impact of expanding access (Fig S6.2).
-Uncertainty in bite model parameters contribute less to variation of estimates of burden.
+These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden (Fig S6.1) and for determining the impact of expanding access (Fig S6.2).
+Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden.
 For the estimates of vial demand, uncertainty around the intercept of the model (i.e.
 the baseline reported bite incidence) has the greatest impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
-Finally, scaling of incidence with population size modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+Finally, scaling of incidence with population size (Fig S6.4) modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
 more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths (Fig S6.5A).
 However, the impact of adding clinics remains broadly the same (Fig S6.5B).
-Overall, while uncertainty in the underlying rabies exposure incidence results in significant variation in the absolute number of rabies burden, our results of the current impact of travel times on burden and of the potential for expanded access to reduce burden are qualitatively similar across parameter estimates.
+Overall, while uncertainty in the underlying rabies exposure incidence results in significant variation in the absolute number of rabies deaths, our estimates of the impact of travel times on burden and of the potential for expanded access to reduce burden are qualitatively similar across parameter assumptions.
 
 # DISCUSSION
-```{r}
-natl_preds %>%
-  filter(scenario == 0) %>%
-  mutate(prop_averted = averted_mean/(deaths_mean + averted_mean)) %>%
-  select(prop_averted) -> prop_averted
-```
 
@@
 Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
-Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic provisioning PEP per district reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`.
-
-However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all major health centers (n = 1676) in the country, between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally.
+Under the status quo (N = 31 clinics), we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
+We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic per district (N = `r one_per_dist`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")` and to one clinic per commune (N = `r one_per_comm`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")`.
+
+However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all additional CSB II in the country (N = `r max(natl_preds$scenario)` additional clinics), between `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))} - {round(max(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))}")` deaths may still occur nationally.
 This result is driven by two key factors.
-Given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}.
-Finally, the travel time effect, while consistently non-zero and negative, does not account for much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {expected range of incidence at low travel times}.
-
-Because of these factors, we find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
-Vial demand reflects these patterns as well, however, vials needed per death averted match predictions from a recent cost effectiveness analysis of PEP expansion, suggesting that expanding PEP access will remain highly cost effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-While the qualitative patterns hold, the actualy # of deaths is highly sensitive to our assumptions of rabies exposure incidence.
+First, given the landscape of clinics and travel times across the country, despite expansion to all major health centers in the country, travel times remain high for a significant proportion of the population (Figure S5.4).
+Secondly, even for locations with low travel times given the range human rabies exposures between 15 - 76 per 100,000 persons, we estimate an average incidence of human rabies deaths of `r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` (Figure S5.5).
+Importantly, for administrative units with similar travel times, we estimate a significant amount of variation that cannot be explained by travel times alone, suggesting other factors may be driving access and reporting of bite incidence.
+
+Because of these factors, we find that as PEP provisioning is expanded to more clinics, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
+Vial demand reflects these patterns as well.
+However, vials needed per death averted are lower than predictions from a recent cost effectiveness analysis of PEP expansion (assuming a cost of 5 USD per vial, a maximum of `r round(max(natl_preds$vials_upper/natl_preds$averted_upper)*5)` USD per death averted in direct costs, similar to the ~ 600 USD per death averted estimated in @whorabiesmodelingconsortiumPotentialImpactImproved2019), suggesting that expanding PEP access will remain highly cost effective given current ranges of reported bite incidence.
+Ultimately, these quantitative findings are highly dependent on the assumption of underlying rabies exposure incidence.
+Data on rabies incidence in dogs and exposures in humans are needed to confirm these estimates, but the qualitative patterns regarding travel time impacts remain robust and therefore can be useful in identifying strategies for expanding PEP and reducing burden.
 
@@
 The travel time estimates from the MAP friction surface underestimate true travel times reported by patients, likely due to discrepancies between assigned transport speeds and modes and the reality of travel in Madagascar.
-These speeds are taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018], and also assumes that given multiple types of transport in a cell, the fastest available mode of transport is used.
+These speeds are taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018], and also assume that given multiple types of transport in a cell, the fastest available mode of transport is used.
 In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
@@
 
+We were able to use multiple sources of data to compare estimates of bite incidence across scales.
+We found that we could effectively disaggregate burden to the commune scale using our travel time metric.
+The Moramanga data demonstrates that sub-district patterns of access are important to account for when estimating potential areas with low access.
+However, we lacked data from other catchments at the commune scale to further confirm and compare finer scale patterns.
+Focusing on routinely collecting data at the sub-district level could help to resolve these patterns.
+
 We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
@@
 In fact, a recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [@leblancclaireRabiesMadagascarThreepronged2019].
-OUTLIERS TO TIE INTO OD & other factors
-
-We included an overdispersion parameter in our model to quantify variation not due to travel times, and for all models, the data were significantly overdispersed, suggesting that variation at the clinic level (i.e.
-differences in clinic functioning such as vaccine availability) and at the administrative level (i.e.
+We also assume that all ARMC reliably provision PEP, but this is unlikely to be the case.
+Leblanc 2019 found that at the time of their KAP survey, three clinics were experiencing vaccine stock-outs.
+In addition, while direct costs of the vaccine are not charged to patients, 22 of 31 current ARMC ask for other fees (ranging from 0.50 - 3.00 USD for consultations, gauze, supplementary medication for wound treatment, etc per visit; @leblancclaireRabiesMadagascarThreepronged2019) which might additionally prevent people from receiving treatment after reporting.
+
+We included an overdispersion parameter in our model to quantify variation not captured by travel times.
+For all models, the data were significantly overdispersed, suggesting that variation at the clinic level (i.e.
+differences in clinic functioning such as vaccine availability and costs charged) and at the administrative level (i.e.
 differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
@@
 Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems.
-
-SCALE SCALE We were able to use multiple sources of data to compare estimates of bite incidence across scales.
-We found that we could effectively disaggregate burden to the commune scale using our travel time metrics.
-The Moramanga data demonstrates that sub-district patterns of access are important to account for when estimating potential areas with low access.
-However, we lacked data from other catchments at the commune scale to further confirm and compare finer scale patterns.
 
@@
 
-We also make several simplifying assumptions regarding PEP compliance and efficacy.
+We make several simplifying assumptions regarding PEP compliance and efficacy.
 We assume perfect compliance of patients and efficacy of PEP.
@@
 
-We also assume that all ARMC reliably provision PEP, but this is unlikely to be the case, as Leblanc 2019 found that at the time of their KAP survey, three clinics were experiencing vaccine stock-outs, which likely contributes to burden.
-PART ABT THEM CHARGING PATIENTS FOR OTHER FEES HERE (UNCLEAR WHETHER THEY ARE IN THE DATA).
-22 ARMC ask for fees (consultation fees to fees for guaze etc for the form.
-ranging from 2000 Ar - 10000 Ar in some cases, 0.50 - 3,00 USD).
-And while PEP is free, KAP bit?
-In fact, 22 clinics still charge patients (on average ~ 2000 Ar but anywhere from 500 - 10000 Ar for costs other than the vaccine) so this could be a factor too.
-PRovisioning and clinic functioning will be as important as expanding to everywhere!
-
-Ideally, many of these assumptions could either be confirmed or replaced with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
+Ideally, many of these assumptions could either be confirmed or updated with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
 Even with these additional data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict the impact of PEP spatially.
-We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior, particularly in the context of PEP provisioning, and the policy implications of these exceptions.
-
-## Box 1: case studies of the challenges of predicting health seeking behavior
+We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior in the context of PEP provisioning, and the policy implications of these exceptions.
+
+## Box 1: case studies of the challenges of predicting health seeking behavior and PEP efficacy
 ```{r}
@@
   summarize(prop_cat1 = sum(known_cat1)/n()) -> prop_cat1_AA
-admin_preds <- fread(here("output/preds/burden_base.gz")) 
+admin_preds <- fread(here("output/preds/admin_preds_base.gz")) 
 admin_preds %>%
@@
   summarize(inc = sum(deaths_mean)/sum(pop)*1e5, bite_inc = sum(bites_mean)/sum(pop)*1e5) -> aa_preds
+
+ranked_clins <- fread(here("output/ttimes/addclinics/clinics_ranked.csv"))
+aa_added <- min(ranked_clins[district == "Anosibe-An'ala"]$scenario)
+gtruth <- fread(here("output/ttimes/gtruth_ttimes.csv"))
+gtruth %>%
+  filter(district == "Anosibe-An'ala") %>%
+  summarize(mean = mean(ttimes_reported, na.rm = TRUE), 
+            min =  min(ttimes_reported, na.rm = TRUE),
+            max = max(ttimes_reported, na.rm = TRUE)) -> aa_ttimes
+admin_base %>%
+  mutate(distcode = substr(names, 1, 7)) %>%
+  filter(distcode == "MG33317") %>%
+  group_by(scale) %>%
+  summarize(deaths_mean = sum(deaths_mean), pop = sum(pop)) -> aa_deaths
+
 ```
@@
 ### Anosibe A'nala
-In the district of Anosibe A'nala (population of ~ `r round(mada_districts$pop[mada_districts$district = "Anosibe-An'ala"])` people) south of the Moramanga District, travel times to the nearest ARMC are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
+In the district of Anosibe A'nala (population of ~ `r round(mada_districts$pop[mada_districts$district == "Anosibe-An'ala"])` people) south of the Moramanga District, travel times to the nearest ARMC are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
 Despite these barriers, we see a moderate incidence of reported bites (~ `r round(aa_bites$inc)` per 100,000 persons).
@@
 people who had been in close proximity or had touched a person that had died from suspected rabies.
-Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 5 deaths in this district over a `r max(mora_bites$month)` month period.
+Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 4 deaths in this district over a `r max(mora_bites$month)` month period.
 Only one animal case was confirmed diagnostically however.
@@
 
-1) This remote community does suffer from rabies exposure and is likely experiencing a high burden of death that we are only seeing the tip of the iceberg for.
-
-2) When people are aware of the risk of death due to rabies, they are willing to travel a long time (in many cases walking 3 days to attend the clinic) to seek the vaccine.
-In addition, surveillance responses, i.e.
-when there is a suspected cases, there is often referral of these people from far away, can work and if these investigations are triggered on exposure rather than death then they could greatly help to prevent deaths.
-There is great potential for community outreach to go far.
-
-3) What do we estimate as the death incidence?
-When would we add a clinic?
-The travel times estimated from the friction surface are much lower than those observed.
-Because while there is a road, road speeds are often < 10km per hour and people hitch rides with transport trucks.
-This is also highly seasonal, during the rainy season it's almost impossible to travel other than by foot.
-And most people use multiple modes of transport depending on cost and availability (i.e.
-boat, on foot, by camion).
-So these metrics may miss key areas that are underserved and warrants improvement.
+1) Given the high travel times and the observed incidence of rabies exposures and human rabies deaths, this remote community is likely experiencing a much higher burden of rabies of which we are observing only a small fraction.
+In our modeling analysis, we estimate that an annual burden of `r round(aa_deaths$deaths_mean[1], 2)` per year.
+
+2)Travel times estimated from the friction surface (average weighted travel times of `r round(mada_districts$ttimes_wtd[mada_districts$district == "Anosibe-An'ala"]/60, 2)`) are much lower than those observed for this district (average patient reported travel times of `r round(aa_ttimes$mean, 2)` hours).
+While there is a road connecting the main commune of Anosibe An'ala to the Moramanga ARMC, road speeds are often < 10km per hour and conditions, particularly during the rainy seasons, make it impossible for anything but large transport trucks to travel this road.
+Most people use multiple modes of transport depending on cost and availability (i.e.
+boat, on foot, by transport truck).
+In our ranking of clinics by reduction in travel times, this clinic was the `r aa_added`th clinic added.
+However, given the recurrent cases of human rabies deaths and the high travel times, this area might deserve higher prioritization.
+Importantly, other remote communities may also be experiencing a high and hidden burden of rabies, and improved surveillance across the country will be necessary to best prioritize resources.
+
+3) Despite these high travel times, when people are aware of the risk of death due to rabies and when PEP is available at no direct costs, they are willing to travel many hours to days (in some cases walking 3 days to attend the clinic) to seek PEP.
+In addition, a strong surveillance response, i.e.
+referral of people in contact with suspect human cases of rabies is possible, and can work to improve PEP efficacy if these investigations were triggered on human exposure or animal cases rather than a human death.
+Community outreach and active surveillance in other remote areas could also greatly improve people's awareness of risk and health seeking behavior.
 
@@
 In 2016, a taxi driver in his 60s died of rabies in suburban Antananarivo.
-He was bitten by an unknown dog returning from work one evening, and the next day he promptly reported to a clinic where he was reffered to the ARMC at IPM.
-His daughter, a prospective medical student, and other family members urged him to seek the vaccine, but he did not believe his risk was high and so in the end did not seek PEP.
+He was bitten by an unknown dog returning from work one evening, and the next day he promptly reported to a clinic where he was reffered to the ARMC at IPM for PEP.
+His daughter, a prospective medical student, and other family members urged him to seek the vaccine, but he did not believe his risk was high and so in the end did not seek further care.
 He developed symptoms 2 weeks later and died within a few days.
-The case was confirmed by the Naional Rabies Reference Laboratory at IPM.
+The case was later confirmed as a rabies death by the Naional Rabies Reference Laboratory at IPM.
 This case points to the other side of the health-seeking behavior coin from Anosibe A'nala.
 Even given prompt reporting and referral, and socioeconomic indicators that would suggest a high probability of seeking PEP, this person did not seek PEP.
-However, the case does point to the fact that community education on rabies risk, as well as availability of PEP at peripheral clinics even in areas with already high access could remove further barriers to PEP and prevent additional deaths.
+This case highlights the need for community education on rabies risk, and how improving availability of PEP at peripheral clinics even in areas with already high access could remove further barriers to PEP and prevent additional deaths.
+It also highlights the limitations of PEP provisioning as the sole intervention for preventing human rabies deaths.
 
 ## Broader context
-
 Our work is in line with previous studies demonstrating the impacts of geographic access to care on health outcomes [@aleganaTreatmentseekingBehaviourLow2017; @nikolayEvaluatingHospitalBasedSurveillance2017; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @Hegde2019dt; @metcalfTransportNetworksInequities2015].
@@
 
-The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa, and largely is in line with previous model estimates of bite incidence based on national statistics.
-USE SPECIFIC ESTIMATES HERE (i.e.
-human development index & proportion of the population that is urban, see @whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
-Ultimately, reported bite incidence is driven by many factors unrelated to rabies including the underlying size of dog populations, barriers to access to PEP (i.e.
+The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa (see @whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
+Ultimately, reported bite incidence is driven by many factors unrelated to rabies incidence in the dog population, including the underlying size of dog populations, barriers to access to PEP (i.e.
 both financial costs and geographic access), and awareness of rabies risk.
 In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are genuine rabies exposures.
-As we did not have location specific estimates of this parameter, we used a range estimated from a previous study in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018], however this approach is likely to be less applicable in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
-in settings where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population).
-
-While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies STATE RANGES HERE this range estimates in our study vs this range estimated in these studies [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
-We estimate a higher burden of rabies deaths here STATE that previously for Madagascar STATE [@rajeevHealthcareUtilizationProvisioning2018].
-While in this study we assumed a lower range of incidence of underlying rabies exposure, previously we assumed a fixed reporting of 85% of rabies exposures across the country.
+As we did not have location specific estimates of this parameter, we used a range estimated from a previous study in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018].
+This approach is likely to be less applicable in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
+where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population) and reporting of non-rabid bites high.
+
+While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies  [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
+We estimate a higher burden of rabies deaths here than we previously estimated for Madagascar [@rajeevHealthcareUtilizationProvisioning2018].
+While in this study we assumed a lower range of incidence of underlying human rabies exposures, previously we assumed a fixed reporting of 85% of rabies exposures across the country.
 Accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning.
@@
 Our estimates of vial demand depend on countrywide adoption of the new abridged intradermal regimen recommended by WHO [@tarantolaIntradermalRabiesPostexposure2018].
-While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019 only 15% of staff interviewed were aware of this regimen [@leblancclaireRabiesMadagascarThreepronged2019].
+While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019, only 15% of staff interviewed were aware of this regimen [@leblancclaireRabiesMadagascarThreepronged2019].
 In addition, the majority of staff were also not aware of the WHO classifications of exposure categories, and vaccination of Category I exposures, i.e.
 those not recommended PEP, is a common practice.
-In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018], and more judicious PEP provisioning could reduce vial demand, particularly if expansion of PEP results in increases in reported bite incidence beyond what we predict.
+In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018], and more judicious PEP provisioning could reduce vial demand accordingly.
 
@@
 drone delivery of emergency medicine, [@boeras2019use]) may be potential ways to mitigate these challenges.
-IMPROVED vacc technology!
 Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine from IPM upon demand.
@@
 
+```{r}
+catch_trends <- fread(here("output/stats/trend_catches.csv"))
+catch_trends %>% 
+  group_by(scenario) %>% 
+  summarize(mean_pops = mean(pops_mean)) %>%
+  as.data.table() -> catch_pops
+```
+
 A Gavi investment can remove this cost barrier.
 In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
-Currently, each clinic in Madagascar serves on average a catchment population of 750,000 persons.
-Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons, which would require approximately 200 additional ARMC in Madagascar.
+Currently, each clinic in Madagascar serves on average a catchment population of `r round(mean(catch_trends[scenario == 0]$pops_mean), -3)` persons.
+Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons, which would require approximately `r min(catch_pops[mean_pops < 1e5]$scenario)` additional ARMC in Madagascar.
 If GAVI investment results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
-Importantly, given our estimates of vial demand and deaths averted and accounting for direct vaccine costs only, PEP expansion remains highly cost-effective (< 500 USD per death averted given average vaccine vial costs, @whorabiesmodelingconsortiumPotentialImpactImproved2019).
-
-Ultimately even given expanded access, achieving 'the last mile', i.e.
-preventing deaths due to rabies in the most remote populations will require a disproportionate level of resources and our results suggest that this will not be achievable through PEP alone [@delriovilasTribulationsLastMile2017].
+Given our estimates of vial demand and deaths averted, PEP expansion remains highly cost-effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019] when considering direct costs only.
+However, our results suggest that expansion alone cannot prevent the majority of deaths and improved outreach and education on the risk of rabies exposure, better surveillance and active case management, and ultimately control in the domestic dog population will be necessary to manage the burden of rabies in Madagascar.
+
+Ultimately even given maximal access, achieving 'the last mile', preventing deaths due to rabies in the most remote and vulnerable populations, will require a disproportionate level of resources [@delriovilasTribulationsLastMile2017].
 IBCM programs and risked based assessments are one way to tackle the issue of managing PEP resources effectively even as demand outpaces reductions in burden [@rysavaPathRabiesElimination2018].
@@
 
-## Conclusion {inline these stats}
+## Conclusion
 Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths.
 In the absence of data on rabies incidence and risk of exposure, targeting expansion of PEP at areas currently underserved may be the best way to reduce rabies burden and provide equitable access to PEP.
-Specifically, choosing clinics serving populations that under the status quo would have to travel greater than 3 hours to seek PEP, or alternatively aiming for clinics to serve a catchment size of approximately 100,000 persons, are two ways that PEP access could be expanded strategically.
+Specifically, exanding provisioning clinics serving populations that under the status quo would have to travel greater than 3 hours to seek PEP, or alternatively aiming for clinics to serve a catchment size of approximately 100,000 persons, are two ways that PEP access could be expanded strategically.
 However, our results also suggest that increasing outreach and awareness of rabies risk, improving clinic functioning and provisioning practices, and removing socioeconomic barriers for persons seeking PEP will be critical to ensure the efficacy of PEP expansion.
@@
 
+We would like to thank all the clinicians and staff at the ARMC across the country.
+We are grateful to IPM and the Ministry of Public Health who work to collect and maintain data on PEP provisioning.
+In particular, on the data management side, we would like to thank the GIS unit for assistance with spatial data and Michael Luciano Tantely for sharing driving time data.
+We would like to thank Claire LeBlanc, Rila Ratovoson, and Daouda Kassie for sharing results of their work in the Moramanga District.
+In addition, we would like to thank Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, Amy Winter, Christian Ranaivoson, John Friar, and Amy Wesolowski for their assistance with data collection.
+This work was funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to CJEM and MR.
+MR is supported by an NSF Graduate Research Fellowship.
+KH is supported by the Wellcome Trust (207569/Z/17/Z).
 \newpage
@@
 ## S1. Estimating travel times to the closest ARMC {#S1}
-[*jump back to main results section: estimating travel times*](#results-ttimes)
-
-```{r fig S1.1, fig.fullwidth=TRUE}
+
+```{r fig S1.1 inputs A friction B pop}
 include_graphics(here("figs/supplementary/S1.1_inputs.jpeg"))
 ```
+
 *__[Figure S1.1.]{#figS1.1}__ Raster inputs to estimate travel times to the closest ARMC for Madagascar __A)__ Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale __B)__ Population estimates resampled to the same friction surface.*
 
-```{r fig S1.2, fig.fullwidth=TRUE}
+```{r fig S1.2 ttimes A Moramanga B Drive}
 include_graphics(here("figs/supplementary/S1.2_raw.jpeg"))
 ```
+
 *__[Figure S1.2.]{#figS1.2}__ Source of true travel time estimates __A)__ Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
@@
 
-```{r fig S1.3, fig.fullwidth=TRUE}
+```{r fig S1.3 travel modes}
 include_graphics(here("figs/supplementary/S1.3_modes.jpeg"))
 ```
+
 *__[Figure S1.3.]{#figS1.3}__ Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
 
-```{r fig S1.4, fig.fullwidth=TRUE}
+```{r fig S1.4 lms ttimes}
 include_graphics(here("figs/supplementary/S1.4_lms.jpeg"))
 ```
+
 *__[Figure S1.4.]{#figS1.4}__ Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) predicted by __A)__ Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) __B)__ Travel time estimates and C) Travel time estimates weighted by population (for commune means only).*
 
-*__[Table S1.1.]{#tabS1.1} R^2 values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
-```{r table S1.1}
+*__[Table S1.1.]{#tabS1.1}__ R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
+```{r table S1.1 lms table}
 S1.1_table <- read.csv(here("output/stats/access_met_r2.csv"), row.names = 1)
@@
 
-```{r fig S1.5, fig.fullwidth=TRUE}
+```{r fig S1.5 output ttimes A district B commune}
 include_graphics(here("figs/supplementary/S1.5_outputs.jpeg"))
 ```
+
 *__[Figure S1.5.]{#figS1.5}__ Estimates of mean travel times weighted by population at the __A)__ District __B)__ Commune scale.*
@@
 ## S2. Estimating average annual bite incidence at the district and commune level
-[*jump back to main results section: Bite patient data*](#results-biteinc)
-
-```{r fig S2.1, fig.fullwidth=TRUE}
+
+```{r fig S2.1 A catchments B closest C Prop bites}
 include_graphics(here("figs/supplementary/S2.1_catchments.jpeg"))
 ```
-*__[Figure S2.1]{#figS2.1} __A)__ Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or which a district (polygon outline).
+
+*__[Figure S2.1]{#figS2.1}A)__ Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or which a district (polygon outline).
 Admin units where the fill and border colors do not match show places where assigned catchments differ at the district vs commune scale.
@@
 
-```{r fig 2.2, fig.fullwidth=TRUE}
+```{r fig 2.2 form sub A time series B reporting C vial comp}
 include_graphics(here("figs/supplementary/S2.2_vialcomps.jpeg"))
 ```
+
 *__[Figure S2.2.]{#figS2.2}A)__ The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
@@
 
-[*jump back to main results section: Travel times as a predictor of bites*](#results-bitemods)
-
-```{r fig S3.1, fig.fullwidth=TRUE}
+```{r fig S3.1 corr pop ttimes}
 include_graphics(here("figs/supplementary/S3.1_corrs.jpeg"))
 ```
-_**Figure S3.1.** Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale._
+_**Figure S3.1.** Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale._\
 
@@
 For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
-```{r table S3.1}
+```{r table S3.1 dic}
 model_ests <- read.csv(here("output/mods/estimates.csv"))
@@
 
-```{r fig S3.2, fig.fullwidth=TRUE}
+```{r fig S3.2 obs vs predicted}
 include_graphics(here("figs/supplementary/S3.2_fitted.jpeg"))
 ```
+
 _**Figure S3.2.
@@
 
-```{r fig S3.3, fig.fullwidth=TRUE}
+```{r fig S3.3 out of fit}
 include_graphics(here("figs/supplementary/S3.3_outfit.jpeg"))
 ```
+
 _**Figure S3.3.
@@
 
-```{r fig S3.4, fig.fullwidth=TRUE}
+```{r fig S3.4 od comp posts}
 include_graphics(here("figs/supplementary/S3.4_posts_all.jpeg"))
 ```
+
 _**Figure S3.4.** Posterior estimates of parameters from models with travel time and population as an offset (flatPop in figures S3.2 & S3.3), comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter.
@@
 
-```{r fig S3.5, fig.fullwidth=TRUE}
+```{r fig S3.5 od comp preds}
 include_graphics(here("figs/supplementary/S3.5_preds_OD.jpeg"))
 ```
+
 _**Figure S3.5.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
@@
 
-```{r fig S3.6, fig.fullwidth=TRUE}
+```{r fig S3.6 random vs od}
 include_graphics(here("figs/supplementary/S3.6_posts_intercepts.jpeg"))
 ```
+
 _**Figure S3.6.** Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter._\
 
-```{r fig S3.7, fig.fullwidth=TRUE}
+```{r fig S3.7 random vs od}
 include_graphics(here("figs/supplementary/S3.7_preds_ODcomp.jpeg"))
 ```
+
 _**Figure S3.7.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons for random intercept model without overdispersion vs.
@@
 
-```{r fig S3.8, fig.fullwidth=TRUE}
+```{r fig S3.8 rep cutoff posts}
 include_graphics(here("figs/supplementary/S3.8_posts_rep_se.jpeg"))
 ```
+
 _**Figure S3.8.** Posterior estimates for models with population as an offset and an overdispersion parameter.
@@
 
-```{r fig S3.9, fig.fullwidth=TRUE}
+```{r fig S3.9 rep cutoff preds}
 include_graphics(here("figs/supplementary/S3.9_preds_rep_se.jpeg"))
@@
 
-_**Figure S3.5.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with population as an offset and an overdispersion parameter.
+_**Figure S3.9.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with population as an offset and an overdispersion parameter.
 Panels show predictions from models fitted to the national dataset 1) correcting for periods of at least 7 days where zero patient forms where submiited (a less stringent cutoff resulting in lower estimates of the proportion of forms submitted and thus higher estimates of reported bite incidence), (2) correcting for periods of 15 days where zero patient forms where submitted, as presented in the main analysis, and (3) with the raw data not correcting for undersubmission (resulting in lower estimates of reported bite incidence).
@@
 
-```{r fig S4.1, fig.fullwidth=TRUE}
+```{r fig S4.1 exp range}
 include_graphics(here("figs/supplementary/S4.1_exprange.jpeg"))
@@
 _**Figure S4.1.** Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis), assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
-The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (the red horizontal line).
+The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (where the red horizontal line and black dashed lines intersect).
 The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@leblancclaireRabiesMadagascarThreepronged2019].
@@
 ## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
-[*jump back to main methods section: Incremental reduction of burden*](#methods-incremental)  
-[*jump back to main results section: Incremental reduction of burden*](#results-incremental)
-
-```{r fig S5.1 add clinics, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.1_ARMCadded.jpeg"))
-```
-_**Figure S5.1.** Map of the location and at what step each clinic was added.
-The circles are each of the CSB II across the country sized by the proportion of the population for which they reduced travel times (excluding those already living within 3 hours of an ARMC) and the colors showing at what step each clinic was added (i.e.
-1 = the first clinic added).
-The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the remaining CSB II in the country (which reduced travel times for less than 0.01% of people living greater than 3 hours from a clinic and thus were not ranked)._
-
-```{r fig S5.2 ttime map shifts, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.2_map_shifts.tiff"))
-```
-_**Figure S5.2.** Maps of how travel times change as clinics are added at the **A)** ~1 x 1 km grid cell **B)** commune and **C)** district scales.
-The columns are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 723 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country).
+
+```{r fig S5.1 metrics}
+include_graphics(here("figs/supplementary/S5.1_tests.jpeg"))
+```
+_**Figure S5.1.** Comparing metrics for ranking clinics for targeted expansion to 83 additional districts (so that each district has an ARMC).
+We simulated expansion using three different ranking metrics: 1) reduction in mean travel times (green line) 2) the proportion of the population for which travel times were reduced (red dashed line) and 3) the proportion of the population for which travel times were reduced weighted by the reduction in travel times (pink dashed line).
+For each of these, we simulated burden using our decision tree framework (y axis is the mean of 1000 simulations of annual deaths at the national level).
+The blue lines show 10 simulations of randomly expanding access on reducing burden as a comparator.
+The panels show to the commune and district model of reported bite incidence.
+_
+
+```{r fig S5.2 add clinics}
+include_graphics(here("figs/supplementary/S5.2_ARMCadded.jpeg"))
+```
+_**Figure S5.2.** Map of the location and at what step each clinic was added.
+The circles are each of the CSB II across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
+The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional CSB II in the country which were added in the final step but not ranked._
+
+```{r fig S5.3 map shifts}
+include_graphics(here("figs/supplementary/S5.3_map_shifts.jpeg"))
+```
+_**Figure S5.3.** Maps of how travel times change as clinics are added at the **A)** ~1 x 1 km grid cell **B)** commune and **C)** district scales.
+The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional CSB IIs in the country).
 Grey pixels show the location of ARMC at each step.
@@
 
-```{r fig S5.3 prop pop ttimes, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.3_prop_ttimes.jpeg"))
-```
-_**Figure S5.4.** As clinics are added, the proportion of the population binned by travel times in hours.
-Panel labels are the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
-
-```{r fig S5.4 prop pop catchment, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.4_prop_catch.jpeg"))
-```
-_**Figure S5.3.** As clinics are added, shifts in distribution of the proportion of population in a given catchment  for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data).
-Rows are ordered by the number or clinics at each step: baseline (N = 31), 100, 200, 300, 472 (when algorithm stops adding clinics), and max (N = 1648 clinics, all CSB IIs in the country)._
-
-```{r fig S5.5 travel times(a) bite inc(b), fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.5_ttimes_shifts.jpeg"))
-```
-_**Figure S5.5.** Shifts in estimated **A)** travel times (hrs), **B)** bite incidence per 100,000 persons.
-The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading) for communes and districts  for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
-
-```{r fig S5.6 reporting(a) death inc(b), fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.6_reporting_shifts.jpeg"))
-```
-_**Figure S5.6.** Shifts in estimated **A)** Reporting, **B)** death incidence per 100,000 persons.
-The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading) for communes and districts for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
-
-```{r fig S5.7 catch(a) vials(b) throughput(c), fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.7_clinic_shifts.jpeg"))
-```
-_**Figure S5.7.** Shifts in clinic average **A)** catchment population size, **B)** and annual vial demand, and **C)**  daily throughput given estimates of bite incidence for the two models (colors, Commune scale model fitted to the Moramanga data and District scale model fitted to the National data).
-The left panels show the shift in distributions of these estimates as clinics are added, and the right panels show the changes in the averages (line) and the 95% quantiles (shading).
-Catchment population sizes are the same for each model._
-
-```{r fig S5.8 prop same admin, fig.fullwidth=TRUE}
-include_graphics(here("figs/supplementary/S5.8_prop_in_admin.jpeg"))
-```
-_**Figure S5.8.** Shifts in the distribution of the proportion of bites reported to a clinic that came from within the same A) commune or B) district as the ARMC was located for the two models (Commune scale model fitted to the Moramanga data and District scale model fitted to the National data)._
+```{r fig S5.4 travel times(a) bite inc(b)}
+include_graphics(here("figs/supplementary/S5.4_ttimes_bites.jpeg"))
+```
+_**Figure S5.4.** Shifts in estimated **A)** travel times (hrs, x-axis is square root transformed), **B)** bite incidence per 100,000 persons and boxplots showing the median for communes and district models (colors) as clinics are added._
+
+```{r fig S5.5 reporting(a) death inc(b)}
+include_graphics(here("figs/supplementary/S5.5_reporting_shifts.jpeg"))
+```
+_**Figure S5.5.** Shifts in estimated **A)** reporting, **B)** death incidence per 100,000 persons for the commune and district models (colors) as clinics are added._
+
+```{r fig S5.6 catch(a) vials(b) throughput(c)}
+include_graphics(here("figs/supplementary/S5.6_clinic_shifts.jpeg"))
+```
+_**Figure S5.6.** Shifts in average clinic **A)** catchment population size, **B)** annual vial demand, and **C)** daily throughput (i.e.
+average number of patients reporting each day) given estimates of bite incidence for the commune and district models (colors).
+For vial demand estimation, catchment population sizes are the same for each model as these populations are allocated at the grid cell level (i.e.
+population in a grid cell is allocated to the clinic catchment it is closest to in terms of travel times regardless of district or commune).
+All x-axes are log transformed._
+
+```{r fig S5.7 comm network}
+include_graphics(here("figs/supplementary/S5.7_comm_map.jpeg"))
+```
+
+_**Figure S5.7.** Shifts in where bites are reported to as clinics are added for the commune model.
+The circles show the clinic locations for each scenario, with size propotional to the annual average bites reported to that clinic.
+Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
+The polygon shading shows the commune level reported bite incidence._
+
+```{r fig S5.8 dist network}
+include_graphics(here("figs/supplementary/S5.8_dist_map.jpeg"))
+```
+
+_**Figure S5.8.** Shifts in where bites are reported to as clinics are added for the district model.
+The circles show the clinic locations for each scenario, with size propotional to the annual average bites reported to that clinic.
+Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
+The polygon shading shows the district level reported bite incidence._
 
 ## S6. Sensitivity of burden estimates to parameter assumptions {#S6}
-
-[*jump back to main methods section: Incremental reduction of burden*](#methods-sensitivity)  
-[*or to main results section: Incremental reduction of burden*](#results-sensitivity)
 
@@
 
-| **Parameter**   | **Description**        | **Commune model (Moramanga data)** | **District model (National data)** | 
+| **Parameter**   | **Description**        | **Commune model** | **District model** | 
 |:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
-| $\beta_t$ | Travel time effect	| `r bites_se$Commune[par == "beta_t"]` | `r bites_se$Commune[par == "beta_t"]`
-| $\beta_0$ | Model intercept	| `r bites_se$Commune[par == "beta_0"]` | `r bites_se$Commune[par == "beta_0"]`
-| $\sigma_0$ | Standard deviation of intercept by catchment	| -- | `r bites_se$Commune[par == "sigma_0"]`
+| $\beta_t$ | Travel time effect	| `r bites_se[par == "beta_ttimes"]$Commune` | `r bites_se[par == "beta_ttimes"]$District`
+| $\beta_0$ | Model intercept	| `r bites_se[par == "beta_0"]$Commune` | `r bites_se[par == "beta_0"]$District`
+| $\sigma_0$ | Overdispersion	| `r bites_se[par == "sigma_e"]$Commune` | `r bites_se[par == "sigma_e"]$District`
 
@@
 |:----------------|:----------------------:|:----------------------------|:---------:|
-| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se$range[par == "exp_max"]`  	
-| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se$range[par == "p_rab_max"]` 	
-| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se$range[par == "rho_max"]`	 
-| $p_{death}$ | The probability of death given a rabies exposure^[1] | `r dtree_se$range[par == "p_death"]`	
-[1] 95% CI of estimate from @changaluchaNeedImproveAccess2018|
-
-```{r fig S6.1 base se, fig.fullwidth=TRUE}
+| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se[par == "exp_max"]$range`  	
+| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se[par == "p_rab_max"]$range` 	
+| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se[par == "rho_max"]$range`	 
+| $p_{death}$ | The probability of death given a rabies exposure* | `r dtree_se[par == "p_death"]$range`	
+_*95% CI of estimate from @changaluchaNeedImproveAccess2018 _
+
+```{r fig S6.1 base se}
 include_graphics(here("figs/supplementary/S6.1_base_se.jpeg"))
 ```
+
 _**Figure S6.1.** A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B)  estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
@@
 
-```{r fig S6.2 add se, fig.fullwidth=TRUE}
+```{r fig S6.2 add se}
 include_graphics(here("figs/supplementary/S6.2_addARMC_se.jpeg"))
 ```
-_**Figure S6.2.** A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+
+_**Figure S6.2.** A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
 See Table S6.1 & 2 for ranges used for each parameter._
 
-```{r fig S6.3 vial se, fig.fullwidth=TRUE}
+```{r fig S6.3 vial se}
 include_graphics(here("figs/supplementary/S6.3_vial_se.jpeg"))
 ```
-_**Figure S6.3.** A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+
+_**Figure S6.3.** A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors) for the two models (colors).
 See Table S6.1 & 2 for ranges used for each parameter._
 
-```{r fig S6.4 scaling factors, fig.fullwidth=TRUE}
+```{r fig S6.4 scaling factors}
 include_graphics(here("figs/supplementary/S6.4_scaling_rels.jpeg"))
 ```
+
 *__Figure S6.4.__ Range of constrained scaling factors generated for district and commune population size so that underlying rabies exposures either A) decreases with increasing population size or B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
@@
 
-```{r fig S6.5 scaling se, fig.fullwidth=TRUE}
+```{r fig S6.5 scaling se}
 include_graphics(here("figs/supplementary/S6.5_scaling_se.jpeg"))
 ```
+
 *__Figure S6.5.__ Predicted relationship between A) travel times and rabies death incidence per 100k persons and B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
@@
 
-## Supplementary Reference (R packages used) {#srefs}
-
-::: {#refs_R}
+## S7. Sotware citations (R Packages used) {#srefs}
+::: {#refs_supp}
 :::

### 2020-09-07 analysis/docs/manuscript.Rmd
some more reorganizing
- safe paths
- and here::here()

--- 2020-05-26 docs/manuscript.Rmd
@@
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
+bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
 bibliography_main: ["bib/refs_all.bib"]
@@
 library(data.table)
-library(rgdal)
+library(sf)
 library(glue)
@@
 mora_bites <- fread(here("output/bites/mora_bites.csv"))
-mada_communes <- readOGR(here("data/processed/shapefiles/mada_communes.shp"))
-mada_districts <- readOGR(here("data/processed/shapefiles/mada_districts.shp"))
+mada_communes <- st_read(here("data/processed/shapefiles/mada_communes.shp"))
+mada_districts <- st_read(here("data/processed/shapefiles/mada_districts.shp"))
 
@@
 missing_comms <- fread(here("output/stats/missing_comms.csv"))
-comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
+comms_no_clins <- nrow(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
 csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
@@
 A target to eliminate deaths due to canine rabies by the year 2030 has been set forth by global health partners, and in 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly expand PEP availability and reduce rabies burden.
-Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited spatially (N = 31 Anti-Rabies Medical Center or ARMC across the country).
+Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited (to just 31 clinics across the country).
 
@@
 We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes burden sub-nationally.
-We estimate an annual burden of `r round(sort(natl_base$deaths_mean)[1])` (95% CI: `r round(sort(natl_base$deaths_lower)[nrow(natl_base)])` - `r round(sort(natl_base$deaths_upper)[nrow(natl_base)])`), with current PEP provisioning averting ~ `r glue("{trans_perc(min(prop_averted))}%")` percent of deaths.
-Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`.
+We estimate an annual burden of `r round(natl_base$deaths_mean[1], -1)` (95% PI: `r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with current PEP provisioning averting an additional `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) deaths.
+Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`.
 However, our results suggest that even with all major health centers in the country provisioning PEP (N = `r max(natl_preds$scenario)` additional clinics), we could still see substantial mortality due to rabies.
@@
 Data on rabies incidence will be neccessary to refine these estimates and to better target interventions.
-While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed & improved clinic functioning & surveillance systems.
+While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed in addition to improved vaccine provisioning and surveillance systems.
 A GAVI investment can propel countries forward on this path combining these life saving interventions.
@@
 In this study, we model estimates of reported bite incidence as a function of travel times and estimates of rabies exposure incidence to extrapolate the burden of human rabies given the current availability of PEP across the country.
-We find that PEP currently averts a signficant number of deaths, but that burden remains high, particularly in areas with high travel times to a clinic.
+We find that PEP currently averts approximately `r round(natl_base$averted_mean[1], -2)` deaths, but that burden remains high, particularly in areas with high travel times to a clinic.
 Expanding PEP availability to more clinics could significantly reduce the current burden of rabies deaths, but our results suggest that expansion alone will not eliminate deaths due to rabies under current levels of reporting.
@@
 So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
+
 $$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
+
 where $\rho_{max}$ is the maximum reporting.
-We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018].
+We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018](canyoulinkthis).
 We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
@@
 We developed a framework to rank clinics by how much they could improve access for people currently underserved and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
-We simulate expansion first to each district (N = `r length(mada_districts)`)and then to each commune in the country for all communes with a clinic.
+We simulate expansion first to each district (N = `r nrow(mada_districts)`)and then to each commune in the country for all communes with a clinic.
 We select the CSB II (Centre de Sante Niveau II, major health centers with capacity for delivering vaccines) in the highest density grid cell of the administrative unit as candidates for expansion.
@@
 ttime_stats %>%
-  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
+  filter(scenario == 0, !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
   sum(prop_pop) -> prop_over3_base
 ttime_stats %>%
-  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
+  filter(scenario == 0, !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
   sum(prop_pop) -> prop_over6_base
@@
 We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A).
-After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).
+After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r nrow(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).
 
@@
 
-Overall, we estimate between `r round(sort(natl_base$deaths_mean)[1])` - `r round(sort(natl_base$deaths_mean)[nrow(natl_base)])` average annual deaths in Madagascar due to rabies depending on the model used (Table 2).
-We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(sort(natl_base$averted_mean)[1])` - `r round(sort(natl_base$averted_mean)[nrow(natl_base)])` deaths due to rabies are prevented through provisioning of PEP each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and  there is significant variation sub-nationally when burden is modeled at the district level, with even stronger patterns of increase in burden with travel times when modeling burden at the commune level (Figure 4).
-However, there is considerable uncertainty in these estimates, in particular given the district model of bite incidence.
+Overall, we estimate approximately `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)`(95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`).
+These results vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Figure 6A).
+We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)` deaths due to rabies are prevented through provisioning of PEP each year.
+In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally when deaths are modeled at the district and commune scale, with the least accessible communities having the highest burden of deaths (Figure 6B-C).
 
@@
 
-*__Figure 4.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
+*__Figure 6.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
 __A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model fit to the National data for all communes in a district.
@@
 We found that strategic expansion of access to these additional `r nrow(clin_per_dist)` clinics (i.e.
-1 clinic per district) reduced burden by `r round(min(1 - props$deaths_mean[props$scenario == one_per_dist])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == one_per_dist])*100, 2)`% (Fig 7A).
-With one clinic per commune (where a clinic was available, N = `r one_per_comm`), we see a further reduction between `r round(min(1 - props$deaths_mean[props$scenario == one_per_comm])*100, 2)` - `r round(max(1 - props$deaths_mean[props$scenario == one_per_comm])*100, 2)`.
+1 clinic per district) reduced burden by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 7A).
+With one clinic per commune (where a clinic was available, N = `r one_per_comm`), we see a further reduction `r round((1 - props$deaths_mean[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`%).
 However, reductions in burden saturate as more clinics are added (Fig S5.2).
-Even when all CSB II provision PEP in the country, our model still predicts `r round(min(1 - natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` - `r round(max(1 - natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` rabies deaths per annum, and average reporting of rabid exposures remains between `r round(min(1 - trends$reporting_mean[trends$scenario == max(trends$scenario)])*100, 2)` - `r round(max(1 - trends$deaths_mean[trends$scenario == max(trends$scenario)])*100, 2)`% (Fig S5.5).
+Even when all CSB II provision PEP in the country, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`) rabies deaths per annum, and average reporting of rabid exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5).
 This is largely due to the fact that as more clinics are added, reported bite incidence saturates (Fig S5.4), and instead there are shifts in which clinic bites are reported to (S5.7 & S5.8).
@@
 Vial demand also outpaces reductions in burden (Fig 7B), with more vials  needed per averted death as PEP access is expanded (Fig 7C).
-Our model predicts an increase from `r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == 0]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
-visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == 250]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` - `r round(max(1 - natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario)]), 2)` vials if all CSB IIs were to provision PEP.
-Importantly, clinic catchment populations and throughput decrease, as well, with clinics serving fewer patients per day (S5.6).
+Our model predicts an increase from `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)})")` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
+visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")` vials with 250 clinics, and approximately `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` vials if all CSB IIs were to provision PEP.
+Importantly, clinic catchment populations and throughput decrease, as well, with clinics seeing fewer patients per day (S5.6).
 
@@
 
-*__Figure 5.
+*__Figure 7.
 A)__ Decrease in deaths due to rabies, __B)__ increase in total # of vials as additional ARMC are added at the national level , and __C)__ increase in vials needed per death averted based on the two models of reported bites.
@@
 Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
-Under the status quo (N = 31 clinics), we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic per district (N = `r one_per_dist`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")` and to one clinic per commune (N = `r one_per_comm`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")`.
-
-However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all additional CSB II in the country (N = `r max(natl_preds$scenario)` additional clinics), between `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))} - {round(max(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))}")` deaths may still occur nationally.
+Under the status quo (N = 31 clinics), we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
+We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic per district (N = `r one_per_dist`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")` and to one clinic per commune (N = `r one_per_comm`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")`.
+
+However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all additional CSB II in the country (N = `r max(natl_preds$scenario)` additional clinics), approximately `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))}")` deaths may still occur nationally.
 This result is driven by two key factors.
 First, given the landscape of clinics and travel times across the country, despite expansion to all major health centers in the country, travel times remain high for a significant proportion of the population (Figure S5.4).
-Secondly, even for locations with low travel times given the range human rabies exposures between 15 - 76 per 100,000 persons, we estimate an average incidence of human rabies deaths of `r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` (Figure S5.5).
+Secondly, even for locations with low travel times given the range human rabies exposures between 15 - 76 per 100,000 persons, we estimate an average incidence of human rabies deaths of `r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` and reporting of approximately `r round(max(admin_base$reporting_mean), 2)*100`% rabies exposures (Figure S5.5).
 Importantly, for administrative units with similar travel times, we estimate a significant amount of variation that cannot be explained by travel times alone, suggesting other factors may be driving access and reporting of bite incidence.
@@
 people who had been in close proximity or had touched a person that had died from suspected rabies.
-Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 4 deaths in this district over a `r max(mora_bites$month)` month period.
+Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 4 deaths in this district over a `r max(mora_bites$nobs, na.rm = TRUE)` month period.
 Only one animal case was confirmed diagnostically however.

### 2020-09-07 analysis/src/01_gis/baseline.R

@@ -1,0 +1,164 @@
+# ------------------------------------------------------------------------------------------------ #
+#' Getting baseline travel time estimates and catchments @ grid cell level
+#' Given the 31 existing clinics in the country
+# ------------------------------------------------------------------------------------------------ #
+
+start <- Sys.time()
+
+# Set-up --------------------------------------------------------------------------------------
+library(sf)
+library(raster)
+library(gdistance)
+library(foreach)
+library(doParallel)
+library(iterators)
+library(data.table)
+library(dplyr)
+library(lubridate)
+library(rmapshaper)
+library(fasterize)
+library(here)
+
+# Source
+source("R/functions/ttime_functions.R")
+source("R/functions/out.session.R")
+
+# Load in GIS files
+mada_districts <- st_read("data-raw/raw/shapefiles/districts/mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp")
+mada_communes <- st_read("data-raw/raw/shapefiles/communes/mdg_admbnda_adm3_BNGRC_OCHA_20181031.shp")
+ctar_metadata <- read.csv("data-raw/raw/ctar_metadata.csv")
+
+# Fix up shapefiles ------------------------------------------------------
+# Get distcodes for both admin levels
+mada_districts %>%
+  mutate(distcode = substring(as.character(ADM2_PCODE), 1, 7)) -> mada_districts
+
+mada_districts %>%
+  group_by(distcode) %>%
+  summarize(geometry = st_union(geometry)) %>%
+  left_join(st_drop_geometry(mada_districts[6:nrow(mada_districts), ])) %>%
+  mutate(ADM2_EN = recode(ADM2_EN, 
+                          `6e Arrondissement` = "Antananarivo Renivohitra")) -> mada_districts
+
+mada_communes %>%
+  mutate(distcode = substring(as.character(ADM2_PCODE), 1, 7),
+         ADM2_EN = case_when(distcode == "MG11101" ~ "Antananarivo Renivohitra",
+                             TRUE ~ ADM2_EN)) -> mada_communes
+
+# Get the minimum ttimes for each clinic ------------------------------------------------------
+# Load in rasters
+pop1x1 <- raster("data-raw/out/rasters/wp_2015_1x1.tif")
+prop_pop <- pop1x1/sum(values(pop1x1), na.rm = TRUE)
+
[…]

### 2020-09-07 data-raw/src/01_process_rasters.R

@@ -1,7 +1,7 @@
 # ------------------------------------------------------------------------------------------------ #
-#' Process all raster files
-#' Get masked friction surface as input for travel time estimates 
-#' Create transition layer for gdistance functions
+#' Fetch raster files clipped to Madagascar for travel time analyses
+#' Friction surface and create transition layer
+#' Resample world pop to friction surface
 # ------------------------------------------------------------------------------------------------ #
 
 # Packages
@@ -10,12 +10,18 @@ library(malariaAtlas) # for friction surface
 library(raster) # for reading in rasters
 require(sf) # for reading in shapefiles
 library(gdistance) # for making transition object
-source("R/out.session.R")
+library(here) # for paths
+library(data.table) # for out session 
+source(here("R", "out.session.R"))
+source(here("R", "match_pop.R"))
 
 # Shapefile for masking to (from OCHA)
 mada_districts <- 
-  here("data-raw/inputs/shapefiles/districts/mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp") %>%
+  here("data-raw", "raw", "shapefiles", "districts", "mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp") %>%
   st_read() 
+
+# World Pop 2015 (Linnaird et al.)
+wp_2015 <- raster(here("data-raw", "raw", "rasters", "WorldPop", "MDG_ppp_2015_adj_v2.tif"))
 
 # Masked friction surface
 friction_masked <- 
@@ -23,12 +29,43 @@ friction_masked <-
     surface = "A global friction surface enumerating land-based travel speed for a nominal year 2015",
     shp = mada_districts
   )
-writeRaster(friction_masked, "data/processed/rasters/friction_mada_masked.tif", overwrite = TRUE)
+writeRaster(friction_masked, 
+            here("data-raw", "out", "rasters", "friction_mada_masked.tif"), 
+            overwrite = TRUE)
 
 # Masked transition surface (geocorrected  transition matrix (i.e., the graph))
 trans <- transition(friction_masked, function(x) 1/mean(x), 8) 
 trans_gc <- geoCorrection(trans)
-saveRDS(trans_gc, "data/processed/rasters/trans_gc_masked.rds")
+saveRDS(trans_gc, 
+        here("data-raw", "out", "rasters", "trans_gc_masked.rds"))
+
[…]

### 2020-09-07 data-raw/src/02_pop_to_pix.R

@@ -1,48 +1,0 @@
-# ------------------------------------------------------------------------------------------------ #
-#' Aggregating wp pop to friction surface
-#' Matching pop with NA friction value to nearest non-NA friction value
-#' and then aggregating to the friction surface
-# ------------------------------------------------------------------------------------------------ #
-
-start <- Sys.time()
-
-# Set-up
-library(here)
-library(raster)
-library(data.table)
-source("R/out.session.R")
-source("R/match_pop.R")
-
-# Load in frition surface and then convert to spatial pixels
-friction_masked <- raster(here("data-raw/outputs/rasters/friction_mada_masked.tif"))
-friction_pix <- as(friction_masked, "SpatialPixelsDataFrame") 
-friction_pix$match_id <- friction_pix@grid.index
-
-# WP 2015 original 
-wp_2015 <- raster(here("data-raw/inputs/rasters/WorldPop/MDG_ppp_2015_adj_v2.tif"))
-wp_2015_pix <- as(wp_2015, "SpatialPixelsDataFrame")
-
-# here minDimension which ranks intersections
-wp_2015_pix$match_id<- over(wp_2015_pix, friction_pix, minDimension = 0)$match_id
-match <- raster(wp_2015_pix["match_id"]) # transform back to raster
-
-# For ones that are missing find the nearest non NA cell id
-wp_2015_unmatched <- wp_2015_pix[is.na(wp_2015_pix$match_id), ]
-match <- match_nearest(unmatched_pix = wp_2015_unmatched, matched_raster = match, 
-                       max_adjacent = 100)
-
-# Summarize pop to this updated raster
-pop_dt <- data.table(pop = wp_2015[], friction_id = match[])
-pop_dt <- pop_dt[, .(pop = sum(pop, na.rm = TRUE)), by = "friction_id"][!is.na(friction_id)]
-friction_pix$pop <- pop_dt$pop[match(friction_pix$match_id, pop_dt$friction_id)]
-pop <- raster(friction_pix["pop"]) # transform back to raster
-
-# should be true (or minimal difference)
-sum(pop[], na.rm = TRUE) - sum(wp_2015[], na.rm = TRUE)
-
-writeRaster(pop, "data-raw/outputs/rasters/wp_2015_1x1.tif", overwrite = TRUE, 
-            options=c("COMPRESS=LZW"))
-
-# Save session info
-out.session(path = "data-raw/src/02_pop_to_pix.R", filename = "output/log_local.csv")
-

### 2020-09-07 data-raw/src/03_clinics.R

@@ -3,7 +3,7 @@
 # ------------------------------------------------------------------------------------------------ #
 
 start <- Sys.time()
-source("R/functions/out.session.R")
+source(here("R", "out.session.R"))
 
 # packages
 library(tidyverse)
@@ -12,10 +12,10 @@ library(sf)
 select <- dplyr::select
 
 # data
-mada_districts <- st_read("data/processed/shapefiles/mada_districts.shp")
-mada_communes <- st_read("data/processed/shapefiles/mada_communes.shp")
-ctar_metadata <- read.csv("data/raw/ctar_metadata.csv")
-csbs <- read.csv("data/raw/csbs.csv", stringsAsFactors = FALSE)
+mada_districts <- st_read(here("data-raw", "out", "shapefiles", "mada_districts.shp"))
+mada_communes <- st_read(here("data-raw", "out", "shapefiles", "mada_communes.shp"))
+ctar_metadata <- read.csv(here("data-raw", "raw", "ctar_metadata.csv"))
+csbs <- read.csv(here("data-raw", "raw", "ipm_data", "csbs.csv"), stringsAsFactors = FALSE)
 
 # get clinic commcodes & distcodes
 csbs <- st_as_sf(csbs, coords = c("xcoor", "ycoor"), 
@@ -45,10 +45,10 @@ csb2 %>%
          long = st_coordinates(.)[, "X"], 
          lat = st_coordinates(.)[, "Y"]) -> csb2
 
-write.csv(st_drop_geometry(csb2), "data/processed/clinics/csb2.csv", row.names = FALSE)
+write.csv(st_drop_geometry(csb2), here("data-raw", "out", "clinics", "csb2.csv"), row.names = FALSE)
 
 # Process additional scenarios -------------------------------------------------------------
-pop_1x1 <- raster("data/processed/rasters/wp_2015_1x1.tif")
+pop_1x1 <- raster(here("data-raw", "out", "rasters", "wp_2015_1x1.tif"))
 csb2$pop_dens <- extract(pop_1x1, csb2[, c("long", "lat")])
 
 # 1 per district
@@ -86,14 +86,15 @@ csb1 %>%
          lat = st_coordinates(.)[, "Y"]) %>%
   bind_rows(clinic_per_comm) ->  clinic_per_comm
 
-write.csv(st_drop_geometry(clinic_per_comm), "data/processed/clinics/clinic_per_comm.csv", 
[…]

### 2020-09-07 data-raw/src/04_run_baseline.R

@@ -1,12 +1,10 @@
 # ------------------------------------------------------------------------------------------------ #
-#' Getting baseline travel time estimates and catchments for grid cells; also summarizing
-#' to admin units 
-#' Given the 31 existing clinics in the country
+#' Cleaning up shapefiles & keeping only columns needed
 # ------------------------------------------------------------------------------------------------ #
 
 start <- Sys.time()
 
-# Set-up --------------------------------------------------------------------------------------
+# Set-up ------------------------------------------------------------
 library(sf)
 library(raster)
 library(gdistance)
@@ -18,18 +16,21 @@ library(dplyr)
 library(lubridate)
 library(rmapshaper)
 library(fasterize)
+library(here)
 
 # Source
-source("R/functions/ttime_functions.R")
-source("R/functions/out.session.R")
+source(here("R", "ttime_functions.R"))
+source(here("R", "out.session.R"))
 
 # Load in GIS files
-mada_districts <- st_read("data/raw/shapefiles/districts/mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp")
-mada_communes <- st_read("data/raw/shapefiles/communes/mdg_admbnda_adm3_BNGRC_OCHA_20181031.shp")
-ctar_metadata <- read.csv("data/raw/ctar_metadata.csv")
+mada_districts <- st_read(here("data-raw", "raw", "shapefiles", "districts", 
+                               "mdg_admbnda_adm2_BNGRC_OCHA_20181031.shp"))
+mada_communes <- st_read(here("data-raw", "raw", "shapefiles", "communes", 
+                              "mdg_admbnda_adm3_BNGRC_OCHA_20181031.shp"))
+ctar_metadata <- read.csv(here("data-raw", "raw", "ctar_metadata.csv"))
 
 # Fix up shapefiles ------------------------------------------------------
-# Get distcodes for both admin levels
+# Get distcodes for both admin levels & dissolve tana polygons
 mada_districts %>%
   mutate(distcode = substring(as.character(ADM2_PCODE), 1, 7)) -> mada_districts
 
@@ -45,120 +46,26 @@ mada_communes %>%
          ADM2_EN = case_when(distcode == "MG11101" ~ "Antananarivo Renivohitra",
                              TRUE ~ ADM2_EN)) -> mada_communes
 
[…]

### 2020-09-19 analysis/paper/manuscript.Rmd
fix names of functions

--- 2020-09-07 analysis/docs/manuscript.Rmd
@@
 ## data
-ctar_metadata <- fread(here("data/processed/clinics/ctar_metadata.csv"))
-district_bites <- fread(here("output/bites/district_bites.csv"))
-mora_bites <- fread(here("output/bites/mora_bites.csv"))
-mada_communes <- st_read(here("data/processed/shapefiles/mada_communes.shp"))
-mada_districts <- st_read(here("data/processed/shapefiles/mada_districts.shp"))
+ctar_metadata <- fread(here("data-raw/out/clinics/ctar_metadata.csv"))
+district_bites <- fread(here("analysis/out/bites/district_bites.csv"))
+mora_bites <- fread(here("analysis/out/bites/mora_bites.csv"))
+mada_communes <- st_read(here("analysis/out/shapefiles/mada_communes.shp"))
+mada_districts <- st_read(here("analysis/out/shapefiles/mada_districts.shp"))
 
 ## All preds (for abstract stats etc)
-natl_preds <- fread(here("output/preds/natl_preds.gz"))
+natl_preds <- fread(here("analysis/out/preds/natl_preds.gz"))
 natl_base <- natl_preds[scenario == 0]
@@
 
-admin_base <- fread(here("output/preds/admin_preds_base.gz"))
+admin_base <- fread(here("analysis/out/preds/admin_preds_base.gz"))
 
 ## candidate points
-csbs <- fread(here("data/processed/clinics/csb2.csv"))
-clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
-clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
-missing_comms <- fread(here("output/stats/missing_comms.csv"))
+csbs <- fread(here("data-raw/out/clinics/csb2.csv"))
+clin_per_comm <- fread(here("data-raw/out/clinics/clinic_per_comm.csv"))
+clin_per_dist <- fread(here("data-raw/out/clinics/clinic_per_dist.csv"))
+missing_comms <- fread(here("analysis/out/stats/missing_comms.csv"))
 comms_no_clins <- nrow(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
@@
   mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
-trends <- read.csv(here("output/stats/trends_adminpreds.csv"))
+trends <- read.csv(here("analysis/out/stats/trends_adminpreds.csv"))
 
@@
 ```{r fig 1}
-include_graphics(here("figs/main/M1_sankey.jpeg"))
+include_graphics(here("analysis/figs/main/M1_sankey.jpeg"))
 ```
@@
 ```{r ttime stats}
-ttime_stats <- read.csv(here("output/stats/prop_pop_ttimes.csv"))
+ttime_stats <- read.csv(here("analysis/out/stats/prop_pop_ttimes.csv"))
 # get prop of pop > 3 hrs away at baseline
@@
 ```{r fig M2}
-include_graphics(here("figs/main/M2_ttimes.jpeg"))
+include_graphics(here("analysis/figs/main/M2_ttimes.jpeg"))
 ```
@@
 ```{r fig M3 catchmaps}
-include_graphics(here("figs/main/M3_bitemaps.jpeg"))
+include_graphics(here("analysis/figs/main/M3_bitemaps.jpeg"))
 ```
@@
 ```{r fig M4 biteinc}
-include_graphics(here("figs/main/M4_biteinc.jpeg"))
+include_graphics(here("analysis/figs/main/M4_biteinc.jpeg"))
 ```
@@
 ```{r}
-outliers <- read.csv(here("output/stats/outliers.csv"))
+outliers <- read.csv(here("analysis/out/stats/outliers.csv"))
 ```
@@
 ```{r}
-convergence <- read.csv(here("output/stats/convergence.csv"))
+convergence <- read.csv(here("analysis/out/stats/convergence.csv"))
 ```
@@
 ```{r fig 5}
-include_graphics(here("figs/main/M5_mods.jpeg"))
+include_graphics(here("analysis/figs/main/M5_mods.jpeg"))
 ```
@@
 ```{r fig 6}
-include_graphics(here("figs/main/M6_burden_base.jpeg"))
+include_graphics(here("analysis/figs/main/M6_burden_base.jpeg"))
 ```
@@
 ```{r fig 7}
-include_graphics(here("figs/main/M7_addARMC.jpeg"))
+include_graphics(here("analysis/figs/main/M7_addARMC.jpeg"))
 ```
@@
             inc = bites/pop*1e5) -> aa_bites
-mora_raw <- read.csv(here("data/processed/bitedata/moramanga.csv"))
+mora_raw <- read.csv(here("data-raw/out/bitedata/moramanga.csv"))
 mora_raw %>%
@@
   summarize(prop_cat1 = sum(known_cat1)/n()) -> prop_cat1_AA
-admin_preds <- fread(here("output/preds/admin_preds_base.gz")) 
+admin_preds <- fread(here("analysis/out/preds/admin_preds_base.gz")) 
 admin_preds %>%
@@
 
-ranked_clins <- fread(here("output/ttimes/addclinics/clinics_ranked.csv"))
+ranked_clins <- fread(here("analysis/out/ttimes/addclinics/clinics_ranked.csv"))
 aa_added <- min(ranked_clins[district == "Anosibe-An'ala"]$scenario)
-gtruth <- fread(here("output/ttimes/gtruth_ttimes.csv"))
+gtruth <- fread(here("analysis/out/ttimes/gtruth_ttimes.csv"))
 gtruth %>%
@@
 ```{r}
-catch_trends <- fread(here("output/stats/trend_catches.csv"))
+catch_trends <- fread(here("analysis/out/stats/trend_catches.csv"))
 catch_trends %>% 
@@
 ```{r fig S1.1 inputs A friction B pop}
-include_graphics(here("figs/supplementary/S1.1_inputs.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S1.1_inputs.jpeg"))
 ```
@@
 ```{r fig S1.2 ttimes A Moramanga B Drive}
-include_graphics(here("figs/supplementary/S1.2_raw.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S1.2_raw.jpeg"))
 ```
@@
 ```{r fig S1.3 travel modes}
-include_graphics(here("figs/supplementary/S1.3_modes.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S1.3_modes.jpeg"))
 ```
@@
 ```{r fig S1.4 lms ttimes}
-include_graphics(here("figs/supplementary/S1.4_lms.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S1.4_lms.jpeg"))
 ```
@@
 ```{r table S1.1 lms table}
-S1.1_table <- read.csv(here("output/stats/access_met_r2.csv"), row.names = 1)
+S1.1_table <- read.csv(here("analysis/out/stats/access_met_r2.csv"), row.names = 1)
 kable(S1.1_table, col.names = c("Predictor", "Driving times", 
@@
 ```{r fig S1.5 output ttimes A district B commune}
-include_graphics(here("figs/supplementary/S1.5_outputs.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S1.5_outputs.jpeg"))
 ```
@@
 ```{r fig S2.1 A catchments B closest C Prop bites}
-include_graphics(here("figs/supplementary/S2.1_catchments.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S2.1_catchments.jpeg"))
 ```
@@
 ```{r fig 2.2 form sub A time series B reporting C vial comp}
-include_graphics(here("figs/supplementary/S2.2_vialcomps.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S2.2_vialcomps.jpeg"))
 ```
@@
 ```{r fig S3.1 corr pop ttimes}
-include_graphics(here("figs/supplementary/S3.1_corrs.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.1_corrs.jpeg"))
 ```
@@
 ```{r table S3.1 dic}
-model_ests <- read.csv(here("output/mods/estimates.csv"))
+model_ests <- read.csv(here("analysis/out/mods/estimates.csv"))
 model_ests %>%
@@
 ```{r fig S3.2 obs vs predicted}
-include_graphics(here("figs/supplementary/S3.2_fitted.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.2_fitted.jpeg"))
 ```
@@
 ```{r fig S3.3 out of fit}
-include_graphics(here("figs/supplementary/S3.3_outfit.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.3_outfit.jpeg"))
 ```
@@
 ```{r fig S3.4 od comp posts}
-include_graphics(here("figs/supplementary/S3.4_posts_all.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.4_posts_all.jpeg"))
 ```
@@
 ```{r fig S3.5 od comp preds}
-include_graphics(here("figs/supplementary/S3.5_preds_OD.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.5_preds_OD.jpeg"))
 ```
@@
 ```{r fig S3.6 random vs od}
-include_graphics(here("figs/supplementary/S3.6_posts_intercepts.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.6_posts_intercepts.jpeg"))
 ```
@@
 ```{r fig S3.7 random vs od}
-include_graphics(here("figs/supplementary/S3.7_preds_ODcomp.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.7_preds_ODcomp.jpeg"))
 ```
@@
 ```{r fig S3.8 rep cutoff posts}
-include_graphics(here("figs/supplementary/S3.8_posts_rep_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.8_posts_rep_se.jpeg"))
 ```
@@
 ```{r fig S3.9 rep cutoff preds}
-include_graphics(here("figs/supplementary/S3.9_preds_rep_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S3.9_preds_rep_se.jpeg"))
 ```
@@
 ```{r fig S4.1 exp range}
-include_graphics(here("figs/supplementary/S4.1_exprange.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S4.1_exprange.jpeg"))
 ```
@@
 ```{r fig S5.1 metrics}
-include_graphics(here("figs/supplementary/S5.1_tests.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.1_tests.jpeg"))
 ```
@@
 ```{r fig S5.2 add clinics}
-include_graphics(here("figs/supplementary/S5.2_ARMCadded.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.2_ARMCadded.jpeg"))
 ```
@@
 ```{r fig S5.3 map shifts}
-include_graphics(here("figs/supplementary/S5.3_map_shifts.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.3_map_shifts.jpeg"))
 ```
@@
 ```{r fig S5.4 travel times(a) bite inc(b)}
-include_graphics(here("figs/supplementary/S5.4_ttimes_bites.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.4_ttimes_bites.jpeg"))
 ```
@@
 ```{r fig S5.5 reporting(a) death inc(b)}
-include_graphics(here("figs/supplementary/S5.5_reporting_shifts.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.5_reporting_shifts.jpeg"))
 ```
@@
 ```{r fig S5.6 catch(a) vials(b) throughput(c)}
-include_graphics(here("figs/supplementary/S5.6_clinic_shifts.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.6_clinic_shifts.jpeg"))
 ```
@@
 ```{r fig S5.7 comm network}
-include_graphics(here("figs/supplementary/S5.7_comm_map.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.7_comm_map.jpeg"))
 ```
@@
 ```{r fig S5.8 dist network}
-include_graphics(here("figs/supplementary/S5.8_dist_map.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S5.8_dist_map.jpeg"))
 ```
@@
 ```{r se tables}
-se_pars <- read.csv(here("output/sensitivity/se_pars.csv"))
+se_pars <- read.csv(here("analysis/out/sensitivity/se_pars.csv"))
 
@@
 ```{r fig S6.1 base se}
-include_graphics(here("figs/supplementary/S6.1_base_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S6.1_base_se.jpeg"))
 ```
@@
 ```{r fig S6.2 add se}
-include_graphics(here("figs/supplementary/S6.2_addARMC_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S6.2_addARMC_se.jpeg"))
 ```
@@
 ```{r fig S6.3 vial se}
-include_graphics(here("figs/supplementary/S6.3_vial_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S6.3_vial_se.jpeg"))
 ```
@@
 ```{r fig S6.4 scaling factors}
-include_graphics(here("figs/supplementary/S6.4_scaling_rels.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S6.4_scaling_rels.jpeg"))
 ```
@@
 ```{r fig S6.5 scaling se}
-include_graphics(here("figs/supplementary/S6.5_scaling_se.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S6.5_scaling_se.jpeg"))
 ```

### 2020-09-19 R/batch_functions.R

@@ -28,8 +28,8 @@ run_scenarios <- function(lookup, ttimes_dir = "analysis/out/ttimes/",
     j = iter(lookup, by = "row"), .combine = multicomb,
     .packages = c("data.table", "foreach", "triangle", "glue"), .options.RNG = rng_seed,
     .export = c(
-      "get.samps", "predict.bites", "predict.deaths", "summarize_mats",
-      "constrained_inc", "get.vials"
+      "get_samps", "predict_bites", "predict_deaths", "summarize_mats",
+      "constrained_inc", "get_vials"
     )
   ) %dorng% { # have to export funcs when in function
 
@@ -46,7 +46,7 @@ run_scenarios <- function(lookup, ttimes_dir = "analysis/out/ttimes/",
 
     # first do bite preds
     if (par_type == "posterior") {
-      posts <- as.data.frame(get.samps(
+      posts <- as.data.frame(get_samps(
         pop_predict = j$pop_predict,
         data_source = j$data_source,
         intercept = j$intercept,
@@ -57,7 +57,7 @@ run_scenarios <- function(lookup, ttimes_dir = "analysis/out/ttimes/",
       posts <- j
     }
 
-    bite_mat <- predict.bites(
+    bite_mat <- predict_bites(
       ttimes = ttimes, pop = comm$pop_admin,
       catch = comm$catchment, names = comm$commcode,
       beta_ttimes = posts$beta_ttimes, beta_0 = posts$beta_0,
@@ -86,7 +86,7 @@ run_scenarios <- function(lookup, ttimes_dir = "analysis/out/ttimes/",
         inc_scaled <- NULL
       }
 
-      all_mats <- predict.deaths(bite_mat,
+      all_mats <- predict_deaths(bite_mat,
         pop = comm$pop_admin,
         p_rab_min = j$p_rab_min, p_rab_max = j$p_rab_max,
         rho_max = j$rho_max, exp_min = j$exp_min, exp_max = j$exp_max,
@@ -145,7 +145,7 @@ run_scenarios <- function(lookup, ttimes_dir = "analysis/out/ttimes/",
 
       catch_mat <- as.matrix(bites_by_catch[, .SD, .SDcols = cols])
 
-      vial_preds <- sapply(catch_mat, get.vials)
+      vial_preds <- sapply(catch_mat, get_vials)
       vials <- matrix(unlist(vial_preds["vials", ]),
         nrow = nrow(catch_mat),
         ncol = ncol(catch_mat)

### 2020-09-19 R/data_functions.R

@@ -1,6 +1,6 @@
 
 # Function for getting days to include via RLEs
-rle.days <- function(vec, threshold = 10) {
+rle_days <- function(vec, threshold = 10) {
   rle(vec) %>%
     unclass() %>%
     as.data.frame() %>%
@@ -58,12 +58,12 @@ clean_natl <- function(national, mada_districts, mada_communes, ctar_metadata,
       fill = list(no_patients = 0)
     ) -> throughput
 
-  # rle.days = Helper function for getting which days to include (moved to functions from data_functions.R)
+  # rle_days = Helper function for getting which days to include (moved to functions from data_functions.R)
   throughput %>%
     group_by(id_ctar) %>%
     arrange(date_reported, .by_group = TRUE) %>%
     mutate(
-      include_day = rle.days(no_patients, threshold = tput_thresh),
+      include_day = rle_days(no_patients, threshold = tput_thresh),
       mean_throughput = mean(no_patients[include_day == 1]),
       sd_throughput = sd(no_patients[include_day == 1]),
       year = year(date_reported)

### 2020-09-19 R/estimate_pars.R

@@ -17,7 +17,7 @@
 #' @param thinning number of iterations to thin at (i.e. save every n steps of each chain)
 #' @param dic whether to estimate the model dic
 #' @param save whether to save the mcmc chains
-#' @param pass_priors a list of priors to pass, uses the \code[pass.priors] functions to gsub out old
+#' @param pass_priors a list of priors to pass, uses the \code[pass_priors] functions to gsub out old
 #' priors
 #' @param seed seed to use for the jags model (if NULL will not be reproducible)
 #' @param suffix another label to pass when saving the object (i.e. to identify the saved files)
@@ -26,7 +26,7 @@
 #'     rjags, foreach
 #'
 
-estimate.pars <- function(data_df, covar_df, model = "random", pars,
+estimate_pars <- function(data_df, covar_df, model = "random", pars,
                           trans = 1e5, chains = 3, adapt = 500, burn = 100,
                           iter = 10000, thinning = 5,
                           dic = TRUE, save = FALSE, pass_priors = NULL,
@@ -51,17 +51,17 @@ estimate.pars <- function(data_df, covar_df, model = "random", pars,
   ), envir = environment())
 
   if (!is.null(pass_priors)) {
-    model <- pass.priors(prior_list = pass_priors, uninformed = "dnorm(0, 10^-3)", model)
+    model <- pass_priors(prior_list = pass_priors, uninformed = "dnorm(0, 10^-3)", model)
   }
 
   cat(model)
 
   if (!is.null(seed)) {
-    mcmc.combine <- function(...) {
+    mcmc_combine <- function(...) {
       return(as.mcmc.list(sapply(list(...), mcmc)))
     }
 
-    foreach(i = 1:chains, .combine = mcmc.combine, .multicombine = TRUE) %do% {
+    foreach(i = 1:chains, .combine = mcmc_combine, .multicombine = TRUE) %do% {
       inits <- list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = seed + i)
       jags_mod <- jags.model(textConnection(model),
         data = data, inits = inits,
@@ -119,7 +119,7 @@ estimate.pars <- function(data_df, covar_df, model = "random", pars,
 #' @section Dependencies:
 #'     glue
 #'
-pass.priors <- function(prior_list, uninformed = "dnorm(0, 10^-3)", model) {
+pass_priors <- function(prior_list, uninformed = "dnorm(0, 10^-3)", model) {
[…]

### 2020-09-19 R/predict_functions.R

@@ -18,7 +18,7 @@
 #' (pred_type = "bites") from poisson given expecatation
 #' @param pred_type If par_type = "point_est" uses point estimate of parameters.
 #' If par_type = "posterior" need to pass vectors (or matrix if passing known_alphas) with length
-#' equal to nsims (use \code[get.samps])
+#' equal to nsims (use \code[get_samps])
 #' for this.
 #' @param error_n set to 1 when generating expectated range of preds for a given covariate value,
 #' defaults to length of the data when predicting for a set of locations
@@ -26,7 +26,7 @@
 #' @return Matrix of simulated predictions (nrow = length of data, ncol = nsims)
 #' @section Dependencies: foreach
 
-predict.bites <- function(ttimes, pop, catch, names,
+predict_bites <- function(ttimes, pop, catch, names,
                           beta_ttimes, beta_0, beta_pop, sigma_0, known_alphas, sigma_e,
                           pop_predict = c("flatPop", "addPop", "onlyPop"),
                           intercept = c("random", "fixed"), trans = 1e5, known_catch = TRUE,
@@ -109,7 +109,7 @@ predict.bites <- function(ttimes, pop, catch, names,
 }
 
 #' Predict deaths from reported bite incidence
-#' Takes matrix of predicted reported bites from \code[predict.bites] and applies decision tree
+#' Takes matrix of predicted reported bites from \code[predict_bites] and applies decision tree
 #' framework to simulate reporting, deaths, and deaths averted.
 #' @param p_rab_min minimum estimate of proportion of reported bites that are rabies exposures
 #' @param p_rab_max maximum estimate of proportion of reported bites that are rabies exposures
@@ -131,7 +131,7 @@ predict.bites <- function(ttimes, pop, catch, names,
 #' to generate those predictions.
 #' @section Dependencies: triangle
 
-predict.deaths <- function(bite_mat, pop, p_rab_min = 0.2, p_rab_max = 0.6, p_rab_mode = NULL,
+predict_deaths <- function(bite_mat, pop, p_rab_min = 0.2, p_rab_max = 0.6, p_rab_mode = NULL,
                            exp_min = 15 / 1e5, exp_max = 76 / 1e5, exp_mode = NULL, rho_max = 0.9,
                            prob_death = 0.16, dist = "uniform", exp_scaled = NULL, ...) {
 
@@ -224,7 +224,7 @@ constrained_inc <- function(slope, pop, max, min) {
 #' @param x numeric number of bites reported in a given year
[…]

### 2020-09-19 R/ttime_functions.R

@@ -1,5 +1,5 @@
 # 1. Get travel times -----------------------------------------------------------------------------
-#' \code{get.ttimes} calculates the minimum travel times/distance for an input raster to
+#' \code{get_ttimes} calculates the minimum travel times/distance for an input raster to
 #' an input set of
 #' GPS points.
 #' This function uses the friction surface from the Malaria Atlas Project. Script adapted
@@ -18,7 +18,7 @@
 #' @section Dependencies:
 #'  Packages: gdistance, raster, sf, sp
 
-get.ttimes <- function(friction, shapefile, coords, trans_matrix_exists = TRUE,
+get_ttimes <- function(friction, shapefile, coords, trans_matrix_exists = TRUE,
                        filename_trans) {
 
   # crop friction surface to shapefile
@@ -49,7 +49,7 @@ get.ttimes <- function(friction, shapefile, coords, trans_matrix_exists = TRUE,
 
 # 2. Getting ranked clinics and district/commune covariates  ----------------------------------
 #' Rank clinics and summarize travel times and catchments at admin level
-#' \code{add.armc} gets ranked ARMC and associated travel times and catchments at the district and
+#' \code{add_armc} gets ranked ARMC and associated travel times and catchments at the district and
 #' commune levels using catchment matrix of travel times at each grid cell (rows) for each clinic
 #' (columns) to rank which order clinics should be added based on how adding them shifts the
 #' distribution of travel times at the population level.
@@ -81,7 +81,7 @@ get.ttimes <- function(friction, shapefile, coords, trans_matrix_exists = TRUE,
 #'     Packages: data.table, foreach
 
 # Pass through base df, otherwise you need all the vectors!
-add.armc <- function(base_df, cand_df, max_clinics, thresh_ttimes, dir_name, rank_metric,
+add_armc <- function(base_df, cand_df, max_clinics, thresh_ttimes, dir_name, rank_metric,
                      base_scenario = 0, thresh_met = 0, overwrite = TRUE, random = TRUE) {
 
   base <- copy(base_df) # so as not to modify global env
@@ -141,13 +141,13 @@ add.armc <- function(base_df, cand_df, max_clinics, thresh_ttimes, dir_name, ran
           (!is.na(new_ttimes) & is.na(ttimes)), clin_chosen$clinic_id, catchment)
       )]
 
[…]

### 2020-09-19 analysis/src/00_install.R

@@ -3,7 +3,7 @@
 # ------------------------------------------------------------------------------------------------ #
 
 # Get list of needed packages
-pkgs_used <- read.csv("output/logs/pkgs_used.csv", stringsAsFactors = FALSE)
+pkgs_used <- read.csv("analysis/out/logs/pkgs_used.csv", stringsAsFactors = FALSE)
 installed <- sapply(pkgs_used$pkgs, require, character.only = TRUE, quietly = TRUE)
 to_install <- pkgs_used$pkgs[!installed]
 

### 2020-09-19 analysis/src/01_ttimes/01_baseline.R

@@ -47,7 +47,7 @@ system.time({
     points = iter(point_mat_base, by = "row"),
     .packages = c("raster", "gdistance", "data.table")
   ) %dopar% {
-    ttimes <- get.ttimes(
+    ttimes <- get_ttimes(
       friction = friction_masked, shapefile = mada_districts,
       coords = points[, c("x", "y")], trans_matrix_exists = TRUE,
       filename_trans = "data-raw/out/rasters/trans_gc_masked.rds"
@@ -105,7 +105,7 @@ write_create(
 )
 
 # District
-district_df <- aggregate.admin(base_df = base_df, admin = "distcode", scenario = 0)
+district_df <- aggregate_admin(base_df = base_df, admin = "distcode", scenario = 0)
 district_maxcatch <- district_df[, .SD[prop_pop_catch == max(prop_pop_catch,
   na.rm = TRUE
 )],
@@ -123,7 +123,7 @@ write_create(
 )
 
 # Commune
-commune_df <- aggregate.admin(base_df = base_df, admin = "commcode", scenario = 0)
+commune_df <- aggregate_admin(base_df = base_df, admin = "commcode", scenario = 0)
 commune_maxcatch <- commune_df[, .SD[prop_pop_catch == max(prop_pop_catch,
   na.rm = TRUE
 )],

### 2020-09-19 analysis/src/02_bitedata/02_bite_incidence_se.R

@@ -29,16 +29,16 @@ national %>%
                                by = "day"), id_ctar,
            fill = list(no_patients = 0)) -> patient_ts
 
-# rle.days = Helper function for getting which days to include
+# rle_days = Helper function for getting which days to include
 patient_ts %>%
   group_by(id_ctar) %>%
   arrange(date_reported, .by_group = TRUE) %>%
-  mutate(include_nozeros = rle.days(no_patients, threshold = 0),
-         include_5 = rle.days(no_patients, threshold = 5),
-         include_10 = rle.days(no_patients, threshold = 10),
-         include_15 = rle.days(no_patients, threshold = 15),
-         include_30 = rle.days(no_patients, threshold = 30),
-         include_all = rle.days(no_patients, threshold = Inf),
+  mutate(include_nozeros = rle_days(no_patients, threshold = 0),
+         include_5 = rle_days(no_patients, threshold = 5),
+         include_10 = rle_days(no_patients, threshold = 10),
+         include_15 = rle_days(no_patients, threshold = 15),
+         include_30 = rle_days(no_patients, threshold = 30),
+         include_all = rle_days(no_patients, threshold = Inf),
          year = year(date_reported)) -> throughput
 throughput$ctar <- ctar_metadata$CTAR[match(throughput$id_ctar,
                                             ctar_metadata$id_ctar)]
@@ -74,24 +74,24 @@ patient_ts %>%
 
 # Do average of 3 doses per patient at days 0, 3, 7 over 4 year period
 # Get completion of subsequent doses by clinic
-get.vials.3 <- function(x) {
+get_vials.3 <- function(x) {
   day0 <- floor(runif(x, min = 1, max = 365*4))
   days <- tabulate(c(day0, day0 + 3, day0 + 7))
   return(sum(ceiling(days)/2))
 }
 
-get.vials.4 <- function(x) {
+get_vials.4 <- function(x) {
   day0 <- floor(runif(x, min = 1, max = 365*4))
   days <- tabulate(c(day0, day0 + 3, day0 + 7, day0 + 28))
   return(sum(ceiling(days)/2))
 }
 
 mean.vials.3 <- function(patients, n){
-  mean(replicate(n, get.vials.3(patients)))
+  mean(replicate(n, get_vials.3(patients)))
 }
 
 mean.vials.4 <- function(patients, n){
-  mean(replicate(n, get.vials.4(patients)))
+  mean(replicate(n, get_vials.4(patients)))
 }
 
 ctar_bites %>%

### 2020-09-19 analysis/src/04_addclinics/02_test_metrics.R

@@ -47,7 +47,7 @@ foreach(i = iter(metrics), .combine = rbind) %do% {
   dir <- fp(paste0("analysis/out/ttimes/tests/", i))
 
   # Rank clinics (need to check base_df not modified!)
-  add.armc(
+  add_armc(
     base_df = base_df, cand_df = clin_per_dist,
     max_clinics = nrow(clin_per_dist),
     rank_metric = rank_metric,

### 2020-09-19 analysis/src/04_addclinics/03_ttimes_added.R

@@ -35,7 +35,7 @@ clin_per_dist <- fread(here_safe("data-raw/out/clinics/clinic_per_dist.csv"))
 clin_per_dist[, candfile := fp(paste0("analysis/out/ttimes/candidates/clinic_",
                                       clinic_id, ".tif"))]
 
-base_df <- add.armc(
+base_df <- add_armc(
   base_df = base_df, cand_df = clin_per_dist,
   max_clinics = nrow(clin_per_dist),
   rank_metric = prop, thresh_met = 1e-4,
@@ -48,7 +48,7 @@ clin_per_comm <- fread(here_safe("data-raw/out/clinics/clinic_per_comm.csv"))
 clin_per_comm[, candfile := fp(paste0("analysis/out/ttimes/candidates/clinic_",
                                       clinic_id, ".tif"))]
 
-base_df <- add.armc(
+base_df <- add_armc(
   base_df = base_df, cand_df = clin_per_comm,
   max_clinics = nrow(clin_per_comm),
   rank_metric = prop,

### 2020-09-19 analysis/src/figures/M3_bitemaps.R

@@ -95,7 +95,7 @@ mada_districts %>%
   rename(ctar = catchment) -> mada_districts
 
 # Bezier curves
-bez_pts <- get.bezier.pts(
+bez_pts <- get_bezier_pts(
   from = data.frame(
     long = ctar_todist_lines$from_long,
     lat = ctar_todist_lines$from_lat
@@ -125,7 +125,7 @@ ctar_todist_bez %>%
     group = interaction(distcode, ctar)
   ) -> ctar_todist_bez
 
-leg_pts <- get.bezleg(
+leg_pts <- get_bezleg(
   bbox = st_bbox(mada_districts), n_pts = 11, size_vec = c(100, 400, 1600, 5000),
   min_size = 5, offset_long = 0.4, offset_lat = -1.5
 )
@@ -226,7 +226,7 @@ comm_pts$from_long <- ctar_metadata$long[ctar_metadata$CTAR == "Moramanga"]
 comm_pts$from_lat <- ctar_metadata$lat[ctar_metadata$CTAR == "Moramanga"]
 
 # Get beziers
-bez_pts <- get.bezier.pts(
+bez_pts <- get_bezier_pts(
   from = data.frame(long = comm_pts$from_long, lat = comm_pts$from_lat),
   to = data.frame(long = comm_pts$to_long, lat = comm_pts$to_lat),
   frac = 0.5, transform = function(x) sqrt(1 / x) * 0.1
@@ -244,7 +244,7 @@ comms_to_plot <- filter(mada_communes, commcode %in% comms_to_plot$commcode)
 mora_district <- filter(mada_districts, district == "Moramanga")
 
 # Get legend
-leg_pts <- get.bezleg(
+leg_pts <- get_bezleg(
   bbox = bounds,
   n_pts = 11, size_vec = c(100, 200, 400, 1600),
   min_size = 5, offset_long = -0.1, offset_lat = -0.65

### 2020-09-19 analysis/src/figures/M5_modpreds.R

@@ -70,8 +70,8 @@ ggplot(
   cowplot::theme_minimal_grid() -> mod_preds
 
 # Plot posterior ests ------------------------------------------------------------------
-all_samps_natl <- summarize.samps(parent_dir = "analysis/out/mods/samps/National/")
-all_samps_mora <- summarize.samps(parent_dir = "analysis/out/mods/samps/Moramanga/")
+all_samps_natl <- summarize_samps(parent_dir = "analysis/out/mods/samps/National/")
+all_samps_mora <- summarize_samps(parent_dir = "analysis/out/mods/samps/Moramanga/")
 all_samps_natl %>%
   filter(pop_predict == "flatPop", intercept == "fixed", OD == TRUE) %>%
   bind_rows(filter(all_samps_mora, pop_predict == "flatPop", OD == FALSE)) -> all_samps

### 2020-09-19 analysis/src/process_session.R

@@ -8,9 +8,9 @@ library(lubridate)
 library(RCurl)
 library(glue)
 
-# Do out.session here so any packages used in this script get included too!
-source("R/functions/out.session.R")
-out.session("R/process_session.R", filename = "output/log_local")
+# Do out_session here so any packages used in this script get included too!
+source(here::here("R", "utils.R"))
+out_session("R/process_session.R", filename = "analysis/out/log_local")
 
 # Find all .R & .Rmd files 
 files <- list.files(recursive = TRUE)
@@ -18,14 +18,14 @@ files <- files[grep(".R$|.Rmd$", files)]
 files <- files[grep("archive", files, invert = TRUE)] # excluding these that are archived
 
 # Pull in local log
-log_local <- read.csv("output/log_local.csv", stringsAsFactors = FALSE)
+log_local <- read.csv("analysis/out/log_local.csv", stringsAsFactors = FALSE)
 log_local <- as.data.frame(apply(log_local, 2, function (x) as.character(x)), 
                            stringsAsFactors = FALSE)
 
 # Pull down log from cluster
 # Make sure you're vpn'd in before doing this
 # system("rsync -rvlzt --update mrajeev@della.princeton.edu:~/MadaAccess/log_cluster.csv ~/Documents/Projects/MadaAccess/output/")
-log_cluster <- read.csv("output/log_cluster.csv", stringsAsFactors = FALSE)
+log_cluster <- read.csv("analysis/out/log_cluster.csv", stringsAsFactors = FALSE)
 log_cluster <- as.data.frame(apply(log_cluster, 2, function (x) as.character(x)),
                              stringsAsFactors = FALSE)
 log_cluster %>%
@@ -43,7 +43,7 @@ log_local %>%
   bind_rows(log_cluster) -> log_scripts
 
 # write log out of when scripts last ran
-write.csv(log_scripts, "output/logs/log_last_ran.csv", row.names = FALSE)
+write.csv(log_scripts, "analysis/out/logs/log_last_ran.csv", row.names = FALSE)
 
 log_scripts %>%
   pivot_longer(-(ran:running), names_to = "pkgs", values_to = "version") %>%
@@ -65,5 +65,5 @@ source_urls[!exists] <- github_lookup[match(names(github_lookup), names(source_u
 pkgs_used$source <- source_urls
 
 # Write table & bib out for citing in supplement 
[…]

### 2020-09-19 data-raw/src/01_rasters.R

@@ -80,4 +80,4 @@ write_create(pop, here_safe("data-raw/out/rasters/wp_2015_1x1.tif"),
 )
 
 # Saving session info
-out_session(logfile = "logs/data_raw.csv", start = start, ncores = 1)
+out_session(logfile = here_safe("logs/data_raw.csv"), start = start, ncores = 1)

### 2020-09-19 data-raw/src/04_match_names.R

@@ -28,7 +28,7 @@ peripheral$distcode <- paste0(
   substring(peripheral$district, 1, 1),
   substring(peripheral$district, 3, 8)
 )
-peripheral_comm_matches <- match.admin(
+peripheral_comm_matches <- match_admin(
   data_names = peripheral$commune,
   data_nest = peripheral$distcode,
   match_names = mada_communes$commune,
@@ -43,7 +43,7 @@ write_create(peripheral_comm_matches,
 
 # Match admin names in IPM data ---------------------------------------------------------------
 load("data-raw/raw/ipm_data/ipm.rda")
-ipm_dist_matches <- match.admin(
+ipm_dist_matches <- match_admin(
   data_names = IPM$fiv, data_nest = NULL,
   match_names = mada_districts$district, match_nest = NULL,
   match_method = "osa", nested = FALSE
@@ -60,7 +60,7 @@ names_matched$distcode <- mada_districts$distcode[match(
   tolower(mada_districts$district)
 )]
 IPM$distcode <- names_matched$distcode[match(tolower(IPM$fiv), names_matched$names_tomatch)]
-ipm_comm_matches <- match.admin(
+ipm_comm_matches <- match_admin(
   data_names = IPM$fir, data_nest = IPM$distcode,
   match_names = mada_communes$commune,
   match_nest = mada_communes$distcode,
@@ -78,7 +78,7 @@ moramanga$commune <- trimws(moramanga$commune, which = "right")
 moramanga$district <- sapply(strsplit(as.character(moramanga$Patient.Home), "\\, "), "[", 2)
 moramanga$district <- gsub(" \\(District\\)", "", moramanga$district)
 
-moramanga_dist_matches <- match.admin(
+moramanga_dist_matches <- match_admin(
   data_names = moramanga$district, data_nest = NULL,
   match_names = mada_districts$district, match_nest = NULL,
   match_method = "osa", nested = FALSE
@@ -99,7 +99,7 @@ moramanga$distcode <- moramanga_dist_matches$distcode[match(
   tolower(moramanga$district),
   moramanga_dist_matches$names_tomatch
 )]
-moramanga_comm_matches <- match.admin(
+moramanga_comm_matches <- match_admin(
   data_names = moramanga$commune,
   data_nest = moramanga$distcode,
   match_names = mada_communes$commune,
[…]

### 2020-09-19 data-raw/src/05_bitedata.R

@@ -186,4 +186,4 @@ write_create(moramanga_ttimes, here_safe("data-raw/out/bitedata/moramanga_ttimes
 )
 
 # Saving session info
-out_session(logfile = "logs/data_raw.csv", start = start, ncores = 1)
+out_session(logfile = here_safe("logs/data_raw.csv"), start = start, ncores = 1)

### 2020-10-02 analysis/paper/manuscript.Rmd

@@
 ---
-title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
+title: "How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
 author:
@@
       correspondence: true
-  - Helene Guis:
-      institute: [IPMepi, CIRAD]
+  - Hélène Guis:
+      institute: [IPMepi, cirad, astre, fofifa]
       correspondence: false
@@
       correspondence: false
+  - José Nely:
+      institute: [smeetn]
   - Jean-Michel Heraud:
@@
       correspondence: false
+      equal_contributor: "yes"
   - C.J.E. Metcalf:
@@
       correspondence: false
+      equal_contributor: "yes"
   - Katie Hampson:
@@
       correspondence: false
+      equal_contributor: "yes"
 institute:
@@
   - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
-  - OMS: World Health Organization, Madagascar
+  - cirad: CIRAD, UMR ASTRE, Antananarivo, Madagascar
+  - astre: ASTRE, Univ Montpellier, CIRAD, INRAE, Montpellier, France 
+  - fofifa: FOFIFA-DRZVP, Antananarivo, Madagascar6 World Health Organization, Madagascar
+  - smeetn: Service contre les Maladies Endémo-épidémiques et Tropicales  Négligées, Ministère de la Santé Publique, Antananarivo, Madagascar
+  - OMS: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
   - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
-  - Geneva: University of Geneva
+  - Geneva: Institute of Global Health, Faculty of Medicine, University of Geneva, Geneva, Switzerland
   - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
   - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - IC: Institut Pasteur de Cambodge
+  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut pasteur de Madagascar)
   - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
 date: ""
+nocite: '@*'
 link-citations: no
-nocite: '@*'
 output:
@@
       - --lua-filter=format/multiple-bibliographies.lua
-      - --reference-doc=format/Reference_Document.docx
+      - --reference-doc=format/reference_supplement.docx
   bookdown::html_document2:
@@
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
-bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
+#bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib", "bib/refs_supextra.bib"]
 bibliography_main: ["bib/refs_all.bib"]
 bibliography_supp: ["bib/refs_pkgs.bib"]
+bibliography_supextra: ["bib/refs_supextra.bib"]
 ---
@@
 library(knitr)
-library(tidyverse)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
 library(data.table)
@@
 library(magrittr)
-opts_chunk$set(echo = FALSE, warning = FALSE, out.width="80%")
+source(here("R", "utils.R"))
+
+opts_chunk$set(echo = FALSE, warning = FALSE, 
+                message=FALSE, out.width="80%")
 
@@
 
-## data
+# data
 ctar_metadata <- fread(here("data-raw/out/clinics/ctar_metadata.csv"))
@@
 
-## All preds (for abstract stats etc)
+# All preds (for abstract stats etc)
 natl_preds <- fread(here("analysis/out/preds/natl_preds.gz"))
@@
 
-## candidate points
+# candidate points
 csbs <- fread(here("data-raw/out/clinics/csb2.csv"))
@@
 clin_per_dist <- fread(here("data-raw/out/clinics/clinic_per_dist.csv"))
-missing_comms <- fread(here("analysis/out/stats/missing_comms.csv"))
-comms_no_clins <- nrow(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
+
+comms_no_csb2 <- nrow(clin_per_comm[type == "CSB1"])
+
+comms_no_clins <- 
+  nrow(mada_communes) - 
+  length(unique(c(clin_per_comm$commcode, 
+                  clin_per_dist$commcode, 
+                  ctar_metadata$commcode)))
+
 csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
-
+csb2_max <- sum(csbs_max$type == "CSB2")
+csb1_max <- sum(csbs_max$type == "CSB1")
 max_added <- max(natl_preds$scenario)
@@
 
-# ABSTRACT
+# Abstract
 
 ## Background
-For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries.
-A target to eliminate deaths due to canine rabies by the year 2030 has been set forth by global health partners, and in 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly expand PEP availability and reduce rabies burden.
-Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited (to just 31 clinics across the country).
+Post-exposure prophylaxis (PEP) is highly effective at preventing human rabies deaths, however access to PEP is limited in many rabies endemic countries.
+The 2018 decision by Gavi to add human rabies vaccine to its investment portfolio should expand PEP availability and reduce rabies deaths.
+We explore how geographic access to PEP impacts the rabies burden in Madagascar and the potential benefits of improved provisioning.
 
 ## Methodology & Principal Findings
-We use travel times to the closest ARMC as a proxy for access to PEP.
-We find that travel times are a strong predictor of reported bite incidence across the country.
-We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes burden sub-nationally.
-We estimate an annual burden of `r round(natl_base$deaths_mean[1], -1)` (95% PI: `r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with current PEP provisioning averting an additional `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) deaths.
-Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`.
-However, our results suggest that even with all major health centers in the country provisioning PEP (N = `r max(natl_preds$scenario)` additional clinics), we could still see substantial mortality due to rabies.
-Our estimates are most sensitive to assumptions of underlying rabies exposure incidence, but the qualitative patterns of the impacts of travel times and expanded access on burden are robust to these assumptions.
+We use travel times to the closest clinic providing PEP (N=31) as a proxy for access.
+We find that travel times strongly predict reported bite incidence across the country.
+Using resulting estimates in an adapted decision tree framework we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
+We estimate 960 human rabies deaths annually (95% Prediction Intervals (PI):`r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with PEP averting an additional 800 deaths (95% PI: `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) each year.
+Under these assumptions, we find that expanding PEP to one clinic per district could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major health centers provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
+Our quantitative estimates are most sensitive to assumptions of underlying rabies exposure incidence, but qualitative patterns of the impacts of travel times and expanded PEP access are robust.
 
 ## Conclusions & Significance
-PEP is effective at preventing deaths and in the absence of rabies data, targeting expansion to populations currently underserved may be the most equitable way to provision PEP.
-Data on rabies incidence will be neccessary to refine these estimates and to better target interventions.
-While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed in addition to improved vaccine provisioning and surveillance systems.
-A GAVI investment can propel countries forward on this path combining these life saving interventions.
-
-# AUTHOR SUMMARY 
-Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low and middle income countries where people have limited access to both human (post-exposure prophylaxis or PEP) and dog rabies vaccine.
-Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by the year 2030, and recently, GAVI, a organization that provides funding for immunization programs to eligible countries, added human rabies vaccine to it's funding agenda.
-In this study, we model estimates of reported bite incidence as a function of travel times and estimates of rabies exposure incidence to extrapolate the burden of human rabies given the current availability of PEP across the country.
-We find that PEP currently averts approximately `r round(natl_base$averted_mean[1], -2)` deaths, but that burden remains high, particularly in areas with high travel times to a clinic.
-Expanding PEP availability to more clinics could significantly reduce the current burden of rabies deaths, but our results suggest that expansion alone will not eliminate deaths due to rabies under current levels of reporting.
-Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other countries like it, forwards on the path to elimination.
-
-\newpage
-
-# INTRODUCTION
+PEP is effective at preventing rabies deaths, and in the absence of strong surveillance, targeting underserved populations may be the most equitable way to provision PEP.
+Our framework could be used to guide PEP expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted.
+While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyse countries' progress towards achieving zero rabies deaths.
+
+# Author Summary 
+Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low- and middle-income countries where people have limited access to both human vaccines (post-exposure prophylaxis or PEP) and dog rabies vaccines.
+Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by 2030, and recently, Gavi, a multilateral organization that aims to improve access to vaccines in the poorest countries, added human rabies vaccine to it’s portfolio.
+In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP, and extrapolate human rabies deaths in Madagascar.
+We find that PEP currently averts around `r round(natl_base$averted_mean[1], -2)` deaths each year, but that the burden remains high (`r round(natl_base$deaths_mean[1], -2)` deaths/ year), particularly in remote, hard-to-reach areas.
+We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone will not eliminate deaths.
+Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other Low- and Middle-Income countries, forward on the path to rabies elimination.
+
+\newpage
+
+# Introduction
 
@@
 
-Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Canine rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans.
-While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018].
-However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
-
-In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero].
+While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid animal [@RabiesVaccinesWHO2018].
+However, access to human vaccine is limited in many countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
+
+In 2015, a global framework to eliminate deaths due to canine rabies by 2030 ('Zero by 30') through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero].
 Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS].
-From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.
-
-A primary challenge in expanding access effectively is the lack of data on rabies exposures and deathss in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
-Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in the most vulnerable and remote communities [@taylor2017difficulties].
-Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites and a decision tree framework to extrapolate deaths, with the key assumption that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e.
-the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018].
+From 2021, Gavi-eligible countries should be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.
+
+A primary challenge in expanding access effectively is the lack of data on rabies exposures and deaths in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
+Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in remote  and marginalized communities [@taylor2017difficulties].
+Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites at clinics provisioning PEP and a decision tree framework to extrapolate deaths, assuming that overall reported bite incidence (i.e.
+both bites due to non-rabid and rabid animals) is proportional to rabies incidence (i.e.
+the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting to clinics for PEP is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2019].
 If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.
 
-In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country [@rajeevHealthcareUtilizationProvisioning2018].
-PEP is not available at any other public clinics or through the private sector.
-In addition, there is limited control of rabies in dog populations and the disease appears to be endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011].
-Due to the spatially restricted nature of PEP distribution points and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
-Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2018].
-Here, we revise estimates of burden by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
-
-# METHODS
-
-## Estimating travel time to closest ARMC {#methods-ttimes}
-To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B.
-
-Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP at an approximately 1 x 1 km scale.
-We then extracted the mean travel times as well as the mean of travel times weighted by population for the district and commune (the administrative unit below the district) level.
-We also compared these metrics to euclidean distance, taking the minimum distance from the administrative unit centroid to any ARMC.
-We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.
-
-We compared these travel time and distance estimates to driving time data collected by IPM during field missions and patient reported travel times from a subset of patients reporting to the Moramanga ARMC (see Fig S1.2 for raw data).
-
-## Estimating average annual bite incidence {#methods-bitedata}
+In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to it's own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019].
+Other than at these 31 anti-rabies medical centers (ARMC), PEP is not available at any other public clinics or through the private sector.
+In addition, there is limited control of rabies in dog populations and the disease is endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011].
+Due to the spatially restricted nature of PEP provisioning and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
+Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
+Here, we provide revised estimates of human rabies deaths by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
+This framework may usefully apply to other countries where PEP availability is currently geographically restricted in considering how to most effectively and equitably provision these life-saving vaccines.
+
+# Methods
+
+## Estimating geographic access to ARMC {#methods-ttimes}
+To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two rasters: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B), which we aggregated to the friction surface.
+
+From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018].
+We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and euclidean distance, i.e.
+the minimum distance from the administrative unit centroid to any ARMC.
+We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018).
+We compared travel times and distance estimates to driving times collected by IPM during field missions and patient reported travel times from a subset of Moramanga ARMC patients (see Fig S1.2 for raw data).
+
+## Estimating bite incidence {#methods-bitedata}
 We used two datasets on bite patients reporting to ARMC:
 
-1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
-These were individual patient data forms that were submitted from clinics as frequently as monthly to annually, which included data on the date of reporting and were resolved to the district level (i.e.
-the district where the patient resided).
-2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level.
-We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.
-
-The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.
-However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted.
-To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1)).
-We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic.
-Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.
-
-We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period.
-Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported.
-For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs.
-4 doses total.
-During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit.
-Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day.
-We did not have the date or month when vials were provisioned, so we compared estimated demand to the total vials provisioned across this four year period.
-
-At both the commune and district scale, we assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit.
-For the national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.
-
-## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
-We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
+1)	A national database of individual bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
+These forms were submitted with frequencies ranging from monthly to annually, included the patient reporting date and were resolved to the district level (patient residence).
+
+2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga District ARMC resolved to commune level.
+
+For the national data, some clinics did not submit any data, or had substantial periods (months to a whole year), with no submitted data.
+To correct for this, we exclude periods of 15 consecutive days with zero submitted records (see Supplementary Appendix, section S2).
+For each clinic we divided the total number of bites reported in a given year by the estimated proportion of forms which were not submitted (under-submission).
+Due to yearly variation in submissions, we took the average of annual bite incidence estimates aggregated to district level.
+We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see Supplementary Appendix, section S2).
+At both the commune and district level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit.
+For national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.
+
+## Modeling reported bite incidence {#methods-bitemods}
+
+We modeled the number of reported bites as a function of travel time ($T$) using a Poisson regression:
+
 $$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
 $$ y_i = Poisson(\mu_i) $$
-where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
-We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level.
-In order to more directly compare estimates between these two datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$) so that:
-
-$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0_j)}P_j$$
-
-As travel times are correlated with human population size (Figure S3.1), we also compared how well bites were predicted by human population size alone, and in combination with travel times.
-For the models with population size, we used a model framework with either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.
-
-For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma_0)$) where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept).
-The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.
-
-We tested whether the catchment random effect was capturing overdispersion in the data (i.e.
-variance greater than the mean--the expectation of variance given a poisson distribution) rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
-
-$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j + \sigma_e)}P_i$$
-where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters and for a given model.
-
-We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’.
-We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10^{-6})$).
-For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$).
-We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model [@gelman2006data].
-
-To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to the data.
-In addition, we used the models to predict the data that were not used to fit the model (i.e.
-we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data).
-Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e.
-assuming that data was completely reported resulting in lower estimatates of bite incidence) and with a lower cut-off for identifying periods of undersubmission (7 days vs.
-15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).
-
-## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}
-We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1).
-To model uncertainty in parameter estimates we used triangular distributions, as with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures.
-The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).
-
-For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons).
-The range for $E_i$ is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons.
-These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019].
-We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate).
-As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
-See figure S4 for further details.
-
-```{r fig 1}
-include_graphics(here("analysis/figs/main/M1_sankey.jpeg"))
-```
-
-*__Figure 1__.
-Decision tree for burden estimation.
-For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$).
+
+where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the ARMC as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
+We fit this model to both the national data (district level) and the Moramanga data (commune level).
+To more directly compare estimates between datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$):
+
+$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ \ + \ \beta_{0j})}P_j$$
+
+As travel times are correlated with population size (Fig S3.1), we also compared how well bites were predicted by population size alone, and in combination with travel times.
+For the models with population size, we removed the offset and used either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors.
+
+For the models fit to the national data, we also modeled variation between clinics with a catchment random effect: $B_{0,k} \sim norm(\mu,\ \sigma_0)$), where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept.
+
+We tested whether the catchment random effect captured overdispersion in the data (i.e.
+variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
+
+$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j \ + \ \sigma_e)}P_i$$
+
+where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters for a given model. 
+We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’ [@R-rjags].
+We used model estimates to generate fitted and out-of-fit predictions, and examined the sensitivity of estimates to adjustments for under-submission of forms (Supplementary Appendix, section S3).
+
+## Modeling human rabies deaths {#methods-burden}
+We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1).
+To model uncertainty in parameter estimates we used triangular distributions, as with previous studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual exposure incidence per administrative unit, and  $p_{rabid}$, the proportion of reported bites that are rabies exposures.
+
+```{r fig 1, out.width="100%"}
+include_graphics(here("data-raw/misc/figM1/figM1_final.tiff"))
+```
+
+### *Fig 1. Decision tree for burden estimation.* 
+*For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted reported bites ($B_i$).
 To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures.
-$p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$).
-$R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$).
-Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e.
+$p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting of patients to clinics cannot exceed a maximum ($p_{max}$).
+$R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{death}$) to get deaths due to rabies ($D_i$).
+Similarly, deaths averted by PEP, $A_i$, are estimated by multiplying $R_i$ by $p_{death}$, i.e.
 those who would have died given exposure, but instead received PEP.
-Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*
-
-$B_i$, the number of total reported bites, is the predicted number of bites from the models of bite incidence as a function of travel times.
-To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6.
-So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
+Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution.
+Parameter values and sources are in Table 1.*
+
+For $E_i$, we center the distribution at the lower end of our estimated exposure incidence from the Moramanga District (42 exposures/100,000 persons), with a range applied assuming 1% rabies incidence in dogs (estimated across a range of human-to-dog ratios between 5 - 25) and that on average a rabid dog exposes 0.39 persons [@whorabiesmodelingconsortiumPotentialImpactImproved2019] (see Fig S4.1).
+As there is little data on dog population size and human exposure incidence in Madagascar[@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
+
+For $p_{rabid}$, we use a range between 0.2 - 0.6 estimated from a study of bite patients in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].
+So that rabid reported bites cannot exceed the total expected number of rabies exposures or a maximum reporting (even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$:
 
@@
 
-where $\rho_{max}$ is the maximum reporting.
-We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018](canyoulinkthis).
-We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.
-
-We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$).
-Deaths due to rabies in each administrative unit ($D_i$) are then estimated by mutliplying $D_i$ times $p_{death}$.
-Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_ip_{death}$).
-Table 1 lists all parameter values and sources used in the decision tree.
-\
-
-Table:**Table 1.** Parameters used in the decision tree to estimate human rabies deaths at the administrative level.\
+where $\rho_{max}$ is the maximum reporting, estimated from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where we find that approximately 2% of rabies exposures go unreported [@rajeevHealthcareUtilizationProvisioning2019].
+
+We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing rabies.
+Fig 1 describes the decision tree, the key inputs, and outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).
+Table 1 list all inputs and their sources.
+
+```{r exps}
+exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
+rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
+```
+
+_**Table 1. Parameters used in the decision tree to estimate human rabies deaths at the administrative level.**_\
 
@@
 |:----------------|:----------------:|:----------------------------------|:---------:|
-| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see Fig S4.1|
-| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
-| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
+| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
+| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
+| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2019] |
+| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2019] |
 | $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|
 
-## Estimating the impact of expanding PEP provisioning to additional clinics {#methods-incremental}
-
-We developed a framework to rank clinics by how much they could improve access for people currently underserved and to estimate the incremental reduction in burden as PEP is expanded to these clinics.
-We simulate expansion first to each district (N = `r nrow(mada_districts)`)and then to each commune in the country for all communes with a clinic.
-We select the CSB II (Centre de Sante Niveau II, major health centers with capacity for delivering vaccines) in the highest density grid cell of the administrative unit as candidates for expansion.
-At the commune level, `r nrow(missing_comms) - comms_no_clins`  communes did not have a CSB II, so we chose the CSB I (Centre de Sante Niveau I, secondary health posts) in the highest density grid cell.
+## Estimating the impact of expanding PEP provisioning {#methods-incremental} 
+We developed a framework to rank clinics by how much their PEP provision improves access for underserved communities, estimating incremental reductions in burden and increases in vaccine demand.
+Specifically, we aggregated our model-predicted estimates of annual bites to the clinic level.
+As multiple clinics may serve a single district or commune, we allocated bites to clinics according to the proportion of the population in each administrative unit which were closest.
+For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
+patients report for all doses) to generate subsequent vaccination dates.
+We use these dates to estimate vial usage given routine vial sharing practices in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], but assuming adoption of the WHO-recommended abridged intradermal regimen (2 x 0.1 ml injections on days 0, 3, and 7 [@tarantolaIntradermalRabiesPostexposure2018]).
+For both burden and vial estimates, we take the mean of 1000 simulations as each clinic is added.
+
+We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic.
+We select the CSB II (Centre de Sante Niveau II, major health centers with immunization capacity) in the highest density grid cell of the administrative unit as candidates for expansion.
+For the `r comms_no_csb2` communes without a CSB II, we chose the CSB I (secondary health posts) in the highest density grid cell.
 `r comms_no_clins` communes lacked any CSB I or II.
-
-We first tested three different metrics for ranking clinics:
-
- 1. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced when a clinic is added.
- 2. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced weighted by the  magnitude of the change in travel times when a clinic is added.
- 3. The mean reduction in travel times for people living > 3 hours away from an existing ARMC when a clinic is added.
-
-For all three metrics, once any clinics added only reduced travel times below that threshold (i.e.
-> 3 hours away) for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons), we reduce the threshold by 25%.
-
-For each of these metrics, we sequentially add the clinic with the maximal metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially until each district had a clinic.
-At each step, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.
-We compared the reductions in burden using the decision tree framework described previously for these three strategies, and also to 10 different scenarios where clinics were added randomly.
-We then use the ranking metric that reduced burden the most effectively as clinics were added to expand iteratively to all communes with a clinic.
-Finally, we explore a final scenario where all additional CSB II in the country provision PEP in addition to the clinics added to each district and commune for a total of `r nrow(csbs_max)`.
-
-We also estimate how expansion to additional clinics may impact vaccine demand.
-Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level.
-As clinics are added, multiple clinics may serve a single district or commune.
-Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest.
-For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
-all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses.
-We use these dates to estimate vial usage given routine vial sharing practices in Madagascar as described previously, but under the assumption of adoption of the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]).
-We take the mean of 1000 simulations of throughput for each catchment and set of clinics.
+Finally, we explore a scenario where all additional CSB II (totalling `r nrow(csbs_max)`) provision PEP.
+
+We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from an existing ARMC for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing ARMC.
+We simulated expansion of ARMC to each district using these three metrics, and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
+For the full simulation of expanded access, once clinics reduced travel times for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` living greater than $x$ hrs away, starting with $x$ = 3 hrs), we reduced the travel time threshold by 25%.
 
 ## Sensitivity analysis {#methods-sensitivity}
-To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and the decision tree parameters.
-For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1).
-For parameters from the models of reported bite incidence, we fixed the estimate at the upper and lower 95% quantiles of their posterior distributions (Table S6.2).
-We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden.
-Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e.
-if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases.
-Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e.
-if dog ownership is more common in less populated, rural areas), we might expect rabies exposure incidence within our range to scale negatively with human population size.
-We use scaling factors to scale incidence either positively or negatively with the population sizes observed at the district and commune levels, while still constraining them within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4) and simulated baseline burden, as well as the impact of expanding PEP access.
+To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analysis of both parameters from the models of bite incidence and the decision tree (see Table S6.1 & S6.2 for parameter ranges used).
+We also examined how systematic variation in rabies incidence with human population size affected burden estimates.
+Specifically, if human-to-dog ratios are positively correlated with human populations (i.e.
+dog ownership/populations are higher in more populated, urban areas), we might expect higher rabies exposure incidence as population size increases.
+Alternatively, if human-to-dog ratios inversely correlate with population size (i.e.
+dog ownership is more common in less populated, rural areas), we might expect exposure incidence to scale negatively with population size.
+We use scaling factors to scale incidence either positively or negatively with observed population sizes at the district and commune levels, while constraining them to the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, Fig S4.2) and simulated baseline burden, as well as expanded PEP access.
 
 ## Data and analyses 
-All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (section S7).
-All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at github.com/mrajeev08/MadaAccess.
+All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (Supplementary appendix, section S7).
+All processed data, code, and outputs are archived (here)[zenodo] and maintained at github.com/mrajeev08/MadaAccess.
 The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC.
 These databases were last queried on September 19, 2019 for these analyses.
-Anonymized raw bite patient data are available from IPM following insitutional data transfer protocols.
+The IPM GIS unit provided the data on geolocated clinics across the country.
+Anonymized raw bite patient data and full data on geolocated clinics are available from IPM following insitutional data transfer protocols.
 Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (wisemonkeyfoundation.org) on the same date and are shared in the archived repository.
@@
 ## Ethics statement
-Data collection from the Moramanga District was approved by the Princeton University IRB (# 7801) and the Madagascar Ministry of Public Health Ethics Committee (# 105-MSANP/CE).
+Data collection from the Moramanga District was approved by the Princeton University IRB (7801) and the Madagascar Ministry of Public Health Ethics Committee (105-MSANP/CE).
 Oral informed consent was obtained from all interviewed participants.
@@
 
-# RESULTS
-
-## Estimating travel times to the closest ARMC {#results-ttimes}
+# Results
+
+## Estimating access to ARMC {#results-ttimes}
 ```{r ttime stats}
 ttime_stats <- read.csv(here("analysis/out/stats/prop_pop_ttimes.csv"))
+
 # get prop of pop > 3 hrs away at baseline
 ttime_stats %>%
-  filter(scenario == 0, !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
+  filter(!(cut_times %in% c("< 1", "1 - 2", "2 - 3"))) %$%
   sum(prop_pop) -> prop_over3_base
-ttime_stats %>%
-  filter(scenario == 0, !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
-  sum(prop_pop) -> prop_over6_base
-```
-
-Estimates of travel times to the closest ARMC vary greatly across the country (Fig 3A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live greater than 3 hrs away from a clinic (Fig 3B).
-However, we found that these estimates from the friction surface generally underestimate true driving times across the country and patient reported travel times to the Moramanga ARMC (Fig 3C), and that patient reported travel times were significantly more variable than estimates from the friction surface (Fig S1.2).
-The friction surface assumes that the fastest available mode of transport is used to travel each grid cell (i.e.
-the presence of a road in a given grid cell indicates that all travel through that grid cell will be at vehicle rather than walking speeds).
-However, patients reporting to the Moramanga ARMC report using mutliple modes of transport, with some cases of individuals walking multiple days to reach a clinic (Fig S1.3).
-Despite these discrepancies, travel time estimates from the friction surface correlate with the reported and groundtruthed driving times (Fig 3C, Fig S1.4).
-Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (Table S1.1).
-Therefore we use this metric as a proxy for access at the commune/district level in subsequent analyses.
+```
+
+Estimates of travel times to the closest ARMC varied greatly (Fig 2A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live over 3 hrs from a clinic (Fig 2B).
+We found that travel time estimates from the friction surface underestimated both driving times across the country and patient-reported travel times to the Moramanga ARMC (Fig 2C), but were correlated with ground truthed driving and patient-reported times (Fig 3C, Fig S1.4).
+Patient-reported travel times were also significantly more variable than estimates from the friction surface (Fig S1.2).
+The friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
+the presence of a road indicates that all travel through that grid cell is by vehicle).
+However, Moramanga ARMC patients reported using multiple modes of transport, with some individuals walking days to a clinic (Fig S1.3).
+Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (Table S1.1).
+Therefore we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
 
@@
 
-*__Figure 2A)__ Travel times to the closest ARMC estimated at an ~ 1 km^2^ scale for Madagascar.
-__B)__ Distribution of the population across travel times.
-__C)__ Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level.
-For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune.
-The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
+### *Fig 2. Travel times to ARMC across Madagascar.*
+*(A) Estimated at an ~ 1 km2 scale.
+(B) Distribution of the population across travel times.
+(C) Correlation between ground-truthed travel times (mean of patient reported travel times to the Moramanga ARMC at the commune level and reported driving times between GPS points) and friction surface travel time estimates.
+The vertical lines show the 95% quantiles for reported travel times and the point size shows the number of observations for each commune.
+The best fit lines (red and grey) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
 The dashed black line is the 1:1 line.*
 
-## Estimating average annual bite incidence at the district and commune level {#results-biteinc}
-We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population.
-While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B).
-In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A), and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).
+## Estimating bite incidence {#results-biteinc}
+
+Most patients from each district reported to their closest ARMC by the weighted travel time metric (Fig 3).
+Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
+While there are discrepancies between commune and district catchment assignments (Fig S2.1A), over 75% of the population in a given district or commune were closest to a single clinic (Fig S2.1C).
+We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms (see Supplementary Appendix Section S2).
+After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D).
 
@@
 
-*__Figure 3A)__ The network of patient presentations at the district level for the __A)__ national data and __B)__ commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
+### *Fig 3. The network of patient presentations and estimates of annual bite incidence.*
+*(A) at the district level for the national data and (B) commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
 Color corresponds to the clinic catchment.
@@
 Out-of-catchment reporting is indicated where points and line colors are mismatched.
-For panel __A)__ districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
-For __B)__ the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), all communes included in it's catchment (red polygons), and other communes where bites were reported to in grey.*
-
-Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1).
-We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms (Figure S2.1B).
-Correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned for the majority of clinics (Figure S2.2A and B).
-We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A).
-After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r nrow(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).
-
-```{r fig M4 biteinc}
-include_graphics(here("analysis/figs/main/M4_biteinc.jpeg"))
-```
-
-*__Figure 4.__ The estimated average annual bite incidence from the national and Moramanga data plotted at the __A)__ district scale (both datasets) and at the __B)__ commune scale (Moramanga dataset).
+For panel (A) districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
+For (B) the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), all communes included in it's catchment (red polygons), and other communes where bites were reported to colored by their catchment  (C) The estimated average annual bite incidence from the national and Moramanga data plotted at the district scale (both datasets) and at the (D) commune scale (Moramanga dataset).
 Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
@@
 
-```{r}
-outliers <- read.csv(here("analysis/out/stats/outliers.csv"))
-```
-
-Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship.
-For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times.
+Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
+For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
 While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.
 
-## Modeling reported bite incidence as a function of travel times {#results-bitemods}
-```{r}
-convergence <- read.csv(here("analysis/out/stats/convergence.csv"))
-```
-
-All models and posterior estimates of parameters converged (Gelman-Rubin diagnostic < `r round(max(convergence$val[convergence$OD != TRUE]), 3)`, @gelman2006data).
-All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2).
+## Modeling reported bite incidence {#results-bitemods}
+
+Travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales (Fig 4).
 Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
 
-For the national data, including a random catchment effect improved predictions of the observed data (Fig S3.2 & Fig S3.3), and catchment level effects were strongly identified in the posterior estimates.
-However, after accounting for overdispersion of the data, these catchment effects were not clearly identifiable and the posterior estimates of the catchment intercepts were broadly similar (Fig S3.6), and the models resulted in similar predictions to the data (Fig S3.7), indicating that the catchment effect could not be differentiated from random variation in the data.
-Similarly, while the commune model fit to the Moramanga data generated stronger estimates of the travel time effect (Fig 5B), after accounting for overdispersion of the data, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
-Finally, for the national data, we also looked at the sensitivity of these models to how we corrected our data for underreporting, and we found that parameter estimates of travel time impacts were similar across models and performed similarly in predicting the data, and instead the correction applied impacted the magnitude of the bite incidence estimates (Fig S3.8 and Fig S3.9).
-
-As the prediction from the model fit to the Moramanga data without accounting for overdispersion falls within the prediction interval for the models fit to the national data (Fig 5A), for subsequent predictions of burden, we use the parameter estimates models fit to the national data, as they encompass the range of travel time effects observed in our datasets.
-In addition, our out of fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at this scale, therefore we use both the district and commune model to disaggregate burden to the finest scale possible given our modeling framework.
-
-```{r fig 5}
-include_graphics(here("analysis/figs/main/M5_mods.jpeg"))
-```
-
-_**Figure 5.** Travel times as a predictor of reported bite incidence per 100,000 persons.
-A) The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).
-The lines are the mean estimates and the envelopes are the 95% prediction intervals for predictions generated by drawing 1000 independent samples from the parameter posterior distributions for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with an overdispersion parameter ($\sigma_e$) and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and unadjusted for overdispersion.
-The envelopes show the 95% prediction interval.
-The points show the data used to fit the models: National data (circles) at the district level used to fit the District and Commune models, and Moramanga data (triangles) at the commune level used to fit the Moramanga model.
-The posterior distribution of parameters from the respective models for the B) model intercept, C) travel time effect, and D) the overdisperson parameter (for national data only)._
-
-## Estimating the burden of human rabies deaths subnationally {#results-burden}
-
-Overall, we estimate approximately `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)`(95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`).
-These results vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Figure 6A).
-We also estimate that under the current system of 31 ARMCs in Madagascar, approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)` deaths due to rabies are prevented through provisioning of PEP each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant variation sub-nationally when deaths are modeled at the district and commune scale, with the least accessible communities having the highest burden of deaths (Figure 6B-C).
-
-*__Table 2.__ Model predictions of average annual reported bite incidence, total deaths, and total deaths averted at the national level for the two models (commune level and district level models with a random catchment effect); 95% prediction interval of national estimate in parentheses.*
+For the national data, including a catchment random effect improved predictions (Fig S3.2 & Fig S3.3).
+However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table S3.1) and the models resulted in similar predictions (Fig S3.6 & S3.7), indicating that catchment effects could not be differentiated from random variation in the data.
+Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
+
+As the predictions from the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
+Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig S3.3), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
+Finally, we examined the sensitivity of models to how we corrected for underreporting of data, and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Fig S3.8 and Fig S3.9).
+
+```{r fig 4}
+include_graphics(here("analysis/figs/main/M4_mods.jpeg"))
+```
+
+### *Fig 4. Travel times as a predictor of reported bite incidence per 100,000 persons.*
+*(A) The estimated relationship between travel time in hours (x-axis) and mean annual reported bite incidence (y-axis).
+The lines are the mean estimates and the envelopes are the 95% prediction intervals generated by drawing 1000 independent samples from the parameter posterior distributions for three candidate models: model with travel times at the 1) commune- and 2) district-level fitted to the national data with an overdispersion parameter ($\sigma_e$) and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and unadjusted for overdispersion.
+The points show the data: National data (circles) at the district level used to fit the District and Commune models, and Moramanga data (triangles) at the commune level used to fit the Moramanga model.
+(B) The posterior distribution of parameters from the respective models for the model intercept, travel time effect, and for overdispersion (national data only).*
+
+## Estimating human rabies deaths {#results-burden}
+Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar.
+Our estimates vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Fig 5A).
+Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year.
+In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
+
+### *Table 2.
+Model predictions of average annual reported bite incidence, total deaths, and deaths averted at the national level for the two models (commune level and district level models with travel time predictor and an overdispersion parameter); 95% prediction interval in parentheses.*
 
@@
          averted = glue("{averted_mean} ({averted_lower} - {averted_upper})")) %>%
-  select(scale, bites_100k, deaths, averted) -> table1
-
-kable(table1, digits = -0, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
+  select(scale, bites_100k, deaths, averted) -> table
+
+kable(table, digits = -1, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
                                         "Deaths averted"))           
@@
 
+```{r fig 5, out.width="50%"}
+include_graphics(here("analysis/figs/main/M5_burden_base.jpeg"))
+```
+
+### *Fig 5. Spatial variation in predicted incidence of human rabies deaths per 100,000 persons.*
+*(A) for each district (y-axis) in Madagascar.
+Diamonds show the predicted incidence for the district model and squares show predicted incidence for the commune model fit to the National data for all communes in a district.
+Points are colored and districts ordered by travel times.
+The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
+Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current ARMC.*
+
+## Estimating the impact of PEP provisioning {#results-incremental}
+
+For a subset of additional clinics (n = 83, up to one per district), we compared three methods of ranking for expanding PEP provisioning.
+We found that targeted expansion of PEP to clinics based on the proportion of the population they reduced travel times for resulted in fewest deaths (Fig S5.1).
+Therefore, we used this approach for simulating expansion of PEP to a larger set of clinics (N = `r max_added` ).
+Here we report results from the commune model, as estimates were consistent across models (Fig 6 and Supplementary appendix, section S5).
+
+We estimated that strategic PEP expansion to these additional 83 clinics (1 per district) reduced rabies deaths by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 6A).
+With one clinic per commune (where available, N = 1696), we see a further reduction of `r round((1 - props$deaths_mean[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`%).
+However, reductions in burden saturate as more clinics are added (Fig S5.2).
+Even when all CSB II provisioning PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5); as more clinics are added, reported bite incidence saturates (Fig S5.4), and patients shift which clinic they report to (S5.7 & S5.8).
+
+Vial demand also outpaces reductions in burden (Fig 6B), with more vials needed per death averted (Fig 6C).
+Our model predicts an increase from `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)})")` per annum under current provisioning but with the abridged intradermal regimen (i.e.
+visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")`  with 250 clinics providing PEP, and ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` if all CSB IIs provision PEP.
+In these scenarios, clinic catchment populations and throughput decrease, with clinics seeing fewer patients per day (S5.6).
+
 ```{r fig 6}
-include_graphics(here("analysis/figs/main/M6_burden_base.jpeg"))
-```
-
-*__Figure 6.__ Spatial variation in predicted incidence of human rabies deaths per 100, 000 persons.
-__A)__ for each district (y-axis) in Madagascar, where the diamonds show the predicted incidence for the district model and the squares show the incidence for the commune model fit to the National data for all communes in a district.
-Points are colored by and districts are ordered by travel times.
-The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
-Incidence mapped to the __B)__ commune level and __C)__ district level from the respective models; grey X’s show locations of current ARMC*
-
-## The impact of expanding PEP access to additional clinics {#results-incremental}
-For a subset of clinics (n = 83, up to one clinic per district), we compared the impact of three different metrics for ranking clinics on reducing annual burden of deaths.
-We found that metrics that targeted expansion at clinics based on the proportion of the population they reduced travel times for (above a threshold of three hours) performed the best (Fig S5.1).
-Therefore, we used this metric for simulating expansion to the larger set of clinics (N = `r max_added` clinics).
-
-We found that strategic expansion of access to these additional `r nrow(clin_per_dist)` clinics (i.e.
-1 clinic per district) reduced burden by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 7A).
-With one clinic per commune (where a clinic was available, N = `r one_per_comm`), we see a further reduction `r round((1 - props$deaths_mean[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`%).
-However, reductions in burden saturate as more clinics are added (Fig S5.2).
-Even when all CSB II provision PEP in the country, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`) rabies deaths per annum, and average reporting of rabid exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5).
-This is largely due to the fact that as more clinics are added, reported bite incidence saturates (Fig S5.4), and instead there are shifts in which clinic bites are reported to (S5.7 & S5.8).
-
-Vial demand also outpaces reductions in burden (Fig 7B), with more vials  needed per averted death as PEP access is expanded (Fig 7C).
-Our model predicts an increase from `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)})")` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e.
-visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")` vials with 250 clinics, and approximately `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` vials if all CSB IIs were to provision PEP.
-Importantly, clinic catchment populations and throughput decrease, as well, with clinics seeing fewer patients per day (S5.6).
-
-```{r fig 7}
-include_graphics(here("analysis/figs/main/M7_addARMC.jpeg"))
-```
-
-*__Figure 7.
-A)__ Decrease in deaths due to rabies, __B)__ increase in total # of vials as additional ARMC are added at the national level , and __C)__ increase in vials needed per death averted based on the two models of reported bites.
+include_graphics(here("analysis/figs/main/M6_addARMC.jpeg"))
+```
+
+### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
+*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level , and (C) increase in vials needed per death averted based on the two models of reported bites.
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The points show when all additional csb (N = `r max_added`) clinics have been added).
-*
-
-## Sensitivity of estimates to bite incidence and decision tree model parameters {#results-sensitivity}
-To quantify which parameters contribute to the uncertainty observed in our estimates, we performed a univariate sensitivity analyses, varying parameters in both the model of bite incidence (Table S6.1) and the decision tree model (Table S6.2).
-These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the greatest to uncertainty in estimates of baseline burden (Fig S6.1) and for determining the impact of expanding access (Fig S6.2).
+The points show when all additional csb (N = `r max_added`) clinics have been added).*
+
+## Sensitivity Analyses {#results-sensitivity}
+To quantify which parameters contribute to the uncertainty in our estimates, we performed univariate sensitivity analyses, varying parameters in both the bite incidence model (Table S6.1) and the decision tree (Table S6.2).
+These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute most uncertainty to burden estimates (Fig S6.1) and impacts of PEP access (Fig S6.2).
 Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden.
-For the estimates of vial demand, uncertainty around the intercept of the model (i.e.
-the baseline reported bite incidence) has the greatest impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
-Finally, scaling of incidence with population size (Fig S6.4) modulates this impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
-more rabies in more populated places) dampening and negative scaling enhancing the relationship between access and deaths (Fig S6.5A).
-However, the impact of adding clinics remains broadly the same (Fig S6.5B).
-Overall, while uncertainty in the underlying rabies exposure incidence results in significant variation in the absolute number of rabies deaths, our estimates of the impact of travel times on burden and of the potential for expanded access to reduce burden are qualitatively similar across parameter assumptions.
-
-# DISCUSSION
+For the estimates of vial demand, uncertainty around the model intercept (i.e.
+the baseline reported bite incidence) has most impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
+Finally, scaling of incidence with population size (Fig S4.2) modulates the impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+more rabies in more populated places) dampening and negative scaling exacerbating the relationship between access and deaths (Fig S6.4A).
+However, the impact of adding clinics remains broadly the same (Fig S6.4B).
+Overall, while uncertainty in the underlying rabies exposure incidence results in considerable variation in our burden estimates, the projected impact of travel times and of access to PEP are qualitatively similar across parameter assumptions.
+
+# Discussion
 
 ## Main findings
-Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar.
-Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP.
-Under the status quo (N = 31 clinics), we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP.
-We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic per district (N = `r one_per_dist`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")` and to one clinic per commune (N = `r one_per_comm`) reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")`.
-
-However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all additional CSB II in the country (N = `r max(natl_preds$scenario)` additional clinics), approximately `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]))}")` deaths may still occur nationally.
-This result is driven by two key factors.
-First, given the landscape of clinics and travel times across the country, despite expansion to all major health centers in the country, travel times remain high for a significant proportion of the population (Figure S5.4).
-Secondly, even for locations with low travel times given the range human rabies exposures between 15 - 76 per 100,000 persons, we estimate an average incidence of human rabies deaths of `r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` and reporting of approximately `r round(max(admin_base$reporting_mean), 2)*100`% rabies exposures (Figure S5.5).
-Importantly, for administrative units with similar travel times, we estimate a significant amount of variation that cannot be explained by travel times alone, suggesting other factors may be driving access and reporting of bite incidence.
-
-Because of these factors, we find that as PEP provisioning is expanded to more clinics, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to.
-Vial demand reflects these patterns as well.
-However, vials needed per death averted are lower than predictions from a recent cost effectiveness analysis of PEP expansion (assuming a cost of 5 USD per vial, a maximum of `r round(max(natl_preds$vials_upper/natl_preds$averted_upper)*5)` USD per death averted in direct costs, similar to the ~ 600 USD per death averted estimated in @whorabiesmodelingconsortiumPotentialImpactImproved2019), suggesting that expanding PEP access will remain highly cost effective given current ranges of reported bite incidence.
-Ultimately, these quantitative findings are highly dependent on the assumption of underlying rabies exposure incidence.
-Data on rabies incidence in dogs and exposures in humans are needed to confirm these estimates, but the qualitative patterns regarding travel time impacts remain robust and therefore can be useful in identifying strategies for expanding PEP and reducing burden.
+
+We find that the burden of rabies in Madagascar is likely concentrated in areas with poor PEP access.
+We estimate that current PEP provisioning (at 31 clinics) averts `r glue("{trans_perc(min(prop_averted))}%")`% of deaths that would otherwise occur, and that expanding PEP access should reduce mortality, with provisioning in one clinic per district (N = `r one_per_dist`), or per commune (N = `r one_per_comm`), expected to reduce mortality by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")` and `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")`, respectively.
+However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all CSB II (N = `r max(natl_preds$scenario)`).
+This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all major health centers.
+With reduced travel times, over 20% of exposures will still not seek PEP (Fig S5.5), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
+PEP is expected to remain cost-effective as provisioning expands, to a maximum of 450 USD per death averted (assuming 5 USD per vial), similar to other estimates [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+While our quantitative predictions depend on assumptions of underlying rabies exposure incidence, the qualitative patterns regarding travel time impacts remain robust and are useful in identifying strategies for provisioning PEP.
 
 ## Limitations
-The travel time estimates from the MAP friction surface underestimate true travel times reported by patients, likely due to discrepancies between assigned transport speeds and modes and the reality of travel in Madagascar.
-These speeds are taken from the Open Street Map user community, or in it's absence, from country cluster data [@weissGlobalMapTravel2018], and also assume that given multiple types of transport in a cell, the fastest available mode of transport is used.
-In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used to travel these roads.
-Patients reporting to the Moramanga ARMC reported using a variety of transport methods and had highly variable travel times even within a single commune.
-Future studies should aim to improve estimates of travel times with country specific data on road speeds and travel conditions [@rayAccessModComputingGeographic2008] to better estimate spatial inequities in access, or by collecting data on patient reported travel times.
-While patient reported travel times may also lack precision due to recall and estimation error, they will likely be more reflective of the range of travel realities that people experience.
-
-We were able to use multiple sources of data to compare estimates of bite incidence across scales.
-We found that we could effectively disaggregate burden to the commune scale using our travel time metric.
-The Moramanga data demonstrates that sub-district patterns of access are important to account for when estimating potential areas with low access.
-However, we lacked data from other catchments at the commune scale to further confirm and compare finer scale patterns.
-Focusing on routinely collecting data at the sub-district level could help to resolve these patterns.
-
-We assume that spatial access to PEP is the primary driver of health-seeking behavior, however many other factors may drive an individual’s decision to seek PEP, such as socioeconomic and educational status and awareness about the danger of rabies [@samboKnowledgeAttitudesPractices2014; @barbosacostaInfluencePovertyRabies2018; @etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008].
-In the Madagascar context where PEP is provided free of charge to patients, the main cost of seeking PEP is for transport and associated time lost for travel.
-In addition, more remote communities tend to be of lower socioecomonic and educational status [@peters2008poverty], so our travel time metric may act as a proxy for these set of correlated variables that impact an individual’s access to PEP.
-In fact, a recent knowledge-attitudes-practices (KAP) study of both community members in the Moramanga District and ARMC staff across the country found that both socioeconomic status and living in an urban area were predictive of knowledge scores [@leblancclaireRabiesMadagascarThreepronged2019].
-We also assume that all ARMC reliably provision PEP, but this is unlikely to be the case.
-Leblanc 2019 found that at the time of their KAP survey, three clinics were experiencing vaccine stock-outs.
-In addition, while direct costs of the vaccine are not charged to patients, 22 of 31 current ARMC ask for other fees (ranging from 0.50 - 3.00 USD for consultations, gauze, supplementary medication for wound treatment, etc per visit; @leblancclaireRabiesMadagascarThreepronged2019) which might additionally prevent people from receiving treatment after reporting.
-
-We included an overdispersion parameter in our model to quantify variation not captured by travel times.
-For all models, the data were significantly overdispersed, suggesting that variation at the clinic level (i.e.
-differences in clinic functioning such as vaccine availability and costs charged) and at the administrative level (i.e.
-differences in regional bite incidence, awareness, outreach, etc.) are likely important drivers of health-seeking behavior and overall vaccine demand.
-While clinic specific effects were identifiable in models without overdispersion, these random effects were not identifiable when accounting for random variance in the data.
-Importantly, the national data we used were underreported and incomplete, but we attempted to correct for these issues (quantifying under-submission of forms and excluding districts from catchments with less than 25% of data reported).
-Improving data submission and quality from clinics across the country will be important to better understand patterns of health seeking behavior and access, but also to monitor how well clinics are functioning and to develop better provisioning systems.
-
-Our decision tree framework makes strong assumptions about baseline rabies exposure incidence, based on a previous study from the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018] and a global study of rabies burden[@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-As incidence of rabies exposures varies significantly over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into these estimates to generate a mean estimate of burden at the national level.
-Even when we assume the extremes of these ranges, or that rabies exposure incidence varies systematically with human populations, we find that the qualitative effects of travel times on risk of rabies deaths remains the same.
-While our sensitivity analyses show that the relative patterns of burden are robust, the magnitude of deaths is most sensitive to our assumption of rabies exposure incidence.
-Thus, our estimates are more representative of patterns of risk than quantitative differences in burden, and spatial estimates of rabies exposure incidence would be necessary to generate more precise subnational estimates of burden.
-
-We make several simplifying assumptions regarding PEP compliance and efficacy.
-We assume perfect compliance of patients and efficacy of PEP.
-Although delays in reporting and poor adherence to PEP contribute to risk of death, PEP compliance is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2018], and these factors are unlikely to be major drivers of burden [@changaluchaNeedImproveAccess2018].
-We do not account for differential risk for severely exposed patients (i.e.
-Category III exposures) that do not receive rabies Immunoglobulin (RIG).
-RIG is only available at IPM in Antananarivo, and while exposed persons may be referred to seek RIG, the majority of Category III exposures that seek care are unlikely to have received RIG.
-We were not able to link patient records between peripheral ARMC and IPM to directly estimate this.
-However, recent studies have shown that even in the absence of RIG, PEP is highly effective at preventing rabies deaths, and only adds a marginal benefit in terms of reducing burden but at very high cost.
-It is both unlikely that availability of RIG contributes significantly to rabies burden in Madagascar, or that expansion of RIG provisioning would be a feasible option given the costs.
-
-Ideally, many of these assumptions could either be confirmed or updated with improved data on canine rabies incidence, health-seeking behavior, and clinic functioning and provisioning of PEP.
-Even with these additional data sources, it will be challenging to incorporate the complex interactions between socioeconomic factors, access to and quality of care, and human behavior to predict the impact of PEP spatially.
-We detail case studies in Box 1 that highlight some of the challenges in capturing variable healthcare seeking behavior in the context of PEP provisioning, and the policy implications of these exceptions.
-
-## Box 1: case studies of the challenges of predicting health seeking behavior and PEP efficacy
-```{r}
+Data limitations introduced bias and uncertainty to our estimates.
+For example, travel times from the MAP friction surface underestimated patient-reported travel times, with discrepancies between assigned transport speeds (from the Open Street Map user community, or country cluster data [@weissGlobalMapTravel2018]) and realities of local travel.
+In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used.
+Moramanga ARMC patients reported various transport methods and highly variable travel times even within a single commune.
+While patient-reported travel times may lack precision from recall and estimation error, they likely better reflect lived experience; validated travel times [@rayAccessModComputingGeographic2008, ihantamalalaImprovingGeographicalAccessibility2020] could improve estimates of spatial health inequities.
+Similarly, modeled estimates of population distribution [@linardPopulationDistributionSettlement2012a] also introduce uncertainty.
+Our analysis of data from the Moramanga District indicate that variation at the sub-district level is high and impacts health seeking behavior.
+However, we lacked fine-scale data from other catchments for comparison.
+Additionally, we had to correct for underreporting and incomplete data; strengthening surveillance and routine data collection should improve understanding of health seeking behavior and access, and support monitoring and evaluation of PEP provisioning.
+
+We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
+In Madagascar, where PEP is free-of-charge, the main cost to patients is transport and time lost.
+More remote communities tend to be of lower socioeconomic and educational status [@peters2008poverty], so travel time may be a proxy for these correlated variables.
+We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
+Most ARMC charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers.
+Significant overdispersion in the data that cannot be explained by travel times suggests that clinic-level variation (e.g.
+vaccine availability and charges) and regional differences (e.g.
+dog populations, outbreaks, awareness) further influence health-seeking behavior and vaccine demand.
+
+Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths.
+Our simplifying assumptions, regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP, are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
+Likewise we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care and human behavior, as illustrated by the complex case studies in Box 1.
+
+## Box 1: case studies of of health seeking behavior for PEP in Madagascar
+```{r case studies}
 mora_bites %>%
@@
 
-```
-
-### Anosibe A'nala
-In the district of Anosibe A'nala (population of ~ `r round(mada_districts$pop[mada_districts$district == "Anosibe-An'ala"])` people) south of the Moramanga District, travel times to the nearest ARMC are high and highly variable, often exceeding 24 hours and consisting of days of travel on foot.
-Despite these barriers, we see a moderate incidence of reported bites (~ `r round(aa_bites$inc)` per 100,000 persons).
-Importantly `r trans_perc(prop_cat1_AA$prop_cat1)`% of these patients were loosely defined Category I exposures, i.e.
-people who had been in close proximity or had touched a person that had died from suspected rabies.
-Approximately 50% of the non-Category I exposures were classified as a probable rabies cases based on clinical criteria, and we recorded 4 deaths in this district over a `r max(mora_bites$nobs, na.rm = TRUE)` month period.
-Only one animal case was confirmed diagnostically however.
-The data from Anosibe A'nala points to three key insights:
-
-1) Given the high travel times and the observed incidence of rabies exposures and human rabies deaths, this remote community is likely experiencing a much higher burden of rabies of which we are observing only a small fraction.
-In our modeling analysis, we estimate that an annual burden of `r round(aa_deaths$deaths_mean[1], 2)` per year.
-
-2)Travel times estimated from the friction surface (average weighted travel times of `r round(mada_districts$ttimes_wtd[mada_districts$district == "Anosibe-An'ala"]/60, 2)`) are much lower than those observed for this district (average patient reported travel times of `r round(aa_ttimes$mean, 2)` hours).
-While there is a road connecting the main commune of Anosibe An'ala to the Moramanga ARMC, road speeds are often < 10km per hour and conditions, particularly during the rainy seasons, make it impossible for anything but large transport trucks to travel this road.
-Most people use multiple modes of transport depending on cost and availability (i.e.
-boat, on foot, by transport truck).
-In our ranking of clinics by reduction in travel times, this clinic was the `r aa_added`th clinic added.
-However, given the recurrent cases of human rabies deaths and the high travel times, this area might deserve higher prioritization.
-Importantly, other remote communities may also be experiencing a high and hidden burden of rabies, and improved surveillance across the country will be necessary to best prioritize resources.
-
-3) Despite these high travel times, when people are aware of the risk of death due to rabies and when PEP is available at no direct costs, they are willing to travel many hours to days (in some cases walking 3 days to attend the clinic) to seek PEP.
-In addition, a strong surveillance response, i.e.
-referral of people in contact with suspect human cases of rabies is possible, and can work to improve PEP efficacy if these investigations were triggered on human exposure or animal cases rather than a human death.
+aa_exposures <- read.csv(here_safe("data-raw/misc/aa_exposures.csv"))
+nrabid <- nrow(aa_exposures[aa_exposures$suspect.exposure %in% c("Probable",
+                                                                 "Confirmed"), ])
+ninterviewed <- nrow(aa_exposures[!(aa_exposures$suspect.exposure == ""), ])
+
+```
+
+1.	Anosibe An’ala District (population ~ `r format(round(mada_districts$pop[mada_districts$district == "Anosibe-An'ala"]), big.mark = ",")`), south of Moramanga, has moderate incidence of bite patients (~ `r round(aa_bites$inc)`/100,000 persons) even though travel times often exceed 24 hours.
+While a road connects the main Anosibe An’ala commune to the Moramanga ARMC, it is only passable by large trucks during much of the rainy season, with speeds usually < 10km per hour.
+Over `r trans_perc(prop_cat1_AA$prop_cat1)`% of patients from Anosibe An’ala had been in close proximity or touched a person that died from rabies (four suspect human rabies deaths of patients who did not receive any PEP), whilst of patients with Category II and III exposures that were interviewed, `r glue("{nrabid}/{ninterviewed}") ` (`r trans_perc(nrabid/ninterviewed)`%) were bitten by probable rabid dogs.
+Given the high travel times (although underestimated by the friction surface) and incidence of reported rabies exposures and deaths, we predict a large but unobserved rabies burden in this remote community (~`r round(aa_deaths$deaths_mean[1], 2)` deaths per year) and we ranked an ARMC provisioning PEP in Anosibe An’ala `r aa_added`th for travel time reductions.
+Other remote communities likely experience similar high and unrecognized burden, but improved surveillance is necessary to identify such areas.
+Notably, bite patients in this district demonstrate willingness to travel for free PEP (in some cases walking 3 days to a clinic) with awareness of rabies risk.
 Community outreach and active surveillance in other remote areas could also greatly improve people's awareness of risk and health seeking behavior.
 
-### Antananarivo 
-In 2016, a taxi driver in his 60s died of rabies in suburban Antananarivo.
-He was bitten by an unknown dog returning from work one evening, and the next day he promptly reported to a clinic where he was reffered to the ARMC at IPM for PEP.
-His daughter, a prospective medical student, and other family members urged him to seek the vaccine, but he did not believe his risk was high and so in the end did not seek further care.
-He developed symptoms 2 weeks later and died within a few days.
-The case was later confirmed as a rabies death by the Naional Rabies Reference Laboratory at IPM.
-This case points to the other side of the health-seeking behavior coin from Anosibe A'nala.
-Even given prompt reporting and referral, and socioeconomic indicators that would suggest a high probability of seeking PEP, this person did not seek PEP.
-This case highlights the need for community education on rabies risk, and how improving availability of PEP at peripheral clinics even in areas with already high access could remove further barriers to PEP and prevent additional deaths.
-It also highlights the limitations of PEP provisioning as the sole intervention for preventing human rabies deaths.
+2.
+Recently, a middle aged taxi driver died of rabies in suburban Antananarivo.
+The day after being bitten by an unknown dog, he reported to a clinic that referred him to the ARMC at IPM, approximately one hour’s drive from his home.
+His family urged him to get PEP, but he did not believe his risk was high and decided not to seek further care.
+He developed symptoms two weeks later and was confirmed as a rabies death by the National Rabies Reference Laboratory.
+Despite prompt reporting and referral, and socioeconomic indicators suggesting a high care-seeking probability, this person did not receive PEP.
+His story highlights the need for sensitization about rabies and how PEP at peripheral clinics (even in areas with reasonable access) could prevent additional deaths, but also how PEP alone is unlikely to prevent all rabies deaths.
 
 ## Broader context
-Our work is in line with previous studies demonstrating the impacts of geographic access to care on health outcomes [@aleganaTreatmentseekingBehaviourLow2017; @nikolayEvaluatingHospitalBasedSurveillance2017; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @Hegde2019dt; @metcalfTransportNetworksInequities2015].
-While many of these studies directly estimated reporting using community based and cross-sectional survey data [@aleganaTreatmentseekingBehaviourLow2017; @nikolayEvaluatingHospitalBasedSurveillance2017], we use model derived estimates of reporting and assumptions of baseline disease incidence to estimate burden, similar to methods developed for other pathogens to estimate burden in data sparse settings [@macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @Hegde2019dt].
-For canine rabies, estimating burden of disease directly is challenging [@scottAddressingDisconnectEstimated2017; @taylorDifficultiesEstimatingHuman2017] and direct empirical estimates of rabies burden have previously relied on exhaustive contact tracing studies [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008] or large-scale mortality surveys [@sudarshanAssessingBurdenHuman2007].
+
 Recent studies have estimated access to health-seeking behavior and PEP completion and adherence, but not directly linked these estimates to burden [@tarantolaRabiesPostexposureProphylaxis2017; @zaidiGeographicVariationAccess2013].
-Our framework for incorporating estimates of access into previously developed burden estimation methods (i.e.
-decision trees) could help to improve and refine existing burden estimates and explore impacts of expanding access, particularly at the sub-national level.
-
-The observed range of reported bite incidence from our data are in line with estimates of reported bite incidence across Africa (see @whorabiesmodelingconsortiumPotentialImpactImproved2019 Supplementary Materials for both a summary of data sources and model estimates).
-Ultimately, reported bite incidence is driven by many factors unrelated to rabies incidence in the dog population, including the underlying size of dog populations, barriers to access to PEP (i.e.
-both financial costs and geographic access), and awareness of rabies risk.
-In our study, we do not attempt to quantify the total non-rabies related bite incidence, but use a range of values for the proportion of reported bites which we think are genuine rabies exposures.
-As we did not have location specific estimates of this parameter, we used a range estimated from a previous study in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018].
-This approach is likely to be less applicable in settings where rabies exposures only make up a small fraction of patients reporting for PEP [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015], i.e.
-where barriers to PEP are low (i.e low or no direct costs, low travel times, high socioeconomic status of the population) and reporting of non-rabid bites high.
-
-While our incidence of human rabies deaths are based on strong assumptions of rabies incidence, they generally fall within the range estimated for a wide range of settings from both empirical and modeling studies  [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011; @lyRabiesSituationCambodia2009, @whorabiesmodelingconsortiumPotentialImpactImproved2019].
-We estimate a higher burden of rabies deaths here than we previously estimated for Madagascar [@rajeevHealthcareUtilizationProvisioning2018].
-While in this study we assumed a lower range of incidence of underlying human rabies exposures, previously we assumed a fixed reporting of 85% of rabies exposures across the country.
-Accounting for spatial variation in access and in reported bite incidence across the country resulted in predictions of increased burden, mirrored by a decrease in deaths averted under baseline provisioning.
-Improved clinic level data, as well as surveillance data on rabies exposures will be necessary to further refine these estimates, particularly given sensitivity of these estimates to assumptions of rabies exposure incidence.
-
-One approach to obtaining this data is integrated bite case management which relies on follow-up of bite patients to determine rabies exposure status, recommend treatment, and identify other potential persons exposed [@lechenneImportanceParticipatoryIntegrated2017; @undurragaCostEffectivenessEvaluationNovel2017; @rajeevHealthcareUtilizationProvisioning2018].
-If these methods were implemented, ARMC could act as sentinels for rabies surveillance across the country, with bite patients reporting to clinics triggering investigations of suspect rabies cases.
-Ultimately, these methods rely on exposed persons to report to clinics, and as such, expanding access to PEP could strengthen clinic based surveillance efforts.
-
-Our estimates of vial demand depend on countrywide adoption of the new abridged intradermal regimen recommended by WHO [@tarantolaIntradermalRabiesPostexposure2018].
-While this new recommendation has been adopted by the Ministry of Health, it will be important to train clinicians and staff, as in 2019, only 15% of staff interviewed were aware of this regimen [@leblancclaireRabiesMadagascarThreepronged2019].
-In addition, the majority of staff were also not aware of the WHO classifications of exposure categories, and vaccination of Category I exposures, i.e.
-those not recommended PEP, is a common practice.
-In the Moramanga district, up to 20% of vial demand was due to these exposures [@rajeevHealthcareUtilizationProvisioning2018], and more judicious PEP provisioning could reduce vial demand accordingly.
-
-We predict that as clinics are expanded, overall throughput (i.e.
-the number of patients reporting daily to a clinic) will decrease.
-This may make provisioning PEP more challenging, as vial demand may become less predictable annually, resulting in stockouts or vaccine wastage.
-Establishing decentralized provisioning mechanisms, for example through combining provisioning of PEP with routine childhood vaccines, or novel methods for vaccine delivery (i.e.
-drone delivery of emergency medicine, [@boeras2019use]) may be potential ways to mitigate these challenges.
-Previous to 2006 when the nerve tissue vaccine was still in use, access to PEP in Madagascar was more widespread, with clinics able to request vaccine from IPM upon demand.
-However, once the more expensive Vero cell vaccine was implemented, vaccine provisioning to all clinics became too costly, resulting in provisioning to a subset of clinics [@andriamandimbySurveillanceControlRabies2013].
-This suggests that wider spread vaccine provisioning is feasible given health infrastructure in Madagascar if cost barriers are removed.
+Our approach for incorporating access to vaccines (echoing [@Hegde2019dt; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @metcalfTransportNetworksInequities2015; @aleganaTreatmentseekingBehaviourLow2017]) into burden estimation methods could guide provisioning of PEP to maximize impacts.
+This approach will have most value in settings with limited PEP access and poor health seeking, but will be less valuable where rabies exposures make up a small fraction of patients reporting for PEP e.g.
+[@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015].
+Our revised estimate of rabies deaths in Madagascar using this approach was higher than previous [@rajeevHealthcareUtilizationProvisioning2019], which assumed uniform reporting of 85%, but remained within the range of other empirical and modeling studies from low-income countries [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @lyRabiesSituationCambodia2009; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011].
+
+Our estimates of vial demand depend on use of the new abridged intradermal regimen [@tarantolaIntradermalRabiesPostexposure2018], which has been adopted by the Ministry of Health in Madagascar.
+However, most ARMC staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moromanga [@rajeevHealthcareUtilizationProvisioning2019].
+We predict that as clinics are expanded, throughput (daily patients reporting to a clinic) will decrease.
+This may make provisioning PEP more challenging and vial demand less predictable, leading to stock outs or wastage.
+Decentralized provisioning mechanisms, for example adopting routine childhood vaccine supply chains, or novel vaccine delivery methods such as drones [@boeras2019use], may mitigate these challenges.
+When nerve tissue vaccines were used in Madagascar, clinics requested vaccine upon demand and PEP access was more widespread, but provisioning the more expensive cell culture vaccines to all clinics became too costly [@andriamandimbySurveillanceControlRabies2013].
+Widespread vaccine provisioning is therefore feasible given Madagascar’s health infrastructure, if cost barriers are removed.
 
@@
 
-A Gavi investment can remove this cost barrier.
-In most rabies endemic countries, access to PEP is geographically limited and/or costly to patients [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
-Currently, each clinic in Madagascar serves on average a catchment population of `r round(mean(catch_trends[scenario == 0]$pops_mean), -3)` persons.
-Latin American countries, where significant progress has been made towards elimination, aim to provide one clinic providing PEP for 100,000 persons, which would require approximately `r min(catch_pops[mean_pops < 1e5]$scenario)` additional ARMC in Madagascar.
-If GAVI investment results in expansion to underserved areas and strengthens existing provisioning and supply chain mechanisms, then rabies burden could be greatly reduced.
-Given our estimates of vial demand and deaths averted, PEP expansion remains highly cost-effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019] when considering direct costs only.
-However, our results suggest that expansion alone cannot prevent the majority of deaths and improved outreach and education on the risk of rabies exposure, better surveillance and active case management, and ultimately control in the domestic dog population will be necessary to manage the burden of rabies in Madagascar.
-
-Ultimately even given maximal access, achieving 'the last mile', preventing deaths due to rabies in the most remote and vulnerable populations, will require a disproportionate level of resources [@delriovilasTribulationsLastMile2017].
-IBCM programs and risked based assessments are one way to tackle the issue of managing PEP resources effectively even as demand outpaces reductions in burden [@rysavaPathRabiesElimination2018].
-Combining PEP provisioning with mass dog vaccination will be key to interrrupting transmission, and eliminating deaths due to rabies.
-These same issues of access, however apply to both dog vaccination programs and surveillance systems, and integrating spatial heterogeneities in vaccination and detection will be critical to understand how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015].
+Gavi investment could greatly reduce the access and cost barriers to PEP [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
+Currently, each clinic in Madagascar serves an average catchment of `r format(round(mean(catch_trends[scenario == 0]$pops_mean), -4), big.mark = ",")` persons.
+Latin American countries, where significant progress has been made towards elimination, aim for one PEP clinic per 100,000 persons.
+In Madagascar this would require around `r min(catch_pops[mean_pops < 1e5]$scenario)` additional ARMC.
+We predict that Gavi investment would be highly cost-effective, greatly reducing deaths by expanding PEP supply to underserved areas.
+
+However, our results suggest that PEP expansion alone cannot prevent the majority of rabies deaths, and even given maximal access, achieving ‘the last mile’, preventing deaths in the most remote populations, will require disproportionate resources [@delriovilasTribulationsLastMile2017].
+To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission, and eliminating deaths.
+Integrated Bite Case Management (IBCM) uses bite patient risk assessments to determine rabies exposure status, guide PEP administration, and trigger investigations of rabid animals, potentially identifying other exposed persons [@lechenneImportanceParticipatoryIntegrated2017; @rajeevHealthcareUtilizationProvisioning2019; @undurragaCostEffectivenessEvaluationNovel2017].
+IBCM is one way to manage PEP effectively [@rysavaPathRabiesElimination2018] and as it relies on exposed persons reporting to clinics, expanding PEP access could strengthen this surveillance framework.
+These same issues of access, however apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].
 
 ## Conclusion
-Geographic access to PEP is a strong predictor of reported bite incidence in Madagascar and expanding PEP availability to more clinics across the country could greatly reduce human rabies deaths.
-In the absence of data on rabies incidence and risk of exposure, targeting expansion of PEP at areas currently underserved may be the best way to reduce rabies burden and provide equitable access to PEP.
-Specifically, exanding provisioning clinics serving populations that under the status quo would have to travel greater than 3 hours to seek PEP, or alternatively aiming for clinics to serve a catchment size of approximately 100,000 persons, are two ways that PEP access could be expanded strategically.
-However, our results also suggest that increasing outreach and awareness of rabies risk, improving clinic functioning and provisioning practices, and removing socioeconomic barriers for persons seeking PEP will be critical to ensure the efficacy of PEP expansion.
-Improved surveillance of animal rabies will also be needed to better understand the landscape of rabies burden across countries and to identify populations at the greatest risk for rabies deaths.
-Expansion of PEP can greatly reduce risk and inequities in access to this life-saving vaccine, and a GAVI investment would allow countries to more equitably provide PEP to populations that currently face immense barriers to seeking this intervention ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies).
-In addition, as PEP alone cannot prevent all rabies deaths, it could facilitate investment in mass dog vaccination to interrupt transmission, and eventually eliminate deaths due to tnis devastating disease.
-
-## Acknowledgements
-
-We would like to thank all the clinicians and staff at the ARMC across the country.
-We are grateful to IPM and the Ministry of Public Health who work to collect and maintain data on PEP provisioning.
-In particular, on the data management side, we would like to thank the GIS unit for assistance with spatial data and Michael Luciano Tantely for sharing driving time data.
-We would like to thank Claire LeBlanc, Rila Ratovoson, and Daouda Kassie for sharing results of their work in the Moramanga District.
-In addition, we would like to thank Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, Amy Winter, Christian Ranaivoson, John Friar, and Amy Wesolowski for their assistance with data collection.
+Our study suggests that rabies deaths in Madagascar disproportionately occur in communities with the poorest access to PEP and that expanding PEP access should reduce deaths.
+Without data on rabies incidence and exposure risk, targeting PEP expansion to underserved areas is a strategic way to reduce  rabies burden and provide equitable access, for example, by expanding provisioning to clinics serving populations that target an evidence-based travel time threshold or catchment size.
+Implementing outreach programs to raise awareness should further increase the efficacy of PEP expansion by improving care seeking.
+Better surveillance is also needed to understand the geographical distribution of rabies exposures and identify populations most at risk.
+Gavi investment could support countries to more equitably provision PEP and overcome barriers to access ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies), but as PEP alone cannot prevent all rabies deaths, investment should be used to catalyse mass dog vaccination to interrupt transmission, and eventually eliminate rabies deaths.
+
+# Acknowledgements
+
+We thank all the clinicians and staff at the ARMC across the country.
+We are grateful to IPM and the Ministry of Public Health who collect and maintain data on PEP provisioning.
+In particular, we thank the GIS unit for assistance with spatial data, Michael Luciano Tantely for sharing driving time data, and Claire Leblanc, Rila Ratovoson, and Daouda Kassie for sharing results of their work in the Moramanga District.
+In addition, we thank Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, Amy Winter, Christian Ranaivoson, John Friar, and Amy Wesolowski for assistance.
 This work was funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to CJEM and MR.
@@
 KH is supported by the Wellcome Trust (207569/Z/17/Z).
-\newpage
-
-## References 
+
+\newpage
+
+# References 
 ::: {#refs_main}
@@
 
-\newpage
-
-# SUPPLEMENTARY FIGURES & TABLES
+<!-- split -->
+
+\newpage
+
+# Supplementary Appendix
 
@@
 ```
-
-*__[Figure S1.1.]{#figS1.1}__ Raster inputs to estimate travel times to the closest ARMC for Madagascar __A)__ Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale __B)__ Population estimates resampled to the same friction surface.*
+### *Fig S1.1. Raster inputs to estimate travel times to the closest ARMC for Madagascar.*
+_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale. (B) Population estimates resampled to the same friction surface._
+
+\newpage
 
@@
 
-*__[Figure S1.2.]{#figS1.2}__ Source of true travel time estimates __A)__ Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
-__B)__ Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
-Paths are bezier curves from origin to destination, but do not show actual paths driven.*
+### *Fig S1.2. Raw patient reported and driving travel time data.*
+_(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
+(B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
+Paths are bezier curves from origin to destination, and do not show actual paths driven._
+
+\newpage
 
@@
 
-*__[Figure S1.3.]{#figS1.3}__ Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
+### *Fig S1.3. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
+\newpage
 
@@
 
-*__[Figure S1.4.]{#figS1.4}__ Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) predicted by __A)__ Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) __B)__ Travel time estimates and C) Travel time estimates weighted by population (for commune means only).*
-
-*__[Table S1.1.]{#tabS1.1}__ R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
+### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations).*
+_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only)._
+\newpage
+
+### *Table S1.1. R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
 ```{r table S1.1 lms table}
@@
 kable(S1.1_table, col.names = c("Predictor", "Driving times", 
-                                "Commune mean of patient reported travel times"))
-```
+                                "Commune mean of patient reported travel times"),
+      digits = 3)
+```
+
+ \newpage
 
@@
 
-*__[Figure S1.5.]{#figS1.5}__ Estimates of mean travel times weighted by population at the __A)__ District __B)__ Commune scale.*
-
-## S2. Estimating average annual bite incidence at the district and commune level
+### *Fig S1.5. Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.*
+
+\newpage
+
+## S2. Estimating bite incidence 
+Vial demand was simulated under simplified assumptions of PEP administration and adherence [@rajeevHealthcareUtilizationProvisioning2019], based on patients reported randomly across the year.
+During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient completing doses on days 0, 3, 7, and 28.
+Vials can be shared within a day between two patients, resulting in 0.1ml wastage per vial shared, plus any additional wastage from unused doses discarded at the end of the day.
+We estimate vial estimates as the midpoint estimate if all patients complete 3 vs.
+4 doses.
+As clinic submission of forms was highly variable from 2014-2017 (Fig S2.2A), we compared estimated demand to the total vials provisioned across this four year period comparing different thresholds for correcting for periods of no form submission (i.e, designating periods of 1, 5, 10, 15, and 30 consecutive days with zero submitted records as missing, compared to no correction).
+
+Estimates of vial demand based on uncorrected bite patient numbers were generally lower than the number of vials provisioned for those clinics with substantial under-submission of forms (Fig S2.2B).
+Correcting patient numbers for under-submission resulted in estimates of vial demand closer to the provisioned vials for most clinics  (Fig S2.2C, Table S2.1).
+
+\newpage
 
@@
 
-*__[Figure S2.1]{#figS2.1}A)__ Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or which a district (polygon outline).
+### *Fig S2.1. Catchment assignments by travel time.*
+*(A) Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or within a district (polygon outline).
 Admin units where the fill and border colors do not match show places where assigned catchments differ at the district vs commune scale.
-__B)__ Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
-__C)__ The proportion of bites reported to each clinic which originated from a district within the assigned catchment.*
+(B) Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
+(C) The proportion of bites reported to each clinic which originated from a district within the assigned catchment.
+The vertical line indicates the proportion of bites from within the assigned catchment for the Moramanga data (~90%).
+*
+
+\newpage
 
@@
 
-*__[Figure S2.2.]{#figS2.2}A)__ The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
-__B)__ Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the  criteria of 15 consecutive days of non-submission/365.
-__C)__ The difference between log(estimated) and log(observed) vials provisioned for the period of 2013 -  2017 for each clinic correcting for undersubmission (squares) using the 15 day cut-off show in A and B, vs.
-not correcting for undersubmission (circles).
+### *Fig S2.2. Estimating undersubmission of patient forms.*
+*(A) The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
+(B) Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the criteria of 15 consecutive days of non-submission/365.
+(C) The difference between log(estimated) and log(observed) vials provisioned for the period of 2013 - 2017 for each clinic correcting for under-submission (squares) using the 15 day cut-off show in A and B, vs.
+not correcting for under-submission (circles).
 We did not have data on vials provisioned for IPM.*
-
-## S3. Modeling reported bite incidence as a function of travel times {#S3}
+\newpage
+
+### *Table S2.1.
+Root mean squared error (MSE) between observed vials provisioned and estimated by the different consecutive day threshold for correcting for periods of no form submission, with the minimum root MSE in bold.*
+```{r table mse stat}
+mse <- read.csv(here("analysis/out/stats/vial_mse.csv"))
+mse %>%
+  mutate(cut_off = case_when(cut_off == 0 ~ "1", 
+                             cut_off == Inf ~ "No correction", 
+                             !(cut_off %in% c(0, Inf)) ~ as.character(cut_off))) %>%
+  select(`Consecutive day threshold` = cut_off, `Root MSE` = mse) -> mse
+mse[which.min(mse$`Root MSE`), ] <- paste0("**", mse[which.min(mse$`Root MSE`), ], "**")
+kable(mse)
+```
+\newpage
+
+## S3. Modeling reported bite incidence
+We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given dataset with a standard deviation of 10.
+For the covariate terms, we centered the priors around zero with a standard deviation of 10.
+For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 5)$).
+We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as the maximum potential scale reduction factors (psrf) for each covariate and the multivariate psrf for the whole model (both are metrics of model convergence, where values < 1.1 are indicative of convergence, Table S3.1) [@gelman2006data].
+
+To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to compare to data.
+In addition, we used the models to predict out-of-fit data (i.e.
+estimates from models fitted to the national data were used to predict the Moramanga data, and estimates from models fitted to the Moramanga data were used to predict the national data).
+Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e.
+assuming forms were completely reported resulting in lower estimates of bite incidence) and with a lower cut-off (7 days, resulting in higher estimates of under-submission of forms).
+
+\newpage
 
@@
 ```
-_**Figure S3.1.** Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale._\
-
-<br/>
-
-*__Table S3.1.__ DIC Estimates for all models.
-Models with the lowest DIC for each dataset and scale of covariate are in bold.
-For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
-For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).*
+### *Fig S3.1. Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale.*
+
+\newpage
+
+### *Table S3.1. DIC and convergence estimates (maximum potential scale reduction factor and multivariate psrf, values < 1.1 indicate convergence) for all models.*
+*For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
+For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).
+The Overdispersion column indicates whether an overdispersion parameter was estimated (yes) or not (no).
+*
 ```{r table S3.1 dic}
 model_ests <- read.csv(here("analysis/out/mods/estimates.csv"))
+
 model_ests %>%
-  filter(OD == FALSE) %>%
-  select(params, Mean, pop_predict, intercept, data_source, scale) %>%
-  spread(key = params, value = Mean, fill = 0) -> model_means
-model_ests %>%
-  group_by(data_source, scale, pop_predict, intercept) %>%
-  summarize(dic = mean(dic)) -> dic_ests
+  group_by(data_source, scale, pop_predict, intercept, OD) %>%
+  summarize(dic = max(dic),
+            psrf_max = max(psrf_upper),
+            mpsrf = max(mpsrf)) -> dic_ests
 dic_ests %>%
-  group_by(data_source) %>%
-  arrange(dic, .by_group = TRUE) %>%
-  mutate(dic = round(dic, 2)) -> dic_ranks
-knitr::kable(dic_ranks, 
-             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type", "DIC"))
-```
-<br/>
-
-```{r fig S3.2 obs vs predicted}
+  group_by(data_source, scale) %>%
+  mutate(min = case_when(dic == min(dic) ~ "Min",
+                         TRUE ~ ""), 
+         OD = case_when(OD == TRUE ~ "yes", 
+                        OD == FALSE ~ "no"),
+         across(where(is.numeric), 
+                function(x) as.character(round(x, 3)))) %>%
+  arrange(dic, .by_group = TRUE) -> dic_ranks
+
+dic_ranks[dic_ranks$min == "Min", ] <- 
+  apply(dic_ranks[dic_ranks$min == "Min", ], 2,  function(x) paste0("**", x, "**"))
+
+knitr::kable(select(dic_ranks, -min), 
+             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type",
+                           "Overdispersion",
+                           "DIC", "Max psrf", "Multivariate psrf"),
+             digits = 3)
+```
+
+\newpage
+
+```{r fig S3.2 obs vs predicted, out.width="65%"}
 include_graphics(here("analysis/figs/supplementary/S3.2_fitted.jpeg"))
 ```
-
-_**Figure S3.2.
-Prediction to data used to fit each model.** Log of the observed bites against the log of predicted bites generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
+### *Fig S3.2. Prediction to data used to fit each model.*
+*Log of the observed bites against the log of predicted bites generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
 Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set  was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level)._\
-
-<br/>
+Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level).*
+
+\newpage
 
@@
 
-_**Figure S3.3.
-Out of fit predictions to data.** Log of the observed bites against the log of predicted bites for data not used to fit the model.
+### *Fig S3.3. Out of fit predictions to data.*
+*Log of the observed bites against the log of predicted bites for data not used to fit the model.
 Predictions were generated by sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
@@
 The third column are predictions from models fitted to the Moramanga data for the national data at the commune and district scale (only fixed intercept models).
-Rows are the type of model structure with respect to the population covariate and colors show which data set was used for fitting as per Figure S3.2._
-
-<br/>
+Rows are the type of model structure with respect to the population covariate and colors show which data set was used for fitting as per Fig S3.2.*
+\newpage
 
@@
 
-_**Figure S3.4.** Posterior estimates of parameters from models with travel time and population as an offset (flatPop in figures S3.2 & S3.3), comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter.
-For the Moramanga model, as data came from a single catchment, models with a random catchment effect ($\sigma_0$) were not fitted._
+### *Fig S3.4. Posterior estimates of parameters from models with travel time and population as an offset.*
+*Comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter  (flatPop in Figs S3.2 & S3.3).
+For the Moramanga model, as data came from a single catchment, models with a random catchment effect ($\sigma_0$) were not fitted.*
+
+ \newpage
 
@@
 
-_**Figure S3.5.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
+### *Fig S3.5. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.* 
+*Generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
 Rows are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and columns are whether the model estimated an overdispersion parameter.
-The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset._\
+The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset.
+Note the different y-axis limits between the fixed and random intercept models.*
+\newpage
 
@@
 
-_**Figure S3.6.** Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter._\
+### *Fig S3.6. Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter.*
+
+\newpage
 
@@
 
-_**Figure S3.7.** Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons for random intercept model without overdispersion vs.
+### *Fig S3.7. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.*
+*For random intercept model without overdispersion vs.
 fixed intercept model with overdispersion, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
-The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset._\
+The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset.*
+
+\newpage
 
@@
 
-_**Figure S3.8.** Posterior estimates for models with population as an offset and an overdispersion parameter.
-Columns show estimates from models fitted to the national dataset (1) corrected for both undersubmission by correcting for periods of at least 7 days where zero patient forms where submiited, (2) correcting for periods of 15 days where zero patient forms where submiited, and (3) with the raw data not correcting for undersubmission.
-._\
+### *Fig S3.8. Posterior estimates for models with population as an offset and an overdispersion parameter.*
+*Columns show estimates from models fitted to the national dataset (1) corrected for both under-submission by correcting for periods of at least 7 days where zero patient forms were submitted, (2) correcting for periods of 15 days where zero patient forms were submitted, and (3) with the raw data not correcting for under-submission.*
+
+ \newpage
 
@@
 
-_**Figure S3.9.** The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis) for models with population as an offset and an overdispersion parameter.
-Panels show predictions from models fitted to the national dataset 1) correcting for periods of at least 7 days where zero patient forms where submiited (a less stringent cutoff resulting in lower estimates of the proportion of forms submitted and thus higher estimates of reported bite incidence), (2) correcting for periods of 15 days where zero patient forms where submitted, as presented in the main analysis, and (3) with the raw data not correcting for undersubmission (resulting in lower estimates of reported bite incidence).
+### *Fig S3.9. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).*
+*For models with population as an offset and an overdispersion parameter.
+Panels show predictions from models fitted to the national dataset 1) correcting for periods of at least 7 days where zero patient forms were submitted (a less stringent cutoff resulting in lower estimates of the proportion of forms submitted and thus higher estimates of reported bite incidence), (2) correcting for periods of 15 days where zero patient forms where submitted, as presented in the main analysis, and (3) with the raw data not correcting for under-submission (resulting in lower estimates of reported bite incidence).
 Predictions were generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
-The points show the data used to fit the models._
+The points show the data used to fit the models.*
+\newpage
 
@@
 
-_**Figure S4.1.** Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis), assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
+### *Fig S4.1. Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis).* 
+*Assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
 The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (where the red horizontal line and black dashed lines intersect).
-The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@leblancclaireRabiesMadagascarThreepronged2019].
-The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2018]._
+The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@sambo2018estimating].
+The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].*
+\newpage
+
+```{r fig S4.2 scaling factors}
+include_graphics(here("analysis/figs/supplementary/S4.2_scaling_rels.jpeg"))
+```
+
+### *Fig S4.2. Range of constrained scaling factors generated for district and commune population size* 
+*Underlying rabies exposures either (A) decreases with increasing population size or (B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
+Lines show the expected relationship, with points showing where administrative units fall along this curve, and maps show how this results in variation in assumed exposure incidence spatially at the commune and district level.*
+\newpage
 
@@
 ```
-_**Figure S5.1.** Comparing metrics for ranking clinics for targeted expansion to 83 additional districts (so that each district has an ARMC).
-We simulated expansion using three different ranking metrics: 1) reduction in mean travel times (green line) 2) the proportion of the population for which travel times were reduced (red dashed line) and 3) the proportion of the population for which travel times were reduced weighted by the reduction in travel times (pink dashed line).
+### *Fig S5.1. Comparing metrics for ranking clinics for targeted expansion.*  
+*We simulated expansion using three different ranking metrics: 1) reduction in mean travel times (green line) 2) the proportion of the population for which travel times were reduced (red dashed line) and 3) the proportion of the population for which travel times were reduced weighted by the reduction in travel times (pink dashed line).
 For each of these, we simulated burden using our decision tree framework (y axis is the mean of 1000 simulations of annual deaths at the national level).
 The blue lines show 10 simulations of randomly expanding access on reducing burden as a comparator.
-The panels show to the commune and district model of reported bite incidence.
-_
+The panels show to the commune and district model of reported bite incidence.*
+\newpage
 
@@
 ```
-_**Figure S5.2.** Map of the location and at what step each clinic was added.
-The circles are each of the CSB II across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
-The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional CSB II in the country which were added in the final step but not ranked._
+### *Fig S5.2. Map of the location and at what step each clinic was added.*
+*The circles are each of the CSB II across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
+The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional CSB II in the country which were added in the final step but not ranked.*
+\newpage
 
@@
 ```
-_**Figure S5.3.** Maps of how travel times change as clinics are added at the **A)** ~1 x 1 km grid cell **B)** commune and **C)** district scales.
+### *Fig S5.3. Maps of how travel times change as clinics are added.*
+*(A) at the ~1 x 1 km grid cell (B) commune and (C) district scales.
 The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional CSB IIs in the country).
 Grey pixels show the location of ARMC at each step.
-Commune and district values are the average grid cell travel times weighted by the population in each cell._
+Commune and district values are the average grid cell travel times weighted by the population in each cell.*
+\newpage
 
@@
 ```
-_**Figure S5.4.** Shifts in estimated **A)** travel times (hrs, x-axis is square root transformed), **B)** bite incidence per 100,000 persons and boxplots showing the median for communes and district models (colors) as clinics are added._
+### *Fig S5.4. Shifts in key metrics as clinics are added.* 
+*(A) travel times (hrs, x-axis is square root transformed), (B) bite incidence per 100,000 persons and boxplots showing the median for communes and district models (colors).*
+ \newpage
 
@@
 ```
-_**Figure S5.5.** Shifts in estimated **A)** reporting, **B)** death incidence per 100,000 persons for the commune and district models (colors) as clinics are added._
+### *Fig S5.5. Shifts in  key metrics as clinics are added.*
+*(A) reporting, (B) death incidence per 100,000 persons for the commune and district models (colors) as clinics are added.*
+\newpage
 
@@
 ```
-_**Figure S5.6.** Shifts in average clinic **A)** catchment population size, **B)** annual vial demand, and **C)** daily throughput (i.e.
+
+### *Fig S5.6. Shifts in key metrics as clinics are added.*
+*(A) catchment population size, (B) annual vial demand, and (C) daily throughput (i.e.
 average number of patients reporting each day) given estimates of bite incidence for the commune and district models (colors).
@@
 population in a grid cell is allocated to the clinic catchment it is closest to in terms of travel times regardless of district or commune).
-All x-axes are log transformed._
+All x-axes are log transformed.*
+\newpage
 
@@
 
-_**Figure S5.7.** Shifts in where bites are reported to as clinics are added for the commune model.
-The circles show the clinic locations for each scenario, with size propotional to the annual average bites reported to that clinic.
+### *Fig S5.7. Shifts in where bites are reported to as clinics are added for the commune model.*
+*The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
 Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the commune level reported bite incidence._
+The polygon shading shows the commune level reported bite incidence.*
+\newpage
 
@@
 
-_**Figure S5.8.** Shifts in where bites are reported to as clinics are added for the district model.
-The circles show the clinic locations for each scenario, with size propotional to the annual average bites reported to that clinic.
+### *Fig S5.8. Shifts in where bites are reported to as clinics are added for the district model.*
+*The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
 Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the district level reported bite incidence._
+The polygon shading shows the district level reported bite incidence.*
+\newpage
 
@@
 
-Table:**Table S6.1.** Range of parameters used in models of bite incidence for the univariate sensitivity analyses.\
+### *Table S6.1. Range of parameters (upper 97.5% and lower 2.5% credible interval of posterior) used in models of bite incidence for the univariate sensitivity analyses.*
 
@@
 
-Table:**Table S6.2.** Range of parameters used in decision tree model for the univariate sensitivity analyses.\
+\newpage
+
+### *Table S6.2. Range of parameters used in decision tree model for the univariate sensitivity analyses.*
 
@@
 | $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se[par == "rho_max"]$range`	 
-| $p_{death}$ | The probability of death given a rabies exposure* | `r dtree_se[par == "p_death"]$range`	
-_*95% CI of estimate from @changaluchaNeedImproveAccess2018 _
+| $p_{death}$ | The probability of death given a rabies exposure^1^ | `r dtree_se[par == "p_death"]$range`	
+^1^*95% CI of estimate from [@changaluchaNeedImproveAccess2018]*
+
+\newpage
 
@@
 
-_**Figure S6.1.** A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B)  estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter._
+### *Fig S6.1. Sensitivity analyses for baseline burden estimates.*
+*(A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter.*
+
+\newpage
 
@@
 
-_**Figure S6.2.** A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter._
+### *Fig S6.2. Sensitivity analysis for PEP expansion.*
+*(A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter.*
+
+ \newpage
 
@@
 
-_**Figure S6.3.** A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter._
-
-```{r fig S6.4 scaling factors}
-include_graphics(here("analysis/figs/supplementary/S6.4_scaling_rels.jpeg"))
-```
-
-*__Figure S6.4.__ Range of constrained scaling factors generated for district and commune population size so that underlying rabies exposures either A) decreases with increasing population size or B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
-Lines show the expected relationship, with points showing where administrative units fall along this curve, and maps show how this results in variation in assumed exposure incidence spatially at the commune and district level.*
-
-```{r fig S6.5 scaling se}
-include_graphics(here("analysis/figs/supplementary/S6.5_scaling_se.jpeg"))
-```
-
-*__Figure S6.5.__ Predicted relationship between A) travel times and rabies death incidence per 100k persons and B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
+### *Fig S6.3. Sensitivity analysis for vial demand.*
+*(A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors) for the two models (colors).
+See Table S6.1 & 2 for ranges used for each parameter.*
+\newpage
+
+```{r fig S6.4 scaling se}
+include_graphics(here("analysis/figs/supplementary/S6.4_scaling_se.jpeg"))
+```
+
+### *Fig S6.4. Sensitivity analysis for exposure scaling with population.*
+*Predicted relationship between (A) travel times and rabies death incidence per 100k persons and (B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
 commune models (colors) with columns comparing the baseline (presented in main analyses) and assumptions of rabies exposure incidence increasing or decreasing with human population size (shapes).
-The points show the mean estimates from 1000 simulations and the lineranges show the 95% prediction intervals.*
-
-## S7. Sotware citations (R Packages used) {#srefs}
+The points show the mean estimates from 1000 simulations and the line ranges show the 95% prediction intervals.*
+
+\newpage
+
+## S7. Sotware citations {#srefs}
 ::: {#refs_supp}
 :::
+
+## S8. Additional supplementary references {#srefextra}
+::: {#refs_supextra}
+:::

### 2020-10-02 analysis/paper/man.Rmd
adding files for archiving on zenodo

--- 2020-10-02 analysis/paper/manuscript.Rmd
@@
 date: ""
-nocite: '@*'
 link-citations: no
@@
       - --lua-filter=format/multiple-bibliographies.lua
-      - --reference-doc=format/reference_supplement.docx
-  bookdown::html_document2:
-    toc: true
-    toc_float: true
-    number_sections: no
-    pandoc_args:
-      - --csl=format/plos-neglected-tropical-diseases.csl
-      - --filter=pandoc-crossref
-      - --lua-filter=format/scholarly-metadata.lua
-      - --lua-filter=format/author-info-blocks.lua
-      - --lua-filter=format/pagebreak.lua
-      - --lua-filter=format/multiple-bibliographies.lua
+      - --reference-doc=format/reference_main.docx
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
-#bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib", "bib/refs_supextra.bib"]
 bibliography_main: ["bib/refs_all.bib"]
-bibliography_supp: ["bib/refs_pkgs.bib"]
 bibliography_supextra: ["bib/refs_supextra.bib"]
 ---
+<!-- Do not edit this file by hand, only for submission! -->
 
 ```{r setupmain, include=FALSE}
+start <- Sys.time()
+
 # Libraries
@@
 
+# for logging dependencies
+out_session("logs/log_local.csv", start = start, 
+            ncores = 1)
 ```
@@
 :::
-
-<!-- split -->
-
-\newpage
-
-# Supplementary Appendix
-
-## S1. Estimating travel times to the closest ARMC {#S1}
-
-```{r fig S1.1 inputs A friction B pop}
-include_graphics(here("analysis/figs/supplementary/S1.1_inputs.jpeg"))
-```
-### *Fig S1.1. Raster inputs to estimate travel times to the closest ARMC for Madagascar.*
-_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale. (B) Population estimates resampled to the same friction surface._
-
-\newpage
-
-```{r fig S1.2 ttimes A Moramanga B Drive}
-include_graphics(here("analysis/figs/supplementary/S1.2_raw.jpeg"))
-```
-
-### *Fig S1.2. Raw patient reported and driving travel time data.*
-_(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
-(B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
-Paths are bezier curves from origin to destination, and do not show actual paths driven._
-
-\newpage
-
-```{r fig S1.3 travel modes}
-include_graphics(here("analysis/figs/supplementary/S1.3_modes.jpeg"))
-```
-
-### *Fig S1.3. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
-\newpage
-
-```{r fig S1.4 lms ttimes}
-include_graphics(here("analysis/figs/supplementary/S1.4_lms.jpeg"))
-```
-
-### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations).*
-_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only)._
-\newpage
-
-### *Table S1.1. R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
-```{r table S1.1 lms table}
-S1.1_table <- read.csv(here("analysis/out/stats/access_met_r2.csv"), row.names = 1)
-kable(S1.1_table, col.names = c("Predictor", "Driving times", 
-                                "Commune mean of patient reported travel times"),
-      digits = 3)
-```
-
- \newpage
-
-```{r fig S1.5 output ttimes A district B commune}
-include_graphics(here("analysis/figs/supplementary/S1.5_outputs.jpeg"))
-```
-
-### *Fig S1.5. Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.*
-
-\newpage
-
-## S2. Estimating bite incidence 
-Vial demand was simulated under simplified assumptions of PEP administration and adherence [@rajeevHealthcareUtilizationProvisioning2019], based on patients reported randomly across the year.
-During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient completing doses on days 0, 3, 7, and 28.
-Vials can be shared within a day between two patients, resulting in 0.1ml wastage per vial shared, plus any additional wastage from unused doses discarded at the end of the day.
-We estimate vial estimates as the midpoint estimate if all patients complete 3 vs.
-4 doses.
-As clinic submission of forms was highly variable from 2014-2017 (Fig S2.2A), we compared estimated demand to the total vials provisioned across this four year period comparing different thresholds for correcting for periods of no form submission (i.e, designating periods of 1, 5, 10, 15, and 30 consecutive days with zero submitted records as missing, compared to no correction).
-
-Estimates of vial demand based on uncorrected bite patient numbers were generally lower than the number of vials provisioned for those clinics with substantial under-submission of forms (Fig S2.2B).
-Correcting patient numbers for under-submission resulted in estimates of vial demand closer to the provisioned vials for most clinics  (Fig S2.2C, Table S2.1).
-
-\newpage
-
-```{r fig S2.1 A catchments B closest C Prop bites}
-include_graphics(here("analysis/figs/supplementary/S2.1_catchments.jpeg"))
-```
-
-### *Fig S2.1. Catchment assignments by travel time.*
-*(A) Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or within a district (polygon outline).
-Admin units where the fill and border colors do not match show places where assigned catchments differ at the district vs commune scale.
-(B) Distribution of the proportion of the population in a given administrative unit (district or commune) served by the catchment assigned.
-(C) The proportion of bites reported to each clinic which originated from a district within the assigned catchment.
-The vertical line indicates the proportion of bites from within the assigned catchment for the Moramanga data (~90%).
-*
-
-\newpage
-
-```{r fig 2.2 form sub A time series B reporting C vial comp}
-include_graphics(here("analysis/figs/supplementary/S2.2_vialcomps.jpeg"))
-```
-
-### *Fig S2.2. Estimating undersubmission of patient forms.*
-*(A) The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
-(B) Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the criteria of 15 consecutive days of non-submission/365.
-(C) The difference between log(estimated) and log(observed) vials provisioned for the period of 2013 - 2017 for each clinic correcting for under-submission (squares) using the 15 day cut-off show in A and B, vs.
-not correcting for under-submission (circles).
-We did not have data on vials provisioned for IPM.*
-\newpage
-
-### *Table S2.1.
-Root mean squared error (MSE) between observed vials provisioned and estimated by the different consecutive day threshold for correcting for periods of no form submission, with the minimum root MSE in bold.*
-```{r table mse stat}
-mse <- read.csv(here("analysis/out/stats/vial_mse.csv"))
-mse %>%
-  mutate(cut_off = case_when(cut_off == 0 ~ "1", 
-                             cut_off == Inf ~ "No correction", 
-                             !(cut_off %in% c(0, Inf)) ~ as.character(cut_off))) %>%
-  select(`Consecutive day threshold` = cut_off, `Root MSE` = mse) -> mse
-mse[which.min(mse$`Root MSE`), ] <- paste0("**", mse[which.min(mse$`Root MSE`), ], "**")
-kable(mse)
-```
-\newpage
-
-## S3. Modeling reported bite incidence
-We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given dataset with a standard deviation of 10.
-For the covariate terms, we centered the priors around zero with a standard deviation of 10.
-For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 5)$).
-We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as the maximum potential scale reduction factors (psrf) for each covariate and the multivariate psrf for the whole model (both are metrics of model convergence, where values < 1.1 are indicative of convergence, Table S3.1) [@gelman2006data].
-
-To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to compare to data.
-In addition, we used the models to predict out-of-fit data (i.e.
-estimates from models fitted to the national data were used to predict the Moramanga data, and estimates from models fitted to the Moramanga data were used to predict the national data).
-Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e.
-assuming forms were completely reported resulting in lower estimates of bite incidence) and with a lower cut-off (7 days, resulting in higher estimates of under-submission of forms).
-
-\newpage
-
-```{r fig S3.1 corr pop ttimes}
-include_graphics(here("analysis/figs/supplementary/S3.1_corrs.jpeg"))
-```
-### *Fig S3.1. Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale.*
-
-\newpage
-
-### *Table S3.1. DIC and convergence estimates (maximum potential scale reduction factor and multivariate psrf, values < 1.1 indicate convergence) for all models.*
-*For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
-For the intercept type: random = random intercept by catchment, fixed = a single fixed intercept was estimated).
-The Overdispersion column indicates whether an overdispersion parameter was estimated (yes) or not (no).
-*
-```{r table S3.1 dic}
-model_ests <- read.csv(here("analysis/out/mods/estimates.csv"))
-
-model_ests %>%
-  group_by(data_source, scale, pop_predict, intercept, OD) %>%
-  summarize(dic = max(dic),
-            psrf_max = max(psrf_upper),
-            mpsrf = max(mpsrf)) -> dic_ests
-dic_ests %>%
-  group_by(data_source, scale) %>%
-  mutate(min = case_when(dic == min(dic) ~ "Min",
-                         TRUE ~ ""), 
-         OD = case_when(OD == TRUE ~ "yes", 
-                        OD == FALSE ~ "no"),
-         across(where(is.numeric), 
-                function(x) as.character(round(x, 3)))) %>%
-  arrange(dic, .by_group = TRUE) -> dic_ranks
-
-dic_ranks[dic_ranks$min == "Min", ] <- 
-  apply(dic_ranks[dic_ranks$min == "Min", ], 2,  function(x) paste0("**", x, "**"))
-
-knitr::kable(select(dic_ranks, -min), 
-             col.names = c("Dataset", "Scale", "Pop effect", "Intercept type",
-                           "Overdispersion",
-                           "DIC", "Max psrf", "Multivariate psrf"),
-             digits = 3)
-```
-
-\newpage
-
-```{r fig S3.2 obs vs predicted, out.width="65%"}
-include_graphics(here("analysis/figs/supplementary/S3.2_fitted.jpeg"))
-```
-### *Fig S3.2. Prediction to data used to fit each model.*
-*Log of the observed bites against the log of predicted bites generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
-Columns are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and rows are the type of model structure with respect to the population covariate (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model).
-Colors show which data set was used for fitting and the scale of the model (Moramanga = Moramanga data with covariates at the commune level, Commune = National data with covariates at the commune level, District = National data with covariates at the district level).*
-
-\newpage
-
-```{r fig S3.3 out of fit}
-include_graphics(here("analysis/figs/supplementary/S3.3_outfit.jpeg"))
-```
-
-### *Fig S3.3. Out of fit predictions to data.*
-*Log of the observed bites against the log of predicted bites for data not used to fit the model.
-Predictions were generated by sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
-The first two columns are the predictions from the commune and district model fitted to the national data for the Moramanga data with fixed and random intercepts.
-The third column are predictions from models fitted to the Moramanga data for the national data at the commune and district scale (only fixed intercept models).
-Rows are the type of model structure with respect to the population covariate and colors show which data set was used for fitting as per Fig S3.2.*
-\newpage
-
-```{r fig S3.4 od comp posts}
-include_graphics(here("analysis/figs/supplementary/S3.4_posts_all.jpeg"))
-```
-
-### *Fig S3.4. Posterior estimates of parameters from models with travel time and population as an offset.*
-*Comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter  (flatPop in Figs S3.2 & S3.3).
-For the Moramanga model, as data came from a single catchment, models with a random catchment effect ($\sigma_0$) were not fitted.*
-
- \newpage
-
-```{r fig S3.5 od comp preds}
-include_graphics(here("analysis/figs/supplementary/S3.5_preds_OD.jpeg"))
-```
-
-### *Fig S3.5. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.* 
-*Generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
-Rows are by the type of model intercept (either a fixed intercept or a random intercept by catchment) and columns are whether the model estimated an overdispersion parameter.
-The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset.
-Note the different y-axis limits between the fixed and random intercept models.*
-\newpage
-
-```{r fig S3.6 random vs od}
-include_graphics(here("analysis/figs/supplementary/S3.6_posts_intercepts.jpeg"))
-```
-
-### *Fig S3.6. Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter.*
-
-\newpage
-
-```{r fig S3.7 random vs od}
-include_graphics(here("analysis/figs/supplementary/S3.7_preds_ODcomp.jpeg"))
-```
-
-### *Fig S3.7. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.*
-*For random intercept model without overdispersion vs.
-fixed intercept model with overdispersion, generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
-The points show the data used to fit the models (the National dataset), as well as the Moramanga dataset.*
-
-\newpage
-
-```{r fig S3.8 rep cutoff posts}
-include_graphics(here("analysis/figs/supplementary/S3.8_posts_rep_se.jpeg"))
-```
-
-### *Fig S3.8. Posterior estimates for models with population as an offset and an overdispersion parameter.*
-*Columns show estimates from models fitted to the national dataset (1) corrected for both under-submission by correcting for periods of at least 7 days where zero patient forms were submitted, (2) correcting for periods of 15 days where zero patient forms were submitted, and (3) with the raw data not correcting for under-submission.*
-
- \newpage
-
-```{r fig S3.9 rep cutoff preds}
-include_graphics(here("analysis/figs/supplementary/S3.9_preds_rep_se.jpeg"))
-```
-
-### *Fig S3.9. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).*
-*For models with population as an offset and an overdispersion parameter.
-Panels show predictions from models fitted to the national dataset 1) correcting for periods of at least 7 days where zero patient forms were submitted (a less stringent cutoff resulting in lower estimates of the proportion of forms submitted and thus higher estimates of reported bite incidence), (2) correcting for periods of 15 days where zero patient forms where submitted, as presented in the main analysis, and (3) with the raw data not correcting for under-submission (resulting in lower estimates of reported bite incidence).
-Predictions were generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
-The points show the data used to fit the models.*
-\newpage
-
-## S4. Range of rabies exposure incidence in people
-
-```{r fig S4.1 exp range}
-include_graphics(here("analysis/figs/supplementary/S4.1_exprange.jpeg"))
-```
-
-### *Fig S4.1. Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis).* 
-*Assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
-The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (where the red horizontal line and black dashed lines intersect).
-The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@sambo2018estimating].
-The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].*
-\newpage
-
-```{r fig S4.2 scaling factors}
-include_graphics(here("analysis/figs/supplementary/S4.2_scaling_rels.jpeg"))
-```
-
-### *Fig S4.2. Range of constrained scaling factors generated for district and commune population size* 
-*Underlying rabies exposures either (A) decreases with increasing population size or (B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
-Lines show the expected relationship, with points showing where administrative units fall along this curve, and maps show how this results in variation in assumed exposure incidence spatially at the commune and district level.*
-\newpage
-
-## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
-
-```{r fig S5.1 metrics}
-include_graphics(here("analysis/figs/supplementary/S5.1_tests.jpeg"))
-```
-### *Fig S5.1. Comparing metrics for ranking clinics for targeted expansion.*  
-*We simulated expansion using three different ranking metrics: 1) reduction in mean travel times (green line) 2) the proportion of the population for which travel times were reduced (red dashed line) and 3) the proportion of the population for which travel times were reduced weighted by the reduction in travel times (pink dashed line).
-For each of these, we simulated burden using our decision tree framework (y axis is the mean of 1000 simulations of annual deaths at the national level).
-The blue lines show 10 simulations of randomly expanding access on reducing burden as a comparator.
-The panels show to the commune and district model of reported bite incidence.*
-\newpage
-
-```{r fig S5.2 add clinics}
-include_graphics(here("analysis/figs/supplementary/S5.2_ARMCadded.jpeg"))
-```
-### *Fig S5.2. Map of the location and at what step each clinic was added.*
-*The circles are each of the CSB II across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
-The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional CSB II in the country which were added in the final step but not ranked.*
-\newpage
-
-```{r fig S5.3 map shifts}
-include_graphics(here("analysis/figs/supplementary/S5.3_map_shifts.jpeg"))
-```
-### *Fig S5.3. Maps of how travel times change as clinics are added.*
-*(A) at the ~1 x 1 km grid cell (B) commune and (C) district scales.
-The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional CSB IIs in the country).
-Grey pixels show the location of ARMC at each step.
-Commune and district values are the average grid cell travel times weighted by the population in each cell.*
-\newpage
-
-```{r fig S5.4 travel times(a) bite inc(b)}
-include_graphics(here("analysis/figs/supplementary/S5.4_ttimes_bites.jpeg"))
-```
-### *Fig S5.4. Shifts in key metrics as clinics are added.* 
-*(A) travel times (hrs, x-axis is square root transformed), (B) bite incidence per 100,000 persons and boxplots showing the median for communes and district models (colors).*
- \newpage
-
-```{r fig S5.5 reporting(a) death inc(b)}
-include_graphics(here("analysis/figs/supplementary/S5.5_reporting_shifts.jpeg"))
-```
-### *Fig S5.5. Shifts in  key metrics as clinics are added.*
-*(A) reporting, (B) death incidence per 100,000 persons for the commune and district models (colors) as clinics are added.*
-\newpage
-
-```{r fig S5.6 catch(a) vials(b) throughput(c)}
-include_graphics(here("analysis/figs/supplementary/S5.6_clinic_shifts.jpeg"))
-```
-
-### *Fig S5.6. Shifts in key metrics as clinics are added.*
-*(A) catchment population size, (B) annual vial demand, and (C) daily throughput (i.e.
-average number of patients reporting each day) given estimates of bite incidence for the commune and district models (colors).
-For vial demand estimation, catchment population sizes are the same for each model as these populations are allocated at the grid cell level (i.e.
-population in a grid cell is allocated to the clinic catchment it is closest to in terms of travel times regardless of district or commune).
-All x-axes are log transformed.*
-\newpage
-
-```{r fig S5.7 comm network}
-include_graphics(here("analysis/figs/supplementary/S5.7_comm_map.jpeg"))
-```
-
-### *Fig S5.7. Shifts in where bites are reported to as clinics are added for the commune model.*
-*The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
-Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the commune level reported bite incidence.*
-\newpage
-
-```{r fig S5.8 dist network}
-include_graphics(here("analysis/figs/supplementary/S5.8_dist_map.jpeg"))
-```
-
-### *Fig S5.8. Shifts in where bites are reported to as clinics are added for the district model.*
-*The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
-Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the district level reported bite incidence.*
-\newpage
-
-## S6. Sensitivity of burden estimates to parameter assumptions {#S6}
-
-```{r se tables}
-se_pars <- read.csv(here("analysis/out/sensitivity/se_pars.csv"))
-
-se_pars %>%
-  pivot_longer(exp_min:beta_0) %>%
-  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
-                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
-  group_by(scale, par = name) %>%
-  summarize(upper = max(value), lower = min(value)) %>%
-  mutate(range = glue("{lower} \U2012 {upper}")) -> se_table_all
-
-se_table_all %>%
-  filter(grepl("beta|sigma", par)) %>%
-  pivot_wider(par, values_from = range, names_from = scale) %>%
-  as.data.table(.) -> bites_se
-se_table_all %>%
-  filter(!grepl("beta|sigma|min", par), scale != "Commune") %>%
-  as.data.table(.) -> dtree_se
-```
-
-### *Table S6.1. Range of parameters (upper 97.5% and lower 2.5% credible interval of posterior) used in models of bite incidence for the univariate sensitivity analyses.*
-
-| **Parameter**   | **Description**        | **Commune model** | **District model** | 
-|:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
-| $\beta_t$ | Travel time effect	| `r bites_se[par == "beta_ttimes"]$Commune` | `r bites_se[par == "beta_ttimes"]$District`
-| $\beta_0$ | Model intercept	| `r bites_se[par == "beta_0"]$Commune` | `r bites_se[par == "beta_0"]$District`
-| $\sigma_0$ | Overdispersion	| `r bites_se[par == "sigma_e"]$Commune` | `r bites_se[par == "sigma_e"]$District`
-
-\newpage
-
-### *Table S6.2. Range of parameters used in decision tree model for the univariate sensitivity analyses.*
-
-| **Parameter**   | **Description**        | **Range**	                 | **Notes**|
-|:----------------|:----------------------:|:----------------------------|:---------:|
-| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se[par == "exp_max"]$range`  	
-| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se[par == "p_rab_max"]$range` 	
-| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se[par == "rho_max"]$range`	 
-| $p_{death}$ | The probability of death given a rabies exposure^1^ | `r dtree_se[par == "p_death"]$range`	
-^1^*95% CI of estimate from [@changaluchaNeedImproveAccess2018]*
-
-\newpage
-
-```{r fig S6.1 base se}
-include_graphics(here("analysis/figs/supplementary/S6.1_base_se.jpeg"))
-```
-
-### *Fig S6.1. Sensitivity analyses for baseline burden estimates.*
-*(A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
-
-\newpage
-
-```{r fig S6.2 add se}
-include_graphics(here("analysis/figs/supplementary/S6.2_addARMC_se.jpeg"))
-```
-
-### *Fig S6.2. Sensitivity analysis for PEP expansion.*
-*(A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
-
- \newpage
-
-```{r fig S6.3 vial se}
-include_graphics(here("analysis/figs/supplementary/S6.3_vial_se.jpeg"))
-```
-
-### *Fig S6.3. Sensitivity analysis for vial demand.*
-*(A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
-\newpage
-
-```{r fig S6.4 scaling se}
-include_graphics(here("analysis/figs/supplementary/S6.4_scaling_se.jpeg"))
-```
-
-### *Fig S6.4. Sensitivity analysis for exposure scaling with population.*
-*Predicted relationship between (A) travel times and rabies death incidence per 100k persons and (B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
-commune models (colors) with columns comparing the baseline (presented in main analyses) and assumptions of rabies exposure incidence increasing or decreasing with human population size (shapes).
-The points show the mean estimates from 1000 simulations and the line ranges show the 95% prediction intervals.*
-
-\newpage
-
-## S7. Sotware citations {#srefs}
-::: {#refs_supp}
-:::
-
-## S8. Additional supplementary references {#srefextra}
-::: {#refs_supextra}
-:::

### 2020-10-02 analysis/paper/manuscript.Rmd
adding files for archiving on zenodo

@@
   - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut pasteur de Madagascar)
+  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut Pasteur de Madagascar)
   - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
@@
 ```{r setupmain, include=FALSE}
+start <- Sys.time()
+
 # Libraries
@@
 Our framework could be used to guide PEP expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted.
-While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyse countries' progress towards achieving zero rabies deaths.
+While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyze progress towards achieving zero rabies deaths.
 
@@
 ```
+
 ### *Fig S1.1. Raster inputs to estimate travel times to the closest ARMC for Madagascar.*
@@
 _Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only)._
+
 \newpage
@@
 ```
+
 ### *Fig S3.1. Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale.*
@@
 ```
+
 ### *Fig S3.2. Prediction to data used to fit each model.*
@@
 ```
+
 ### *Fig S5.1. Comparing metrics for ranking clinics for targeted expansion.*  
@@
 ```
+
 ### *Fig S5.2. Map of the location and at what step each clinic was added.*
@@
 ```
+
 ### *Fig S5.3. Maps of how travel times change as clinics are added.*
@@
 ```
+
 ### *Fig S5.4. Shifts in key metrics as clinics are added.* 
@@
 ```
+
 ### *Fig S5.5. Shifts in  key metrics as clinics are added.*

### 2020-10-02 analysis/paper/supplement.Rmd
adding files for archiving on zenodo

--- 2020-10-02 analysis/paper/manuscript.Rmd
@@
 ---
-title: "How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
+title: "Supplementary Appendix"
+subtitle: "How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
 author:
@@
   - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
-  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut Pasteur de Madagascar)
+  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut pasteur de Madagascar)
   - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
@@
   bookdown::word_document2:
+    toc: true
     pandoc_args:
@@
       - --reference-doc=format/reference_supplement.docx
-  bookdown::html_document2:
-    toc: true
-    toc_float: true
-    number_sections: no
-    pandoc_args:
-      - --csl=format/plos-neglected-tropical-diseases.csl
-      - --filter=pandoc-crossref
-      - --lua-filter=format/scholarly-metadata.lua
-      - --lua-filter=format/author-info-blocks.lua
-      - --lua-filter=format/pagebreak.lua
-      - --lua-filter=format/multiple-bibliographies.lua
 keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
-#bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib", "bib/refs_supextra.bib"]
-bibliography_main: ["bib/refs_all.bib"]
 bibliography_supp: ["bib/refs_pkgs.bib"]
@@
 ---
-
+<!-- Do not edit this file by hand, only for submission! -->
 ```{r setupmain, include=FALSE}
@@
 
-\newpage
-
-# Abstract
-
-## Background
-Post-exposure prophylaxis (PEP) is highly effective at preventing human rabies deaths, however access to PEP is limited in many rabies endemic countries.
-The 2018 decision by Gavi to add human rabies vaccine to its investment portfolio should expand PEP availability and reduce rabies deaths.
-We explore how geographic access to PEP impacts the rabies burden in Madagascar and the potential benefits of improved provisioning.
-
-## Methodology & Principal Findings
-We use travel times to the closest clinic providing PEP (N=31) as a proxy for access.
-We find that travel times strongly predict reported bite incidence across the country.
-Using resulting estimates in an adapted decision tree framework we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
-We estimate 960 human rabies deaths annually (95% Prediction Intervals (PI):`r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with PEP averting an additional 800 deaths (95% PI: `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) each year.
-Under these assumptions, we find that expanding PEP to one clinic per district could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major health centers provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
-Our quantitative estimates are most sensitive to assumptions of underlying rabies exposure incidence, but qualitative patterns of the impacts of travel times and expanded PEP access are robust.
-
-## Conclusions & Significance
-PEP is effective at preventing rabies deaths, and in the absence of strong surveillance, targeting underserved populations may be the most equitable way to provision PEP.
-Our framework could be used to guide PEP expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted.
-While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyze progress towards achieving zero rabies deaths.
-
-# Author Summary 
-Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low- and middle-income countries where people have limited access to both human vaccines (post-exposure prophylaxis or PEP) and dog rabies vaccines.
-Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by 2030, and recently, Gavi, a multilateral organization that aims to improve access to vaccines in the poorest countries, added human rabies vaccine to it’s portfolio.
-In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP, and extrapolate human rabies deaths in Madagascar.
-We find that PEP currently averts around `r round(natl_base$averted_mean[1], -2)` deaths each year, but that the burden remains high (`r round(natl_base$deaths_mean[1], -2)` deaths/ year), particularly in remote, hard-to-reach areas.
-We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone will not eliminate deaths.
-Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other Low- and Middle-Income countries, forward on the path to rabies elimination.
-
-\newpage
-
-# Introduction
-
-Inequities in access to care are a major driver of disease burden globally [@world2018world].
-Often, the populations at greatest risk of a given disease are the most underserved [@peters2008poverty].
-Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to allocate limited resources [@mainaSpatialDatabaseHealth2019].
-
-Canine rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans.
-While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid animal [@RabiesVaccinesWHO2018].
-However, access to human vaccine is limited in many countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
-
-In 2015, a global framework to eliminate deaths due to canine rabies by 2030 ('Zero by 30') through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero].
-Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS].
-From 2021, Gavi-eligible countries should be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.
-
-A primary challenge in expanding access effectively is the lack of data on rabies exposures and deaths in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017].
-Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in remote  and marginalized communities [@taylor2017difficulties].
-Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites at clinics provisioning PEP and a decision tree framework to extrapolate deaths, assuming that overall reported bite incidence (i.e.
-both bites due to non-rabid and rabid animals) is proportional to rabies incidence (i.e.
-the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting to clinics for PEP is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2019].
-If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.
-
-In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to it's own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019].
-Other than at these 31 anti-rabies medical centers (ARMC), PEP is not available at any other public clinics or through the private sector.
-In addition, there is limited control of rabies in dog populations and the disease is endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011].
-Due to the spatially restricted nature of PEP provisioning and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
-Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
-Here, we provide revised estimates of human rabies deaths by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
-This framework may usefully apply to other countries where PEP availability is currently geographically restricted in considering how to most effectively and equitably provision these life-saving vaccines.
-
-# Methods
-
-## Estimating geographic access to ARMC {#methods-ttimes}
-To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two rasters: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B), which we aggregated to the friction surface.
-
-From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018].
-We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and euclidean distance, i.e.
-the minimum distance from the administrative unit centroid to any ARMC.
-We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018).
-We compared travel times and distance estimates to driving times collected by IPM during field missions and patient reported travel times from a subset of Moramanga ARMC patients (see Fig S1.2 for raw data).
-
-## Estimating bite incidence {#methods-bitedata}
-We used two datasets on bite patients reporting to ARMC:
-
-1)	A national database of individual bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
-These forms were submitted with frequencies ranging from monthly to annually, included the patient reporting date and were resolved to the district level (patient residence).
-
-2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga District ARMC resolved to commune level.
-
-For the national data, some clinics did not submit any data, or had substantial periods (months to a whole year), with no submitted data.
-To correct for this, we exclude periods of 15 consecutive days with zero submitted records (see Supplementary Appendix, section S2).
-For each clinic we divided the total number of bites reported in a given year by the estimated proportion of forms which were not submitted (under-submission).
-Due to yearly variation in submissions, we took the average of annual bite incidence estimates aggregated to district level.
-We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see Supplementary Appendix, section S2).
-At both the commune and district level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit.
-For national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.
-
-## Modeling reported bite incidence {#methods-bitemods}
-
-We modeled the number of reported bites as a function of travel time ($T$) using a Poisson regression:
-
-$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
-$$ y_i = Poisson(\mu_i) $$
-
-where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the ARMC as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
-We fit this model to both the national data (district level) and the Moramanga data (commune level).
-To more directly compare estimates between datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$):
-
-$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ \ + \ \beta_{0j})}P_j$$
-
-As travel times are correlated with population size (Fig S3.1), we also compared how well bites were predicted by population size alone, and in combination with travel times.
-For the models with population size, we removed the offset and used either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors.
-
-For the models fit to the national data, we also modeled variation between clinics with a catchment random effect: $B_{0,k} \sim norm(\mu,\ \sigma_0)$), where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept.
-
-We tested whether the catchment random effect captured overdispersion in the data (i.e.
-variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
-
-$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j \ + \ \sigma_e)}P_i$$
-
-where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters for a given model. 
-We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’ [@R-rjags].
-We used model estimates to generate fitted and out-of-fit predictions, and examined the sensitivity of estimates to adjustments for under-submission of forms (Supplementary Appendix, section S3).
-
-## Modeling human rabies deaths {#methods-burden}
-We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1).
-To model uncertainty in parameter estimates we used triangular distributions, as with previous studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual exposure incidence per administrative unit, and  $p_{rabid}$, the proportion of reported bites that are rabies exposures.
-
-```{r fig 1, out.width="100%"}
-include_graphics(here("data-raw/misc/figM1/figM1_final.tiff"))
-```
-
-### *Fig 1. Decision tree for burden estimation.* 
-*For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted reported bites ($B_i$).
-To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures.
-$p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting of patients to clinics cannot exceed a maximum ($p_{max}$).
-$R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{death}$) to get deaths due to rabies ($D_i$).
-Similarly, deaths averted by PEP, $A_i$, are estimated by multiplying $R_i$ by $p_{death}$, i.e.
-those who would have died given exposure, but instead received PEP.
-Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution.
-Parameter values and sources are in Table 1.*
-
-For $E_i$, we center the distribution at the lower end of our estimated exposure incidence from the Moramanga District (42 exposures/100,000 persons), with a range applied assuming 1% rabies incidence in dogs (estimated across a range of human-to-dog ratios between 5 - 25) and that on average a rabid dog exposes 0.39 persons [@whorabiesmodelingconsortiumPotentialImpactImproved2019] (see Fig S4.1).
-As there is little data on dog population size and human exposure incidence in Madagascar[@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
-
-For $p_{rabid}$, we use a range between 0.2 - 0.6 estimated from a study of bite patients in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].
-So that rabid reported bites cannot exceed the total expected number of rabies exposures or a maximum reporting (even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$:
-
-$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
-
-where $\rho_{max}$ is the maximum reporting, estimated from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where we find that approximately 2% of rabies exposures go unreported [@rajeevHealthcareUtilizationProvisioning2019].
-
-We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing rabies.
-Fig 1 describes the decision tree, the key inputs, and outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).
-Table 1 list all inputs and their sources.
-
-```{r exps}
-exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
-rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
-```
-
-_**Table 1. Parameters used in the decision tree to estimate human rabies deaths at the administrative level.**_\
-
-| **Parameter**   | **Value**        | **Description**	                 | **Source**|
-|:----------------|:----------------:|:----------------------------------|:---------:|
-| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
-| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
-| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|
-
-## Estimating the impact of expanding PEP provisioning {#methods-incremental} 
-We developed a framework to rank clinics by how much their PEP provision improves access for underserved communities, estimating incremental reductions in burden and increases in vaccine demand.
-Specifically, we aggregated our model-predicted estimates of annual bites to the clinic level.
-As multiple clinics may serve a single district or commune, we allocated bites to clinics according to the proportion of the population in each administrative unit which were closest.
-For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
-patients report for all doses) to generate subsequent vaccination dates.
-We use these dates to estimate vial usage given routine vial sharing practices in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], but assuming adoption of the WHO-recommended abridged intradermal regimen (2 x 0.1 ml injections on days 0, 3, and 7 [@tarantolaIntradermalRabiesPostexposure2018]).
-For both burden and vial estimates, we take the mean of 1000 simulations as each clinic is added.
-
-We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic.
-We select the CSB II (Centre de Sante Niveau II, major health centers with immunization capacity) in the highest density grid cell of the administrative unit as candidates for expansion.
-For the `r comms_no_csb2` communes without a CSB II, we chose the CSB I (secondary health posts) in the highest density grid cell.
-`r comms_no_clins` communes lacked any CSB I or II.
-Finally, we explore a scenario where all additional CSB II (totalling `r nrow(csbs_max)`) provision PEP.
-
-We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from an existing ARMC for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing ARMC.
-We simulated expansion of ARMC to each district using these three metrics, and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
-For the full simulation of expanded access, once clinics reduced travel times for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` living greater than $x$ hrs away, starting with $x$ = 3 hrs), we reduced the travel time threshold by 25%.
-
-## Sensitivity analysis {#methods-sensitivity}
-To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analysis of both parameters from the models of bite incidence and the decision tree (see Table S6.1 & S6.2 for parameter ranges used).
-We also examined how systematic variation in rabies incidence with human population size affected burden estimates.
-Specifically, if human-to-dog ratios are positively correlated with human populations (i.e.
-dog ownership/populations are higher in more populated, urban areas), we might expect higher rabies exposure incidence as population size increases.
-Alternatively, if human-to-dog ratios inversely correlate with population size (i.e.
-dog ownership is more common in less populated, rural areas), we might expect exposure incidence to scale negatively with population size.
-We use scaling factors to scale incidence either positively or negatively with observed population sizes at the district and commune levels, while constraining them to the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, Fig S4.2) and simulated baseline burden, as well as expanded PEP access.
-
-## Data and analyses 
-All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (Supplementary appendix, section S7).
-All processed data, code, and outputs are archived (here)[zenodo] and maintained at github.com/mrajeev08/MadaAccess.
-The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC.
-These databases were last queried on September 19, 2019 for these analyses.
-The IPM GIS unit provided the data on geolocated clinics across the country.
-Anonymized raw bite patient data and full data on geolocated clinics are available from IPM following insitutional data transfer protocols.
-Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (wisemonkeyfoundation.org) on the same date and are shared in the archived repository.
-
-## Ethics statement
-Data collection from the Moramanga District was approved by the Princeton University IRB (7801) and the Madagascar Ministry of Public Health Ethics Committee (105-MSANP/CE).
-Oral informed consent was obtained from all interviewed participants.
-Data collected from bite patients at the national level are maintained jointly by the Ministry of Health and IPM as a routine part of PEP provisioning.
-
-# Results
-
-## Estimating access to ARMC {#results-ttimes}
-```{r ttime stats}
-ttime_stats <- read.csv(here("analysis/out/stats/prop_pop_ttimes.csv"))
-
-# get prop of pop > 3 hrs away at baseline
-ttime_stats %>%
-  filter(!(cut_times %in% c("< 1", "1 - 2", "2 - 3"))) %$%
-  sum(prop_pop) -> prop_over3_base
-```
-
-Estimates of travel times to the closest ARMC varied greatly (Fig 2A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live over 3 hrs from a clinic (Fig 2B).
-We found that travel time estimates from the friction surface underestimated both driving times across the country and patient-reported travel times to the Moramanga ARMC (Fig 2C), but were correlated with ground truthed driving and patient-reported times (Fig 3C, Fig S1.4).
-Patient-reported travel times were also significantly more variable than estimates from the friction surface (Fig S1.2).
-The friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
-the presence of a road indicates that all travel through that grid cell is by vehicle).
-However, Moramanga ARMC patients reported using multiple modes of transport, with some individuals walking days to a clinic (Fig S1.3).
-Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (Table S1.1).
-Therefore we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
-
-```{r fig M2}
-include_graphics(here("analysis/figs/main/M2_ttimes.jpeg"))
-```
-
-### *Fig 2. Travel times to ARMC across Madagascar.*
-*(A) Estimated at an ~ 1 km2 scale.
-(B) Distribution of the population across travel times.
-(C) Correlation between ground-truthed travel times (mean of patient reported travel times to the Moramanga ARMC at the commune level and reported driving times between GPS points) and friction surface travel time estimates.
-The vertical lines show the 95% quantiles for reported travel times and the point size shows the number of observations for each commune.
-The best fit lines (red and grey) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
-The dashed black line is the 1:1 line.*
-
-## Estimating bite incidence {#results-biteinc}
-
-Most patients from each district reported to their closest ARMC by the weighted travel time metric (Fig 3).
-Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
-While there are discrepancies between commune and district catchment assignments (Fig S2.1A), over 75% of the population in a given district or commune were closest to a single clinic (Fig S2.1C).
-We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms (see Supplementary Appendix Section S2).
-After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D).
-
-```{r fig M3 catchmaps}
-include_graphics(here("analysis/figs/main/M3_bitemaps.jpeg"))
-```
-
-### *Fig 3. The network of patient presentations and estimates of annual bite incidence.*
-*(A) at the district level for the national data and (B) commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
-Color corresponds to the clinic catchment.
-Circles with a white outline are the total number of bites reported for that administrative unit (plotted as the centroid).
-Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded.
-Out-of-catchment reporting is indicated where points and line colors are mismatched.
-For panel (A) districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey.
-For (B) the inset of Madagascar shows the location of the enlarged area plotted, which shows the district of Moramanga (outlined in black), all communes included in it's catchment (red polygons), and other communes where bites were reported to colored by their catchment  (C) The estimated average annual bite incidence from the national and Moramanga data plotted at the district scale (both datasets) and at the (D) commune scale (Moramanga dataset).
-Colors correspond to the clinic catchment, shape indicates the dataset, and the size of the point indicates the number of observations (i.e.
-the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).
-The point lines indicate the range of estimated bite incidence for each district.*
-
-Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
-For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
-While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.
-
-## Modeling reported bite incidence {#results-bitemods}
-
-Travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales (Fig 4).
-Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
-
-For the national data, including a catchment random effect improved predictions (Fig S3.2 & Fig S3.3).
-However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table S3.1) and the models resulted in similar predictions (Fig S3.6 & S3.7), indicating that catchment effects could not be differentiated from random variation in the data.
-Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
-
-As the predictions from the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
-Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig S3.3), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
-Finally, we examined the sensitivity of models to how we corrected for underreporting of data, and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Fig S3.8 and Fig S3.9).
-
-```{r fig 4}
-include_graphics(here("analysis/figs/main/M4_mods.jpeg"))
-```
-
-### *Fig 4. Travel times as a predictor of reported bite incidence per 100,000 persons.*
-*(A) The estimated relationship between travel time in hours (x-axis) and mean annual reported bite incidence (y-axis).
-The lines are the mean estimates and the envelopes are the 95% prediction intervals generated by drawing 1000 independent samples from the parameter posterior distributions for three candidate models: model with travel times at the 1) commune- and 2) district-level fitted to the national data with an overdispersion parameter ($\sigma_e$) and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept and unadjusted for overdispersion.
-The points show the data: National data (circles) at the district level used to fit the District and Commune models, and Moramanga data (triangles) at the commune level used to fit the Moramanga model.
-(B) The posterior distribution of parameters from the respective models for the model intercept, travel time effect, and for overdispersion (national data only).*
-
-## Estimating human rabies deaths {#results-burden}
-Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar.
-Our estimates vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Fig 5A).
-Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
-
-### *Table 2.
-Model predictions of average annual reported bite incidence, total deaths, and deaths averted at the national level for the two models (commune level and district level models with travel time predictor and an overdispersion parameter); 95% prediction interval in parentheses.*
-
-```{r Table 2}
-pop <- sum(mada_districts$pop)
-natl_base %>%
-  mutate_at(vars(starts_with("bites")), function(x) x/pop*1e5) %>%
-  mutate_at(vars(bites_mean:averted_lower), round) %>%
-  mutate(bites_100k = glue("{bites_mean} ({bites_lower} - {bites_upper})"),
-         deaths = glue("{deaths_mean} ({deaths_lower} - {deaths_upper})"),
-         averted = glue("{averted_mean} ({averted_lower} - {averted_upper})")) %>%
-  select(scale, bites_100k, deaths, averted) -> table
-
-kable(table, digits = -1, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
-                                        "Deaths averted"))           
-```
-
-```{r fig 5, out.width="50%"}
-include_graphics(here("analysis/figs/main/M5_burden_base.jpeg"))
-```
-
-### *Fig 5. Spatial variation in predicted incidence of human rabies deaths per 100,000 persons.*
-*(A) for each district (y-axis) in Madagascar.
-Diamonds show the predicted incidence for the district model and squares show predicted incidence for the commune model fit to the National data for all communes in a district.
-Points are colored and districts ordered by travel times.
-The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
-Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current ARMC.*
-
-## Estimating the impact of PEP provisioning {#results-incremental}
-
-For a subset of additional clinics (n = 83, up to one per district), we compared three methods of ranking for expanding PEP provisioning.
-We found that targeted expansion of PEP to clinics based on the proportion of the population they reduced travel times for resulted in fewest deaths (Fig S5.1).
-Therefore, we used this approach for simulating expansion of PEP to a larger set of clinics (N = `r max_added` ).
-Here we report results from the commune model, as estimates were consistent across models (Fig 6 and Supplementary appendix, section S5).
-
-We estimated that strategic PEP expansion to these additional 83 clinics (1 per district) reduced rabies deaths by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 6A).
-With one clinic per commune (where available, N = 1696), we see a further reduction of `r round((1 - props$deaths_mean[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`%).
-However, reductions in burden saturate as more clinics are added (Fig S5.2).
-Even when all CSB II provisioning PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5); as more clinics are added, reported bite incidence saturates (Fig S5.4), and patients shift which clinic they report to (S5.7 & S5.8).
-
-Vial demand also outpaces reductions in burden (Fig 6B), with more vials needed per death averted (Fig 6C).
-Our model predicts an increase from `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)})")` per annum under current provisioning but with the abridged intradermal regimen (i.e.
-visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")`  with 250 clinics providing PEP, and ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` if all CSB IIs provision PEP.
-In these scenarios, clinic catchment populations and throughput decrease, with clinics seeing fewer patients per day (S5.6).
-
-```{r fig 6}
-include_graphics(here("analysis/figs/main/M6_addARMC.jpeg"))
-```
-
-### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
-*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level , and (C) increase in vials needed per death averted based on the two models of reported bites.
-Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The points show when all additional csb (N = `r max_added`) clinics have been added).*
-
-## Sensitivity Analyses {#results-sensitivity}
-To quantify which parameters contribute to the uncertainty in our estimates, we performed univariate sensitivity analyses, varying parameters in both the bite incidence model (Table S6.1) and the decision tree (Table S6.2).
-These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute most uncertainty to burden estimates (Fig S6.1) and impacts of PEP access (Fig S6.2).
-Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden.
-For the estimates of vial demand, uncertainty around the model intercept (i.e.
-the baseline reported bite incidence) has most impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
-Finally, scaling of incidence with population size (Fig S4.2) modulates the impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
-more rabies in more populated places) dampening and negative scaling exacerbating the relationship between access and deaths (Fig S6.4A).
-However, the impact of adding clinics remains broadly the same (Fig S6.4B).
-Overall, while uncertainty in the underlying rabies exposure incidence results in considerable variation in our burden estimates, the projected impact of travel times and of access to PEP are qualitatively similar across parameter assumptions.
-
-# Discussion
-
-## Main findings
-
-We find that the burden of rabies in Madagascar is likely concentrated in areas with poor PEP access.
-We estimate that current PEP provisioning (at 31 clinics) averts `r glue("{trans_perc(min(prop_averted))}%")`% of deaths that would otherwise occur, and that expanding PEP access should reduce mortality, with provisioning in one clinic per district (N = `r one_per_dist`), or per commune (N = `r one_per_comm`), expected to reduce mortality by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")` and `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")`, respectively.
-However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all CSB II (N = `r max(natl_preds$scenario)`).
-This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all major health centers.
-With reduced travel times, over 20% of exposures will still not seek PEP (Fig S5.5), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
-PEP is expected to remain cost-effective as provisioning expands, to a maximum of 450 USD per death averted (assuming 5 USD per vial), similar to other estimates [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-While our quantitative predictions depend on assumptions of underlying rabies exposure incidence, the qualitative patterns regarding travel time impacts remain robust and are useful in identifying strategies for provisioning PEP.
-
-## Limitations
-Data limitations introduced bias and uncertainty to our estimates.
-For example, travel times from the MAP friction surface underestimated patient-reported travel times, with discrepancies between assigned transport speeds (from the Open Street Map user community, or country cluster data [@weissGlobalMapTravel2018]) and realities of local travel.
-In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used.
-Moramanga ARMC patients reported various transport methods and highly variable travel times even within a single commune.
-While patient-reported travel times may lack precision from recall and estimation error, they likely better reflect lived experience; validated travel times [@rayAccessModComputingGeographic2008, ihantamalalaImprovingGeographicalAccessibility2020] could improve estimates of spatial health inequities.
-Similarly, modeled estimates of population distribution [@linardPopulationDistributionSettlement2012a] also introduce uncertainty.
-Our analysis of data from the Moramanga District indicate that variation at the sub-district level is high and impacts health seeking behavior.
-However, we lacked fine-scale data from other catchments for comparison.
-Additionally, we had to correct for underreporting and incomplete data; strengthening surveillance and routine data collection should improve understanding of health seeking behavior and access, and support monitoring and evaluation of PEP provisioning.
-
-We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
-In Madagascar, where PEP is free-of-charge, the main cost to patients is transport and time lost.
-More remote communities tend to be of lower socioeconomic and educational status [@peters2008poverty], so travel time may be a proxy for these correlated variables.
-We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
-Most ARMC charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers.
-Significant overdispersion in the data that cannot be explained by travel times suggests that clinic-level variation (e.g.
-vaccine availability and charges) and regional differences (e.g.
-dog populations, outbreaks, awareness) further influence health-seeking behavior and vaccine demand.
-
-Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths.
-Our simplifying assumptions, regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP, are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
-Likewise we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care and human behavior, as illustrated by the complex case studies in Box 1.
-
-## Box 1: case studies of of health seeking behavior for PEP in Madagascar
-```{r case studies}
-mora_bites %>%
-  filter(district == "Anosibe-An'ala") %>%
-  summarize(pop = sum(pop, na.rm = TRUE), bites = sum(avg_bites, na.rm = TRUE), 
-            inc = bites/pop*1e5) -> aa_bites
-mora_raw <- read.csv(here("data-raw/out/bitedata/moramanga.csv"))
-mora_raw %>%
-  filter(distcode == "MG33317") %>%
-  summarize(prop_cat1 = sum(known_cat1)/n()) -> prop_cat1_AA
-admin_preds <- fread(here("analysis/out/preds/admin_preds_base.gz")) 
-admin_preds %>%
-  mutate(distcode = substr(names, 1, 7)) %>%
-  filter(distcode == "MG33317") %>%
-  group_by(scale) %>%
-  summarize(inc = sum(deaths_mean)/sum(pop)*1e5, bite_inc = sum(bites_mean)/sum(pop)*1e5) -> aa_preds
-
-ranked_clins <- fread(here("analysis/out/ttimes/addclinics/clinics_ranked.csv"))
-aa_added <- min(ranked_clins[district == "Anosibe-An'ala"]$scenario)
-gtruth <- fread(here("analysis/out/ttimes/gtruth_ttimes.csv"))
-gtruth %>%
-  filter(district == "Anosibe-An'ala") %>%
-  summarize(mean = mean(ttimes_reported, na.rm = TRUE), 
-            min =  min(ttimes_reported, na.rm = TRUE),
-            max = max(ttimes_reported, na.rm = TRUE)) -> aa_ttimes
-admin_base %>%
-  mutate(distcode = substr(names, 1, 7)) %>%
-  filter(distcode == "MG33317") %>%
-  group_by(scale) %>%
-  summarize(deaths_mean = sum(deaths_mean), pop = sum(pop)) -> aa_deaths
-
-aa_exposures <- read.csv(here_safe("data-raw/misc/aa_exposures.csv"))
-nrabid <- nrow(aa_exposures[aa_exposures$suspect.exposure %in% c("Probable",
-                                                                 "Confirmed"), ])
-ninterviewed <- nrow(aa_exposures[!(aa_exposures$suspect.exposure == ""), ])
-
-```
-
-1.	Anosibe An’ala District (population ~ `r format(round(mada_districts$pop[mada_districts$district == "Anosibe-An'ala"]), big.mark = ",")`), south of Moramanga, has moderate incidence of bite patients (~ `r round(aa_bites$inc)`/100,000 persons) even though travel times often exceed 24 hours.
-While a road connects the main Anosibe An’ala commune to the Moramanga ARMC, it is only passable by large trucks during much of the rainy season, with speeds usually < 10km per hour.
-Over `r trans_perc(prop_cat1_AA$prop_cat1)`% of patients from Anosibe An’ala had been in close proximity or touched a person that died from rabies (four suspect human rabies deaths of patients who did not receive any PEP), whilst of patients with Category II and III exposures that were interviewed, `r glue("{nrabid}/{ninterviewed}") ` (`r trans_perc(nrabid/ninterviewed)`%) were bitten by probable rabid dogs.
-Given the high travel times (although underestimated by the friction surface) and incidence of reported rabies exposures and deaths, we predict a large but unobserved rabies burden in this remote community (~`r round(aa_deaths$deaths_mean[1], 2)` deaths per year) and we ranked an ARMC provisioning PEP in Anosibe An’ala `r aa_added`th for travel time reductions.
-Other remote communities likely experience similar high and unrecognized burden, but improved surveillance is necessary to identify such areas.
-Notably, bite patients in this district demonstrate willingness to travel for free PEP (in some cases walking 3 days to a clinic) with awareness of rabies risk.
-Community outreach and active surveillance in other remote areas could also greatly improve people's awareness of risk and health seeking behavior.
-
-2.
-Recently, a middle aged taxi driver died of rabies in suburban Antananarivo.
-The day after being bitten by an unknown dog, he reported to a clinic that referred him to the ARMC at IPM, approximately one hour’s drive from his home.
-His family urged him to get PEP, but he did not believe his risk was high and decided not to seek further care.
-He developed symptoms two weeks later and was confirmed as a rabies death by the National Rabies Reference Laboratory.
-Despite prompt reporting and referral, and socioeconomic indicators suggesting a high care-seeking probability, this person did not receive PEP.
-His story highlights the need for sensitization about rabies and how PEP at peripheral clinics (even in areas with reasonable access) could prevent additional deaths, but also how PEP alone is unlikely to prevent all rabies deaths.
-
-## Broader context
-
-Recent studies have estimated access to health-seeking behavior and PEP completion and adherence, but not directly linked these estimates to burden [@tarantolaRabiesPostexposureProphylaxis2017; @zaidiGeographicVariationAccess2013].
-Our approach for incorporating access to vaccines (echoing [@Hegde2019dt; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @metcalfTransportNetworksInequities2015; @aleganaTreatmentseekingBehaviourLow2017]) into burden estimation methods could guide provisioning of PEP to maximize impacts.
-This approach will have most value in settings with limited PEP access and poor health seeking, but will be less valuable where rabies exposures make up a small fraction of patients reporting for PEP e.g.
-[@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015].
-Our revised estimate of rabies deaths in Madagascar using this approach was higher than previous [@rajeevHealthcareUtilizationProvisioning2019], which assumed uniform reporting of 85%, but remained within the range of other empirical and modeling studies from low-income countries [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @lyRabiesSituationCambodia2009; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011].
-
-Our estimates of vial demand depend on use of the new abridged intradermal regimen [@tarantolaIntradermalRabiesPostexposure2018], which has been adopted by the Ministry of Health in Madagascar.
-However, most ARMC staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moromanga [@rajeevHealthcareUtilizationProvisioning2019].
-We predict that as clinics are expanded, throughput (daily patients reporting to a clinic) will decrease.
-This may make provisioning PEP more challenging and vial demand less predictable, leading to stock outs or wastage.
-Decentralized provisioning mechanisms, for example adopting routine childhood vaccine supply chains, or novel vaccine delivery methods such as drones [@boeras2019use], may mitigate these challenges.
-When nerve tissue vaccines were used in Madagascar, clinics requested vaccine upon demand and PEP access was more widespread, but provisioning the more expensive cell culture vaccines to all clinics became too costly [@andriamandimbySurveillanceControlRabies2013].
-Widespread vaccine provisioning is therefore feasible given Madagascar’s health infrastructure, if cost barriers are removed.
-
-```{r}
-catch_trends <- fread(here("analysis/out/stats/trend_catches.csv"))
-catch_trends %>% 
-  group_by(scenario) %>% 
-  summarize(mean_pops = mean(pops_mean)) %>%
-  as.data.table() -> catch_pops
-```
-
-Gavi investment could greatly reduce the access and cost barriers to PEP [@changaluchaNeedImproveAccess2018; @liDescriptiveAssessmentRabies2019; @sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017].
-Currently, each clinic in Madagascar serves an average catchment of `r format(round(mean(catch_trends[scenario == 0]$pops_mean), -4), big.mark = ",")` persons.
-Latin American countries, where significant progress has been made towards elimination, aim for one PEP clinic per 100,000 persons.
-In Madagascar this would require around `r min(catch_pops[mean_pops < 1e5]$scenario)` additional ARMC.
-We predict that Gavi investment would be highly cost-effective, greatly reducing deaths by expanding PEP supply to underserved areas.
-
-However, our results suggest that PEP expansion alone cannot prevent the majority of rabies deaths, and even given maximal access, achieving ‘the last mile’, preventing deaths in the most remote populations, will require disproportionate resources [@delriovilasTribulationsLastMile2017].
-To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission, and eliminating deaths.
-Integrated Bite Case Management (IBCM) uses bite patient risk assessments to determine rabies exposure status, guide PEP administration, and trigger investigations of rabid animals, potentially identifying other exposed persons [@lechenneImportanceParticipatoryIntegrated2017; @rajeevHealthcareUtilizationProvisioning2019; @undurragaCostEffectivenessEvaluationNovel2017].
-IBCM is one way to manage PEP effectively [@rysavaPathRabiesElimination2018] and as it relies on exposed persons reporting to clinics, expanding PEP access could strengthen this surveillance framework.
-These same issues of access, however apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].
-
-## Conclusion
-Our study suggests that rabies deaths in Madagascar disproportionately occur in communities with the poorest access to PEP and that expanding PEP access should reduce deaths.
-Without data on rabies incidence and exposure risk, targeting PEP expansion to underserved areas is a strategic way to reduce  rabies burden and provide equitable access, for example, by expanding provisioning to clinics serving populations that target an evidence-based travel time threshold or catchment size.
-Implementing outreach programs to raise awareness should further increase the efficacy of PEP expansion by improving care seeking.
-Better surveillance is also needed to understand the geographical distribution of rabies exposures and identify populations most at risk.
-Gavi investment could support countries to more equitably provision PEP and overcome barriers to access ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies), but as PEP alone cannot prevent all rabies deaths, investment should be used to catalyse mass dog vaccination to interrupt transmission, and eventually eliminate rabies deaths.
-
-# Acknowledgements
-
-We thank all the clinicians and staff at the ARMC across the country.
-We are grateful to IPM and the Ministry of Public Health who collect and maintain data on PEP provisioning.
-In particular, we thank the GIS unit for assistance with spatial data, Michael Luciano Tantely for sharing driving time data, and Claire Leblanc, Rila Ratovoson, and Daouda Kassie for sharing results of their work in the Moramanga District.
-In addition, we thank Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, Amy Winter, Christian Ranaivoson, John Friar, and Amy Wesolowski for assistance.
-This work was funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to CJEM and MR.
-MR is supported by an NSF Graduate Research Fellowship.
-KH is supported by the Wellcome Trust (207569/Z/17/Z).
-
-\newpage
-
-# References 
-::: {#refs_main}
-:::
-
 <!-- split -->
@@
 \newpage
-
-# Supplementary Appendix
 
@@
 _Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only)._
-
 \newpage

### 2020-10-02 analysis/src/05_predictions/03_preds_se.R

@@ -3,7 +3,7 @@
 #' impact of expanding access
 # ------------------------------------------------------------------------------
 
-# sub_cmd=-t 12 -n 31 -jn preds_se -wt 5m
+# sub_cmd=-t 12 -n 41 -jn preds_se -wt 5m
 
 # Set it up
 source(here::here("R", "utils.R"))

### 2020-10-03 analysis/paper/man.Rmd
adding renv & other project files
+ some updates to paper pre-sub (removed toc from supplement)

@@
 All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (Supplementary appendix, section S7).
-All processed data, code, and outputs are archived (here)[zenodo] and maintained at github.com/mrajeev08/MadaAccess.
+All processed data, code, and outputs are archived on Zenodo (http://doi.org/10.5281/zenodo.4064312 and https://doi.org/10.5281/zenodo.4064304),  and maintained at https://github.com/mrajeev08/MadaAccess.
 The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC.

### 2020-10-03 analysis/paper/supplement.Rmd
adding renv & other project files
+ some updates to paper pre-sub (removed toc from supplement)

@@
   bookdown::word_document2:
-    toc: true
     pandoc_args:

### 2020-10-03 analysis/src/00_install.R

@@ -3,9 +3,9 @@
 # ------------------------------------------------------------------------------
 
 # Get list of needed packages
-pkgs_used <- read.csv("analysis/out/logs/pkgs_used.csv", stringsAsFactors = FALSE)
-installed <- sapply(pkgs_used$pkgs, require, character.only = TRUE, quietly = TRUE)
-to_install <- pkgs_used$pkgs[!installed]
+pkgs_used <- read.csv("logs/pkgs_used.csv", stringsAsFactors = FALSE)
+installed <- sapply(pkgs_used$packages, require, character.only = TRUE, quietly = TRUE)
+to_install <- pkgs_used$packages[!installed]
 
 # Install needed packages
 install.packages(to_install, repos = "http://cloud.r-project.org")

### 2020-10-03 analysis/src/make_paper.R

@@ -12,7 +12,7 @@ split_render(manuscript = "analysis/paper/manuscript.Rmd",
 # Internal steps
 # Convert to pdf (using system & this:
 # stack exchange answer)
-system("rm analysis/paper/subfiles/*.pdf") # clean first
+system("rm -r analysis/paper/subfiles/*.pdf") # clean first
 system("bash analysis/paper/subfiles/word2pdf.sh")
 
 # Push up to google drive (single document)

### 2020-10-03 analysis/src/process_session.R

@@ -41,7 +41,7 @@ log_scripts %>%
   dplyr::distinct() -> log_last_ran
 
 # write log out of when scripts last ran
-write_create(log_scripts,
+write_create(log_last_ran,
              here_safe("logs/log_last_ran.csv"),
              write.csv,
              row.names = FALSE)
@@ -64,9 +64,9 @@ pkgs_used %>%
 
 # Write table & bib out for citing in supplement
 write_create(pkgs_used,
-             here_safe("analysis/out/logs/pkgs_used.csv"),
+             here_safe("logs/pkgs_used.csv"),
              write.csv,
              row.names = FALSE)
-write_create(pkgs_used$pkgs,
-             here_safe("docs/bib/refs_pkgs.bib"),
+write_create(pkgs_used$packages,
+             here_safe("analysis/paper/bib/refs_pkgs.bib"),
              write_bib)

### 2020-10-03 bin/zenodo.sh

@@ -3,18 +3,21 @@
 
 
 # And make sure all cluster results are pulled down (may take a bit!) (only will pull down if newer than local ones
-rsync -rLvzt --update --exclude "*.DS_Store*"  mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/MadaAccess/analysis/out/* ~/Documents/Projects/MadaAccess/analysis/out
+# rsync -rLvzt --update --exclude "*.DS_Store*"  mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/MadaAccess/analysis/out/* ~/Documents/Projects/MadaAccess/analysis/out
 
 if [ ! -d "zenodo" ];
 then
 mkdir -p zenodo/analysis
 mkdir -p zenodo/data-raw
-cp -r analysis/out  zenodo/analysis/
-else
-rsync -rLvzt --update --exclude "*.DS_Store*" analysis/out  zenodo/analysis/
 fi
 
-# Do data raw (don't copy as you have to exclude various bits
-rsync -rLvzt --update --exclude "*.DS_Store*" --exclude data-raw/raw/ipm_data --exclude data-raw/out --exclude data-raw/src --exclude "data-raw/log*" data-raw zenodo/data-raw/
+# clean it out
+rm -r zenodo/analysis/out/*
+rm -r zenodo/data-raw
 
+# Move over newest files
+cp -r analysis/out  zenodo/analysis/
+
+# Do data raw (don't copy as you have to exclude various bits
+rsync -rLvzt --update --exclude "*.DS_Store*" --exclude data-raw/raw/ipm_data --exclude data-raw/out --exclude data-raw/src --exclude "data-raw/log*" data-raw zenodo/
 

### 2020-10-03 bin/subrsync.sh
updating readme

@@ -10,7 +10,7 @@
 ## -q for quietly
 
 # Push up the scripts to home
-rsync -rLvztq --update --exclude '*.git' --exclude '.gitignore' --exclude "data-raw/raw*" --exclude "*.DS_Store*" --exclude ".Rproj*" --exclude analysis/out --exclude analysis/paper --exclude figs --exclude docs --exclude archive ~/Documents/Projects/MadaAccess mrajeev@della.princeton.edu:~/
+rsync -rLvztq --update --exclude '*.git' --exclude '.gitignore' --exclude "data-raw/raw*" --exclude "*.DS_Store*" --exclude ".Rproj*" --exclude analysis/out --exclude analysis/paper --exclude "renv*" --exclude ".Rprofile" figs --exclude docs --exclude archive --exclude ~/Documents/Projects/MadaAccess mrajeev@della.princeton.edu:~/
 
 # Push up the outputs to scratch (so always working from there for outputs)
 rsync -rLvztq --update --exclude "*.DS_Store*" ~/Documents/Projects/MadaAccess/analysis/out mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/MadaAccess/analysis/

### 2020-10-03 bin/zenodo.sh

@@ -3,7 +3,7 @@
 
 
 # And make sure all cluster results are pulled down (may take a bit!) (only will pull down if newer than local ones
-# rsync -rLvzt --update --exclude "*.DS_Store*"  mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/MadaAccess/analysis/out/* ~/Documents/Projects/MadaAccess/analysis/out
+# rsync -rLvzt --update --exclude "*.DS_Store*" --exclude "ttimes/candidates/*" mrajeev@della.princeton.edu:/scratch/gpfs/mrajeev/MadaAccess/analysis/out/* ~/Documents/Projects/MadaAccess/analysis/out
 
 if [ ! -d "zenodo" ];
 then

### 2020-10-05 analysis/paper/man.Rmd
fixing dependencies & adding striking image

@@
 if(no_figs) include_graphics <- function(x, ...) { return(invisible(NULL)) }
-
-# for logging dependencies
-out_session("logs/log_local.csv", start = start, 
-            ncores = 1)
 ```
@@
 In addition, we thank Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, Amy Winter, Christian Ranaivoson, John Friar, and Amy Wesolowski for assistance.
-This work was funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to CJEM and MR.
-MR is supported by an NSF Graduate Research Fellowship.
-KH is supported by the Wellcome Trust (207569/Z/17/Z).
-
 \newpage

### 2020-10-05 analysis/src/00_install.R

@@ -4,8 +4,20 @@
 
 # Get list of needed packages
 pkgs_used <- read.csv("logs/pkgs_used.csv", stringsAsFactors = FALSE)
-installed <- sapply(pkgs_used$packages, require, character.only = TRUE, quietly = TRUE)
+installed <- sapply(pkgs_used$packages, require, character.only = TRUE,
+                    quietly = TRUE)
 to_install <- pkgs_used$packages[!installed]
 
 # Install needed packages
 install.packages(to_install, repos = "http://cloud.r-project.org")
+
+# Comparing installed versions
+sapply(pkgs_used$packages, require, character.only = TRUE, quietly = TRUE) # run this again to load pkgs
+version_comp <- unlist(lapply(pkgs_used$packages, function(x) as.character(packageVersion(x))))
+version_installed <- gsub("-", ".", as.character(pkgs_used$version))
+
+# If any installed are older, update these
+to_update <- pkgs_used$pkgs[version_comp < version_installed]
+
+# Run the next line if you want to update installed packages
+# update.packages(to_update, repos = "http://cloud.r-project.org")

### 2020-10-05 analysis/src/04_addclinics/01_ttimes_candidates.R

@@ -4,7 +4,6 @@
 # ------------------------------------------------------------------------------
 
 # sub_cmd:=-t 12 -n 31 -jn candidates -wt 2m  -md "gdal"
-# skipit
 
 # Set it up
 source(here::here("R", "utils.R"))

### 2020-10-05 analysis/src/04_addclinics/02_test_metrics.R

@@ -3,7 +3,6 @@
 # ------------------------------------------------------------------------------
 
 # sub_cmd=-sn -t 12 -n 10 -jn testmets -wt 5m
-# skipit
 
 # Set it up
 source(here::here("R", "utils.R"))

### 2020-10-05 analysis/src/04_addclinics/03_ttimes_added.R

@@ -6,7 +6,6 @@
 # ------------------------------------------------------------------------------
 
 # sub_cmd=-sn -t 12 -n 12 -jn addclinics -wt 5m
-# skipit
 
 # Set it up
 source(here::here("R", "utils.R"))

### 2020-10-05 analysis/src/04_addclinics/04_postprocess.R

@@ -3,7 +3,6 @@
 # ------------------------------------------------------------------------------
 
 # sub_cmd=-sn -t 12 -n 5 -jn process -wt 2m
-# skipit
 
 # Set it up
 source(here::here("R", "utils.R"))

### 2020-10-05 analysis/src/figures/M3_bitemaps.R

@@ -214,6 +214,54 @@ mada_map_A <- ggplot() +
   ) +
   theme_map(font_size = 10)
 
+
+striking_img <- ggplot() +
+  geom_sf(
+    data = mada_districts,
+    aes(fill = ctar),
+    color = "white", alpha = 0.5
+  ) +
+  geom_bezier2(
+    data = filter(ctar_todist_bez, count > 4, distcode != dist_ctar),
+    aes(
+      x = long, y = lat, group = fct_reorder(group, total_lines, .desc = TRUE),
+      color = ctar, size = ifelse(index == 1, 0.05, size_lines(count))
+    ), n = 1000,
+    alpha = 0.8
+  ) +
+  geom_point(
+    data = ctar_pts, aes(
+      x = long_cent, y = lat_cent,
+      size = size_pts(count), fill = ctar
+    ),
+    shape = 21, color = "black", stroke = 1.2
+  ) +
+  geom_point(
+    data = dist_pts, aes(
+      x = long, y = lat, fill = ctar,
+      size = size_pts(count)
+    ),
+    shape = 21, color = "white", alpha = 1, stroke = 0.5
+  ) +
+  geom_point(
+    data = filter(ctar_all, exclude == 1), aes(x = long_cent, y = lat_cent),
+    shape = 1, stroke = 1, color = "black"
+  ) +
+  scale_discrete_manual(aes = "stroke", values = c(1.2, 0.5), guide = "none") +
+  scale_color_manual(values = catch_fills, guide = "none") +
+  scale_fill_manual(values = catch_cols, guide = "none") +
+  scale_size_identity("Reported bites", guide = "none") +
+  theme_map(font_size = 10)
+
+# for plos
+write_create(
+  striking_img,
+  here_safe("analysis/figs/striking_img.tiff"),
+  ggsave_it,
+  dpi = 400, height = 8.75, width = 7.5,
+  compression = "lzw", type = "cairo"
+)
+
 # Mapping raw data: Moramanga at commune level ---------------------------------
 moramanga %>%
   mutate(date_reported = ymd(date_reported)) %>%

### 2020-10-05 analysis/src/figures/SM1_ttimes.R

@@ -1,6 +1,6 @@
-# ------------------------------------------------------------------------------------------------ #
+# ------------------------------------------------------------------------------
 #' SM 1. GIS inputs and comparison to data
-# ------------------------------------------------------------------------------------------------ #
+# ------------------------------------------------------------------------------
 
 source(here::here("R", "utils.R"))
 start <- Sys.time()
@@ -29,7 +29,7 @@ mada_districts <- st_read(here_safe("analysis/out/shapefiles/mada_districts_simp
 friction <- raster(here_safe("data-raw/out/rasters/friction_mada_masked.tif"))
 pop1x1 <- raster(here_safe("data-raw/out/rasters/wp_2015_1x1.tif"))
 
-# Plotting GIS inputs (friction + pop) --------------------------------------------------------
+# Plotting GIS inputs (friction + pop) -----------------------------------------
 friction_df <- as.data.frame(friction, xy = TRUE)
 pop_df <- as.data.frame(pop1x1, xy = TRUE)
 
@@ -80,7 +80,7 @@ write_create(
   width = 12
 )
 
-# Raw data: groundtruthing ----------------------------------------------------------------------
+# Raw data: groundtruthing -----------------------------------------------------
 gtruth <- read.csv(here_safe("analysis/out/ttimes/gtruth_ttimes.csv"))
 baseline <- fread(here_safe("analysis/out/ttimes/base/grid_df.gz"))
 
@@ -167,7 +167,7 @@ write_create(
   height = 7
 )
 
-# Mode of travel ------------------------------------------------------------------------------
+# Mode of travel ---------------------------------------------------------------
 gtruth %>%
   filter(type != "point", !is.na(mode)) %>%
   group_by(mode) %>%
@@ -198,7 +198,7 @@ write_create(
   width = 6
 )
 
-# Comparing different metrics of access -------------------------------------------------------
+# Comparing different metrics of access ----------------------------------------
 # Plots comparing ttimes_wtd / unwtd / distance @ commune/district/grid scales
 gtruth %>%
   filter(type == "commune_wtd", !is.na(ttimes_reported)) %>%
@@ -286,7 +286,7 @@ metric <- c("Weighted travel times (hrs)", "Unweighted travel times (hrs)", "Dis
 comp <- data.frame(metric, points, communes)
 write.csv(comp, "analysis/out/stats/access_met_r2.csv")
 
[…]

### 2020-10-05 analysis/src/make_paper.R

@@ -2,7 +2,9 @@
 #' Compile paper for submission
 # ------------------------------------------------------------------------------
 
+start <- Sys.time()
 library(here)
+library(knitr)
 source("R/split_render.R")
 
 # Get the submission files
@@ -10,8 +12,8 @@ split_render(manuscript = "analysis/paper/manuscript.Rmd",
              output_dir ="analysis/paper/subfiles")
 
 # Internal steps
-# Convert to pdf (using system & this:
-# stack exchange answer)
+# Convert to pdf (using system & this applescript: https://superuser.com/a/1559043
+# superuser answer)
 system("rm -r analysis/paper/subfiles/*.pdf") # clean first
 system("bash analysis/paper/subfiles/word2pdf.sh")
 
@@ -19,3 +21,7 @@ system("bash analysis/paper/subfiles/word2pdf.sh")
 rmarkdown::render("analysis/paper/manuscript.Rmd",
                   output_format = "bookdown::word_document2")
 googledrive::drive_update('MadaAccess', 'analysis/paper/manuscript.docx')
+
+# cover letter
+rmarkdown::render(here("analysis/paper/cover_letter.Rmd"),
+                  output_dir = here("analysis/paper/subfiles"))

### 2021-01-13 analysis/paper/minor_revision_1/response_reviewers.Rmd
minor revisions for plos ntds

---
title: "Response to reviewers"
author: "Malavika Rajeev, Princeton University"
date: "1/12/2020"
output:
  word_document: default
  pdf_document:
    keep_tex: no
---

We thank the editor and the reviewers for their constructive feedback and close reading of the manuscript and have revised our manuscript accordingly. Both a document with track changes (of both the manuscript & the supplement) and clean separate documents have been submitted, and we have also verified our figures meet the requirements per the PACE tool. The reviewer comments are below with our response in italics.

### Reviewer 1

-   The authors performed the well-designed epidemiological analysis; given the distance from the residence to the PEP accessibility and assess the rabies risk. There is only once concern arising how much the prevalence (or exposure to residence) of rabid dogs in each area was included into this study. The authors also pointed out the uncertainty of underlying rabies expose according to the sensitivity analysis. This should be more clearly describe in the manuscript, with a validity of the corresponding parameters used in the current study.

    *We thank the reviewer for their helpful feedback. We have revised the methods and results section based on this and another reviewer's feedback. The methods section describing the decision tree now forefronts the parameter estimates with their sources, The results section now maps onto the major findings of each of our analyses, and hopefully clarifies the results of each section. We have also added a section to the discussion to regarding the validity of our assumptions (see lines xx - yy).* <!-- Done -->

-   Regarding Lines 275-281, the authors indicate the bases with adequate references how to reach the idea that human-to-dog rabies ratio are both positively and negatively correlated with human populations.

    *Thank you.* <!-- Done -->

-   Overall, the statements in Conclusion part is adequate and convince readers.

    *Thank you.* <!-- Done -->

-   In Lines 523-525, the authors discuss the potential risk factor of human rabies in Madagascar. These are very general and commonly argued in anywhere. Don't the authors can find the supportive reasons or data for these assumptions? Specifically, box1 may provide the example of association between area and awareness to the risk. We would like to know how these gaps may arise.

    *We have reorganized this discussion point, and and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines xx - yy). We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines xx).* <!-- Done -->

### Reviewer 2

-   The authors describe an approach to model the efficacy of different PEP distribution scenarios scenarios interesting and new. The approach is based on several steps and each steps is complex and relies on a set of assumptions. Because it would be difficult to collect empirical data (e.g. in a cluster randomized trial), I think the study is worth publishing. I appreciate that a lot of details and supplementary information is provided and that the source code and data is public. My main critic points are: i) due to the complexity and amount and diversity of information (the appendix alone has 47 pages) the authors should guide the reader better through the manuscript. Some suggestions are provided below, ii) the results are based on many assumptions -- I am not sure if the authors are not a bit too optimistic with their approach. I think the next step should be to try to validate the model before "Our framework could be used to guide PEP expansion and improve targeting of interventions."

    *We thank the reviewer for their helpful feedback and have revised the manuscript accordingly (see specific points regarding revisions to the methods and results in response to points below). Additionally, we have revised the text in the abstract (lines xx - yy) and in the conclusion (lines xx - xx) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP. We also added a point that in other settings, addressing other barriers to PEP access may be more effective (i.e. direct costs of PEP in settings that charge for PEP) (see lines xx - yy in the 'Broader context' section of the discussion).* <!-- Done -->

-   Ideas to guide the reader a bit better through the manuscript: - One weakness of the paper is that the reader has to jump back and forth to get all the information. Different parameters are introduced over 5 pages. Maybe the method section could be rearranged and a bit sharpened. E.g. p[rabid] is introduced in line 201. In line 208 (a figure legend) it is explained that R[rabid] is constrained. In line 225 it is explained which range was used to express uncertainty about P[rabies], in line 229 the constraining formula is provided and finally in Table 1 the values distribution and reference for this parameter is provided. I don't know if the constraint is so important that it has to be mentioned to this extend. ;Maybe a footnote in table1 would be sufficient. If not, there is certainly a better way to arrange the method section that it can be introduced at only one (maybe 2) occasion.

    *We have revised the methods section accordingly (lines xx - yy). Now all parameters are introduced in Table 1 with their sources and details, and Figure 1 describes how they are used in the decision tree framework. We have moved Table 1 up to the top of the section and describe the constraint on p[rabid] in a footnote and removed this from the figure description. The minor details about both p[rabid] and rho[max] are now described in the `Description` column of the table.* <!-- Done -->

-   Same with bite incidence. The result section shows some graphs and a lot of information related to model fit. But the reader is likely more interested in the estimated mean incidence. But this value is reported is not in this but in the next paragraph.

    *For the results section, we have revised each section to better state the main finding up front and focus on details in subsequent paragraphs. We combined the sections describing the relationship between bite incidence and travel times and moved some of the details regarding bite incidence estimation to the Supplementary Appendix section S2 and some of the details on model fit and comparisons to section S3.* <!-- Done -->

-   Terminology could be simplified. E.g. using consistently either commune or CSB2 in results and discussion to describe this level.

    *We have added a short glossary to guide the reader through the manuscript (lines xx - yy) and also changed all mentions of CSB-II to primary clinic and CSB-I to secondary clinic, and use the term clinics throughout to refer to health facilities.* <!-- Done -->

-   The appendix is quite long. A table of content would be helpful

    *Done.* <!-- Done -->

-   Isn't it possible to integrate the information of tables S6.1/2 into figures S6.1/2.

    *We added a panel into Figure S6.1 combining the information from Tables S6.1/2 and removed the tables.* <!-- Done -->

-   The most important result is "Estimating the impact of PEP provisioning." I think it would be good to have these results summarized in a table. This table could include/replace table 2.

    *We include an estimate of the deaths averted under current PEP provisioning in Table 2 and clarified the column names to make this more clear.* <!-- Done -->

-   Out-of-stock is briefly mentioned in this section but I think it is worth to discuss also that the supply chain is getting much more complicated if PEP hast o be delivered to much more clinics.

    *We expand on this later in the discussion, and we have rephrased the sentence in the discussion to point directly to complications in the supply chain. We also point to potential solutions for managing these supply chains (i.e. piggybacking on EPI programs, see lines xx - yy).* <!-- Done -->

-   Abstract - After reading the abstract I had no clear impression what the authors exatly did, e.g. ehat kind of data they used to estimate travel time, bite incidence and rabies deaths.

    *We have revised the abstract to more clearly describe the data and methods we used (lines xx - yy).* <!-- Done -->

-   "expanding PEP to one clinic per district could reduce deaths by 19%" Please add the number of districts (114) or the number of additional clinics (83).

    *Done.* <!-- Done -->

-   Authors summary " but our results suggest that expansion alone will not eliminate deaths." Well, how could PEP eliminate human rabies deaths? It does not interrupt transmission. I think this statement should be rephrased.

    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved..* <!-- Done -->

-   Travel time: The authors used geographical information to estimate theoretical travel time as an proxi for access. In addition they tried to validate the approach by comparing the estimates with travel time reported by patients and by travel time reported by the Institute Pasteur. However, very basic information is lacking, e.g. I could even not find the number of patients interviewed and it remains unclear what " driving times collected by IPM during field missions " actually means. Of course, reported and estimated travel times are correlated -- both are associated with distance -- but if I look at figure 2c or the appendix the variation is remarkable. Some estimate of fit would be nice. Maybe mean/median residual or something similar. By the way: I don't think the shaded areas (CIs around the regression lines) does not add a lot tot he interpretation. I would remove them because the points below are difficult to see.

    *We have clarified both the methods (lines xx-yy) and the results (lines xx - yy) based on the reviewer's helpful feedback. We point to some reasons for the observed variance in the results (lines xx - yy) and in the discussion (lines xx - yy), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data. We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines xx - yy). See Table S1.1 for R^2^ estimates for the linear models we used to compare the metrics, which we now also report in the main results text (line xx). We removed the CIs around the regression lines for both Fig 2C and Fig S1.4.* <!-- Done -->

-   Figure2.2: A indicates " (A) The daily time series of the number of forms submitted by each clinic," I don't think that the figure presents time series data. B boxplots for 4 points are (literally?) a bit pointless. Present only the 4 points or the 4 points + mean or median.

    *Thanks for **pointing** this out, fig S2.2 and it's caption have been revised accordingly, with S2.2A showing the number of forms submitted for each clinic with excluded periods shown in grey and S2.2B showing the four point estimates of reporting and the line showing the range.* <!-- Done -->

    #### Minor points :

-   Is the overdisperion parameter presented in Fig 4 the estimate or log(estimate)? If it is the estimate, an estimate below 1 is rather unusual.

    *The overdispersion estimate is not on a log scale, but we incorrectly describe it's form in the methods and have corrected it. The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines xx - yy and Gelman reference) and it is accordingly scaled by the offset (population size). This means that the estimated overdispersion is actually quite large, as described in the text and vizualised in the figures. For example, given a travel time of zero with a population of 100,000 and taking approximately the mean of the posteriors, we would predict exp(-5.5 +/- 0.9)100000 = 170 - 1000 bites per 100,000 persons. We apologize for the error in the text and thank the reviewer for catching it!* <!-- Done -->

-   Lines 131 133: please use superscript instead of caret to indicate exponentiation

    *Corrected.* <!-- Done -->

-   Lines 493: 45%%

    *Corrected.* <!-- Done -->

-   Lines 494-494: Please keep the same order.

    *Corrected.* <!-- Done -->

### Reviewer 3

-   The analysis presented match the analysis plan and the results are clearly and completely presented. The figures have sufficient quality.

    *Thank you.* <!-- Done -->

-   All the criterias are met. Minor points are the following: - Please indicate in S1.1A that the black crosses represent ARMCs.

    *Done.* <!-- Done -->

-   Delete in line 132 'the'

    *Corrected* <!-- Done -->

-   Fig. 1 with its description is confusing. I would rather recommend a list with the variables and their description.

    *We have revised this section based on this and another reviewers feedback (lines xx - yy). Briefly, we moved the table with the parameters and description to the top of the section, use the figure to show how these parameters were used, and additional choices regarding parameter ranges in the subsequent paragraph.* <!-- Done -->

-   Line 234, can this really be assumed? A baseline study in West Africa demonstrated, that only half of the bite victims complete PEP. However, based on the information regarding the high patient compliance in Madagaskar (line 539), I think you can make this assumption here.

    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (lines xx - yy) to ground this assumption up front (line xx).* <!-- Done -->

-   Remove a % in line 493

    *Corrected.* <!-- Done -->

-   This work nicely demonstrates that even with provisioning PEP all over the country, there will still be rabies deaths in the country. So besides the provision of PEP for humans, the disease needs to be eliminated in the animal reservoir.

    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines xx - yy).* <!-- Done -->

-   It was a pleasure to read this fantastic work.

    *We thank the reviewer for their kind and helpful review!* <!-- Done -->

### 2021-01-13 analysis/paper/manuscript.Rmd
minor revisions for plos ntds

@@
       - --csl=format/plos-neglected-tropical-diseases.csl
-      - --filter=pandoc-crossref
       - --lua-filter=format/scholarly-metadata.lua
@@
       - --lua-filter=format/multiple-bibliographies.lua
-      - --reference-doc=format/reference_supplement.docx
+      - --reference-doc=format/reference_main.docx
   bookdown::html_document2:
@@
 ## Methodology & Principal Findings
-We use travel times to the closest clinic providing PEP (N=31) as a proxy for access.
+We use spatially resolved data on numbers of bite patients seeking PEP across Madagascar and estimates of travel times to the closest clinic providing PEP (N = 31) in a Bayesian regression framework to estimate how geographic access predicts reported bite incidence.
 We find that travel times strongly predict reported bite incidence across the country.
-Using resulting estimates in an adapted decision tree framework we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
+Using resulting estimates in an adapted decision tree framework, we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
 We estimate 960 human rabies deaths annually (95% Prediction Intervals (PI):`r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with PEP averting an additional 800 deaths (95% PI: `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) each year.
-Under these assumptions, we find that expanding PEP to one clinic per district could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major health centers provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
+Under these assumptions, we find that expanding PEP to one clinic per district (N = 83 additional clinics) could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major primary clinics provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
 Our quantitative estimates are most sensitive to assumptions of underlying rabies exposure incidence, but qualitative patterns of the impacts of travel times and expanded PEP access are robust.
@@
 PEP is effective at preventing rabies deaths, and in the absence of strong surveillance, targeting underserved populations may be the most equitable way to provision PEP.
-Our framework could be used to guide PEP expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted.
+Given the potential for countries to use Gavi funding to expand access to PEP in the coming years, this framework could be used as a first step to guide expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted and baseline data on rabies risk is lacking.
 While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyze progress towards achieving zero rabies deaths.
@@
 Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by 2030, and recently, Gavi, a multilateral organization that aims to improve access to vaccines in the poorest countries, added human rabies vaccine to it’s portfolio.
-In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP, and extrapolate human rabies deaths in Madagascar.
+In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP and extrapolate human rabies deaths in Madagascar.
 We find that PEP currently averts around `r round(natl_base$averted_mean[1], -2)` deaths each year, but that the burden remains high (`r round(natl_base$deaths_mean[1], -2)` deaths/ year), particularly in remote, hard-to-reach areas.
-We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone will not eliminate deaths.
+We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone is unlikely to achieve the global goal of zero human deaths from dog-mediated rabies by 2030.
 Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other Low- and Middle-Income countries, forward on the path to rabies elimination.
@@
 Often, the populations at greatest risk of a given disease are the most underserved [@peters2008poverty].
-Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to allocate limited resources [@mainaSpatialDatabaseHealth2019].
+Delivering interventions to these groups is challenging due to financial and infrastructural limitations and requires careful consideration of how best to allocate limited resources [@mainaSpatialDatabaseHealth2019].
 
@@
 Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans.
-While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid animal [@RabiesVaccinesWHO2018].
-However, access to human vaccine is limited in many countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
+While dog vaccination can interrupt transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid animal [@RabiesVaccinesWHO2018].
+However, access to the human rabies vaccine is limited in many countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].
 
@@
 
-In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to it's own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019].
+In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to its own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019].
 Other than at these 31 anti-rabies medical centers (ARMC), PEP is not available at any other public clinics or through the private sector.
@@
 Due to the spatially restricted nature of PEP provisioning and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
-Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
+Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, but did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
 Here, we provide revised estimates of human rabies deaths by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
@@
 
+### Box 1: Glossary
+- **Institut Pasteur de Madagascar (IPM):** private Malagasy research institution located in the capital of Antananarivo; part of the Institut Pasteur network and also houses the National Rabies Reference Laboratory (NRRL), as well as the vaccination clinic (one of the 31 clinics where PEP is provided).
+- **Antirabies medical center (ARMC):** public clinics (N = 31) in Madagascar where PEP is provided free of charge to patients.
+IPM provides PEP to these clinics, and the clinical staff implement vaccination (can be through emergency, immunizaton, or primary care departments).
+- **Primary clinic:** a primary health clinic formally known as 'Centre de Sante Niveau II' or CSB-II in French; these are major health centers that should theoretically have immunization capacity.
+- **Secondary clinic:** a secondary health clinic formally known as 'Centre de Sante Niveau I' or CSB-I in French; these are lower level health posts that are often staffed by a single clinician and often without formal immunization capacity.
+- **District:** second-level administrative divisions of Madagascar. There are 114 districts in Madagascar.
+- **Commune:** third-level administrative division of Madagascar below the district level. There are 1579 communes in Madagascar.
+
 # Methods
 
+```{r}
+
+ttime_data <- read.csv(here("data-raw/out/bitedata/moramanga_ttimes.csv"))
+patient_nobs <- sum(table(ttime_data$hours))
+ttime_driving <- read.csv(here("data-raw/misc/ttimes_IPM.csv"))
+
+```
+
 ## Estimating geographic access to ARMC {#methods-ttimes}
-To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two rasters: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B), which we aggregated to the friction surface.
+To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two raster datasets: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2^ scale (Fig S1.1A) and 2) the population estimates from the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2^ resolution, Fig S1.1B), which we aggregated to the friction surface.
 
 From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018].
-We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and euclidean distance, i.e.
+We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and Euclidean distance, i.e.
 the minimum distance from the administrative unit centroid to any ARMC.
 We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018).
-We compared travel times and distance estimates to driving times collected by IPM during field missions and patient reported travel times from a subset of Moramanga ARMC patients (see Fig S1.2 for raw data).
+To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e.
+time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best in terms of a linear model.
 
@@
 
-## Modeling reported bite incidence {#methods-bitemods}
+## Modeling reported bite incidence as a function of access {#methods-bitemods}
 
@@
 We tested whether the catchment random effect captured overdispersion in the data (i.e.
-variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter, $\sigma_e$ [@gelman2006data]:
-
-$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j \ + \ \sigma_e)}P_i$$
+variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter: $\epsilon_i \sim norm(0,\ \sigma_e)$, where $\sigma_e$ is the standard deviation around a random variable with mean of zero [@gelman2006data]:
+
+$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j \ + \ \epsilon_i)}P_i$$
 
@@
 ## Modeling human rabies deaths {#methods-burden}
-We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1).
-To model uncertainty in parameter estimates we used triangular distributions, as with previous studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual exposure incidence per administrative unit, and  $p_{rabid}$, the proportion of reported bites that are rabies exposures.
+We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework.
+Table 1 lists all parameter values and their sources.
+Fig 1 describes how these parameters are used in the decision tree and the key outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).
+
+```{r exps}
+exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
+rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
+```
+
+_**Table 1. Parameters used in the decision tree to estimate human rabies deaths at the administrative level.**_\
+
+| **Parameter**   | **Value**        | **Description**	                 | **Source**|
+|:----------------|:----------------:|:----------------------------------|:---------:|
+| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
+| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
+| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures^1^ | [@rajeevHealthcareUtilizationProvisioning2019] |
+| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga ARMC (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
+| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|
+
+^1^ $p_{rabid}$ is constrained so that rabid reported bites cannot exceed the total expected number of rabies exposures ($E_i$) or maximum reporting in a given simulation ($\rho_max$).
 
@@
 To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures.
-$p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting of patients to clinics cannot exceed a maximum ($p_{max}$).
 $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{death}$) to get deaths due to rabies ($D_i$).
@@
 As there is little data on dog population size and human exposure incidence in Madagascar[@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
-
-For $p_{rabid}$, we use a range between 0.2 - 0.6 estimated from a study of bite patients in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].
-So that rabid reported bites cannot exceed the total expected number of rabies exposures or a maximum reporting (even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$:
-
-$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
-
-where $\rho_{max}$ is the maximum reporting, estimated from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where we find that approximately 2% of rabies exposures go unreported [@rajeevHealthcareUtilizationProvisioning2019].
-
-We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing rabies.
-Fig 1 describes the decision tree, the key inputs, and outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).
-Table 1 list all inputs and their sources.
-
-```{r exps}
-exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
-rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
-```
-
-_**Table 1. Parameters used in the decision tree to estimate human rabies deaths at the administrative level.**_\
-
-| **Parameter**   | **Value**        | **Description**	                 | **Source**|
-|:----------------|:----------------:|:----------------------------------|:---------:|
-| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
-| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
-| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|
+Given previously high observed compliance in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], we assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing rabies.
 
@@
 We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic.
-We select the CSB II (Centre de Sante Niveau II, major health centers with immunization capacity) in the highest density grid cell of the administrative unit as candidates for expansion.
-For the `r comms_no_csb2` communes without a CSB II, we chose the CSB I (secondary health posts) in the highest density grid cell.
-`r comms_no_clins` communes lacked any CSB I or II.
-Finally, we explore a scenario where all additional CSB II (totalling `r nrow(csbs_max)`) provision PEP.
+We select the primary clinic (primary health facility, see Glossary) in the highest density grid cell of the administrative unit as candidates for expansion.
+For the `r comms_no_csb2` communes without a primary clinic, we chose the secondary clinic (secondary health facility, see Glossary) in the highest density grid cell.
+`r comms_no_clins` communes lacked any health facilities.
+Finally, we explore a scenario where all additional primary clinics (totalling `r nrow(csbs_max)`) provision PEP.
 
 We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from an existing ARMC for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing ARMC.
-We simulated expansion of ARMC to each district using these three metrics, and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
+We simulated expansion of ARMC to each district using these three metrics and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
 For the full simulation of expanded access, once clinics reduced travel times for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` living greater than $x$ hrs away, starting with $x$ = 3 hrs), we reduced the travel time threshold by 25%.
@@
 The IPM GIS unit provided the data on geolocated clinics across the country.
-Anonymized raw bite patient data and full data on geolocated clinics are available from IPM following insitutional data transfer protocols.
+Anonymized raw bite patient data and full data on geolocated clinics are available from IPM following institutional data transfer protocols.
 Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (wisemonkeyfoundation.org) on the same date and are shared in the archived repository.
@@
 
-## Estimating access to ARMC {#results-ttimes}
+## Travel times to ARMC are high and variable across Madagascar. {#results-ttimes}
 ```{r ttime stats}
 ttime_stats <- read.csv(here("analysis/out/stats/prop_pop_ttimes.csv"))
+ttime_fit <- read.csv(here("analysis/out/stats/access_met_r2.csv"), row.names = 1)
 
@@
   sum(prop_pop) -> prop_over3_base
-```
-
-Estimates of travel times to the closest ARMC varied greatly (Fig 2A), with approximately `r round(prop_over3_base*100, 0)`% of the population estimated to live over 3 hrs from a clinic (Fig 2B).
-We found that travel time estimates from the friction surface underestimated both driving times across the country and patient-reported travel times to the Moramanga ARMC (Fig 2C), but were correlated with ground truthed driving and patient-reported times (Fig 3C, Fig S1.4).
-Patient-reported travel times were also significantly more variable than estimates from the friction surface (Fig S1.2).
-The friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
+
+ttime_stats %>%
+  filter(cut_times %in% c("8 - 10", "10 - 15", "15 +")) %$%
+  sum(prop_pop) -> prop_over6_base
+```
+
+Based on the estimates from the friction surface, approximately `r round(prop_over3_base*100, 0)`% of the population of Madagascar are estimated to live over 3 hours from a clinic (Fig 2).
+However, we found that these estimates underestimated both driving times across the country and patient-reported travel times to the Moramanga ARMC (Fig 2C).
+Patient reported travel times were highly variable for a given commune compared to the estimated travel times (Fig S1.2), potentially due to the fact that the friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
 the presence of a road indicates that all travel through that grid cell is by vehicle).
 However, Moramanga ARMC patients reported using multiple modes of transport, with some individuals walking days to a clinic (Fig S1.3).
-Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (Table S1.1).
-Therefore we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
+
+While the travel time estimates may not reflect exact distributions of travel times, they were correlated with ground-truthed driving and patient-reported times and likely reflect patterns of access over the country (Fig 3C, Fig S1.4).
+Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (R^2^ = `r round(ttime_fit$communes[ttime_fit$metric == "Weighted travel times (hrs)"], 2)`, Table S1.1), therefore, we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
 
@@
 The vertical lines show the 95% quantiles for reported travel times and the point size shows the number of observations for each commune.
-The best fit lines (red and grey) and 95% confidence interval (shading) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
+The best fit lines (red and grey) from a linear model where observed travel times are predicted by estimated travel times for each data source are also shown.
 The dashed black line is the 1:1 line.*
 
-## Estimating bite incidence {#results-biteinc}
-
-Most patients from each district reported to their closest ARMC by the weighted travel time metric (Fig 3).
-Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
-While there are discrepancies between commune and district catchment assignments (Fig S2.1A), over 75% of the population in a given district or commune were closest to a single clinic (Fig S2.1C).
-We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms (see Supplementary Appendix Section S2).
+## As travel times increase, reported bite incidence decreases. {#results-biteinc}
+
+Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
 After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D).
+For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
+While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.
 
@@
 
-Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
-For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
-While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.
-
-## Modeling reported bite incidence {#results-bitemods}
-
-Travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales (Fig 4).
-Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
-
-For the national data, including a catchment random effect improved predictions (Fig S3.2 & Fig S3.3).
-However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table S3.1) and the models resulted in similar predictions (Fig S3.6 & S3.7), indicating that catchment effects could not be differentiated from random variation in the data.
-Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
-
+Our modeling results show that travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales with the best fitting models including travel times and an overdispersion parameter (Fig 4, see Supplementary Appendix section S3 for comparisons to models with catchment effects and with population size as a covariate).
 As the predictions from the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
 Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig S3.3), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
-Finally, we examined the sensitivity of models to how we corrected for underreporting of data, and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Fig S3.8 and Fig S3.9).
+Finally, we examined the sensitivity of models to how we corrected for underreporting of data and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Fig S3.8 and Fig S3.9).
 
@@
 
-## Estimating human rabies deaths {#results-burden}
+## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access to PEP. {#results-burden}
+In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
+Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year.
 Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar.
 Our estimates vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Fig 5A).
-Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year.
-In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
 
@@
 kable(table, digits = -1, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
-                                        "Deaths averted"))           
+                                        "Deaths averted by current PEP provisioning"))           
 ```
@@
 
-## Estimating the impact of PEP provisioning {#results-incremental}
-
-For a subset of additional clinics (n = 83, up to one per district), we compared three methods of ranking for expanding PEP provisioning.
+## Expanding PEP access to underserved populations is effective at reducing human rabies deaths, but this effect saturates as clinics are added. {#results-incremental}
+
 We found that targeted expansion of PEP to clinics based on the proportion of the population they reduced travel times for resulted in fewest deaths (Fig S5.1).
-Therefore, we used this approach for simulating expansion of PEP to a larger set of clinics (N = `r max_added` ).
 Here we report results from the commune model, as estimates were consistent across models (Fig 6 and Supplementary appendix, section S5).
-
 We estimated that strategic PEP expansion to these additional 83 clinics (1 per district) reduced rabies deaths by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 6A).
@@
 However, reductions in burden saturate as more clinics are added (Fig S5.2).
-Even when all CSB II provisioning PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5); as more clinics are added, reported bite incidence saturates (Fig S5.4), and patients shift which clinic they report to (S5.7 & S5.8).
+Even when all primary clinics provision PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5); as more clinics are added, reported bite incidence saturates (Fig S5.4), and patients shift which clinic they report to (S5.7 & S5.8).
 
@@
 Our model predicts an increase from `r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 0 & natl_preds$scale == 'Commune'], -2)})")` per annum under current provisioning but with the abridged intradermal regimen (i.e.
-visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")`  with 250 clinics providing PEP, and ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` if all CSB IIs provision PEP.
+visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")`  with 250 clinics providing PEP, and ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` if all primary clinics provision PEP.
 In these scenarios, clinic catchment populations and throughput decrease, with clinics seeing fewer patients per day (S5.6).
@@
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The points show when all additional csb (N = `r max_added`) clinics have been added).*
-
-## Sensitivity Analyses {#results-sensitivity}
-To quantify which parameters contribute to the uncertainty in our estimates, we performed univariate sensitivity analyses, varying parameters in both the bite incidence model (Table S6.1) and the decision tree (Table S6.2).
-These analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute most uncertainty to burden estimates (Fig S6.1) and impacts of PEP access (Fig S6.2).
-Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden.
+The points show when all additional primary clinics and secondary clinics (N = `r max_added`) clinics have been added).*
+
+## Burden estimates are most sensitive to assumptions of underlying rabies incidence. {#results-sensitivity}
+While qualitative patterns of the current impact of geographic access on human rabies deaths and the impact of expanding access to PEP on reducing these deaths is robust across a wide range of parameter estimates, our sensitivity analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the most uncertainty to our quantitative estimates (Fig S6.1 & 2).
+Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden or impacts of expanded access.
 For the estimates of vial demand, uncertainty around the model intercept (i.e.
@@
 However, the impact of adding clinics remains broadly the same (Fig S6.4B).
-Overall, while uncertainty in the underlying rabies exposure incidence results in considerable variation in our burden estimates, the projected impact of travel times and of access to PEP are qualitatively similar across parameter assumptions.
 
@@
 We find that the burden of rabies in Madagascar is likely concentrated in areas with poor PEP access.
-We estimate that current PEP provisioning (at 31 clinics) averts `r glue("{trans_perc(min(prop_averted))}%")`% of deaths that would otherwise occur, and that expanding PEP access should reduce mortality, with provisioning in one clinic per district (N = `r one_per_dist`), or per commune (N = `r one_per_comm`), expected to reduce mortality by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")` and `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")`, respectively.
-However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all CSB II (N = `r max(natl_preds$scenario)`).
-This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all major health centers.
+We estimate that current PEP provisioning (at 31 clinics) averts `r glue("{trans_perc(min(prop_averted))}%")`of deaths that would otherwise occur, and that expanding PEP access should reduce mortality, with provisioning in one clinic per district (N = `r one_per_dist`), or per commune (N = `r one_per_comm`), expected to reduce mortality by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))}%")` and `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == one_per_comm]))}%")`, respectively.
+However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all primary clinics (N = `r max(natl_preds$scenario)`).
+This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all major primary clinics.
 With reduced travel times, over 20% of exposures will still not seek PEP (Fig S5.5), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
@@
 
+Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths.
+While we rely on a number of assumptions, they are based on data specific to Madagascar or from other similar settings and consistent with estimates from the literature more broadly (see Table 1 and section S4).
+Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
+Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
+
 We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
+Most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
+[@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers to PEP access.
 In Madagascar, where PEP is free-of-charge, the main cost to patients is transport and time lost.
 More remote communities tend to be of lower socioeconomic and educational status [@peters2008poverty], so travel time may be a proxy for these correlated variables.
-We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
-Most ARMC charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers.
 Significant overdispersion in the data that cannot be explained by travel times suggests that clinic-level variation (e.g.
@@
 dog populations, outbreaks, awareness) further influence health-seeking behavior and vaccine demand.
-
-Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths.
-Our simplifying assumptions, regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP, are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
-Likewise we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care and human behavior, as illustrated by the complex case studies in Box 1.
-
-## Box 1: case studies of of health seeking behavior for PEP in Madagascar
+Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care, and human behavior, as illustrated by the case studies in Box 1.
+
+### Box 2: case studies of of health seeking behavior for PEP in Madagascar
 ```{r case studies}
@@
 2.
-Recently, a middle aged taxi driver died of rabies in suburban Antananarivo.
-The day after being bitten by an unknown dog, he reported to a clinic that referred him to the ARMC at IPM, approximately one hour’s drive from his home.
-His family urged him to get PEP, but he did not believe his risk was high and decided not to seek further care.
+Recently, a middle-aged taxi driver died of rabies in suburban Antananarivo.
+The day after being bitten by an unknown dog, he reported to a clinic that referred him to the ARMC at IPM for PEP, approximately one hour’s drive from his home.
+His family urged him to go, but he did not believe his risk was high and decided not to seek further care.
 He developed symptoms two weeks later and was confirmed as a rabies death by the National Rabies Reference Laboratory.
-Despite prompt reporting and referral, and socioeconomic indicators suggesting a high care-seeking probability, this person did not receive PEP.
-His story highlights the need for sensitization about rabies and how PEP at peripheral clinics (even in areas with reasonable access) could prevent additional deaths, but also how PEP alone is unlikely to prevent all rabies deaths.
+Despite prompt reporting, appropriate referral, and socioeconomic indicators suggesting a high care-seeking probability, this person did not receive PEP.
+His story highlights the need for sensitization about rabies, how PEP provisioning at peripheral clinics (even in areas with reasonable access) could prevent additional deaths, and ultimately that PEP alone is unlikely to prevent all rabies deaths.
 
@@
 [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015].
+In other settings, similar statistical approaches could be used to identify and quantifying key barriers to PEP seeking behavior.
+For example, reducing the direct cost of PEP is likely to be of more importance than increasing geographic access where PEP costs are high.
+
 Our revised estimate of rabies deaths in Madagascar using this approach was higher than previous [@rajeevHealthcareUtilizationProvisioning2019], which assumed uniform reporting of 85%, but remained within the range of other empirical and modeling studies from low-income countries [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @lyRabiesSituationCambodia2009; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011].
-
 Our estimates of vial demand depend on use of the new abridged intradermal regimen [@tarantolaIntradermalRabiesPostexposure2018], which has been adopted by the Ministry of Health in Madagascar.
-However, most ARMC staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moromanga [@rajeevHealthcareUtilizationProvisioning2019].
+However, most ARMC staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moramanga [@rajeevHealthcareUtilizationProvisioning2019].
+
 We predict that as clinics are expanded, throughput (daily patients reporting to a clinic) will decrease.
-This may make provisioning PEP more challenging and vial demand less predictable, leading to stock outs or wastage.
+This may complicate the supply chain and make provisioning PEP more challenging as vial demand becomes less predictable, leading to stock outs or wastage.
 Decentralized provisioning mechanisms, for example adopting routine childhood vaccine supply chains, or novel vaccine delivery methods such as drones [@boeras2019use], may mitigate these challenges.
@@
 IBCM is one way to manage PEP effectively [@rysavaPathRabiesElimination2018] and as it relies on exposed persons reporting to clinics, expanding PEP access could strengthen this surveillance framework.
-These same issues of access, however apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].
+These same issues of access, however, apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].
 
@@
 Implementing outreach programs to raise awareness should further increase the efficacy of PEP expansion by improving care seeking.
-Better surveillance is also needed to understand the geographical distribution of rabies exposures and identify populations most at risk.
-Gavi investment could support countries to more equitably provision PEP and overcome barriers to access ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies), but as PEP alone cannot prevent all rabies deaths, investment should be used to catalyse mass dog vaccination to interrupt transmission, and eventually eliminate rabies deaths.
+Better surveillance is also needed to understand the geographical distribution of rabies exposures and identify populations most at risk, and to evaluate the effectiveness of PEP expansion at preventing human rabies deaths.
+Gavi investment could support countries to more equitably provision PEP and overcome barriers to access ([@wentworthSocialJusticePerspective2019], see Box 1 for case studies), but as PEP alone cannot prevent all rabies deaths, investment should be used to catalyze mass dog vaccination to interrupt transmission, and eventually eliminate rabies deaths.
 
@@
 ### *Fig S1.1. Raster inputs to estimate travel times to the closest ARMC for Madagascar.*
-_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale. (B) Population estimates resampled to the same friction surface._
+_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current ARMC (N = 31) shown with black x's.
+(B) Population estimates resampled to the same friction surface._
 
@@
 (B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
-Paths are bezier curves from origin to destination, and do not show actual paths driven._
+Paths are Bezier curves from origin to destination, and do not show actual paths driven._
 
@@
 
-### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations).*
-_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only)._
+### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) vs. estimates from friction surface.*
+_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only) ._
 
@@
 ## S2. Estimating bite incidence
+
+Most patients from each district reported to their closest ARMC by the weighted travel time metric (Fig 3).
+Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
+While there are discrepancies between commune and district catchment assignments (Fig S2.1A), over 75% of the population in a given district or commune were closest to a single clinic (Fig S2.1C).
+We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms.
+
 Vial demand was simulated under simplified assumptions of PEP administration and adherence [@rajeevHealthcareUtilizationProvisioning2019], based on patients reported randomly across the year.
@@
 
-### *Fig S2.2. Estimating undersubmission of patient forms.*
-*(A) The daily time series of the number of forms submitted by each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in white).
-(B) Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year) calculated as the # of days in a year which were not excluded based on the criteria of 15 consecutive days of non-submission/365.
+### *Fig S2.2. Estimating under-submission of patient forms.*
+*(A) Number of patient forms submitted to IPM for each clinic over the study period for each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in grey); (B) Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year and the line is the range), calculated as the # of days in a year which were not excluded based on the criteria of 15 consecutive days of non-submission/365.
 (C) The difference between log(estimated) and log(observed) vials provisioned for the period of 2013 - 2017 for each clinic correcting for under-submission (squares) using the 15 day cut-off show in A and B, vs.
@@
                              cut_off == Inf ~ "No correction", 
-                             !(cut_off %in% c(0, Inf)) ~ as.character(cut_off))) %>%
+                             !(cut_off %in% c(0, Inf)) ~ as.character(cut_off)), 
+         mse = round(mse, 2)) %>%
   select(`Consecutive day threshold` = cut_off, `Root MSE` = mse) -> mse
@@
 assuming forms were completely reported resulting in lower estimates of bite incidence) and with a lower cut-off (7 days, resulting in higher estimates of under-submission of forms).
+
+For the national data, including a catchment random effect improved predictions (Fig S3.2 & Fig S3.3).
+However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table S3.1) and the models resulted in similar predictions (Fig S3.6 & S3.7), indicating that catchment effects could not be differentiated from random variation in the data.
+Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
+Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
 
@@
 ### *Fig S5.2. Map of the location and at what step each clinic was added.*
-*The circles are each of the CSB II across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
-The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional CSB II in the country which were added in the final step but not ranked.*
+*The circles are each of the primary clinics across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
+The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional primary clinics in the country which were added in the final step but not ranked.*
 \newpage
@@
 *(A) at the ~1 x 1 km grid cell (B) commune and (C) district scales.
-The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional CSB IIs in the country).
+The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional primary clinics in the country).
 Grey pixels show the location of ARMC at each step.
@@
 
-```{r se tables}
-se_pars <- read.csv(here("analysis/out/sensitivity/se_pars.csv"))
-
-se_pars %>%
-  pivot_longer(exp_min:beta_0) %>%
-  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
-                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
-  group_by(scale, par = name) %>%
-  summarize(upper = max(value), lower = min(value)) %>%
-  mutate(range = glue("{lower} \U2012 {upper}")) -> se_table_all
-
-se_table_all %>%
-  filter(grepl("beta|sigma", par)) %>%
-  pivot_wider(par, values_from = range, names_from = scale) %>%
-  as.data.table(.) -> bites_se
-se_table_all %>%
-  filter(!grepl("beta|sigma|min", par), scale != "Commune") %>%
-  as.data.table(.) -> dtree_se
-```
-
-### *Table S6.1. Range of parameters (upper 97.5% and lower 2.5% credible interval of posterior) used in models of bite incidence for the univariate sensitivity analyses.*
-
-| **Parameter**   | **Description**        | **Commune model** | **District model** | 
-|:----------------|:----------------------:|:----------------------------------:|:----------------------------------:|
-| $\beta_t$ | Travel time effect	| `r bites_se[par == "beta_ttimes"]$Commune` | `r bites_se[par == "beta_ttimes"]$District`
-| $\beta_0$ | Model intercept	| `r bites_se[par == "beta_0"]$Commune` | `r bites_se[par == "beta_0"]$District`
-| $\sigma_0$ | Overdispersion	| `r bites_se[par == "sigma_e"]$Commune` | `r bites_se[par == "sigma_e"]$District`
-
-\newpage
-
-### *Table S6.2. Range of parameters used in decision tree model for the univariate sensitivity analyses.*
-
-| **Parameter**   | **Description**        | **Range**	                 | **Notes**|
-|:----------------|:----------------------:|:----------------------------|:---------:|
-| $E_i$ |	Annual exposures per 100,000 persons| `r dtree_se[par == "exp_max"]$range`  	
-| $p_{rabid}$	| Proportion of reported bites that are rabies exposures | `r dtree_se[par == "p_rab_max"]$range` 	
-| $\rho_{max}$ | The maximum reporting possible for any location | `r dtree_se[par == "rho_max"]$range`	 
-| $p_{death}$ | The probability of death given a rabies exposure^1^ | `r dtree_se[par == "p_death"]$range`	
-^1^*95% CI of estimate from [@changaluchaNeedImproveAccess2018]*
-
-\newpage
-
 ```{r fig S6.1 base se}
@@
 ### *Fig S6.1. Sensitivity analyses for baseline burden estimates.*
-*(A) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
+*(A) Range of parameters used in the univariate sensitivity analysis (upper 97.5% and lower 2.5% credible interval of posterior for the parameters in the bite incidence model, 95% of CI from 95% CI of estimate from [@changaluchaNeedImproveAccess2018] for $\rho_max$, and fixed at the minimum and maximum of the range used in the main analysis for all other parameters) (B) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (C) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
 See Table S6.1 & 2 for ranges used for each parameter.*
@@
 *(A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
+See Fig S6.1A for ranges used for each parameter.*
 
@@
 The points show the mean estimates from 1000 simulations and the line ranges show the 95% prediction intervals.*
-
 \newpage
@@
 
+\newpage
+
 ## S8. Additional supplementary references {#srefextra}

### 2021-01-13 analysis/paper/subfiles/cover_letter.Rmd
minor revisions for plos ntds

---
title: ""
author: ""
date: "10/2/2020"
output: 
  pdf_document:
    keep_tex: false
---

Dear Editor,

We are submitting our manuscript entitled “How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar” for consideration for publication in PLoS Neglected Tropical Disease. Our article presents a modeling framework to estimate how geographic access to post-exposure prophylaxis (PEP) shapes the burden of human rabies deaths.

We use data from clinics across Madagascar and at multiple scales and travel time estimates to clinics provisioning PEP to fit models of reported bite incidence. We then use the predictions these models and integrate them into a decision tree framework to extrapolate rabies burden and vial demand. Finally, we examine how expanding access strategically can reduce human rabies burden. This work is inline with a body of work that shows the impacts of geography on access to care and disease outcomes. It also provides a modeling framework that could be usefully applied to other countries where PEP access is currently geographically restricted in how to effectively increase expansion, particularly in light of the recent addition of human rabies vaccine to the Gavi investment portfolio.

We thank you for your consideration.

Sincerely,  
Malavika Rajeev, Princeton University

### 2021-01-13 analysis/src/figures/SM1_ttimes.R

@@ -236,7 +236,7 @@ S1.4_lms <- ggplot(
   aes(x = value, y = mean, color = type, shape = type), alpha = 0.75
 ) +
   geom_point() +
-  geom_smooth(method = "lm", aes(color = type, fill = type)) +
+  geom_smooth(method = "lm", aes(color = type), fill = NA) +
   scale_shape_discrete(
     name = "Type of estimate",
     labels = c(

### 2021-01-13 analysis/src/figures/SM2_bitedata.R

@@ -130,21 +130,30 @@ throughput_cols <- c("0" = "white", "< 10" = '#edf8fb', "< 20" = '#b3cde3',
                      "< 40" = '#8c96c6', "< 80" = '#8856a7', "< 100" = '#810f7c')
 throughput_brks <- c(-0.1, 0.1, 10, 20, 40, 80, 100)
 throughput$date_reported <- ymd(throughput$date_reported)
+throughput$include_15[throughput$include_15 == 0] <- NA
+throughput$include_15[is.na(throughput$no_patients) & throughput$include_15 == 1] <- 0
 
 throughput_A <- ggplot(data = throughput, aes(x = date_reported, y = reorder(ctar, include_15))) +
-  geom_tile(aes(fill = ifelse(include_15 == 0, NA, as.character(ctar)))) +
-  scale_fill_manual(values = catch_cols, guide = "none") +
+  geom_tile(aes(fill = no_patients)) +
   xlim(ymd("2014-01-01"), ymd("2017-12-31")) +
+  scale_fill_distiller(trans = "sqrt", breaks = c(0, 10, 50), direction = 1,
+                       name = "Number of patients") +
   theme_minimal_hgrid() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
   xlab("Year") +
   ylab("ARMC") +
   labs(tag = "A")
 
-rep_B <- ggplot(data = reporting, aes(x = reorder(ctar, include_15), y = include_15,
+reporting %>%
+  group_by(ctar) %>%
+  summarize(min = min(include_15), max = max(include_15),
+            include_15 = mean(include_15),) -> rep_summ
+
+rep_B <- ggplot(data = rep_summ, aes(x = reorder(ctar, include_15), y = include_15,
                                               color = ctar)) +
-  geom_boxplot() +
-  geom_point(alpha = 0.5) +
+  geom_linerange(aes(ymin = min, ymax = max, color = ctar)) +
+  geom_point(data = reporting, aes(x = ctar, y = include_15, color = ctar),
+             alpha = 0.5, size = 3) +
   labs(y = "Estimated submission \n of forms", x = "", tag = "B") +
   geom_hline(yintercept = 0.25, linetype = 2, color = "grey") +
   coord_flip() +

### 2021-01-13 analysis/src/figures/SM6_burden_se.R

@@ -18,6 +18,7 @@ library(glue)
 # Baseline burden (univariate) ----------------------------------------------------------------
 base_se_admin <- fread(here_safe("analysis/out/sensitivity/burden_baseline_se.gz"))
 baseline <- fread(here_safe("analysis/out/preds/admin_preds.gz"))[scenario == 0]
+se_pars <-  fread(here_safe("analysis/out/sensitivity/se_pars.csv"))
 
 # At national level ----------------------------------------------------------------
 # Labels and colors
@@ -53,20 +54,65 @@ base_se$vary <- factor(base_se$vary, levels = rev(c(
   "p_rabid", "p_death", "human_exp"
 )))
 
-base_natl_A <- ggplot(
+# range of pars used
+se_pars %>%
+  pivot_longer(exp_min:beta_0) %>%
+  mutate(value = case_when(name %in% c("exp_min", "exp_max") ~ round(value*1e5, 2),
+                           !(name %in% c("exp_min", "exp_max")) ~ round(value, 2))) %>%
+  filter(!(name %in% c("exp_max", "p_rab_max"))) %>%
+  mutate(name = case_when(name %in% c("exp_min") ~ "human_exp",
+                          name %in% c("p_rab_min") ~ "p_rabid",
+                          TRUE ~ name)) %>%
+  group_by(scale, par = name) %>%
+  summarize(upper = max(value), lower = min(value),
+            dummy = "") -> pars
+# order them correctly & name them correctly
+pars$par <- factor(pars$par,
+                   levels = c("beta_ttimes", "beta_0", "sigma_e", "rho_max",
+                              "p_rabid", "p_death", "human_exp")
+)
+pars$par <- fct_recode(pars$par,
+                       `beta[t]` = "beta_ttimes", `beta[0]` = "beta_0",
+                       `sigma[e]` = "sigma_e", `rho[max]` = "rho_max",
+                       `p[rabid]` = "p_rabid", `p[death]` = "p_death",
+                       `E[i]` = "human_exp"
+)
+
+base_pars_A <-
+  ggplot(
+    data = pars,
+    aes(x = dummy, color = scale, fill = scale)
+  ) +
+  geom_errorbar(aes(ymax = upper, ymin = lower), position = position_dodge(width = 0.1)) +
+  scale_y_continuous(labels = function(y) round(y, 2)) +
[…]

### 2021-01-13 analysis/paper/minor_revision_1/response_reviewers.Rmd
updating with figs & files and making paper instructions

@@
 We have revised the methods and results section based on this and another reviewer's feedback.
-The methods section describing the decision tree now forefronts the parameter estimates with their sources, The results section now maps onto the major findings of each of our analyses, and hopefully clarifies the results of each section.
-We have also added a section to the discussion to regarding the validity of our assumptions (see lines xx - yy).* <!-- Done -->
+The methods section describing the decision tree now forefronts the parameter estimates with their sources.
+The results section now maps onto the major findings of each of our analyses, and hopefully clarifies the results of each section.
+We have also added a sentence to the discussion to regarding the validity of our assumptions (see lines 624 - 626).* <!-- Done -->
 
@@
 
-    *We have reorganized this discussion point, and and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines xx - yy).
-We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines xx).* <!-- Done -->
+    *We have reorganized this discussion point, and and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines 644 - 649).
+We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines 652 - 656).* <!-- Done -->
 
@@
     *We thank the reviewer for their helpful feedback and have revised the manuscript accordingly (see specific points regarding revisions to the methods and results in response to points below).
-Additionally, we have revised the text in the abstract (lines xx - yy) and in the conclusion (lines xx - xx) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP.
+Additionally, we have revised the text in the abstract (lines 63 - 68) and in the conclusion (line 764) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP.
 We also added a point that in other settings, addressing other barriers to PEP access may be more effective (i.e.
-direct costs of PEP in settings that charge for PEP) (see lines xx - yy in the 'Broader context' section of the discussion).* <!-- Done -->
+direct costs of PEP in settings that charge for PEP) (see lines 701 - 704 in the 'Broader context' section of the discussion).* <!-- Done -->
 
@@
 
-    *We have revised the methods section accordingly (lines xx - yy).
+    *We have revised the methods section accordingly (lines 242 - 277).
 Now all parameters are introduced in Table 1 with their sources and details, and Figure 1 describes how they are used in the decision tree framework.
@@
 
-    *We have added a short glossary to guide the reader through the manuscript (lines xx - yy) and also changed all mentions of CSB-II to primary clinic and CSB-I to secondary clinic, and use the term clinics throughout to refer to health facilities.* <!-- Done -->
+    *We have added a short glossary to guide the reader through the manuscript (lines 42 - 60) and also changed all mentions of CSB-II to primary clinic and CSB-I to secondary clinic, and use the term clinics throughout to refer to health facilities.* <!-- Done -->
 
@@
 We also point to potential solutions for managing these supply chains (i.e.
-piggybacking on EPI programs, see lines xx - yy).* <!-- Done -->
+piggybacking on EPI programs, see lines 714 - 730).* <!-- Done -->
 
@@
 
-    *We have revised the abstract to more clearly describe the data and methods we used (lines xx - yy).* <!-- Done -->
+    *We have revised the abstract to more clearly describe the data and methods we used (lines 42 - 46).* <!-- Done -->
 
@@
 
-    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved..* <!-- Done -->
+    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved (lines 84 - 85).* <!-- Done -->
 
@@
 
-    *We have clarified both the methods (lines xx-yy) and the results (lines xx - yy) based on the reviewer's helpful feedback.
-We point to some reasons for the observed variance in the results (lines xx - yy) and in the discussion (lines xx - yy), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data.
-We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines xx - yy).
-See Table S1.1 for R^2^ estimates for the linear models we used to compare the metrics, which we now also report in the main results text (line xx).
+    *We have clarified both the methods and added sample sizes(lines 162 - 180), as well as the results the results (lines 356 - 382) based on the reviewer's helpful feedback.
+We point to some reasons for the observed variance in the results (lines 360 - 363) and in the discussion (lines 603 - 617), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data.
+We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines 614 - 615).
+See Table S1.1 for R^2^ estimates for the linear models we used to compare the metrics, which we now also report in the main results text (line 381).
 We removed the CIs around the regression lines for both Fig 2C and Fig S1.4.* <!-- Done -->
@@
     *The overdispersion estimate is not on a log scale, but we incorrectly describe it's form in the methods and have corrected it.
-The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines xx - yy and Gelman reference) and it is accordingly scaled by the offset (population size).
+The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines 234 - 235) and it is accordingly scaled by the offset (population size).
 This means that the estimated overdispersion is actually quite large, as described in the text and vizualised in the figures.
@@
 
-    *We have revised this section based on this and another reviewers feedback (lines xx - yy).
+    *We have revised this section based on this and another reviewers feedback (lines 242 - 277).
 Briefly, we moved the table with the parameters and description to the top of the section, use the figure to show how these parameters were used, and additional choices regarding parameter ranges in the subsequent paragraph.* <!-- Done -->
@@
 
-    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (lines xx - yy) to ground this assumption up front (line xx).* <!-- Done -->
+    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (now line 275) to ground this assumption up front.* <!-- Done -->
 
@@
 
-    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines xx - yy).* <!-- Done -->
+    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines 84 - 85).* <!-- Done -->
 

### 2021-01-13 analysis/paper/manuscript.Rmd
updating with figs & files and making paper instructions

@@
 To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e.
-time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best in terms of a linear model.
+time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best to predict estimated travel times in a linear model.
 
@@
 ## Modeling human rabies deaths {#methods-burden}
+
+```{r exps}
+exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
+rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
+```
+
 We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework.
@@
 Fig 1 describes how these parameters are used in the decision tree and the key outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).
-
-```{r exps}
-exps <- read.csv(here_safe("data-raw/misc/moramanga_exposures.csv"))
-rho_max <- nrow(exps[exps$reported == "Yes", ])/nrow(exps)
-```
 
@@
 | $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]| 
-
-^1^ $p_{rabid}$ is constrained so that rabid reported bites cannot exceed the total expected number of rabies exposures ($E_i$) or maximum reporting in a given simulation ($\rho_max$).
+- ^1^ $p_{rabid}$ is constrained so that rabid reported bites cannot exceed the total expected number of rabies exposures ($E_i$) or maximum reporting in a given simulation ($\rho_{max}$).
 
@@
 `r comms_no_clins` communes lacked any health facilities.
-Finally, we explore a scenario where all additional primary clinics (totalling `r nrow(csbs_max)`) provision PEP.
+Finally, we explore a scenario where all additional primary clinics (totaling `r nrow(csbs_max)`) provision PEP.
 
@@
 Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
-After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D).
+After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D, see Supplementary Appendix section S2 for more details).
 For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
@@
 
-## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access to PEP. {#results-burden}
+## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access. {#results-burden}
 In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
@@
 ### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
-*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level , and (C) increase in vials needed per death averted based on the two models of reported bites.
+*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bites.
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
@@
 
+While we rely on a number of assumptions, they are based on data specific to Madagascar or from other similar settings and consistent with estimates from the literature more broadly (see Table 1 and section S4).
 Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths.
-While we rely on a number of assumptions, they are based on data specific to Madagascar or from other similar settings and consistent with estimates from the literature more broadly (see Table 1 and section S4).
 Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
@@
 We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
-Most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
+For example, most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
 [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers to PEP access.
@@
 However, our results suggest that PEP expansion alone cannot prevent the majority of rabies deaths, and even given maximal access, achieving ‘the last mile’, preventing deaths in the most remote populations, will require disproportionate resources [@delriovilasTribulationsLastMile2017].
-To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission, and eliminating deaths.
+To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission and eliminating deaths.
 Integrated Bite Case Management (IBCM) uses bite patient risk assessments to determine rabies exposure status, guide PEP administration, and trigger investigations of rabid animals, potentially identifying other exposed persons [@lechenneImportanceParticipatoryIntegrated2017; @rajeevHealthcareUtilizationProvisioning2019; @undurragaCostEffectivenessEvaluationNovel2017].

### 2021-01-20 analysis/paper/minor_revision_1/response_reviewers.Rmd
final files for resubmission

@@
 author: "Malavika Rajeev, Princeton University"
-date: "1/12/2020"
+date: "1/19/2020"
 output:
@@
 Both a document with track changes (of both the manuscript & the supplement) and clean separate documents have been submitted, and we have also verified our figures meet the requirements per the PACE tool.
-The reviewer comments are below with our response in italics.
+The reviewer comments are below with our response in italics (line numbers refer to combined and tracked document).
 
@@
 The results section now maps onto the major findings of each of our analyses, and hopefully clarifies the results of each section.
-We have also added a sentence to the discussion to regarding the validity of our assumptions (see lines 624 - 626).* <!-- Done -->
+We have also added a sentence to the discussion to regarding the validity of our assumptions (see lines 728 - 730).* <!-- Done -->
 
@@
 
-    *We have reorganized this discussion point, and and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines 644 - 649).
-We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines 652 - 656).* <!-- Done -->
+    *We have reorganized this discussion point, and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines 769 - 771).
+We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines 777 - 783).* <!-- Done -->
 
@@
     *We thank the reviewer for their helpful feedback and have revised the manuscript accordingly (see specific points regarding revisions to the methods and results in response to points below).
-Additionally, we have revised the text in the abstract (lines 63 - 68) and in the conclusion (line 764) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP.
+Additionally, we have revised the text in the abstract (lines 48 - 51) and in the conclusion (line 896) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP.
 We also added a point that in other settings, addressing other barriers to PEP access may be more effective (i.e.
-direct costs of PEP in settings that charge for PEP) (see lines 701 - 704 in the 'Broader context' section of the discussion).* <!-- Done -->
+direct costs of PEP in settings that charge for PEP) (see lines 832 - 835 in the 'Broader context' section of the discussion).* <!-- Done -->
 
@@
 
-    *We have revised the methods section accordingly (lines 242 - 277).
+    *We have revised the methods section accordingly (lines 250 - 293).
 Now all parameters are introduced in Table 1 with their sources and details, and Figure 1 describes how they are used in the decision tree framework.
@@
 
-    *We have added a short glossary to guide the reader through the manuscript (lines 42 - 60) and also changed all mentions of CSB-II to primary clinic and CSB-I to secondary clinic, and use the term clinics throughout to refer to health facilities.* <!-- Done -->
+    *We have removed acronyms as much as possible, now referring to CSB-II as primary clinics, CSB-I as secondary clinics, and use the term clinics provisioning PEP or alternatively PEP clinic throughout to refer to health facilities provisioning PEP.
+We've also tried to clarify when using commune and district that we are referring to administrative units.* <!-- Done -->
 
@@
 
-    *Done.* <!-- Done -->
+    *We added a ToC (not included in the tracked version, but included in the separate clean supplement resubmitted).* <!-- Done -->
 
@@
 We also point to potential solutions for managing these supply chains (i.e.
-piggybacking on EPI programs, see lines 714 - 730).* <!-- Done -->
+piggybacking on EPI programs, see lines 855 - 857).* <!-- Done -->
 
@@
 
-    *We have revised the abstract to more clearly describe the data and methods we used (lines 42 - 46).* <!-- Done -->
+    *We have revised the abstract to more clearly describe the data and methods we used (lines 48 - 50).* <!-- Done -->
 
@@
 
-    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved (lines 84 - 85).* <!-- Done -->
+    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved (lines 70 - 75).* <!-- Done -->
 
@@
 
-    *We have clarified both the methods and added sample sizes(lines 162 - 180), as well as the results the results (lines 356 - 382) based on the reviewer's helpful feedback.
-We point to some reasons for the observed variance in the results (lines 360 - 363) and in the discussion (lines 603 - 617), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data.
-We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines 614 - 615).
+    *We have clarified both the methods and added sample sizes(lines 171 - 187), as well as the results the results (lines 404 - 432) based on the reviewer's helpful feedback.
+We point to some reasons for the observed variance in the results (lines 421 - 436) and in the discussion (lines 706 - 717), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data.
+We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines 718 - 720).
 See Table S1.1 for R^2^ estimates for the linear models we used to compare the metrics, which we now also report in the main results text (line 381).
@@
     *The overdispersion estimate is not on a log scale, but we incorrectly describe it's form in the methods and have corrected it.
-The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines 234 - 235) and it is accordingly scaled by the offset (population size).
+The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines 241 - 243) and it is accordingly scaled by the offset (population size).
 This means that the estimated overdispersion is actually quite large, as described in the text and vizualised in the figures.
@@
 
-    *We have revised this section based on this and another reviewers feedback (lines 242 - 277).
+    *We have revised this section based on this and another reviewers feedback (lines 250 - 293).
 Briefly, we moved the table with the parameters and description to the top of the section, use the figure to show how these parameters were used, and additional choices regarding parameter ranges in the subsequent paragraph.* <!-- Done -->
@@
 
-    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (now line 275) to ground this assumption up front.* <!-- Done -->
+    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (now line 291 - 293) to ground this assumption up front.* <!-- Done -->
 
@@
 
-    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines 84 - 85).* <!-- Done -->
+    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines 96 - 97).* <!-- Done -->
 

### 2021-01-20 analysis/paper/manuscript.Rmd
final files for resubmission

@@
       equal_contributor: "yes"
-  - C.J.E. Metcalf:
+  - C. Jessica E. Metcalf:
       institute: [Princeton]
@@
   - astre: ASTRE, Univ Montpellier, CIRAD, INRAE, Montpellier, France 
-  - fofifa: FOFIFA-DRZVP, Antananarivo, Madagascar6 World Health Organization, Madagascar
+  - fofifa: FOFIFA-DRZVP, Antananarivo, Madagascar
+  - OMS: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
   - smeetn: Service contre les Maladies Endémo-épidémiques et Tropicales  Négligées, Ministère de la Santé Publique, Antananarivo, Madagascar
-  - OMS: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
   - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
@@
 We find that travel times strongly predict reported bite incidence across the country.
-Using resulting estimates in an adapted decision tree framework, we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
+Using resulting estimates in an adapted decision tree, we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally.
 We estimate 960 human rabies deaths annually (95% Prediction Intervals (PI):`r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with PEP averting an additional 800 deaths (95% PI: `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) each year.
-Under these assumptions, we find that expanding PEP to one clinic per district (N = 83 additional clinics) could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major primary clinics provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
+Under these assumptions, we find that expanding PEP to one clinic per district (83 additional clinics) could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major primary clinics provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality.
 Our quantitative estimates are most sensitive to assumptions of underlying rabies exposure incidence, but qualitative patterns of the impacts of travel times and expanded PEP access are robust.
@@
 Given the potential for countries to use Gavi funding to expand access to PEP in the coming years, this framework could be used as a first step to guide expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted and baseline data on rabies risk is lacking.
-While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyze progress towards achieving zero rabies deaths.
+While better PEP access should save many lives, improved outreach, surveillance, and dog vaccination will be necessary, and if rolled out with Gavi investment, could catalyze progress towards achieving zero rabies deaths.
 
@@
 Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by 2030, and recently, Gavi, a multilateral organization that aims to improve access to vaccines in the poorest countries, added human rabies vaccine to it’s portfolio.
-In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP and extrapolate human rabies deaths in Madagascar.
+In this study, we estimated reported incidence of patients seeking PEP in relation to travel times to clinics provisioning PEP and extrapolate human rabies deaths in Madagascar.
 We find that PEP currently averts around `r round(natl_base$averted_mean[1], -2)` deaths each year, but that the burden remains high (`r round(natl_base$deaths_mean[1], -2)` deaths/ year), particularly in remote, hard-to-reach areas.
-We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone is unlikely to achieve the global goal of zero human deaths from dog-mediated rabies by 2030.
+We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results reaffirm that expansion alone is will not achieve the global goal of zero human deaths from dog-mediated rabies by 2030.
 Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other Low- and Middle-Income countries, forward on the path to rabies elimination.
@@
 In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to its own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019].
-Other than at these 31 anti-rabies medical centers (ARMC), PEP is not available at any other public clinics or through the private sector.
+Other than at these 31 clinics, PEP is not available at any other public clinics or through the private sector.
 In addition, there is limited control of rabies in dog populations and the disease is endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011].
@@
 
-### Box 1: Glossary
-- **Institut Pasteur de Madagascar (IPM):** private Malagasy research institution located in the capital of Antananarivo; part of the Institut Pasteur network and also houses the National Rabies Reference Laboratory (NRRL), as well as the vaccination clinic (one of the 31 clinics where PEP is provided).
-- **Antirabies medical center (ARMC):** public clinics (N = 31) in Madagascar where PEP is provided free of charge to patients.
-IPM provides PEP to these clinics, and the clinical staff implement vaccination (can be through emergency, immunizaton, or primary care departments).
-- **Primary clinic:** a primary health clinic formally known as 'Centre de Sante Niveau II' or CSB-II in French; these are major health centers that should theoretically have immunization capacity.
-- **Secondary clinic:** a secondary health clinic formally known as 'Centre de Sante Niveau I' or CSB-I in French; these are lower level health posts that are often staffed by a single clinician and often without formal immunization capacity.
-- **District:** second-level administrative divisions of Madagascar. There are 114 districts in Madagascar.
-- **Commune:** third-level administrative division of Madagascar below the district level. There are 1579 communes in Madagascar.
-
 # Methods
@@
 
-## Estimating geographic access to ARMC {#methods-ttimes}
-To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two raster datasets: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2^ scale (Fig S1.1A) and 2) the population estimates from the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2^ resolution, Fig S1.1B), which we aggregated to the friction surface.
-
-From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018].
-We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and Euclidean distance, i.e.
-the minimum distance from the administrative unit centroid to any ARMC.
-We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018).
+## Estimating geographic access to PEP {#methods-ttimes}
+To estimate mean and population weighted travel times to the nearest clinic, we used two raster datasets: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2^ scale (Fig S1.1A) and 2) the population estimates from the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2^ resolution, Fig S1.1B), which we aggregated to the friction surface.
+
+From GPS locations of the 31 clinics that currently provision PEP, we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018].
+We then extracted the mean and population-weighted mean travel times for each district (2nd level administrative division, N = 114) and commune (the administrative unit below the district, N = 1579), and Euclidean distance, i.e.
+the minimum distance from the administrative unit centroid to any clinic.
+We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the district and communes boundaries (as of October 31, 2018).
 To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e.
-time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best to predict estimated travel times in a linear model.
+time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga clinic bite patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best to predict estimated travel times in a linear model.
 
 ## Estimating bite incidence {#methods-bitedata}
-We used two datasets on bite patients reporting to ARMC:
-
-1)	A national database of individual bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017.
-These forms were submitted with frequencies ranging from monthly to annually, included the patient reporting date and were resolved to the district level (patient residence).
-
-2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga District ARMC resolved to commune level.
+We used two datasets on bite patients reporting to clinics for PEP:
+
+1)	A national database of individual bite patient forms from the 31 clinics provisioning PEP across the country between 2014 - 2017.
+These forms were submitted to IPM with frequencies ranging from monthly to annually, included the patient reporting date and were resolved to the district level (patient residence).
+
+2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga clinic resolved to the commune level.
 
@@
 We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see Supplementary Appendix, section S2).
-At both the commune and district level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit.
+At both the commune and district administrative level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit.
 For national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.
@@
 
-where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the ARMC as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
+where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the clinic as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
 We fit this model to both the national data (district level) and the Moramanga data (commune level).
@@
 |:----------------|:----------------:|:----------------------------------|:---------:|
-| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
+| $B_i$ |	Function of travel time to closest clinic provisioning PEP | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
 | $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
 | $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures^1^ | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga ARMC (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
+| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga PEP clinic (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
 | $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]| 
@@
 As there is little data on dog population size and human exposure incidence in Madagascar[@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
-Given previously high observed compliance in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], we assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing rabies.
+Given previously high observed compliance in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], we assume that all rabies exposed patients who report to a clinic receive and complete PEP, and PEP is completely effective at preventing rabies.
 
@@
 We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic.
-We select the primary clinic (primary health facility, see Glossary) in the highest density grid cell of the administrative unit as candidates for expansion.
-For the `r comms_no_csb2` communes without a primary clinic, we chose the secondary clinic (secondary health facility, see Glossary) in the highest density grid cell.
+We select the primary clinic (primary health facility, usually with capacity to provision vaccines) in the highest density grid cell of the administrative unit as candidates for expansion.
+For the `r comms_no_csb2` communes without a primary clinic, we chose the secondary clinic (secondary health facility, often without formal vaccination capacity) in the highest density grid cell.
 `r comms_no_clins` communes lacked any health facilities.
@@
 
-We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from an existing ARMC for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing ARMC.
-We simulated expansion of ARMC to each district using these three metrics and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
+We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from a existing PEP clinic that provisions PEP for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing PEP clinic.
+We simulated expansion of clinics provisioning PEP to each district using these three metrics and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
 For the full simulation of expanded access, once clinics reduced travel times for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` living greater than $x$ hrs away, starting with $x$ = 3 hrs), we reduced the travel time threshold by 25%.
@@
 All processed data, code, and outputs are archived on Zenodo (http://doi.org/10.5281/zenodo.4064312 and https://doi.org/10.5281/zenodo.4064304),  and maintained at https://github.com/mrajeev08/MadaAccess.
-The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral ARMC.
+The raw bite patient data at the national level are maintained in two secure REDCap (project-redcap.org) databases, one for IPM and another for all peripheral clinics provisioning PEP.
 These databases were last queried on September 19, 2019 for these analyses.
@@
 
-## Travel times to ARMC are high and variable across Madagascar. {#results-ttimes}
+## Estimates of travel times to clinics are high and variable across Madagascar. {#results-ttimes}
 ```{r ttime stats}
@@
 Based on the estimates from the friction surface, approximately `r round(prop_over3_base*100, 0)`% of the population of Madagascar are estimated to live over 3 hours from a clinic (Fig 2).
-However, we found that these estimates underestimated both driving times across the country and patient-reported travel times to the Moramanga ARMC (Fig 2C).
+However, we found that these estimates underestimated both driving times across the country and patient-reported travel times to the Moramanga PEP clinic (Fig 2C).
 Patient reported travel times were highly variable for a given commune compared to the estimated travel times (Fig S1.2), potentially due to the fact that the friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
 the presence of a road indicates that all travel through that grid cell is by vehicle).
-However, Moramanga ARMC patients reported using multiple modes of transport, with some individuals walking days to a clinic (Fig S1.3).
+However, patients reported using multiple modes of transport, with some individuals walking days to the Moramanga PEP clinic (Fig S1.3).
 
@@
 
-### *Fig 2. Travel times to ARMC across Madagascar.*
+### *Fig 2. Travel times to clinics provisioning PEP across Madagascar.*
 *(A) Estimated at an ~ 1 km2 scale.
 (B) Distribution of the population across travel times.
-(C) Correlation between ground-truthed travel times (mean of patient reported travel times to the Moramanga ARMC at the commune level and reported driving times between GPS points) and friction surface travel time estimates.
+(C) Correlation between ground-truthed travel times (mean of patient reported travel times to the Moramanga PEP clinic at the commune level and reported driving times between GPS points) and friction surface travel time estimates.
 The vertical lines show the 95% quantiles for reported travel times and the point size shows the number of observations for each commune.
@@
 
-Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
+Bite incidence estimates generally increased with decreasing weighted travel times at both administrative scales (district and commune), although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
 After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D, see Supplementary Appendix section S2 for more details).
@@
 ### *Fig 3. The network of patient presentations and estimates of annual bite incidence.*
-*(A) at the district level for the national data and (B) commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each ARMC for which we have data.
+*(A) at the district level for the national data and (B) commune level for the Moramanga data: circles with a black outline represent the total number of patients reporting to each clinic for which we have data.
 Color corresponds to the clinic catchment.
 Circles with a white outline are the total number of bites reported for that administrative unit (plotted as the centroid).
-Lines show which ARMC those patients reported to, with the line width proportional to number of patients from that district reporting to the ARMC; flows of less than 5 patients were excluded.
+Lines show which clinic those patients reported to, with the line width proportional to number of patients from that district reporting to the clinic; flows of less than 5 patients were excluded.
 Out-of-catchment reporting is indicated where points and line colors are mismatched.
@@
 Our modeling results show that travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales with the best fitting models including travel times and an overdispersion parameter (Fig 4, see Supplementary Appendix section S3 for comparisons to models with catchment effects and with population size as a covariate).
-As the predictions from the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
+As the predictions of the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
 Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig S3.3), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
@@
 
-## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access. {#results-burden}
+## Current provisioning of PEP substantially reduces human rabies deaths, but incidence of deaths remains high in areas with poor access {#results-burden}
+
 In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
-Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year.
+We estimate that under the current system of 31 clinics in Madagascar provisioning PEP that approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies are prevented through PEP each year.
 Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar.
@@
 
-kable(table, digits = -1, col.names = c("Model", "Reported bite incidence per 100k", "Burden of deaths", 
-                                        "Deaths averted by current PEP provisioning"))           
+kable(table, digits = -1, col.names = c("**Model**", "**Reported bite incidence per 100k**", "**Burden of deaths**", 
+                                        "**Deaths averted by current PEP provisioning**"))           
 ```
@@
 The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
-Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current ARMC.*
-
-## Expanding PEP access to underserved populations is effective at reducing human rabies deaths, but this effect saturates as clinics are added. {#results-incremental}
+Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current clinics provisioning PEP*
+
+## Expanding PEP access to underserved populations is effective at reducing human rabies deaths, but this effect saturates as more clinics provision PEP {#results-incremental}
 
@@
 ### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
-*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bites.
+*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional clinics provisioning PEP are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bites.
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
@@
 However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all primary clinics (N = `r max(natl_preds$scenario)`).
-This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all major primary clinics.
+This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all primary clinics.
 With reduced travel times, over 20% of exposures will still not seek PEP (Fig S5.5), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
@@
 Data limitations introduced bias and uncertainty to our estimates.
-For example, travel times from the MAP friction surface underestimated patient-reported travel times, with discrepancies between assigned transport speeds (from the Open Street Map user community, or country cluster data [@weissGlobalMapTravel2018]) and realities of local travel.
+For example, travel times from the Malaria Atlas Project friction surface underestimated patient-reported travel times, with discrepancies between assigned transport speeds (from the Open Street Map user community, or country cluster data [@weissGlobalMapTravel2018]) and realities of local travel.
 In Madagascar, the presence of paved roads does not necessarily reflect their quality or the modes of transport used.
-Moramanga ARMC patients reported various transport methods and highly variable travel times even within a single commune.
+Patients seeking PEP at the Moramanga clinic reported various transport methods and highly variable travel times even within a single commune.
 While patient-reported travel times may lack precision from recall and estimation error, they likely better reflect lived experience; validated travel times [@rayAccessModComputingGeographic2008, ihantamalalaImprovingGeographicalAccessibility2020] could improve estimates of spatial health inequities.
 Similarly, modeled estimates of population distribution [@linardPopulationDistributionSettlement2012a] also introduce uncertainty.
-Our analysis of data from the Moramanga District indicate that variation at the sub-district level is high and impacts health seeking behavior.
+Our analysis of data from the Moramanga District indicates that variation at the sub-district level is high and impacts health seeking behavior.
 However, we lacked fine-scale data from other catchments for comparison.
@@
 Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
-Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
-We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
-
-We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
-For example, most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
+Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in the captial of Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+We also assume that clinics reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
+
+We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014; castillo2020behavioral] all play a role.
+For example, most PEP clinics also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
 [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers to PEP access.
@@
 Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care, and human behavior, as illustrated by the case studies in Box 1.
+However, it is very likely that the impacts of improving access to PEP could be further increased with outreach and awareness raising efforts that we were unable to parameterize.
 
@@
 1.	Anosibe An’ala District (population ~ `r format(round(mada_districts$pop[mada_districts$district == "Anosibe-An'ala"]), big.mark = ",")`), south of Moramanga, has moderate incidence of bite patients (~ `r round(aa_bites$inc)`/100,000 persons) even though travel times often exceed 24 hours.
-While a road connects the main Anosibe An’ala commune to the Moramanga ARMC, it is only passable by large trucks during much of the rainy season, with speeds usually < 10km per hour.
+While a road connects the main Anosibe An’ala commune to the Moramanga PEP clinic, it is only passable by large trucks during much of the rainy season, with speeds usually < 10km per hour.
 Over `r trans_perc(prop_cat1_AA$prop_cat1)`% of patients from Anosibe An’ala had been in close proximity or touched a person that died from rabies (four suspect human rabies deaths of patients who did not receive any PEP), whilst of patients with Category II and III exposures that were interviewed, `r glue("{nrabid}/{ninterviewed}") ` (`r trans_perc(nrabid/ninterviewed)`%) were bitten by probable rabid dogs.
-Given the high travel times (although underestimated by the friction surface) and incidence of reported rabies exposures and deaths, we predict a large but unobserved rabies burden in this remote community (~`r round(aa_deaths$deaths_mean[1], 2)` deaths per year) and we ranked an ARMC provisioning PEP in Anosibe An’ala `r aa_added`th for travel time reductions.
+Given the high travel times (although underestimated by the friction surface) and incidence of reported rabies exposures and deaths, we predict a large but unobserved rabies burden in this remote community (~`r round(aa_deaths$deaths_mean[1], 2)` deaths per year) and we ranked a clinic provisioning PEP in Anosibe An’ala `r aa_added`th for travel time reductions.
 Other remote communities likely experience similar high and unrecognized burden, but improved surveillance is necessary to identify such areas.
@@
 Recently, a middle-aged taxi driver died of rabies in suburban Antananarivo.
-The day after being bitten by an unknown dog, he reported to a clinic that referred him to the ARMC at IPM for PEP, approximately one hour’s drive from his home.
+The day after being bitten by an unknown dog, he reported to a clinic that referred him to the closest clinic provisioning PEP, approximately one hour’s drive from his home.
 His family urged him to go, but he did not believe his risk was high and decided not to seek further care.
@@
 
-Recent studies have estimated access to health-seeking behavior and PEP completion and adherence, but not directly linked these estimates to burden [@tarantolaRabiesPostexposureProphylaxis2017; @zaidiGeographicVariationAccess2013].
+Recent studies have estimated access to health-seeking behavior and PEP completion and adherence, but not directly linked these estimates to burden [@tarantolaRabiesPostexposureProphylaxis2017; @zaidiGeographicVariationAccess2013; @delapuente-leon2020].
 Our approach for incorporating access to vaccines (echoing [@Hegde2019dt; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @metcalfTransportNetworksInequities2015; @aleganaTreatmentseekingBehaviourLow2017]) into burden estimation methods could guide provisioning of PEP to maximize impacts.
@@
 [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015].
-In other settings, similar statistical approaches could be used to identify and quantifying key barriers to PEP seeking behavior.
+In other settings, similar statistical approaches could be used to identify and quantify key barriers to PEP seeking behavior.
 For example, reducing the direct cost of PEP is likely to be of more importance than increasing geographic access where PEP costs are high.
 
-Our revised estimate of rabies deaths in Madagascar using this approach was higher than previous [@rajeevHealthcareUtilizationProvisioning2019], which assumed uniform reporting of 85%, but remained within the range of other empirical and modeling studies from low-income countries [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @lyRabiesSituationCambodia2009; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011].
+Our revised estimate of rabies deaths in Madagascar using this approach was higher than previously estimated (between 280 - 750 deaths/year) [@rajeevHealthcareUtilizationProvisioning2019], which assumed uniform reporting of 85%, but remained within the range of other empirical and modeling studies from low-income countries [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008; @lyRabiesSituationCambodia2009; @sudarshanAssessingBurdenHuman2007; @tenzinDogBitesHumans2011].
 Our estimates of vial demand depend on use of the new abridged intradermal regimen [@tarantolaIntradermalRabiesPostexposure2018], which has been adopted by the Ministry of Health in Madagascar.
-However, most ARMC staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moramanga [@rajeevHealthcareUtilizationProvisioning2019].
+However, most clinic staff were not aware of WHO classifications of exposure categories, and vaccination of Category I exposures (those not requiring PEP) remains common practice, comprising 20% of vial demand in Moramanga [@rajeevHealthcareUtilizationProvisioning2019].
 
@@
 Decentralized provisioning mechanisms, for example adopting routine childhood vaccine supply chains, or novel vaccine delivery methods such as drones [@boeras2019use], may mitigate these challenges.
-When nerve tissue vaccines were used in Madagascar, clinics requested vaccine upon demand and PEP access was more widespread, but provisioning the more expensive cell culture vaccines to all clinics became too costly [@andriamandimbySurveillanceControlRabies2013].
+When nerve tissue vaccines were used in Madagascar, clinics requested vaccines upon demand and PEP access was more widespread, but provisioning the more expensive cell culture vaccines to all clinics became too costly [@andriamandimbySurveillanceControlRabies2013].
 Widespread vaccine provisioning is therefore feasible given Madagascar’s health infrastructure, if cost barriers are removed.
@@
 Latin American countries, where significant progress has been made towards elimination, aim for one PEP clinic per 100,000 persons.
-In Madagascar this would require around `r min(catch_pops[mean_pops < 1e5]$scenario)` additional ARMC.
+In Madagascar this would require around `r min(catch_pops[mean_pops < 1e5]$scenario)` additional clinics provisioning PEP.
 We predict that Gavi investment would be highly cost-effective, greatly reducing deaths by expanding PEP supply to underserved areas.
@@
 
-We thank all the clinicians and staff at the ARMC across the country.
+We thank all the clinicians and staff at the clinics across the country.
 We are grateful to IPM and the Ministry of Public Health who collect and maintain data on PEP provisioning.
@@
 
-## S1. Estimating travel times to the closest ARMC {#S1}
+## S1. Estimating travel times to the closest clinic provisioning PEP {#S1}
 
@@
 
-### *Fig S1.1. Raster inputs to estimate travel times to the closest ARMC for Madagascar.*
-_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current ARMC (N = 31) shown with black x's.
+### *Fig S1.1. Raster inputs to estimate travel times to the closest clinic provisioning PEP for Madagascar.*
+_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current clinics provisioning PEP (N = 31) shown with black x's.
 (B) Population estimates resampled to the same friction surface._
@@
 ### *Fig S1.2. Raw patient reported and driving travel time data.*
-_(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga ARMC.
-(B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
+_(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga PEP clinic (B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
 Paths are Bezier curves from origin to destination, and do not show actual paths driven._
@@
 
-### *Fig S1.3. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga ARMC.*
+### *Fig S1.3. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga PEP clinic*
 \newpage
@@
 ### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) vs. estimates from friction surface.*
-_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga ARMC for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only) ._
+_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga PEP clinic for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only) ._
 
@@
 
-Most patients from each district reported to their closest ARMC by the weighted travel time metric (Fig 3).
+Most patients from each district reported to their closest clinic provisioning PEP by the weighted travel time metric (Fig 3).
 Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
@@
 *The circles are each of the primary clinics across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
-The large white crosses show the location of the existing 31 ARMC in Madagascar and the smaller white crosses are the additional primary clinics in the country which were added in the final step but not ranked.*
+The large white crosses show the location of the existing 31 clinics provisioning PEP in Madagascar and the smaller white crosses are the additional primary clinics in the country which were added in the final step but not ranked.*
 \newpage
@@
 The columns are ordered by the number or clinics at each step: baseline (N = 31), + 83 (1 per district), + 200, + 600, + 1406 (1 per commune), and max (+ 1696 clinics, all additional primary clinics in the country).
-Grey pixels show the location of ARMC at each step.
+Grey pixels show the location of clinics provisioning PEP at each step.
 Commune and district values are the average grid cell travel times weighted by the population in each cell.*

### 2021-01-20 analysis/paper/make_paper.R

@@ -12,7 +12,7 @@ rmarkdown::render(here("analysis/paper/minor_revision_1/response_reviewers.Rmd")
                   output_dir = here("analysis/paper/minor_revision_1"))
 
 # upload to google drive for feedback
-googledrive::drive_update('MadaAccess', here('analysis/paper/minor_revision_1/manuscript_tracked.docx'))
+googledrive::drive_update('MadaAccess', here('analysis/paper/minor_revision_1/manuscript_supplement_tracked.docx'))
 googledrive::drive_put(here('analysis/paper/minor_revision_1/response_reviewers.docx'), 'MadaAccess_response')
 
 # For submitting

### 2021-01-20 analysis/paper/minor_revision_1/word2pdf.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+cd analysis/paper/minor_revision_1
+
+find . -type f \( -iname \*.doc -o -iname \*.docx \) -print0 | while read -d $'\0' file
+ do
+     automator -i "$file" ~/Desktop/w2pdf.app
+ done

### 2021-01-20 analysis/src/figures/M3_bitemaps.R

@@ -136,11 +136,11 @@ leg_pts <- get_bezleg(
   min_size = 5, offset_long = 0.4, offset_lat = -1.5
 )
 leg_pts$ptcol <- case_when(
-  leg_pts$index == 3 ~ "ARMC",
+  leg_pts$index == 3 ~ "Clinic",
   leg_pts$index == 1 ~ "District"
 )
 leg_cols <- c("black", "white")
-names(leg_cols) <- c("ARMC", "District")
+names(leg_cols) <- c("Clinic", "District")
 catch_fills <- c(catch_fills, leg_cols)
 
 mada_map_A <- ggplot() +
@@ -197,7 +197,7 @@ mada_map_A <- ggplot() +
   ) +
   geom_text(
     data = filter(ungroup(leg_pts), long == min(long) | long == max(long)),
-    aes(x = long, y = lat, label = c("District", "ARMC")), hjust = c(1, 1),
+    aes(x = long, y = lat, label = c("District", "Clinic")), hjust = c(1, 1),
     angle = 90,
     nudge_y = -0.5
   ) +
@@ -304,11 +304,13 @@ leg_pts <- get_bezleg(
   min_size = 5, offset_long = -0.1, offset_lat = -0.65
 )
 leg_pts$ptcol <- case_when(
-  leg_pts$index == 3 ~ "ARMC",
-  leg_pts$index == 1 ~ "District"
+  leg_pts$index == 3 ~ "Clinic",
+  leg_pts$index == 2 ~ "Commune",
+  leg_pts$index == 1 ~ "Commune"
 )
 leg_cols <- c("black", "white")
-names(leg_cols) <- c("ARMC", "District")
+names(leg_cols) <- c("Clinic", "Commune")
+catch_fills <- c(catch_fills, leg_cols)
 
 mora_map_B <- ggplot() +
   geom_sf(
@@ -360,9 +362,9 @@ mora_map_B <- ggplot() +
   ) +
   geom_text(
     data = filter(ungroup(leg_pts), long == min(long) | long == max(long)),
-    aes(x = long, y = lat, label = c("District", "ARMC")), hjust = c(1, 1),
+    aes(x = long, y = lat, label = c("Commune", "Clinic")), hjust = c(1, 1),
     angle = 90,
-    nudge_y = -0.09
+    nudge_y = -0.08
   ) +
   scale_discrete_manual(aes = "stroke", values = c(1.2, 0.5), guide = "none") +
[…]

### 2021-01-20 analysis/src/figures/M6_incremental.R

@@ -92,7 +92,7 @@ vials_per_death <- ggplot(
   scale_color_manual(values = model_cols, labels = scale_labs, name = "Scale") +
   scale_fill_manual(values = model_cols, labels = scale_labs, name = "Scale") +
   labs(
-    x = "# Additional ARMC", y = "Average vials per \n death averted",
+    x = "# Additional clinics \n provisioning PEP", y = "Average vials per \n death averted",
     tag = "C"
   ) +
   coord_cartesian(clip = "off") +

### 2021-01-20 analysis/src/figures/SM2_bitedata.R

@@ -141,7 +141,7 @@ throughput_A <- ggplot(data = throughput, aes(x = date_reported, y = reorder(cta
   theme_minimal_hgrid() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
   xlab("Year") +
-  ylab("ARMC") +
+  ylab("Clinic provisioning PEP") +
   labs(tag = "A")
 
 reporting %>%

### 2021-01-20 analysis/src/figures/SM5_shifts.R

@@ -87,7 +87,7 @@ ggplot(
     labels = mets, name = "Rank metric"
   ) +
   scale_linetype(labels = mets, name = "Rank metric") +
-  labs(x = "# Additional ARMC", y = "Average annual deaths") +
+  labs(x = "# Additional clinics \n provisioning PEP", y = "Average annual deaths") +
   facet_wrap(~scale, nrow = 1) +
   theme_minimal_hgrid() +
   theme(
@@ -157,7 +157,7 @@ ggplot(data = clins_ranked, aes(x = scenario)) +
     color = "#0B775E",
     size = 4, alpha = 1
   ) +
-  labs(x = "# Additional ARMC", y = "Average annual deaths \n (smoothed)") +
+  labs(x = "# Additional clinics \n provisioning PEP", y = "Average annual deaths \n (smoothed)") +
   theme_minimal_hgrid(font_size = 11) +
   theme(
     axis.text.x = element_text(angle = 45, hjust = 1),
@@ -320,7 +320,7 @@ ggplot(
   ) +
   scale_y_continuous(trans = "sqrt", breaks = c(0.5, 1, 3, 5, 10, 15)) +
   scale_x_discrete(labels = scenario_labs) +
-  labs(x = "# Additional ARMC", y = "Travel times (hrs)", tag = "A") +
+  labs(x = "# Additional clinics \n provisioning PEP", y = "Travel times (hrs)", tag = "A") +
   coord_flip() +
   theme_minimal_hgrid(color = "grey50") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) -> ttimes_dist
@@ -382,7 +382,7 @@ ggplot(
     guide = "none"
   ) +
   scale_x_discrete(labels = scenario_labs) +
-  labs(x = "# Additional ARMC", y = "Reporting", tag = "A") +
+  labs(x = "# Additional clinics \n provisioning PEP", y = "Reporting", tag = "A") +
   coord_flip() +
   theme_minimal_hgrid(color = "grey50") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) -> reporting_dist
@@ -478,7 +478,7 @@ ggplot(
   ) +
   scale_y_continuous(trans = "log", breaks = c(100, 1e3, 1e4, 1e5, 1e6)) +
   scale_x_discrete(labels = scenario_labs) +
-  labs(x = "# Additional ARMC", y = "Catchment pop size", tag = "A") +
[…]

### 2021-02-10 analysis/paper/manuscript.Rmd
fixing so no figs for sub & word output only

@@
 link-citations: no
+params:
+  nofig: FALSE
 output:
@@
 
-no_figs <- FALSE
+no_figs <- params$nofig
 if(no_figs) include_graphics <- function(x, ...) { return(invisible(NULL)) }

### 2021-03-08 analysis/paper/manuscript.Rmd
fixing rendering for final proof submission

@@
 no_figs <- params$nofig
-if(no_figs) include_graphics <- function(x, ...) { return(invisible(NULL)) }
+
+# Dummy for typesetting submission files
+if(no_figs) {
+  include_graphics <- function(x, ...) { return(invisible(NULL)) }
+  kable <- function(x, ...) { return(invisible(NULL)) }
+} else {
+  # explicitly call the alternative otherwise if you're rendered with no figs = FALSE
+  # still wont knit figs
+  include_graphics <- knitr::include_graphics
+  kable <- knitr::kable
+}
 ```
@@
 Due to the spatially restricted nature of PEP provisioning and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country.
-Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, but did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
+Previously, we estimated the burden of rabies in Madagascar using data from a single district to extrapolate to the national scale, but did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019].
 Here, we provide revised estimates of human rabies deaths by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.
@@
 
+## Ethics statement
+Data collection from the Moramanga District was approved by the Princeton University IRB (7801) and the Madagascar Ministry of Public Health Ethics Committee (105-MSANP/CE).
+Oral informed consent was obtained from all interviewed participants.
+Data collected from bite patients at the national level are maintained jointly by the Ministry of Health and IPM as a routine part of PEP provisioning.
+
 ## Estimating geographic access to PEP {#methods-ttimes}
-To estimate mean and population weighted travel times to the nearest clinic, we used two raster datasets: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2^ scale (Fig S1.1A) and 2) the population estimates from the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2^ resolution, Fig S1.1B), which we aggregated to the friction surface.
+To estimate mean and population weighted travel times to the nearest clinic, we used two raster datasets: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2^ scale and 2) the population estimates from the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2^ resolution), which we aggregated to the friction surface (Fig A in S1 Text).
 
@@
 To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e.
-time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga clinic bite patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best to predict estimated travel times in a linear model.
+time it took to travel by car between two locations excluding break times (N = `r nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga clinic bite patients (N = `r patient_nobs`), see Fig B in S1 Text for raw data) by seeing which had the highest R^2^ when predicting estimated travel times in a linear model.
 
@@
 For the national data, some clinics did not submit any data, or had substantial periods (months to a whole year), with no submitted data.
-To correct for this, we exclude periods of 15 consecutive days with zero submitted records (see Supplementary Appendix, section S2).
+To correct for this, we exclude periods of 15 consecutive days with zero submitted records (see S2 Text).
 For each clinic we divided the total number of bites reported in a given year by the estimated proportion of forms which were not submitted (under-submission).
 Due to yearly variation in submissions, we took the average of annual bite incidence estimates aggregated to district level.
-We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see Supplementary Appendix, section S2).
+We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see S2 Text).
 At both the commune and district administrative level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit.
@@
 
-where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the clinic as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune).
+where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the clinic as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location $i$ (district or commune).
 We fit this model to both the national data (district level) and the Moramanga data (commune level).
@@
 
-As travel times are correlated with population size (Fig S3.1), we also compared how well bites were predicted by population size alone, and in combination with travel times.
+As travel times are correlated with population size (Fig A in S3 Text), we also compared how well bites were predicted by population size alone, and in combination with travel times.
 For the models with population size, we removed the offset and used either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors.
@@
 We tested whether the catchment random effect captured overdispersion in the data (i.e.
-variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter: $\epsilon_i \sim norm(0,\ \sigma_e)$, where $\sigma_e$ is the standard deviation around a random variable with mean of zero [@gelman2006data]:
+whether variance equals the mean, the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter: $\epsilon_i \sim norm(0,\ \sigma_e)$, where $\sigma_e$ is the standard deviation around a random variable with mean of zero [@gelman2006data]:
 
@@
 
-where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters for a given model. 
+and where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of all parameters for a given model. 
 We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’ [@R-rjags].
-We used model estimates to generate fitted and out-of-fit predictions, and examined the sensitivity of estimates to adjustments for under-submission of forms (Supplementary Appendix, section S3).
+We used model estimates to generate fitted and out-of-fit predictions and examined the sensitivity of estimates to adjustments for under-submission of forms (S3 Text).
 
@@
 | $B_i$ |	Function of travel time to closest clinic provisioning PEP | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
-| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
+| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig A in S4 Text|
 | $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures^1^ | [@rajeevHealthcareUtilizationProvisioning2019] |
-| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga PEP clinic (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
+| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga clinic (average travel time estimate of ~ 3.12 minutes)| [@rajeevHealthcareUtilizationProvisioning2019] |
 | $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]| 
@@
 Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution.
-Parameter values and sources are in Table 1.*
-
-For $E_i$, we center the distribution at the lower end of our estimated exposure incidence from the Moramanga District (42 exposures/100,000 persons), with a range applied assuming 1% rabies incidence in dogs (estimated across a range of human-to-dog ratios between 5 - 25) and that on average a rabid dog exposes 0.39 persons [@whorabiesmodelingconsortiumPotentialImpactImproved2019] (see Fig S4.1).
-As there is little data on dog population size and human exposure incidence in Madagascar[@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
+Parameter values and sources are listed in Table 1.*
+
+For $E_i$, we center the distribution at the lower end of our estimated exposure incidence from the Moramanga District (42 exposures/100,000 persons), with a range applied assuming 1% rabies incidence in dogs (estimated across a range of human-to-dog ratios between 5 - 25) and that on average a rabid dog exposes 0.39 persons [@whorabiesmodelingconsortiumPotentialImpactImproved2019] (see Fig A in S4 Text).
+As there is little data on dog population size and human exposure incidence in Madagascar [@andriamandimbySurveillanceControlRabies2013; @ratsitorahinaDogEcologyDemography2009], the range we used encompasses both observed human-to-dog ratios across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and generates similar exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008].
 Given previously high observed compliance in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], we assume that all rabies exposed patients who report to a clinic receive and complete PEP, and PEP is completely effective at preventing rabies.
@@
 Specifically, we aggregated our model-predicted estimates of annual bites to the clinic level.
-As multiple clinics may serve a single district or commune, we allocated bites to clinics according to the proportion of the population in each administrative unit which were closest.
+As multiple clinics may serve a single district or commune, we allocated bites to clinics according to the proportion of the population in each administrative unit which were closest to that clinic.
 For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e.
@@
 
-We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from a existing PEP clinic that provisions PEP for which travel times were reduced; 2) This proportion (1), weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing PEP clinic.
-We simulated expansion of clinics provisioning PEP to each district using these three metrics and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts for the full expansion of PEP.
+We tested three metrics for ranking additional clinics: 1) The proportion of people living >3 hours from an existing clinic that provisions PEP for which travel times were reduced; 2) This proportion weighted by the magnitude of the change in travel times and 3) The mean reduction in travel times for people living >3 hours from an existing PEP clinic.
+We simulated expansion of clinics provisioning PEP to each district using these three metrics and chose the metric which decreased burden the most compared to simulations (N = 10) where clinics were added randomly to districts.
 For the full simulation of expanded access, once clinics reduced travel times for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` living greater than $x$ hrs away, starting with $x$ = 3 hrs), we reduced the travel time threshold by 25%.
@@
 ## Sensitivity analysis {#methods-sensitivity}
-To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analysis of both parameters from the models of bite incidence and the decision tree (see Table S6.1 & S6.2 for parameter ranges used).
+To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analysis of both parameters from the models of bite incidence and the decision tree (see Fig A in S6 Text for parameter ranges used).
 We also examined how systematic variation in rabies incidence with human population size affected burden estimates.
@@
 dog ownership is more common in less populated, rural areas), we might expect exposure incidence to scale negatively with population size.
-We use scaling factors to scale incidence either positively or negatively with observed population sizes at the district and commune levels, while constraining them to the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, Fig S4.2) and simulated baseline burden, as well as expanded PEP access.
+We use scaling factors to scale incidence either positively or negatively with observed population sizes at the district and commune levels, while constraining them to the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, Fig B in S4 Text), and then simulate baseline burden, as well as expanded PEP access.
 
 ## Data and analyses 
-All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (Supplementary appendix, section S7).
+All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (S7 Text).
 All processed data, code, and outputs are archived on Zenodo (http://doi.org/10.5281/zenodo.4064312 and https://doi.org/10.5281/zenodo.4064304),  and maintained at https://github.com/mrajeev08/MadaAccess.
@@
 
-## Ethics statement
-Data collection from the Moramanga District was approved by the Princeton University IRB (7801) and the Madagascar Ministry of Public Health Ethics Committee (105-MSANP/CE).
-Oral informed consent was obtained from all interviewed participants.
-Data collected from bite patients at the national level are maintained jointly by the Ministry of Health and IPM as a routine part of PEP provisioning.
-
 # Results
@@
 However, we found that these estimates underestimated both driving times across the country and patient-reported travel times to the Moramanga PEP clinic (Fig 2C).
-Patient reported travel times were highly variable for a given commune compared to the estimated travel times (Fig S1.2), potentially due to the fact that the friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
-the presence of a road indicates that all travel through that grid cell is by vehicle).
-However, patients reported using multiple modes of transport, with some individuals walking days to the Moramanga PEP clinic (Fig S1.3).
-
-While the travel time estimates may not reflect exact distributions of travel times, they were correlated with ground-truthed driving and patient-reported times and likely reflect patterns of access over the country (Fig 3C, Fig S1.4).
-Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (R^2^ = `r round(ttime_fit$communes[ttime_fit$metric == "Weighted travel times (hrs)"], 2)`, Table S1.1), therefore, we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
+Patient reported travel times were highly variable for a given commune compared to the estimated travel times (Fig B in S1 Text).
+This is likely due to the fact that the friction surface assumes that the fastest available mode of transport is used across each grid cell (i.e.
+the presence of a road indicates that all travel through that grid cell is by vehicle), while patients reported using multiple modes of transport, with some individuals walking days to the Moramanga PEP clinic (Fig C in S1 Text).
+
+While the travel time estimates may not reflect exact distributions of travel times, they were correlated with ground-truthed driving and patient-reported times and likely reflect patterns of access over the country (Fig 3C, Fig D in S1 Text).
+Travel times weighted by population at the grid cell level were a better predictor than unweighted travel times or distance (R^2^ = `r round(ttime_fit$communes[ttime_fit$metric == "Weighted travel times (hrs)"], 2)`, Table A in S1 Text), therefore, we use population-weighted travel time as a proxy for access at the commune/district level in subsequent analyses.
 
@@
 ### *Fig 2. Travel times to clinics provisioning PEP across Madagascar.*
-*(A) Estimated at an ~ 1 km2 scale.
+*(A) ) Estimated at an ~ 1 km2 scale using the MAP friction surface.
+.
 (B) Distribution of the population across travel times.
@@
 Bite incidence estimates generally increased with decreasing weighted travel times at both administrative scales (district and commune), although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D).
-After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D, see Supplementary Appendix section S2 for more details).
+After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D, see S2 Text for more details).
 For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times.
@@
 
-Our modeling results show that travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales with the best fitting models including travel times and an overdispersion parameter (Fig 4, see Supplementary Appendix section S3 for comparisons to models with catchment effects and with population size as a covariate).
-As the predictions of the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
-Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig S3.3), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
-Finally, we examined the sensitivity of models to how we corrected for underreporting of data and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Fig S3.8 and Fig S3.9).
+Our modeling results show that travel times were a strong and consistent predictor of reported bite incidence in both datasets and across scales with the best fitting models including travel times and an overdispersion parameter (Fig 4, see S3 Text for comparisons to models with catchment effects and with population size as a covariate).
+As the predictions from the model fit to the Moramanga data without accounting for overdispersion fall within the prediction intervals for the models fit to the national data (Fig 4A), for subsequent predictions, we used the parameter estimates from models fit to the national data, which encompass the range of travel time effects observed in our datasets.
+Moreover, our out-of-fit predictions to the data across scales suggest that the commune model is able to capture travel time impacts at the commune level (Fig C in S3 Text), therefore we use both the district and commune model to disaggregate burden to the finest scale possible.
+Finally, we examined the sensitivity of models to how we corrected for underreporting of data and found that parameter estimates of travel time impacts were similar across models and performed similarly in prediction (Figs H and I in S3 Text).
 
@@
 In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C).
-We estimate that under the current system of 31 clinics in Madagascar provisioning PEP that approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies are prevented through PEP each year.
+We estimate that approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies are prevented through PEP each year under the current system of 31 PEP clinics in Madagascar .
 Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar.
@@
 
-kable(table, digits = -1, col.names = c("**Model**", "**Reported bite incidence per 100k**", "**Burden of deaths**", 
+knitr::kable(table, digits = -1, 
+             col.names = c("**Model**", "**Reported bite incidence per 100k**", "**Burden of deaths**", 
                                         "**Deaths averted by current PEP provisioning**"))           
@@
 *(A) for each district (y-axis) in Madagascar.
-Diamonds show the predicted incidence for the district model and squares show predicted incidence for the commune model fit to the National data for all communes in a district.
+Diamonds show the predicted incidence for the district model and squares show predicted incidence for the commune model fit to the National data for all communes in a given district.
 Points are colored and districts ordered by travel times.
@@
 
-We found that targeted expansion of PEP to clinics based on the proportion of the population they reduced travel times for resulted in fewest deaths (Fig S5.1).
-Here we report results from the commune model, as estimates were consistent across models (Fig 6 and Supplementary appendix, section S5).
+We found that targeted expansion of PEP to clinics based on the proportion of the population they reduced travel times for resulted in fewest deaths (Fig A in S5 Text).
+Here we report results from the commune model, as estimates were consistent across models (Fig 6 and S5 Text).
 We estimated that strategic PEP expansion to these additional 83 clinics (1 per district) reduced rabies deaths by `r round((1 - props$deaths_mean[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_dist & props$scale == "Commune"])*100, 0)`%) (Fig 6A).
 With one clinic per commune (where available, N = 1696), we see a further reduction of `r round((1 - props$deaths_mean[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`% (95% PI: `r round((1 - props$deaths_upper[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)` - `r round((1 - props$deaths_lower[props$scenario == one_per_comm & props$scale == "Commune"])*100, 0)`%).
-However, reductions in burden saturate as more clinics are added (Fig S5.2).
-Even when all primary clinics provision PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig S5.5); as more clinics are added, reported bite incidence saturates (Fig S5.4), and patients shift which clinic they report to (S5.7 & S5.8).
+However, reductions in burden saturate as more clinics are added (Fig B in S5 Text).
+Even when all primary clinics provision PEP, our model still predicts `r round(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` (95% PI:  `r round(natl_preds$deaths_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)` - `r round(natl_preds$deaths_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == "Commune"], -2)`)  deaths per annum, and average reporting of rabies exposures remains approximately `r round(trends$reporting_mean[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`% (95% PI: `r round(trends$reporting_upper[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)` - `r round(trends$reporting_lower[trends$scenario == max(trends$scenario) & trends$scale == "Commune"]*100, 0)`%) (Fig E in S5 Text).
+Our model predicts that as more clinics are added, reported bite incidence saturates (Fig D in S5 Text), and patients shift which clinic they report to (Figs G & H in S5 Text).
 
@@
 visits on days 0, 3, 7), to ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == 250 & natl_preds$scale == 'Commune'], -2)})")`  with 250 clinics providing PEP, and ~`r glue("{round(natl_preds$vials_mean[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} vials (95% PI: {round(natl_preds$vials_lower[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)} -  {round(natl_preds$vials_upper[natl_preds$scenario == max(natl_preds$scenario) & natl_preds$scale == 'Commune'], -2)})")` if all primary clinics provision PEP.
-In these scenarios, clinic catchment populations and throughput decrease, with clinics seeing fewer patients per day (S5.6).
+In these scenarios, clinic catchment populations and throughput decrease, with clinics seeing fewer patients per day (Fig F in S5 Text).
 
@@
 ### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
-*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional clinics provisioning PEP are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bites.
+*(A) Decrease in deaths due to rabies, (B) increase in totalnumberof vials as additional clinics provisioning PEP are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bite incidence.
 Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals.
-The points show when all additional primary clinics and secondary clinics (N = `r max_added`) clinics have been added).*
+The points show the scenario in which all additional primary clinics and secondary clinics (N = `r max_added`) clinics have been added).*
 
 ## Burden estimates are most sensitive to assumptions of underlying rabies incidence. {#results-sensitivity}
-While qualitative patterns of the current impact of geographic access on human rabies deaths and the impact of expanding access to PEP on reducing these deaths is robust across a wide range of parameter estimates, our sensitivity analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the most uncertainty to our quantitative estimates (Fig S6.1 & 2).
+While qualitative patterns of the current impact of geographic access on human rabies deaths and the impact of expanding access to PEP on reducing these deaths is robust across a wide range of parameter estimates, our sensitivity analyses show that assumptions of the underlying rabies exposure incidence ($E_i$) contribute the most uncertainty to our quantitative estimates (Fig A and B in S6 Text).
 Uncertainty in bite model parameters contribute less to uncertainty in estimates of burden or impacts of expanded access.
 For the estimates of vial demand, uncertainty around the model intercept (i.e.
-the baseline reported bite incidence) has most impact, rather than the travel time effect or the overdispersion parameter (Fig S6.3).
-Finally, scaling of incidence with population size (Fig S4.2) modulates the impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
-more rabies in more populated places) dampening and negative scaling exacerbating the relationship between access and deaths (Fig S6.4A).
-However, the impact of adding clinics remains broadly the same (Fig S6.4B).
+the average reported bite incidence) has most impact, rather than the travel time effect or the overdispersion parameter (Fig C in S6 Text).
+Finally, scaling of incidence with population size (Fig B in S4 Text) modulates the impact of travel times on deaths, with positive scaling of rabies incidence with population size (i.e.
+more rabies in more populated places) dampening and negative scaling exacerbating the relationship between access and deaths (Fig D in S6 Text).
+However, the impact of adding clinics remains broadly the same.
 
@@
 However, improved PEP provisioning is unlikely to eliminate rabies deaths; with over `r glue("{round(min(natl_preds$deaths_mean[natl_preds$scenario == max(natl_preds$scenario)]), -2)}")` deaths expected even with PEP at all primary clinics (N = `r max(natl_preds$scenario)`).
-This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig S5.4) even after expanding PEP to all primary clinics.
-With reduced travel times, over 20% of exposures will still not seek PEP (Fig S5.5), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
+This is partly because travel times remain high (> 2 hrs as estimated by the friction surface for over 10% of the population, Fig D in S5 Text) even after expanding PEP to all primary clinics.
+With reduced travel times, over 20% of exposures will still not seek PEP (Fig E in S5 Text), resulting in ~`r round(min(admin_base$deaths_mean/admin_base$pop*1e5), 2)` rabies deaths per 100,000 people.
 PEP is expected to remain cost-effective as provisioning expands, to a maximum of 450 USD per death averted (assuming 5 USD per vial), similar to other estimates [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
@@
 Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018].
-Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in the captial of Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
+Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in the capital of Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].
 We also assume that clinics reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].
 
-We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014; castillo2020behavioral] all play a role.
+We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.
 For example, most PEP clinics also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc.
@@
 His family urged him to go, but he did not believe his risk was high and decided not to seek further care.
-He developed symptoms two weeks later and was confirmed as a rabies death by the National Rabies Reference Laboratory.
+He developed symptoms two weeks later and was confirmed as a rabies death by the National Rabies Reference Laboratory at IPM.
 Despite prompt reporting, appropriate referral, and socioeconomic indicators suggesting a high care-seeking probability, this person did not receive PEP.
@@
 Our approach for incorporating access to vaccines (echoing [@Hegde2019dt; @macphersonDisparitiesAccessDiagnosis2019; @polettiHiddenBurdenMeasles2018; @metcalfTransportNetworksInequities2015; @aleganaTreatmentseekingBehaviourLow2017]) into burden estimation methods could guide provisioning of PEP to maximize impacts.
-This approach will have most value in settings with limited PEP access and poor health seeking, but will be less valuable where rabies exposures make up a small fraction of patients reporting for PEP e.g.
+This approach will have most value in settings with limited PEP access, but will be less valuable where rabies exposures make up a small fraction of patients reporting for PEP e.g.
 [@rysavaPathRabiesElimination2018; @wallaceEstablishmentCanineRabies2015].
@@
 
-# Supplementary Appendix
-
-## S1. Estimating travel times to the closest clinic provisioning PEP {#S1}
+## S1 Text. Estimating travel times to the closest clinic provisioning PEP
 
@@
 
-### *Fig S1.1. Raster inputs to estimate travel times to the closest clinic provisioning PEP for Madagascar.*
+### *Fig A. Raster inputs to estimate travel times to the closest clinic provisioning PEP for Madagascar.*
 _(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current clinics provisioning PEP (N = 31) shown with black x's.
@@
 
-### *Fig S1.2. Raw patient reported and driving travel time data.*
+### *Fig B. Raw patient reported and driving travel time data.*
 _(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga PEP clinic (B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
@@
 
-### *Fig S1.3. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga PEP clinic*
+### *Fig C. Reported modes of transport used compared to reported travel times for patients reporting to the Moramanga PEP clinic*
 \newpage
@@
 
-### *Fig S1.4. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) vs. estimates from friction surface.*
+### *Fig D. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) vs. estimates from friction surface.*
 _Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga PEP clinic for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only) ._
@@
 
-### *Table S1.1. R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
+### *Table A. R^2^ values from linear models with estimated access metrics predicting either driving times or commune means of patient reported travel times.*
 ```{r table S1.1 lms table}
@@
 
-### *Fig S1.5. Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.*
-
-\newpage
-
-## S2. Estimating bite incidence
-
-Most patients from each district reported to their closest clinic provisioning PEP by the weighted travel time metric (Fig 3).
+### *Fig E. Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.*
+
+\newpage
+
+<!-- split -->
+
+## S2 Text. Estimating bite incidence
+
+Most patients from each district reported to their closest clinic provisioning PEP by the weighted travel time metric (main text Fig 3).
 Accordingly, we assigned catchments based on which clinic was the closest for the majority of the population.
-While there are discrepancies between commune and district catchment assignments (Fig S2.1A), over 75% of the population in a given district or commune were closest to a single clinic (Fig S2.1C).
-We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms.
+While there are discrepancies between commune and district catchment assignments (Fig A.1), over 75% of the population in a given district or commune were closest to a single clinic (Fig A.3).
+We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, main text Fig 3A grey polygons) and corrected for periods where clinics did not submit any forms.
 
@@
 4 doses.
-As clinic submission of forms was highly variable from 2014-2017 (Fig S2.2A), we compared estimated demand to the total vials provisioned across this four year period comparing different thresholds for correcting for periods of no form submission (i.e, designating periods of 1, 5, 10, 15, and 30 consecutive days with zero submitted records as missing, compared to no correction).
-
-Estimates of vial demand based on uncorrected bite patient numbers were generally lower than the number of vials provisioned for those clinics with substantial under-submission of forms (Fig S2.2B).
-Correcting patient numbers for under-submission resulted in estimates of vial demand closer to the provisioned vials for most clinics  (Fig S2.2C, Table S2.1).
+As clinic submission of forms was highly variable from 2014-2017 (Fig B.1), we compared estimated demand to the total vials provisioned across this four year period comparing different thresholds for correcting for periods of no form submission (i.e, designating periods of 1, 5, 10, 15, and 30 consecutive days with zero submitted records as missing, compared to no correction).
+
+Estimates of vial demand based on uncorrected bite patient numbers were generally lower than the number of vials provisioned for those clinics with substantial under-submission of forms (Fig B.2).
+Correcting patient numbers for under-submission resulted in estimates of vial demand closer to the provisioned vials for most clinics  (Fig B.3, Table A).
 
@@
 
-### *Fig S2.1. Catchment assignments by travel time.*
+### *Fig A. Catchment assignments by travel time.*
 *(A) Catchments as assigned by closest clinic for the majority of the population within a commune (polygon fill) or within a district (polygon outline).
@@
 
-### *Fig S2.2. Estimating under-submission of patient forms.*
+### *Fig B. Estimating under-submission of patient forms.*
 *(A) Number of patient forms submitted to IPM for each clinic over the study period for each clinic, with periods of time where no forms were submitted for >= 15 days excluded (in grey); (B) Estimates for the proportion of forms submitted for each clinic (points are the estimate for each year and the line is the range), calculated as the # of days in a year which were not excluded based on the criteria of 15 consecutive days of non-submission/365.
@@
 
-### *Table S2.1.
+### *Table A.
 Root mean squared error (MSE) between observed vials provisioned and estimated by the different consecutive day threshold for correcting for periods of no form submission, with the minimum root MSE in bold.*
@@
 
-## S3. Modeling reported bite incidence
+<!-- split -->
+
+## S3 Text. Modeling reported bite incidence
 We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given dataset with a standard deviation of 10.
@@
 For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 5)$).
-We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as the maximum potential scale reduction factors (psrf) for each covariate and the multivariate psrf for the whole model (both are metrics of model convergence, where values < 1.1 are indicative of convergence, Table S3.1) [@gelman2006data].
+We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as the maximum potential scale reduction factors (psrf) for each covariate and the multivariate psrf for the whole model (both are metrics of model convergence, where values < 1.1 are indicative of convergence, Table A) [@gelman2006data].
 
@@
 
-For the national data, including a catchment random effect improved predictions (Fig S3.2 & Fig S3.3).
-However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table S3.1) and the models resulted in similar predictions (Fig S3.6 & S3.7), indicating that catchment effects could not be differentiated from random variation in the data.
-Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig S3.4), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig S3.5).
-Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3).
+For the national data, including a catchment random effect improved predictions (Fig B & Fig C).
+However, after accounting for overdispersion, catchment effects were not clearly identifiable (Table A) and the models resulted in similar predictions (Fig F & G), indicating that catchment effects could not be differentiated from random variation in the data.
+Similarly, while the commune model fit to the Moramanga data generated stronger travel time effects (Fig 4B), after accounting for data overdispersion, the posterior estimates of the parameters overlapped for the commune and district models fit to the national data (Fig D), and the model estimates were in general less robust to overdispersion than for the national data, particularly at low travel times (Fig E).
+Population size alone was the poorest fit to the data as estimated by DIC (Table A), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs B and C).
 
@@
 
-### *Fig S3.1. Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale.*
-
-\newpage
-
-### *Table S3.1. DIC and convergence estimates (maximum potential scale reduction factor and multivariate psrf, values < 1.1 indicate convergence) for all models.*
+### *Fig A. Correlation between travel time in hours (the average weighted by the population) and population size of administrative units at the district and commune scale.*
+
+\newpage
+
+### *Table A. DIC and convergence estimates (maximum potential scale reduction factor and multivariate psrf, values < 1.1 indicate convergence) for all models.*
 *For the column pop effect, addPop = models with population size as additional covariate, onlyPop = models with population as only covariate, flatPop = models with population as offset in model.
@@
 
-knitr::kable(select(dic_ranks, -min), 
+kable(select(dic_ranks, -min), 
              col.names = c("Dataset", "Scale", "Pop effect", "Intercept type",
@@
 
-### *Fig S3.2. Prediction to data used to fit each model.*
+### *Fig B. Prediction to data used to fit each model.*
 *Log of the observed bites against the log of predicted bites generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the points the mean of the predictions and the linerange the 95% prediction intervals.
@@
 
-### *Fig S3.3. Out of fit predictions to data.*
+### *Fig C. Out of fit predictions to data.*
 *Log of the observed bites against the log of predicted bites for data not used to fit the model.
@@
 
-### *Fig S3.4. Posterior estimates of parameters from models with travel time and population as an offset.*
+### *Fig D. Posterior estimates of parameters from models with travel time and population as an offset.*
 *Comparing models accounting for overdispersion ($\sigma_e$) compared to models with no overdispersion parameter  (flatPop in Figs S3.2 & S3.3).
@@
 
-### *Fig S3.5. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.* 
+### *Fig E. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.* 
 *Generated from sampling 1000 independent draws from the posterior distributions for each parameter, with the line the mean of the predictions and the envelopes showing the 95% prediction intervals.
@@
 
-### *Fig S3.6. Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter.*
+### *Fig F. Posterior estimates of the catchment intercepts ($\alpha$ parameters, with $\beta_0$ the estimated mean intercept) for models with and without an overdispersion parameter.*
 
@@
 
-### *Fig S3.7. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.*
+### *Fig G. Predicted relationship between travel times (in hours) and reported bite incidence per 100,000 persons.*
 *For random intercept model without overdispersion vs.
@@
 
-### *Fig S3.8. Posterior estimates for models with population as an offset and an overdispersion parameter.*
+### *Fig H. Posterior estimates for models with population as an offset and an overdispersion parameter.*
 *Columns show estimates from models fitted to the national dataset (1) corrected for both under-submission by correcting for periods of at least 7 days where zero patient forms were submitted, (2) correcting for periods of 15 days where zero patient forms were submitted, and (3) with the raw data not correcting for under-submission.*
@@
 
-### *Fig S3.9. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).*
+### *Fig I. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis).*
 *For models with population as an offset and an overdispersion parameter.
@@
 
-## S4. Range of rabies exposure incidence in people
+<!-- split -->
+
+## S4 Text. Range of rabies exposure incidence in people
 
 ```{r fig S4.1 exp range}
-include_graphics(here("analysis/figs/supplementary/S4.1_exprange.jpeg"))
-```
-
-### *Fig S4.1. Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis).* 
+include_graphics(here("analysis/figs/supplementary/S4_1_exprange.jpeg"))
+```
+
+### *Fig A. Estimated exposures per 100,000 given a range of human to dog ratios (HDRs, x-axis) and annual dog rabies incidence (y axis).* 
 *Assuming that each dog on average exposes 0.39 persons [@hampsonSurveillanceEstablishElimination2016].
@@
 ```{r fig S4.2 scaling factors}
-include_graphics(here("analysis/figs/supplementary/S4.2_scaling_rels.jpeg"))
-```
-
-### *Fig S4.2. Range of constrained scaling factors generated for district and commune population size* 
+include_graphics(here("analysis/figs/supplementary/S4_2_scaling_rels.jpeg"))
+```
+
+### *Fig B. Range of constrained scaling factors generated for district and commune population size* 
 *Underlying rabies exposures either (A) decreases with increasing population size or (B) increases with increasing population size across a fixed range of exposure incidence (15.6 - 76 exposures/100k persons).
@@
 
-## S5. Estimating the impact of expanding PEP provisioning to additional clinics {#S5}
+<!-- split -->
+
+## S5 Text. Estimating the impact of expanding PEP provisioning to additional clinics
 
@@
 
-### *Fig S5.1. Comparing metrics for ranking clinics for targeted expansion.*  
+### *Fig A. Comparing metrics for ranking clinics for targeted expansion.*  
 *We simulated expansion using three different ranking metrics: 1) reduction in mean travel times (green line) 2) the proportion of the population for which travel times were reduced (red dashed line) and 3) the proportion of the population for which travel times were reduced weighted by the reduction in travel times (pink dashed line).
@@
 
-### *Fig S5.2. Map of the location and at what step each clinic was added.*
+### *Fig B. Map of the location and at what step each clinic was added.*
 *The circles are each of the primary clinics across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
@@
 
-### *Fig S5.3. Maps of how travel times change as clinics are added.*
+### *Fig C. Maps of how travel times change as clinics are added.*
 *(A) at the ~1 x 1 km grid cell (B) commune and (C) district scales.
@@
 
-### *Fig S5.4. Shifts in key metrics as clinics are added.* 
+### *Fig D. Shifts in key metrics as clinics are added.* 
 *(A) travel times (hrs, x-axis is square root transformed), (B) bite incidence per 100,000 persons and boxplots showing the median for communes and district models (colors).*
@@
 
-### *Fig S5.5. Shifts in  key metrics as clinics are added.*
+### *Fig E. Shifts in  key metrics as clinics are added.*
 *(A) reporting, (B) death incidence per 100,000 persons for the commune and district models (colors) as clinics are added.*
@@
 
-### *Fig S5.6. Shifts in key metrics as clinics are added.*
+### *Fig F. Shifts in key metrics as clinics are added.*
 *(A) catchment population size, (B) annual vial demand, and (C) daily throughput (i.e.
@@
 
-### *Fig S5.7. Shifts in where bites are reported to as clinics are added for the commune model.*
+### *Fig G. Shifts in where bites are reported to as clinics are added for the commune model.*
 *The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
@@
 
-### *Fig S5.8. Shifts in where bites are reported to as clinics are added for the district model.*
+### *Fig H. Shifts in where bites are reported to as clinics are added for the district model.*
 *The circles show the clinic locations for each scenario, with size proportional to the annual average bites reported to that clinic.
@@
 
-## S6. Sensitivity of burden estimates to parameter assumptions {#S6}
+<!-- split -->
+
+## S6 Text. Sensitivity of burden estimates to parameter assumptions
 
@@
 
-### *Fig S6.1. Sensitivity analyses for baseline burden estimates.*
-*(A) Range of parameters used in the univariate sensitivity analysis (upper 97.5% and lower 2.5% credible interval of posterior for the parameters in the bite incidence model, 95% of CI from 95% CI of estimate from [@changaluchaNeedImproveAccess2018] for $\rho_max$, and fixed at the minimum and maximum of the range used in the main analysis for all other parameters) (B) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (C) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
+### *Fig A. Sensitivity analyses for baseline burden estimates.*
+*(A) Range of parameters used in the univariate sensitivity analysis (upper 97.5% and lower 2.5% credible interval of posterior for the parameters in the bite incidence model, 95% of CI from 95% CI of estimate from [@changaluchaNeedImproveAccess2018] for $\rho_max$, and fixed at the minimum and maximum of the range used in the main analysis for all other parameters) (B) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (C) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).*
 
@@
 
-### *Fig S6.2. Sensitivity analysis for PEP expansion.*
+### *Fig B. Sensitivity analysis for PEP expansion.*
 *(A) Estimates of the maximum proportion reduction of deaths from the baseline for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how human rabies decreases proportional to the baseline as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors).
-See Fig S6.1A for ranges used for each parameter.*
+See Fig A for ranges used for each parameter.*
 
@@
 
-### *Fig S6.3. Sensitivity analysis for vial demand.*
+### *Fig C. Sensitivity analysis for vial demand.*
 *(A) Estimates of the baseline vial demand for where each parameter was set to the upper and lower estimat of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (B) estimates for how vial demand increases as clinics are added for these same parameter ranges (line shows the mean estimate presented in the main analyses and the envelope shows the mean estimates for the parameters fixed at the upper and lower range) for the two models (colors) for the two models (colors).
-See Table S6.1 & 2 for ranges used for each parameter.*
+See Fig A for ranges used for each parameter.*
 \newpage
@@
 
-### *Fig S6.4. Sensitivity analysis for exposure scaling with population.*
+### *Fig D. Sensitivity analysis for exposure scaling with population.*
 *Predicted relationship between (A) travel times and rabies death incidence per 100k persons and (B) the proportional reduction in deaths from the baseline as clinics are added for the district model vs.
@@
 
-## S7. Sotware citations {#srefs}
+<!-- split -->
+
+## S7 Text. Additional citations
+
+### Software references
 ::: {#refs_supp}
@@
 
-## S8. Additional supplementary references {#srefextra}
+### Additional supplementary references {#srefextra}
 ::: {#refs_supextra}

### 2021-03-08 analysis/paper/subfiles/cover_letter.Rmd
fixing rendering for final proof submission

@@
 author: ""
-date: "10/2/2020"
+date: "20/02/2020"
 output: 
@@
 
-We are submitting our manuscript entitled “How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar” for consideration for publication in PLoS Neglected Tropical Disease.
-Our article presents a modeling framework to estimate how geographic access to post-exposure prophylaxis (PEP) shapes the burden of human rabies deaths.
-
-We use data from clinics across Madagascar and at multiple scales and travel time estimates to clinics provisioning PEP to fit models of reported bite incidence.
-We then use the predictions these models and integrate them into a decision tree framework to extrapolate rabies burden and vial demand.
-Finally, we examine how expanding access strategically can reduce human rabies burden.
-This work is inline with a body of work that shows the impacts of geography on access to care and disease outcomes.
-It also provides a modeling framework that could be usefully applied to other countries where PEP access is currently geographically restricted in how to effectively increase expansion, particularly in light of the recent addition of human rabies vaccine to the Gavi investment portfolio.
+We are submitting our manuscript entitled “Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District” for consideration for publication in Tropical Medicine and Infectious Disease (particularly for consideration for the Special Issue on One Health and NTDs, per communication between editors and co-author K.
+Hampson).
+Our article presents results from two pilot dog rabies vaccination campaigns in the Moramanga District and integrates the data we collected during the campaign into quantitative analyses of costs, coverage, and potential future vaccination strategies.
+We feel that this work will be very useful for not only Madagascar, but sub-Saharan African countries in general hoping to scale up dog vaccination, and to one health practioners more broadly in thinking about how best to translate strategies for public health implementation across human and animal sectors.
 

### 2021-03-08 R/get_d3_data.R

@@ -13,6 +13,9 @@ library(readr)
 library(ggplot2)
 library(raster)
 
+test <- st_read("analysis/out/access-story/world.geojson")
+plot(test, max.plot = 1)
+
 # out dir
 fp <- function(x) here::here("analysis/out/access-story", x)
 
@@ -27,9 +30,22 @@ garc_data %>%
 # ctar points ----
 ctar_pts <- read_csv("data-raw/out/clinics/ctar_metadata.csv")
 ctar_pts %>%
-  select(CTAR, color, distcode, lat, long) %>%
+  dplyr::select(CTAR, color, distcode, lat, long) %>%
   st_as_sf(coords = c("long", "lat")) %>%
-  st_write(fp("ctar_pts.geojson"))
+  st_write(fp("ctar_pts.json"), layer_options = "NATIVE_DATA=YES",
+           append = FALSE)
+
+# bite data ----
+bites <- read_csv("analysis/out/bites/district_bites.csv")
+exps <- read_csv("analysis/out/mods/preds/expectations.csv")
+exps %>%
+  filter(interaction(data_source, intercept, OD) %in% "National.fixed.TRUE" &
+           scale %in% "Commune") %>%
+  write_json(fp("mod.json"))
+
+bites %>%
+  dplyr::select(distcode, ttimes = ttimes_wtd, catchment, avg_bites) %>%
+  write_json(fp("bites.json"))
 
 # circle packing data frame ----
 circle_pack <- bind_rows(data.frame(reported = rbinom(100, 1, 0.3), rabid = TRUE),

### 2021-03-08 R/split_render.R

@@ -1,47 +1,72 @@
 
 split_render <- function(manuscript = "analysis/paper/manuscript.Rmd",
-                         output_dir ="analysis/paper/subfiles") {
-  single <- readLines(manuscript)
-  split <- grep("<!-- split -->", single)
-  man <- single[1:(split-1)]
-  supplement <- single[(split-1):length(single)]
-
-  # take out yaml from first one
-  inds <- grep("---", man)
-  man <- man[-c(inds[1]:inds[2])]
-  man <- c(readLines(here::here("analysis/paper/format/main.yaml")),
-           man)
-
-  writeLines(man, here::here("analysis/paper/man.Rmd"))
-
-  # and level 1 header from supplement
-  inds <- grep("# Supplementary Appendix", supplement)
-  supplement <- supplement[-c(inds[1])]
-
-  # Also add the first set-up code chunk!
-  inds <- grep("```", single)
-  supplement <- c(single[inds[1]:inds[2]], supplement)
-  supplement <- c(readLines(here::here("analysis/paper/format/supplement.yaml")),
-                  supplement)
-
-  writeLines(supplement, here::here("analysis/paper/supplement.Rmd"))
+                         output_dir ="analysis/paper/subfiles",
+                         write_man = TRUE) {
 
   # and render both
   if(!dir.exists(here::here(output_dir))) {
     dir.create(here::here(output_dir))
   }
 
-  rmarkdown::render(here::here("analysis/paper/supplement.Rmd"),
-                    output_dir = here::here(output_dir))
-  rmarkdown::render(here::here("analysis/paper/man.Rmd"),
-                    output_dir = here::here(output_dir),
-                    params = list(nofig = TRUE))
-  rmarkdown::render(here::here("analysis/paper/supplement.Rmd"),
-                    output_file = "supplement_textonly",
-                    output_dir = here::here(output_dir),
-                    params = list(nofig = TRUE))
+  single <- readLines(manuscript)
+  split <- grep("<!-- split -->", single)
 
-  # clean up files
-  unlink(here::here("analysis/paper/supplement.Rmd"))
-  unlink(here::here("analysis/paper/man.Rmd"))
+  stitles <- single[grep("## S[0-9]", single)]
+  stitles <- gsub("#", "", stitles)
+
+  for(i in seq_len(length(split))) {
+
+    start <- split[i]
+    end <- ifelse(i == length(split), length(single), split[i + 1])
+    supplement <- single[start:end]
[…]

### 2021-03-08 analysis/paper/make_paper.R

@@ -2,26 +2,16 @@
 #' Compile paper for submission
 # ------------------------------------------------------------------------------
 
-start <- Sys.time()
 library(here)
 library(knitr)
 source("R/split_render.R")
 
-# For revisions
-rmarkdown::render(here("analysis/paper/minor_revision_1/response_reviewers.Rmd"),
-                  output_dir = here("analysis/paper/minor_revision_1"))
-
-# upload to google drive for feedback
-googledrive::drive_update('MadaAccess', here('analysis/paper/minor_revision_1/manuscript_supplement_tracked.docx'))
-googledrive::drive_put(here('analysis/paper/minor_revision_1/response_reviewers.docx'), 'MadaAccess_response')
-
-# For submitting
-split_render(manuscript = "analysis/paper/manuscript.Rmd",
-             output_dir ="analysis/paper/minor_revision_1")
+# For submitting final proofs (supplementary materials only!)
+split_render(output_dir = "analysis/paper/final_proof", write_man = TRUE)
 
 # Internal steps
 # Convert to pdf (using system & this applescript: https://superuser.com/a/1559043
 # superuser answer)
-system("rm -r analysis/paper/minor_revision_1/*.pdf") # clean first
-system("bash analysis/paper/minor_revision_1/word2pdf.sh")
+system("rm -r analysis/paper/final_proof/*.pdf") # clean first
+system("bash analysis/paper/final_proof/word2pdf.sh")
 

### 2021-04-02 analysis/paper/manuscript.Rmd
updating with final (hopefully) proof formatting

@@
 output:
+  bookdown::pdf_document2:
+    pandoc_args:
+      - --csl=format/plos-neglected-tropical-diseases.csl
+      - --lua-filter=format/multiple-bibliographies.lua
   bookdown::word_document2:
@@
 the minimum distance from the administrative unit centroid to any clinic.
-We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the district and communes boundaries (as of October 31, 2018).
+We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs (OCHA) available through the Humanitarian Data Exchange for the district and communes boundaries (retreived on October 31, 2018, https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries).
 To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e.
@@
 ```{r fig 1, out.width="100%"}
-include_graphics(here("data-raw/misc/figM1/figM1_final.tiff"))
+# include_graphics(here("data-raw/misc/figM1/figM1_final.tiff"))
 ```
@@
 ## Data and analyses 
-All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the supplementary references (S7 Text).
+All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using a number of additional packages (see S7 Text for details).
 All processed data, code, and outputs are archived on Zenodo (http://doi.org/10.5281/zenodo.4064312 and https://doi.org/10.5281/zenodo.4064304),  and maintained at https://github.com/mrajeev08/MadaAccess.
@@
 ### *Fig 2. Travel times to clinics provisioning PEP across Madagascar.*
-*(A) ) Estimated at an ~ 1 km2 scale using the MAP friction surface.
-.
+*(A) ) Estimated at an ~ 1 km2 scale using the estimated from the Malaria Atlas Project friction surface (https://malariaatlas.org/research-project/accessibility-to-cities/, CC-BY 3.0) and on human population from WorldPop (https://www.worldpop.org/geodata/summary?id=70, CC-BY 4.0 license).
 (B) Distribution of the population across travel times.
@@
 the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).
-The point lines indicate the range of estimated bite incidence for each district.*
+The point lines indicate the range of estimated bite incidence for each district.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO)*
 
@@
 The vertical lines show the average national incidence of human rabies deaths for the commune (grey) and district (black) models.
-Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current clinics provisioning PEP*
+Incidence mapped to the (B) commune- and (C) district-level from the respective models; grey X’s show locations of current clinics provisioning PEP.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO)*
 
@@
 <!-- split -->
-
 \newpage
@@
 ### *Fig A. Raster inputs to estimate travel times to the closest clinic provisioning PEP for Madagascar.*
-_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current clinics provisioning PEP (N = 31) shown with black x's.
-(B) Population estimates resampled to the same friction surface._
+_(A) Friction surface of travel speeds (in minutes per meter) at an ~ 1x1 km scale, with location of current clinics provisioning PEP (N = 31) shown with black x's (original source: Malaria Atlas Project friction surface, (https://malariaatlas.org/research-project/accessibility-to-cities/, CC-BY 3.0).
+(B) Population estimates resampled to the same friction surface (original source: WorldPop, https://www.worldpop.org/geodata/summary?id=70, CC-BY 4.0 license)._
 
@@
 _(A) Distribution of travel times estimated at the grid cell level and reported by patients for each commune where patient data were available from the Moramanga PEP clinic (B) Reported driving times between locations, with the color corresponding to the total driving time and the size of the line showing the direction of travel (narrow to wide ~ origin to destination).
-Paths are Bezier curves from origin to destination, and do not show actual paths driven._
+Paths are Bezier curves from origin to destination, and do not show actual paths driven.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO)._
 
@@
 ### *Fig D. Observed estimates of travel times (commune means of patient reported travel times and driving times between point locations) vs. estimates from friction surface.*
-_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga PEP clinic for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only) ._
+_Predicted by (A) Distance (km) (Euclidean distance between origin and destination for driving times and distance from the commune centroid to the Moramanga PEP clinic for commune means) (B) Travel time estimates and (C) Travel time estimates weighted by population (for commune means only).
+Administrative boundaries are from OCHA (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries,
+CC-BY-IGO)._
 
@@
 
-### *Fig E. Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.*
-
-\newpage
+### *Fig E.
+Estimates of mean travel times weighted by population at the (A) District (B) Commune scale.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 
@@
 The vertical line indicates the proportion of bites from within the assigned catchment for the Moramanga data (~90%).
-*
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 
@@
 ```
-\newpage
-
+
+### References
+::: {#refs_supextra}
+:::
 <!-- split -->
@@
 The points show the data used to fit the models.*
-\newpage
-
+
+### References
+::: {#refs_supextra}
+:::
 <!-- split -->
@@
 ```{r fig S4.1 exp range}
-include_graphics(here("analysis/figs/supplementary/S4_1_exprange.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S4.1_exprange.jpeg"))
 ```
@@
 The black dashed lines show the range of human to dog ratios (HDRs) we use in the main analysis to estimate the range of human exposure incidence (where the red horizontal line and black dashed lines intersect).
-The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@sambo2018estimating].
-The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].*
+The grey dashed lines show the HDRs estimated from the Moramanga district from a recent study [@leblancclaireRabiesMadagascarThreepronged2019].
+The cells with red outlines show the range of estimated exposure incidence from a previous study of bite patients in Moramanga District [@rajeevHealthcareUtilizationProvisioning2019].
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 \newpage
@@
 ```{r fig S4.2 scaling factors}
-include_graphics(here("analysis/figs/supplementary/S4_2_scaling_rels.jpeg"))
+include_graphics(here("analysis/figs/supplementary/S4.2_scaling_rels.jpeg"))
 ```
@@
 Lines show the expected relationship, with points showing where administrative units fall along this curve, and maps show how this results in variation in assumed exposure incidence spatially at the commune and district level.*
-\newpage
-
+
+### References
+::: {#refs_supextra}
+:::
 <!-- split -->
@@
 *The circles are each of the primary clinics across the country sized by the resulting average reduction in burden (based on smoothed annual burden estimates from the commune model, see inset).
-The large white crosses show the location of the existing 31 clinics provisioning PEP in Madagascar and the smaller white crosses are the additional primary clinics in the country which were added in the final step but not ranked.*
+The large white crosses show the location of the existing 31 clinics provisioning PEP in Madagascar and the smaller white crosses are the additional primary clinics in the country which were added in the final step but not ranked.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 \newpage
@@
 Grey pixels show the location of clinics provisioning PEP at each step.
-Commune and district values are the average grid cell travel times weighted by the population in each cell.*
+Commune and district values are the average grid cell travel times weighted by the population in each cell.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 \newpage
@@
 Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the commune level reported bite incidence.*
+The polygon shading shows the commune level reported bite incidence.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO).*
 \newpage
@@
 Lines show where the bites are reported from (commune centroid) also proportional to the number of bites.
-The polygon shading shows the district level reported bite incidence.*
-\newpage
+The polygon shading shows the district level reported bite incidence.
+Administrative boundaries from OCHA via HDX (https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries, CC-BY-IGO)*
 
@@
 ### *Fig A. Sensitivity analyses for baseline burden estimates.*
-*(A) Range of parameters used in the univariate sensitivity analysis (upper 97.5% and lower 2.5% credible interval of posterior for the parameters in the bite incidence model, 95% of CI from 95% CI of estimate from [@changaluchaNeedImproveAccess2018] for $\rho_max$, and fixed at the minimum and maximum of the range used in the main analysis for all other parameters) (B) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (C) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).*
+*(A) Range of parameters used in the univariate sensitivity analysis (upper 97.5% and lower 2.5% credible interval of posterior for the parameters in the bite incidence model, 95% CI of estimate from [@changaluchaNeedImproveAccess2018] for $\rho_{max}$, and fixed at the minimum and maximum of the range used in the main analysis for all other parameters) (B) Range of estimates of the estimate average deaths at the national level for range of parameter estimates (point shows the estimate presented in the main analyses and the ends show the upper and lower estimates for the parameter range) and (C) estimates for the predicted relationship between travel times and incidence of human rabies deaths across the same parameter ranges (line shows the estimate presented in the main analyses and the envelope shows the upper and lower estimates for the parameter range) for the two models (colors).*
 
@@
 The points show the mean estimates from 1000 simulations and the line ranges show the 95% prediction intervals.*
-\newpage
-
+
+### References
+::: {#refs_supextra}
+:::
 <!-- split -->
 
-## S7 Text. Additional citations
-
-### Software references
+## S7 Text. Details on software packages used
+```{r}
+tbl_bib <- bib2df::bib2df(here("analysis/paper/bib/refs_pkgs.bib"))
+
+tbl_bib %>%
+  filter(CATEGORY == "MANUAL") %>%
+  mutate(citekey = BIBTEXKEY, name = gsub("R-", "", BIBTEXKEY), 
+         .keep = "none") %>%
+  mutate(vals = glue("`{name}` [@{citekey}]")) %$% 
+  paste(vals, collapse = ", ") -> pkg_names
+
+```
+
+Our analyses relied on a number of open-source software packages for the `R` programming language [@RLang2019] including the following: `r pkg_names`.
+For full details on implementation please see the archived repositories on [Zenodo](http://doi.org/10.5281/zenodo.4064312) or on [Github](https://github.com/mrajeev08/MadaAccess).
+
+### References
 ::: {#refs_supp}
 :::
-
-\newpage
-
-### Additional supplementary references {#srefextra}
-::: {#refs_supextra}
-:::

### 2021-04-02 R/split_render.R

@@ -47,8 +47,8 @@ split_render <- function(manuscript = "analysis/paper/manuscript.Rmd",
 
     # handle the supplement but with (no fig = TRUE & toc = false)
     suppl <- single[split[1]:length(single)]
-    suppl <- suppl[-grep("newpage|:::|## S7 Text|### Software references|### Additional supplementary",
-                         suppl)]
+
+    # Filter out all text that's not a header!
     suppl <- c("## List of supplementary figures and tables", suppl)
 
     # take out yaml from first one

### 2021-04-02 analysis/paper/make_paper.R

@@ -7,11 +7,10 @@ library(knitr)
 source("R/split_render.R")
 
 # For submitting final proofs (supplementary materials only!)
-split_render(output_dir = "analysis/paper/final_proof", write_man = TRUE)
+split_render(output_dir = "analysis/paper/final_proof", write_man = FALSE)
 
 # Internal steps
 # Convert to pdf (using system & this applescript: https://superuser.com/a/1559043
 # superuser answer)
-system("rm -r analysis/paper/final_proof/*.pdf") # clean first
 system("bash analysis/paper/final_proof/word2pdf.sh")
 

### 2021-04-02 data-raw/src/06_doc_data.R

@@ -307,7 +307,7 @@ mada_districts %>%
     sources = list(
       list(
         title = "United Nations Office for the Coordination of Humanitarian Affairs (OCHA) Regional Office for Southern Africa",
-        path = "https://data.world/ocha-rosa/26fa506b-0727-4d9d-a590-d2abee21ee22"
+        path = "https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries"
       )
     )
   ) %>%
@@ -339,7 +339,7 @@ mada_communes %>%
     sources = list(
       list(
         title = "United Nations Office for the Coordination of Humanitarian Affairs (OCHA) Regional Office for Southern Africa",
-        path = "https://data.world/ocha-rosa/26fa506b-0727-4d9d-a590-d2abee21ee22"
+        path = "https://data.humdata.org/dataset/madagascar-administrative-level-0-4-boundaries"
       )
     )
   ) %>%
